### 2021-12-14 src/ms.tex
Prototype different chapters and sections.

\RequirePackage[l2tabu, orthodox]{nag} 
% Define document class
\documentclass[]{report}
\usepackage{style}

% Begin!
\begin{document}

% Title
\title{\vspace{-15pt}\begin{center}\includegraphics[width=0.4\linewidth]{static/misc/logo.jpg}\end{center}
\vspace{20pt}
\begin{center}
School of Physics \& Astronomy
\end{center}
\vspace{20pt}
\begin{center}
\LARGE\textbf{Probabilistic modeling of astrophysical time series:
gravitational microlensing and occultation mapping of planets and moons}
\end{center}
\vspace{20pt}
\begin{center}
Supervisor: Dr. Martin Dominik
\end{center}
}

\maketitle
\tableofcontents

% Abstract 
\begin{abstract}
    Here goes the abstract.
\end{abstract}

\clearpage
\thispagestyle{empty}

% CHAPTER 1: Introduction
\chapter{Introduction}
\section{Context}
\section{Gravitational microlensing}
\section{Occultation and phase curve mapping}

% CHAPTER 2: The theoretical minimum
\chapter{The theoretical minimum}

\section{Gravitational microlensing}
\subsection{Deflection of light by gravity}
\subsection{The different scales of lensing}
%% Read the "scales of lensing" paper by that Italian guy
\subsection{Microlensing}
\subsection{The point mass lens and the lens equation}
\subsection{Parallax}
\subsection{Finite source effects}
\subsection{The binary lens}
\subsection{Multiple lenses}
\subsection{Solving the lens equation}

\section{Occultation and phase curve mapping}
\subsection{The Starry framework}
\subsection{Mapping Solar System bodies}
\subsection{Mapping stars and exoplanets}
\subsection{The future of mapping exoplanets}

% In this and the following sections make sure to use microlensing and occultation
% mapping as examples
\section{Bayesian statistics}
\subsection{Probability theory}
\subsection{The meaning of probability}
\subsection{Probability as frequency of events in repeated trials}
\subsection{Probability as degrees of belief}
\subsection{The likelihood function}
% mention the likelihood principle
\subsection{Priors}
\subsection{On machine learning and the difference between explanation and prediction}
% mention that even if we have a model focused on inference, it might be best if we judge 
% its quality through its predictive performance

\section{Inference}
% check that paper on physics and ML by that MIT guy, there's a good stuff there. 
% check also Ian Murray's lecture notes  on MLPR
\subsection{The curse of dimensionality}
\subsection{There's no free lunch}
% exploration vs. explotation 
\subsection{Optimization vs. sampling}

\subsection{Sampling and optimization methods}
\subsubsection{Maximum likelihood}
\subsubsection{The Laplace approximation}
\subsubsection{Variational inference}
\subsubsection{Importance sampling}
\subsubsection{Rejection sampling}
\subsubsection{Markov Chain Monte Carlo}
\subsubsection{Hamiltonian Monte Carlo}
\subsubsection{Other notable methods}

\section{Model validation and comparison}
\subsection{Cross validation}
% discuss vanilla k-fold CV, stress out just how important and central CV is to science
\subsubsection{Pareto smoothed importance sampling LOO}
\subsection{Model comparison}
% don't do it!

\section{Putting it all together -- \emph{probabilistic programming}}
\subsection{The success of machine learning}
% that quote from Yann LeCun on differentiable code 
\subsection{Automatic differentiation}
\subsubsection{Forward mode autodiff}
\subsubsection{Reverse mode autodiff or backpropagation}
\subsubsection{Jacobian vector products}
\subsection{JAX}
\subsection{Numpyro}

% CHAPTER 3: Modeling microlensing events 
\chapter{Modeling microlensing events}
\section{Dealing with multimodal posteriors}
\section{A differentiable solver for the lens equation}
\section{Population level inference in microlensing}

% CHAPTER 4: Mapping the surface of Io 
\chapter{Mapping the surface of Io}
\section{The data}
\section{Information content}
\section{Pixel sampling}
\section{A static map model}
\section{A dynamic map model}
\section{Future work}
\subsection{Mapping Pluto and Charon}

% CHAPTER 5: Mapping the surfaces of exoplanets
\chapter{Mapping the surfaces of exoplanets}
\section{Previous work}
\section{What can we learn using eclipse mapping?}
\section{Time dependent maps}

% CHAPTER 6: Conclusion
\chapter{Conclusion}
\section{Summary of the thesis}
\subsection{Microlensing}
\subsection{Occultation mapping}
\section{Future work}
\subsection{Microlensing in the era of the Roman telescope}
\subsection{Mapping exoplanets with JWST and beyond}

\appendix

\chapter{First appendix}
\chapter{Second appendix}

\end{document}

### 2022-05-18 src/ms.tex
Update ToC
minor
add abstract

@@
 \begin{abstract}
-    Here goes the abstract.
+    Scientific progress in modern astronomy research commonly relies on gathering large quantities of
+    data using exceedingly precise instruments.
+    The process which ``generates'' these data consists of the physical phenomenon of
+    interest -- for instance, an exoplanet blocking or twisting the light of a distant star, and
+    the noise introduced by the measurement process and the presence of an atmosphere
+    The task for an astronomer is to first construct a \emph{model} which describes the entire process
+    which generated the data and to then ``fit'' that model to data.
+    All models only approximate reality and the researcher has to make a series of decisions during the
+    model building process, everything from how to process the raw data to which results to put in
+    the abstract of a paper.
+    Advancements in computational statistics and machine learning in the past decade or
+    so have made it possible to fit ever more complex models to data.
+    These models are generally expressed in computer code which may contain complex numerical algorithms,
+    such as iterative solvers and numerical integrals.
+    In this thesis, I mostly focused on developing methods which enable \emph{statistical inference}
+    with these kinds of complex models in two particular domains within astronomy, gravitational
+    microlensing and occultation mapping.
+    The common theme between these two topics is that both deal with accurately measuring
+    the brightness of distant stars as a function of time with the goal of inferring properties of
+    exoplanets and stars.
+    Broadly speaking, I believe the biggest contribution of this thesis is providing a new lens
+    for looking at a particular set of old problems, a lens which incorporates recent advancements
+    from statistics, machine learning and computer science.
+    More specifically, I have developed a an open-source software
+    package \textsf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which enables
+    fast and accurate computation of binary lens and triple lens microlensing light curves and
+    simultaneously provides exact \emph{gradients} of the code outputs with respect to all its inputs.
+    This is significant because it for the first time enables the use of modern gradient based
+    statistical inference algorithms such as Hamiltonian Monte Carlo with microlensing light curves.
+    Microlensing is one of the major goals for the upcoming \emph{NASA Roman} telescope and the existing
+    modeling methods are completely inadequate for dealing with the scale of data which will come
+    from \emph{Roman}.
+    I also propose a framework for dealing with issues which have plagued the field for
+    decades -- various pathologies in microlensing models and questions about the interpretation
+    of statistical results.
+    Besides microlensing, I have also delved into the field of occultation mapping of Solar System
+    objects and exoplanets.
+    Together with collaborators, I have developed  a novel statistical method for reconstructing
+    spatial maps of volcanic emission on Jupiter's moon Io using infra-red occultation
+    light curves.
+    I applied the same method to exoplanets to explore the exciting possibility of
+    detecting weather changes on Hot Jupiters by reconstructing two dimensional maps of
+    the emission surface from simulated \emph{JWST} secondary eclipse light curves.
+    I found that planetary scale changes in the emission pattern should be detectable with
+    \emph{JWST}.
+
 \end{abstract}
@@
 \section{Occultation and phase curve mapping}
+\subsection{History}
 \subsection{The Starry framework}
-\subsection{Mapping Solar System bodies}
-\subsection{Mapping stars and exoplanets}
-\subsection{The future of mapping exoplanets}
+\subsection{Reflected vs. emitted light}
 
@@
 \section{A dynamic map model}
-\section{Future work}
-\subsection{Mapping Pluto and Charon}
 
@@
 \subsection{Microlensing in the era of the Roman telescope}
+\subsection{Mapping volcanic activity on Io}
 \subsection{Mapping exoplanets with JWST and beyond}

### 2022-05-22 src/ms.tex
introduction chapter

@@
-\RequirePackage[l2tabu, orthodox]{nag}
 % Define document class
-\documentclass[]{report}
-\usepackage{style}
+\documentclass[11pt]{report}
+\usepackage{aas_macros}
+
+% Citations
+\usepackage{natbib}
+\bibliographystyle{abbrvnat}
+\setcitestyle{authoryear}
+
+% Language 
+\usepackage[utf8x]{inputenc}
+
+% Layout and margins 
+\usepackage[margin=1in]{geometry}
+\usepackage{layout}
+\usepackage{lastpage}
+\usepackage{titling}
+
+% Figures & tables 
+\usepackage{graphicx,xcolor}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{booktabs} % much better looking tables
+
+% Math 
+\usepackage{amsmath,amssymb,bm}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\usepackage{siunitx} % for proper units
+
+% Other packages
+\usepackage{url}
+\usepackage{epstopdf}
+\usepackage[font=scriptsize]{caption} % smaller font for fig captions
+\usepackage{placeins}
+%\usepackage{import}
+% of references
+
+% Settings
+\hfuzz=6pt % tolerance on hbox overflows
+
+% Custom command
+\newcommand{\ud}{\,\mathrm{d}}
+\renewcommand{\vec}[1]{\bm{#1}}
+
+\hypersetup{
+    colorlinks=true,
+    allcolors=cyan,
+}
 
@@
 % Abstract 
-\begin{abstract}
-    Scientific progress in modern astronomy research commonly relies on gathering large quantities of
-    data using exceedingly precise instruments.
-    The process which ``generates'' these data consists of the physical phenomenon of
-    interest -- for instance, an exoplanet blocking or twisting the light of a distant star, and
-    the noise introduced by the measurement process and the presence of an atmosphere
-    The task for an astronomer is to first construct a \emph{model} which describes the entire process
-    which generated the data and to then ``fit'' that model to data.
-    All models only approximate reality and the researcher has to make a series of decisions during the
-    model building process, everything from how to process the raw data to which results to put in
-    the abstract of a paper.
-    Advancements in computational statistics and machine learning in the past decade or
-    so have made it possible to fit ever more complex models to data.
-    These models are generally expressed in computer code which may contain complex numerical algorithms,
-    such as iterative solvers and numerical integrals.
-    In this thesis, I mostly focused on developing methods which enable \emph{statistical inference}
-    with these kinds of complex models in two particular domains within astronomy, gravitational
-    microlensing and occultation mapping.
-    The common theme between these two topics is that both deal with accurately measuring
-    the brightness of distant stars as a function of time with the goal of inferring properties of
-    exoplanets and stars.
-    Broadly speaking, I believe the biggest contribution of this thesis is providing a new lens
-    for looking at a particular set of old problems, a lens which incorporates recent advancements
-    from statistics, machine learning and computer science.
-    More specifically, I have developed a an open-source software
-    package \textsf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which enables
-    fast and accurate computation of binary lens and triple lens microlensing light curves and
-    simultaneously provides exact \emph{gradients} of the code outputs with respect to all its inputs.
-    This is significant because it for the first time enables the use of modern gradient based
-    statistical inference algorithms such as Hamiltonian Monte Carlo with microlensing light curves.
-    Microlensing is one of the major goals for the upcoming \emph{NASA Roman} telescope and the existing
-    modeling methods are completely inadequate for dealing with the scale of data which will come
-    from \emph{Roman}.
-    I also propose a framework for dealing with issues which have plagued the field for
-    decades -- various pathologies in microlensing models and questions about the interpretation
-    of statistical results.
-    Besides microlensing, I have also delved into the field of occultation mapping of Solar System
-    objects and exoplanets.
-    Together with collaborators, I have developed  a novel statistical method for reconstructing
-    spatial maps of volcanic emission on Jupiter's moon Io using infra-red occultation
-    light curves.
-    I applied the same method to exoplanets to explore the exciting possibility of
-    detecting weather changes on Hot Jupiters by reconstructing two dimensional maps of
-    the emission surface from simulated \emph{JWST} secondary eclipse light curves.
-    I found that planetary scale changes in the emission pattern should be detectable with
-    \emph{JWST}.
+\chapter*{Abstract}
 
-\end{abstract}
+Scientific progress in modern astronomy research commonly relies on gathering
+large quantities of data using exceedingly precise instruments. The process
+which ``generates'' these data consists of the physical phenomenon of interest
+-- for instance, an exoplanet blocking or twisting the light of a distant star,
+and the noise introduced by the measurement process and the presence of an
+atmosphere The task for an astronomer is to first construct a \emph{model}
+which describes the entire process which generated the data and to then ``fit''
+that model to data. All models only approximate reality and the researcher has
+to make a series of decisions during the model building process, everything
+from how to process the raw data to which results to put in the abstract of a
+paper. Advancements in computational statistics and machine learning in the
+past decade or so have made it possible to fit ever more complex models to
+data. These models are generally expressed in computer code which may contain
+complex numerical algorithms, such as iterative solvers and numerical
+integrals. In this thesis, I mostly focused on developing methods which enable
+\emph{statistical inference} with these kinds of complex models in two
+particular domains within astronomy, gravitational microlensing and occultation
+mapping. The common theme between these two topics is that both deal with
+accurately measuring the brightness of distant stars as a function of time with
+the goal of inferring properties of exoplanets and stars. Broadly speaking, I
+believe the biggest contribution of this thesis is providing a new lens for
+looking at a particular set of old problems, a lens which incorporates recent
+advancements from statistics, machine learning and computer science. More
+specifically, I have developed a an open-source software package
+\textsf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which
+enables fast and accurate computation of binary lens and triple lens
+microlensing light curves and simultaneously provides exact \emph{gradients} of
+the code outputs with respect to all its inputs. This is significant because it
+for the first time enables the use of modern gradient based statistical
+inference algorithms such as Hamiltonian Monte Carlo with microlensing light
+curves. Microlensing is one of the major goals for the upcoming \emph{NASA
+    Roman} telescope and the existing modeling methods are completely inadequate
+for dealing with the scale of data which will come from \emph{Roman}. I also
+propose a framework for dealing with issues which have plagued the field for
+decades -- various pathologies in microlensing models and questions about the
+interpretation of statistical results. Besides microlensing, I have also delved
+into the field of occultation mapping of Solar System objects and exoplanets.
+Together with collaborators, I have developed a novel statistical method for
+reconstructing spatial maps of volcanic emission on Jupiter's moon Io using
+infra-red occultation light curves. I applied the same method to exoplanets to
+explore the exciting possibility of detecting weather changes on Hot Jupiters
+by reconstructing two dimensional maps of the emission surface from simulated
+\emph{JWST} secondary eclipse light curves. I found that planetary scale
+changes in the emission pattern should be detectable with \emph{JWST}.
 
-\clearpage
+\newpage
 \thispagestyle{empty}
@@
 \section{Context}
+
+The key idea of the scientific revolution in the 16th century was to carefully
+observe the world, build \emph{models} which describe some aspect of the
+observed phenomenon, and finally and most importantly -- test those models to
+see if they provide an accurate description of reality. Since the time of
+Galileo, this process gave rise to the modern world and revolutionised our
+understanding of the universe. Today we have amazingly accurate models of the
+universe. Einstein's theory of \emph{General Relativity} which describes the
+universe at a large scale and the \emph{Standard Model} which describes the
+universe at the atomic and subatomic scales, both of which were developed in
+the 20th century. In my opinion, the two most important questions for the
+physical sciences in the 21st century and beyond are the following:
+\begin{itemize}
+    \item How do we combine the Standard Model and General Relativity to get a complete
+          physical theory of the Universe?
+    \item What is the origin and distribution of complex life in the Universe?
+\end{itemize}
+
+It used to be the case that conducting experiments with the potential to to
+change our understanding of fundamental physics was relatively straightforward
+and could be conducted by a single person or a small group of researchers. I
+have in mind, for instance, Kepler's observations of planetary orbits, Ernest
+Rutherford's experiments with atoms or Arthur Eddington's observations of the
+Solar eclipse with the goal of testing General Relativity. The data gathering
+process consisted of writing notes in a physical notebook and the analysis
+hardly required complex statistics. Today the situation is more complicated
+because much of the ``low-hanging fruit'' of scientific discovery (in the
+physical sciences) has been exhausted and the use of computers is absolutely
+central to the process. Progress today usually (but not always) requires
+coordination between many scientists, engineers and software engineers building
+highly complex experiments. Consider this list of some of the most notable
+scientific discoveries in the past few decades:
+\begin{itemize}
+    \item  Discovery of accelerated expansion of the Universe
+    \item  First sequencing of the human genome
+    \item  Detection of the Higgs Boson
+    \item  Paleogenomics studies of the origin of Homo Sapiens
+    \item  Detection of gravitational waves by LIGO
+    \item  Reconstruction of the first image of a Black Hole
+\end{itemize}
+All of these discoveries required gathering substantial amounts of data and the use of
+relatively complex statistical analysis techniques.
+Computation and statistical analysis are now absolutely central the process of scientific
+discovery.
+
+Simultaneously, in the past decade there's been a complete revolution in the
+field of machine learning/AI thanks to the advent of deep learning and neural
+networks. Besides the incredible progress on predicting patterns in language
+and vision, deep learning has also been used in science for things like solving
+the protein folding problem \citep{2021Natur.596..583J} and in mathematics for
+the purpose of discovering novel conjectures and theorems
+\citet{2021Natur.600...70D}. Deep learning has been less useful in physics and
+astronomy so far but as I will argue in this thesis, some of the technologies
+which underlie deep learning such as automatic differentiation and GPU
+computing are going to be (if they aren't already) crucial for processing and
+understanding complex datasets in physics and astronomy.
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=0.8\linewidth]{figures/ads_keyword_bayesian.pdf}
+        \caption{Number of NASA/ADS entries containing the keyword ``Bayesian'' per year.}
+        \label{fig:ads_keyword_bayesian}
+    \end{centering}
+\end{figure}
+
+There is another revolution worth mentioning. It is slightly less obvious than
+the one in machine learning but nevertheless significant. In the past two
+decades there has been a substantial increase in popularity of Bayesian
+statistics. Fig~\ref{fig:ads_keyword_bayesian} shows that number of entries in
+the NASA/ADS database containing the keyword ``Bayesian'' each year is growing
+almost exponentially. One of the reasons that these methods are so popular now
+even though they have been invented many decades ago is that they used to be
+very computationally expensive and the algorithms necessary to do proper
+Bayesian analysis were somewhat underdeveloped. This has changed drastically in
+the past decade. I will discuss these methods in detail in
+\S~\ref{sec:bayesian_statistics}. Much of this thesis is about applying
+Bayesian methods to problems in astronomy.
+
+Having situated the work presented this thesis in the present moment and point
+out some scientific and technological changes that I think are relevant, I will
+now focus on astronomy in particular. The kinds of questions in astronomy that
+most excite me are those which lead us closer to answering one of those two
+fundamental questions I stated at the beginning of this chapter. The question
+about the origin of life in the universe in particular. Since the first
+discoveries of planets outside of our Solar System in the early 90s
+\citep{1992Natur.355..145W,1995Natur.378..355M} thousands more have been
+confirmed\footnote{At the time of writing the
+    \href{https://exoplanetarchive.ipac.caltech.edu/}{NASA Exoplanet Archive}
+    contains more than 5000 exoplanets.} using methods such as transits, radial
+velocity, microlensing and direct imaging. Thanks to gravitational microlensing
+we now know \citep{2012Natur.481..167C} that there is an average one planet per
+star in the Milky Way. In addition to detecting the presence of the planets and
+inferring their properties such as mass, radius, and orbital period; it is now
+possible to measure the transmission and emission spectra of their atmospheres
+and even reconstruct crude maps of their surfaces\footnote{The subject of
+    Chapter~\ref{ch:mapping_exoplanets}.}
+\citep{2007Natur.447..183K,2012ApJ...747L..20M}. With the James Webb Space
+Telescope, we might even be able to detect biosignatures in the atmospheres of
+Earth size planets.
+
+Answering a grand question such as ``are there biologically produced complex
+molecules in this exoplanet atmosphere'' will not be easy even with
+cutting-edge instruments such as the James Webb Space Telescope. It will also
+almost certainly not be a clear yes or no kind of answer; rather, it will
+require a deep understanding (i.e. good \emph{models}) of the physics of
+exoplanet atmospheres, stellar variability, the response of the instrument,
+sophisticated statistical analysis of the drawing on multiple independent
+pieces of evidence and clear definitions of what it means to have detected
+something. To build a good model for the thing we really care about requires
+understanding also the the things we may intrinsically care less about
+(instrumental systematics, details of stellar variability, variations in
+Earth's atmosphere etc.).
+
+A large fraction of this thesis (with the exception of
+Chapter~\ref{ch:mapping_exoplanets}) is focused on building the necessary
+infrastructure (computation, statistics, interpretation) which should enable
+new discoveries in exoplanet (and planetary) science. Wherever possible I try
+to approach these problems from first principles thinking but with a heavily
+computational/statistical approach with a healthy dose of pragmatism. In the
+next two sections I will provide a brief introduction to specific areas I
+worked on and end the chapter with an outline of the following chapters.
+
 \section{Gravitational microlensing}
+\emph{``Do not bodies act upon light at a distance and by their action bend its rays,
+    and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
+Much later in 1912, even before he published his theory of General Relativity, Einstein
+predicted the that a distant light source could be magnified by a foreground mass
+relative to an observer on Earth if the two objects were very precisely aligned
+\citep{Renn1997}.
+At the time Einstein concluded that this \emph{gravitational lensing} effect was
+unobservable and did not even bother
+publishing the result until 24 years later a young amateur scientist R.W. Mandl convinced
+him to write a short paper on the subject \citep{1936Sci....84..506E}.
+In that paper he
+stated that \emph{``Of course, there is no hope of observing this phenomenon directly.
+    First, we shall scarcely ever approach closely enough to such a central line.
+    Second, [the angles] will defy the resolving power of our instruments''}.
+Still, the consensus at the time was that this was unobservable.
+
+The first definitive observation came in 1979 by \citet{1980ApJ...241..507Y}
+who observed a double image of the quasar Q0957+561 and concluded that the two
+images correspond to the same object whose light was distorted by a foreground
+gravitational lens in the form of a massive galaxy. That the same effect could
+be used to detect the presence of planets orbiting around their host stars was
+first theorized by \citet{1964PhRv..133..835L} who wrote that \emph{``the
+    primary effect of planetary deflectors bound to stars other than the Sun will
+    be to slightly perturb the lens action of these stars''} although he was also
+sceptical about the possibility of detection, saying that \emph{``associated
+    pulses would be so weak and infrequent and of such fleeting duration – perhaps
+    a few hours – as to defy detection''}. Gravitational lensing as a method for
+discovering exoplanets really took off with the work of Paczyński
+\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
+coined the term \emph{microlensing}, which refers to gravitational lensing in a
+regime where the images of the background object cannot be resolved but one can
+nevertheless measure its magnification as a function of time.
+
+\begin{figure}
+    \begin{centering}
+        \includegraphics[width=0.5\linewidth]{static/microlensing/crowded_field.jpg}
+        \caption{
+            Stellar field of a microlensing event GLE-2012-BLG-0406 (centered),
+            imaged by one of Las Cumbres Observatory's 2m telescopes,
+            showing the high density of stars typical in microlensing observations.
+            Credit: Y. Tsapras. Taken from
+            \url{http://microlensing-source.org/pictures/}.}
+        \label{fig:crowded_field}
+    \end{centering}
+\end{figure}
+
+Microlensing as a method for detecting exoplanets has some unique aspects.
+First, it's a one-off event which happens on a timescale of a few minutes up to
+several months depending on the distances to the background star and the
+lensing star and the mass of the lens. In addition to the fact that there's
+only one chance to observe such an event for it to happen at all requires
+extremely precise alignment between the lens and the background star and the
+chance of observing a stellar microlensing event is about one-in-a-million
+(CITE) for a typical star within the Milky Way. To observe a planetary signal
+is about an order of magnitude less likely than that. Hence, obtaining a decent
+sample of planetary events requires continuous monitoring on the order of
+$10^{8}$ stars at the time which way microlensing observations that to focus on
+the densest region of the Milky Way -- the galactic bulge.
+Figure~\ref{fig:crowded_field} a picture from of such a dense stellar field.
+Finally, in the vast majority of cases we do not detect any light from the lens
+itself, the collected photons are from a background star completely unrelated
+and distant from the lensing star. This is very different from other exoplanet
+discovery methods and it means that we only obtain dynamical properties of
+planets such their masses and periods.
+
+The aspects that make microlensing events that make them difficult to observe
+also mean that it provides a unique lens onto exoplanet systems. Relative to
+other methods such as transits and radial velocity, microlensing is sensitive
+to planets located at substantially greater distances, well outside of our
+Solar System neighborhood and even potentially to Milky Way's satellite
+galaxies such as the Magellanic Clouds and the nearby Andromeda galaxy (CITE).
+Microlensing is also sensitive to very to very small planets and planets which
+are further out from the star than those typically detected using transits and
+radial velocity. Microlensing surveys such as OGLE \citep{1993AcA....43..289U}
+and MOA \citep{1999PThPS.133..233M} have been continuously monitoring crowded
+stellar fields in the Milky Way since the 90s\footnote{Initially the focus was
+    on finding dark matter candidate particles -- so called MACHOs (Massive Compact
+    Halo Objects).} discovering thousands of stellar events and dozens of planetary
+events. In order to detect planetary deviations in the observed light curve it
+essential to have high cadence observations of the source star. The way surveys
+have traditionally worked is that once a particular star started to become
+magnified many additional small telescopes would start observing it to obtain
+denser coverage and sometimes space-based observatories get involved as well.
+The vast majority of microlensing events analyzed so far consist of
+observations from multiple observatories, each with its own unique aspects such
+as noise properties, cadence and photometric quality.
+
+Future surveys such as the ground based Rubin Observatory
+\citep{2019ApJ...873..111I} telescope and the space based Roman Telescope
+\citep{2019ApJS..241....3P} and Euclid \citep{2022arXiv220209475B} will detect
+tens of thousands of events in total. Although most of past work in the field
+focused on characterizing individual events binary lens events, answering
+questions about \emph{populations} with these new (but also existing) datasets
+requires scalable data analysis methods and algorithms and a clear set of
+guidelines on how to interpret the analysis products. This is a substantial
+challenge because microlensing events are notoriously difficult to model. Even
+though the datasets are relatively simple, consisting of multiple time series
+photometric light curves in different bands, the parameter space of even the
+simplest models is highly non-linear, correlated, relatively high dimensional
+and there are often near perfect degeneracies in the solutions.
+
+The assumptions that existing methods for modeling microlensing events rely on
+are often opaque and unquestioned. Discussions on model ``degeneracies''
+\citep{2014MNRAS.437.4006S,2019AJ....157...23H,2018AcA....68...43S,2009MNRAS.393..816D},
+correlated noise \citep{2015ApJ...812..136B,2019MNRAS.488.3308L} and model
+comparison \citep{2018AJ....155..259H,2019MNRAS.484.5608D} have been ongoing in
+the microlensing literature for decades without a clear solution and proper
+framing of the issue. In this thesis I will revisit these kinds of questions
+while taking into account many recent developments in the fields of
+computational statistics and machine learning. My main contributions are the
+following:
+\begin{itemize}
+    \item I wrote an open-source Python package \textsf{caustics} which enables fast and
+          accurate computation of binary and triple lens microlensing light curves using
+          contour integration. The code runs on both CPU and GPU architectures, and
+          crucially, for reasons that I will elaborate in subsequent chapters, it
+          supports \emph{automatic differentiation} of all outputs with respect to all
+          input parameters which enables the use of statistical methods which are orders
+          of magnitude more efficient than existing approaches. The code can easily be
+          extended to support quadruple lensing and arbitrary intensity profiles for the
+          source star.
+    \item I have revisited the topic of degeneracies in the posterior probability
+          distributions which appear in the context of microlensing and propose solutions
+          to these problems.
+    \item I have extended the complex polynomial root solver from \citet{Cameron2021}
+          such that it can be executed on GPU architectures and that it supports
+          automatic differentiation. This enables the evaluation of ~1M lens equation
+          solutions for point source binary and triple lenses in seconds using a CPU and
+          miliseconds using a GPU.
+    \item I propose an approach to statistical inference on a population of microlensing
+          events which different from existing methods and has some advantages.
+    \item I briefly discuss the issue of correlated noise in microlensing light curves
+          and propose ways to account for it using Gaussian Processes.
+\end{itemize}
+
 \section{Occultation and phase curve mapping}
+Interestingly, the second topic I will cover in this thesis has a history not
+unlike that of microlensing. It was proposed in the early 20th century and it
+was only much later that technology caught up with the idea. Back in 1906,
+\citet{1906ApJ....24....1R} pointed out that certain features in light curves
+of Solar System satellites may be attributed to inhomogeneities of their
+surfaces. The key idea is that although at any given time we only observe the
+total light from an unresolved satellite or planet, different portions of the
+surface are visible at different times so we may expect that some of the
+information about the surface intensity (be it emission from heat or reflected
+Sunlight) gets imprinted onto the light curve. \citet{1906ApJ....24....1R} also
+considered the inverse problem -- can we learn something about the surface of
+these objects starting from a light curve? The method he proposed is now known
+as \emph{phase curve mapping} and it was first attempted by
+\citet{1972ApJ...174..449L} who analyzed photometric light curves of Pluto in
+reflected light attempting to constrain variations in the \emph{albedo} of the
+surface with inconclusive results.
+
+Later works such as
+\citet{1992Icar...97..211B,1993Icar..102..134Y,1999AJ....117.1063Y} went a step
+further by using not only phase curves but also light curves of mutual
+\emph{occultations} of Pluto by its moon Charon to reconstruct albedo maps with
+greater success. The major advantage of occultations relative to just phase
+curves is that more information about the surface is encoded into the light
+curve because of the sharp limb of the occultor sweeping over the disc of an
+occulted body and blocking the reflected light. More importantly for this
+thesis, \citet{1994Icar..107..195S} was the first to observe the occultations
+of Jupiter's moon Io by another of Jupiter's moons -- Europa, and also
+occultations of Io by Jupiter itself. These observations were conducted using
+near-infrared telescopes such as NASA's Infrared Telescope Facility (IRTF) to
+observe emitted light from Io's surface which is covered with many time-varying
+and bright volcanic features. The observing campaign of Io has yielded insights
+into the the nature of its volcanic activity and it continues to this day. I
+will return to the subject of Io in great detail in
+Chapter~\ref{ch:mapping_io}.
+
+A natural question arises, can we do this with objects outside of the Solar
+System? The answer is yes. \citet{2007Natur.447..183K},
+\citet{2012ApJ...747L..20M}, and \citet{arXiv:1202.3829} used Spitzer
+mid-infrared observations of secondary eclipses of the Hot Jupiter HD189733b
+and found that surface emission is best described by the presence of a large
+hot spot on the dayside of the planet which is longitudinally offset from the
+substellar point. Similarly, \citet{2014Sci...346..838S}, produced temperature
+maps of the Hot Jupiter WASP-43b, \citet{2013ApJ...776L..25D} mapped the Hot
+Jupiter Kepler-7b in reflected light and \citet{2016Natur.532..207D} mapped the
+thermal emission from the Super Earth 55 Cancri e. These studies were only able
+to capture longitudinal variations in intensity. Real exoplanet atmospheres of
+are certain to have three-dimensional spatial inhomogeneities in emission more
+complex than a single hot spot due to the presence of clouds, zonal jets,
+storms, waves etc. \citep{2020SSRv..216..139S}.
+
+In recent years there have been significant advances in statistical modeling of
+phase curves and eclipse light curves. Most notably,
+\citet{2019AJ....157...64L} introduced the \textsf{starry} code which enables
+analytic computation of phase curves and occultation light curves for bodies
+with arbitrary emission maps expressed in a spherical harmonic basis (an idea
+dating back to \citet{1906ApJ....24....1R} ) and \citet{2021arXiv210306275L}
+expanded the algorithm for the (considerably more complicated) case of
+reflected light. In this thesis, I will present the work I've done in
+collaboration with other researchers from the planetary science and exoplanet
+communities on using \textsf{starry} to map the surface of Io and investigate
+the prospects for detecting fine spatial structure in the atmospheres of Hot
+Jupiters using JWST phase curves and eclipse light curves. My contributions to
+this area are the following:
+\begin{itemize}
+    \item I have developed a novel model for inferring emission maps of Io from
+          occultation light curves. First part of this project is published in
+          \citet{2022PSJ.....3...67B} though I haven't managed to complete the second
+          part.
+    \item I have investigated the problem of eclipse mapping of exoplanets using the
+          \textsf{starry} framework, focusing in particular on the possibility of
+          detecting planetary scale storms on Hot Jupiters using JWST. I found that it
+          extremely difficult to infer maps of higher resolution than a dipole order but
+          that should still be sufficient for detecting large scale weather and climate
+          change in the atmospheres of these planets.
+\end{itemize}
+The application of the occultation/eclipse mapping method to Io can be seen as the
+best case scenario for the application of the same method to exoplanets.
+I will show that even in the case of observing a bright object right in our neighborhood,
+this is by no means and easy task. Our ability to reconstruct the surface features of
+Io sets a sort of upper limit on what is possible with exoplanets.
+
+%\section{Outline}
+%The outline of the thesis is as follows. Chapter~\ref{ch:theoretical_min}
+%contains the theoretical background needed to understand the subsequent
+%chapters, covering the theory behind microlensing and occultation mapping, a
+%summary of Bayesian statistics in particular and statistical inference more
+%generally, and probabilistic programming which ties everything together.
+%Chapter~\ref{ch:microlensing} is the most extensive chapter in the thesis
+%covering all of my work on microlensing. I start by discussing the case of
+%single lens microlensing models, showing that it contains many of the
+%properties of more complex models. In particular I focus on the issue of
+%multi-modal posterior distributions which is a ubiquitous problem in
+%microlensing. I propose several solutions to this and demonstrate their
+%effectiveness. I then move on to the considerably more complex case of binary
+%and triple lens models starting with numerical methods for solving the lens
+%equation (finding roots of a complex polynomial equation). I demonstrate the
+%advantage of the Ehrlich-Aberth algorithm over previous methods and explain how
+%to make this algorithm differentiable and orders of magnitude faster through
+%the use of GPUs. Next, I cover the contour integration algorithm for computing
+%the magnification of a limb-darkened source star and its implementation in
+%\textsf{caustics}.
 
@@
 \chapter{The theoretical minimum}
+\label{ch:theoretical_min}
 
 \section{Gravitational microlensing}
+\label{sec:microlensing}
+
 \subsection{Deflection of light by gravity}
+
 \subsection{The different scales of lensing}
@@
 \section{Occultation and phase curve mapping}
+\label{sec:occultations}
 \subsection{History}
@@
 \section{Bayesian statistics}
+\label{sec:bayes}
 \subsection{Probability theory}
@@
 \section{Inference}
+\label{sec:inference}
 % check that paper on physics and ML by that MIT guy, there's a good stuff there. 
@@
 \subsection{Optimization vs. sampling}
-
 \subsection{Sampling and optimization methods}
@@
 \section{Model validation and comparison}
+\label{sec:model_comparison}
 \subsection{Cross validation}
@@
 \section{Putting it all together -- \emph{probabilistic programming}}
+\label{sec:programming}
 \subsection{The success of machine learning}
@@
 \chapter{Modeling microlensing events}
-\section{Dealing with multimodal posteriors}
-\section{A differentiable solver for the lens equation}
-\section{Population level inference in microlensing}
+\label{ch:microlensing}
+\section{Single lens events}
+\section{Modeling single lens events}
+\section{Numerical solutions to the lens equation}
+\section{\textsf{caustics} -- computing the magnification of an extended limb-darkened source}
+\section{Comparison with previous work}
+\section{Modeling binary lens events}
+\section{Population-level inference using hierarchical modeling}
 
@@
 \chapter{Mapping the surface of Io}
+\label{ch:mapping_io}
 \section{The data}
@@
 \chapter{Mapping the surfaces of exoplanets}
+\label{ch:mapping_exoplanets}
 \section{Previous work}
@@
 
+\bibliography{bib.bib}
+
 \end{document}

### 2022-05-22 src/figures/ads_keyword_bayesian.py

@@ -1,0 +1,17 @@
+import numpy as np
+from matplotlib import pyplot as plt
+
+
+data = np.loadtxt("../data/misc/ads_keyword_bayes.csv", delimiter=",")
+year, count, count_ref = data.T
+
+fig, ax = plt.subplots(figsize=(9, 5))
+
+ax.bar(year, count, color="white", edgecolor="k", linewidth=1.5)
+ax.set_xlim(1980, 2021)
+ax.set_ylabel("Number of entries for the keyword\n 'Bayesian' on NASA ADS")
+ax.set_xlabel("year")
+
+# Save as pdf
+fig.savefig("ads_keyword_bayesian.pdf", bbox_inches="tight")
+

### 2022-05-23 src/ms.tex
minor updates to introduction

@@
     and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
-Much later in 1912, even before he published his theory of General Relativity, Einstein
-predicted the that a distant light source could be magnified by a foreground mass
-relative to an observer on Earth if the two objects were very precisely aligned
-\citep{Renn1997}.
-At the time Einstein concluded that this \emph{gravitational lensing} effect was
-unobservable and did not even bother
-publishing the result until 24 years later a young amateur scientist R.W. Mandl convinced
-him to write a short paper on the subject \citep{1936Sci....84..506E}.
-In that paper he
-stated that \emph{``Of course, there is no hope of observing this phenomenon directly.
+Much later in 1911, even before he published his theory of General Relativity (GR), Einstein
+considered the deflection of light from a distant star passing close
+to a massive object in the foreground and derived an expression for the deflection angle of the light ray
+which was off by a factor of 2 relative to the correct value.
+He corrected the error in 1915 using GR.
+The first experimental confirmation of this prediction came in 1919 during the
+Solar eclipse when the astronomer Arthur Eddington famously measured the
+deflection of light from distant stars passing close to the limb of the Sun
+\footnote{Weather the data was actually good enough to make the claim that the
+    deflection angle is equal to the one Einstein predicted is contested}.
+Much later in 1936 a young amateur scientist R.W. Mandl convinced Einstein to
+write a short a paper on the lensing effect by a massive star acting as the lens
+instead of the Sun.
+In the paper \citep{1936Sci....84..506E} Einstein derived an expression for the
+magnification of the distant star when it is closely aligned to the foreground lens
+star relative to an observer and he predicted that a luminous ring would form
+if the two stars were perfectly aligned.
+He considered this effect curious but useless, stating that
+\emph{``Of course, there is no hope of observing this phenomenon directly.
     First, we shall scarcely ever approach closely enough to such a central line.
     Second, [the angles] will defy the resolving power of our instruments''}.
-Still, the consensus at the time was that this was unobservable.
 
-The first definitive observation came in 1979 by \citet{1980ApJ...241..507Y}
-who observed a double image of the quasar Q0957+561 and concluded that the two
-images correspond to the same object whose light was distorted by a foreground
-gravitational lens in the form of a massive galaxy. That the same effect could
-be used to detect the presence of planets orbiting around their host stars was
-first theorized by \citet{1964PhRv..133..835L} who wrote that \emph{``the
-    primary effect of planetary deflectors bound to stars other than the Sun will
-    be to slightly perturb the lens action of these stars''} although he was also
-sceptical about the possibility of detection, saying that \emph{``associated
-    pulses would be so weak and infrequent and of such fleeting duration – perhaps
-    a few hours – as to defy detection''}. Gravitational lensing as a method for
-discovering exoplanets really took off with the work of Paczyński
-\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
-coined the term \emph{microlensing}, which refers to gravitational lensing in a
-regime where the images of the background object cannot be resolved but one can
-nevertheless measure its magnification as a function of time.
+In this matter, Einstein was wrong. The famous astronomer Fritz Zwicky noticed
+the phenomenon is likely to be observable at galactic scales if the lens was a
+massive galaxy \citep{1937PhRv...51..290Z,1937PhRv...51..679Z} and that one
+could use the measurement of the deflection angle to weigh the lens. Zwicky was
+right and the first definitive observation came in 1979 by
+\citet{1980ApJ...241..507Y} who observed a double image of the quasar Q0957+561
+and concluded that the two images correspond to the same object whose light was
+distorted by a massive galaxy. Many other observations followed, including
+visible Einstein rings. Far from being just a curious phenomenon, galactic
+scale lensing is now one of the key methods of observational cosmology, used
+for inferring the parameters of the standard model and the distribution of dark
+matter in the universe.
 
@@
 
-Microlensing as a method for detecting exoplanets has some unique aspects.
-First, it's a one-off event which happens on a timescale of a few minutes up to
-several months depending on the distances to the background star and the
-lensing star and the mass of the lens. In addition to the fact that there's
-only one chance to observe such an event for it to happen at all requires
-extremely precise alignment between the lens and the background star and the
-chance of observing a stellar microlensing event is about one-in-a-million
-(CITE) for a typical star within the Milky Way. To observe a planetary signal
-is about an order of magnitude less likely than that. Hence, obtaining a decent
-sample of planetary events requires continuous monitoring on the order of
-$10^{8}$ stars at the time which way microlensing observations that to focus on
-the densest region of the Milky Way -- the galactic bulge.
-Figure~\ref{fig:crowded_field} a picture from of such a dense stellar field.
-Finally, in the vast majority of cases we do not detect any light from the lens
-itself, the collected photons are from a background star completely unrelated
-and distant from the lensing star. This is very different from other exoplanet
-discovery methods and it means that we only obtain dynamical properties of
-planets such their masses and periods.
+That the same effect could be used to detect the presence of planets orbiting
+around the lens star was first theorized by \citet{1964PhRv..133..835L} who
+wrote that \emph{``the primary effect of planetary deflectors bound to stars
+    other than the Sun will be to slightly perturb the lens action of these
+    stars''} although he was also sceptical about the possibility of detection,
+saying that \emph{``associated pulses would be so weak and infrequent and of
+    such fleeting duration – perhaps a few hours – as to defy detection''}.
+Gravitational lensing as a method for discovering exoplanets really took off
+with the work of Paczyński
+\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
+coined the term \emph{microlensing}, which refers to gravitational lensing in a
+regime where the images of the background object cannot be resolved but one can
+nevertheless measure its magnification as a function of time. Microlensing as a
+method for detecting exoplanets has some unique aspects. First, it is a one-off
+event which happens on a timescale of a few minutes up to several months
+depending on the distances to the background star and the lensing star and the
+mass of the lens. In addition to the fact that there's only one chance to
+observe such an event for it to happen at all requires extremely precise
+alignment between the lens and the background star and the chance of observing
+a stellar microlensing event is about one-in-a-million (CITE) for a typical
+star within the Milky Way. To observe a planetary signal is about an order of
+magnitude less likely than that. Hence, obtaining a decent sample of planetary
+events requires continuous monitoring on the order of $10^{8}$ stars at the
+time which way microlensing observations that to focus on the densest region of
+the Milky Way -- the galactic bulge. Figure~\ref{fig:crowded_field} a picture
+from of such a dense stellar field. Finally, in the vast majority of cases we
+do not detect any light from the lens itself, the collected photons are from a
+background star completely unrelated and distant from the lensing star. This is
+very different from other exoplanet discovery methods and it means that we only
+obtain dynamical properties of planets such their masses and periods.
 
@@
 \label{sec:microlensing}
+\subsection{Deflection of light by gravity}
+Gravitational lensing is a phenomenon fully described by Einstein's General
+Theory of Relativity (GR) in which light changes its direction of propagation
+when passing close to a massive body. GR predicts the following:
+\begin{itemize}
+    \item the presences of mass changes the spacetime geometry from the flat (Minkowski)
+          metric $\eta_{\mu\nu}$ to a curved metric, specified by the tensor $g_{\mu\nu}$
+    \item massless particles such as photons follow the null geodesics of the curved
+          spacetime which are obtained by solving the \emph{geodesic equation}
+\end{itemize}
+If we restrict ourselves to the regime where the metric is time independent
+and the particles are allowed to travel at any velocity less than $c$,
+also known as the \emph{weak field} or \emph{linearized approximation }
+of GR, the metric tensor
+can be written as
+\begin{equation}
+    g_{\mu\nu}=\eta_{\mu\nu}+h_{\mu\nu}\quad,
+\end{equation}
+where $h_{\mu\nu}=
+    \textrm{diag}(-2\Phi,-2\Phi,-2\Phi,-2\Phi)$\footnote{Assuming the
+    the metric sign convention $(-,+,+,+)$} \citep{carroll_2019}.
+The (static) Newtonian gravitational potential $\Phi$ obeys the Poisson equation
+\begin{equation}
+    \nabla^2\Phi=4\pi G\rho
+\end{equation}
+where $\rho$ is the mass density of the distribution of mass located between the observer
+and the light source.
+The geometry of the lensing system consisting of a distant source of light, the
+observer, and a distribution of mass with density $\rho$ located between the
+source and the observer is shown in Figure~\ref{fig:deflection_angle}.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.6\linewidth]{static/microlensing/deflection_angle.pdf}
+    \caption{Geometry of gravitational lensing.
+        A geodesic curve $x^\mu(\lambda)$ is deflected
+        by an angle $\hat\alpha$ from its initial trajectory
+        due to the presence of a massive body with mass density $\rho$. Figure
+        adapted from \citet{carroll_2019}.}
+    \label{fig:deflection_angle}
+\end{figure}
+A photon is deflected as it travels from a source to an observer by a
+\emph{deflection angle} $\hat\alpha$, a vector in the
+plane  perpendicular to the wave vector $\vec{k}$ pointing in the direction of
+photon propagation.
+The deflection angle is given by \citep{carroll_2019}
+\begin{equation}
+    \hat\alpha=2\int\nabla_\perp\Phi ds
+    \label{eq:deflection_angle_general}
+\end{equation}
+where $\nabla_\perp\Phi$ is the gradient of the potential in the direction transverse
+to the path of the photon and $s$ is the spatial distance traveled.
+For a point mass $M$, the potential is given by the potential is
+\begin{equation}
+    \Phi=- \frac{GM}{(b^2+x^2)^{1/2}}
+\end{equation}
+where $x$ parametrizes the straight line distance between the observer and the lens and
+$b$ is the impact parameter of the light ray.
+The deflection angle (in SI units) is then given by
+\begin{equation}
+    \hat\alpha= \frac{4GM}{c^2b}= \frac{2R_s}{b}
+\end{equation}
+where the integral has been taken from $-\infty$ to $\infty$ (i.e. assuming that both
+the source and the observer are located far away from the deflecting mass) and
+$R_s=2GM/c^2$
+is the Schwarzschild radius.
+The deflection angle $\hat\alpha$ is directly proportional to the mass of the lens and
+it does not depend on the wavelength of the light.
+It also increases with the proximity of
+the light ray to the lensing mass
+which is the opposite behaviour to that of a
+classical convex lens for which the deflection angle
+vanishes at the center of the lens.
+The same year Albert Einstein published his General theory of Relativity (1916),
+English
+physicist Sir Oliver Lodge suggested the light bending phenomenon could produce
+a \emph{gravitational lens}.
+He warned that it wouldn't be a lens in the usual sense,  because it does not have a
+focal length.
+In 1936, Einstein showed that a bright background light source perfectly aligned
+with a foreground massive body would distort the source in the form of a ring
+around the massive body.
+However, he concluded that that there is no great chance of discovering such a
+phenomenon.
+Once astronomers started finding such ring like images in their data the name
+\emph{gravitational lensing} caught on.
 
-\subsection{Deflection of light by gravity}
+The resulting deflection angle is very small, for example, for the Sun we have
+$GM/c^2=1.48\times 10^5\si{\centi\meter}$ (2.95 km), $R=6.96\times
+    10^{10}\si{\centi\meter}$ and $\hat\alpha=1.75$ arc seconds. This deflection
+effect was famously claimed to have been observed by Eddington when during a
+total solar eclipse in 1919 \citet{2001ASPC..252...21C}, he observed the
+position of background stars very close to the Sun's radius in the sky and
+compared them to previous measurements taken when the stars weren't close to
+the Sun in the sky. He concluded that the observed deflection was in agreement
+with GR, although it is doubtful that his data is actually good enough to
+distinguish between the GR prediction for the deflection of light and the
+Newtonian prediction\footnote{ Gravitational bending on light can be predicted
+    in Newtonian mechanics, assuming that light is a massless particle and using
+    the equivalence principle which says that that all particles independent of
+    their mass accelerate towards massive bodies.} , which is half of the GR value.
+
+It is important to note that Eq.~\ref{eq:deflection_angle_general} is only
+valid when the linearized approximation of GR is sufficient. Also, in this
+approximation, the deflection angles are additive and the total deflection due
+to several massive bodies is just the sum of the individual deflection angles.
+The linearized approximation breaks down when the impact parameter $b$
+approaches the Schwarzschild radius of the lensing mass and the linearized
+metric is no longer sufficient. For most cases of astrophysical interest this
+is not an issue.
 

### 2022-05-25 src/ms.tex
chapter 2 progress

--- 2022-05-22 src/ms.tex
@@
     and is not this action strongest at the least distance?''} asked \citet{Newton1704}.
-Much later in 1912, even before he published his theory of General Relativity, Einstein
-predicted the that a distant light source could be magnified by a foreground mass
-relative to an observer on Earth if the two objects were very precisely aligned
-\citep{Renn1997}.
-At the time Einstein concluded that this \emph{gravitational lensing} effect was
-unobservable and did not even bother
-publishing the result until 24 years later a young amateur scientist R.W. Mandl convinced
-him to write a short paper on the subject \citep{1936Sci....84..506E}.
-In that paper he
-stated that \emph{``Of course, there is no hope of observing this phenomenon directly.
-    First, we shall scarcely ever approach closely enough to such a central line.
-    Second, [the angles] will defy the resolving power of our instruments''}.
-Still, the consensus at the time was that this was unobservable.
+Much later in 1911, even before he published his theory of General Relativity (GR), Einstein
+considered the deflection of light from a distant star passing close
+to a massive object in the foreground and derived an expression for the deflection angle of the light ray
+which was off by a factor of 2 relative to the correct value.
+He corrected the error in 1915 using GR.
+The same year Albert Einstein published his General theory of Relativity (1916),
+English physicist Sir Oliver Lodge suggested the light bending phenomenon could produce
+a \emph{gravitational lens}.
+He warned that it wouldn't be a lens in the usual sense,  because it does not have a
+focal length.
 
-The first definitive observation came in 1979 by \citet{1980ApJ...241..507Y}
-who observed a double image of the quasar Q0957+561 and concluded that the two
-images correspond to the same object whose light was distorted by a foreground
-gravitational lens in the form of a massive galaxy. That the same effect could
-be used to detect the presence of planets orbiting around their host stars was
-first theorized by \citet{1964PhRv..133..835L} who wrote that \emph{``the
-    primary effect of planetary deflectors bound to stars other than the Sun will
-    be to slightly perturb the lens action of these stars''} although he was also
-sceptical about the possibility of detection, saying that \emph{``associated
-    pulses would be so weak and infrequent and of such fleeting duration – perhaps
-    a few hours – as to defy detection''}. Gravitational lensing as a method for
-discovering exoplanets really took off with the work of Paczyński
-\citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
-coined the term \emph{microlensing}, which refers to gravitational lensing in a
-regime where the images of the background object cannot be resolved but one can
-nevertheless measure its magnification as a function of time.
+The first experimental confirmation of the deflection of light by a mass came
+in 1919 during the Solar eclipse when the astronomer Arthur Eddington famously
+measured the deflection of light from distant stars passing close to the limb
+of the Sun \footnote{He concluded that the observed deflection was in agreement
+    with GR, although it is doubtful that the data was actually good enough to
+    distinguish between the GR prediction for the deflection of light and the
+    Newtonian prediction which is derived using the equivalence principle and is
+    equal to one half the GR value.}. Much later in 1936 a young amateur scientist
+R.W. Mandl convinced Einstein to write a short a paper on the lensing effect by
+a massive star acting as the lens instead of the Sun. In the paper
+\citep{1936Sci....84..506E} Einstein derived an expression for the
+magnification of the distant star when it is closely aligned to the foreground
+lens star relative to an observer and he predicted that a luminous ring would
+form if the two stars were perfectly aligned. He considered this effect curious
+but useless, stating that \emph{``Of course, there is no hope of observing this
+    phenomenon directly. First, we shall scarcely ever approach closely enough to
+    such a central line. Second, [the angles] will defy the resolving power of our
+    instruments''}.
+
+In this matter, Einstein was wrong. The famous astronomer Fritz Zwicky noticed
+the phenomenon is likely to be observable at galactic scales if the lens was a
+massive galaxy \citep{1937PhRv...51..290Z,1937PhRv...51..679Z} and that one
+could use the measurement of the deflection angle to weigh the lens. Zwicky was
+right and the first definitive observation came in 1979 by
+\citet{1980ApJ...241..507Y} who observed a double image of the quasar Q0957+561
+and concluded that the two images correspond to the same object whose light was
+distorted by a massive galaxy. Many other observations followed, including
+visible Einstein rings. Far from being just a curious phenomenon, galactic
+scale lensing is now one of the key methods of observational cosmology, used
+for inferring the parameters of the standard model and the distribution of dark
+matter in the universe.
 
@@
 \label{sec:microlensing}
+Gravitational lensing is generally divided into multiple classes depending
+on weather the effect is discernible at the level of individual objects or a
+a statistical sample of object. The main division is between \emph{weak lensing}
+and \emph{strong lensing}.
+The former refers to lensing by galaxies and clusters of galaxies on cosmological
+scales where the deflection  is impossible to detect in a single background source
+but one can tease out the effect in a statistical sense.
+Strong lensing refers to lensing of individual objects and it is further divided into
+\emph{macrolensing} -- lensing of galaxies where the multiple images of the source
+are resolved, and \emph{microlensing}  -- the images are generally not resolved,
+and the observable is the total magnification of the source as a function of time (a light curve).
+In case of microlensing the light source is either a quasar or a star (sometimes a binary star)
+and the lenses are stars, brown dwarfs, planets, and compact objects such as black holes, neutron
+stars and white dwarfs. In this thesis we focus specifically on microlensing
+with stars as the light source instead of quasars\footnote{Quasar microlensing
+    is a somewhat separate community from the rest of the microlensing community.}.
 
 \subsection{Deflection of light by gravity}
+Gravitational lensing is a phenomenon fully described by Einstein's General
+Theory of Relativity (GR) in which light changes its direction of propagation
+when passing close to a massive body. GR predicts the following:
+\begin{itemize}
+    \item the presences of mass changes the spacetime geometry from the flat (Minkowski)
+          metric $\eta_{\mu\nu}$ to a curved metric, specified by the tensor $g_{\mu\nu}$
+    \item massless particles such as photons follow the null geodesics -- minimum
+          distance paths in curved spacetime which are obtained by solving the
+          \emph{geodesic equation}
+\end{itemize}
+If we restrict ourselves to the regime where the metric is time independent
+and the particles are allowed to travel at any velocity less than $c$,
+also known as the \emph{weak field} or \emph{linearized approximation }
+of GR, the metric tensor
+can be written as
+\begin{equation}
+    g_{\mu\nu}=\eta_{\mu\nu}+h_{\mu\nu}\quad,
+\end{equation}
+where $h_{\mu\nu}=
+    \textrm{diag}(-2\Phi,-2\Phi,-2\Phi,-2\Phi)$\footnote{Assuming the
+    the metric sign convention $(-,+,+,+)$} \citep{carroll_2019}.
+The (static) Newtonian gravitational potential $\Phi$ obeys the Poisson equation
+\begin{equation}
+    \nabla^2\Phi=4\pi G\rho\quad,
+\end{equation}
+where $\rho$ is the mass density of the distribution of mass located between the observer
+and the light source.
+The geometry of the lensing system consisting of a distant source of light, the
+observer, and a distribution of mass with density $\rho$ located between the
+source and the observer is shown in Figure~\ref{fig:deflection_angle}.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.6\linewidth]{static/microlensing/deflection_angle.pdf}
+    \caption{Geometry of gravitational lensing.
+        A geodesic curve $x^\mu(\lambda)$ is deflected
+        by an angle $\hat\alpha$ from its initial trajectory
+        due to the presence of a massive body with mass density $\rho$. Figure
+        adapted from \citet{carroll_2019}.}
+    \label{fig:deflection_angle}
+\end{figure}
+A photon is deflected as it travels from a source to an observer by a
+\emph{deflection angle} $\hat\alpha$, a vector in the
+plane  perpendicular to the wave vector $\vec{k}$ pointing in the direction of
+photon propagation.
+The deflection angle is given by \citep{carroll_2019}
+\begin{equation}
+    \hat\alpha=2\int\nabla_\perp\Phi \ud s\quad,
+    \label{eq:deflection_angle_general}
+\end{equation}
+where $\nabla_\perp\Phi$ is the gradient of the potential in the direction transverse
+to the path of the photon and $s$ is the spatial distance traveled.
+For a point mass $M$, the potential is given by the potential is
+\begin{equation}
+    \Phi=- \frac{GM}{(b^2+x^2)^{1/2}}\quad,
+\end{equation}
+where $x$ parametrizes the straight line distance between the observer and the lens and
+$b$ is the impact parameter of the light ray.
+Integrating from $-\infty$ to $\infty$ (i.e. assuming that both
+the source and the observer are located far away from the deflecting mass), we obtain
+the deflection angle:
+\begin{equation}
+    \hat\alpha= \frac{4GM}{c^2b}= \frac{2R_s}{b}\quad,
+\end{equation}
+where $R_s=2GM/c^2$ is the Schwarzschild radius.
+$\hat\alpha$ is directly proportional to the mass of the lens and
+it does not depend on the wavelength of the light. It also increases with the proximity of
+the light ray to the lensing mass which is the opposite behavior to that of a
+classical convex lens for which the deflection angle vanishes at the center of the lens.
+The gravitational lens does not have a focal length and so it is not a lens in the usual
+sense.
+The resulting deflection angle is very small, for example, for the Sun we have
+$GM/c^2=1.48\times 10^5\si{\centi\meter}$ (2.95 km), $R=6.96\times
+    10^{10}\si{\centi\meter}$ and $\hat\alpha=1.75$ arc seconds. This is the angle
+that was claimed to have been observed by Eddington when during the 1919 total
+solar eclipse.
 
-\subsection{The different scales of lensing}
-%% Read the "scales of lensing" paper by that Italian guy
-\subsection{Microlensing}
-\subsection{The point mass lens and the lens equation}
+Equation~\ref{eq:deflection_angle_general} is only valid when the linearized
+approximation of GR is sufficient. In this approximation, the deflection angles
+are additive and the total deflection caused by a group massive bodies is just
+the sum of the individual deflection angles. The linearized approximation
+breaks down when the impact parameter $b$ approaches the Schwarzschild radius
+of the lensing mass and the linearized metric is no longer sufficient. Except
+for modeling lensing in the close vicinity of a black, the linearized
+approximation is sufficient.
+
+\subsection{The magnification of a point source by a point lens}
+Consider a system consisting of an observer $O$, a point mass $M$, and a point
+light source $S$. The geometry of this system is shown in
+Figure~\ref{fig:lens_geometry}. The lens is located in the \emph{lens plane}, a
+plane perpendicular to the observer--lens axis, at a distance $D_L$ away from
+the source. The light source is located in the \emph{source plane}
+perpendicular to the observer--source axis, at a distance $D_S$ away from the
+observer, and the observer is in the \emph{observer plane}. The assumption that
+the location of the source, observer, and the lens can be parametrized using
+angular coordinates (or Cartesian coordinates if the distances to the lens and
+the source are known) in their respective planes is justifiable because the
+deflection angles involved are very small and the distance between the observer
+and any point of interest in the lens plane is approximately constant.
+\begin{figure}[!ht]
+    \centering
+    \includegraphics[width=0.4\linewidth]{static/microlensing/lens_geometry.pdf}
+    \caption{The geometry of a system consisting of a point mass lens $M$ located at distance
+        $D_L$ at angular separation $\beta$ from the observer, and a single light source $S$ located at distance $D_S$, emitting a light ray which gets
+        deflected by an angle $\hat\alpha$ from its
+        original trajectory. As a result the observer sees the source at angular separation
+        $\theta$. Figure adapted from \citet{1992grle.book.....S}.}
+
+    \label{fig:lens_geometry}
+\end{figure}
+Let $\theta$ be the apparent angular position of the source in the sky with respect to the
+observer--lens axis and $\beta$ be the actual angular position of the source.
+Assuming a metric for spacetime in between the source and the observer that is
+approximately Euclidian\footnote{This is a valid assumption for galactic sources but not for
+    extragalactric source such as quasars and galaxies which require a cosmological
+    model for the metric.} the distance from the lens to the
+source is $D_S-D_L$.
+
+From Figure~\ref{fig:lens_geometry} it follows that the relationship between
+the observed and the actual location of the source is given by
+\begin{equation}
+    \beta=\theta-2 R_s \frac{D_S-D_L}{D_LD_S}
+    \frac{1}{\theta}\quad,
+    \label{eq:lens_equation}
+\end{equation}
+which is known as the \emph{lens equation} or sometimes also the \emph{ray-tracing}
+equation. If the source and the lens are perfectly aligned with respect to the observer,
+then $\beta=0$ and we have
+\begin{equation}
+    \theta \equiv \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}
+\end{equation}
+The distance term $D_L^{-1}-D_\textrm{s}^{-1}$ is proportional to the
+lens-source parallax.
+
+The image of the source forms a ring (the Einstein ring) around the lens star
+with angular radius $\theta_E$ -- the \emph{Einstein radius}. $\theta_E$
+depends on the mass of the lens and the distances to the lens and the source.
+The angular Einstein radius sets the characteristic scale of a microlensing
+event, it is a natural unit for most microlensing parameters. For an M-dwarf
+star lens in the galactic disc ($D_L\sim 3\,\textrm{kpc}$) and a source star in
+the galactic bulge ($D_S\sim 8\,\textrm{kpc}$) we have $\theta_E\sim
+    1\,\textrm{mas}$ so the typical scale of microlensing is on the order of
+miliarcseconds. The images of the source star are only rarely observable using
+the most advanced optical interferometers such as the GRAVITY instrument on
+ESO's Very Large Telescope whose lower resolution limit is $2\,\textrm{mas}$
+\citep{arXiv:1705.02345}. \cite{2019ApJ...871...70D} were the first to resolve
+the images of a microlensing event, finding the value of
+$\theta_E=1.87\,\textrm{mas}$ using the GRAVITY instrument.
+
+We can rewrite Equation~\ref{eq:lens_equation} by defining new dimensionless
+angular coordinates scaled by the angular Einstein radius:
+\begin{equation}
+    x= \frac{\theta}{D_L\theta_E}, \quad y=\frac{\beta}{D_S\theta_E}
+\end{equation}
+The lens equation then takes the very simple form
+\begin{equation}
+    y= x- \frac{ 1}{x}
+    \label{eq:lens_equation_dimensionless}
+\end{equation}
+Equation~\ref{eq:lens_equation_dimensionless} is a nonlinear mapping between the
+source plane $y$ and the image plane $x$.
+If we solve for $x$, we obtain two solutions for the position of the images, given by
+
+\begin{equation}
+    x_\pm= \frac{1}{2} \left( y\pm\sqrt{y^2 +4}\right)
+    \label{eq:images_location}
+\end{equation}
+For non zero $y$ the so called \emph{minor image} is always located
+within the Einstein radius ($\lvert x_-\rvert<1$) while the \emph{major image}
+is located outside of it ($\lvert x_+\rvert>1$).
+
+Since gravitational lensing does not involve any additional emission or
+absorption along a deflected ray of light and since the wavelength of the light
+ray does not change when it encounters the lens, the \emph{surface brightness}
+(flux density per unit angular area) $I$ of the source image is identical to
+the surface brightness of the unlensed source. The flux of an infinitesimal
+source is a product of the surface brightness and the solid angle $\Delta
+    \omega$ subtended by the source in the sky which changes throughout the
+microlensing event because the source and the lens are in motion relative to an
+observer. The ratio of the original and the lensed source flux, the
+\emph{magnification} $A$, is then given by the ratio of the two solid angles
+\begin{equation}
+    A= \frac{\Delta \omega}{(\Delta\omega)_0}\quad,
+    \label{eq:magnification}
+\end{equation}
+where $0$ denotes the unlensed solid angle.
+Assuming an infinitesimal source size at angular location $\mathbf y =(y_1,y_2)$ subtending an angle
+$\Delta (\omega)_0$, and an image of the source at location $\mathbf x=(x_1,x_2)$ subtending an angle
+$\Delta \omega$; the ratio between the solid angle of the source and the image
+is given by the determinant of the Jacobian matrix of the
+lens mapping $\mathbf{x}\rightarrow\mathbf{y}$, evaluated at the location of the images
+\begin{equation}
+    \frac{(\Delta\omega)_0}{\Delta\omega} =\left\lvert\textrm{det}
+    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert
+\end{equation}
+The magnification $A$ is then given by the inverse Jacobian determinant of the lens mapping:
+\begin{equation}
+    A= \left\lvert\det
+    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert^{-1} \label{eq:magnification_general}
+\end{equation}
+The images for which the determinant of the Jacobian of the lens mapping is
+positive are said to have positive \emph{parity} and vice versa.
+
+Looking at Equation~\ref{eq:magnification_general}, we see that the
+magnification diverges when the Jacobian determinant of the lens mapping
+vanishes. Curves in the lens plane for which the determinant of the lens
+mapping vanishes, that is,
+\begin{equation}
+    \left\lvert\det
+    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert=0
+\end{equation}
+are called \emph{critical curves}.
+The critical curves in the lens plane can be mapped to the source plane using
+the lens equation $\mathbf{x}\rightarrow \mathbf{y}$ and these curves are
+are called \emph{caustic curves}.
+Although the magnification factor in Equation~\ref{eq:magnification_general} formally
+diverges at the points of critical curves or caustics,
+the divergence is not physical because light sources aren't truly point like.
+If the finite angular size of the source is taken into
+account, the magnification is an integral of Equation~\ref{eq:magnification_general} over the
+extent of the source, weighted by the source brightness and that integral is always finite.
+Both the critical and the caustic curves are closed and one can prove that \textbf{the
+    number of images changes by two if and only if the source crosses a caustic curve}
+\citep[][Chapter 6]{1992grle.book.....S}.
+
+We can derive a simpler expression for Equation~\ref{eq:magnification} by
+switching to polar coordinates $(v,\phi)$ in the lens plane. The lens equation
+for the two vector components of the apparent source position is given by
+\begin{align}
+    y_1= & x_1- \frac{x_1}{x_1^2+x_2^2} \\
+    y_2= & x_2- \frac{x_2}{x_1^2+x_2^2}
+\end{align}
+Introducing polar coordinates $x_1=v\cos\phi,\;x_2=v\sin\phi$, we have
+\begin{equation}
+    \left\lvert\textrm{det}
+    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert =1- \frac{1}{v^4}\quad,
+\end{equation}
+By substituting in the locations of the source images Equation~\ref{eq:images_location} into the
+above expression, we obtain the magnification of the source for both images
+\begin{equation}
+    A_\pm=\left(1- \frac{1}{x^4_\pm} \right)^{-1}
+\end{equation}
+The total magnification is then the sum of the two magnifications, and is given by
+\citep{1936Sci....84..506E}
+\begin{equation}
+    A(u)=\lvert A_-\rvert+\lvert A_+\rvert= \frac{u^2+2}{u\sqrt{u^2+4}}\quad,
+    \label{eq:magnification_point_lens}
+\end{equation}
+where $u = \sqrt{y_1^2+y_2^2}$ is the the magnitude of the position
+vector of the source star.
+The critical curve of the point lens is the Einstein
+ring corresponding to $v=\sqrt{x_1^2+x_2^2}=1$ and the caustic curve is mapped to a
+a single point in the source plane at $u=0$.
+For source separations much smaller than the angular Einstein radius ($u\ll 1$), the
+magnification is approximately $A(u)\simeq 1/u$, in the opposite case ($u\gg 1$)
+we have $A(u)\simeq 1+ 2/u^4$, that is, the magnification falls off rapidly the further
+away the source is from the lens.
+
+%The angular Einstein radius determines a characteristic timescale
+%$t_E=\theta_E/|\boldsymbol{\mu}_{LS}|$ on which the magnification of of the
+%source changes, where $\boldsymbol{\mu}_{LS}$ is the relative proper motion
+%vector between the source and the lens
+%$\boldsymbol{\mu}_{LS}\equiv\boldsymbol{\mu}_L- \boldsymbol{\mu}_S$. The
+%predicted flux of a star undergoing a microlensing event is then given by the
+%following expression
+%\begin{equation}
+%    f=F_SA(t)+F_B
+%    \label{eq:flux_paczynski}
+%\end{equation}
+%where $F_S$ is the intrinsic flux of the star and
+%$F_B$ is any background flux due to other stars in the field which
+%might be blended with the source star, usually called the \emph{blending flux}.
+%$A(t)$ is a non-linear time-dependant magnification mapping which depends
+%on the properties of the lens and the geometry of the lensing event.
+%In the simplest case, that of a point source being magnified by a point lens,
+%the magnification is given by
+%\begin{equation}
+%    A(t)= \frac{u(t)^2+2}{u(t)\sqrt{u(t)^2+4}}
+%\end{equation}
+%where $u(t)$ is the separation between the lens and the source expressed in
+%units of the angular Einstein radius.
+
+To be able to evaluate Equation~\ref{eq:magnification_point_lens} in practice
+we have to parametrize the position of the source on the sky $u$ as a function
+of time. To derive an expression for $u(t)$ we assume the the motion of the
+observer, lens and the source is rectilinear (acceleration can be neglected).
+We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
+on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
+vector. Likewise for the lens. We thus have
+\begin{align}
+    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
+    \left(t-t_{0}\right) \boldsymbol{\mu}_S
+    \label{eq:source_position}                              \\
+    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
+    +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
+    \label{eq:lens_position}
+\end{align}
+where $t_0$ is some reference time.
+The relative position vector of the lens with respect to the source is then
+\begin{equation}
+    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
+        -\boldsymbol{\theta}_S(t)}{\theta_E}=
+    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
+    +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
+    \label{eq:relative_trajectory_no_parallax}
+\end{equation}
+where
+$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
+is the relative position at $t_0$ and
+$\boldsymbol{\mu}_{LS}\equiv \boldsymbol{\mu}_{L}- \boldsymbol{\mu}_{S}$ is the
+relative proper motion.
+Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
+perpendicular to each other, it follows that the magnitude of the relative
+separation is
+\begin{equation}
+    u(t)=\sqrt{u_0^2+ \left(\frac{t-t_0}{t_E}\right)^2}
+\end{equation}
+where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
+and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
+The model for the light curve is then characterized by
+the parameters $\{F_S,F_B, t_0, u_0, t_E\}$ and is known in the
+literature as the Paczy\'nski curve.
+
+The magnification as a function of time for different impact parameters $u_0$
+is shown in Figure~\ref{fig:single_lens_mag}. Notice the very steep fall off in
+magnification as as we move away from the Einstein ring ($A(u)\propto
+    1+2u^{-4}$). The curves shown in Figure~\ref{fig:paczynski_curve} are often
+called \emph{Paczyński curves} \citep{1986ApJ...304....1P,1986ApJ...301..503P},
+in honour of the late Polish astronomer Bohdan Paczyński, one of the key
+figures of gravitational microlensing.
+
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=0.7\linewidth]{figures/paczynski_curve.pdf}
+        \caption{
+            Magnification of a point source by a point lens as a function of time. The
+            various magnification curves correspond to different impact parameters $u_0$. The inset figure
+            shows the source trajectories in the lens plane, the dashed circle corresponds
+            to the Einstein radius $v=1$.}
+        \label{fig:paczynski_curve}
+    \end{centering}
+\end{figure}
+
+\subsection{Magnification of an extended source}
+For a relatively small subset of microlensing events the source star passes
+very close to the caustic and the variation of magnification over the extent of
+the source star disc is non-negligible. For those events the point source
+approximation breaks down and we have to take into account the finite angular
+radius $\theta_\star$. This will generally happen when $u_0 \lesssim
+    \rho_\star/2$ \citep{1997ApJ...477..580G} where
+$\rho_\star=\theta_\star/\theta_E$. The effect of finite source effects on the
+magnification curve thus matters only near the peak of an event and it results
+in a rounder top of the curves shown in Figure~\ref{fig:paczynski_curve}.
+Unfortunately there are no analytic solutions for the magnification of an
+extended source lensed by a single point lens with a general surface brightness
+profile. \citet{1994ApJ...421L..71G} derived an approximate solution valid for
+uniform brightness sources with $\rho_\star\lesssim 0.1$ and
+\citet{1994ApJ...430..505W} derived a general solution (also for a uniform
+brightness source) involving solutions to elliptic integrals of the first,
+second and third kind. A general solution involving Fast Fourier Transforms
+which is applicable to any source profile was recently proposed by
+\citet{2022arXiv220306637S}.
+
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/single_lens_images.pdf}
+        \caption{Images of an extended limb-darkened source star lensed by a single
+            point lens for varying position of the source star. The top panels show the magnification map with a logarithmic scale
+            colormap  consisting of a single point caustic. The semi-transparent circle
+            is the source star disc with radius $\rho_\star=0.15$. The bottom row shows
+            the two images merging into the Einstein ring as the source moves over the caustic.}
+        \label{fig:single_lens_images}
+    \end{centering}
+\end{figure}
+
+In Figure~\ref{fig:single_lens_images} we show the magnification map and the
+images of an extended limb-darkened source with $\rho_\star=0.15$ as it
+approaches the point caustic for a single lens. The top panels show the
+magnification map in the source plane using a logarithmic scale colormap. The
+source disc is shown (semi-transparent grey circle) at different positions
+relative to the caustic. The bottom panels show the resulting images in the
+image plane, this is what we would see on the sky if we could resolve the
+microlensing event. As the source moves closer to the caustic the area (and
+hence the magnification) of the two images increases and eventually they merge
+forming an Einstein ring.
+
+To produce these plots I evaluated the lens mapping
+(Eq.~\ref{eq:lens_equation_dimensionless}) on a regular grid in the image plane
+and then plotted the surface brightness of the corresponding point on the
+source disc in the source plane. I will discuss all of this in far greater
+detail In Chapter~\ref{ch:microlensing} but for now I should mention a few key
+points:
+\begin{itemize}
+    \item Every point inside the two images maps to a single point on the source disc via
+          the lens equation. The inverse mapping is one-to-many because each point on the
+          source disc maps to two points (point lens images) in the image plane. The
+          points on the limbs of the two images correspond to the points on the limb of
+          the source disc. We cannot predict the location of the images without inverting
+          the lens mapping
+    \item To numerically compute the magnification of an extended source, we could either
+          integrate the magnification function convolved with the source brightness
+          profile over the entire source disc (a two-dimensional integral), or we could
+          do the same integral in the image plane. The first approach is highly
+          numerically unstable because we would need to integrate over a function which
+          diverges as $1/r$ approaching the caustic and this would require a very large
+          number of lens equation evaluations. Hence it is much better to integrate in
+          the image plane where the function is smooth.
+    \item Decreasing the source radius $\rho_\star$ results in narrower arcs of the
+          images (smaller radial extent).
+\end{itemize}
+
 \subsection{Parallax}
-\subsection{Finite source effects}
-\subsection{The binary lens}
-\subsection{Multiple lenses}
+In Equation~\ref{eq:relative_trajectory_no_parallax} we have assumed that the
+relative motion of the lens with respect to the source on the plane of the sky
+is rectilinear. This is a reasonable approximation in a barycentric frame of
+reference if the acceleration of the source and the lens can be neglected.
+However, most microlensing events are observed from a non-inertial geocentric
+frame so that the observed trajectory is modified due the local acceleration of
+the Earth and this effect needs to be taken into account in a realistic model.
+In order to include the parallax effect, we need to modify
+Equations~\ref{eq:source_position} and \ref{eq:lens_position} to include the
+projected motion of the Sun relative to the Earth on the plane of the sky. We
+project $\mathbf s$ -- the position vector of the Sun relative to Earth, onto a
+plane perpendicular to the line of sight towards the source star (plane of the
+sky) which is defined by the unit vector $\boldsymbol{\hat n}$ normal to the
+plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky coordinates of
+the source star $(\alpha,\delta)$ (right ascension and declination).
+
+We choose to work in geocentric equatorial coordinates defined by the spherical
+unit vectors $\hat{\mathbf e}_n$, pointing North, and $\hat{\mathbf e}_e$
+pointing eastward such that the coordinate system is right-handed. The unit
+vectors are defined by
+\begin{align}
+    \hat{\mathbf e}_e & = \hat{\mathbf z}\times \hat{\mathbf n}  \\
+    \hat{\mathbf e}_n & = \hat{\mathbf n}\times\hat{\mathbf e}_e
+\end{align}
+The two components of $\mathbf s$ projected onto the plane of the sky are then
+\begin{align}
+    \zeta_e(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_e \\
+    \zeta_n(t;\alpha,\delta)\equiv \mathbf s\cdot \hat{\mathbf e}_n
+\end{align}
+In practice, we can retrieve $\mathbf{s}$ using NASA's JPL Horizons system and compute
+the projected separation at any time $t$.
+The angular positions of the source and the lens
+(Equations~\ref{eq:source_position} and \ref{eq:lens_position}) then become
+\begin{align}
+    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
+    +\pi_S\,\boldsymbol{\zeta}(t)                                                    \\
+    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
+    +\pi_L\,\boldsymbol{\zeta}(t)
+\end{align}
+where  $\pi_S\equiv 1\,\mathrm{au}/D_S$ is the source parallax, and $\pi_L$ is the
+lens parallax.
+The relative separation is then
+\begin{equation}
+    \boldsymbol{u}(t)= \frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
+    +\frac{t-t_0}{\theta_E}\boldsymbol{\mu}_{LS}+\pi_{E}\,\boldsymbol{\zeta}(t)
+    \label{eq:relative_separation_parallax}
+\end{equation}
+with $\pi_E\equiv\pi_{LS}/\theta_E\equiv (\pi_L-\pi_S)/\theta_E$.
+
+Since annual parallax is a higher order effect affecting the apparent
+trajectory of the lens on the sky, it makes sense to decompose the trajectory
+as a sum of rectilinear motion plus a deviation due to parallax. Following
+\citet{2004ApJ...606..319G} we define the position offset of the Sun on the
+plane of the sky relative to its position at a particular time $t_0'$ as
+
+\begin{equation}
+    \delta\boldsymbol \zeta (t)=\boldsymbol \zeta (t)-\boldsymbol \zeta (t_0')-(t-t_0')
+    \boldsymbol{\dot \zeta} (t_0')
+\end{equation}
+By construction, we have $\delta\boldsymbol \zeta (t_0')=0$ and
+$\delta\dot{\boldsymbol \zeta} (t_0')=0$.
+Equation~\ref{eq:relative_separation_parallax} then takes the form
+\begin{equation}
+    \boldsymbol{u}(t)=\mathbf{u}_0 + (t-t_0')\,\dot{\mathbf{u}}_0 +
+    \pi_E\,\delta\boldsymbol \zeta(t)
+    \label{eq:relative_separation_parallax_decomposed}
+\end{equation}
+where
+\begin{align}
+    \mathbf{u}_0       & \equiv \mathbf{u}(t_0) =\frac{\boldsymbol\theta_{LS, 0}}{\theta_E}
+    + \frac{(t_0'-t_0)}{\theta_E}\boldsymbol\mu_{LS}+
+    \pi_E\,\boldsymbol \zeta(t_0)                                                           \\
+    \dot{\mathbf{u}}_0 & \equiv\dot{\mathbf{u}}(t_0)=
+    \frac{\boldsymbol\mu_{LS}}{\theta_E}  + \pi_E\,\dot{\boldsymbol \zeta}
+    (t_0')
+\end{align}
+In what follows, we set the reference time to $t_0'\equiv t_0$ in order to ensure that
+the vectors $\mathbf{u}_0$ and $\dot{\mathbf{u}}_0$ are perpendicular to each other,
+which is the case at $t=t_0$ because at $t_0$ the trajectory has a local extremum
+\footnote{In CITE, $t_0'$ is fixed to a particular value. It is not clear
+    to me how this ensures that the two vectors are perpendicular to each other
+    (TODO: show that this is true).}
+
+\subsubsection{$\mathbf{u}(t)$ in the $(\mathbf{\hat e}_\bot,
+        \mathbf{\hat e}_\parallel)$ coordinate system}%
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.3\linewidth]{static/microlensing/microlensing_parallax_coordinates.pdf}
+    \caption{A coordinate system
+        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the lens trajectory at time $t_0$. The orange curve
+        represents the lens trajectory relative to the source star at the origin. The
+        coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to the equatorial coordinates $(\mathbf{\hat
+                e}_e,\mathbf{\hat e}_n)$ by a rotation through an angle $\psi$.}
+    \label{fig:parallax2}
+\end{figure}
+
+To evaluate Equation~\ref{eq:relative_separation_parallax_decomposed}, we need
+to choose a suitable basis. A natural coordinate system for describing the
+trajectory of the lens relative to the source is one defined by unit vectors
+$(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ where $\mathbf{\hat e}_\parallel$ is parallel to the trajectory $\mathbf{u}(t)$ at time $t_0$
+(Figure~\ref{fig:parallax2}). We define the unit vectors as
+\begin{equation}
+    \mathbf{\hat e}_\bot\equiv \frac{\mathbf{u}_0}{|\mathbf{u}_0|}\quad
+    \mathbf{\hat e}_\parallel\equiv \frac{\mathbf{\hat n}\times\mathbf{u}_0}{|\mathbf{u}_0|}\quad
+\end{equation}
+and the coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to ecliptic coordinates via a simple rotation through an
+angle $\psi$
+\begin{equation}
+    \begin{pmatrix}
+        \mathbf{\hat e}_\bot \\
+        \mathbf{\hat e}_\parallel
+    \end{pmatrix}
+    =
+    \begin{pmatrix}
+        \cos\psi & -\sin\psi \\
+        \sin\psi & \cos\psi
+    \end{pmatrix}
+    \begin{pmatrix}
+        \mathbf{\hat e}_e \\
+        \mathbf{\hat e}_n
+    \end{pmatrix}
+    \label{eq:ecliptic_to_parallel}
+\end{equation}
+By construction, at time $t_0$ we have
+$\mathbf{\dot u}_0\,\bot\,\dot{\mathbf{u}}_0$ and the two components of
+$\mathbf{u}(t)$ are then
+\begin{align}
+    u_\bot(t)      & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\bot= u_0 +
+    \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\bot                                                                                                                                       \\
+    u_\parallel(t) & \equiv \mathbf{u}(t)\cdot \mathbf{\hat e}_\parallel= (t-t_0)\,\dot{\mathbf{u}}_0\cdot\mathbf{\hat e}_\parallel+ \pi_E\,\delta\boldsymbol \zeta(t)\cdot\mathbf{\hat e}_\parallel
+\end{align}
+using the definitions of the unit vectors and
+$\delta\boldsymbol \zeta(t)=
+    \delta \zeta_e(t)\,\mathbf{\hat e}_e+\delta \zeta_n(t)\,\mathbf{\hat e}_n$, we have
+\begin{align}
+    u_\bot(t)      & = u_0 + \pi_E\,\cos\psi\,\delta \zeta_e(t) - \pi_E\,\sin\psi\,\delta \zeta_n(t)
+    \label{eq:u_t_parallel1}                                                                         \\
+    u_\parallel(t) & =(t-t_0)/t_E' + \pi_E\,\sin\psi\,\delta \zeta_e(t) +
+    \pi_E\,\cos\psi\,\delta \zeta_n(t) \label{eq:u_t_parallel2}
+\end{align}
+where $t_E' =|\dot{\mathbf{u}}_0|=|\boldsymbol\mu_{LS}/\theta_E  + \pi_E\,\dot{\boldsymbol \zeta}(t_0')|$.
+The model parameters which determine the magnification $A(t)$ are then
+$\left(u_0,t_0,t'_E,\pi_E,\psi\right)$.
+An alternative parametrization which is more commonly used is obtained by
+defining components of a "microlensing parallax vector" as
+\begin{align}
+    \pi_{E,N} & \equiv \pi_E\cos\psi \\
+    \pi_{E,E} & \equiv \pi_E\sin\psi
+\end{align}
+in which case we have
+\begin{align}
+    u_\bot(t)      & = u_0 + \pi_{E,N}\,\delta \zeta_e(t) - \pi_{E,E}\,\delta \zeta_n(t) \\
+    u_\parallel(t) & =(t-t_0)\omega_E + \pi_{E,E}\,\delta\zeta_e(t) +
+    \pi_{E,N}\,\delta\zeta_n(t)
+\end{align}
+and the model parameters are $\left(u_0,t_0,t_E',\pi_{E,E},\pi_{E,N}\right)$.
+
+For reference, we can also write down the components of $\mathbf{u}(t)$ in
+equatorial coordinates by inverting the rotation matrix in
+Equation~\ref{eq:ecliptic_to_parallel} and applying it to
+Equations~\ref{eq:u_t_parallel1} and \ref{eq:u_t_parallel2}, the two components
+are then
+\begin{align}
+    u_e & =u_0\cos\psi + (t-t_0)\omega_E\sin\psi + \pi_E\,\delta\zeta_e(t)  \\
+    u_n & =-u_0\sin\psi + (t-t_0)\omega_E\cos\psi + \pi_E\,\delta\zeta_n(t)
+\end{align}
+
+\subsubsection{$\mathbf{u}(t)$ in the $(\mathbf{\hat e}_\parallel, \mathbf{\hat e}_\bot)$ coordinates using acceleration parameters} There
+exist another parametrization of the trajectory in which we fit for the local
+acceleration of the lens at $t_0$ instead of $\pi_E$ and the angle $\psi$. We
+define the position, velocity and acceleration such that at $t=t_0$ we have
+\begin{align}
+    \mathbf{u}(t_0)       & \equiv \widetilde{u}_0 \,\hat{\mathbf e}_\bot
+    \label{eq:u_primed_def}                                               \\
+    \dot{\mathbf{u}}(t_0) & \equiv \frac{1}{\widetilde{t_E'}}\,
+    \hat{\mathbf e}_\parallel \label{eq:u_dot_primed_def}                 \\ \ddot{\mathbf{u}}(t_0) & \equiv
+       a_\bot\,\hat{\mathbf e}_\bot + a_\parallel\,\hat{\mathbf e}_\parallel \label{eq:u_ddot_primed_def}
+\end{align}
+where $a_\parallel$ and $a_\bot$ are the two components of the instantaneous acceleration of
+the lens at $t_0$. From Equations~\ref{eq:u_t_parallel1} and
+\ref{eq:u_t_parallel2} it follows that
+\begin{align}
+    \mathbf{u}(t_0)       & = u_0\,\hat{\mathbf e}_\bot                    \\
+    \dot{\mathbf{u}}(t_0) & =\frac{1}{t_E'}\,\hat{\mathbf e}_\parallel     \\ \ddot{\mathbf{u}}(t_0) &
+       =\left[\pi_E\cos\psi\,\delta\ddot{\zeta}_e(t_0)
+    -\pi_E\sin\psi\,\delta\ddot{\zeta}_n(t_0)\right]\,\hat{\mathbf e}_\bot \\  &
+       +\left[\pi_E\sin\psi\,\delta\ddot{\zeta}_e(t_0)
+           +\pi_E\cos\psi\,\delta\ddot{\zeta}_n(t_0)\right]\,\hat{\mathbf e}_\parallel
+\end{align}
+By equating the components of the position, velocity and acceleration vectors, we obtain the expressions for the old parameters in terms of the new parameters as
+\begin{equation}
+    \widetilde{u}_0=u_0,\quad \widetilde{t_E'}=t_E'\,\quad
+    \pi_E=\sqrt{\frac{a_\parallel^2 + a_\bot^2}{1} \left[\delta\ddot{\boldsymbol{\zeta}}_e(t_0)\right]^2 +
+        \left[\delta\ddot{\boldsymbol{\zeta}}_n(t_0)\right]^2}
+\end{equation}
+Similarly, we have
+\begin{align}
+    \pi_{E,N}=\pi_E\cos\psi & = \frac{a_\parallel\,\delta\ddot{\zeta}_n(t_0) + a_\bot\,\delta\ddot{\zeta}_e(t_0)
+    }{[\delta\ddot{\zeta}_e(t_0)]^2 + [\delta\ddot{\zeta}_n(t_0)]^2}                                              \\
+    \pi_{E,E}=\pi_E\sin\psi & = \frac{ a_\parallel\,\delta\ddot{\zeta}_e(t_0) - a_\bot\,\delta\ddot{\zeta}_n(t_0)
+    }{[\delta\ddot{\zeta}_e(t_0)]^2 + [\delta\ddot{\zeta}_n(t_0)]^2}
+\end{align}
+To obtain the trajectory in terms of these new parameters, we plug in the above expressions
+into Equations~\ref{eq:u_t_parallel1} and \ref{eq:u_t_parallel2}.
+The new parameter set is then $\left(u_0,t_0,t_E',a_\parallel,a_\bot\right)$. This parametrization makes it obvious that the parallax
+effect depends on the apparent local acceleration of the lens at time $t_0$.
+
+%\subsection{Binary and triple lens}
+%Microlensing with more than one lens is considerably more complex than the case
+%of a single lens. Consider a system with $N$ point mass lenses with a total
+%mass $M=\sum_{i=1}^Nm_i$. As before, we have the angular source position in the
+%source plane, given by the the dimensionless vector
+%$\mathbf{y}=\bm\beta/(D_\textrm{S}\theta_E)$ and the position of the images in
+%the lens plane, given by given by
+%$\mathbf{x}=\bm\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now refers to
+%the angular Einstein radius corresponding to the total mass $M$. The lens
+%equation then contains a sum over the deflection angles $\bm\alpha_i$
+%corresponding to each point mass $m_i$\footnote{This is because we working
+%    within the linearized GR framework so the deflection angles are additive.}:
+%\begin{equation}
+%    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\bm\alpha_i(\mathbf{x},\mathbf{x}_i)
+%\end{equation}
+%where $\bm\alpha_i$ is the deflection angle due to the $i$th lens.
+%Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
+%\begin{equation}
+%    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
+%    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
+%    \label{eq:lens_equation_general}
+%\end{equation}
+%where $\epsilon_i\equiv m_i/M$.
+
+%It is useful to rewrite Equation~\ref{eq:lens_equation_general} in complex form
+%by introducing complex variables \citep{1990A&A...236..311W}
+%\begin{equation}
+%    w= y_1+iy_2,\quad z=x_1+ix_2
+%\end{equation}
+%The lens equation then takes the form
+%\begin{equation}
+%    w=z-\sum_{i=1}^N \frac{\epsilon_i}{\bar z-\bar z_i}
+%    \label{eq:lens_equation_complex}
+%\end{equation}
+%The mapping from the image plane $z$ to the source plane $w$ is one-to-one and it is
+%straightforward to compute using Equation~\ref{eq:lens_equation_complex}. The inverse
+%mapping is not as easy. It turns out that it is possible to rewrite Equation~\ref{eq:lens_equation_complex}
+%as a complex polynomial of degree $N^2 + 1$ (see Appendix~\ref{app:complex_poly}).
+
 \subsection{Solving the lens equation}
@@
 \chapter{First appendix}
+\label{app:complex_poly}
+
 \chapter{Second appendix}

### 2022-05-25 src/figures/paczynski_curve.py

@@ -1,0 +1,54 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+tau = np.linspace(-1.1, 1.1, 1000)
+
+u0 = np.array([0.1, 0.3, 0.5, 0.7, 0.9, 1.1])
+colors = [
+    "C0",
+]
+
+fig, ax = plt.subplots(figsize=(10, 8))
+
+for i, u0_ in enumerate(u0):
+    u = np.sqrt(u0_ ** 2 + tau ** 2)
+    A = (u ** 2 + 2) / (u * np.sqrt(u ** 2 + 4))
+    color = "C" + str(i)
+    ax.plot(tau, A, color, label=r"$u_0=" + str(u0_) + "$")
+
+# Inset plot
+a = plt.axes([0.15, 0.5, 0.32, 0.32])
+# a.grid(True)
+a.set_xlim(-1.2, 1.2)
+a.set_ylim(-1.2, 1.2)
+circle = plt.Circle(
+    (0, 0), 1.0, facecolor="white", edgecolor="k", linestyle="dashed", lw=1
+)
+a.add_artist(circle)
+circle2 = plt.Circle((0, 0), 0.03, facecolor="black")
+a.add_artist(circle2)
+a.set_aspect(1)
+x_ = np.linspace(0, 0.79)
+a.plot(x_, 0.85 * x_, "k--", lw=1)
+a.scatter([0.0], [0.0], marker="o", color="black")
+
+plt.text(-0.8, -1.3, r"source trajectory", fontsize=16)
+plt.text(0.8, 0.7, r"$\theta_E$", fontsize=16)
+plt.text(-0.2, 0.1, r"L", fontsize=16)
+plt.axis("off")
+
+for i, u0_ in enumerate(u0):
+    u = np.sqrt(u0_ ** 2 + tau ** 2)
+    A = (u ** 2 + 2) / (u * np.sqrt(u ** 2 + 4))
+    color = "C" + str(i)
+    a.axhline(-u0_, color=color)
+
+ax.grid(True)
+ax.set_xlabel(r"$\tau=(t-t_0)/t_E$")
+ax.set_ylabel(r"$A(u)$")
+ax.set_xticks([-1.0, -0.5, 0.0, 0.5, 1.0])
+ax.legend(prop={"size": 16})
+
+# Save to disk
+fig.savefig("paczynski_curve.pdf", bbox_inches="tight")

### 2022-05-25 src/figures/single_lens_images.py

@@ -1,0 +1,76 @@
+import numpy as np
+import jax.numpy as jnp
+from jax import jit, vmap
+
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+
+
+def mag(z):
+    u = jnp.abs(z) ** 2
+    return (u ** 2 + 2) / (u * jnp.sqrt(u ** 2 + 4))
+
+
+@jit
+def brightness_profile(z, rho, w_center, u=0.1):
+    w = z - 1 / jnp.conjugate(z)
+    r = jnp.abs(w - w_center) / rho
+
+    B_r = jnp.where(r <= 1.0, jnp.sqrt(1 - r ** 2), 0.0,)
+    I = 3.0 / (3.0 - u) * (u * B_r + 1.0 - u)
+    return I
+
+
+# Source plane
+xgrid, ygrid = jnp.meshgrid(np.linspace(-1, 1, 200), jnp.linspace(-1, 1, 200))
+wgrid = xgrid + 1j * ygrid
+
+mag = vmap(vmap(mag))(wgrid)
+
+# Lens plane
+rho = 0.1
+w_center = 0.2
+
+l = 1.5
+xgrid_im, ygrid_im = jnp.meshgrid(np.linspace(-l, l, 500), jnp.linspace(-l, l, 500))
+zgrid = xgrid_im + 1j * ygrid_im
+I_eval = brightness_profile(zgrid, rho, w_center, u=0.2)
+
+fig, ax = plt.subplots(2, 3, figsize=(15, 10), gridspec_kw={"hspace": 0.4})
+
+for a in ax[0]:
+    a.pcolormesh(
+        xgrid, ygrid, mag, cmap="Greys", norm=colors.LogNorm(vmax=1e04), zorder=-1
+    )
+
+rho = 0.15
+w1, w2, w3 = -0.4, -0.2, 0.0
+
+I_eval = brightness_profile(zgrid, rho, w1, u=0.2)
+ax[1, 0].pcolormesh(xgrid_im, ygrid_im, I_eval, cmap="Greys", zorder=-1)
+
+I_eval = brightness_profile(zgrid, rho, w2, u=0.2)
+ax[1, 1].pcolormesh(xgrid_im, ygrid_im, I_eval, cmap="Greys", zorder=-1)
+
+I_eval = brightness_profile(zgrid, rho, w3, u=0.2)
+ax[1, 2].pcolormesh(xgrid_im, ygrid_im, I_eval, cmap="Greys", zorder=-1)
+
[…]

### 2022-05-30 src/ms.tex
Finish writing the microlensing part of Chapter 2 (for now)

@@
 % Define document class
-\documentclass[11pt]{report}
+\documentclass[12pt]{report}
 \usepackage{aas_macros}
@@
 Gravitational lensing as a method for discovering exoplanets really took off
-with the work of Paczyński
+with the work of Paczy\'nski
 \citep{1986ApJ...304....1P,1986ApJ...301..503P,1991ApJ...374L..37M} who also
@@
 equation. If the source and the lens are perfectly aligned with respect to the observer,
-then $\beta=0$ and we have
-\begin{equation}
-    \theta \equiv \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}
-\end{equation}
-The distance term $D_L^{-1}-D_\textrm{s}^{-1}$ is proportional to the
-lens-source parallax.
+then $\beta=0$ and $\theta\equiv\theta_E$ where
+\begin{equation}
+    \theta_E= \sqrt{ \frac{4GM}{c^2} \left( \frac{1}{D_L} - \frac{1}{D_S} \right)}=
+    \sqrt{\kappa M\pi_{LS}}
+    \label{eq:angular_einstein_radius}
+\end{equation}
+where $\kappa \equiv \frac{4 G}{c^{2} \mathrm{au}} \simeq 8.1 \frac{\mathrm{mas}}{M_{\odot}}$
+and
+\begin{equation}
+    \pi_{LS}\equiv\pi_L - \pi_S=\frac{1\mathrm{au}}{D_L}-\frac{1\mathrm{au}}{D_S}
+\end{equation}
+is the relative lens-source parallax.
 
@@
 away the source is from the lens.
-
-%The angular Einstein radius determines a characteristic timescale
-%$t_E=\theta_E/|\boldsymbol{\mu}_{LS}|$ on which the magnification of of the
-%source changes, where $\boldsymbol{\mu}_{LS}$ is the relative proper motion
-%vector between the source and the lens
-%$\boldsymbol{\mu}_{LS}\equiv\boldsymbol{\mu}_L- \boldsymbol{\mu}_S$. The
-%predicted flux of a star undergoing a microlensing event is then given by the
-%following expression
-%\begin{equation}
-%    f=F_SA(t)+F_B
-%    \label{eq:flux_paczynski}
-%\end{equation}
-%where $F_S$ is the intrinsic flux of the star and
-%$F_B$ is any background flux due to other stars in the field which
-%might be blended with the source star, usually called the \emph{blending flux}.
-%$A(t)$ is a non-linear time-dependant magnification mapping which depends
-%on the properties of the lens and the geometry of the lensing event.
-%In the simplest case, that of a point source being magnified by a point lens,
-%the magnification is given by
-%\begin{equation}
-%    A(t)= \frac{u(t)^2+2}{u(t)\sqrt{u(t)^2+4}}
-%\end{equation}
-%where $u(t)$ is the separation between the lens and the source expressed in
-%units of the angular Einstein radius.
 
@@
 and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
-The model for the light curve is then characterized by
-the parameters $\{F_S,F_B, t_0, u_0, t_E\}$ and is known in the
-literature as the Paczy\'nski curve.
-
+The magnification is then a function of three parameters $(t_0, u_0, t_E)$
+and the resulting curve as a function of time is often called the Paczy\'nski curve
+\citep{1986ApJ...304....1P,1986ApJ...301..503P}.
 The magnification as a function of time for different impact parameters $u_0$
-is shown in Figure~\ref{fig:single_lens_mag}. Notice the very steep fall off in
+is shown in Figure~\ref{fig:paczynski_curve}. Notice the very steep fall off in
 magnification as as we move away from the Einstein ring ($A(u)\propto
-    1+2u^{-4}$). The curves shown in Figure~\ref{fig:paczynski_curve} are often
-called \emph{Paczyński curves} \citep{1986ApJ...304....1P,1986ApJ...301..503P},
-in honour of the late Polish astronomer Bohdan Paczyński, one of the key
-figures of gravitational microlensing.
+    1+2u^{-4}$).
+In the limits of $u_0\rightarrow 0$ and $u_0\gg 1$  there is a continuous
+mathematical degeneracy between the parameters $t_0$, $u_0$ and $t_E$
+\citep{1997ApJ...487...55W}.
 
@@
             Magnification of a point source by a point lens as a function of time. The
-            various magnification curves correspond to different impact parameters $u_0$. The inset figure
+            various magnification curves correspond to different impact parameters $u_0$.
+            The inset figure
             shows the source trajectories in the lens plane, the dashed circle corresponds
@@
 \end{figure}
+
+\subsection{Observed flux}
+The magnification $A(t)$ is not a direct observable in microlensing events. We
+can only measure the flux of the source star which is usually contaminated with
+the flux of other nearby stars in the dense stellar field and potentially also
+with light from the lens and possible companions to the lens. The observed flux
+is thus given by
+\begin{equation}
+    F(t)=F_S(t)+F_B(t)
+    \label{eq:flux_observed}
+\end{equation}
+where $F_s$ is the flux of the source star and $F_B$ is the contamination flux also
+called the \emph{blending flux}. To quantify the strength of the blending e can
+define the emph{source flux fraction} $b_S$ as the fraction of total (unmagnified)
+flux that is coming from the source star:
+\begin{equation}
+    b_S\equiv F_S/(F_S + F_B)
+\end{equation}
+For highly blended events $b_S\approx 0$ and in absence of blending $b_S=1$.
+The source flux $F_S$ and the blending flux $F_B$ are generally highly correlated
+and it is often preferable to use a slightly different parametrization proposed by
+\citet{2009MNRAS.393..816D}. In this parametrization we use the
+\emph{baseline flux} $F_\mathrm{base}$ and the difference between
+the baseline flux and the flux at peak magnification $\Delta F\equiv F_S[A(u_0)-1]$
+as the new parameters.
+Equation~\ref{eq:flux_observed} then takes the form
+\begin{equation}
+    f=\Delta F\frac{A(t) - 1}{A(t_0)-1}+F_\mathrm{base}
+    \label{eq:flux_dominik}
+\end{equation}
+It is generally straightforward to measure the flux difference $\Delta F$, the
+baseline flux $F_\mathrm{base}$ and the center of the curve $t_0$ but getting a
+good estimate of $t_E$ even outside of the regime $u_0\rightarrow 0$ and
+$u_0\gg 1$ requires good photometric sampling along the ``wings'' of the light
+curve \citep{2009MNRAS.393..816D}.
 
@@
 
-\subsection{Parallax}
+\subsection{Annual parallax}
+\label{ssec:single_lens_parallax}
 In Equation~\ref{eq:relative_trajectory_no_parallax} we have assumed that the
 relative motion of the lens with respect to the source on the plane of the sky
-is rectilinear. This is a reasonable approximation in a barycentric frame of
+is rectilinear.
+This is a reasonable approximation in a barycentric frame of
 reference if the acceleration of the source and the lens can be neglected.
 However, most microlensing events are observed from a non-inertial geocentric
-frame so that the observed trajectory is modified due the local acceleration of
-the Earth and this effect needs to be taken into account in a realistic model.
-In order to include the parallax effect, we need to modify
-Equations~\ref{eq:source_position} and \ref{eq:lens_position} to include the
-projected motion of the Sun relative to the Earth on the plane of the sky. We
-project $\mathbf s$ -- the position vector of the Sun relative to Earth, onto a
-plane perpendicular to the line of sight towards the source star (plane of the
-sky) which is defined by the unit vector $\boldsymbol{\hat n}$ normal to the
-plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky coordinates of
-the source star $(\alpha,\delta)$ (right ascension and declination).
-
-We choose to work in geocentric equatorial coordinates defined by the spherical
-unit vectors $\hat{\mathbf e}_n$, pointing North, and $\hat{\mathbf e}_e$
-pointing eastward such that the coordinate system is right-handed. The unit
-vectors are defined by
+frame (Earth) and sometimes also from multiple locations simultaneously. In those
+cases we have to take into account parallax effects.
+We differentiate between two kinds of parallax. First is the \emph{annual
+    parallax} (sometimes also called the orbital parallax) which modifies the
+lens-source relative motion vector $\boldsymbol\mu_{LS}$ due to the local
+acceleration of Earth in its orbit. It is important for long timescale events
+when the event timescale is equal to some substantial fraction of a year. The
+effect of annual parallax is usually a slight modification of the shape of the
+Paczy\'nski curve.
+
+The second kind of parallax is the \emph{satellite parallax} which refers to a
+difference in viewpoint when conducting simultaneous observations of the source
+star from different locations separated by a substantial fraction of the
+Einstein ring projected onto the observer plane $\tilde{r}_E\equiv
+    D_{LS}\theta_E$ (where $D_{LS}^{-1}=D_L^{-1} - D_S^{-1}$). In practice this
+means observing a microlensing event from Earth and a space based telescope
+simultaneously. It was first proposed by \citet{1966MNRAS.134..315R}. There is
+also the ``terrestrial parallax'' which is identical to the satellite parallax
+except that it involves only ground-based observations separated by a large
+fraction of Earth's diameter. A measurement of these parallax effects allows
+for a partial breaking of the degeneracy in the event timescale $t_E$ by
+providing a relationship between the mass and the distance to the lens
+(assuming one can estimate the distance to the source star).
+
+In this section I will describe the annual parallax effect. We start by
+modifying Equations~\ref{eq:source_position} and \ref{eq:lens_position} to
+include the projected motion of the Sun relative to the Earth on the plane of
+the sky. We project $\mathbf s$ -- the position vector of the Sun relative to
+Earth, onto a plane perpendicular to the line of sight towards the source star
+(plane of the sky) which is defined by the unit vector $\boldsymbol{\hat n}$
+normal to the plane. The unit vector $\boldsymbol{\hat n}$ depends on the sky
+coordinates of the source star $(\alpha,\delta)$ (right ascension and
+declination). We choose to work in geocentric equatorial coordinates defined by
+the spherical unit vectors $\hat{\mathbf e}_n$, pointing North, and
+$\hat{\mathbf e}_e$ pointing eastward such that the coordinate system is
+right-handed. The unit vectors are defined by
 \begin{align}
@@
 \end{equation}
-with $\pi_E\equiv\pi_{LS}/\theta_E\equiv (\pi_L-\pi_S)/\theta_E$.
+where
+\begin{equation}
+    \pi_E\equiv \frac{\pi_{LS}}{\theta_E}
+    \label{eq:pi_E}
+\end{equation}
 
 Since annual parallax is a higher order effect affecting the apparent
-trajectory of the lens on the sky, it makes sense to decompose the trajectory
+trajectory of the source on the sky, it makes sense to decompose the trajectory
 as a sum of rectilinear motion plus a deviation due to parallax. Following
@@
     \caption{A coordinate system
-        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the lens trajectory at time $t_0$. The orange curve
-        represents the lens trajectory relative to the source star at the origin. The
+        $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ parallel to the source trajectory at time $t_0$. The orange curve
+        represents the source trajectory relative to the lens at the origin. The
         coordinate system $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ is related to the equatorial coordinates $(\mathbf{\hat
@@
 to choose a suitable basis. A natural coordinate system for describing the
-trajectory of the lens relative to the source is one defined by unit vectors
+trajectory of the source relative to the lens is one defined by unit vectors
 $(\mathbf{\hat e}_\bot,\mathbf{\hat e}_\parallel)$ where $\mathbf{\hat e}_\parallel$ is parallel to the trajectory $\mathbf{u}(t)$ at time $t_0$
@@
 \begin{equation}
-    \mathbf{\hat e}_\bot\equiv \frac{\mathbf{u}_0}{|\mathbf{u}_0|}\quad
+    \mathbf{\hat e}_\bot\equiv \frac{\mathbf{u}_0}{|\mathbf{u}_0|}\quad,
     \mathbf{\hat e}_\parallel\equiv \frac{\mathbf{\hat n}\times\mathbf{u}_0}{|\mathbf{u}_0|}\quad
@@
 By construction, at time $t_0$ we have
-$\mathbf{\dot u}_0\,\bot\,\dot{\mathbf{u}}_0$ and the two components of
+$\mathbf{u}_0\,\bot\,\dot{\mathbf{u}}_0$ and the two components of
 $\mathbf{u}(t)$ are then
@@
 The model parameters which determine the magnification $A(t)$ are then
-$\left(u_0,t_0,t'_E,\pi_E,\psi\right)$.
+$\left(u_0,t_0,t'_E,\pi_E,\psi\right)$. Notice that $u_0$ in this case can also be negative.
+
 An alternative parametrization which is more commonly used is obtained by
@@
     u_\bot(t)      & = u_0 + \pi_{E,N}\,\delta \zeta_e(t) - \pi_{E,E}\,\delta \zeta_n(t) \\
-    u_\parallel(t) & =(t-t_0)\omega_E + \pi_{E,E}\,\delta\zeta_e(t) +
+    u_\parallel(t) & =(t-t_0)/t_E' + \pi_{E,E}\,\delta\zeta_e(t) +
     \pi_{E,N}\,\delta\zeta_n(t)
@@
 \begin{align}
-    u_e & =u_0\cos\psi + (t-t_0)\omega_E\sin\psi + \pi_E\,\delta\zeta_e(t)  \\
-    u_n & =-u_0\sin\psi + (t-t_0)\omega_E\cos\psi + \pi_E\,\delta\zeta_n(t)
+    u_e & =u_0\cos\psi + (t-t_0)/t_E'\sin\psi + \pi_E\,\delta\zeta_e(t)  \\
+    u_n & =-u_0\sin\psi + (t-t_0)/t_E'\cos\psi + \pi_E\,\delta\zeta_n(t)
 \end{align}
@@
 
-%\subsection{Binary and triple lens}
-%Microlensing with more than one lens is considerably more complex than the case
-%of a single lens. Consider a system with $N$ point mass lenses with a total
-%mass $M=\sum_{i=1}^Nm_i$. As before, we have the angular source position in the
-%source plane, given by the the dimensionless vector
-%$\mathbf{y}=\bm\beta/(D_\textrm{S}\theta_E)$ and the position of the images in
-%the lens plane, given by given by
-%$\mathbf{x}=\bm\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now refers to
-%the angular Einstein radius corresponding to the total mass $M$. The lens
-%equation then contains a sum over the deflection angles $\bm\alpha_i$
-%corresponding to each point mass $m_i$\footnote{This is because we working
-%    within the linearized GR framework so the deflection angles are additive.}:
-%\begin{equation}
-%    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\bm\alpha_i(\mathbf{x},\mathbf{x}_i)
-%\end{equation}
-%where $\bm\alpha_i$ is the deflection angle due to the $i$th lens.
-%Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
-%\begin{equation}
-%    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
-%    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
-%    \label{eq:lens_equation_general}
-%\end{equation}
-%where $\epsilon_i\equiv m_i/M$.
-
-%It is useful to rewrite Equation~\ref{eq:lens_equation_general} in complex form
-%by introducing complex variables \citep{1990A&A...236..311W}
-%\begin{equation}
-%    w= y_1+iy_2,\quad z=x_1+ix_2
-%\end{equation}
-%The lens equation then takes the form
-%\begin{equation}
-%    w=z-\sum_{i=1}^N \frac{\epsilon_i}{\bar z-\bar z_i}
-%    \label{eq:lens_equation_complex}
-%\end{equation}
-%The mapping from the image plane $z$ to the source plane $w$ is one-to-one and it is
-%straightforward to compute using Equation~\ref{eq:lens_equation_complex}. The inverse
-%mapping is not as easy. It turns out that it is possible to rewrite Equation~\ref{eq:lens_equation_complex}
-%as a complex polynomial of degree $N^2 + 1$ (see Appendix~\ref{app:complex_poly}).
-
-\subsection{Solving the lens equation}
+\subsection{Measuring the lens mass}
+Notice that only quantity of physical interest in single lens microlensing --
+the lens mass $M$, is buried inside of the definition of the angular Einstein
+radius $\theta_E$ which, when combined with the magnitude of the relative
+proper motion $|\boldsymbol\mu_{LS}|$ forms the observable $t_E$. In practice,
+depending on the microlensing event there are several possible channels which
+enable an estimate of the lens mass. For example, a measurement of $\theta_E$
+from finite-source effects or from a direct measurement of
+$|\boldsymbol\mu_{LS}|$ can be combined with a measurement of $\pi_E$ to yield
+(via Equations \ref{eq:angular_einstein_radius} and \ref{eq:pi_E})
+\begin{equation}
+    M = \frac{\theta_E}{\kappa \pi_E}
+\end{equation}
+With an estimate of the distance to the source star one can also obtain the distance
+to the lens.
+An alternative, less direct approach is to use a measurement of $\theta_E$ combined
+with the measurement of the lens flux to estimate the mass and distance to the lens
+\citep{2007ApJ...660..781B}.
+It should be noted that either of these approaches is possible for only a small subset
+of all detected microlensing events.
+
+\subsection{A system with $N$ lenses}
+Microlensing with more than one lens is considerably more complex than the case
+of a single lens. Consider a system with $N$ point mass lenses with a total
+mass $M=\sum_{i=1}^Nm_i$. As before, we have the angular source position in the
+source plane, given by the the dimensionless vector
+$\mathbf{y}=\bm\beta/(D_\textrm{S}\theta_E)$ and the position of the images in
+the lens plane, given by given by
+$\mathbf{x}=\bm\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now refers to
+the angular Einstein radius corresponding to the total mass $M$. The lens
+equation then contains a sum over the deflection angles $\bm\alpha_i$
+corresponding to each point mass $m_i$\footnote{This is because we are working
+    within the linearized GR framework so the deflection angles are additive.}:
+\begin{equation}
+    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\bm\alpha_i(\mathbf{x},\mathbf{x}_i)
+\end{equation}
+where $\bm\alpha_i$ is the deflection angle due to the $i$th lens.
+Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
+\begin{equation}
+    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
+    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
+    \label{eq:lens_equation_general}
+\end{equation}
+where $\epsilon_i\equiv m_i/M$.
+
+It is useful to rewrite Equation~\ref{eq:lens_equation_general} in complex form
+by introducing complex variables \citep{1990A&A...236..311W}
+\begin{equation}
+    w= y_1+iy_2,\quad z=x_1+ix_2
+\end{equation}
+The lens equation then takes the form
+\begin{equation}
+    w=z-\sum_{i=1}^N \frac{\epsilon_i}{\bar z-\bar z_i}
+    \label{eq:lens_equation_complex}
+\end{equation}
+The mapping from the image plane $z$ to the source plane $w$ is one-to-one and it is
+straightforward to compute using Equation~\ref{eq:lens_equation_complex}. The inverse
+mapping is not as easy. It turns out that it is possible to rewrite Equation~\ref{eq:lens_equation_complex}
+as a complex polynomial of degree $N^2 + 1$ (see Appendix~\ref{app:complex_poly}).
+From the Fundamental Theorem of Algebra, it follows that such a polynomial has exactly $N^2+1$ complex roots.
+However, not all of these roots are necessarily solutions to the lens equations so in practice one first has to
+compute all of the polynomial roots and then discard those which do not satisfy
+Equation~\ref{eq:lens_equation_complex}.
+In the case of binary lens, there are always either 3 or 5 real images and for $N\geq 2$ the maximum number
+of images is $5(N-1)$ \citep{arXiv:astro-ph/0103463,astro-ph/0305166,arXiv:math/0401188v2}.
+As before, the magnification is given by the inverse Jacobian determinant of the mapping $z\rightarrow w$
+evaluated at the images. We have
+\begin{equation}
+    \mathbf{J}=
+    \renewcommand\arraystretch{2}
+    \begin{pmatrix}
+        \frac{\partial w}{\partial z} & \frac{\partial w}{\partial \bar{z}} \\ \frac{\partial \bar{w}}{\partial z} & \frac{\partial \bar{w}}{\partial \bar{z}}
+    \end{pmatrix}
+\end{equation}
+and
+\begin{equation}
+    \mathrm{det}\,\mathbf{J}=\frac{\partial w}{\partial z}\frac{\partial \bar{w}}{\partial \bar{z}} - \frac{\partial w}{\partial \bar{z}}\frac{\partial \bar{w}}{\partial z} = \left|\frac{\partial w}{\partial z}\right|^2 - \left|\frac{\partial w}{\partial \bar{z}}\right|^2
+\end{equation}
+where we have used the identities
+$\overline{\frac{\partial\bar a}{\partial{\bar b}}}=\frac{\partial a}{\partial{b}}$ and $a\bar a=|a|^2$. Finally, by evaluating the partial derivatives
+using Equation~\ref{eq:lens_equation_complex} we obtain
+\begin{equation}
+    \mathrm{det} \,\mathbf J=1-\left|\sum_{i=0}^{N} \frac{\epsilon_{i}}{\left(\bar{z}-\bar{z}_i\right)^{2}}\right|^{2}
+\end{equation}
+so the magnification is given by
+\begin{equation}
+    A = \sum_j \frac{1}{\left|\mathrm{det}\,\mathbf J\right|_j}
+\end{equation}
+where $j$ denotes the $j$-th image. The points in the image plane where the Jacobian determinant vanishes
+($\mathrm{det}\,\mathbf J=0$) are the critical curves:
+\begin{equation}
+    \left|\sum_{i=0}^{N} \frac{\epsilon_{i}}{\left(\bar{z}-\bar{z}_i\right)^{2}}\right|^{2}=1
+    \label{eq:critical_curves_complex}
+\end{equation}
+The points on the critical curve mapped to the source plane via
+the lens equation (Equation~\ref{eq:lens_equation_complex}) are the caustic curves.
+The difference compared to the case of a single lens lens equation is that the caustics are closed curves
+rather than a single point. The caustic curves are comprised of concave segments called \emph{folds} which are connected
+at points called \emph{cusps}.
+Equation~\ref{eq:critical_curves_complex} can also be written as a complex polynomial, of
+degree $2N$ (see Appendix~\ref{app:complex_poly_crit}) so there are at most $2N$ critical and caustic curves.
+The complexity of microlensing events involving multiple lenses is a direct consequence of the
+non-smooth nature of caustic curves.
+
+\subsection{Binary lens}
+In this section we will focus specifically on the binary lens. We choose a
+coordinate system whose origin is at the midpoint of the line connecting the
+two lenses and both lenses are on the real axis. If the first lens is at
+distance $a$ ($z_1=a$) then the second lens is at $-a$ and the lens equation
+takes the form
+\begin{equation}
+    w=z+\frac{\epsilon_{1}}{\bar{z} - a}+\frac{1 - \epsilon_{1}}{\bar{z} + a}
+\end{equation}
+The above equation can be rewritten as a 5th order complex polynomial
+(see Appendix~\ref{app:complex_poly}). Either 3 or 5 of the complex roots of that
+polynomial are also solutions to the lens equation
+(5 inside the caustics and 3 outside).
+It is common to use the mass ratio $q\equiv \epsilon_2/\epsilon_1$ with
+$\epsilon_2 \leq \epsilon_1$ instead  of $\epsilon_1$ as a parameter and the
+separation between the lenses $s\equiv 2a$ instead of half the separation $a$.
+As a reminder all angular quantities are expressed in the units of the angular Einstein
+radius of the total mass of the system.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.8\linewidth]{figures/binary_lens_topology.pdf}
+    \caption{The three different topologies of the binary lens. The left panels show the
+        critical curves in the lens plane, the right panels show the caustic curves in the
+        source plane. The figure shows all three topologies and the transitions between
+        them for an equal mass binary lens with $q=1$. In this case the transition between
+        the close and intermediate topology occurs at $s=1/\sqrt{2}$ and the one between
+        the intermediate and wide topology at $s=2.0$. Figure adapted from
+        \citet{dominik1999}.
+    }
+    \label{fig:binary_lens_topology}
+\end{figure}
+
+Depending on the separation between the two lenses, $s$, the critical and the
+caustic curves can have three distinct topologies which are labeled
+\emph{close}, \emph{intermediate}, and \emph{wide}. These are shown in
+Figure~\ref{fig:binary_lens_topology} for a system with $q=1$ where on the left
+we see the critical curves in the lens plane, and on the right the caustic
+curves in the source plane. The critical and caustic curves are symmetric with
+respect to the x-axis. The sharp structure of the caustic curves compared to
+the smooth critical curves is due to the nonlinearity of the lens mapping. The
+number of different topologies does not depend on the mass ratio.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{figures/planetary_caustics.pdf}
+    \caption{Caustic structure for a binary lens with $q=0.003$, shown for different
+        values of the separation $s$. The orange star denotes the position of the star and
+        the blue dot designates the planet.
+        The transition between the close and intermediate
+        topologies occurs at $s\approx 0.92$ and the one between intermediate and wide
+        topologies at $s\approx 1.21$. The vertical gray dashed line shows the angular
+        Einstein radius at $y_1=1$. }
+    \label{fig:planetary_caustics}
+\end{figure}
+
+Caustics shown in Figure~\ref{fig:binary_lens_topology} corresponds to an equal
+mass binary lens, systems with $q\ll 1$ are of more interest because these
+correspond to planetary systems. The lower mass ratio does not change the
+topological structure of the caustics (except for shifting the boundaries
+between the different topologies) but it does change their shape and size. In
+Figure~\ref{fig:planetary_caustics} we show point source \emph{magnification
+    maps} (computed using the \textsf{caustics} code which is the subject of
+Chapter~\ref{ch:microlensing}) for a binary lens with $q=5\times 10^{-3}$,
+roughly corresponding to a gas giant planet orbiting an M dwarf. For the close
+topology (first two panels from the top in Figure~\ref{fig:planetary_caustics})
+we see one caustic centered on the star which is called a \emph{central
+    caustic} and two additional caustics on the opposite side of the planet,
+symmetric with respect to the star-planet axis, which are called
+\emph{planetary caustics} because they are associated with the planet rather
+than the star. Notice that the magnification in the vicinity of these planetary
+caustics is significantly smaller than the magnification around the central
+caustic. Because of this it is easier to search for binary events with
+trajectory passing close to the central caustic rather than the planetary
+caustics. As the separation $s$ approaches $s_c$, the critical value between
+the close and wide topologies, the two planetary caustics merge with the
+central caustic into a single caustic, often called the \emph{resonant
+    caustic}, with a much larger cross section than either the central or the
+planetary caustics (third panel from the top). Finally, as the separation $s$
+increases further beyond the intermediate/wide boundary, the resonant caustic
+separates into a smaller central caustic centered on the star and a single,
+larger, planetary caustic located on the star-planet axis (bottom panel). The
+size of this planetary caustic scales as $q^{1/2}s^{-2}$ \citep[see references
+    in review by][]{Gaudi2012}.
+
+Given caustics such as the ones shown in Figure~\ref{fig:planetary_caustics}
+the magnification as a function of time depends on the source trajectory in the
+source plane. Absent parallax, it is just a slice through the magnification map
+convolved with the source star brightness profile. \textbf{Herein lies the
+    complexity of microlensing. The trajectory of the source star over the caustic
+    patterns is completely random and the magnification response is highly
+    non-linear}. The consequence of this fact is there is a large variety of
+possible light curve morphologies and the parameter space of the models is
+highly complex.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{figures/close_wide_degeneracy.pdf}
+    \caption{Illustration of the ``close/wide'' degeneracy in binary microlensing events.
+        The two panels on the left show the magnification maps (on a logarithmic scale) for
+        a binary lens with $q=5\times 10^{-3}$, $s=0.7$ (top panel), and $s=1/0.7$
+        (bottom panel). The dashed grey line indicates the source trajectory. The panel
+        on the right shows the corresponding magnification as a function of time.
+        We see that trajectories through two very different configurations of the binary lens
+        result in nearly identical magnification curves. In absence of good
+        photometric coverage near the central caustic crossing this approximate degeneracy
+        can become exact.}
+    \label{fig:close_wide_degeneracy}
+\end{figure}
+
+One notable feature of the binary lens is that in the limit $q\ll 1$ and
+$\lvert s-1\rvert \gg q$, the structure of the caustic curves are invariant to
+the $s\rightarrow s^{-1}$ transformation \citet{dominik1999} which is known in
+the literature as the \emph{close/wide degeneracy} and it often arises for
+microlensing events involving planets. Figure~\ref{fig:close_wide_degeneracy}
+illustrates this (approximate) degeneracy. The two panels on the left show the
+magnification maps of a binary lens with $q=5\times 10^{-3}$ and two different
+values of the separation $s$ related by the transformation $s\rightarrow
+    s^{-1}$ . The dashed grey line shows the trajectory of the source passing close
+to the central caustic. The panel on the right shows that the magnification as
+a function of time is nearly identical for the two different configurations of
+the binary lens. In the absence of good photometric coverage near the peak of
+the event such approximate degeneracies can become exact. These kinds of data
+dependent degeneracies (rather than exact mathematical symmetries) are a very
+common feature of microlensing \citep{erdl1993}.
+
+\subsubsection{Parametrizing the trajectory}
+To parametrize the trajectory $\mathbf u(t)$ including annual parallax effect
+we have use an additional angle to specify the orientation of the axis
+containing both lenses. We can build on the work done in
+Section~\ref{ssec:single_lens_parallax} and instead of interpreting the origin
+of the coordinate system as the location of the single lens, we interpret it as
+the midpoint between the two lenses which lie on the same axis. $\mathbf u_0$
+is then the point on the trajectory closest to the midpoint and $\dot{\mathbf
+        u}_0$ is the velocity vector at that point. To specify the trajectory of the
+source in this new coordinate system we just need to rotate the coordinate
+system $(\hat{\mathbf e}_\parallel, \hat{\mathbf e}_\bot)$ by $\alpha$ where $\alpha$ is the angle between
+the $\hat{\mathbf e}_\parallel$ and the axis containing the two lenses. Using
+Equations~\ref{eq:u_t_parallel1} and~\ref{eq:u_t_parallel2} we have
+\begin{equation}
+    \mathbf u(t)  =
+    \mathbf R(\alpha)
+    \begin{pmatrix}
+        u_\bot (t) \\
+        u_\parallel (t)
+    \end{pmatrix}
+    =
+    \mathbf R(\alpha)
+    \renewcommand\arraystretch{2}
+    \begin{pmatrix}
+        u_0 + \pi_E\,\cos\psi\,\delta \zeta_e(t) - \pi_E\,\sin\psi\,\delta \zeta_n(t) \\
+        (t-t_0)/t_E' + \pi_E\,\sin\psi\,\delta \zeta_e(t) +
+        \pi_E\,\cos\psi\,\delta \zeta_n(t)
+    \end{pmatrix}
+\end{equation}
+where $\mathbf R (\alpha)$ is the rotation matrix  given by
+\begin{equation}
+    \mathbf R (\alpha)=
+    \begin{pmatrix}
+        \cos\alpha & -\sin\alpha \\
+        \sin\alpha & \cos\alpha
+    \end{pmatrix}
+\end{equation}
+The magnification of a binary lens is then fully parametrized by the following parameters
+\begin{equation}
+    (u_0,t_0,t'_E,\pi_E,\psi, a, q, \alpha)
+\end{equation}
+
+\subsection{Triple lens}
+The first detected triple lens microlensing event was OGLE-2006-BLG-109Lb,c
+\citep{2008Sci...319..927G,2010ApJ...713..837B}, a system consisting of two
+massive planets and a star, similar to Jupiter and Saturn in the Solar System.
+A circumbinary planet has also been detected \citep{2016AJ....152..125B} and
+there is a also possibility of detecting exomoons although none have been
+confirmed so far \citep{2010A&A...520A..68L}. To parametrize a triple lens
+system, we use the same coordinate system as for the binary lens with the
+addition of a third lens at an arbitrary location $z_3$ in the source plane.
+The lens equation is then
+\begin{equation}
+    w=z+\frac{\epsilon_{1}}{\bar{z} - a}+\frac{\epsilon_{2}}{\bar{z} + a} +
+    \frac{1 - \epsilon_1 - \epsilon_2}{\bar{z} - \bar{z}_3}
+\end{equation}
+The complex polynomial derived from the triple lens lens equation is a 10th degree
+polynomial.
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.8\linewidth]{figures/triple_lens_caustics.pdf}
+    \caption{Caustic structure for a triple lens with
+        $\epsilon_1=0.9$, $\epsilon_2=\epsilon_e=0.05$, $s=0.8$ and $z_3=0.3-i0.8$.
+        The caustics are considerably more complex than in the binary case and they
+        can be nested and self-intersecting.
+    }
+    \label{fig:triple_lens_caustics}
+\end{figure}
+
+The caustic/critical curve structure of a triple lens system is vastly more
+complex than the binary lens system. \citet{2019ApJ...880...72D} investigate
+three different kinds of systems: equal masses for all three lenses, two equal
+mass lenses and a low-mass third component and a hierarchical combination of
+the three masses. They find 11 different kinds of topologies of the caustics.
+Depending on the mass ratios, there could even be more kinds of topologies. In
+Figure~\ref{fig:triple_lens_caustics} we illustrate an example caustic
+structure for a triple lens system with $\epsilon_1=0.9$,
+$\epsilon_2=\epsilon_e=0.05$, $s=0.8$ and $z_3=0.3-i0.8$. The key features to
+note are that the caustic pattern is no longer symmetric about the x-axis and
+the caustics can be nested and self-intersecting.
+
+\subsection{Other effects}
+There are a few other effects which often need to be taken into account when
+modeling real microlensing events. I won't cover these in detail except for the
+first one because the focus on this thesis is on core but I describe them below
+for completeness.
+\begin{itemize}
+    \item \textbf{Finite source effects:} accurately computing the magnification
+          of a limb-darkened extended source is highly non-trivial, this is the subject
+          of Chapter~\ref{ch:microlensing}.
+    \item \textbf{Multiple sources:} instead of a single source star we could have
+          a multiple source stars \citep{1998A&A...333..893D}.  Events with a binary source
+          star can sometimes mimic binary lens events with a single source star.
+          Modeling these events requires specifying the trajectory and separation of
+          the binary and the total flux is then the sum of the flux from the primary and the
+          secondary star.
+          There is a paucity of binary source events in the microlensing literature, partly due to
+          their intrinsic rarity \citep{1998MNRAS.301..231H} and partly because there is a
+          bias in the community towards fitting binary lens models and
+          simply not putting as much effort into testing binary source models as
+          for the binary lens models \citep{2017AJ....153..129J,2019MNRAS.484.5608D}.
+    \item \textbf{Orbital motion of the lenses:} if the orbital timescale of
+          the lens in binary or triple lens systems is comparable to the event timescale
+          (this is true for a small subset of events) then one needs to take into account
+          the orbital motion of the lens projected onto the plane of the sky.
+          In the general case of a Keplerian orbit five additional parameters are needed
+          for the binary lens: the mass of the primary lens, the three components of the
+          secondary's velocity relative to the primary and the projected separation
+          of the secondary along the line of sight in units of $\theta_E$
+          \citep{1998A&A...329..361D}. In practice only two additional parameters
+          can be measured: the sky-projection of the two components of the velocity of
+          the secondary relative to the primary. These are parametrized by
+          $\gamma_\parallel\equiv \dot{s}/s$ -- the fractional rate of change of the projected
+          separation between the two lenses, and $\gamma_{\perp} \equiv-\dot{\alpha}$ --
+          the angular rotation rate of the projected separation axis. The effect of
+          $\gamma_\perp$ is to simply rotate the magnification pattern on the sky while a
+          nonzero value of $\gamma_\parallel$ changes the magnification pattern itself. These new parameters are
+          partially degenerate with other parameters such as parallax and it is very
+          rarely possible to constrain the full Keplerian orbit
+          \citep{2011ApJ...738...87S,2020A&A...633A..98W}.
+    \item \textbf{Orbital motion of the sources:} the effect of the orbital motion of a binary
+          source star is less dramatic than for the orbital motion of lens stars because
+          the caustic structure stays fixed. \citet{1998A&A...329..361D} showed that six
+          additional parameters are needed to describe the full Keplerian orbit of a
+          binary source star. \citet{1992ApJ...397..362G,1993ApJ...407..440G} pointed out
+          that such events should be rare although it is not clear if that is still the
+          case today with the advent of new microlensing surveys.
+          There is also a possibility that the secondary source is not a star but a large
+          planet resulting in subtle deviations in the primary
+          source position as it orbits the barycentar. This is an alternative channel for
+          detecting planets using microlensing and it is often
+          called \emph{xallarap} (inverse parallax) in the literature
+          \citep{2009MNRAS.392.1193R,2021AJ....162...59R,2021AJ....161...84M}.
+    \item \textbf{Fine structure in the source profiles:} microlensing in principle allows for
+          the measurement of the intensity and shape profile
+          of source stars and their possible planets. \citet{1997ApJ...490...38H} investigates
+          microlensing of elliptical sources by a single lens as a model for oblate stars and inclined
+          accretion discs. \citet{1999ApJ...513..619G} discusses the sensitivity of binary and single lens
+          caustic crossing events to spatial and spectral structure stellar atmospheres.
+          \citet{2003ApJ...586..527G} focused specifically on the
+          possibility of inferring
+          intensity and shape profiles of caustic crossings \emph{planets} in the source plane. Deviations
+          in the light curve due to a non-uniform surface brightness profile or shape of the planet are only
+          resolved during a short time window when the planet is within about one radius from the caustic.
+          \citet{2003ApJ...586..527G} find that there is a possibility of detecting ring-like structures
+          around the planet with a $\sim 30$m class telescope and high cadence observations but
+          detecting features such as spots and zonal bands will be very difficult to detect. Detecting
+          differences in \emph{phase} of the planet may be easier \citep{2001MNRAS.325..305A}.
+\end{itemize}
 
@@
 \label{sec:occultations}
-\subsection{History}
 \subsection{The Starry framework}
@@
 \label{sec:model_comparison}
+
+``Furthermore, in the future, a blind search for planetary signals will not
+turn up all potential competing models unless those models
+are specifically considered.''\citep{2017AJ....153..129J}.
+\citet{2021AJ....162...59R}
+\citet{2020AJ....160...17H}
+
 \subsection{Cross validation}
@@
 
-\chapter{First appendix}
+\chapter{Complex polynomial coefficients}
 \label{app:complex_poly}
-
-\chapter{Second appendix}
+\section{Lens equation}
+In this appendix we (partially) derive the coefficients of the complex
+polynomial derived from the lens equation
+(Equation~\ref{eq:lens_equation_complex}). Taking the complex conjugate of
+Equation~\ref{eq:lens_equation_complex}, have
+\begin{equation}
+    w=z-\sum_{i=1}^{N} \frac{\epsilon_{i}}{\bar{z}-\bar{z}_{i}}
+\end{equation}
+Let's define $z_i\equiv z - z_j$ and two polynomials $G$ and $H$
+\begin{align}
+    G & =\sum_{k=0}^{N-1} G_{k} z^{k} =\sum_{j=1}^{N} \epsilon_{j} \prod_{i=1, i \neq j}^{N} z_{i} \\
+    H & =\sum_{k=0}^{N} H_{k} z^{k}   =\prod_{i=1}^{N} z_{i}
+\end{align}
+it follows that
+\begin{equation}
+    \frac{G}{H}=\frac{\sum_{j=1}^{N} \epsilon_{j}
+        \prod_{i \neq j} z_{i}}{\prod_{j=1}^{N} z_{j}}=\sum_{j=1}^{N} \frac{\epsilon_{j}}{z_{j}}
+\end{equation}
+and
+\begin{equation}
+    \bar{z}=\bar{w}+G / H
+\end{equation}
+Defining $\omega_{j} \equiv \bar{r}_{j}-\bar{w}_{i}$, we obtain
+\begin{equation}
+    z-w=\sum_{j=1}^{N} \frac{\epsilon_{j}}{G / H+\bar{w}-\bar{r}_{j}}
+\end{equation}
+Multiplying the above equation with $\prod_{j=1}^{N}\left(G-\varpi_{j} H\right)$
+results in
+\begin{equation}
+    0=(z-w) \prod_{j=1}^{N}\left(G-\varpi_{j} H\right)-
+    H \sum_{j=1}^{N}\left[\epsilon_{j} \prod_{i=1, j \neq i}^{N}\left(G-\varpi_{i} H\right)\right]
+    \label{eq:complex_poly}
+\end{equation}
+Equation~\ref{eq:complex_poly} is complex polynomial of degree $N^2+1$. To obtain coefficients of this
+polynomial in terms of fundamental parameters we use a computer algebra software package \textsf{SymPy}\footnote{
+    \url{https://www.sympy.org}
+}\citep{10.7717/peerj-cs.103}. A Jupyter notebook with the complete derivation is available
+\href{https://github.com/fbartolic/caustics/blob/main/notebooks/ComplexPolynomialCoefficients.ipynb}{here}.
+I have managed to derive the coefficient for the binary and triple lens case but \textsf{SymPy} fails to
+converge for the quadruple lens case because the number of coefficients becomes very large.
+It is possible that an alternative computer algebra system such as Wolfram Mathematica would work.
+
+\section{Critical curve equation}
+\label{app:complex_poly_crit}
+Similarly to the lens equation, we can derive the coefficients of the complex
+polynomial equation whose roots are points on the critical curve. To solve for
+these points we need to solve the following equation
+\citep{1990A&A...236..311W}:
+\begin{equation}
+    \sum_{i=1}^{N} \frac{\epsilon_{i}}{\left(\bar{z}-\bar{r}_{i}\right)^{2}}=e^{i \phi}
+\end{equation}
+for each value of the parameter $\phi\in[0,2\pi]$.
+Taking the complex conjugate of the above equation we have
+\begin{equation}
+    \sum_{i=1}^{N} \frac{\epsilon_{i}}{\left(z-r_{i}\right)^{2}}=e^{-i \phi}
+\end{equation}
+We define polynomials $G'$ and $H'$
+\begin{align}
+    G' & = \sum_{j=1}^N\epsilon_j\prod_{i=1,\;i\neq j}^Nz^2_i \\
+    H' & = \prod_{i=1}^Nz^2_i
+\end{align}
+and
+\begin{equation}
+    e^{-i\phi}=\frac{G'}{H'}
+\end{equation}
+Finally, we obtain the polynomial
+\begin{equation}
+    e^{-i\phi}H' - G' = 0
+    \label{eq:complex_poly_critical}
+\end{equation}
+Equation~\ref{eq:complex_poly_critical} is a complex polynomial of degree $2N$. We can
+obtain its coefficients with \textsf{SymPy} as before and their values are available
+\href{https://github.com/fbartolic/caustics/blob/main/notebooks/ComplexPolynomialCoefficients.ipynb}{here}.
 

### 2022-05-30 src/figures/binary_lens_topology.py

@@ -1,0 +1,104 @@
+import numpy as np
+import jax.numpy as jnp
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+import matplotlib.pyplot as plt
+
+from caustics import critical_and_caustic_curves_binary
+
+s_array = jnp.array([0.6, 1 / np.sqrt(2), 1.2, 2.0, 3.2])
+
+fig, ax = plt.subplots(5, 2, figsize=(10, 18), sharex=True)
+fig.subplots_adjust(hspace=0.15, wspace=0.3)
+
+for i, s_ in enumerate(s_array):
+    critical_curves, caustic_curves = critical_and_caustic_curves_binary(
+        0.5 * s_, 0.5, npts=1000
+    )
+
+    ax[i, 0].scatter(
+        critical_curves.real, critical_curves.imag, s=0.8, c="k", alpha=0.8
+    )
+    ax[i, 1].scatter(caustic_curves.real, caustic_curves.imag, s=0.8, c="k", alpha=0.8)
+
+
+for a in ax.ravel():
+    #    a.grid(True)
+    a.set_xlim(-2.5, 2.5)
+    a.set_ylim(-1.6, 1.6)
+    a.set_aspect(1)
+
+for a in ax[:, 0]:
+    a.set_ylabel(r"$\mathrm{Im}(z)$")
+
+for a in ax[:, 1]:
+    a.set_ylabel(r"$\mathrm{Im}(w)$")
+
+ax[-1, 0].set_xlabel(r"$\mathrm{Re}(z)$")
+ax[-1, 0].set_xticks([-2, -1, 0, 1, 2])
+ax[-1, 1].set_xlabel(r"$\mathrm{Re}(w)$")
+ax[-1, 1].set_xticks([-2, -1, 0, 1, 2])
+
+ax[0, 0].set_title(r"close", x=1.15, pad=10)
+ax[0, 0].text(
+    0.17,
+    0.1,
+    "$s=0.6$",
+    horizontalalignment="center",
+    verticalalignment="center",
+    transform=ax[0, 0].transAxes,
+    fontsize=16,
+)
+ax[1, 0].set_title(r"close$\;\rightarrow\;$intermediate", x=1.15, pad=10)
+ax[1, 0].text(
+    0.17,
+    0.1,
+    "$s=1/\sqrt{2}$",
+    horizontalalignment="center",
+    verticalalignment="center",
+    transform=ax[1, 0].transAxes,
+    fontsize=16,
+)
[…]

### 2022-05-30 src/figures/close_wide_degeneracy.py

@@ -1,0 +1,105 @@
+import numpy as np
+import jax.numpy as jnp
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+
+from caustics import mag_point_source_binary
+
+q = 5e-03
+e1 = q / (1 + q)
+
+s_array = jnp.array([0.7, 1 / 0.7])
+
+t = jnp.linspace(-10.0, 10.0, 600)
+tE = 30
+
+
+mosaic = """
+    AC
+    BC
+    """
+fig = plt.figure(figsize=(14, 6), constrained_layout=True)
+ax_dict = fig.subplot_mosaic(mosaic, gridspec_kw={"wspace": 0.05})
+
+ax_maps = [ax_dict["A"], ax_dict["B"]]
+
+xgrid, ygrid = jnp.meshgrid(jnp.linspace(-1.2, 0.2, 1000), jnp.linspace(-0.3, 0.3, 300))
+wgrid = xgrid + 1j * ygrid
+
+for i, s_ in enumerate(s_array):
+    a = 0.5 * s_
+    x_cm = (2 * e1 - 1) * a
+    mag = mag_point_source_binary(wgrid, a, e1)
+    im = ax_maps[i].pcolormesh(
+        xgrid,
+        ygrid,
+        mag,
+        cmap="Greys",
+        norm=colors.LogNorm(vmin=1, vmax=200),
+        zorder=-1,
+    )
+
+
+# Trajectory s
+t0 = -1.5
+u0 = -0.36
+alpha = jnp.deg2rad(10.0)
+
+x_traj = jnp.cos(alpha) * u0 - jnp.sin(alpha) * (t - t0) / tE
+y_traj = jnp.sin(alpha) * u0 + jnp.cos(alpha) * (t - t0) / tE
+w_traj = x_traj + 1j * y_traj
+mag_traj = mag_point_source_binary(w_traj, 0.5 * s_array[0], e1)
+ax_maps[0].plot(x_traj, y_traj, color="k", alpha=0.5, linestyle="dashed")
+ax_dict["C"].plot(t, mag_traj, color=f"k", lw=2.5, label="s = 0.7")
+
+# Trajectory 1/s
+t0 = -3.7
+u0 = -0.7167
+alpha = jnp.deg2rad(10.8)
+
+x_traj = jnp.cos(alpha) * u0 - jnp.sin(alpha) * (t - t0) / tE
[…]

### 2022-05-30 src/figures/planetary_caustics.py

@@ -1,0 +1,78 @@
+import numpy as np
+import jax.numpy as jnp
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+
+from caustics import mag_point_source_binary
+
+q = 5e-03
+e1 = q / (1 + q)
+s_array = [0.8, 0.9, 1.0, 1.2, 1.4]
+
+fig, ax = plt.subplots(len(s_array), figsize=(12, 14), sharex=True)
+fig.subplots_adjust(hspace=0.05)
+
+for i, s_ in enumerate(s_array):
+
+    xgrid, ygrid = jnp.meshgrid(
+        jnp.linspace(-1, 1.0, 1500), jnp.linspace(-0.2, 0.2, 300)
+    )
+    wgrid = xgrid + 1j * ygrid
+    a = 0.5 * s_
+    x_cm = (2 * e1 - 1) * a
+    mag = mag_point_source_binary(wgrid, a, e1)
+    im = ax[i].pcolormesh(
+        xgrid,
+        ygrid,
+        mag,
+        cmap="Greys",
+        norm=colors.LogNorm(vmin=1, vmax=200),
+        zorder=-1,
+    )
+
+
+for i, s_ in enumerate(s_array):
+    ax[i].scatter(0.5 * s_, [0.0], color="C0", marker="o", label=r"planet", alpha=0.7)
+    ax[i].scatter(
+        -0.5 * s_, [0.0], color="C1", marker="*", label=r"star", s=100, alpha=0.7
+    )
+    ax[i].text(
+        0.98,
+        0.12,
+        f"$s={s_}$",
+        horizontalalignment="right",
+        verticalalignment="center",
+        transform=ax[i].transAxes,
+        fontsize=16,
+    )
+
+
+for a in ax.ravel():
+    a.set_aspect(1)
+
+    # Formatting each yaxis
+    minorLocator = plt.MultipleLocator(0.1)
+    minorFormatter = plt.FormatStrFormatter("%1.1f")
+    a.xaxis.set_minor_locator(minorLocator)
+    a.set_rasterization_zorder(0)
+
+
+ax[0].legend(prop={"size": 12})
+ax[2].set_ylabel(r"$\mathrm{Im}(w)$")
+ax[-1].set_xlabel(r"$\mathrm{Re}(w)$")
+
+# Add colorbar on the right side
+plt.colorbar(
[…]

### 2022-05-30 src/figures/triple_lens_caustics.py

@@ -1,0 +1,49 @@
+import numpy as np
+import jax.numpy as jnp
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+
+from caustics import mag_point_source_triple
+
+# Lens postions
+a = 0.4
+r3 = 0.3 - 0.8j
+
+# Lens mass ratios
+e1 = 0.9
+e2 = 0.05
+e3 = 1.0 - e1 - e2
+
+# Define 2D grid in the source plane
+npts = 1000
+
+x = jnp.linspace(-0.1, 1, npts)
+y = jnp.linspace(-0.45, 0.5, npts)
+xgrid, ygrid = jnp.meshgrid(x, y)
+wgrid = xgrid + 1j * ygrid
+
+mag = mag_point_source_triple(wgrid, a, r3, e1, e2)
+
+fig, ax = plt.subplots(figsize=(12, 8))
+im = ax.pcolormesh(
+    jnp.real(wgrid),
+    jnp.imag(wgrid),
+    mag,
+    cmap="Greys",
+    norm=colors.LogNorm(vmax=150),
+    zorder=-1,
+)
+
+ax.set_aspect(1)
+ax.set(xlabel="$\mathrm{Re}(w)$", ylabel="$\mathrm{Im}(w)$")
+plt.colorbar(im, label="magnification")
+ax.set_rasterization_zorder(0)
+
+# Save to disc
+fig.savefig("triple_lens_caustics.pdf", bbox_inches="tight", dpi=200)
+

### 2022-06-04 src/ms.tex
finish writing the occultation section of chapter 2

@@
 \usepackage{siunitx} % for proper units 
+\usepackage{esint} % various fancy integral symbols
+\usepackage[bbgreekl]{mathbbol}
 
@@
 \usepackage[font=scriptsize]{caption} % smaller font for fig captions
-\usepackage{placeins}
+\usepackage[section]{placeins}
 %\usepackage{import}
@@
 Later works such as
-\citet{1992Icar...97..211B,1993Icar..102..134Y,1999AJ....117.1063Y} went a step
-further by using not only phase curves but also light curves of mutual
-\emph{occultations} of Pluto by its moon Charon to reconstruct albedo maps with
-greater success. The major advantage of occultations relative to just phase
-curves is that more information about the surface is encoded into the light
-curve because of the sharp limb of the occultor sweeping over the disc of an
-occulted body and blocking the reflected light. More importantly for this
+\citet{1986AJ.....92.1201D,1992Icar...97..211B,1993Icar..102..134Y,1999AJ....117.1063Y}
+went a step further by using not only phase curves but also light curves of
+mutual \emph{occultations} of Pluto by its moon Charon to reconstruct albedo
+maps with greater success. The major advantage of occultations relative to just
+phase curves is that more information about the surface is encoded into the
+light curve because of the sharp limb of the occultor sweeping over the disc of
+an occulted body and blocking the reflected light. More importantly for this
 thesis, \citet{1994Icar..107..195S} was the first to observe the occultations
@@
 \subsection{Magnification of an extended source}
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/single_lens_images.pdf}
+        \caption{Images of an extended limb-darkened source star lensed by a single
+            point lens for varying position of the source star. The top panels show the magnification map with a logarithmic scale
+            colormap  consisting of a single point caustic. The semi-transparent circle
+            is the source star disc with radius $\rho_\star=0.15$. The bottom row shows
+            the two images merging into the Einstein ring as the source moves over the caustic.}
+        \label{fig:single_lens_images}
+    \end{centering}
+\end{figure}
 For a relatively small subset of microlensing events the source star passes
@@
 \citet{2022arXiv220306637S}.
-
-\begin{figure}[t]
-    \begin{centering}
-        \includegraphics[width=\linewidth]{figures/single_lens_images.pdf}
-        \caption{Images of an extended limb-darkened source star lensed by a single
-            point lens for varying position of the source star. The top panels show the magnification map with a logarithmic scale
-            colormap  consisting of a single point caustic. The semi-transparent circle
-            is the source star disc with radius $\rho_\star=0.15$. The bottom row shows
-            the two images merging into the Einstein ring as the source moves over the caustic.}
-        \label{fig:single_lens_images}
-    \end{centering}
-\end{figure}
 
@@
 \subsection{Binary lens}
-In this section we will focus specifically on the binary lens. We choose a
-coordinate system whose origin is at the midpoint of the line connecting the
-two lenses and both lenses are on the real axis. If the first lens is at
-distance $a$ ($z_1=a$) then the second lens is at $-a$ and the lens equation
-takes the form
-\begin{equation}
-    w=z+\frac{\epsilon_{1}}{\bar{z} - a}+\frac{1 - \epsilon_{1}}{\bar{z} + a}
-\end{equation}
-The above equation can be rewritten as a 5th order complex polynomial
-(see Appendix~\ref{app:complex_poly}). Either 3 or 5 of the complex roots of that
-polynomial are also solutions to the lens equation
-(5 inside the caustics and 3 outside).
-It is common to use the mass ratio $q\equiv \epsilon_2/\epsilon_1$ with
-$\epsilon_2 \leq \epsilon_1$ instead  of $\epsilon_1$ as a parameter and the
-separation between the lenses $s\equiv 2a$ instead of half the separation $a$.
-As a reminder all angular quantities are expressed in the units of the angular Einstein
-radius of the total mass of the system.
-
-\begin{figure}[t]
+\begin{figure}[ht!]
     \centering
@@
 \end{figure}
+In this section we will focus specifically on the binary lens. We choose a
+coordinate system whose origin is at the midpoint of the line connecting the
+two lenses and both lenses are on the real axis. If the first lens is at
+distance $a$ ($z_1=a$) then the second lens is at $-a$ and the lens equation
+takes the form
+\begin{equation}
+    w=z+\frac{\epsilon_{1}}{\bar{z} - a}+\frac{1 - \epsilon_{1}}{\bar{z} + a}
+\end{equation}
+The above equation can be rewritten as a 5th order complex polynomial
+(see Appendix~\ref{app:complex_poly}). Either 3 or 5 of the complex roots of that
+polynomial are also solutions to the lens equation
+(5 inside the caustics and 3 outside).
+It is common to use the mass ratio $q\equiv \epsilon_2/\epsilon_1$ with
+$\epsilon_2 \leq \epsilon_1$ instead  of $\epsilon_1$ as a parameter and the
+separation between the lenses $s\equiv 2a$ instead of half the separation $a$.
+As a reminder all angular quantities are expressed in the units of the angular Einstein
+radius of the total mass of the system.
 
@@
           caustic crossing events to spatial and spectral structure stellar atmospheres.
-          \citet{2003ApJ...586..527G} focused specifically on the
-          possibility of inferring
+          \citet{2003ApJ...586..527G} focused specifically on the possibility of inferring
           intensity and shape profiles of caustic crossings \emph{planets} in the source plane. Deviations
@@
 \label{sec:occultations}
+In this section I will briefly review the theory behind modeling phase curves and
+occultation light curves of spherical bodies for the case of emitted light
+(isotropic blackbody emission from the body itself) and reflected light
+(starlight scattered from the surface or atmosphere of the body). The following is
+mostly a short summary of the papers introducing the \textsf{starry} framework
+\citep{2019AJ....157...64L,2021arXiv210306275L}.
 \subsection{The Starry framework}
-\subsection{Reflected vs. emitted light}
-
-% In this and the following sections make sure to use microlensing and occultation
-% mapping as examples
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/spherical_harmonics.pdf}
+        \caption{Real spherical harmonics up to degree $l=5$ computed from
+            Equation~\ref{eq:spherical_harmonics}. Figure adapted from Figure 1 in
+            \citet{2019AJ....157...64L}.}
+        \label{fig:spherical_harmonics}
+    \end{centering}
+\end{figure}
+\subsubsection{Emitted light}
+Consider a spherical body of unit radius emitting light isotropically (a
+Lambertian emitter) at each point $(\theta,\phi)$ where $\theta$ is the
+inclination angle and $\phi$ is the azimuthal angle. To compute a phase curve
+or an occultation light curve we need to be able to integrate the flux over the
+entire projected disc of the body given the distribution of specific intensity
+$I(\theta, \phi)$, the 3D orientation of the body in space, and the location of
+a possible occultor relative to the body. These two-dimensional integrals can
+always be computed numerically (see for instance the approaches presented in
+\citet{2018AJ....156..146F,2018MNRAS.477.2613L}) but
+\citet{2019AJ....157...64L} showed that it is possible to compute them
+analytically if the specific intensity is first expanded in a basis of
+spherical harmonics and then mapped to a different basis more suitable for
+evaluating the integrals.
+
+Following \citet{2019AJ....157...64L}, we start by setting up a cartesian
+coordinate system on the units sphere, such that
+\begin{align}
+     & x=\sin \theta \cos \phi \\
+     & y=\sin \theta \sin \phi \\
+     & z=\cos \theta
+\end{align}
+and the observer is located on the $z$-axis at $\infty$ such that the projected disc
+of the body is centered at the origin of the $xy$ plane with the unit vector
+$\hat{\mathbf{x}}$ pointing to the right and $\hat{\mathbf{y}}$ pointing up.
+We introduce spherical harmonics $Y_{l m}(\theta, \phi)$ of degree $l\geq 0$ and order
+$m\in [-l, l]$ defined  as
+\begin{equation}
+    Y_{l m}(\theta, \phi)= \begin{cases}\bar{P}_{l m}(\cos \theta) \cos (m \phi) & m \geqslant 0 \\ \bar{P}_{l|m|}(\cos \theta) \sin (|m| \phi) & m<0\end{cases}
+    \label{eq:spherical_harmonics}
+\end{equation}
+When the spherical harmonics defined above are rewritten in terms of $x$, $y$, and $z$
+they become polynomials in these variables (see Appendix A in \citet{2019AJ....157...64L}).
+We can expand the specific intensity distribution $I(x,y)$ in a spherical harmonic basis
+as
+\begin{equation}
+    I(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
+    \label{eq:sh_expansion}
+\end{equation}
+where $\tilde{\mathbf{y}}$ is the basis vector of spherical harmonics  arranged in
+increasing order:
+\begin{align}
+     & \tilde{\mathbf{y}}=\left(\begin{array}{lllll}
+                                    Y_{0,0} & Y_{1,-1} & Y_{1,0} & Y_{1,1} & Y_{2,-2}
+                                \end{array}\right. \\
+     & \left.\begin{array}{lllll}
+                 Y_{2,-1} & Y_{2,0} & Y_{2,1} & Y_{2,2} & \cdots
+             \end{array}\right)^{\top} \quad,
+\end{align}
+and $\mathbf{y}$ is a vector of scalar \emph{spherical harmonic coefficients}.
+$\mathbf{y}$ is a central quantity which defines the \emph{map} (the specific
+intensity at every point) of the body.
+\citet{2019AJ....157...64L} shows that we can represent $y$ in a polynomial basis (see
+\citet{2019AJ....157...64L} for the general expression)
+\begin{equation}
+    \tilde{\boldsymbol{p}}=\left(\begin{array}{llllllllll}
+        1 & x & z & y & x^{2} & x z & x y & y z & y^{2} & \cdots
+    \end{array}\right)^{\top}
+\end{equation}
+such that
+\begin{align}
+    I(x, y) & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{p}                    \\
+            & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{y}\quad,
+    \label{eq:intensity_poly_basis}
+\end{align}
+and also in the Green's basis
+\begin{equation}
+    \tilde{\boldsymbol{g}}=\left(\begin{array}{llllllllll}
+        1 & 2 x & z & y & 3 x^{2} & -3 x z & 2 x y & 3 y z & y^{2} & \cdots
+    \end{array}\right)^{\top}
+\end{equation}
+such that
+\begin{align}
+    I(x, y) & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{g}              \\
+            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A}_2 \mathbf{p} \\
+            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A} \mathbf{y}
+\end{align}
+where $\mathbf{A}_1$ is the change-of-basis matrix from $\mathbf{y}$ to $\mathbf{p}$,
+$\mathbf{A}_2$ is the change-of-basis matrix from $\mathbf{p}$ to $\mathbf{g}$, and
+$\mathbf{A}\equiv \mathbf{A}_1 \mathbf{A}_2$.
+
+To compute rotational light curves we also need to be able to specify the
+orientation of a surface map with coefficients $\mathbf y$. We can rotate the
+map using the Wigner rotation matrix $\mathbf{R}(\mathrm{I}, \Lambda, \Theta)$
+that rotates $y$ given the body's inclination $\mathrm{I}$, obliquity $\Lambda$
+and rotational phase $\Theta$. The rotated map is then $\mathbf{R}(\mathrm{I},
+    \Lambda, \Theta) \mathbf{y}$ and Equation~\ref{eq:intensity_poly_basis} can be
+rewritten more generally as
+\begin{equation}
+    I(x, y)=\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{R}\mathbf{y}
+\end{equation}
+The total flux measured by an observer is is given by an integral of the specific
+intensity over a region S of the projected disc of the body:
+\begin{align}
+    F & =\oiint I(x, y) \ud S                                                            \\
+      & =\oiint \tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_{1} \mathbf{R} \mathbf{y} d S \\
+      & =\mathbf{r}^{\top} \mathbf{A}_{1} \mathbf{R} \mathbf{y}
+\end{align}
+where $\mathbf{r}$ is a column vector whose $n$-th element is \citep{2019AJ....157...64L}
+\begin{equation}
+    r_{n} \equiv \oiint \tilde{p}_{n}(x, y) \ud S
+\end{equation}
+When the entire disc is visible (no occultor)
+\begin{equation}
+    r_{n}=\int_{-1}^{1} \int_{-\sqrt{1-x^{2}}}^{\sqrt{1+x^{2}}} \tilde{p}_{n}(x, y) \mathrm{d}y \mathrm{d} x
+\end{equation}
+and the solution to this integral can be expressed in terms of Gamma functions
+\citep[Equation 20 in ][]{2019AJ....157...64L}.
+$\mathbf{r}$ and $\mathbf{A}_1$ are independent of the map coefficients $\mathbf{y}$
+so they can be pre-computed.
+
+Computing the occultation light curves is more complicated because the the body
+is partially occulted by an occultor of radius $r$, centered at $(x_o, y_o)$
+and the exposed portion of the disc is a function of $(r, x_o, y_o)$. The
+general expression for the flux is
+\begin{align}
+    F & =\oiint I(x, y) \mathrm{d} S                         \\
+      & =\mathbf{s}^\top  \mathbf{A} \mathbf{R} \mathbf{y} .
+\end{align}
+where the vector
+\begin{equation}
+    \mathbf{s}^\top \equiv \oiint \tilde{\mathbf{g}}^{\top}(x, y) \mathrm{d} S
+\end{equation}
+is defined to be the solution to the integral.
+\citet{2019AJ....157...64L} solves this integral in two steps.  First, we rotate
+the coordinate system about the  $z$-axis  so that the occultor lies at along the
+$+y$ axis and its center is at
+a distance $b=\sqrt{x_o^{2}+y_o^{2}}$ from the origin. This substantially simplifies
+the limits of integration.
+The second step is to use Green's theorem to express the surface integral of
+$\tilde{\mathbf{g}}_n$ as a line integral of a vector function $\mathbf{G}_n$
+along the boundary of the visible portion of the occulted disc.
+The $n$-th element of the solution vector $\mathbf{s}^\top$ is
+\begin{equation}
+    s_{n}=\oiint \tilde{g}_{n}(x, y) \mathrm{d} S=\oint \mathbf{G}_{n}(x, y) \cdot \ud \mathbf{r}
+\end{equation}
+where
+$\mathbf{G}_{n}(x, y)=G_{n x}(x, y) \hat{\mathbf{x}}+G_{n y}(x, y) \hat{\mathbf{y}}$
+is constructed such that
+\begin{equation}
+    \mathbf{D} \wedge \mathbf{G}_{n}=\tilde{g}_{n}(x, y)
+    \label{eq:G_n}
+\end{equation}
+were $\mathbf{D} \wedge \mathbf{G}_{n}$ is the exterior derivative of $\mathbf{G}_n$
+given by
+\begin{equation}
+    \mathbf{D} \wedge \mathbf{G}_{n} \equiv \frac{\ud G_{n y}}{\ud x}-\frac{\ud G_{n_{x}}}{\ud y}
+\end{equation}
+in Cartesian coordinates.
+\citet{2019AJ....157...64L} provides one possible expression for $\mathbf{G}_n$
+which satisfies Equation~\ref{eq:G_n} and shows that the final solution can be written as
+\begin{equation}
+    s_{n}=\mathcal{Q}\left(\mathbf{G}_{n}\right)-\mathcal{P}\left(\mathbf{G}_{n}\right)
+\end{equation}
+where $\mathcal{P}(\mathbf{G}_n)$ is the  line integral along the arc of the occultor
+of radius $r$ and $\mathcal{Q}(\mathbf{G}_n)$ is the line integral along the arc of the
+occulted body of unit radius. Solutions to these integrals are long but they consist of
+analytic functions such as sines, cosines and complete elliptic integrals and importantly
+they only need to be evaluated once for an arbitrary map with fixed geometry of an
+occultation \citep{2019AJ....157...64L}.
+
+The final expression for the total flux for a particular geometrical
+arrangement of the occulted body and the occultor can then be written
+succinctly as
+\begin{equation}
+    f = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{y}
+    \label{eq:starry}
+\end{equation}
+Where $\mathbf{R}'$ is a rotation matrix which rotates the map such that the occultor
+is placed symmetrically on the $+y$ axis at distance $b$.
+Notice that Equation~\ref{eq:starry} defines a linear operation acting on a vector
+of spherical harmonic coefficients $\mathbf{y}$.
+It is also trivial to generalize this expression to a case of \emph{spectral map}. In
+that case we define a matrix $Y$ whose columns are the spherical harmonic coefficients
+at different wavelengths and the total flux (spectrum) is then
+\begin{equation}
+    \mathbf{f} = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{Y}
+    \label{eq:starry_mw}
+\end{equation}
+where $\mathbf{f}$ is a vector of fluxes, one per wavelength bin.
+
+The formalism above can also be extended to account for limb-darkening which is
+important when modeling transits of planets across stars or the light curves of
+eclipsing binaries. \citet{2020AJ....159..123A} show how a limb-darkening
+profile that is an order $l$ polynomial function of
+$\mu=z=\sqrt{1-x^{2}-y^{2}}$ can be expressed in terms of the $m=0$ spherical
+harmonics up to order $l$. In the case of quadratic limb-darkening
+$(l_\mathrm{max}=2)$, the limb darkening profile is
+\begin{equation}
+    \frac{I(\mu)}{I(1)}=1-u_{1}(1-\mu)-u_{2}(1-\mu)^{2}
+\end{equation}
+which can be expressed as a sum of spherical harmonics
+\citep[Equation 38 in][]{2019AJ....157...64L}. Limb-darkening needs to be
+treated separately from the map coefficients $\mathbf{y}$ because it does not rotate along with the map
+when rotations $\mathbf{R}$ and $\mathbf{R}'$ are applied. The limb-darkening
+coefficients are applied to the map $\mathbf{y}$ as a multiplicative filter
+following any rotations which results in another set of spherical harmonic
+coefficients because products of spherical harmonics are also spherical
+harmonics. Applying limb-darkening raises the degree of the map by the degree
+of the limb-darkening.
+
+\subsubsection{Reflected light}
+The results derived in the previous section are valid only for light emitted
+from the occulted body. Equation~\ref{eq:starry} is a good model for stellar
+light curves and secondary eclipse light curves and phase curves of planets
+observed in the mid to far infrared wavelengths where the contribution of
+scattered starlight to total flux is negligible. Modeling reflected light phase
+curves and occultations is far more complicated because one has to take into
+account the the nonuniform illumination of the body and the possible presence
+of a terminator line (boundary between day and night).
+\citet{2021arXiv210306275L} extends the formalism described in the previous
+section to the case of reflected light phase curves and occultations for a body
+whose spatial \emph{albedo} (specifically the \emph{spherical albedo} -- the
+fraction of power incident on a body at a given wavelength that is scattered
+into space in all directions) distribution is expanded in a basis of spherical
+harmonics. Thus, analogous to Equation~\ref{eq:sh_expansion} we have
+\begin{equation}
+    A(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
+\end{equation}
+where $A$ is the albedo. We work under the assumption of isotropic (Lambertian)
+scattering of light at every point on the body which means that the illumination
+profile $\mathcal{I}$ of the surface is given by Lambert's law:
+\begin{equation}
+    \mathcal{I}\left(\vartheta_{\mathrm{i}}\right)=\mathcal{I}_{0} \max \left(0, \cos \vartheta_{\mathrm{i}}\right)
+\end{equation}
+where $\vartheta_i$ is the angle between the incident radiation at the surface normal
+and $\mathcal{I}_0$ is the maximum illumination. In case of, for example, a planet illuminated
+by its host star \citep[Appendix A.2 in][]{2021arXiv210306275L}
+\begin{equation}
+    \mathcal{I}_{0}=\frac{f_{s}}{\pi r_{\mathrm{s}}^{2}}
+\end{equation}
+where $r_s$ is the distance between the planet and the star in units of planet's radius
+and $f_s$ is the stellar flux measured at the observer in some arbitrary units.
+We assume that $f_s=1$ so that all fluxes are defined as the fraction of the flux
+of the illumination source at the observer.
+It is the presence of the day/night terminator that complicates the calculation
+of total flux the most. The limits of integration end up depending on a
+solution to a quartic equation specifying the points of intersection between
+the occultor and the day/night terminator line and the solution to those
+integrals are a function of \emph{incomplete} elliptic integrals as opposed to
+complete elliptic integrals as was the case for occultations in emitted light.
+The full derivation of the total flux is provided by \citet{2021arXiv210306275L} and
+I will only briefly summarize  the result here for completeness.
+
+Under the Lambertian scattering assumption the observed intensity at any point
+of the surface of the body is is proportional to the cosine of the angle
+$\vartheta_i$ between the incident light rays and the surface normal. All
+points for which $\vartheta_i\geq \pi/2$ have intensity of zero (they're
+unilluminated). Let's assume that the point source is placed at sky coordinates
+$\left(x_{\mathrm{s}}, y_{\mathrm{s}}, z_{\mathrm{s}}\right)$ in units of the
+radius of the illuminated body. The day/night terminator on the body is a
+half-ellipse with a semi-major axis equal to unity and a (signed) semi-minor
+axis equal to
+\begin{equation}
+    b=-\frac{z_{\mathrm{s}}}{r_{\mathrm{s}}}
+\end{equation}
+where $r_{\mathrm{s}}=\sqrt{x_{\mathrm{s}}^{2}+y_{\mathrm{s}}^{2}+z_{\mathrm{s}}^{2}}$
+is the distance to the source. The angle by which the semi-major axis of this ellipse is
+rotated away from the $+x$-axis is  is
+\begin{equation}
+    \theta=-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
+\end{equation}
+Under the assumption that $rs \gg 1$, the illumination $\mathcal{I}$ at point
+$(x,y)$ on the projected disc is \citep{2021arXiv210306275L}
+\begin{equation}
+    \mathcal{I}\left(b, \theta, r_{\mathrm{s}} ; x, y\right)=\max \left(0, I\left(b, \theta, r_{\mathrm{s}} ; x, y\right)\right)
+    \label{eq:illumination}
+\end{equation}
+where
+\begin{align}
+    I\left(b, \theta, r_{\mathrm{s}} ; x, y\right) & =\frac{1}{\pi r_{\mathrm{s}}^{2}} \cos \vartheta_{\mathrm{i}}                                                      \\
+                                                   & =\frac{1}{\pi r_{\mathrm{s}}^{2}}\left(-b_{\mathrm{c}} \sin \theta x+b_{\mathrm{c}} \cos \theta y-b z(x, y)\right)
+\end{align}
+with  $b_{\mathrm{c}} \equiv \sqrt{1-b^{2}}$  and $z(x, y)=\sqrt{1-x^{2}-y^{2}}$.
+The illumination $\mathcal{I}$ is a dimensionless quantity normalized such that the
+integral of $A\,\mathcal{I}$ over the units disc is equal to the flux measured by the
+observer as a fraction of the flux of the illumination source.
+To compute the total flux, we need to weigh each of the terms in the Green's basis
+and integrate them over the visible portion of the body's disc to obtain
+the reflected light solution vector $\mathbb{s}^\top$.
+Evaluating these integrals is unfortunately intractable because of the piecewise
+function in Equation~\ref{eq:illumination}. It is more tractable to weight the basis
+terms by the function $I$ and modify the limits of integration such that the nightside
+of the body is excluded. \citet{2021arXiv210306275L} shows that $I$ can be expressed
+in the form of a linear operator $\mathbf{I}$ to weight a map vector  in the polynomial
+basis $\tilde{\mathbf{p}}$ by the illumination profile. The total flux is then
+\begin{equation}
+    f=\mathbb{s}^\top\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right) \mathbf{A}_{\mathbf{2}} \mathbf{I}\left(b, \theta^{\prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime}\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+\end{equation}
+where
+\begin{equation}
+    \theta^{\prime}=\arctan 2\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right)-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
+    \label{eq:starry_ref}
+\end{equation}
+is the angle of the terminator in frame $\mathcal{F}^\prime$.
+In case the illuminated body is not occulted we have
+\begin{equation}
+    f_{0}=\mathbb{r}^{\top}(b) \mathbf{I}\left(b, \theta^{\prime \prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime \prime}\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+    \label{eq:starry_ref_phase}
+\end{equation}
+where $\theta^{\prime\prime}=0$ is the angle of the terminator in frame
+$\mathcal{F}^{\prime\prime}$ by construction. $\mathbf{R}^{\prime\prime}$ rotates the
+body through an angle $\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)$ so the
+semi-major axis of the terminator is aligned with the $x^{\prime \prime}$ axis.
+The solutions to integrals
+$\mathbb{r}^{\top}(b)$ and $\mathbb{s}^{\top}\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right)$
+are anything but trivial, they are derived in the appendices of
+\citet{2021arXiv210306275L}. The key point is that the model is once again linear and
+the matrices in Equations~\ref{eq:starry_ref} and \ref{eq:starry_ref_phase} need to be
+computed only once for a specific occultation/illumination geometry.
+\citet{2021arXiv210306275L} also derives a solution for the total flux in case the
+light source is extended (the assumption that $r_s\gg 1$ is not satisfied) and
+in case the body is not a perfect Lambertian reflector.
+
+\subsection{The starry code}
+The method described in the previous section is implemented in the
+\textsf{Python} package
+\textsf{starry}\footnote{\url{https://starry.readthedocs.io/en/latest/}}.
+\textsf{starry} is many orders of magnitude faster and more accurate than
+similar codes which rely on numerical integration. It enables computation of
+rotational light curves, planetary transits, secondary eclipse light curves and
+planet-planet occultations in both emitted and reflected light. Recent
+extensions to \textsf{starry}, not mentioned in Section~\ref{sec:occultations},
+extended the spherical harmonic formalism to the problem of Doppler imaging
+\citep{2021arXiv211006271L} and add support for computing occultation light
+curves of oblate bodies \citep{2022ApJ...925..185D}. In
+Chapters~\ref{ch:mapping_io} and \ref{ch:mapping_exoplanets} we make extensive
+use of \textsf{starry} to explore solutions to the \emph{inverse problem} --
+how can we obtain an estimate of the map coefficients $\mathbf{y}$ which best
+explain the observed light curve? \textsf{starry} abstracts the complexities of
+computing the total flux and allows us to treat this problem as a linear
+problem of the form
+\begin{equation}
+    f = \mathbf{X}\mathbf{y}
+\end{equation}
+where  $\mathbf{X}$ is computed by \textsf{starry} depending on the geometry of an
+occultation or transit event. The inverse problem is highly degenerate
+and making progress requires  imposing some prior structure on the solution
+$\mathbf{y}$.
+
+In Chapter~\ref{ch:mapping_io} we use \textsf{starry} to model occultation
+light curves of a Solar System object -- Jupiter's moon Io, as it is occulted
+by Jupiter. Thanks to the comparatively close distance to Io relative to
+objects outside of the Solar System, these light curves are very high quality
+we are able to fit very high order maps $l=20$ and test our models in a high
+signal-to-noise regime. We also discuss how possible approaches to modeling
+time-dependent maps in which case the spherical harmonic coefficients are
+time-dependent: $\mathbf{y}=\mathbf{y}(t)$. In
+Chapter~\ref{ch:mapping_exoplanets} we tackle a very similar problem but in a
+very different regime of exoplanet eclipse mapping where the signal-to-noise is
+orders of magnitude worse than in the case of Io. In both cases we mostly focus
+on thermal light curves rather than light curves in reflected light.
+
 \section{Bayesian statistics}
 \label{sec:bayes}
+Having covered the basic physics behind gravitational lensing and occultation mapping I
+now turn to the subject of statistical inference in general and more focus more
+specifically on Bayesian statistics in particular because it provides a straightforward
+and elegant theoretical framework for solving inverse problems of the kind we mentioned
+in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}.
+
 \subsection{Probability theory}
@@
 \label{sec:inference}
+\subsection{The curse of dimensionality}
 % check that paper on physics and ML by that MIT guy, there's a good stuff there. 
 % check also Ian Murray's lecture notes  on MLPR
-\subsection{The curse of dimensionality}
+
 \subsection{There's no free lunch}

### 2022-06-04 src/figures/spherical_harmonics.py

@@ -1,0 +1,62 @@
+import starry
+import matplotlib.pyplot as plt
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+ydeg = 5
+fig, ax = plt.subplots(ydeg + 1, 2 * ydeg + 1, figsize=(12, 6))
+fig.subplots_adjust(hspace=0)
+for axis in ax.flatten():
+    axis.set_xticks([])
+    axis.set_yticks([])
+    axis.spines["top"].set_visible(False)
+    axis.spines["right"].set_visible(False)
+    axis.spines["bottom"].set_visible(False)
+    axis.spines["left"].set_visible(False)
+for l in range(ydeg + 1):
+    ax[l, 0].set_ylabel(
+        "l = %d" % l,
+        rotation="horizontal",
+        labelpad=20,
+        y=0.38,
+        fontsize=10,
+        alpha=0.5,
+    )
+for j, m in enumerate(range(-ydeg, ydeg + 1)):
+    ax[-1, j].set_xlabel("m = %d" % m, labelpad=10, fontsize=10, alpha=0.5)
+
+# Loop over the orders and degrees
+map = starry.Map(ydeg=ydeg)
+for i, l in enumerate(range(ydeg + 1)):
+    for j, m in enumerate(range(-l, l + 1)):
+
+        # Offset the index for centered plotting
+        j += ydeg - l
+
+        # Compute the spherical harmonic
+        # with no rotation
+        map.reset()
+        if l > 0:
+            map[l, m] = 1
+
+        # Plot circle
+        circ = plt.Circle((0, 0), 1.01, edgecolor="k", facecolor="none", linewidth=1.0)
+        ax[i, j].add_patch(circ)
+
+        # Plot the spherical harmonic
+        ax[i, j].imshow(
+            map.render(),
+            cmap="OrRd",
+            interpolation="none",
+            origin="lower",
+            extent=(-1, 1, -1, 1),
+            zorder=-1,
+        )
+        ax[i, j].set_xlim(-1.1, 1.1)
+        ax[i, j].set_ylim(-1.1, 1.1)
+        ax[i, j].set_aspect(1)
+        ax[i, j].set_rasterization_zorder(0)
+
+# Save to file
+plt.savefig("spherical_harmonics.pdf", bbox_inches="tight", dpi=200)

### 2022-06-06 src/ms.tex
making progress on the stats section

@@
 \newcommand{\ud}{\,\mathrm{d}}
-\renewcommand{\vec}[1]{\bm{#1}}
+\renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}
 
@@
 observer, lens and the source is rectilinear (acceleration can be neglected).
-We take $\boldsymbol{\theta}_S$ to be the angular position of the source star
-on the plane of the sky and $\boldsymbol{\mu}_L$ to be its proper motion
-vector. Likewise for the lens. We thus have
+We take $\vec{x}_S$ to be the angular position of the source star on the plane
+of the sky and $\boldsymbol{\mu}_L$ to be its proper motion vector. Likewise
+for the lens. We thus have
 \begin{align}
-    \boldsymbol{\theta}_S(t) & =\boldsymbol{\theta}_{S, 0}+
+    \vec{x}_S(t) & =\vec{x}_{S, 0}+
     \left(t-t_{0}\right) \boldsymbol{\mu}_S
     \label{eq:source_position}      \\
-    \boldsymbol{\theta}_L(t) & =\boldsymbol{\theta}_{L, 0}
+    \vec{x}_L(t) & =\vec{x}_{L, 0}
     +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
@@
 \begin{equation}
-    \boldsymbol{u}(t) \equiv \frac{\boldsymbol{\theta}_L(t)
-        -\boldsymbol{\theta}_S(t)}{\theta_E}=
-    \frac{\boldsymbol{\theta}_{LS, 0}}{\theta_E}
+    \boldsymbol{u}(t) \equiv \frac{\vec{x}_L(t)
+        -\vec{x}_S(t)}{\theta_E}=
+    \frac{\vec{x}_{LS, 0}}{\theta_E}
     +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
@@
 where
-$\boldsymbol{\theta}_{L S, 0}\equiv\boldsymbol{\theta}_{L,0}-\boldsymbol{\theta}_{S, 0}$
+$\vec{x}_{L S, 0}\equiv\vec{x}_{L,0}-\vec{x}_{S, 0}$
 is the relative position at $t_0$ and
@@
 relative proper motion.
-Since $\boldsymbol{\mu}_{LS}$ and $\boldsymbol{\theta}_{L S, 0}$ are
+Since $\boldsymbol{\mu}_{LS}$ and $\vec{x}_{L S, 0}$ are
 perpendicular to each other, it follows that the magnitude of the relative
@@
 \end{equation}
-where we have defined $u_0\equiv |\boldsymbol{\theta}_{L S, 0}|/\theta_E$
+where we have defined $u_0\equiv |\vec{x}_{L S, 0}|/\theta_E$
 and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
@@
 \begin{align}
-    \boldsymbol{\theta}_S(t) & = \boldsymbol{\theta}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
+    \vec{x}_S(t) & = \vec{x}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
     +\pi_S\,\boldsymbol{\zeta}(t)                            \\
-    \boldsymbol{\theta}_L(t) & = \boldsymbol{\theta}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
+    \vec{x}_L(t) & = \vec{x}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
     +\pi_L\,\boldsymbol{\zeta}(t)
@@
 source plane, given by the the dimensionless vector
-$\mathbf{y}=\bm\beta/(D_\textrm{S}\theta_E)$ and the position of the images in
-the lens plane, given by given by
-$\mathbf{x}=\bm\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now refers to
-the angular Einstein radius corresponding to the total mass $M$. The lens
-equation then contains a sum over the deflection angles $\bm\alpha_i$
-corresponding to each point mass $m_i$\footnote{This is because we are working
-    within the linearized GR framework so the deflection angles are additive.}:
-\begin{equation}
-    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\bm\alpha_i(\mathbf{x},\mathbf{x}_i)
-\end{equation}
-where $\bm\alpha_i$ is the deflection angle due to the $i$th lens.
+$\mathbf{y}=\boldsymbol\beta/(D_\textrm{S}\theta_E)$ and the position of the
+images in the lens plane, given by given by
+$\mathbf{x}=\boldsymbol\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now
+refers to the angular Einstein radius corresponding to the total mass $M$. The
+lens equation then contains a sum over the deflection angles
+$\boldsymbol\alpha_i$ corresponding to each point mass $m_i$\footnote{This is
+    because we are working within the linearized GR framework so the deflection
+    angles are additive.}:
+\begin{equation}
+    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\boldsymbol\alpha_i(\mathbf{x},\mathbf{x}_i)
+\end{equation}
+where $\boldsymbol\alpha_i$ is the deflection angle due to the $i$th lens.
 Using Equation~\ref{eq:lens_equation_dimensionless}, we obtain
@@
 \citep{2019AJ....157...64L,2021arXiv210306275L}.
-\subsection{The Starry framework}
+\subsection{The starry algorithm}
 \begin{figure}[t]
@@
 
-\section{Bayesian statistics}
-\label{sec:bayes}
+\section{Statistical inference -- theory}
+\label{sec:inference_theory}
 Having covered the basic physics behind gravitational lensing and occultation mapping I
-now turn to the subject of statistical inference in general and more focus more
-specifically on Bayesian statistics in particular because it provides a straightforward
-and elegant theoretical framework for solving inverse problems of the kind we mentioned
-in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}.
+now turn to the subject of statistical inference in general and more specifically --
+Bayesian statistical inference. Bayesian (as opposed to frequentist)
+provides a straightforward and elegant theoretical framework for solving inverse
+problems of the kind we mentioned
+in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}. These are problems
+where we have little data per object of interest, the data is very noisy,
+the physical models are relatively well understood but their parameters don't map
+to observables in a straightforward way and there are multiple competing, sometimes
+physically  quite different, models which provide a good explanation for the data.
+Having said that, I am not particularly dogmatic about the use of Bayesian methods
+and I will occasionally steer away from the classical approach and point out where
+a less Bayesian approach is often superior in practice.  What is most important is the
+use of \emph{probabilities} for encoding uncertainty about the physical world.
+
+I start by reviewing basic probability theory and briefly mention the
+frequentist interpretation. I then introduce two key concepts, the likelihood
+function -- a recipe for generating plausible datasets given model parameters,
+and priors -- what we know about the model parameters prior to gathering any
+data. Next, I briefly summarize fundamental concepts such as the curse of
+dimensionality and the no-free-lunch-theorem. Finally I end the section by
+comparing machine learning to statistical inference within the physical
+sciences.
 
 \subsection{Probability theory}
-\subsection{The meaning of probability}
-\subsection{Probability as frequency of events in repeated trials}
-\subsection{Probability as degrees of belief}
-\subsection{The likelihood function}
-% mention the likelihood principle
-\subsection{Priors}
-\subsection{On machine learning and the difference between explanation and prediction}
-% mention that even if we have a model focused on inference, it might be best if we judge 
-% its quality through its predictive performance
-
-\section{Inference}
-\label{sec:inference}
-\subsection{The curse of dimensionality}
-% check that paper on physics and ML by that MIT guy, there's a good stuff there. 
-% check also Ian Murray's lecture notes  on MLPR
-
-\subsection{There's no free lunch}
-% exploration vs. explotation 
-\subsection{Optimization vs. sampling}
+There are two different interpretations of probability. In the
+\emph{frequentist} interpretation probabilities are \emph{frequencies of
+    events} which are repeated in a large number of trials (think of a repeated
+coin tosses). An alternative to the frequentist interpretation is the
+\emph{Bayesian} interpretation in which probabilities represent \emph{degrees
+    of belief} about the subject of interest. In both cases probabilities are
+necessarily positive numbers between zero and one. The major advantage of the
+Bayesian interpretation is that it allows one to reason about the probability
+of one-off events such as ``person X wins the election'' or ``this microlensing
+event is a binary lens event''. The differences between the two interpretations
+are not just philosophical because which interpretation one subscribes to can
+determine how one approaches a particular statistical inference problem and
+which sets of tools one uses. Both approaches are commonly used within the
+physical sciences depending on the field and the problem at hand. For instance,
+in case of particle physics frequentist methods are much more common because
+particle physics experiments often involve billions of repeated experiments in
+particle accelerators. One the other hand, Bayesian methods are more popular in
+astronomy because we cannot do true repeated experiments with astrophysical
+objects.
+
+The mathematical rules for manipulating probabilities are quite
+straightforward. In probability theory we use \emph{random variables} which can
+be either discrete or continuous and their possible values are described with
+probability distributions over all possible values the variable is allowed to
+take. For discrete random variables we can talk of probabilities that a given
+variable assumes a particular value, say $5$, but with continuous random
+variables we can only talk of probabilities that a given variable is in a
+particular range of values, say $[0,1]$. Now let's consider a continuous
+\emph{random variable} $x$ with an associated probability density function
+(pdf) $p(x)$\footnote{The notation $p(x)$ we use is somewhat overloaded. In
+    statistics, the proper notation for a probability density function is
+    $p_{X}(x)$ where $X$ is a random variable and $x$ is a particular realization
+    of that random variable.} Since $x$ has to take on \emph{some values} within
+its domain we have the requirement that
+\begin{equation}
+    \int p(x)\,\ud x=1
+    \label{eq:normalization_condition}
+\end{equation}
+where the integral is taken over the entire domain of $x$.
+In order for Equation~\ref{eq:normalization_condition} to hold,
+$p(x)$ has to have units of $x^{-1}$ so that
+the integral is a dimensionless number.
+
+In general, we are usually dealing with several parameters at once in which
+case we are interested in \emph{joint} probability distributions over all the
+parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
+density function $p(x_1,x_2)$ which when integrated over a particular interval
+gives the probability that both $x_1$ and $x_2$ are contained within that
+interval. These joint probabilities can be decomposed into \emph{conditional
+    probabilities} using the so called product rule:
+\begin{align}
+    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
+    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
+    \label{eq:product_rule}
+\end{align}
+where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
+    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
+valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
+Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
+\begin{equation}
+    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
+    \label{eq:bayes_theorem}
+\end{equation}
+Bayes's theorem is simply a rule for converting one kind of a conditional probability
+into another.
+Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
+\emph{marginalize} parameters that we are not interested in to obtain a distribution over
+parameters of interest\footnote{
+    It is very common in astronomy to have a model
+    which consists of a set of parameters $\boldsymbol x_1$ of physical signal of interest and another set
+    of ``nuisance parameters'' $\boldsymbol x_2$ which often describe the noise properties. In those cases
+    we need to marginalize over parameters $\boldsymbol x_2$ to obtain a distribution over $x_1$.}:
+\begin{equation}
+    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
+    x_2)\,p(x_2)\textrm{d}x_2
+\end{equation}
+where we applied Bayes's theorem in the second inequality.
+In cases where the parameter space is high dimensional these integrals might be
+impossible to calculate analytically but they can (sometimes) be estimated using sampling
+algorithms as we shall se in subsequent sections.
+In some cases, we may be able to factor out $p(x_1,x_2)$ as
+If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
+\begin{equation}
+    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
+\end{equation}
+Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
+of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
+of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
+\begin{equation}
+    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
+    \label{eq:likelihood_indep}
+\end{equation}
+If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
+identically distributed (``iid''). If the data is independent but not identically distributed
+we say that the uncertainties are \emph{heteroscedastic}.
+The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
+uncertainties.
+
+\subsubsection{Expectation values and the mode}
+The \emph{mean} or the \emph{expected value} of a distribution $p(\vec{x})$,
+usually denoted by $\mu$, is a measure of the ``center'' of a distribution and
+it is defined as
+\begin{equation}
+    \mathbb{E}[x] \equiv\int x \,p(x) \ud x
+\end{equation}
+Similarly, the \emph{variance} is a measure of the ``spread'' of a distribution:
+\begin{align}
+    \mathbb{V}[x] & \equiv\mathbb{E}\left[(x-\mu)^{2}\right]=\int(x-\mu)^{2} p(x) d x                                           \\
+                  & =\int x^{2} p(x) \ud x+\mu^{2} \int p(x) \ud x-2 \mu \int x p(x) \ud x=\mathbb{E}\left[x^{2}\right]-\mu^{2}
+\end{align}
+The variance is usually denoted by $\sigma^2$ and the square root of $\sigma^2$ is the
+\emph{standard deviation}.
+For a \emph{multivariate} random variable with $D$ dimensions then the variance generalizes to
+a symmetric, positive semi definite matrix called the \emph{covariance matrix}:
+\begin{align}
+    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\top}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                                \\
+                                & =\left(\begin{array}{cccc}
+                                             \mathbb{V}\left[x_{1}\right]                & \operatorname{Cov}\left[x_{1}, x_{2}\right] & \cdots & \operatorname{Cov}\left[x_{1}, x_{D}\right] \\
+                                             \operatorname{Cov}\left[x_{2}, x_{1}\right] & \mathbb{V}\left[x_{2}\right]                & \cdots & \operatorname{Cov}\left[x_{2}, x_{D}\right] \\
+                                             \vdots                                      & \vdots                                      & \ddots & \vdots                                      \\
+                                             \operatorname{Cov}\left[x_{D}, x_{1}\right] & \operatorname{Cov}\left[x_{D}, x_{2}\right] & \cdots & \mathbb{V}\left[x_{D}\right]
+                                         \end{array}\right)
+\end{align}
+
+In the general case, we can compute expectation values of an arbitrary function
+$g(\vec{x})$ under the probability distribution $p(\vec{x})$:
+\begin{equation}
+    \mathbb{E}\left[g(\vec{x})\right]=\int g(\vec{x})\,p(\vec{x})\ud x
+    \ud\theta
+\end{equation}
+Some examples of $g$ include higher moments of the distribution, the median and
+quantile estimates. The \emph{mode} of $p(\vec{x})$
+\begin{equation}
+    \vec{x}^{*}=\underset{\vec{x}}{\operatorname{argmax}}\,p(\vec{x})
+\end{equation}
+on the other hand is not an expectation value. It is the global maximum of the
+function $p(\vec{x})$.
+Estimating the mode involves computing the \emph{derivatives} of the density
+$p(\vec{x})$, whereas expectation values require computing
+\emph{integrals} over the space $\vec{x}$. Computing either
+the mode or expectation values
+for nontrivial densities $p(\vec{x})$ is a very hard and often impossible problem
+(more on this in Section~\ref{sec:inference_practice}).
+
+\subsubsection{Central Limit Theorem}
+Consider $N$ \emph{independent and identically distributed} random variables
+with pdf's $p_n(x)$ which are not necessarily Gaussian. Let $S_N$ be the sum
+the random variables. The \emph{Central Limit Theorem} (CLT) says that as $N$
+increases, the distributions of this sum approaches the Gaussian distribution
+\begin{equation}
+    p\left(S_{N}=u\right)=\frac{1}{\sqrt{2 \pi N \sigma^{2}}} \exp \left(-\frac{(u-N \mu)^{2}}{2 N \sigma^{2}}\right)
+\end{equation}
+and the distribution of the quantity
+\begin{equation}
+    Z_{N} \equiv \frac{S_{N}-N \mu}{\sigma \sqrt{N}}=\frac{\bar{x}-\mu}{\sigma / \sqrt{N}}
+\end{equation}
+where $\bar{x}$ is the \emph{sample mean} $\frac{1}{N}\sum_{i=1}^Nx_i=S_N/N$
+converges to the Gaussian distribution with zero mean and unit variance.
+Because lots of random variables representing real world processes can be seen as sums
+of i.i.d. independent random variables the Gaussian distribution is all over the place
+in statistics. One should keep in mind that the assumptions underlying the CLT are
+not always satisfied and also the convergence towards a Gaussian is
+more rapid in the bulk of the distribution around the mean than in the tails of the
+distribution.
+
+\subsubsection{Change of variables}
+Given a univariate random variable $x$ the effect of some deterministic
+transformation $y(x)$ is to stretch and shift the distribution $p(x)$. For an
+infinitesimal interval the \emph{probability mass} is the same in both
+variables so $p(x) \ud x=p(y) \ud y$ which yields the \emph{change of variables
+    formula}
+\begin{equation}
+    p_{y}(y)=p_{x}(x)\left|\frac{\ud x}{\ud y}\right|
+\end{equation}
+In the multivariate case, if $\vec{f}$ is an invertible function mapping $\mathbb{R}^n$
+to $\mathbb{R}^n$ with inverse $\vec{g}$, the pdf of $\vec{y}=\vec{f}(\vec{x})$ is
+\citep{murphy_book_2022}
+\begin{equation}
+    p_{y}(\vec{y})=p_{x}(\vec{g}(\vec{y}))\left|\operatorname{det}\mathbf{J}_{g}(\vec{y})\right|
+\end{equation}
+where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\top}}$ is the Jacobian of $\vec{g}$
+and $\operatorname{det}|\mathbf{J}(\vec{y})|$ is the determinant of the Jacobian.
+
+\subsubsection{Kullback Leibler divergence}
+A common way of measuring ``similarity'' between two distributions $p(x)$ and
+$q(x)$ is using the \emph{Kullback-Leibler divergence} (KL divergence for
+short):
+\begin{equation}
+    D_{\mathrm{KL}}(p \| q)=\int_{-\infty}^{\infty} p(x) \ln \left(\frac{p(x)}{q(x)}\right) \ud x
+    \label{eq:kl_divergence}
+\end{equation}
+where $D_{\mathrm{KL}}(p \| q)$ should be read as ``the KL divergence from p to q''.
+This is not a measure of ``distance'' between the two distributions because
+it is not a symmetric quantity. The KL divergence is best understood as a quantity which
+quantifies \emph{how surprised one would be if one expected to encounter distribution $p(x)$
+    but instead saw distribution
+    $q(x)$}\footnote{\url{https://wiki.santafe.edu/images/a/a8/IT-for-Intelligent-People-DeDeo.pdf}}.
+This quantity is important in the context of variational inference which is covered in
+Section~\ref{sec:inference_practice}.
+
+\subsection{Bayes' theorem in practice}
+\label{ssec:likelihood_function}
+Consider a general inference problem where we collect some data $\mathcal{D}$ and we are interested
+in the joint probability distribution for parameters $\boldsymbol\theta$ of particular model.
+We can then rewrite Bayes' theorem as
+\begin{equation}
+    p(\vec{\theta} | \mathcal{D})=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{p(\mathcal{D})}=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{\int p\left(\vec{\theta}^{\prime}\right) p\left(\mathcal{D} |\vec{\theta}^{\prime}\right) d \vec{\theta}^{\prime}}
+    \label{eq:bayes_theorem2}
+\end{equation}
+In Equation~\ref{eq:bayes_theorem2} the term $p(\mathcal{D}|\vec{\theta})$  is a probability
+distribution over the data $\mathcal{D}$ given the parameters of the model $\vec{\theta}$.
+When this term is seen as a function of the parameters given a fixed dataset, it is
+called the \emph{likelihood function }  or the \emph{likelihood} for short. $p(\boldsymbol\theta)$
+is to so called \emph{prior} distribution over $\vec{\theta}$ and it represents our beliefs
+about $\vec{\theta}$ prior to having observed the data $\mathcal{D}$. When the prior distribution is
+multiplied by the likelihood we obtain, up to a normalizing constant $p(\mathcal{D})$, the posterior
+distribution over $\vec{\theta}$ given the data $\mathcal{D}$.
+The best way to think about Equation~\ref{eq:bayes_theorem2} is seeing at as a rule
+for updating beliefs. Our prior beliefs about the parameters of the model
+$p(\mathrm{parameters})$ should be updated to posterior beliefs $
+    p(\mathrm{parameters} | \mathrm{data})$ by making use of the likelihood
+$p(\mathrm{data} | \mathrm{parameters})$. If the data is useful, the likelihood is generally
+much narrower than the prior pdf and the choice of priors does not significantly
+influence the final results.
+
+\subsubsection{The likelihood}
+Let's cover each of these three terms in more detail, starting with the most
+important one -- the likelihood. The likelihood can be seen as a \emph{recipe
+    for generating plausible datasets $\mathcal{D}$} (a \emph{generative model})
+given a specific realization of the parameters $\vec{\theta}$. For the kinds of
+models we work with in astronomy, the likelihood consists of two parts. The
+first part is (usually deterministic) function which describes the physical
+system of interest. This is for instance the predicted flux for a binary
+microlensing event with a particular trajectory. The second part is the
+\emph{noise model} which describes the statistical properties of the noise
+inherent in any physical measurement. We often assume that the noise is
+independent and Gaussian. It is very common to use the likelihood as is and
+optimize it with respect to the model parameters $\vec{\theta}$ in a procedure
+called \emph{maximum likelihood estimation} (MLE) to find a single point in the
+parameter space $\vec{\theta}^\star$ which maximizes the likelihood (more on
+this in Section~\ref{ssec:mle_estimation}).
+
+We can also marginalize the likelihood over a subset of parameters. Say that we
+have a likelihood function $p(\vec{\theta} ,\vec{\alpha}|\mathcal{D})$ where
+$\vec{\alpha}$ is a set of \emph{nuisance parameters} which we are not
+particularly interested in. To marginalize over these nuisance parameters we
+need to introduce a prior\footnote{Not introducing a prior leads to a
+    dimensionally incorrect result.} $p(\vec{\theta}|\vec{\theta})$ which may
+potentially also depend on the $\vec{\theta}$ and the marginalized likelihood
+is then
+\begin{equation}
+    p(\mathcal{D} | \vec{\theta})=\int p(\mathcal{D} | \vec{\theta}, \vec{\alpha})\,p(\vec{\alpha} |\vec{\theta}) \ud \vec{\alpha}\quad .
+\end{equation}
+The problem structure where where $\vec{\alpha}$ is some set of parameters
+we don't particularly care about such and $\vec{\theta}$ are the physical
+parameters of interest is very common in astronomy. For instance, in single lens gravitational microlensing estimating
+the distribution $p(t_E)$ over the parameter of interest $t_E$ requires marginalizing over
+parameters such as $F_S, F_B, t_0$ and $u_0$.
+
+Not all statistical models can be written in terms of the likelihood function.
+Most machine learning models for example do not have a simple probabilistic
+interpretation\footnote{Indeed, finding probabilistic interpretations of
+    complex machine learning models such as deep neural networks is an active
+    research area.}. If we can write down the likelihood function for a certain
+problem, we say that we have a \emph{generative model for the data} because the
+likelihood can be used to generate mock data sets.
+
+\subsubsection{Priors}
+The second term of interest is the prior $p(\vec{\theta})$. In principle, the
+prior pdf-s should capture all of the prior information we might have about the
+model parameters prior to gathering new data; in practice, we don't want to
+bias our results so we often try to use ``weakly informative'' priors -- priors
+that are informative enough that they exclude unreasonable parameter values but
+aren't so narrow that they rule out sensible values which may be favored by the
+likelihood. For example, we might have prior information such as ``the mass of
+the planet should be positive'' in which case a suitable weakly informative
+prior on the mass parameter should first of all exclude negative values but
+also potentially exclude values which are nonsensically large for the problem
+at hand. There are also the so called ``non-informative prior'' such as
+Jeffrey's prior and maximum entropy priors. Although these priors can be of use
+in some cases, in general \textbf{there no such thing as a truly non
+    informative prior}. Although a prior for a particular parameter $\theta$ might
+be uninformative in one set of coordinates, it is often informative for a
+certain transformation of the parameter $g(\theta)$. For example, a uniform
+prior in $\theta$ is not uniform in $\ln\theta$. Having said that, it is often
+worth the effort to construct priors which respect certain symmetries of the
+problem the we know to be true.
+
+Priors should not be seen in isolation, rather, they should be seen in the
+context of the likelihood \citep{arXiv:1708.07487}. To see that this is the
+case consider drawing samples from the prior $\vec{\theta}\sim
+    p(\vec{\theta})$, plugging those values into the likelihood distribution
+$p(\mathcal{D}|\vec{\theta})$ and then generating mock datasets, a process
+called \emph{prior predictive checking}. This process makes it easy to spot
+priors which in combination with the likelihood yield predictions of what the
+data should like like which do not match our expectations. An excellent highly
+in depth study on choosing priors is available
+\href{https://betanalpha.github.io/assets/case_studies/prior_modeling.html}{here}.
+
+Depending on how informative the data is we differentiate between two different
+regimes. We say that the inference process is \emph{data driven} if the data
+constrains the model parameters so well that the choice of the priors is
+effectively irrelevant. The other regime is said to be \emph{prior driven},
+when the data is so weakly informative that the shape of the posterior pdf-s is
+similar to the prior pdf. In general, in any statistical analysis involving
+priors, one should check the sensitivity of the scientific conclusion to
+assumptions of different priors. No statistical inference procedure is free of
+assumptions but the use of priors and likelihoods makes those assumptions more
+explicit.
+
+Priors are sometimes the goal of the inference. For example, if the prior on
+the prior pdf $p(\vec{\theta})$ itself depends on some other set of parameters
+$\vec{\beta}$, then we can marginalize out all of the $\vec{\theta}$ parameters
+and we are left with a likelihood for the parameters $\vec{\beta}$ of the
+prior.
+\begin{equation}
+    p(\mathcal{D}\lvert\vec{\beta})=\int p(\mathcal{D}\lvert\vec{\theta})\,p(\vec{\theta}\lvert
+    \vec{\beta})\ud \vec{\theta}
+\end{equation}
+This can then be used, using Bayes's theorem, to calculate the posterior
+over the prior parameters $\bm\beta$.
+\begin{equation}
+    p(\bm\beta\lvert\mathcal{D})\propto p(\mathcal{D}\lvert\vec{\beta})
+    \,p(\bm\beta)
+\end{equation}
+The parameters $\bm\beta$ characterizing the prior are usually called
+\emph{hyperparameters} or \emph{population level parameters}. Similarly, the priors
+for those parameters are called \emph{hyperpriors}.
+The whole process is called \emph{hierarchical Bayesian inference}
+(in astronomy literature),  or \emph{multilevel modelling, hierarchical modelling,
+    nested modelling, mixed models or random effects models}
+(in statistics and adjacent fields).
+Hierarchical modelling is an very useful method for estimating
+population-wide properties given data about individual objects. For example,
+in microlensing observations of exoplanets the individual parameters could be the
+masses of individual planets while the hyperparameters would describe the
+population the shape of the planetary mass function. We shall return to this topic
+in Chapter~\ref{ch:microlensing}.
+
+\subsubsection{The evidence}
+The final component of Equation~\ref{eq:bayes_theorem2} is the normalization
+constant $ p(\mathcal{D})$ in the denominator:
+\begin{equation}
+    \mathcal{Z}\equiv p(\mathcal{D})=\int p(\mathcal D\lvert\boldsymbol\theta)\,p(\boldsymbol\theta)\textrm{d}\boldsymbol\theta
+    \label{eq:evidence}
+\end{equation}
+is often called the \emph{fully marginalized likelihood} because
+it is a likelihood marginalized over all of the parameters of the model.
+Sometimes
+it is also called the \emph{Bayesian evidence}.
+$\mathcal{Z}$ is general extremely difficult to calculate because it is an
+integral over a potentially very high dimensional parameter space $\vec{\theta}$.
+Fortunately, in most cases, we can obtain the posterior distribution without having
+to calculate $\mathcal{Z}$.
+
+%% mention the likelihood principle
+\section{Statistical inference}
+\label{sec:inference_practice}
+In the previous section I briefly summarized some key aspects of probability theory
+and Bayesian statistics but I haven't discussed how to apply these ideas in practice. That
+is the subject of statistical inference. I start by introducing some very fundamental
+concepts such as the \emph{curse of dimensionality} and the \emph{no free lunch theorems}.
+I then cover linear models which are the foundation of all statistics and also relevant
+for the occultation mapping work since the problem is linear.
+Next, I discuss various computational inference methods such
+as optimization methods, Markov chain Monte Carlo and Variational Inference.
+Finally, I discuss model comparison and evaluation.
+
+\subsection{The curse of dimensionality and the no-free-lunch theorem}
+The \emph{curse of dimensionality} refers to a phenomenon where the
+\emph{volume} of \emph{possible} explanations for the data grows exponentially
+with increasing dimension of the problem. Consider the following
+example\footnote{ Taken from the excellent paper by \citet{arXiv:2104.00008}.
+}. There are images consisting of $n$ pixels where each of the pixels is either
+a black (0) or white (1). The number of possible images is $2^n$ which grows
+exponentially. If each image out of the $2^n$ possible images is described by
+one of two labels $A$, or $B$ then the total number of possible ways of
+labeling the set of images is $2^{2^n}$ which is a stupidly large number. If
+$n=9$, the number of ways to classify the set of all 9-bit black and white
+images is greater than the number of atoms in the observable universe. If the
+labels don't correlate with the properties of the images then one strategy for
+solving this problem would be the lookup table approach -- memorize each image
+with its label. The famous \emph{no-free-lunch} (NFLT) theorem proved by the
+mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
+impossible to construct an algorithm which does better than the lookup table
+approach.
+
+Of course the assumption that the labels aren't correlated to features in the
+images is silly for any meaningful problem. For instance, all images of cats
+have certain things in common as do images of galaxies. The no-free-lunch
+theorem doesn't say that it is impossible to construct an algorithm to classify
+images of cats, it only says that there is no best algorithm \emph{if we
+    average over all possible inputs to the problem}. To construct an algorithm
+which works in practice we need to impose structure on the problem, what
+machine learning researchers call \emph{inductive bias}. A set of assumptions
+which work in one domain may not work well in another domain and no single
+model will work well in all domains.
+
+\subsection{Maximum likelihood estimation}
+One of the most popular inference methods is called \emph{maximum likelihood
+    estimation} (MLE). Maximum likelihood estimation answers the question:
+\textbf{what choice of parameters $\vec{x}$ makes my data most likely} as
+opposed to the question \textbf{what are the most likely values of the
+    parameters $\vec{x}$ having observed the data $\mathcal{D}$}\footnote{The
+    distinction between the two questions is subtle, but important. The parameters
+    which maximize the likelihood might be very different from those which are most
+    ``sensible'' and which best describe the data.} which necessitates the
+multiplication of the likelihood with a prior. Since we don't incorporate the
+prior, it is not a Bayesian method. The output of MLE is a \emph{point
+    estimate} of the parameters $\hat{\vec{\theta}}$. In a Bayesian viewpoint, the
+MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.
+
+\subsubsection{MLE for a Gaussian linear model}
+\subsection{The Laplace approximation}
+\subsection{Gaussian linear models}
+
+%\subsection{Machine learning and the difference between inference and prediction}
+%% mention that even if we have a model focused on inference, it might be best if we judge 
+%% its quality through its predictive performance
+%
 \subsection{Sampling and optimization methods}
-\subsubsection{Maximum likelihood}
-\subsubsection{The Laplace approximation}
-\subsubsection{Variational inference}
-\subsubsection{Importance sampling}
-\subsubsection{Rejection sampling}
-\subsubsection{Markov Chain Monte Carlo}
-\subsubsection{Hamiltonian Monte Carlo}
-\subsubsection{Other notable methods}
-
-\section{Model validation and comparison}
-\label{sec:model_comparison}
-
+%\subsubsection{The Laplace approximation}
+%\subsubsection{Variational inference}
+%\subsubsection{Importance sampling}
+%\subsubsection{Rejection sampling}
+%\subsubsection{Markov Chain Monte Carlo}
+%\subsubsection{Hamiltonian Monte Carlo}
+%\subsubsection{Other notable methods}
+%
+%\subsection{Model validation and comparison}
+%\label{ssec:model_comparisonGaussian linear }
+%
 ``Furthermore, in the future, a blind search for planetary signals will not
@@
 
-\subsection{Cross validation}
-% discuss vanilla k-fold CV, stress out just how important and central CV is to science
-\subsubsection{Pareto smoothed importance sampling LOO}
-\subsection{Model comparison}
-% don't do it!
-
-\section{Putting it all together -- \emph{probabilistic programming}}
-\label{sec:programming}
-\subsection{The success of machine learning}
-% that quote from Yann LeCun on differentiable code 
-\subsection{Automatic differentiation}
-\subsubsection{Forward mode autodiff}
-\subsubsection{Reverse mode autodiff or backpropagation}
-\subsubsection{Jacobian vector products}
-\subsection{JAX}
-\subsection{Numpyro}
+%\subsubsection{Cross validation}
+%% discuss vanilla k-fold CV, stress out just how important and central CV is to science
+%\subsubsubsection{Pareto smoothed importance sampling LOO}
+%\subsubsection{Model comparison}
+%% don't do it!
+%
+%\section{Putting it all together -- \emph{probabilistic programming}}
+%\label{sec:programming}
+%\subsection{The success of machine learning}
+%% that quote from Yann LeCun on differentiable code 
+%\subsection{Automatic differentiation}
+%\subsubsection{Forward mode autodiff}
+%\subsubsection{Reverse mode autodiff or backpropagation}
+%\subsubsection{Jacobian vector products}
+%\subsection{JAX}
+%\subsection{Numpyro}
 
@@
 
-\end{document}
+\end{document}ii

### 2022-06-06 src/ms.tex
minor

@@
 solving this problem would be the lookup table approach -- memorize each image
-with its label. The famous \emph{no-free-lunch} (NFLT) theorem proved by the
+with its label. The famous \emph{no-free-lunch theorem} (NFLT) proved by the
 mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
@@
 \subsubsection{MLE for a Gaussian linear model}
-\subsection{The Laplace approximation}
+
 \subsection{Gaussian linear models}
@@
 \subsection{Sampling and optimization methods}
-%\subsubsection{The Laplace approximation}
-%\subsubsection{Variational inference}
-%\subsubsection{Importance sampling}
-%\subsubsection{Rejection sampling}
-%\subsubsection{Markov Chain Monte Carlo}
-%\subsubsection{Hamiltonian Monte Carlo}
-%\subsubsection{Other notable methods}
+\subsubsection{The Laplace approximation}
+\subsubsection{Variational inference}
+\subsubsection{Importance sampling}
+\subsubsection{Rejection sampling}
+\subsubsection{Markov Chain Monte Carlo}
+\subsubsection{Hamiltonian Monte Carlo}
+%\subsubsection{Other }
 %

### 2022-06-07 src/ms.tex
linear regression section

@@
 \usepackage{esint} % various fancy integral symbols
-\usepackage[bbgreekl]{mathbbol}
+%\usepackage[bbgreekl]{mathbbol}
 
@@
 \emph{marginalize} parameters that we are not interested in to obtain a distribution over
-parameters of interest\footnote{
-    It is very common in astronomy to have a model
-    which consists of a set of parameters $\boldsymbol x_1$ of physical signal of interest and another set
-    of ``nuisance parameters'' $\boldsymbol x_2$ which often describe the noise properties. In those cases
-    we need to marginalize over parameters $\boldsymbol x_2$ to obtain a distribution over $x_1$.}:
+parameters of interest:
 \begin{equation}
@@
 MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.
-
-\subsubsection{MLE for a Gaussian linear model}
-
-\subsection{Gaussian linear models}
+The MLE estimate is obtained by maximizing the logarithm\footnote{Because
+    probabilities in most region of the parameter space are very small numbers we
+    always use the log transform when optimizing likelihoods or posteriors.} of the
+likelihood $p(\mathcal{D}\lvert\vec{\theta})$ (or equivalently, minimizing the
+negative log likelihood) with respect to the parameters $\vec{\theta}$:
+\begin{equation}
+    \hat{\boldsymbol{\theta}}_{\mathrm{mle}}=\underset{\boldsymbol{\theta}}{\operatorname{argmax}}
+    \ln p\left(\mathcal{D}\lvert \boldsymbol{\theta}\right)
+\end{equation}
+
+\subsubsection{MLE for a multivariate Gaussian}
+The log-likelihood of multivariate Gaussian distribution without the
+unnecessary constants is
+\begin{equation}
+    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\log p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \log |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\top} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
+\end{equation}
+where $\vec{y}_n$ are the data points and $\vec{\Sigma}^{-1}$ is
+the inverse covariance matrix, also knowns as the \emph{precision matrix}.
+The MLE for the mean parameter $\hat{\vec{\mu}}$
+is given by the sample mean (empirical mean) of the data \citep{murphy_book_2022}:
+\begin{equation}
+    \hat{\boldsymbol{\mu}}=\frac{1}{N} \sum_{n=1}^{N} \boldsymbol{y}_{n}=\overline{\boldsymbol{y}}\quad .
+\end{equation}
+Similarly, the MLE of the the covariance matrix $\hat{\vec{\Sigma}}$ is equal to
+the empirical covariance matrix
+\begin{equation}
+    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\top}
+\end{equation}
+Thus, whenever we compute an empirical mean and the covariance for a given datasets we
+are implicitly assuming that the data is normally distributed and obtaining an MLE
+estimate of that distribution.
+
+\subsection{Least squares linear regression}
+In this section we cover the classical non Bayesian formulation of linear
+regression, also known as \emph{least squares regression} and in the next
+section we generalize to the Bayesian case. In a (multiple) linear regression
+model we have (noisy) measurements $\vec{y}\in \mathbb{R}^n$ (\emph{dependent
+    variable}), an $n\times p$ matrix $\vec{M}\in \mathbb{R}^{n\times p}$ of inputs
+(\emph{independent variables}, \emph{covariates}, or \emph{features}) and a
+vector of \emph{regression coefficients} or \emph{weights} $\vec{\theta}\in
+    \mathbb{R}^p$. This is a very general model because the matrix $\vec{M}$ can
+consist of basis functions such as polynomials, spherical harmonics, Fourier
+modes, wavelets etc. so although the model is linear in the parameters
+(regression coefficients), it is not generally linear in the raw input data. If
+we assume that the observations $\vec{y}$ have Gaussian noise we can write down
+the likelihood of this model as
+\begin{equation}
+    p(\vec{y}\lvert \vec{\theta})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta}, \vec{C})
+\end{equation}
+where $\vec{C}$ is the data covariance matrix and $\mathcal{N}$ is shorthand notation for
+the multivariate normal distribution. In astronomical application it is usually
+assumed to be diagonal and we might have some idea for the scale of the variance
+of the diagonal elements (the ``errorbars'').
+The negative log likelihood (NLL) is then equal (ignoring constants)  to
+\begin{equation}
+    \operatorname{NLL}(\vec{\theta})=\frac{1}{2}(\vec{M} \vec{\theta}-\vec{y})^{\top}\vec{C}^{-1}(\vec{M} \vec{\theta}-\vec{y})
+\end{equation}
+which is just the well known least squares objective or (half the) $\chi^2$ as it is known
+in astronomy.
+
+\subsubsection{Underdetermined case ($p < n$)}
+For now, let's assume that the inverse data covariance matrix $\vec{C}^{-1}$ is
+a diagonal matrix with the the same variance $\sigma^2$ for every data point.
+In that case the MLE solution for the weights $\vec{\theta}$ is the solution to
+the following (convex) optimization problem:
+\begin{equation}
+    \hat{\vec{\theta}}=
+    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2}
+    \label{eq:ols_objective}
+\end{equation}
+There is a closed form solution for Equation~\ref{eq:ols_objective} as long as the number
+of parameters and features $p$ is less than the number of data points $n$ and the solution
+is \citep{arXiv:2101.07256}
+\begin{equation}
+    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M})^{-1} \vec{M}^{\top} \vec{y}
+    \label{eq:ols_solution}
+\end{equation}
+under the assumption that $\vec{M}\vec{M}^\top$ is invertible which is usually the case.
+Equation~\ref{eq:ols_solution} is called the \emph{ordinary least squares} (OLS) solution.
+For a nontrivial data covariance matrix $\vec{C}$ Equation~\ref{eq:ols_solution} generalizes
+to
+\begin{equation}
+    \hat{\vec{\theta}}=(\vec{M}^{\top}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\top}\vec{C}^{-1} \vec{y}
+    \label{eq:wls_solution}
+\end{equation}
+which is the solution to the \emph{weighted least squares} (WLS) problem.
+It is common to modify the least squares objective functions with the addition of a
+\emph{regularization} term to prevent overfitting. For instance, we can modify the OLS
+objective (Equation~\ref{eq:ols_objective}) by adding an \emph{L2 regularization} term:
+\begin{equation}
+    \hat{\vec{\theta}}=
+    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2
+    \label{eq:ols_objective_l2}
+\end{equation}
+where $\lambda >0$ is a regularization parameter
+which penalizes large magnitudes of the coefficients $\vec{\theta}$. This is also a convex
+optimization problem and the solution is
+\begin{equation}
+    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\top} \vec{y}
+    \label{eq:ols_solution_l2}
+\end{equation}
+where $\vec{I}$ is the identity matrix. Regularized OLS regression with this particular
+regularization term is also known as \emph{ridge regression} in the statistics
+literature\footnote{One of the many things that make statistics confusing is that very
+    similar methods often have special names which obscures the big picture. The vast majority
+    of statistical methods can be seen as a linear model with the data represented in a particular
+    basis, and with a particular choice of the likelihood (noise model), priors (regularization)
+    and  a method for obtaining the solution (usually a convex optimization problem).}
+L2 regularization (as well as most other forms of regularization terms) can be seen
+as a particular choice of priors for the regression coefficients in the Bayesian paradigm.
+The above also naturally generalizes to the case of weighted least squares.
+
+\subsubsection{Overdetermined case ($p > n$)}
+Contrary to popular belief within the astronomy community it is possible and
+often desirable to have a model with more parameters than data points (see
+Chapter~\ref{ch:mapping_io} for a real world example). In the overparametrized
+case of linear regression $(p > n)$ there are multiple choices for
+$\vec{\theta}$ which perfectly fit the data. The solution in that case is
+defined to be a minimum-norm parameter vector $\vec{\theta}$ that interpolates
+the data ($\vec{y}=\vec{M} \vec{\theta}$), out of possibly many degenerate
+solutions. The OLS objective in Equation~\ref{eq:ols_objective} can be modified
+such that it is valid for both the $p<n$ and the $p>n$ case by making use of a
+limit in which an L2 regularization term tends to zero
+\citep{arXiv:2101.07256}:
+\begin{equation}
+    \hat{\vec{\theta}}=\lim _{\lambda \rightarrow 0^{+}}\left[
+    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2\right]\quad .
+\end{equation}
+In this case the OLS solution becomes
+\begin{equation}
+    \hat{\vec{\theta}}=\vec{M}^{\top} (\vec{M} \vec{M}^{\top})^{-1} \vec{y}
+    \label{eq:ols_solution_overdet}
+\end{equation}
+Equations~\ref{eq:ols_solution} and \ref{eq:ols_solution_overdet} can be unified into a
+single equation by making use of the \emph{Moore-Penrose pseudoinverse} $\vec{M}^\dagger$.
+The pseudoinverse is defined by taking the singular-value decomposition (SVD)
+of $\vec{M}$:
+\begin{equation}
+    \vec{M}^\dagger\equiv\vec{U} \vec{S}^\dagger \vec{V}^\dagger
+\end{equation}
+Then $\vec{M}^\dagger\equiv  \vec{V}^{\top} \vec{S}^{\dagger} \vec{U}^{\top}$.
+
+\subsection{Bayesian linear regression}
+In the Bayesian paradigm we can write down the posterior over the regression
+weights $\vec{\theta}$ assuming a Gaussian likelihood and a Gaussian prior over
+the weights. The posterior is then proportional to
+\begin{align}
+    p(\vec{\theta}|\vec{y}) & \propto p(\vec{y}\lvert \vec{\theta})\,p(\vec{\theta})                                                          \\
+                            & =\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta},\vec{C})\,\mathcal{N}(\vec{\theta}\lvert \vec{\mu},\vec{\Lambda})
+    \label{eq:bayes_lin_regression}
+\end{align}
+where $\vec{C}$ is the data covariance matrix as before, $\vec{\mu}$ is the mean
+of the Gaussian prior on coefficients and $\vec{\Lambda}$ is the the prior covariance matrix.
+The product of two multivariate Gaussian distributions is also a multivariate Gaussian distribution
+and the solution to Equation~\ref{eq:bayes_lin_regression} is \citep{arXiv:2005.14199}
+\begin{equation}
+    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{\theta}, \vec{C}) \,\mathcal{N}(\vec{\theta} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{\theta} \lvert\vec{a}, \vec{A}) \,\mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
+\end{equation}
+where
+\begin{align}
+     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\top} \vec{C}^{-1} \vec{M}                                                         \\
+     & \vec{a}         =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\top} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
+     & \mathbf{B}      =\mathbf{C}+\vec{M} \boldsymbol{\Lambda} \vec{M}^{\top}                                                        \\
+     & \boldsymbol{b}  =\vec{M} \boldsymbol{\mu}
+\end{align}
 

### 2022-06-21 src/ms.tex
Almost done with Chapter 2

@@
 % Define document class
-\documentclass[12pt]{report}
+\documentclass[12pt,dvipsnames]{report}
 \usepackage{aas_macros}
@@
 % Figures & tables 
-\usepackage{graphicx,xcolor}
+\usepackage{graphicx}
 \usepackage{caption}
@@
 %\usepackage[bbgreekl]{mathbbol}
+
+% Algorithms 
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 
+% Links 
+\hypersetup{
+    hidelinks,
+    colorlinks=true,
+    allcolors=NavyBlue,
+}
+
 % Settings
@@
 \renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}
-
-\hypersetup{
-    colorlinks=true,
-    allcolors=cyan,
-}
+\newcommand{\hquad}{~~}
 
@@
     \item I briefly discuss the issue of correlated noise in microlensing light curves
-          and propose ways to account for it using Gaussian Processes.
+          and propose ways to account for it using Gaussian processes.
 \end{itemize}
@@
 \begin{equation}
-    F(t)=F_S(t)+F_B(t)
-    \label{eq:flux_observed}
+    F(t)=F_S\,A(t)+F_B
+    \label{eq:flux_observed_micro}
 \end{equation}
@@
 as the new parameters.
-Equation~\ref{eq:flux_observed} then takes the form
+Equation~\ref{eq:flux_observed_micro} then takes the form
 \begin{equation}
@@
 \begin{equation}
-    I(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
+    I(x, y)=\tilde{\mathbf{y}}^{\intercal}(x, y) \mathbf{y}
     \label{eq:sh_expansion}
@@
                  Y_{2,-1} & Y_{2,0} & Y_{2,1} & Y_{2,2} & \cdots
-             \end{array}\right)^{\top} \quad,
+             \end{array}\right)^{\intercal} \quad,
 \end{align}
@@
         1 & x & z & y & x^{2} & x z & x y & y z & y^{2} & \cdots
-    \end{array}\right)^{\top}
+    \end{array}\right)^{\intercal}
 \end{equation}
@@
 \begin{align}
-    I(x, y) & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{p}                    \\
-            & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{y}\quad,
+    I(x, y) & =\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{p}                    \\
+            & =\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_1 \mathbf{y}\quad,
     \label{eq:intensity_poly_basis}
@@
         1 & 2 x & z & y & 3 x^{2} & -3 x z & 2 x y & 3 y z & y^{2} & \cdots
-    \end{array}\right)^{\top}
+    \end{array}\right)^{\intercal}
 \end{equation}
@@
 \begin{align}
-    I(x, y) & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{g}              \\
-            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A}_2 \mathbf{p} \\
-            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A} \mathbf{y}
+    I(x, y) & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{g}              \\
+            & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{A}_2 \mathbf{p} \\
+            & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{A} \mathbf{y}
 \end{align}
@@
 \begin{equation}
-    I(x, y)=\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{R}\mathbf{y}
+    I(x, y)=\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_1 \mathbf{R}\mathbf{y}
 \end{equation}
@@
     F & =\oiint I(x, y) \ud S                                                                 \\
-      & =\oiint \tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_{1} \mathbf{R} \mathbf{y} d S \\
-      & =\mathbf{r}^{\top} \mathbf{A}_{1} \mathbf{R} \mathbf{y}
+      & =\oiint \tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_{1} \mathbf{R} \mathbf{y} d S \\
+      & =\mathbf{r}^{\intercal} \mathbf{A}_{1} \mathbf{R} \mathbf{y}
 \end{align}
@@
     F & =\oiint I(x, y) \mathrm{d} S                              \\
-      & =\mathbf{s}^\top  \mathbf{A} \mathbf{R} \mathbf{y} .
+      & =\mathbf{s}^\intercal  \mathbf{A} \mathbf{R} \mathbf{y} .
 \end{align}
@@
 \begin{equation}
-    \mathbf{s}^\top \equiv \oiint \tilde{\mathbf{g}}^{\top}(x, y) \mathrm{d} S
+    \mathbf{s}^\intercal \equiv \oiint \tilde{\mathbf{g}}^{\intercal}(x, y) \mathrm{d} S
 \end{equation}
@@
 along the boundary of the visible portion of the occulted disc.
-The $n$-th element of the solution vector $\mathbf{s}^\top$ is
+The $n$-th element of the solution vector $\mathbf{s}^\intercal$ is
 \begin{equation}
@@
 \begin{equation}
-    A(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
+    A(x, y)=\tilde{\mathbf{y}}^{\intercal}(x, y) \mathbf{y}
 \end{equation}
@@
 and integrate them over the visible portion of the body's disc to obtain
-the reflected light solution vector $\mathbb{s}^\top$.
+the reflected light solution vector $\mathbb{s}^\intercal$.
 Evaluating these integrals is unfortunately intractable because of the piecewise
@@
 \begin{equation}
-    f=\mathbb{s}^\top\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right) \mathbf{A}_{\mathbf{2}} \mathbf{I}\left(b, \theta^{\prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime}\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+    f=\mathbb{s}^\intercal\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right) \mathbf{A}_{\mathbf{2}} \mathbf{I}\left(b, \theta^{\prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime}\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
 \end{equation}
@@
 \begin{equation}
-    f_{0}=\mathbb{r}^{\top}(b) \mathbf{I}\left(b, \theta^{\prime \prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime \prime}\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+    f_{0}=\mathbb{r}^{\intercal}(b) \mathbf{I}\left(b, \theta^{\prime \prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime \prime}\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
     \label{eq:starry_ref_phase}
@@
 The solutions to integrals
-$\mathbb{r}^{\top}(b)$ and $\mathbb{s}^{\top}\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right)$
+$\mathbb{r}^{\intercal}(b)$ and $\mathbb{s}^{\intercal}\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right)$
 are anything but trivial, they are derived in the appendices of
@@
     f = \mathbf{X}\mathbf{y}
+    \label{eq:starry_linear_model}
 \end{equation}
@@
 
-\section{Statistical inference -- theory}
-\label{sec:inference_theory}
+\section{Statistical inference -- introduction}
+\label{sec:inference_intro}
 Having covered the basic physics behind gravitational lensing and occultation mapping I
@@
     x_2)\,p(x_2)\textrm{d}x_2
+    \label{eq:marginalization}
 \end{equation}
@@
 \begin{align}
-    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\top}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                                \\
+    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\intercal}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                           \\
                                 & =\left(\begin{array}{cccc}
@@
 \end{equation}
-where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\top}}$ is the Jacobian of $\vec{g}$
+where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\intercal}}$ is the Jacobian of $\vec{g}$
 and $\operatorname{det}|\mathbf{J}(\vec{y})|$ is the determinant of the Jacobian.
@@
 \end{equation}
-is often called the \emph{fully marginalized likelihood} because
+is often called the \textsl{fully marginalized likelihood} because
 it is a likelihood marginalized over all of the parameters of the model.
@@
 %% mention the likelihood principle
-\section{Statistical inference}
-\label{sec:inference_practice}
-In the previous section I briefly summarized some key aspects of probability theory
-and Bayesian statistics but I haven't discussed how to apply these ideas in practice. That
-is the subject of statistical inference. I start by introducing some very fundamental
-concepts such as the \emph{curse of dimensionality} and the \emph{no free lunch theorems}.
-I then cover linear models which are the foundation of all statistics and also relevant
-for the occultation mapping work since the problem is linear.
-Next, I discuss various computational inference methods such
-as optimization methods, Markov chain Monte Carlo and Variational Inference.
-Finally, I discuss model comparison and evaluation.
-
-\subsection{The curse of dimensionality and the no-free-lunch theorem}
-The \emph{curse of dimensionality} refers to a phenomenon where the
-\emph{volume} of \emph{possible} explanations for the data grows exponentially
-with increasing dimension of the problem. Consider the following
-example\footnote{ Taken from the excellent paper by \citet{arXiv:2104.00008}.
-}. There are images consisting of $n$ pixels where each of the pixels is either
-a black (0) or white (1). The number of possible images is $2^n$ which grows
-exponentially. If each image out of the $2^n$ possible images is described by
-one of two labels $A$, or $B$ then the total number of possible ways of
-labeling the set of images is $2^{2^n}$ which is a stupidly large number. If
-$n=9$, the number of ways to classify the set of all 9-bit black and white
-images is greater than the number of atoms in the observable universe. If the
-labels don't correlate with the properties of the images then one strategy for
-solving this problem would be the lookup table approach -- memorize each image
-with its label. The famous \emph{no-free-lunch theorem} (NFLT) proved by the
-mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
-impossible to construct an algorithm which does better than the lookup table
-approach.
-
-Of course the assumption that the labels aren't correlated to features in the
-images is silly for any meaningful problem. For instance, all images of cats
-have certain things in common as do images of galaxies. The no-free-lunch
-theorem doesn't say that it is impossible to construct an algorithm to classify
-images of cats, it only says that there is no best algorithm \emph{if we
-    average over all possible inputs to the problem}. To construct an algorithm
-which works in practice we need to impose structure on the problem, what
-machine learning researchers call \emph{inductive bias}. A set of assumptions
-which work in one domain may not work well in another domain and no single
-model will work well in all domains.
 
@@
 \begin{equation}
-    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\log p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \log |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\top} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
+    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\ln p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \ln |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\intercal} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
 \end{equation}
@@
 \begin{equation}
-    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\top}
+    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\intercal}
 \end{equation}
@@
 \subsection{Least squares linear regression}
+\label{ssec:least_squares}
 In this section we cover the classical non Bayesian formulation of linear
@@
     variable}), an $n\times p$ matrix $\vec{M}\in \mathbb{R}^{n\times p}$ of inputs
-(\emph{independent variables}, \emph{covariates}, or \emph{features}) and a
-vector of \emph{regression coefficients} or \emph{weights} $\vec{\theta}\in
-    \mathbb{R}^p$. This is a very general model because the matrix $\vec{M}$ can
-consist of basis functions such as polynomials, spherical harmonics, Fourier
-modes, wavelets etc. so although the model is linear in the parameters
-(regression coefficients), it is not generally linear in the raw input data. If
-we assume that the observations $\vec{y}$ have Gaussian noise we can write down
-the likelihood of this model as
-\begin{equation}
-    p(\vec{y}\lvert \vec{\theta})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta}, \vec{C})
+(\emph{design matrix} consisting of \emph{independent variables},
+\emph{covariates}, or \emph{features}) and a vector of \emph{regression
+    coefficients} or \emph{weights} $\vec{w}\in \mathbb{R}^p$. This is a very
+general model because the matrix $\vec{M}$ can consist of basis functions such
+as polynomials, spherical harmonics, Fourier modes, wavelets etc. so although
+the model is linear in the parameters (regression coefficients), it is not
+generally linear in the raw input data. If we assume that the observations
+$\vec{y}$ have Gaussian noise we can write down the likelihood of this model as
+\begin{equation}
+    p(\vec{y}\lvert \vec{w})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{w}, \vec{C})
 \end{equation}
@@
 \begin{equation}
-    \operatorname{NLL}(\vec{\theta})=\frac{1}{2}(\vec{M} \vec{\theta}-\vec{y})^{\top}\vec{C}^{-1}(\vec{M} \vec{\theta}-\vec{y})
+    \operatorname{NLL}(\vec{w})=\frac{1}{2}(\vec{M} \vec{w}-\vec{y})^{\intercal}\vec{C}^{-1}(\vec{M} \vec{w}-\vec{y})
 \end{equation}
@@
 a diagonal matrix with the the same variance $\sigma^2$ for every data point.
-In that case the MLE solution for the weights $\vec{\theta}$ is the solution to
-the following (convex) optimization problem:
-\begin{equation}
-    \hat{\vec{\theta}}=
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2}
+In that case the MLE solution for the weights $\vec{w}$ is the solution to the
+following (convex) optimization problem:
+\begin{equation}
+    \hat{\vec{w}}=
+    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2}
     \label{eq:ols_objective}
@@
 \begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M})^{-1} \vec{M}^{\top} \vec{y}
+    \hat{\vec{w}}=(\vec{M}^{\intercal} \vec{M})^{-1} \vec{M}^{\intercal} \vec{y}
     \label{eq:ols_solution}
 \end{equation}
-under the assumption that $\vec{M}\vec{M}^\top$ is invertible which is usually the case.
+under the assumption that $\vec{M}\vec{M}^\intercal$ is invertible which is usually the case.
 Equation~\ref{eq:ols_solution} is called the \emph{ordinary least squares} (OLS) solution.
@@
 \begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\top}\vec{C}^{-1} \vec{y}
+    \hat{\vec{w}}=(\vec{M}^{\intercal}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\intercal}\vec{C}^{-1} \vec{y}
     \label{eq:wls_solution}
@@
 \begin{equation}
-    \hat{\vec{\theta}}=
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2
+    \hat{\vec{w}}=
+    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2} + \lambda \left|\vec{w}\right|^2
     \label{eq:ols_objective_l2}
@@
 where $\lambda >0$ is a regularization parameter
-which penalizes large magnitudes of the coefficients $\vec{\theta}$. This is also a convex
+which penalizes large magnitudes of the coefficients $\vec{w}$. This is also a convex
 optimization problem and the solution is
 \begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\top} \vec{y}
+    \hat{\vec{w}}=(\vec{M}^{\intercal} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\intercal} \vec{y}
     \label{eq:ols_solution_l2}
@@
 literature\footnote{One of the many things that make statistics confusing is that very
-    similar methods often have special names which obscures the big picture. The vast majority
+    similar methods often have special names wich obscures the big picture. The vast majority
     of statistical methods can be seen as a linear model with the data represented in a particular
@@
 Chapter~\ref{ch:mapping_io} for a real world example). In the overparametrized
-case of linear regression $(p > n)$ there are multiple choices for
-$\vec{\theta}$ which perfectly fit the data. The solution in that case is
-defined to be a minimum-norm parameter vector $\vec{\theta}$ that interpolates
-the data ($\vec{y}=\vec{M} \vec{\theta}$), out of possibly many degenerate
-solutions. The OLS objective in Equation~\ref{eq:ols_objective} can be modified
-such that it is valid for both the $p<n$ and the $p>n$ case by making use of a
-limit in which an L2 regularization term tends to zero
-\citep{arXiv:2101.07256}:
-\begin{equation}
-    \hat{\vec{\theta}}=\lim _{\lambda \rightarrow 0^{+}}\left[
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2\right]\quad .
+case of linear regression $(p > n)$ there are multiple choices for $\vec{w}$
+which perfectly fit the data. The solution in that case is defined to be a
+minimum-norm parameter vector $\vec{w}$ that interpolates the data
+($\vec{y}=\vec{M} \vec{w}$), out of possibly many degenerate solutions. The OLS
+objective in Equation~\ref{eq:ols_objective} can be modified such that it is
+valid for both the $p<n$ and the $p>n$ case by making use of a limit in which
+an L2 regularization term tends to zero \citep{arXiv:2101.07256}:
+\begin{equation}
+    \hat{\vec{w}}=\lim _{\lambda \rightarrow 0^{+}}\left[
+    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2} + \lambda \left|\vec{w}\right|^2\right]\quad .
 \end{equation}
@@
 \begin{equation}
-    \hat{\vec{\theta}}=\vec{M}^{\top} (\vec{M} \vec{M}^{\top})^{-1} \vec{y}
+    \hat{\vec{w}}=\vec{M}^{\intercal} (\vec{M} \vec{M}^{\intercal})^{-1} \vec{y}
     \label{eq:ols_solution_overdet}
@@
 \begin{equation}
-    \vec{M}^\dagger\equiv\vec{U} \vec{S}^\dagger \vec{V}^\dagger
-\end{equation}
-Then $\vec{M}^\dagger\equiv  \vec{V}^{\top} \vec{S}^{\dagger} \vec{U}^{\top}$.
+    \vec{M}\equiv\vec{U} \vec{S} \vec{V}
+\end{equation}
+Then $\vec{M}^\dagger =  \vec{V}^{\intercal} \vec{S}^{\dagger} \vec{U}^{\intercal}$.
 
@@
 In the Bayesian paradigm we can write down the posterior over the regression
-weights $\vec{\theta}$ assuming a Gaussian likelihood and a Gaussian prior over
-the weights. The posterior is then proportional to
+weights $\vec{w}$ assuming a Gaussian likelihood and a Gaussian prior over the
+weights. The posterior is then
 \begin{align}
-    p(\vec{\theta}|\vec{y}) & \propto p(\vec{y}\lvert \vec{\theta})\,p(\vec{\theta})                                                          \\
-                            & =\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta},\vec{C})\,\mathcal{N}(\vec{\theta}\lvert \vec{\mu},\vec{\Lambda})
+    p(\vec{w}|\vec{y}) & = \frac{p(\vec{y}\lvert \vec{w})\,p(\vec{w})}{\int p(\vec{y}\lvert \vec{w})p(\vec{w}) \ud\vec{w}}                                                                                                                                                 \\
+                       & = \frac{    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{w}, \vec{C}) \,\mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda})}{\int \mathcal{N}(\vec{y} \lvert\vec{M} \vec{w}, \vec{C}) \,\mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda}) \ud\vec{w}}
     \label{eq:bayes_lin_regression}
@@
 of the Gaussian prior on coefficients and $\vec{\Lambda}$ is the the prior covariance matrix.
-The product of two multivariate Gaussian distributions is also a multivariate Gaussian distribution
-and the solution to Equation~\ref{eq:bayes_lin_regression} is \citep{arXiv:2005.14199}
-\begin{equation}
-    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{\theta}, \vec{C}) \,\mathcal{N}(\vec{\theta} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{\theta} \lvert\vec{a}, \vec{A}) \,\mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
+Using the fact that the product of two multivariate Gaussian distributions is also a
+multivariate Gaussian distribution, one can show that
+the solution to Equation~\ref{eq:bayes_lin_regression} is
+\citep[see for example][]{arXiv:2005.14199}
+\begin{equation}
+    p(\vec{w}|\vec{y}) =\mathcal{N}(\vec{w} \lvert\vec{a}, \vec{A})
+    \label{eq:linear_regression_posterior}
 \end{equation}
@@
 \begin{align}
-     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\top} \vec{C}^{-1} \vec{M}                                                         \\
-     & \vec{a}         =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\top} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
-     & \mathbf{B}      =\mathbf{C}+\vec{M} \boldsymbol{\Lambda} \vec{M}^{\top}                                                        \\
-     & \boldsymbol{b}  =\vec{M} \boldsymbol{\mu}
+     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\intercal} \vec{C}^{-1} \vec{M}                                                     \\
+     & \vec{a}     =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\intercal} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
 \end{align}
-
-%\subsection{Machine learning and the difference between inference and prediction}
+Since all \textsf{starry} models for occultation light curves are linear
+(Equation~\ref{eq:starry_linear_model}) we can use Equation~\ref{eq:linear_regression_posterior}
+to directly obtain the posterior over the map coefficients $\mathbf{y}$ but the cost
+of doing so is imposing a Gaussian prior on the map coefficients.
+
+\subsection{Marginalizing a likelihood over linear parameters}
+A model structure which is quite common in astronomy and physics is a model
+which depends on some linear, and some nonlinear parameters. This is the
+structure of the equation for the observed flux in microlensing
+(Equations~\ref{eq:flux_observed_micro} or \ref{eq:flux_dominik}) where the
+flux parameters are linear and the magnification parameters are nonlinear. The
+likelihoods for such models can be written as
+\begin{equation}
+    p(\vec{y}\lvert \vec{w}, \vec{\theta})
+\end{equation}
+where $\vec{w}$ are the linear parameters and $\vec{\theta}$ are the nonlinear parameters.
+In some circumstances, it is possible to analytically marginalize over the linear
+parameters $\vec{w}$. Doing so is often advantageous because it reduces the size of the
+parameter space  and the computational cost of doing inference.
+The marginalized likelihood is given by (Equation~\ref{eq:marginalization})
+\begin{equation}
+    p(\vec{y}\lvert \vec{\theta})=\int p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta}) \ud\vec{w}
+\end{equation}
+Marginalization requires the introduction of a prior  $p(\vec{w}\lvert \vec{\theta})$
+which can potentially also depend on the nonlinear parameters $\vec{\theta}$.
+The alternative to doing this integral is re-factorizing the expression in the integrand
+using Bayes' theorem:
+\begin{equation}
+    p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta})  = p(\vec{w}\lvert \vec{y},\vec{\theta})\,p(\vec{y}\lvert \vec{\theta})
+\end{equation}
+which yields the marginalized likelihood without us having to do the integral:
+\begin{equation}
+    p(\vec{y}\lvert \vec{\theta}) = \frac{    p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta})}{p(\vec{w}\lvert \vec{y},\vec{\theta})}
+    \label{eq:likelihood_factorization}
+\end{equation}
+If both the prior and the likelihood are multivariate Gaussians with the following form:
+\begin{align}
+    p(\vec{y}\lvert \vec{w}, \vec{\theta}) & = \mathcal{N}(\vec{y}\lvert \vec{M}\vec{w},\vec{C})  \\
+    p(\vec{w}\lvert \vec{\theta})          & = \mathcal{N}(\vec{w}\lvert \vec{\mu},\vec{\Lambda})
+\end{align}
+then Equation~\ref{eq:likelihood_factorization} becomes
+\citep[for the proof see][]{arXiv:2005.14199}
+\begin{equation}
+    \mathcal{N}(\vec{y} \lvert \vec{M} \vec{w}, \vec{C})\, \mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{w} \lvert\vec{a}, \vec{A})\, \mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
+\end{equation}
+where
+\begin{align}
+     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\intercal} \vec{C}^{-1} \vec{M}                                                     \\
+     & \vec{a}     =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\intercal} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
+     & \mathbf{B}=\mathbf{C}+\vec{M} \vec{\Lambda} \vec{M}^{\intercal}                                                                 \\
+     & \vec{b}=\vec{M} \vec{\mu}
+\end{align}
+Thus the marginalized likelihood is
+\begin{equation}
+    p(\vec{y}\lvert \vec{\theta}) = \mathcal{N}(\vec{y} \lvert \vec{b}, \vec{B})
+    \label{eq:likelihood_lin_params_marginalized}
+\end{equation}
+
+\subsubsection{Application to microlensing}
+Let's apply this trick to the microlensing models
+(Equation~\ref{eq:flux_dominik}). In the general case, the dataset consists of
+$J$ flux vectors $\vec{f}_j$, each of lengths $N_j$ where $j$ indexes different
+spectral bands, often from different observatories as well.
+Equation~\ref{eq:flux_dominik} can then be written in vector form as
+\begin{equation}
+    \vec{f}_j=\vec{M}_j\vec{w}_j
+    \label{eq:microlensing_flux_general}
+\end{equation}
+where the matrix $\vec{M}$ is defined as
+\begin{equation}
+    \vec{M}_j\equiv \begin{pmatrix}
+        \tilde{A}(t_1;\vec{\theta})     & 1      \\
+        \tilde{A}(t_2;\vec{\theta})     & 1      \\
+        \vdots                          & \vdots \\
+        \tilde{A}(t_{N_j};\vec{\theta}) & 1
+    \end{pmatrix}
+\end{equation}
+and $t=t_1,\ldots, t_{N_j}$ are the times of observations in band $j$,
+$\tilde{A}(t_i) \equiv (A(t_i)-1) /(A\left(t_{0}\right)-1)$, $\vec{\theta}$ are the
+nonlinear parameters, and the weights vector $\vec{w}_j$ is defined as
+\begin{equation}
+    \vec{w}_j\equiv    \left( \Delta F_j\; F_{\mathrm{base},j}\right)^\intercal
+\end{equation}
+Equation~\ref{eq:microlensing_flux_general} is general enough to encompass
+either single or multiple lens microlensing models. Assuming independence of
+noise properties between observations in different spectral bands\footnote{This
+    is certainly true if we have observations from different telescopes. It is not
+    strictly true if the observations are from the same telescope in different
+    spectral bands but it is often a good enough assumption.} the likelihood for
+the complete dataset
+$\mathcal{D}=\left\{f_{j}\right\}_{j=1}^{J}$ factorizes as
+\begin{equation}
+    p(\mathcal{D}\lvert \vec{w}, \vec{\theta})=\prod_{j=1}^J p(\vec{f}_j\lvert \vec{w}_j, \vec{\theta})
+\end{equation}
+Using Equation~\ref{eq:likelihood_lin_params_marginalized} we can then write down the
+negative log-likelihood (ignoring constants) for the $j$-th band as
+\begin{align}
+    \operatorname{NLL}(\vec{\theta}) & =\frac{1}{2}\left( \vec{f}_j-\vec{M}_j\vec{\mu}\right)^\intercal\left(\vec{C}_j + \vec{M}_j\vec{\Lambda}\vec{M}_j^\intercal\right)^{-1}
+    \left( \vec{f}_j-\vec{M}_j\vec{\mu}\right)                                                                                                                                 \\
+                                     & + \frac{1}{2}\ln\left| \vec{C}_j + \vec{M}_j\vec{\Lambda}\vec{M}_j^\intercal\right|
+    \label{eq:ll_marginalized_microlensing}
+\end{align}
+where we have assumed a common mean $\vec{\mu}$ and covariance $\vec{\Lambda}$ for the
+prior for the linear flux parameters.
+To avoid computing an inverse and a determinant of $J$ matrices with shape $N_j\times N_j$,
+we can simplify the terms in Equation~\ref{eq:ll_marginalized_microlensing} using the
+matrix inversion lemma \citep[see for example Appendix A3 in][]{rasmussen2006}
+\begin{align}
+     & \left(\mathbf{C}_j+\vec{M}_j \boldsymbol{\Lambda} \vec{M}_j^{\intercal}\right)^{-1}=\mathbf{C}_j^{-1}-\mathbf{C}_j^{-1} \vec{M}_j\left(\boldsymbol{\Lambda}^{-1}+\vec{M}_j^{\intercal} \mathbf{C}^{-1} \vec{M_j}\right)^{-1} \vec{M}_j^{\intercal} \mathbf{C}_j^{-1} \\
+     & \ln\left|\mathbf{C}_j+\vec{M}_j \vec{\Lambda} \vec{M}_j^{\intercal}\right|=\ln|\mathbf{C}_j| +\ln|\vec{\Lambda}| + \ln\left|\boldsymbol{\Lambda}^{-1}+\vec{M}_j^{\intercal} \mathbf{C}_j^{-1} \vec{M}_j\right|
+\end{align}
+In most cases we can assume that $\vec{\mu}=\vec{0}$, $\vec{\Lambda}$ is diagonal, and
+in absence of correlated noise all matrices $\vec{C}_j$ are also diagonal so the expression
+in Equation~\ref{eq:ll_marginalized_microlensing} isn't usually expensive to compute.
+
+Although Equation~\ref{eq:ll_marginalized_microlensing} is the formally correct
+way to marginalize the linear parameters of a general microlensing model, it is
+not used in the microlensing literature. What most people do instead is using
+MLE to maximize the joint likelihood $p(\vec{f}\lvert\vec{w},\vec{\theta})$
+with respect to the linear parameters by solving a least squares problem at
+every MCMC step in the nonlinear parameters (we'll cover MCMC in the next
+section). This procedure is approximately equivalent to using the marginalized
+likelihood from Equation~\ref{eq:ll_marginalized_microlensing} with two
+important caveats. The first it that one shouldn't optimize for the maximum
+likelihood flux parameters conditional on nonlinear parameters, but rather the
+MAP values of those parameters under Gaussian priors (although, the difference
+between the MLE and MAP estimates is negligible under broad priors). The second
+caveat is that the optimization can be cast as a linear least squares problem
+under the assumption that the covariance matrix $\vec{C}$ for the data is
+diagonal. This is a fairly restrictive assumptions because it implies that
+there is no correlated noise in the light curve which isn't true for most real
+world microlensing datasets.
+
+Although we have marginalized the likelihood over the linear parameters which
+substantially reduces the complexity of the inference problem we can still
+compute the posterior distribution for those parameters \emph{conditional on
+    the value of the nonlinear parameters}:
+\begin{equation}
+    p(\vec{w}\lvert \vec{y}, \vec{\theta})=\mathcal{N}(\vec{w}\lvert \vec{y}, \vec{\theta})
+\end{equation}
+which we can use to obtain the full posterior over the linear parameters if we happen
+to have samples from the posterior over the nonlinear parameters $p(\vec{\theta}|\vec{y})$
+(obtained using MCMC for example).
+
+\subsection{Gaussian processes}
+% https://peterroelants.github.io/posts/gaussian-process-tutorial/
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/gp_samples.pdf}
+        \caption{A few samples from a Gaussian process prior with a squared exponential
+            kernel function for two different values of the length scale parameter $l$.}
+        \label{fig:gp_samples}
+    \end{centering}
+\end{figure}
+A final topic that we need to cover in this section is that of \textsl{Gaussian
+    Processes}. In Section~\ref{ssec:least_squares} we have discussed linear
+regression with $p$ components (basis functions) and we have mentioned that one
+fit many more components than there are data points and that these models can
+be quite useful. A Gaussian process (GP) is what happens when let the number of
+components $p$ go to infinity\footnote{See \citet{arXiv:2101.07256} for a
+    demonstration of this point.} which is why they belong to a class of methods
+called \textsl{nonparametric regression} methods, although a better name would
+perhaps be infinite-parameter regression methods. This is not the only
+formulation of GPs. GPs can also be seen as \emph{prior distributions over the
+    space of functions} such that each sample from that GP prior is a function.
+More formally, \emph{a Gaussian process is a collection of random variables,
+    any finite number of which have a joint multivariate Gaussian
+    distribution} \citep{rasmussen2006}.
+
+GPs are completely specified by a \textsl{mean function} $\mu (x)$ and a
+positive-definite \textsl{covariance function} $k(x,x^\prime)$ which are
+functions of the input locations $x\in \mathbb{R}^D$ (for example, a time
+coordinate):
+\begin{align}
+    \mu (x)        & =\mathbb{E}[f(x)]                                                    \\
+    k(x, x^\prime) & =\mathbb{E}\left[(f(x) -\mu (x))(f(x^\prime) -\mu (x^\prime))\right]
+\end{align}
+where $x$ and $x^\prime$ are any two different points in the input space.
+We write
+\begin{equation}
+    f(x) \sim \mathcal{G} \mathcal{P}\left(\mu(x), k\left(x, x^{\prime}\right)\right)
+\end{equation}
+Usually the mean function $\mu(x)$  is taken to be zero for simplicity and all the
+magic happens in the covariance function $k(x,x^\prime)$ which describes
+\emph{how  any two points in the input space related to each other}.
+
+One very popular choice for the kernel function is the \textsl{squared
+    exponential} kernel:
+\begin{equation}
+    k(x, x^\prime) = \exp\left(-\frac{1}{2}\frac{\left|x-x^\prime\right|^2}{l^2}\right)\hquad .
+    \label{eq:sqexp}
+\end{equation}
+The above choice of a covariance function implies that the covariance of the outputs
+$f(x)$ evaluated at any two points  will decay exponentially with the square of
+the distance between the two points.
+In other words, function values evaluated at points close to each tend to go hand in hand
+while function values evaluated at points far away from each other are practically
+independent.
+The parameters $l$ sets the length scale which determines how quickly the covariance decays
+to zero with increasing distance between the two points.
+Figure~\ref{fig:gp_samples} shows samples from a GP prior with the squared exponential
+kernel function in the one-dimensional case for two choices of the length sale parameter
+$l$. One special property of this kernel is that it is \textsl{stationary}, meaning
+that the mean and the variance of the samples from the GP prior don't depend on the
+absolute value of the input coordinate $x$, they are the same everywhere because
+the covariance only depends on the difference between any two points $x$ and
+$x^\prime$\footnote{Stationarity is also a general property of time series.
+    A stationary  time series is one whose properties (mean, variance) do not depend
+    on the time at which the series is observed. A GP with a stationary kernel is a very flexible
+    model for stationary time series.}. Not all kernels are stationary. An example of
+a non-stationary kernel would be a kernel similar to the squared exponential but with
+a variance growing linearly with time.
+Kernels can be combined in any number of ways through for example addition and
+multiplication and the result is also a valid kernel.
+\href{https://www.cs.toronto.edu/~duvenaud/cookbook/}{This} pages provides a
+nice visualization of different kernels and their combinations.
+
+The key property of a GP is that for any finite subset
+$\vec{X}=\left\{\mathbf{x}_{1} \ldots \mathbf{x}_{n}\right\}$ of the domain of
+$x$ (that is, any marginal distribution of $x$) is a multivariate Gaussian:
+\begin{equation}
+    f(\vec{x})\sim \mathcal{N}\,(\vec{\mu}(\vec{X}), \vec K(\vec{X}, \vec{X}))
+\end{equation}
+where $\vec K$ is a covariance matrix of shape $n\times n$ which is constructed by
+evaluating  a kernel such as the one defined in Equation~\ref{eq:sqexp} elementwise.
+This is precisely how we plotted
+the functions in Figure~\ref{fig:gp_samples}, we simply chose a dense number of linearly
+spaced grid points, computed the covariance matrix $K$ and then sampled the multivariate
+Gaussian distribution.
+\subsubsection{Predictions from posterior distribution}
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/gp_samples_conditioned.pdf}
+        \caption{Samples from a Gaussian process with squared exponential kernel $l=1$
+            conditioned on observations simulated from a simple sinusoid.
+            The left panel shows the case when we assume that
+            the observations are noiseless in which case the GP perfectly interpolates the data.
+            The right panel shows the case when we assume that the observations are generated
+            from the GP with the addition of some white Gaussian noise with variance $0.1^2$.}
+        \label{fig:gp_samples_conditioned}
+    \end{centering}
+\end{figure}
+We can go a step beyond drawing samples from the GP prior and condition the GP
+on a set of observations in order fit the GP to data. Let $\vec{f}(\vec{X})$ be
+the vector of observed data points evaluated at $\vec{X}$ with length $n_1$. If
+we are interested in predicting the function values $\vec{f}^*(\vec{X}^*)$ at a
+new vector of data points $\vec{X}^*$ with length $n_2$ (so we could, for
+example, make a plot), then we need to compute the posterior distribution
+$p(\vec{f}^*\lvert \vec{f}, \vec{X},\vec{X}^*)$. In order to compute the
+conditional distribution, we start with the join distribution over $\vec{f}$
+and $\vec{f}^*$. Since $\vec{f}$ and $\vec{f}^*$ both come from the same
+multivariate Gaussian distribution, the joint distribution $p(\vec{f},
+    \vec{f}^*)$ is also a multivariate Gaussian:
+\begin{equation}
+    \begin{pmatrix}
+        \vec{f} \\
+        \vec{f}^*
+    \end{pmatrix} \sim \mathcal{N}\left[\begin{pmatrix}
+            \vec{\mu} \\
+            \vec{\mu}^*
+        \end{pmatrix},\begin{pmatrix}
+            \vec K(\vec{X},\vec{X})    & \vec K(\vec{X},\vec{X}^*)   \\
+            \vec K(\vec{X}^*, \vec{X}) & \vec K(\vec{X}^*,\vec{X}^*)
+        \end{pmatrix}\right]
+\end{equation}
+From this joint distribution, we can obtain the conditional  distribution
+\citep[see appendix A2 in ][]{rasmussen2006} which is also a multivariate Gaussian\footnote{
+    This is one of many special properties of Gaussians.
+}:
+\begin{align}
+    p(\mathbf{f}^{*} \lvert \vec X^{*}, \vec X, \mathbf{f} )\sim \mathcal{N}( & \vec K\left(\vec X^*, \vec X\right) \vec K(\vec X, \vec X)^{-1} \mathbf{f}                                                          \\
+                                                                              & \vec K\left(\vec X^*, \vec X^*\right)  -\vec K\left(\vec X^*, \vec X \right) \vec K(\vec X, \vec X)^{-1} \vec K (\vec X, \vec X^*))
+    \label{eq:gp_posterior}
+\end{align}
+
+Samples from this posterior GP distribution given by
+Equation~\ref{eq:gp_posterior} with a squared exponential kernel conditioned on
+a set of simulated observations $\vec{f}$ generated from a simple sinusoid are
+shown in the left panel of Figure~\ref{fig:gp_samples_conditioned}. Each draw
+from the distribution perfectly interpolates all the data points. The variance
+is naturally larger where the data is sparse and the variance of the
+predictions grows rapidly in the extrapolative regime outside of the range of
+the data. The squared exponential kernel is not a great choice here because the
+covariance rapidly decays to zero with the distance from the first and last
+data point so the predictions will look flat futher away from the range of the
+data. A much better choice would be a kernel with a periodic covariance
+structure which would not have this problem.
+
+The predictions shown in the left panel assume that the observations are
+noisless which is why each sample perfectly interpolates the data. It is more
+common to assume that there is an additional white Gaussian noise component in
+the observed data. In that case we add a diagonal matrix (because white noise
+is independent Gaussian noise) to the covariance matrix
+$\vec{K}(\vec{X},\vec{X})$ and we have
+\begin{equation}
+    \vec{K}(\vec{X},\vec{X}) + \sigma^2\vec{I}
+\end{equation}
+where $\vec{I}$ is the identity matrix. Predictions from a GP with this additional
+noise are shown in the right panel of Figure~\ref{fig:gp_samples_conditioned}.
+
+\subsubsection{Fitting for the parameters of a GP kernel}
+We shown how to sample a GP prior and how to compute predictions from a GP
+conditioned on a set of observations but we haven't discssed the parameters
+which define the kernel function, such as the length scale $l$ in the squared
+exponential kernel. These parameters are often called the
+\textsl{hyperparameters} of the kernel. Given a set of observations $\vec{f}$
+evaluated at $\vec{X}$ with length $n$, the likelihood of a GP model for the
+kernel hyperparameters $\vec{\theta}$ is
+\begin{equation}
+    p(\vec{f}\lvert \vec{\theta})\sim\,
+    \mathcal{N}(\vec{f}\lvert \vec{\mu}(\vec{\theta}),
+    \vec{K}(\vec X, \vec X; \vec\theta))
+    \label{eq:gp_marginal_likelihood}
+\end{equation}
+where $\vec{\theta}$ is a set of parameters which define the kernel function. We
+can treat these kernel parameters as we would any other parameters, for example we
+can optimize the likelihood in Equation~\ref{eq:gp_marginal_likelihood} and find the
+MLE parameters $\hat{\vec{\theta}}_\mathrm{MLE}$. The likelihood is not linear in these
+hyperaparamters and the the optimization problem is generally not convex.
+
+The major challenge with GPs and the reason they still aren't as popular as for
+instance linear regression is that sampling the GP prior or computing the
+marginal likelihood (Equation~\ref{gp:marginal_likelihood}) requires an
+inversion of a dense $n\times n$ covariance matrix, an operation which scales
+as $\mathcal{O}(n^3)$. This operation becomes prohibitively expensive for
+datasets with a few thousand data points. Thankfully, there has been a lot of
+work on faster algorithms for inverting GP covariance matrices. Most of these
+algorithms rely on the assumption of some special structure for the covariance
+matrix which can be exploited to speed up the linear algebra operations. The
+most notable example is the \textsf{celerite} algorithm
+\citep{2017AJ....154..220F} which enables the use of 1D GPs with linear
+$\mathcal{O}(n)$ scaling as long as we restrict ourselved to kernels which are
+sums of complex exponential functions. These kernels are sufficiently flexible
+for modeling physical time-series and the sums of complex exponentials also
+have a physical interpretation as mixtures of stochastically drive, dampled
+simple harmonic oscillators. Thanks to \textsf{celerite} it is possible to fit
+GPs to astronomical light curves with thousands of datapoints.
+
+\subsubsection{Use of GPs in astronomy}
+The use of Gaussian process models within astronomy has skyrocketed in recent
+years. GPs within astronomy are by far most commonly used to model correlated
+noise in astronomical light curves. Practically, this is a small step from just
+assuming independent Gaussian noise. All we have to do is to replace a
+likelihood with a diagonal covariance matrix whose entries are given by the
+variance estimates for individual data points (``errorbars'') with a dense
+covariance matrix whose generated using some kernel function (while still
+including the independent noise term). The mean of the multivariate Gaussian
+likelihood is usually an astrophysical model such as a transit, a microlensing
+flux prediction or a secondary eclipse prediction. In some cases, the mean is
+set to zero and the hyperparameters of the kernel have physical meaning. For
+instance, GPs have been used to model stellar rotation periods using a
+quasi-periodic kernel specified with multiple parameters which describe the
+stochastic variability of the star \citep{2018MNRAS.474.2094A}.
+\citet{2021AJ....162..124L} went a step beyond, starting with properties of
+distributions of stellar spots expressed in the \textsf{starry} formalism they
+derived a closed-form expression for a GP which describes a light curve of a
+rotating, evolving stellar surface conditioned on a distribution of stellar
+spot sizes, contrasts and latitudes.
+
+The advantage of GP models relative to alternatives such as linear regression
+with a fixed number of basis functions is that they are far more flexible and
+allow us to make less rigid assumptions about the phenomenon of interest. This
+section was just a brief summary of their key properties and there is much more
+to GPs that we haven't discussed (for example, fitting GPs to multi-dimensional
+data). In Chapter~\ref{ch:mapping_exoplanets} we will show two very different
+use cases of GPs in the context of occultation mapping.
+
+\section{Statistical inference - computation}
+\label{sec:inference_practice}
+In the previous section I briefly summarized some key aspects of probability theory
+and Bayesian statistics, I discussed maximum likelihood estimation and
+linear models and GPs for which inference can be done analytically.
+In most most cases, models of real world interest are not so well behaved and maximum
+likelihood estimation or obtaining samples from the posterior
+distribution becomes very challenging because of complex likelihood and posterior
+landscapes. In those cases computation needs goes hand-in-hand with model building and it
+is very important to understand how the strucuture of the model and the data maps to the
+structure of the likelihood.
+I start by introducing some very fundamental
+concepts such as the \emph{curse of dimensionality} and the \emph{no free lunch theorems}.
+Next, I discuss various computational inference methods such
+as optimization methods, Markov chain Monte Carlo and Variational Inference.
+Finally, I discuss model comparison through Bayes factors and cross validation.
+
+\subsection{The curse of dimensionality and the no-free-lunch theorem}
+The \emph{curse of dimensionality} refers to a phenomenon where the
+\emph{volume} of \emph{possible} explanations for the data grows exponentially
+with increasing dimension of the problem. Consider the following
+example\footnote{ Taken from the excellent paper by \citet{arXiv:2104.00008}.
+}. There are images consisting of $n$ pixels where each of the pixels is either
+a black (0) or white (1). The number of possible images is $2^n$ which grows
+exponentially. If each image out of the $2^n$ possible images is described by
+one of two labels $A$, or $B$ then the total number of possible ways of
+labeling the set of images is $2^{2^n}$ which is a stupidly large number. If
+$n=9$, the number of ways to classify the set of all 9-bit black and white
+images is greater than the number of atoms in the observable universe. If the
+labels don't correlate with the properties of the images then one strategy for
+solving this problem would be the lookup table approach -- memorize each image
+with its label. The famous \emph{no-free-lunch theorem} (NFL) proved by the
+mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
+impossible to construct an algorithm which does better than the lookup table
+approach.
+
+Of course the assumption that the labels aren't correlated to features in the
+images is silly for any meaningful problem. For instance, all images of cats
+have certain things in common as do images of galaxies. The no-free-lunch
+theorem doesn't say that it is impossible to construct an algorithm to classify
+images of cats, it only says that there is no best algorithm \emph{if we
+    average over all possible inputs to the problem}. To construct an algorithm
+which works in practice we need to impose structure on the problem, what
+machine learning researchers call \emph{inductive bias}. A set of assumptions
+which work in one domain may not work well in another domain and no single
+model will work well in all domains.
+
+\subsection{Sampling vs. optimization}
+%When fitting a model to a scientific dataset we are usually interested in
+%finding some ``best'' estimate of the parameters of that model\footnote{This is
+%    opposed to machine learning where in the vast majority of cases the goal is
+%    \emph{prediction} and the parameters of the model (for instance, weights of a
+%    neural network) are not of interest.}. In science, simply having an estimate of
+%the parameters is not sufficient, we also require estimates of the
+%\emph{uncertainty} associated with those values. In the Bayesian formalism this 
+%means estimating the posterior distribution $p(\vec \theta \lvert \mathcal{D})$.
+
+Broadly speaking, in statistical inference we can differentiate between two
+classes of methods for fitting models to data. Optimization methods cast the
+problem of estimating the likelihood or the posterior as an (generally
+non-convex) optimization problem. The goal in optimization problems is to find
+a single point which minimizes or maximizes some scalar function called a
+\textsl{cost function} or an \textsl{objective function}. The space of possible
+cost functions is infinite but in probabilitic inference these will be (usually
+unnormalized) probability distributions such as likelihood and posteriors. An
+example would be MLE and MAP estimation, the Laplace approximation and
+variational inference.
+
+Sampling methods are a fundamentally different class of algorithms whose
+purpose is accurately computing multidimensional integrals such as with the
+goal of numerically computing \emph{integrals} such as expectation values over
+the distribution $p(\vec{\theta})$:
+\begin{equation}
+    \mu\equiv \mathbb{E}_{p(\vec \theta)}[g(\vec \theta)]=\int g(\vec \theta) p(\vec \theta) \ud \vec \theta\hquad ,
+    \label{eq:general_expectation}
+\end{equation}
+where $g(\vec \theta)$ is some arbitrary function such as the mean.
+These integrals are intractable when the dimension of $\vec{\theta}$ is greater than
+a few because  the volume of the parameter space grows exponentially and finding
+the the regions in parameter space where the integral is nonzero becomes ever harder.
+We can get around this problem by generating samples from the probability distribution
+$\vec{\theta}^{(s)}\sim p(\vec{\theta})$ and computing the integral using
+\textsl{Monte Carlo Integration} in regions where there is non-negligible probability:
+\begin{equation}
+    \hat{\mu}_\mathrm{MC}=\mathbb{E}_{p(\vec \theta)}[g(\vec \theta)] \approx \frac{1}{S} \sum_{i=1}^{S} g\left(\vec \theta^{(s)}\right)\hquad .
+    \label{eq:monte_carlo_estimator}
+\end{equation}
+where $\hat{\mu}_\mathrm{MC}$ is the simple Monte Carlo estimator of the expectation
+$\mu$.
+It can be shown that if the expectation $\mu$ exists, the estimator
+$\hat{\mu}_\mathrm{MC}$ is consistent (it asymptotically converges toward $\mu$ with
+increasing $S$) and unbiased (the mean of the sampling distribution of the estimator
+is equal to $\mu$). If the expectation of $g^2$ is also finite the central limit
+theorem holds and the error of the simple  Monte Carlo estimator decreases as
+$\sqrt{S}$. This error is independent of the dimension!
+The major challenge is how to generate those samples from $p(\vec{\theta})$. This
+is usually accomplished
+sampling numbers from probability distributions that are easy to sample from
+(most commonly multivariate normal distributions) and then reweighting those
+samples at each iteration of the algorithm so that they match the target
+distribution. Examples of sampling methods include Markov chain Monte Carlo
+(MCMC), importance sampling and rejection sampling.
+
+%The difference between optimization and sampling algorithms can often be fuzzy
+%because sampling problems can sometimes be cast as optimization problems. In
+%both cases the NFL theorem holds, meaning that no sampling or optimization
+%method works best across all problems. The fundamental problem in both cases as
+%that we can only evaluate the target density some finite number of times and
+%the challenge is to find high probability regions in a multi-dimensional
+%parameter space. Even in moderately high-dimensional spaces (on the order of 10
+%dimensions) this is an extremely challening problem which is why methods which
+%work well in low dimensions often fail spectacularly in high dimensions. Both
+%classes of methods often use information about the gradient of target
+%distribution to aid the exploration of the parameter space. In fact, optimizing
+%or sampling high dimensional distributions is often impossible without access
+%to (exacy) gradients of the target density.
+%
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/typical_sets.pdf}
+        \caption{Histograms showing the distance from the MAP point for a set of 
+        samples from $D$-dimensional multivariate normal distribution with 
+        zero mean and unit variance. With increasing number of dimensions the 
+        probability mass concentrates in a thin shell centered at the MAP point 
+        called the typical set. For $D=50$ the posterior samples are coming from 
+        a region more than 5 sigma away from the MAP point.}     
+               \label{fig:typical_sets}
+    \end{centering}
+\end{figure}
+Contrary to what we might intuitively think, in all but the lowest dimensional models
+samples from the posterior distribution are not located anywhere 
+near a point estimate such as the MAP point in the parameter space, despite the fact 
+that this is where the posterior density is highest.
+This happens because expectation values  depend not only on the density of the pdf 
+$p(\vec\theta)$ but also on the volume term 
+$\ud \vec\theta$ and the value of the integral is non-negligible only when both are 
+jointly maximized. In high dimensions the volume grows exponentially and most 
+of the volume ends up being concentrated far away from the mode so that 
+the product of the density and the  volume, the \textsl{posterior mass}, ends up 
+being concentrated in a thin shell around the mode called the \textsl{typical set} 
+\citep[for a good overview of this issue see][]{arXiv:1701.02434}.
+This point is illustrated in Figure~\ref{fig:typical_sets} using samples from 
+a $D$-dimensional multivariate normal distribution with zero mean and unit 
+variance. 
+The reason this happens is that  although the most likely points are located 
+in the neighbourhood of the mode, 
+there are many more points further away in the typical set and those are the 
+ones that contribute to the integral involved in the calculation of expectation 
+values.
+For a toy example shown in Figure~\ref{fig:typical_sets} the fact this does not 
+matter much because the posterior has spherical symmetry but  for non Gaussian
+posteriors we can get very different results  when computing the mean (via 
+sampling) and the MAP point (via optimization).
+
+\subsection{Optimization methods}
+In this section we provide a brief overview of two most relevant methods for
+estimating the posterior distribution $p(\vec{\theta}\lvert \mathcal{D})$ using
+optimization algorithms.
+
+\subsubsection{The Laplace approximation}
+One of the simplest way of approximating a posterior distribution is the
+\textsl{Laplace approximation}. We start with the general posterior density of
+the form
+\begin{equation}
+    p(\vec{\theta}\lvert\mathcal{D})=\frac{1}{\mathcal{Z}}e^{-\mathcal{E}(\vec{\theta})}
+\end{equation}
+where $\mathcal{E}(\vec{\theta}) \equiv \ln\,p(\vec{\theta}, \mathcal{D})$
+is called an \textsl{energy function} and $\mathcal{Z}$ is the fully marginalized
+likelihood. Following \citet{murphy_book_2022}, we Taylor expand $\mathcal{E}(\vec{\theta})$ around the mode
+$\hat{\vec{\theta}}$ as follows
+\begin{equation}
+    \mathcal{E}(\vec{\theta}) =\ln\,p(\vec{\theta}, \mathcal{D})\approx \mathcal{E}(\hat{\vec{\theta}})+(\vec{\theta}-\hat{\vec{\theta}})^{\intercal} \vec{g}+\frac{1}{2}(\vec{\theta}-\hat{\vec{\theta}})^{\intercal} \mathbf{H}(\vec{\theta}-\hat{\vec{\theta}})
+\end{equation}
+where $\vec{g}$ is the gradient at the mode (the Jacobian matrix), and $\vec{H}$
+is the \textsl{Hessian} evaluated at the mode. The gradient term by definition
+vanishes at the mode $\hat{\vec{\theta}}$ so we are left with
+\begin{equation}
+    \hat{p}(\vec{\theta}, \mathcal{D})=e^{-\mathcal{E}(\hat{\vec{\theta}})} \exp \left[-\frac{1}{2}(\boldsymbol{\theta}-\hat{\vec{\theta}})^{\intercal} \mathbf{H}(\vec{\theta}-\hat{\vec{\theta}})\right]
+\end{equation}
+from which it follows that
+\begin{equation}
+    \hat{p}(\vec{\theta} \lvert\mathcal{D})=\frac{1}{\mathcal{Z}}\, \hat{p}(\vec{\theta}, \mathcal{D})=\mathcal{N}\left(\vec{\theta} \lvert\hat{\vec{\theta}}, \mathbf{H}^{-1}\right)\hquad .
+\end{equation}
+Thus, we obtain the well known result that a Taylor expansion of an arbitrary
+distribution  $p(\vec{\theta} \lvert\mathcal{D})$ around the mode $\hat{\vec{\theta}}$
+truncated at the first non-vanishing term is simply a multivariate normal distribution
+centered at the mode whose covariance matrix is given by the inverse Hessian (which specifies
+the \emph{curvature} of the distribution) evaluated at the mode.
+
+Laplace approximation works very well with posteriors which are well described
+by a multivariate normal distribution. It is straightforward to compute because
+we can use an optimizer to find the MAP point and then estimate the Hessian
+numerically using finite difference approximation for the gradient or use
+automatic differentiation to obtain the exact Hessian (we'll cover automatic
+differentiation in the next section).
+
+\subsubsection{Variational inference}
+Another method which uses optimization to obtain an approximation to the true
+posterior is \emph{Variational inference} (VI). The advantage of VI over the
+Laplace approximation is that we can fit any distribution we like to the
+posterior. The key idea is to model an intractable density $p(\vec{\theta}
+    \lvert\mathcal{D})$ with some off-the-shelf distribution $q(\vec{\theta})$ such
+that \emph{we minimize a measure of discrepancy between the two distributions}.
+We optimize for the parameters of the distribution $q(\vec{\theta})$
+$\vec{\psi}$ which are called the \textsl{variational parameters}. The most
+common discrepancy measure is the KL divergence
+(Equation~\ref{eq:kl_divergence})from $q$ to $p$. Thus, the optimization
+problem is to find the parameters $\vec{\psi}^*$ which minimize the KL
+divergence:
+\begin{equation}
+    \boldsymbol{\psi}^{*}=\underset{\boldsymbol{\psi}}{\operatorname{argmin}}\,D_{\mathbb{K} \mathbb{L}}(q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})\, \|\, p(\boldsymbol{\theta} \lvert\mathcal{D}))
+    \label{eq:variational_inference}
+\end{equation}
+Making use of Equation~\ref{eq:kl_divergence}, we can expand the above equation as \citep{murphy_book_2022}
+\begin{align}
+    \vec{\psi}^* & =\underset{\vec{\psi}}{\operatorname{argmin}}\, \mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}\left[\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})-\ln \left(\frac{p(\mathcal{D} \lvert\boldsymbol{\theta}) p(\boldsymbol{\theta})}{p(\mathcal{D})}\right)\right]                           \\
+                 & =\underset{\vec{\psi}}{\operatorname{argmin}} \,\underbrace{\mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}[-\ln p(\mathcal{D} \lvert\boldsymbol{\theta})-\ln p(\boldsymbol{\theta})+\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})]}_{-\mathrm{ELBO}(\boldsymbol{\psi})}+\ln p(\mathcal{D})
+\end{align}
+Since the evidence $\ln\,p(\mathcal{D})$ is not a function of the variational parameters
+$\vec{\psi}$, the problem of solving Equation~\ref{eq:variational_inference} reduces
+to maximizing the term
+\begin{equation}
+    \mathrm{ELBO}(\boldsymbol{\psi}) \equiv \mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}[\ln p(\mathcal{D} \lvert\boldsymbol{\theta})+\ln p(\boldsymbol{\theta})-\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})]
+\end{equation}
+which is called the \textsl{evidence lower bound} (ELBO) because
+$D_{\mathbb{K} \mathbb{L}}(q \| p) \geq 0$ implies that
+$\mathrm{ELBO}(\boldsymbol{\psi}) \leq \ln p(\mathcal{D})$.
+The most common choice for the distribution $q$ is, unsurprisingly, a multivariate
+Gaussian. VI with a multivariate Gaussian differs from  the Laplace aproximation
+because  in VI we are optimizing for both the mean and the covariance matrix rather
+than equating the covariance matrix to the inverse Hessian at the MAP point.
+VI is most commonly used with very large datasets and high dimensional parameter
+spaces because it is computationally cheaper than MCMC (although VI and MCMC are not
+doing the same thing). It has seen little use within astronomy.
+
+\subsection{Sampling methods}
+\label{ssec:sampling_methods}
+\subsubsection{Rejection sampling}
+Rejection sampling is one of the simplest algorithms for generating samples
+from a probability distribution. Consider a normalized pdf
+\begin{equation}
+    p(\vec{\theta})=\frac{\tilde{p}(\vec{\theta})}{\mathcal{Z}}
+\end{equation}
+where $\tilde{\vec{\theta}}$ is the unnormalized pdf and $\mathcal{Z}$ is normalization
+constant which isn't necessarily known.
+If a proposal distribution $q(\vec{\theta})$ (which we know how to sample from)
+satisfies the inequality
+\begin{equation}
+    Cq(\vec{\theta})\geq \tilde{p}(\vec{\theta})
+    \label{eq:rejection_sampling_constraint}
+\end{equation}
+for some constant $C$, the function $Cq(\vec{\theta})$ provides an upper envelope for
+$\tilde{p}$. We can use the proposal distribution to generate samples from the target
+distribution using the following algorithm:
+\begin{itemize}
+    \item Sample $\vec{\theta}_0\sim q(\vec{\theta})$ from the proposal distribution.
+    \item Sample $u_0\sim \mathrm{Unif}(0, Cq(\vec{\theta}_0))$ which corresponds to
+          uniformly drawing a height under the envelope.
+    \item If $u_0>\tilde{p}(\vec{\theta}_0)$ reject the sample, otherwise accept
+\end{itemize}
+See \citet{murphy_book_2023} for the proof that this procedure works.
+If $\tilde{p}$ is a normalized target distribution the acceptence probability at
+each iteration is  $1/C$ (TODO: show this) so  $C$ should be as small as possible
+while satisfying the constraint in Equation~\ref{eq:rejection_sampling_constraint}.
+In practice, if the distribution $p(\vec{\theta})$ represents a posterior distribution
+$p(\vec{\theta}\lvert \mathcal{D})$ rejection sampling can be accomplished by using
+the prior as the proposal distribution, sampling a very large number of parameter
+vectors from the prior, computing the value of the likeliood
+$p(\mathcal{D} \lvert\vec{\theta})$ for each and then setting $C$ to be equal to the
+maximum value of the likelihood across all samples.
+
+The major drawback of rejection sampling is that it scales badly with
+increasing dimensionality of the target distribution. For instance, if
+$p(\vec{\theta})=\mathcal{N}(\vec{0}, \sigma_p^2\vec{I})$ and
+$q(\vec{\theta})=\mathcal{N}(\vec{0}, \sigma_q^2\vec{I})$ we need to have
+$\sigma_q^2\sigma_p^2>\sigma_p^2$ to satisfy the bound. In $D$ dimensions the
+optimal value of $C$ is $C=(\sigma_q/\sigma_p)^D$ and since the acceptance
+probability is $1/C$ this decreases exponentially fast with dimension $D$. If
+the the proposal distribution $q(\vec{\theta})$ approaches $0$ in the tails at
+a faster rate than $p(\vec{\theta})$ as $\vec{\theta}\rightarrow \infty$ than
+their ratio $p(\vec{\theta})/q(\vec{\theta})$ approaches $\infty$ and the
+constraint in Equation~\ref{eq:rejection_sampling_constraint} will not
+satisfied. Thus, we must ensure that the target distribution does not have
+heavier tails than the proposal distribution.
+
+Even though it is the simplest sampling algorithm, rejection sampling can be
+very useful for problems with small numbers of parameters and small data sets
+because it can deal with very pathological distributions (for example highly
+multi-modal distributions). The most notable application of rejection sampling
+in astronomy is the \textsf{The Joker} sampler \citep{2017ApJ...837...20P} for
+sampling posterior pdfs for with very sparse binary-star and exoplanet radial
+velocity measurements. \citet{2020ApJ...895....2P} apply this sampler to obtain
+posterior pdfs in Keplerian parameters for $232 495$ sources from the APOGEE
+catalog. Becuase these measurements are very sparse (a few data points per
+star), the posterior pdf over the period prameter is highly multi-modal. The
+brute-force approach of rejection sampling works very well and produces
+independent samples from the distribution. For any individual object this pdf
+doesn't tell us much, however, when the analysis is repeated for thousands of
+objects in the framework of \textsl{hierarchical Bayesian inference} it is
+nevertheless possible to obtain very tight constrain on the population-level
+parameters. This is because even a highly multi-modal posterior pdf represents
+a major contraction of the prior parameter space and thus excludes certain
+parameter values.
+
+The radial veclocity model in \citet{2017ApJ...837...20P} interesting because
+structurally it very closely models the single lens microlensing model with
+parallax effects. Both models have few parameters 7 default parameters, two of
+which are linear and in both cases the posterior can be multi-modal (see
+Chapter~\ref{ch:microlensing} for more on this point). In
+\citet{2017ApJ...837...20P} they use the marginalized likelihood over the
+linear parameters and perform rejection sampling with the dimensionally reduced
+model. However, this requires an extremely large number of prior samples, on
+the order of hundreds of millions! Evaluating the likelihood this many times is
+much more computationally expensive in the microlensing case primarily because
+microlensing light curves have a lot more data points than stellar RV
+measurements. Neverthless, as I will show in Chapter~\ref{ch:microlensing}
+rejection sampling is intractable with the parallax microlensing model but it
+is tractable with the classic Paczynski model with 3 nonlinear parameters
+$(t_0, t_E, u_0)$.
+
+\subsubsection{Importance sampling}
+\textsl{Importance sampling} is not a an algorithm for generating samples from
+$p(\vec{\theta})$ but rather a method for estimating expectations
+$\mathbb{E}_p[g(\vec{\theta})]$ given an existing set of samples from
+some other distribution $q(\vec{\theta})$. We start by rewriting the integral 
+in Equation~\ref{eq:general_expectation} as 
+\begin{equation}
+    \mu = \mathbb{E}_{p}[g(\vec \theta)]=\int g(\vec \theta)\,\frac{p(\vec \theta)}{q(\vec \theta)}\,q(\vec \theta) \ud \vec \theta\hquad ,
+\end{equation}
+where we have introduced some other distribution $q(\vec \theta)$.
+The imporance sampling estimator of $\mu$ is then defined as
+\begin{align}
+    \hat{\mu}_{\mathrm{IS}}&=\frac{1}{S} \sum_{s=1}^{S} \frac{p\left(\boldsymbol{\theta}^{(s)}\right)}{q\left(\boldsymbol{\theta}^{(s)}\right)} g\left(\boldsymbol{\theta}^{(s)}\right)\\
+    &=\frac{1}{S} \sum_{s=1}^{S} w^{(s)} g\left(\boldsymbol{\theta}^{(s)}\right)
+\end{align}
+where  $\boldsymbol{\theta}^{(s)} \sim q(\boldsymbol{\theta})$ are samples from the 
+proposal distribution $q$ and $w^{(s)}$ are the \textsl{importance weights}. 
+$w^{(s)}$  are the ratios between the target distribution $p(\vec \theta)$ and 
+the proposal distribution $q(\vec \theta)$ evaluated at each $\vec{\theta}^{(s)}$.
+As in the case of rejection sampling, the requirement for the proposal distribution $q$
+is that it has support (is nonzero) wherever $p(\boldsymbol{\theta}) g(\boldsymbol{\theta})$
+is nonzero.
+Similarly to $\hat{\mu}_{\mathrm{MC}}$,  $\hat{\mu}_{\mathrm{IS}}$ is also a 
+consistent and  unbiased estimator of $\mu$ but its variance is heavily dependent on 
+how well the proposal distribution $q$ matches $p$. 
+If we do not know the normalization constants of $p$ and $q$ we can instead use the 
+self-normalized importance sampling (SNIS)  estimator 
+\begin{equation}
+    \hat{\mu}_{\mathrm{SNIS}}=\frac{\sum_{s=1}^{S} w^{(s)} h\left(\boldsymbol{\theta}^{(s)}\right)}{\sum_{s=1}^{S} w^{(s)}}
+\label{eq:importance_sampling_estimator}
+\end{equation}
+which is also consistent but it has a small bias of order $\mathcal{O}(1/S)$.
+There are also adaptive versions of importance sampling  which all rely on iteratively 
+adapting the poposal distribution $q$ based on the importance weights at each iteration.
+
+When the proposal distribution $q$ is a poor approximation of $p$ the distribution of 
+importance weights can have a heavy right tail which results in a large and 
+large and possibly infinite variance of $\hat{\mu}_{\mathrm{IS}}$.
+This is especially true in high dimensions when the importance weights $w^{(s)}$ can 
+vary by several orders of magnitude and the estimator in 
+Equation~\ref{eq:importance_sampling_estimator} becomes dominated by a few draws from 
+proposal distribution. One solution to this  problem is to fit a generalized 
+Pareto distribution\footnote{Pareto distributions are commonly used in \textsl{Extreme 
+Value Analysis} (EVA). EVA is a branch of statistics dealing with heavy and fat-tailed 
+distributions, a regime where many classical methods built around Gaussian assumptions 
+fall apart.} 
+to the distribution of importance weights and then use that distribution to produce 
+a uniformly spaced set of importance weights. This procedure is called 
+\textsl{Pareto Smoothed Importance Sampling} (PSIS) \citep{arXiv:1507.02646}.
+PSIS forms the basis of a very popular \textsf{R} package \textsf{loo} for 
+high-dimensional leave-one-out cross validation \citep{arXiv:1507.04544}.
+The PSIS algorithm also reports the estimate of the smoothness parameter in the
+generalized Pareto distribution (\textsl{Pareto $\hat{k}$}). $\hat{k}$ acts as a 
+diagnostic  because the PSIS estimate is likely to be unstable  if $\hat{k}$ is 
+too large \citep{arXiv:1507.02646}.
+
+What if we are not just interested in computing expectations but also want to obtain 
+posterior samples from $p(\vec{\theta})$ in order to, say, plot a historgram? We can 
+use \textsl{importance resampling} which works by resampling the original samples 
+$\vec{\theta}^{(1)} \dots \vec{\theta}^{(S)}$
+with replacament with probabilities proportional to  the importance weights.
+Practically, this meanse drawing a set of indices $i_1, \dots, i_S$ from a 
+multinomial distribution  where the event probabilities are the normalized importance 
+weights $\tilde{w}_{i}=w_{i} / \sum_{j} w_{j}$:
+\begin{equation}
+i_{1}, \ldots, i_{S} \sim \operatorname{Multinomial}\left(1, \tilde{w}_{1}, \ldots, \tilde{w}_{S}\right)\hquad .
+\end{equation}
+Samples with larger weights will appear more often than those with smaller weights. 
+We can use Pareto smoothing when computing the weights.
+
+In the context of astronomy, one  interesting (and relevant for the work presented 
+in thesis) application of importance sampling is for simplifying Bayesian hierarchical 
+inference. \citet{2010ApJ...725.2166H} showed that given posterior samples for a 
+collection of objects (obtained using MCMC for example) which were obtained with some 
+set of vague priors on the parameters of interest, we can re-weight those samples using 
+importance resampling under a different hierarchical prior which depends on some 
+population level parameters and in the process obtain a posterior over the population 
+level parameters. \citet{2010ApJ...725.2166H} applied this method to infer parameters 
+describing the eccentricity  distribution for a catalog of $N$ exoplanets given 
+posterior samples for the eccentricity for each individual planet. I will describe 
+this method in more detail in Chapter~\ref{ch:microlensing} and show how it can be
+applied to model the distribution of the $t_E$ parameter for a catalog of microlensing 
+events.
+
+\subsubsection{Rosenbluth-Metropolis-Hastings algorithm}
+The most popular class of algorithms used for generating samples 
+from a general pdf are 
+\textsl{Markov Chain Monte Carlo} algorithms or MCMC for short. The simplest 
+MCMC algorithm and the very first one that was developed is the
+\textsl{Rosenbluth-Metropolis-Hasting algorithm} (RMH) (also known as the 
+\textsl{Metropolis algorithm}). The RMH algorithm is one of 
+the most famous algorithms developed in the 20th century. The original algorithm was 
+published in \citet{1953JChPh..21.1087M} and it was extended by 
+\citet{1970Bimka..57...97H}. The original version is extremely simple. 
+Given a most recent sample $\vec\theta^{(s)}$, to generate the next sample
+\begin{itemize}
+    \item Draw a sample $\vec{\theta}^\prime$ from a proposal distribution $q(\bm\theta^\prime\lvert\vec\theta^{(s)})$ 
+    \item Draw a uniform random number $r\sim\mathcal{U}(0,1)$
+    \item If $\frac{p(\bm\theta^\prime)}{p(\vec\theta^{(k)})}$ then $\bm\theta^{(k+1)}\leftarrow \bm\theta^\prime$, else $\bm\theta^{(k+1)}\leftarrow\bm\theta^{(k)}$
+\end{itemize}
+The main characteristic of this algorithm is that it defines a biased random walk
+through the parameter space (usually moving in only one parameter at the time) 
+in such that the amount of time spent at a particular location $\vec\theta$ is 
+proportional to the target density $p(\vec\theta)$.
+These samples can then be used to approximate some expectation value of interest 
+by means of Equation~\ref{eq:monte_carlo_estimator}. The samples are always correlated 
+to en extent so estimates of the expectations don't converge as quickly with increasing 
+number of samples as they would if the samples were independent.
+
+The reason this algorithm works is because it defines a process called a 
+\textsl{Markov Chain} in which the probability of moving from a current state to a
+a successive state in the chain depends only on the current state and not any
+past states.
+Markov chains have stationary probability distributions over states and the stationary
+distribution for the Markov chain defined by the RMH algorithm turns out to be
+the target distribution $p(\bm\theta)$.
+For this to happen, the proposal distribution $q$ must satisfied a property
+called \textsl{detailed balance} which is defined as
+\begin{equation}
+    q(\bm\theta^\prime\lvert\bm\theta)=q(\bm\theta\lvert\bm\theta^\prime)\hquad .
+\end{equation}
+That is, the proposal has to be reversible so that the probability of going in either 
+direction is the same.
+
+The proposal distribution $q$ is one of the main tunable parameters
+of the RMH algoritm.
+The simplest choice for $q$ is a multivariate Gaussian distribution
+ $q(\bm\theta^\prime\lvert\bm\theta)\sim\mathcal{N}(\bm\theta,\vec \Sigma)$.
+ The covariance matrix $\vec \Sigma$ sets a characteristic 
+\textsl{step size} in the parameter space. If the step size is
+too large most proposals will get rejected and the chain might miss regions with high
+probability mass.
+If on the other hand the step size is too small, most steps are likely to be accepted
+but we might not explore the parameter space fully.
+RMH (and MCMC samplers in general) don't converge to the stationary distribution  
+of the target distribution immediately, instead there is always an initial exploration 
+phase before the chain reaches the relevant parts of the parameter space 
+(typical set).  The time it 
+takes to reach stationarity is called the \textsl{mixing time} of the sampler 
+or the \textsl{burn-in} period.
+
+All MCMC algorithms have the same goal, namely, generating samples from a
+target pdf such that the least number of samples $S$ are needed to accurately
+approximate the integral in Equation~\ref{eq:general_expectation}. Metrics
+used to judge the quality of different samplers include 
+\begin{itemize}
+    \item Efficiency -- what is the acceptance rate of the proposals?
+    \item Coverage -- is the target distribution explored completely?
+    \item Correlation -- how correlated are individual samples with past samples?
+    \item High dimensions -- can the sampler handle high dimensional spaces?
+    \item Derivatives -- does the sampler require gradients of the target density? 
+\end{itemize}
+In real world problems MCMC samplers require very careful tuning and initalization 
+to work well. Although RMH and other MCMC samplers should in principle converge to the posterior distribution 
+independent of the initialization point in the parameter space and the particular 
+choice of parametrization; in practice intialization, careful tuning of the proposal 
+distribution and how we parametrize the model is crucial for the performance 
+of the sampler. It is not an usual that a minor reparametrization of a given model leads 
+to an increase in sampling efficiency of several orders of magnitude\footnote{For this
+reason it is often much better to invest time into making a given model more friendly 
+to MCMC than to improve the evaluation time of the model by a factor of a few.}.
+
+\subsubsection{Hamiltonian Monte Carlo}
+Hamiltonian Monte Carlo (HMC) \citep{1987PhLB..195..216D,arXiv:1206.1901} is a 
+variant of  Markov Chain Monte Carlo which uses the gradient of the target 
+density to substantially  improve the sampling efficiency compared to 
+gradient-free samplers such as the RMH sampler, especially in high dimensions.
+The gradient information is especially useful for high-dimensional densities 
+because it enables the sampler to better 
+explore the typical set of parameter space by making proposals along the 
+the thin shell comprising the typical set\footnote{In high dimensions and 
+petrurbation away from the thin shell of the typical sets is likely to be 
+outside the tpical set which is why samplers which are not aware of the 
+geometry of the target density such as RMH work very poorly in high dimension.}.
+
+HMC works by transforming the sampling problem to a classical mechanics 
+problem expressed using the theory of Hamiltonian mechanics. 
+We start by transforming the parameter space $\boldsymbol{\theta}$ into phase space by 
+introducing a set of momentum parameters $\vec{p}$ (thus doubling the number of 
+parameters) and define a \textsl{Hamiltonian} of the system to be the negative logarithm 
+of the  joint density over $p(\vec{\theta}, \vec{p})$:
+\begin{equation}
+    \mathcal{H}(\boldsymbol{\theta},\mathbf p)\equiv -\ln p(\boldsymbol{\theta},\mathbf p)=-\ln p(\boldsymbol{\theta})-\ln p(\mathbf p\lvert\boldsymbol{\theta})\hquad .
+\end{equation}
+We can rewrite this Hamiltonian as
+\begin{equation}
+    \mathcal{H}(\boldsymbol{\theta},\mathbf p)= \mathcal{U}(\boldsymbol{\theta})  + 
+    \mathcal{K}(\boldsymbol{\theta},\mathbf p)
+\end{equation}
+where $\mathcal{U}(\boldsymbol{\theta})=-\ln p(\vec \theta)$ is a potential energy term,
+ $\mathcal{K}(\boldsymbol{\theta},\vec p)$ is a kinetic energy term and 
+ $p(\vec\theta)$ is the target  density.
+By mapping a posterior density to a physical system, we can use the machinery of 
+classical mechanics by first sampling the momenta conditional on the parameters 
+$\mathbf{p}\sim p(\mathbf{p}\lvert\boldsymbol{\theta})$
+ and then solving Hamilton's equations
+\begin{align}
+  \dot{\theta}_i&=\frac{\partial\mathcal{H}}{\partial p_i}=\frac{\partial\mathcal{K}}{\partial p_i}\\
+  \dot{p}_i&=-\frac{\partial\mathcal{H}}{\partial q_i}=-\frac{\partial\mathcal{U}}{\partial q_i} - 
+  \frac{\partial\mathcal{K}}{\partial q_i}
+\end{align}
+\textsl{Symplectic integrators} (integrators which respect Liouville's theorem of 
+classical mechanics) are used to integrate Hamilton's equation because of their 
+high accuracy. The most common choice for an integrator is the symplectic 
+(and reversible) \textsl{Leapfrog integrator}, which is defined by 
+\begin{align}
+\vec{p}_{t+1 / 2} &=\vec{p}_{t}-\frac{\eta}{2} \frac{\partial \mathcal{U}\left(\boldsymbol{\theta}_{t}\right)}{\partial \boldsymbol{\theta}} \\
+\boldsymbol{\theta}_{t+1} &=\boldsymbol{\theta}_{t}+\eta \frac{\partial \mathcal{K}\left(\boldsymbol{p}_{t+1 / 2}\right)}{\partial \boldsymbol{p}} \\
+\vec{p}_{t+1} &=\boldsymbol{p}_{t+1 / 2}-\frac{\eta}{2} \frac{\partial \mathcal{U}\left(\boldsymbol{\theta}_{t+1}\right)}{\partial \boldsymbol{\theta}}
+\end{align}
+where $\eta$ is the \textsl{step size}. The Leapfrog integrator works by performing a
+half-update of the momentum, followed by a full update of position and another half-update 
+of momentum.
+Because no integrator t preserve energy perfectly an RMH acceptence criterion of the form 
+$\alpha=\mathrm{min}\{1,\exp (\mathcal{H}(\boldsymbol{\theta}, \mathbf p) -\mathcal{H}(\boldsymbol{\theta}^\prime, \vec p^\prime))\}$ 
+is needed at the end of the inegration step in order to restore the detailed balance 
+property of Markov chains.
+For the pseudocode of a complete algorithm using the Leapfrog integrator see 
+for example \citet{murphy_book_2023}.
+
+The most challenging part of implementing HMC is tuning the various hyperaparamters of 
+the algorithm, namely, the number of leapfrog integration steps $L$, the step size
+$\eta$ and the kinetic energy term $\mathcal{K}$.
+The most common choice for the kinetic energy term is the Euclidian-Gaussian kinetic 
+energy distribution where  the conditional momentum distribution is simply a 
+multivariate normal distribution independent of the position:
+\begin{equation}
+\mathcal{K}(\vec\theta, \vec p) = \mathcal{N}(\mathbf p ; 0, \mathbf M)
+\end{equation}
+where the covariance matrix $\mathbf M$ is usually called the \textsl{mass matrix}.
+The choice of the mass matrix $\mathbf{M}$ is important for efficient sampling because 
+every transformation of the parameter space induces an inverse transformation on 
+momentum space \citep{arXiv:1701.02434}. Consider for example a a cigar shaped target 
+density.
+We can either transform the posterior such that it looks more Gaussian and make it
+ easier to sample from, or equivalently we can choose the momentum distribution such 
+ that the momenta are aligned with the shape of the cigar by having higher velocity 
+ proposals along the longer axis.
+The optimal choice for $\mathbf{M}$ is to choose the matrix to be as close as 
+possible to the covariance matrix of the parameter space.
+The mass matrix is usually estimated empirically from the covariance matrix of the 
+parameters during a burn-in phase of the sampling.
+
+As for the leapfrog parameters, we want to choose the number of leapfrog steps $L$ to be 
+large enough so that the algorithm explores large swathes of the parameter space.
+The most popular variant of HMC, the No U-Turn Sampler (NUTS) algorithm 
+\citep{arXiv:1111.4246} adaptively chooses $L$ such that Leapfrog integration is 
+automatically halted when the trajectory starts going back to its starting point (the
+the momentum vector points in the direction of the starting point).
+This trick ensures that parameter space updates are as large as possible and it 
+reduces autocorrelation between successive samples.
+In addition to automatically selecting $L$, \citet{arXiv:1111.4246} also propose a 
+scheme for adapting the step size $\eta$ on the fly. Adding the automatic adaptation
+of the mass matrix by running short chains during the burn-in period means that 
+NUTS removes any need to tuning. The lack of complicated tuning parameters and 
+the fact that NUTS works very well across a large set of problems are one of the 
+reasons why NUTS (and its derivations) is the work-horse algorithm 
+in many popular statistical modeling packages such as \textsf{Stan}
+\footnote{\url{https://mc-stan.org/}} \citep{2017JSS....76....1C}, 
+\textsf{PyMC}\footnote{\url{https://www.pymc.io/welcome.html}}, \textsf{numpyro}
+\footnote{\url{https://num.pyro.ai/}} and \textsf{TensorFlow Probability}
+\footnote{\url{https://www.tensorflow.org/probability}}.
+
+How well MCMC works depends almost entirely on the geometry of the target density
+(which is parametrization dependent). If the target density is non-Gaussian 
+HMC requires lots of leapfrog integration steps per iteration. It is entirely 
+possible to have a 3 dimensional density with poor geometry (for example 
+a multi-modal density with well seperated modes) for which HMC will 
+fail but if the geometry of the distribution is well behaved HMC can scale to 
+even millions of parameters.
+
+One other notable variant of Hamiltonian Monte Carlo which works better for 
+densities with highly complex geometries is 
+Riemannian Manifold Hamiltonian Monte Carlo (RMHMC) \citep{girolami2011}.
+RMHMC uses a position dependent kinetic energy term $\mathcal{K}$ and second order 
+gradient information (quantifying the curvature of the target density) to improve 
+sampling efficiency with highly correlated densities by adapting to the local geometry 
+of the target density at every point. RMHMC is very rarely used and there are no 
+good open-source implementations of the algorithm. The reason for this is that 
+it is  difficult to construct numerically stable integrators for RMHMC
+\footnote{\href{https://discourse.mc-stan.org/t/riemann-manifold-hmc-in-stan/19466/5}{This}
+discussion on the \textsf{Stan} forums is particularly illuminating}.
+Instead of using RMHMC, we can find better parametrizations for the target density
+and use NUTS to sample it which can end up being equivalent to RMHMC 
+\citep{arXiv:1910.09407}.
+
+For an excellent visualization of different MCMC samplers see 
+\href{http://chi-feng.github.io/mcmc-demo/app.html}{here}.
+
+\subsubsection{Diagnosing MCMC chains}
+How can we tell that our MCMC chains have converged to target density?
+Unfortunately there is no point at which we can say that the chain definitively
+converged to the target density and that we are sampling from the typical set of 
+the posterior density faithfully. At best, we can
+tell when the chain has not converged. The prime diagnostic for the convergence
+of MCMC chains is the \textsl{integrated autocorrelation time}.
+Given $S$ independent samples, the variance $\sigma^2$ for a Monte Carlo 
+estimate of an expectation value $g(\vec\theta)$  is 
+\begin{equation}
+    \sigma^2= \frac{1}{S} \mathbb{V}_{p}\left[g(\vec\theta)\right]
+    \label{eq:mcmc_error_indep}
+\end{equation}
+and the standard deviation of this estimator will decrease
+$\sim 1/\sqrt{S}$ with the number of samples.
+However, samples from $p(\vec\theta)$ produced by MCMC are not truly independent.
+The way we measure this correlation is by computing the integrated autocorrelation
+time $\tau_f$. If the samples are correlated, the the variance in 
+Equation~\ref{eq:mcmc_error_indep} becomes
+\begin{equation}
+    \sigma^2= \frac{\tau_f}{S} \mathbb{V}_{p}\left[g(\vec\theta)\right]
+    \label{eq:mcmc_error}
+\end{equation}
+where $\tau_f$ is the average number of steps needed before the chain ``forgets''
+its initial position. It is defined as
+\begin{equation}
+    \tau_f = \sum_{\tau=-\infty}^\infty\rho_f(\tau)
+\end{equation}
+where $\rho_f(\tau)$ is the normalized \textsl{autocorrelation function} (ACF) of 
+the stochastic process which  generated the chain for
+$f$ \footnote{See this blog post by Daniel Foreman-Mackey for more details on 
+autocorrelation functions: \url{https://dfm.io/posts/autocorr/}.}.
+Estimating integrated autocorrelation times is difficult and in general
+requires long chains because the estimate of $\tau_f$ is slow to converge.
+
+Using an estimate of integrated autocorrelation time  we can compute the 
+\textsl{Effective Sample Size} (ESS) for a given chain \citep{sokal1997}
+\begin{equation}
+\mathrm{ESS}\equiv\frac{S}{\tau_f}
+\end{equation}
+The integrated autocorrelation time (and hence also the ESS) depends on the function 
+$g(\vec\theta)$ and will be different depending which expectation value we 
+are interested in.
+For example if we are interested in calculating the mean and require a certain
+fractional precision we need to know the integrated autocorrelation time to estimate 
+how many effective samples are necessary to achieve this precision.
+\citet{arXiv:1903.08008} suggest computing ESS for both the bulk of the distribution 
+(captured by the median) and \textsl{tail-ESS} which they define to be the minimum 
+of the ESS for the 5\% and 95\% quantiles of the distribution. Tail-ESS will generally 
+be smaller  than the ESS for the median because MCMC converges at a slower rate 
+in the tail of the distribution.
+
+Since MCMC samples are correlated, some users discard all except every $n-th$ sample in a process called
+\textsl{thinning} the chains. Except for reasons of storage in memory, this
+procedure should be discouraged because discarding information never helps with
+better estimates and the Monte Carlo error in Equation~\ref{eq:mcmc_error} 
+already incorporates the fact that the samples are correlated.
+
+Another popular statistic for the diagnostic of MCMC chains is the
+\textsl{Gelman-Rubin} $\hat R$ statistic \citep{1992StaSc...7..457G} which uses 
+information pooled from multiple chains to judge convergence. It is 
+defined as
+\begin{equation}
+    \hat R= \frac{\hat V}{W}
+\end{equation}
+where $W$ is the sample variance within a single chain, and $\hat V$ is the
+variance estimate pooled across several independent chains.
+The $\hat R$ statistic converges to 1 when each of the chains accurately 
+sample the target density. A value greater than 1 indicates that one or more 
+chains have not yet converged.
+
+Finally, the most sophisticated diagnostic for the convergence of MCMC chains
+attempts to detect the the breakdown of the geometric ergodicity property of
+MCMC. Geometric ergodicity is a necessary condition for MCMC estimators to
+follow the central limit theorem ensures that the estimator in 
+Equation~\ref{eq:monte_carlo_estimator} is consistent and unbiased.
+HMC algorithms enable the detection of so called
+\textsl{divergences} 
+\footnote{\url{http://mc-stan.org/users/documentation/case-studies/divergences_and_bias.html}.}.
+ in the parameter space -- points in the parameter space 
+where the target density has very large gradient indicating 
+a suspected  breakdown of geometric ergodicity.
+The presence of a large number of divergences in a model can lead to 
+biased estimators. 
+Modern statistical modeling packages include functions for computing ESS 
+estimators,  $\hat R$ and divergences. Without the use of these diagnostic 
+tools we cannot be sure that the MCMC chains have converged and that our results 
+are correct.
+
+\subsubsection{Failure modes for samplers}
+MCMC methods work really well for distributions with relatively well-behaved 
+geometries but tend to fail with more complex distributions, particularlly 
+multi-modal distributions with well seperated modes. 
+ \citet{arXiv:2101.09675} lists 
+key properties of densities for which MCMC methods tend to fail:
+\begin{itemize}
+    \item \textbf{Highly non-Gaussian distributions}: For example banana shaped
+    distributions, distributions with heavy tails.
+    \item \textbf{Multimodal distributions}: If there are multiple well 
+    separated  modes in the target densities, even if each mode is individually 
+    well described by a Gaussian MCMC methods will tend to get stuck in one of
+    the modes and fail to jump between modes.
+    \item \textbf{High dimensionality}: High dimensional problems can break pretty 
+    much every algorithm, they are however less of a problem for HMC like algorithms.
+    \item \textbf{Highly informative distributions}: If the data is particularly 
+    informative the posterior occupies an extremely small portion of the total 
+    prior volume and it may be challenging to discover those regions. Even HMC 
+    can fail here if the posterior mass is concentrated  in a very narrow 
+    ``canyon'' if the momentum of the proposals is too large.
+    \item \textbf{Phase transitions}: These are abrupt changes in the structure 
+    of the likelihood as we move up constant likelihod contours.
+\end{itemize}
+The first and second failure modes are particularly important for this thesis 
+because likelihoods for microlensing models are highly non-Gaussian and 
+multimodal. In the next section we cover a class of algorithms which (on paper!)
+addresses almost all of these challenges.
+
+\subsubsection{Nested Sampling}
+\textsl{Nested sampling} \citep{2004AIPC..735..395S}  is fundamentally different 
+from MCMC algorithms  because it was designed for the purpose  of estimating the
+Bayesian evidence  $\mathcal{Z}$ (Equation~\ref{eq:evidence}) instead of drawing samples 
+from  a target distribution.
+Since $\mathcal{Z}$ is just a normalization constant, it is usually ignored 
+in traditional MCMC algorithms such as RMH which don't require the target 
+density to be properly normalized.
+However, $\mathcal{Z}$  becomes important in the context of model comparison 
+using \textsl{Bayes factors} (see Section~\ref{ssec:model_comparison}) which
+are proportional to the ratios of the evidences of two models.
+
+In addition to estimating $\mathcal{Z}$ Nested Sampling produces samples 
+from posterior $p(\vec\theta\lvert \mathcal{D})$ as a sort of byproduct. 
+The major advantage of NS over MCMC methods is that acts as a sort of 
+global search algorithm working from the outside-in which improves its 
+performance on highly multi-modal distributions\footnote{
+    The extent to which this claim is true will be investigated in 
+    Chapter~\ref{ch:microlensing}.
+}. Nested Sampling has been extensively used in cosmology in particular, mostly 
+thanks to the popularity of the \textsf{MultiNest} package \citep{arXiv:0809.3437}. 
+\citet{arXiv:2205.15570} is great introductory review to these methods while 
+\citet{arXiv:2101.09675} goes into a lot more depth.
+
+We start by rewriting the evience $\mathcal{Z}$ from Equation~\ref{eq:evidence}
+as 
+\begin{equation}
+\mathcal{Z}=\int \mathcal{L}(\vec\theta) \pi(\vec \theta)\ud\vec\theta
+\label{eq:evidence2}
+\end{equation}
+where  $\mathcal{L}\equiv p(\mathcal{D}\lvert\vec \theta)$ is the likelihood 
+and $\pi(\vec \theta)$ is the prior. The key idea behind NS is that it is possible 
+to replacec the high-dimensional integral in Equation~\ref{eq:evidence2} with 
+a one-dimensional integral over the entire prior volume: 
+\begin{equation}
+    \mathcal{Z}=\int_{0}^{1} \mathcal{L}(X) \ud X
+    \label{eq:nested_sampling}
+\end{equation}
+where $\ud X$ is the volume element of points in the parameters space which 
+share the same likelihood $\mathcal{L}(X)$ weighted by the prior $\pi(\vec\theta)$:
+\begin{equation}
+X\left(\mathcal{L}^{\star}\right)=\int_{\mathcal{L}>\mathcal{L}^{\star}} \pi(\vec\theta) \ud \vec\theta\hquad .
+\end{equation}
+$\mathcal{L}=\mathcal{L}^\star$ defines a contour of constant likelihood and 
+$X(\mathcal{L}^\star)$ is the prior volume enclosed by that contour.
+\emph{Nested sampling thus replaces the original problem of sampling from the posterior 
+to  sampling from prior with a likelihood constraint}.
+For a derivation of Equation~\ref{eq:nested_sampling} see \citet{arXiv:2205.15570}.
+
+The standard version of the NS algorithm works by drawing $K$ \textsl{live points} 
+(points in the parameter space) from the prior at each iteration $i$, removing 
+the point with the lowest likelihood value $\mathcal{L}_i$ and replacing it 
+with a new live point sampled from the prior and subject to the 
+constraint $\mathcal{L}_{i+1}>\mathcal{L}_i$ (see \citet{arXiv:2205.15570} 
+for details). This process continuously shrinks the volume of the prior by an 
+approximately constant factor, effectively scanning the parameter space from 
+contours with a lowest likelihood to contours with a highest likelihood.
+The algorithm terminates when some fractional error tolerance on the estimation 
+of the evidence is reached. 
+At the end of the iterative procedure we are left with 
+estimates of the volume $X(\mathcal{L}^\star)$ at each iteration and we can 
+estimate the integral in Equation~\ref{eq:nested_sampling} using the trapezoidal 
+rule:
+\begin{equation}
+\mathcal{Z}=\sum_{i=1}^{K}w_i\,\mathcal{L}_i^\star\hquad ,
+\end{equation}
+where the weights $w_i$ are 
+\begin{equation}
+w_{i}=\frac{1}{2}\left(X_{i-1}-X_{i+1}\right)\hquad .
+\end{equation}
+The posterior samples can be obtained from the from the discarded live points 
+by assigning weights $\hat{w}_i$ to each live point:
+\begin{equation}
+    \hat{w}_i\equiv \frac{\mathcal{L}_i\,w_i}{\hat{\mathcal{Z}}}
+\end{equation}
+To compute the ESS for these samples, we can use the exectation value dependent 
+estimator provided by \citet{arXiv:1809.04129}:
+\begin{equation}
+    \widehat{\mathrm{ESS}}=\frac{1}{\sum_{s=1}^S\tilde{w}_i^2}
+\end{equation}
+where
+\begin{equation}
+\tilde{w}_{i}\equiv\frac{\left|g\left(\vec{\theta}^{(s)}\right)\right| \hat w_{s}}{\sum_{i=1}^{S}\left|g\left(\vec{\theta}^{(i)}\right)\right| \hat w_{i}}
+\end{equation}
+
+\cite{arXiv:2101.09675} identifies three distinct phases in the behavior 
+of the NS algorithm. In the first phase the prior volume shrinks towards the 
+bulk of the posterior (the typical set) and during this phase the live points 
+vary in likelihood by many orders of magnitude. All of the probability mass is 
+associated with a single live point because the live points are all associated 
+with an equal volume. 
+The typical set is resolved in the second phase and the live points have 
+comparable wieghts.  In the final stage NS reaches the region around the MLE 
+with large values of the likelihood but very small volume.
+Unlike MCMC, NS has a well defined termination criterion and running the algorithm 
+for a longer time does not produce more posterior samples (there are other ways 
+of producing more samples after an NS run has terminated).
+NS is different compared to HMC in that HMC explores level sets of energy where 
+the Hamiltonian is constant while NS scans through a range of energy levels. 
+It's this feature that enables it to discover multiple modes in the density.
+
+Historically NS has been  extensively used by physicists and almost completely 
+ignored by statisticians. This has been changing in recent years and there 
+is now more work from both communities aimed at understanding and improving the 
+algorithm and better diagnosing its failure modes (see for example 
+\citet{ arXiv:1407.5459,arXiv:1804.06406,arXiv:1704.03459,arXiv:2101.09675}).
+For example, a recent paper by \citet{arXiv:1805.03924}  reframes the NS 
+algorithm as a special case of \textsl{Sequantial Monte Carlo}, a class of 
+algorithms that have been extensively studied by statisticians for decades.
+There are also lots of modern open-source implementations of NS such as 
+\textsf{dynesty} \citep{arXiv:1904.02180}, \textsf{JAXNS} \citep{arXiv:2012.15286}
+and \textsf{UltraNest} \citep{2021JOSS....6.3001B}. 
+\textsf{UltraNest} in particular is the most developed modeling framework 
+focused on correctness of posterior inferences. Ultimately, the only way 
+to check whether NS is doing what it promises to do is to apply it to a real 
+world problem. In Chapter~\ref{ch:microlensing} we do just that and compare 
+the results and performance of NS to HMC.
+
+\subsubsection{Ensemble samplers with affine invariance}
+By far the most popular sampler in physics and astronomy is an MCMC method 
+called \textsl{affine invariant ensamble sampler} (AIES)
+\citep{2010CAMCS...5...65G} which was popularized by an excellent \textsf{Python}
+implementation of the algorithm \textsf{emcee} 
+\citep{arXiv:1202.3665,arXiv:1911.07688}. The method is similar 
+to RMH  in the sense that it does not use gradient information and it requires 
+only that the evaluation of the log-probability function. The special property 
+of affine invariant MCMC is that it is invariante to 
+\textsl{affine transformations} (linear transformations such as 
+scale transformations and rotations) of the parameter space. It uses multiple 
+MCMC chains (called \textsl{walkers}) to explore the parameter space which 
+are correlated by construction.
+
+\textsf{emcee} is so popular that the 
+original paper has been cited almost 8000 times so far.  
+The success of \textsf{emcee} has less to do with its actual suitability 
+to a wide range  of problems and more to do with the fact that it is very easy to use with the 
+kinds of models astronomers and physicists work with. \textsf{emcee}  works with
+black-box physics simulators and it is also easily parallelizable. 
+This feature also makes it easy to misuse and apply to problems for which it is 
+not suitable. For example, \citet{arXiv:1509.02230} showed that \textsf{emcee} 
+does not work well in high dimensions (greater than about 10) and it can appear 
+as if the sampler converged when in reality it has not. Nevertheless 
+\textsf{emcee} can still be very useful and fast for very low dimensional 
+problems. As with other MCMC samplers, it will fail with multi-modal 
+likelihoods.
+
+\subsubsection{Other methods}
+That there are many many algorithms in the  literature besides the ones I've covered here. 
+The vast majority of these  algorithms are variants of RHMC, HMC or  
+NS. Most of these algorithms don't have well tested opens-source implementations, 
+the examples presented in the papers are usually low-dimensional toy 
+problems, and the algorithms require very careful tuning to work at all. This 
+is why the development of general purpose statistical modeling 
+(probabilistic programming) packages such as \textsf{Stan} and 
+\textsf{PyMC} has been so important. They demonstrated that methods such as 
+NUTS work well for a wide variety of problems.
+
+\subsection{Model validation and comparison}
+\label{ssec:model_comparison}
+
+So far I have discussed inference of model parameters using either optimization 
+methods such as maximum likelihood or sampling methods such as MCMC and Nested 
+Sampling.
+Using these methods we can obtain some estimates of the parameters of a specific  
+model but this  does not tell us if the model is any good in the first place.
+Every time we are doing parameter inference we assume that the particular model
+is true.
+A major challenge in statistical inference is comparing different 
+models which can explain the same dataset, so called \textsl{model comparison}.
+What we mean by different models is somewhat fuzzy. We could for example have 
+a model specified by the same likelihood function and a set of continous 
+parameters and then call different realizations of those parameters different 
+models. This does not make much sense in the Bayesian context. In most cases
+when a model comparison problem can be cast as a parameter inference problem it 
+is easier to treat it as a parameter inference problem\footnote{Although, as we
+will see in Chapter~\ref{ch:microlensing} it may be advantageous  to use model 
+comparison methods when comparing different modes in a multi-modal posterior.}.
+This happens when two \emph{models are nested}, meaning they share one or 
+several parameters.
+For example, let's say we are fitting a microlensing model and trying to decide 
+whether  or not to include the finite source effect parametrized by a continuous 
+parameter $\rho_\star$.
+We could fit two separate models, a model which uses a point source 
+aproximation for the magnification and one which includes the finite source 
+effects via the $\rho_\star$ parameter and then decide which model better 
+describes the data. The alternative is to simply fit the more complex model, 
+obtain a posterior for the $\rho_\star$ parameter and check if it differs from 
+zero in a meaningful way. Although the former approach is more elegant sometimes
+it is practically intractable because we might end up having trouble sampling 
+the posterior if the additional parameter is not well constrained by the data\footnote{In those
+cases is may nake sense to put a strong prior pushing the parameter to zero which 
+helps with inference. If the parameter is well constrained by the data the 
+likelihood will push the posterior away from zero.} or if the model with the 
+extra parameter is much more computationally expensive. 
+If the models are nested but the paramer which differentiates between models is 
+discrete it is sometimes still possible to cast the comparison problem as a parameter 
+inference problem using \emph{dimension jumping} methods such as 
+\textsl{Reversible-jump MCMC} (RJ-MCMC)
+\citep[See][for an application of a method similar to RJ-MCMC to an 
+astronomy problem]{2015MNRAS.448.3206B}.
+However, these methods are computationally very difficult to implement.
+
+When the models are not nested we have no choice but to do model comparison. 
+An example would be fitting a single lens microlensing model vs. a binary lens 
+or a binary source star model. Statistics alone does not tell us what to put in the 
+abstract of a paper after we have completed the analysis for multiple competing 
+models\footnote{This is the subject of decision theory which relies on 
+specifying \textsl{utility functions} (which depend on our posterior beliefs) 
+to quantify expected loss given a certain decision}. Best we can hope for is 
+assigning probabilities to each of the models taking into account how well they 
+fit the data and how parsimonious they are. Model 
+comparison is one of the key problems with analyzing microlensing light curves 
+because there are often multiple very different models which make near-identical 
+predictions  
+\citep[see][for examples from microlensing]{ 2017AJ....153..129J,2020AJ....160...17H,2021AJ....162...59R}.
+Especially problematic is the fact that when multiple models equally well 
+describe the data, \emph{some models will be more interesting than others}.
+For example one model might be a triple lens model implying the discovery of 
+an \textsl{exomoon} in a planetary system. The other model could be a binary 
+source star model. The major issue here is that the incentives of researchers 
+are not always aligned with truth-seeking and the outcome of a model 
+comparison test can be a strong function of how much effort was invested into 
+a given model\footnote{This is one of the causes of the replication crisis in 
+many scientific disciplines.}. In the extreme case, if we only consider one interesting model then
+there is no chance we can conclude that  a more mundane model describes the 
+data equally well. This is why it is very important to have a clearly defined 
+framework for assesing different models.
+
+\subsubsection{Frequentist model comparison (hypothesis testing)}
+The question of model comparison is different depending on whether
+one uses the frequentist or Bayesian statistics. In classical frequentist
+statistics, one has to form the \textsl{null hypothesis} $\textrm H_0$ which 
+represents a baseline (null) model and an \textsl{alternative hypothesis} which is 
+the hypothesis of interest. For instance, the two hypotheses could be
+\begin{align}
+    \textrm H_0: & \; \textrm{there is no planet in the data} \notag\\
+    \textrm H_1: & \;\textrm{there is a planet in the data}\notag
+\end{align}
+We then form a \textsl{hypothesis test} under some \textsl{test statistic} 
+and based on the outcome of that hypothesis test we either reject the null 
+hypothesis or we don't, depending on an arbitrary threshold called the 
+\textsl{p-value}.
+
+The p-value is the probability of obtaining data equal to or more extreme than 
+what is observed, \emph{assuming that the null hypothesis $\textrm H_0$ is true}.
+It is the probability for obtaining the observed
+dataset in an infinite set of draws of the data, \emph{it is not a statement 
+about the probability of the hypothesis}\footnote{The definition of the p-value
+is so confusing that even some statistics textbooks get it wrong.}.
+All hypothesis testing does is answering the following question:
+assuming there's nothing interesting in the data, how likely was I to observe 
+the data I was given.
+If a certain threshold p-value is reached ($0.05$ in most scientific disciplines, 
+substantially less in physics and astronomy), the null hypothesis hypothesis 
+is rejected.
+The fact the null hypothesis is rejected cannot be interpreted as evidence in
+favour of the alternative hypothesis because such a statement would involve
+using Bayes' theorem to invert the probabilities. However, it is almost always
+(mis)interpreted in that way.
+Given a set of models $\{\mathcal{M}_1,\mathcal{M}_2,\dots,\mathcal{M}_n\}$, all
+we are formally allowed to do is to  reject some of those models based on 
+a p-value, and keep the rest.
+
+One might wonder if there is as an absolute way of judging the quality of a
+single isolated model. In astronomical literature, this usually falls under the term of
+a \textsl{goodness of fit} statistic. These statistics implicitly use the 
+hypothesis testing procedure described above, rejecting the null hypothesis 
+under some p-value. Popular choices are the \textsl{Chi-squared test} 
+(not to be confused with the $\chi^2$ loss function),
+\textsl{Kolmogorov-Smirnov test}, \textsl{Anderson-Darling test}, etc.
+These tests rely on strong assumptions about the data generating process 
+and they are often misused. 
+It is often preferable (if possible) to just use visual checks on the predictions 
+of the model 
+\citep[see][for a visualization guide in the Bayesian paradigm]{arXiv:1709.01449}. 
+For example plotting the predictions of the model in data space, plotting 
+the residuals, etc.  Probabilistic modeling is almost always an iterative 
+process and visualization helps us spot where the model is failing to describe 
+the data adequatelly.
+
+\subsubsection{Bayesian model comparison using Bayes factors}
+In the Bayesian paradigm, the most natural way of comparing different 
+models is by making use of  Bayes' theorem. We can assign the probability of a 
+given model $\mathcal{M}$ as 
+\begin{equation}
+    p(\mathcal{M}\lvert\mathcal D)\propto p(\mathcal{M})\,p(\mathcal D\lvert\mathcal{M})
+\end{equation}
+where $p(\mathcal{M})$ is the prior probability of the model $\mathcal{M}$ and
+$p(\mathcal D\lvert\mathcal{M})$ is the Bayesian evidence. If we are comparing two 
+models, $\mathcal{M}_o$ with a set of parameters $\vec\theta_0$ and
+and $\mathcal{M}_1$ with a set of parameters $\vec\theta_1$, we can form the ratio of 
+the their posterior probabilities given the data
+\begin{equation}
+    \frac{p(\mathcal{M}_0\lvert\mathcal D)}{p(\mathcal{M}_1\lvert\mathcal D)} =
+    \frac{\mathcal{Z}_0}{\mathcal{Z}_1} \, \frac{p(\mathcal{M}_0)}{p(\mathcal{M}_1)}\hquad .
+\end{equation}
+The ratio of the two evidences on the right hand side is called the 
+\textsl{Bayes factor}, and the ratio of prior probabilities is the 
+\textsl{prior odds}.
+Prior odds can be set to unity under the assmuption that both models are equally 
+likely a priori.
+A large Bayes factor $\textrm{B}_{01}\equiv \mathcal{Z}_0/\mathcal{Z}_1$ should 
+be interpreted as evidence in favour of model 0 versus model 1, given the 
+observed data $\mathcal D$.
+As in the case of hypothesis testing, one can then define arbitrary criteria for
+the strength of the evidence in favour of model 0 conditional on the specific value
+of the Bayes factor.
+The popular choice is using Jeffreys scale \citep{1939thpr.book.....J}.
+Unlike the p-value, Bayes factors (multiplied by prior odds)
+are actual probabilities for the models
+themselves, rather than probabilities for the data in an infinite set of trials.
+One useful aspect of Bayesian model comparison is the fact that it automatically
+implements \emph{Occam's razor}, penalizing overly complicated models if there
+is a simpler alternative \citep[see Chapter 28 of][]{2003itil.book.....M}.
+The goal when using Bayes factor is to either select a single ``best'' model
+$\mathcal{M}_i$ or, the more Bayesian option, to average over a set of 
+multiple models using their  posterior probabilities 
+$p(\mathcal{M}_i\lvert\mathcal D)$.
+
+There are two major problems with Bayes' factor. The first one is the fact that
+whereas the Bayesian evidence $\mathcal{Z}$ could often be ignored when inferring the 
+parameters of the models, it is a crucial quantity in Bayesian model comparison. 
+Computing $\mathcal{Z}$ is very computationally expensive, error prone, and it
+ requires methods such as Nested Sampling.
+The other problem with Bayes
+factors is that the estimates of $\mathcal{Z}$ are particularly sensitive to the 
+width and the shape of te prior distribution for the parameters. For example, if 
+the prior for parameter $\theta_k$
+in model $\mathcal{M}_0$ is $\theta_k\sim \mathcal{N}(0, s^2)$, where $s$ is
+very large, the evidence will scale as $\mathcal{Z}\sim 1/s$ practically
+independently of the data. This kind of dependence of the evidences on the
+parameter priors is not unexpected, it means that models which ``waste'' prior
+parameter space will be penalized, but it is often undesirable\footnote{See the 
+excellent essay on Bayes factors by \citet{navarro_2020}.}.
+
+Thus, in order to use Bayes factors for model comparison, one has to be able to
+compute evidences, and think very carefully about the choice of priors. Despite
+these issues, Bayes factors have been used extensively in staistics and the 
+natural sciences, most notably in cosmology. For example,\citet{arXiv:1312.3529} 
+used Bayesian model comparison to estimate Bayes factors of 193 different 
+inflation models using the Cosmic Microwave Background
+data from the Planck satellite and Higgs inflation as the reference model.
+
+\subsubsection{Cross validation}
+The more tractable and robust alternative to estimating Bayes factors is using 
+\textsl{cross validation}. The key idea behind cross validation (CV) is to judge 
+different models based on how well they predict unseen data\footnote{This is 
+principle method of model comparison in machine learning. In machine learning 
+it is customary to split the original dataset into a \textsl{training dataset} 
+and a \textsl{validation dataset}. The training dataset is used to optimize 
+the main parameters of the model and the loss function evaluated on the 
+validation dataset is used to compare different models or set the hyperparameters
+of a given model.}. 
+CV can be used to assess the predictive power of a single model (as a sort of 
+goodness of fit metric) to compare or average multiple models. 
+The idea behind CV is to split the dataset into $K$ parts or \textsl{folds}, 
+re-fit the model including $K-1$ folds, evaluate the objective function (log 
+likelihood or posterior) and then repeat the procedure for each fold. We 
+can then for example add up the $K$ values and compute an estimate 
+the predictive performance of the model. 
+If we set $K$ to be equal to the number of data points $N$, then we 
+are talking about \textsl{leave-one-out cross validation} (LOO-CV).
+
+One measure of the predictive performance of a given model $\mathcal{M}$ which has 
+good theoretical properties is ther so-called 
+\textsl{exptected log predictive density} (ELPD) 
+\citep[see for example][]{10.1214/12-SS102}. It is defined as 
+\begin{equation}
+\operatorname{ELPD}(\mathcal{M})\equiv \int \ln p\left(\tilde{\vec y}_{i} \lvert \vec y, \mathcal{M}\right)\,p_{t}\left(\tilde{\vec y}_{i}\right) \ud \tilde{\vec y}_{i}
+\end{equation}
+where 
+\begin{equation}
+p(\tilde{\vec y}_i\lvert\vec y)=\int p\left(\tilde{\vec y}_{i} \lvert \vec \theta,\mathcal{M}\right)\,p(\vec\theta \lvert \vec y) \ud \vec\theta
+\end{equation}
+is the posterior predictive distribution for a new observation $\tilde{\vec y}_i$ that 
+was generated from some unknown true distribution $p_t(\tilde{\vec y}_i)$.
+ELPD captures the expected predictive performance of the model for some future data
+$\tilde{\vec y}$.  It tells us how well the model will generalize to new data.
+
+One method for estimating ELPD is using LOO-CV. In LOO-CV we compute $N$ leave-one-out 
+posterior distributions $p(\vec\theta\lvert \vec{y}_{-i})$ where $\vec y_{-i}$ 
+denotes all of the observations except the $i$-th observation. Using these 
+posterior estimates we can estimate the ELPD as 
+ \begin{align}
+\mathrm{ELPD}_{\mathrm{LOO}} &\approx\frac{1}{N} \sum_{i=1}^{N} \ln p\left(\vec y_{i} \lvert \vec y_{-i},\mathcal{M}\right) \\
+\label{eq:elpd_loo}
+&=\frac{1}{N} \sum_{i=1}^{L} \ln\int p\left(\vec y_{i} \lvert \vec \theta, \mathcal{M}\right) p\left(\vec\theta \lvert \vec y_{-i},\mathcal{M}\right) \ud \vec\theta \\
+\end{align}
+where $p(\vec y_i\lvert \vec \theta)$ is the likelihood for the $i$-th observation 
+and $p(\vec\theta\lvert \vec y_{-i},\mathcal{M})$ is the posterior for $\vec\theta$ 
+obtained by fitting the full dataset except the held-out observation $\vec y_{-i}$.
+
+A naive implementation LOO-CV is very computationally expensive because it 
+involves re-fitting the model $N$ times. Fortunately, there are reliable methods for 
+approximating it. 
+The most popular approximation uses importance sampling such that the target 
+distribution is the LOO posterior and the proposal distribution is the full
+posterior. Given $S$ draws from the full posterior 
+$\vec\theta\sim p(\vec\theta|\vec y, \mathcal{M})$ we can estimate 
+$p(\vec y_i\lvert \vec y_{-i},\mathcal{M})$ as 
+\begin{align}
+\ln p\left(\vec y_{i} \lvert \vec y_{-i},\mathcal{M}\right)&\approx\ln\left(\frac{\frac{1}{S} \sum_{s=1}^{S} p\left(\vec y_{i} \lvert\vec \theta^{(s)},\mathcal{M}\right) w\left(\vec\theta^{(s)}\right)}{\frac{1}{S} \sum_{s=1}^{S} w\left(\vec\theta^{(s)}\right)}\right) \\
+\label{eq:prob_loocv_pointwise}
+w\left(\vec\theta_{s}\right)&=\frac{p\left(\vec\theta^{(s)} \lvert \vec y_{-i},\mathcal{M}\right)}{p\left(\vec\theta^{(s)} \lvert\vec y,\mathcal{M}\right)} \propto \frac{1}{p\left(\vec y_{i} \lvert\vec\theta^{(s)},\mathcal{M}\right)}
+\end{align}
+If particular observations are highly influential (removing one data point 
+noticeably changes the results) the target LOO posterior can differ 
+from the full posterior which is why Pareto smoothed importance sampling (PSIS) 
+(see Section~\ref{ssec:sampling_methods}) is used to improve the estimate. 
+Using Equation~\ref{eq:prob_loocv_pointwise} and Equation~\ref{eq:elpd_loo}
+we define \citep{2015arXiv150704544V} 
+\begin{equation}
+\mathrm{ELPD}_\mathrm{PSIS-LOO}\equiv\sum_{i=1}^N\ln\left(\frac{\frac{1}{S} \sum_{s=1}^{S} p\left(\vec y_{i} \lvert\vec \theta^{(s)},\mathcal{M}\right) w\left(\vec\theta^{(s)}\right)}{\frac{1}{S} \sum_{s=1}^{S} w\left(\vec\theta^{(s)}\right)}\right) \hquad .
+\label{eq:elpd_psis_loo}
+\end{equation}
+We can use the value of the Pareto $\hat k$ from PSIS to determine if the 
+PSIS-LOO ELPD estimator of $p(\vec\theta\lvert\vec y_{-i})$ for the $i$-th data 
+point potentially do a full re-fit of the model only for the most problematic 
+points. \citet{arXiv:1004.2316} showed that LOO is a consistent and unbiased estimator of 
+the ELPD.
+
+Equation~\ref{eq:elpd_psis_loo} thus enables us to estimate the expected log pointwise 
+predictive density for unseen data using only posterior samples without having 
+to re-fit the model (unless Paretko $\hat k$ is large for certain data points). 
+This equation forms the basis for the widely popular \textsf{loo} statistical package 
+\citep{loo_package}.
+$\mathrm{ELPD}_\mathrm{PSIS-LOO}$ is closely related to the WAIC information 
+criterion \citep{arXiv:1004.2316} which is an alternative approach to estimating 
+ELPD. LOO and WAIC are asymptotically equivalent but LOO is a more robust and less 
+biased estimator with finite data \citep{arXiv:1507.04544}.
+Other information criteria such as AIC, BIC and DIC are inferior to LOO and WAIC 
+estimates of ELPD because they use point estimates (maximum likelihood) to assign 
+a score to a model and they attempt to quantify model complexity in an ad-hoc 
+way (by including a term proportional to the number of parameters in the model). 
+LOO-CV is also less sensitive to the choice of priors and often more reliable 
+for model comparison than Bayes factors \citep{10.1214/17-BA1091,arXiv:2003.04026}.
+
+What if our model includes a Gaussian Process and the likelihood is not 
+factorizable? In that case we cannot use Equation~\ref{eq:elpd_psis_loo} but 
+\citet{arXiv:1810.10559} showed how to evaluate 
+$\ln p(\vec y_i\lvert\vec y_{-i},\vec\theta,\mathcal{M})$ if the likelihood is 
+multivariate normal with a dense covariance or a Student-t likelihood so it is 
+still possible to use LOO-CV cheaply. 
+If we are dealing with time-series data and we are only interested in predicting 
+the next few points conditional on the previous data points then it may not 
+much sense to use LOO-CV because we are using future data points when evaluating 
+predictions on past data points. \textsl{leave-future-out cross validation} 
+(LFO-CV) \citep{arXiv:1902.06281} is a generalization of LOO which takes into 
+account the time-ordering of the data points in a time-series.
+
+For a more information about Cross Validation and its approximations see the excellent
+\href{https://avehtari.github.io/modelselection/CV-FAQ.html}{Cross-validation FAQ}
+by Aki Vehtari.
+In Chapter~\ref{ch:microlensing} I will demonstrate the utility of these methods by applying 
+them to very practical problems in the analysis of microlensing light curves.
+
+\subsection{What about machine learning?}
 %% mention that even if we have a model focused on inference, it might be best if we judge 
@@
 %
-\subsection{Sampling and optimization methods}
-\subsubsection{The Laplace approximation}
-\subsubsection{Variational inference}
-\subsubsection{Importance sampling}
-\subsubsection{Rejection sampling}
-\subsubsection{Markov Chain Monte Carlo}
-\subsubsection{Hamiltonian Monte Carlo}
-%\subsubsection{Other }
-%
-%\subsection{Model validation and comparison}
-%\label{ssec:model_comparisonGaussian linear }
-%
-``Furthermore, in the future, a blind search for planetary signals will not
-turn up all potential competing models unless those models
-are specifically considered.''\citep{2017AJ....153..129J}.
-\citet{2021AJ....162...59R}
-\citet{2020AJ....160...17H}
-
-%\subsubsection{Cross validation}
-%% discuss vanilla k-fold CV, stress out just how important and central CV is to science
-%\subsubsubsection{Pareto smoothed importance sampling LOO}
-%\subsubsection{Model comparison}
-%% don't do it!
-%
+\section{Automatic differentiation}
 %\section{Putting it all together -- \emph{probabilistic programming}}
@@
 % CHAPTER 3: Modeling microlensing events 
+%% PLOT POINTWISE LOO 
 \chapter{Modeling microlensing events}
@@
 
-\end{document}ii
+\end{document}

### 2022-06-21 src/figures/typical_sets.py

@@ -1,0 +1,32 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+
+def draw_samples(ndim, nsamples=5000):
+    mean = np.zeros(ndim)
+    cov = np.diag(np.ones(ndim))
+    return np.random.multivariate_normal(mean, cov, nsamples)
+
+
+distance1 = np.linalg.norm(draw_samples(2), axis=1)
+distance2 = np.linalg.norm(draw_samples(10), axis=1)
+distance3 = np.linalg.norm(draw_samples(50), axis=1)
+
+
+fig, ax = plt.subplots(1, 3, figsize=(12, 2), sharey=True)
+ax[0].hist(distance1, bins=25, density=True, histtype="step", color="k", lw=2.0)
+ax[1].hist(distance2, bins=25, density=True, histtype="step", color="k", lw=2.0)
+ax[2].hist(distance3, bins=25, density=True, histtype="step", color="k", lw=2.0)
+
+ax[0].set_title("D = 2")
+ax[1].set_title("D = 10")
+ax[2].set_title("D = 50")
+
+for a in ax.ravel():
+    a.set_xlim(0.0, 10.0)
+    a.set_xticks(np.arange(11, step=2))
+    a.set_xlabel("Distance from MAP")
+
+# Save as pdf
+fig.savefig("typical_sets.pdf", bbox_inches="tight")
+

### 2022-06-22 src/ms.tex
Finished Chapter 2 draft

@@
 \subsection{What about machine learning?}
-%% mention that even if we have a model focused on inference, it might be best if we judge 
-%% its quality through its predictive performance
-%
+How is \textsl{machine learning} (ML) different from the topics we have discussed so far? 
+I would say the key difference is that ML and particularly \textsl{deep learning} (DL) 
+\emph{ML is much more focused on predicting future data} conditional on observing 
+past data  than on estimating the parameters of some model, which is the subject of 
+statistical inference. Classic example is that of neural networks which 
+today routinely consist of millions, even billions of parameters (weights) but 
+the  values of these weights are not important, only the prediction error 
+(on unseen data) is what matters. For this reason optimization methods such as 
+\textsl{stochastic gradient descent} (SGD) which are computationally much 
+cheaper to implement are dominant over sampling methods such as MCMC.
+
+Despite the fact that the deep learning revolution started in 2012, 
+deep learning has not had as much impact  in the physical sciences as many have 
+expected  despite the fact that  there are many papers on the topic.
+DL methods tend to be black boxes which are very efficient at 
+the task they were trained to do, but their internal structure is usually 
+illegible. This is not as much of an issue in industry applications as it is 
+in the sciences where we care deeply about the precise values of very few 
+parameters.
+Nevertheless, all of this is changing rapidly and there is  a lot of work 
+going on the boundary of physics and machine learning. One example is the field 
+of \textsl{likelihood free inference} (LFI) which aims to emulate expensive 
+physics simulators (such as massive hydrodynamics simulations) with neural 
+networks so that one can do inference with much faster emulators 
+\citep[see for example the review paper by]{2020PNAS..11730055C}. There has 
+also been a lot of work on neural networks which respect certain physical 
+symmetries (see for example Hamiltonian and Lagrangian neural 
+networks\footnote{Or \textsl{geometric deep learning} more generally.})
+and even automated discovery of succinct mathematical descriptions of 
+the dynamical behavior of a physical system through observations of its 
+behaviour \citep{arXiv:2006.11287}.
+
+However, the biggest significance of the atmospheric rise of machine learning 
+in my opinion is that the computational infrastructure developed for training 
+large neural networks can also be used for scientific computing. In particular,
+two of the key technologies responsible for the deep learning  revolution
+are \textsl{Graphical Processing Units} (GPUs) and 
+\textsl{automatic differentiation} (AD), are also very useful 
+for building scientific models.
+
 \section{Automatic differentiation}
-%\section{Putting it all together -- \emph{probabilistic programming}}
-%\label{sec:programming}
-%\subsection{The success of machine learning}
-%% that quote from Yann LeCun on differentiable code 
-%\subsection{Automatic differentiation}
-%\subsubsection{Forward mode autodiff}
-%\subsubsection{Reverse mode autodiff or backpropagation}
-%\subsubsection{Jacobian vector products}
-%\subsection{JAX}
-%\subsection{Numpyro}
-
-% CHAPTER 3: Modeling microlensing events 
+\label{sec:autodiff}
+\textsl{Automatic differentiation} (AD or autodiff for short)\citep{10.1145/355586.364791} is an old idea 
+that has revolutionized the fields of machine learning, AI and optimization in the 
+recent decade. 
+The key idea behind automatic differentiation 
+is that any model implemented in computer code (such as \textsf{C++} or 
+\textsf{Python}) is fundamentally a composition 
+of simple functions such as additions, multiplications and trig functions 
+which can be easily differentiated and these gradients evaluated at each step 
+of the computation can be combined using the chain rule from calculus to 
+yield the exact (up to machine precision) derivative of the complete model.
+Specialized automatic differentiation libraries build a representation of the 
+code as graph, evaluate the derivatives at each node and propagate them through 
+the entire graph.
+
+A variant of automatic differentiation called reverse mode automatic differentiation 
+is widely used in machine learning (where it goes under the name of 
+\emph{backpropagation}) to differentiate scalar functions with respect to 
+millions if not billions of parameters. 
+Backpropagation and autodiff in general tells us how to change the inputs of a 
+model (even if there are billions of inputs) to obtain a particular output.
+Autodiff has been used to differentiate through all sorts of things, including 
+complex simulations of physical systems such as molecular dynamics simulations,
+\citep{arXiv:1912.04232}, cosmological simulations \citep{2016MNRAS.463.2273F},
+ ODE and PDE solvers etc. It has even been applied to differentiating code at 
+ the level of a compiler using the LLVM compiler infrastructure \citep{arXiv:2010.01709}.
+ Because the wide variatey of autodiff applications
+the famous AI researcher Yann LeCun recently popularized the term 
+\textsl{differentiable programming} to refer to any computer code which is 
+automatically differentiable.
+
+The reason why I am discussing this topic in the first place is that implementing AD 
+for astrophysical models enables the use  of advanced MCMC methods such as 
+Hamiltonian Monte Carlo and much faster and more stable numerical optimization  
+(for obtaining the MLE estimate for example).
+
+\subsection{Three ways of differentiating computer programs}
+There are three ways of computing derivatives:
+\begin{itemize}
+    \item \textbf{Symbolic differentiation}: computing \emph{exact} derivatives of 
+    functions using either a pen and paper or a computer algebra system such as 
+    \textsf{Wolfram Mathematica} or \textsf{SymPy} \citep{10.7717/peerj-cs.103}.
+    \item \textbf{Numerical differentiation}: computing \emph{approximate} derivatives 
+    using \textsl{finite difference}  methods.
+    \item \textbf{Automatic differentiation}: computing \emph{exact} derivatives of entire 
+    computer programs containing things like conditionals, loops, etc.
+\end{itemize}
+Symbolic differentiation is familiar to everyone from basic calculus. Using either 
+pen and paper  or Computer Algebra Systems (CAS) we can compute exact 
+derivatives of functions. 
+If our model is expressible as a straightforward closed-form function it is 
+often not to difficult to compute the derivatives of interest and indeed this 
+is what people have been doing for decades to improve the performance of their
+models\footnote{Before the advent of easy to use open-source AD libraries 
+the contribution of many papers in applied AI would be to compute the analytic 
+derivatives for some specific model.}
+
+The challenge with symbolic differentiation is that it is not always possible 
+to easily compute the gradients, especially if the model is implemented in terms 
+of a code which contains control flow constructs such as loops and conditionals, 
+or iterative solvers. In those cases gradients of the output of the model 
+are with respect to the input are usually computed using finite-difference 
+methods. In finite-difference methods the derivative of some function $f(x)$
+is defined as 
+\begin{equation}
+    f^\prime(x)=\lim _{h \rightarrow 0} \frac{f(x+h / 2)-f(x-h / 2)}{h}
+\end{equation}
+for some small step size $h$. Numerical gradients are easy to compute even with 
+complicated models but they have two key drawbacks. First, obtaining a 
+good approximation of the derivatives requires a small step size $h$ which is 
+can lead to errors due to numerical cancellations.
+The second is that if we want to evaluate the Jacobian of a function with lots 
+of input parameters we would need to compute a finite difference gradient 
+for each partial derivative which is prohibitively expensive in high dimensions.
+
+Automatic differentiation solves these problems. If AD is properly implemented 
+evaluating the exact gradient is only a factor of a few slower than evaluating 
+the function itself, even for models with millions of parameters.
+The drawback is that it is usually
+necessary to express our model in terms of constructs which are implemented in 
+specialized AD libraries. It also requires a slightly different programming style
+than what most people are used to. Fortunately there are now lots very flexible 
+and easy to use high performance AD libraries.
+
+\subsection{Forward and reverse-mode autodiff and the chain rule}
+What follows is mostly a summary of the chapter on autodiff 
+from \citet{murphy_book_2023}. First, we define the notation
+\begin{equation}
+\partial x_1 f\left(x_{1}, x_{2}, \ldots, x_{n}\right)
+\end{equation} 
+which denotes the partial derivative of function 
+$f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}$ with respect to argument $x_1$ 
+evaluated at the point $(x_{1}, x_{2}, \ldots, x_{n})$.
+We define the derivative operator $\partial$ and denote the derivative of a 
+differentiable function $f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^m$ as 
+\begin{equation}
+    \partial f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^{m\times n}
+    \label{eq:general_derivative}
+\end{equation}
+The function $\partial f$ is a \textsl{linear map} which maps a point 
+$\vec x\in \mathbb{R}^n$ to the \textsl{Jacobian} of all partial derivatives 
+evaluated at $\vec x$.  $\partial f$ evaluated at point $\vec x$,
+$\partial f (\vec x)$ is the \textsl{linearization} of function $f$ at point $\vec x$.
+The Jacobian can be be represent as a matrix $\vec J$ whose 
+elements $J_{ij}$ are given by 
+\begin{equation}
+J_{i j}=\frac{\partial f_{i}}{\partial x_{j}}
+\end{equation}
+  To make this slighly less abstract, consider the case when
+$m=n=1$, the function takes a scalar and returns a scalar, then 
+Equation~\ref{eq:general_derivative} is just the familiar derivative $f'(x)$:
+\begin{equation}
+    f^\prime\,:\,\mathbb{R} \rightarrow \mathbb{R}
+\end{equation}
+If the $f$ takes a vector and returns a scalar ($m=1$) then derivative operator is 
+the \textsl{gradient} $\nabla f$:
+\begin{equation}
+    \partial f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^{1\times n}
+\end{equation}
+
+We now introduce two key ingredients for an automatic differentiation framework 
+which enable us to efficiently compute the Jacobian for functions of arbitrary 
+complexity.
+
+\subsubsection{Jacobian-vector products (JVPs)}
+Now consider consider a perturbation $\vec v\in\mathbb{R}^n$ to the function input 
+$\vec x$. The  mapping 
+\begin{equation}
+(\vec{x}, \vec{v}) \mapsto \partial f(\vec{x})\,\vec v
+\label{eq:jvp}
+\end{equation}
+is called the \textsl{Jacobian-vector product} (JVP). For example, if $f=\cos(x)$ 
+for a scalar $x$ and $v$ is also a scalar, the JVP would be 
+\begin{equation}
+(x, v) \mapsto \partial \cos (x) v=-v \sin (x)\hquad .
+\end{equation}
+Why is this useful? With a function $f\,:\,\mathbb{R}^n\rightarrow\mathbb{R}^m$
+and a \textsl{one-hot} vector $\vec v=(1, 0, \ldots, 0)^\intercal$ we can use  the JVP to 
+evaluate the Jacobian matrix one column at a time:
+\begin{equation}
+    \partial f(\vec x)\,\vec v = 
+    \begin{pmatrix}
+\frac{\partial f_{1}}{\partial x_{1}} & \cdots & \frac{\partial f_{1}}{\partial x_{n}} \\
+\vdots & \ddots & \vdots \\
+\frac{\partial f_{m}}{\partial x_{1}} & \cdots & \frac{\partial f_{m}}{\partial x_{n}}
+    \end{pmatrix}
+    \begin{pmatrix}
+        1\\
+        0\\
+        \vdots\\
+        0
+    \end{pmatrix}
+    = \begin{pmatrix}
+        \frac{\partial f_{1}}{\partial x_{1}}\\
+        \frac{\partial f_{2}}{\partial x_{1}}\\
+        \vdots\\
+        \frac{\partial f_{m}}{\partial x_{1}}
+    \end{pmatrix}
+\end{equation}
+Evaluating one column of the Jacobian in this way costs about the same as one function evaluation
+so this is efficient for fucntions with ``tall Jacobians'' (functions with many outputs and few
+inputs) but innefficient for functions with ``wide Jacobians''.
+
+\subsubsection{Vector-Jacobian products (VJPs)}
+Similarly, if we consider 
+a perturbation to $\vec u\in\mathbb{R}^m$ to the output of the $f$, we obtain
+the \textsl{vector-Jacobian product} (VJP):
+\begin{equation}
+(\vec{x}, \vec{u}) \mapsto \partial f(\vec{x})^{\intercal}\,\vec u
+\label{eq:vjp}
+\end{equation}
+Whereas JVPs are useful for computing the Jacobian matrix one column at a time, we can use VJPs
+to compute the Jacobian one row at a time:
+\begin{equation}
+    \vec v^\intercal\,\partial f(\vec x)= 
+    (1\,0\, \cdots \, 0)\,
+    \begin{pmatrix}
+\frac{\partial f_{1}}{\partial x_{1}} & \cdots & \frac{\partial f_{1}}{\partial x_{n}} \\
+\vdots & \ddots & \vdots \\
+\frac{\partial f_{m}}{\partial x_{1}} & \cdots & \frac{\partial f_{m}}{\partial x_{n}}
+    \end{pmatrix}
+        = \begin{pmatrix}
+        \frac{\partial f_{1}}{\partial x_{1}}\\
+        \frac{\partial f_{1}}{\partial x_{2}}\\
+        \vdots\\
+        \frac{\partial f_{1}}{\partial x_{n}}
+    \end{pmatrix}
+\end{equation}
+For a scalar function $f\,:\,\mathbb{R}^n\rightarrow\mathbb{R}$, the VJP is simply the gradient 
+$\nabla f$:
+\begin{equation}
+    \vec v^\intercal\,\partial f(\vec x)\equiv \nabla f(\vec x)=
+    \left(1\right)\,\left(  \frac{\partial f}{\partial x_{1}} \,\cdots\,\frac{\partial f}{\partial x_{n}}   \right)
+    =\begin{pmatrix}
+        \frac{\partial f}{\partial x_{1}}\\
+        \vdots\\
+        \frac{\partial f}{\partial x_{n}}
+    \end{pmatrix}
+\end{equation}
+
+\subsubsection{Chain rule}
+Let's now consider a function $f\,:\, \mathbb{R}^n\rightarrow\mathbb{R}^m$ which is a composition
+three functions $a$, $b$ and $c$:
+\begin{equation}
+f=c \circ b \circ a
+\end{equation}
+The \textsl{chain rule} from calculus says that we can differentiate $f$ if we know how to differentiate 
+each of its constituent parts:
+\begin{align}
+\partial f(\vec{x})&=\partial c(b(a(\vec{x}))) \circ \partial b(a(\vec{x})) \circ \partial a(\vec{x})\\
+&=\partial c(b(a(\vec{x})))[\partial b(a(\vec{x}))[\partial a(\vec{x})]]
+\label{eq:chain_rule}
+\end{align}
+The idea behind AD is to decompose complicated functions $f$ into constituent parts (sometimes 
+called \textsl{primitive functions}) which we know how to differentiate  and then use the chain 
+rule to obtain the derivative of $f$.
+Equation~\ref{eq:chain_rule} consists of a series of matrix multiplications. Although the order
+in which we multiply those matrices does not matter because of the associativity of matrix 
+multiplication, it matters a great deal when we are implementing AD on a computer.
+Let's say that we want to compute the JVP $\partial f(\vec{x})\,\vec v$ of this composite 
+function:
+\begin{equation}
+ \partial f(\vec{x})\,\vec v = \partial c(b(a(\vec{x})))[\partial b(a(\vec{x}))[\partial a(\vec{x})\vec v]]
+ \label{eq:jvp_chain_rule}
+\end{equation}
+Right-to-left evaluation of the expression on the right-hand side of 
+Equation~\ref{eq:jvp_chain_rule} is known  as \textsl{forward-mode automatic differentiation}.
+In forward-mode AD we evaluate \textsl{primal terms} $\vec{x}, a(\vec{x}), b(a(\vec{x}))$ 
+alongside the \textsl{tangent terms} 
+$\partial a(\vec{x}), \partial b(a(\vec{x})), \partial c(b(a(\vec{x})))$.
+
+If instead we are interested in evaluating the VJP the order of operations is reversed:
+\begin{equation}
+\partial f(\vec{x})^{\intercal}\,\vec u=\partial a(\vec{x})^{\intercal}\left[\partial b(a(\vec{x}))^\intercal\left[\partial c(b(a(\vec{x})))^{\intercal}\vec u\right]\right]
+\label{eq:vjp_chain_rule}
+\end{equation}
+To evaluate  Equation~\label{eq:vjp_chain_rule} we first have to traverse the whole chain 
+computing the primals $\vec{x}, a(\vec{x}), b(a(\vec{x}))$ as we go and then we proceed in reverse order 
+computing the tangents 
+$\partial c(b(a(\vec{x})))^{\intercal}, \partial b(a(\vec{x}))^{\intercal}, \partial a(\vec{x})^{\intercal}$.
+This process is known as \textsl{reverse-mode automatic differentiation} or 
+\textsl{backpropagation} (since we are propagating the gradients backwards from the function output).
+
+Using JVPs and VJPs is much more memory efficient than computing the Jacobian matrix directly 
+because all intermediate results computed when applying the chain rule are 1D vectors that 
+don't take up a lot of memory. The key difference between reverse mode and  forward mode is that 
+in the case of forward-mode AD we can just evaluate the JVPs as we evaluate the function itself 
+whereas in reverse-mode AD we need to store the primal terms for the entire chain before 
+backpropagating so reverse-mode AD is more memory intensive.
+Reverse-mode is usually far more popular than forward-mode in optimization and machine learning 
+because the function $f$ is some cost function with  a high-dimensional input and a scalar 
+output so we only need one VJP to obtain the entire Jacobian matrix. There is also mixed-mode 
+automatic differentiation which combines forward-mode and reverse-mode to optimize memory use.
+
+In the general case functions can be more complex than a simple chain of compositions and in AD 
+frameworks they are represented as \textsl{Directed Acyclic Graphs} (DAGs), sometimes also 
+called \textsl{computation graphs} or \textsl{circuits}. We point the reader to the Section 
+6.2 in \citep{murphy_book_2023} for details on how to generalize AD to graphs and also higher 
+order AD which is also possible using the same formalism.
+
+\subsection{AD libraries}
+The big challenge in automatic differentiation is creating a library wich can knows how to compute 
+JVPs and VJPs for each primitive operation (for example, multiplication, division, 
+matrix multiplication, array slicing, etc.).
+Because the primitive operations need to be coded 
+manually, expressing a given computation (say an astrophysics code written in \textsf{C++}) in 
+an AD library often requires some effort and it is a bit like programming in a different 
+language. 
+Popular AD libraries are generally developed by large groups of researchers and with the backing 
+of major tech companies.
+
+There are two main kinds of AD libraries:
+\begin{itemize}
+    \item \textbf{Static graph libraries} 
+    \item \textbf{Dynamic graph libraries}
+\end{itemize}
+The static graph libraries first build the entire computation graph for a given program before 
+the graph is evaluated. Examples include \textsf{Theano}\footnote{Theano is no longer 
+maintained but it has recently been ressurected and turned into \textsf{aesara} 
+(\url{https://github.com/aesara-devs/aesara})} \citep{arXiv:1605.02688}, 
+\textsf{TensorFlow}\footnote{\url{https://www.tensorflow.org/}} \citep{arXiv:1603.04467}, 
+and the \textsf{Stan Math Library} \citep{arXiv:1509.07164}.
+Dynamic graph libraries build the graph as the program is is being executed. Examples are 
+\textsf{Tensorflow 2}, \textsf{JAX}\footnote{\url{https://jax.readthedocs.io/en/latest/}.} 
+\citep{jax2018github} and \textsf{PyTorch}\footnote{\url{https://pytorch.org/}.} \citep{pytorch}.
+These libraries have many features in addition to autodiff, most notably support for
+for GPUs. They are written in Python but the primitive operations are usually 
+expressed in a lower level language like \textsf{C++} and \textsf{CUDA C++} 
+for GPU support.
+
+The most recent and most research oriented of these libraries is Google's \textsf{JAX} 
+library.
+\textsf{JAX} is written in \textsf{Python} and it offers nearly identical functionallity 
+to the widely popular \textsf{NumPy} \textsf{Python} library for manipulating 
+multi-dimensional arrays. \textsf{JAX}takes Python code as an input, converts it 
+to an intermediate language (IR) representation called \textsf{jaxpr} which is 
+used to perform autodiff, and finally the resulting code for evaluating functions 
+and their gradients is \textsl{Just-In-Time} (JIT) compiled to a lower level 
+assembly like language called \textsf{HLO-IR} using the 
+\textsf{XLA}\footnote{\url{https://www.tensorflow.org/xla}.} compiler in such a way 
+that the code can be executed on different hardware architecutres such  as 
+CPUs, GPUs and TPUs (Tensor Processing Units).
+In practice this means that one can write \textsf{Python} code using a \textsf{NumPy} 
+like library, easily compute gradients of the code through autodiff and get 
+the performance of a compiled language like $C++$ through on-the-fly JIT 
+compilation. \textsf{JAX} natively supports \textsf{Python} language constructs such 
+as loops, conditionals and data structures such as tuples and dictionaries.
+Almost everything you could write in \textsf{Python} is differentiable with 
+\textsf{JAX} although there is no guarantee that the gradients are meaningful 
+if the function represented by the code is not itself differentiable.
+Implementing a real-world model in an AD framework often requires making individual 
+functions ``smoother'' so they are more easily differentiable (see 
+Chapter~\ref{ch:microlensing2} for more on this point).
+
+Many of the statistical algorithms I have mentioned  in 
+Section~\ref{ssec:sampling_methods} are now implemented in \textsf{JAX} and 
+\textsf{JAX} has also been used within astronomy. \citet{2022ApJS..258...31K} built 
+an autodifferentiable model for exoplanet spectrum modeling, 
+the \textsf{exoplanet} code \citet{2021JOSS....6.3285F} for modeling transit light 
+curves and radial velocity time series uses \textsf{JAX} as a backend, and 
+\citet{2022arXiv220207663G} used \textsf{JAX} to build a differentiable model 
+for strong lensing  which produces samples from a Bayesian posterior by  first
+exploring the posterior using an optimizer, running variational inference to 
+obtain an estimate of the posterior covariance followed by Hamiltonian Monte Carlo
+to obtain the final samples. Importantly, \textsf{JAX} makes it relatively 
+straightforward to implement custom primitives. One of the contributions of this 
+thesis is an implementation of a custom JAX primitive for computing the roots 
+of complex polynomials (see Chapter~\ref{ch:microlensing2} for more details).
+The process of building custom primitives involves specifying the JVP rule 
+(Equation~\ref{eq:jvp}). Interestingly, to support reverse-mode AD we it is 
+not necessary to also specify the VJP rule (Equation~\ref{eq:vjp}) because 
+\textsf{JAX} is able to deduce the transpose of the JVP automatically\footnote{
+    Essentially, this is possible because the implementation of the JVP rule 
+    in \textsf{Python} consists of several \textsf{JAX} primitives such as addition
+    and multiplication and \textsf{JAX} already knows how to transpose those 
+    primitives.} (see \citet{arXiv:2105.09469} for details).
+
+\subsection{Probability + programming + autodiff = probabilistic programming}
+The final topic in this chapter nicely ties up everything we have discussed so far.
+The term \textsl{probabilistic programming} refers to the use of software libraries 
+which  provide a high-level API for constructing statistical models (usually in 
+the Bayesian paradigm). Specifically, 
+the key components of modern probabilistic programming languages (PPLs) include a backend
+autodiff framework, methods for initializing \emph{random variables} with 
+associated pdf or pmf distribution functions and evaluation the (log) probability, 
+and finally an one or several optimizers and samplers. The idea behind these 
+frameworks is to provide an easy to use interface for constructing probabilistic 
+models and to abstract away the inference as much as possible.
+Why is this useful for scientists? Because these frameworks simplify the process  
+of building probabilistic  models and because they include well-tested and high
+performance implementations of state-of-the-art inference methods such as 
+Hamiltonian Monte Carlo.
+
+The most popular such library is \textsf{Stan} \citep{2017JSS....76....1C}. 
+Stan is built in \textsf{C++}, it uses the \textsf{Stan Math Library} for automatic 
+differentiation, it uses a variant of the NUTS HMC sampler and also has some 
+support for Variational Inference. \textsf{Stan} is oriented primarily towards 
+applications in the social sciences and it is not as flexible and easily extensible 
+as some other PPLs. \textsf{PyMC} is another PPL which is very similar to 
+\textsf{Stan} except that it is implemented in \textsf{Python} and it uses 
+the \textsf{aesara}\footnote{\url{https://github.com/aesara-devs/aesara}.} library 
+for AD (it used to be called \textsf{PyMC3} and use \textsf{Theano} for AD) and it 
+also supports \textsf{JAX}. 
+\textsf{PyMC} is somewhat more popular in the physical sciences and easier to 
+extend  with custom functionality. The \textsf{exoplanet} code 
+\citep{2021JOSS....6.3285F} is a good example of an astrophysics code which 
+interfaces with \textsf{PyMC}.  There is also 
+\textsf{numpyro}\footnote{\url{https://num.pyro.ai/}.}
+\citep{arXiv:1912.11554} which is also built in \textsf{Python} and uses 
+\textsf{JAX} for AD and \textsf{Tensorflow Probability} \citep{arXiv:1711.10604} 
+which supports bot \textsf{TensorFlow} and \textsf{JAX} for AD. \textsf{PyMC}, \textsf{numpyro}
+and \textsf{Tensorflow Probability} all use a variant of NUTS similar to the one 
+implemented in \textsf{Stan} as the core algorithm for inference but also 
+include support for Variational Inference and additional samplers. 
+I use PPLs (mostly \textsf{numpyro}) extensively throughout the rest of this thesis.
+
+% CHAPTER 3: Modeling single lens microlensing events 
 %% PLOT POINTWISE LOO 
-\chapter{Modeling microlensing events}
+\chapter{Modeling single lens microlensing events}
 \label{ch:microlensing}
-\section{Single lens events}
-\section{Modeling single lens events}
-\section{Numerical solutions to the lens equation}
+% single lens events are important because...
+\section{A differentiable single lens extended source model}
+\section{Practical solutions for sampling multi-modal posteriors}
+\section{Hierarchical inference}
+
+% CHAPTER 4: Modeling binary and triple microlensing events
+\chapter{Modeling binary and triple lens microlensing events}
+\label{ch:microlensing2}
+\section{A differentiable complex polynomial root solver}
 \section{\textsf{caustics} -- computing the magnification of an extended limb-darkened source}
@@
 
-% CHAPTER 4: Mapping the surface of Io 
+% CHAPTER 5: Mapping the surface of Io 
 \chapter{Mapping the surface of Io}
@@
 
-% CHAPTER 5: Mapping the surfaces of exoplanets
+% CHAPTER 6: Mapping the surfaces of exoplanets
 \chapter{Mapping the surfaces of exoplanets}
@@
 
-% CHAPTER 6: Conclusion
+% CHAPTER 7: Conclusion
 \chapter{Conclusion}

### 2022-07-23 src/ms.tex
Add section on polynomial root solver

--- 2022-06-07 src/ms.tex
@@
 \newcommand{\ud}{\,\mathrm{d}}
+\newcommand{\ssf}[1]{\textsf{#1}}
 \renewcommand{\vec}[1]{\boldsymbol{\mathbf{#1}}}
-
-\hypersetup{
-    colorlinks=true,
-    allcolors=cyan,
-}
+\newcommand{\hquad}{~~}
 
@@
 specifically, I have developed a an open-source software package
-\textsf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which
+\ssf{caustics}\footnote{\url{https://github.com/fbartolic/caustics}} which
 enables fast and accurate computation of binary lens and triple lens
@@
 \begin{itemize}
-    \item I wrote an open-source Python package \textsf{caustics} which enables fast and
+    \item I wrote an open-source Python package \ssf{caustics} which enables fast and
           accurate computation of binary and triple lens microlensing light curves using
@@
 phase curves and eclipse light curves. Most notably,
-\citet{2019AJ....157...64L} introduced the \textsf{starry} code which enables
+\citet{2019AJ....157...64L} introduced the \ssf{starry} code which enables
 analytic computation of phase curves and occultation light curves for bodies
@@
 collaboration with other researchers from the planetary science and exoplanet
-communities on using \textsf{starry} to map the surface of Io and investigate
-the prospects for detecting fine spatial structure in the atmospheres of Hot
+communities on using \ssf{starry} to map the surface of Io and investigate the
+prospects for detecting fine spatial structure in the atmospheres of Hot
 Jupiters using JWST phase curves and eclipse light curves. My contributions to
@@
     \item I have investigated the problem of eclipse mapping of exoplanets using the
-          \textsf{starry} framework, focusing in particular on the possibility of
-          detecting planetary scale storms on Hot Jupiters using JWST. I found that it
-          extremely difficult to infer maps of higher resolution than a dipole order but
-          that should still be sufficient for detecting large scale weather and climate
-          change in the atmospheres of these planets.
+          \ssf{starry} framework, focusing in particular on the possibility of detecting
+          planetary scale storms on Hot Jupiters using JWST. I found that it extremely
+          difficult to infer maps of higher resolution than a dipole order but that
+          should still be sufficient for detecting large scale weather and climate change
+          in the atmospheres of these planets.
 \end{itemize}
@@
 %generally, and probabilistic programming which ties everything together.
-%Chapter~\ref{ch:microlensing} is the most extensive chapter in the thesis
+%Chapter~\ref{ch:microlensing_modeling} is the most extensive chapter in the thesis
 %covering all of my work on microlensing. I start by discussing the case of
@@
 %the magnification of a limb-darkened source star and its implementation in
-%\textsf{caustics}.
+%\ssf{caustics}.
 
@@
 \begin{equation}
-    x= \frac{\theta}{D_L\theta_E}, \quad y=\frac{\beta}{D_S\theta_E}
+    z\equiv \frac{\theta}{D_L\theta_E}, \quad w\equiv\frac{\beta}{D_S\theta_E}
 \end{equation}
@@
 \begin{equation}
-    y= x- \frac{ 1}{x}
+    z= w- \frac{ 1}{w}
     \label{eq:lens_equation_dimensionless}
@@
 Equation~\ref{eq:lens_equation_dimensionless} is a nonlinear mapping between the
-source plane $y$ and the image plane $x$.
-If we solve for $x$, we obtain two solutions for the position of the images, given by
-
-\begin{equation}
-    x_\pm= \frac{1}{2} \left( y\pm\sqrt{y^2 +4}\right)
+source plane $w$ and the image plane $z$.
+If we solve for $z$, we obtain two solutions for the position of the images, given by
+
+\begin{equation}
+    z_\pm= \frac{1}{2} \left( w\pm\sqrt{w^2 +4}\right)
     \label{eq:images_location}
 \end{equation}
-For non zero $y$ the so called \emph{minor image} is always located
-within the Einstein radius ($\lvert x_-\rvert<1$) while the \emph{major image}
-is located outside of it ($\lvert x_+\rvert>1$).
+For non zero $w$ the so called \emph{minor image} is always located
+within the Einstein radius ($\lvert z_-\rvert<1$) while the \emph{major image}
+is located outside of it ($\lvert z_+\rvert>1$).
 
@@
 where $0$ denotes the unlensed solid angle.
-Assuming an infinitesimal source size at angular location $\mathbf y =(y_1,y_2)$ subtending an angle
-$\Delta (\omega)_0$, and an image of the source at location $\mathbf x=(x_1,x_2)$ subtending an angle
+Assuming an infinitesimal source size at angular location
+$\mathbf w \equiv (w_1,w_2)$ subtending an angle
+$\Delta (\omega)_0$, and an image of the source at location
+$\mathbf z\equiv(z_1,z_2)$ subtending an angle
 $\Delta \omega$; the ratio between the solid angle of the source and the image
 is given by the determinant of the Jacobian matrix of the
-lens mapping $\mathbf{x}\rightarrow\mathbf{y}$, evaluated at the location of the images
+lens mapping $\mathbf{z}\rightarrow\mathbf{w}$, evaluated at the location of
+the images
 \begin{equation}
     \frac{(\Delta\omega)_0}{\Delta\omega} =\left\lvert\textrm{det}
-    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert
-\end{equation}
-The magnification $A$ is then given by the inverse Jacobian determinant of the lens mapping:
+    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert
+\end{equation}
+The magnification $A$ is then given by the inverse Jacobian determinant
+of the lens mapping:
 \begin{equation}
     A= \left\lvert\det
-    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert^{-1} \label{eq:magnification_general}
+    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert^{-1} \label{eq:magnification_general}
 \end{equation}
@@
     \left\lvert\det
-    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert=0
+    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert=0
 \end{equation}
@@
 The critical curves in the lens plane can be mapped to the source plane using
-the lens equation $\mathbf{x}\rightarrow \mathbf{y}$ and these curves are
+the lens equation $\mathbf{z}\rightarrow \mathbf{w}$ and these curves are
 are called \emph{caustic curves}.
@@
 \begin{align}
-    y_1= & x_1- \frac{x_1}{x_1^2+x_2^2} \\
-    y_2= & x_2- \frac{x_2}{x_1^2+x_2^2}
+    w_1= & z_1- \frac{z_1}{z_1^2+z_2^2} \\
+    w_2= & z_2- \frac{z_2}{z_1^2+z_2^2}
 \end{align}
-Introducing polar coordinates $x_1=v\cos\phi,\;x_2=v\sin\phi$, we have
+Introducing polar coordinates $z_1=v\cos\phi,\;z_2=v\sin\phi$, we have
 \begin{equation}
     \left\lvert\textrm{det}
-    \frac{\partial \mathbf y}{\partial \mathbf x} \right\rvert =1- \frac{1}{v^4}\quad,
-\end{equation}
-By substituting in the locations of the source images Equation~\ref{eq:images_location} into the
+    \frac{\partial \mathbf w}{\partial \mathbf z} \right\rvert =1- \frac{1}{v^4}\quad,
+\end{equation}
+By substituting in the locations of the source images
+Equation~\ref{eq:images_location} into the
 above expression, we obtain the magnification of the source for both images
 \begin{equation}
-    A_\pm=\left(1- \frac{1}{x^4_\pm} \right)^{-1}
-\end{equation}
+    A_\pm = \frac{1}{2}\left(\frac{u^{2}+2}{u \sqrt{u^{2}+4}} \pm 1\right)
+    \label{eq:magnification_per_image}
+\end{equation}
+where $u = \sqrt{w_1^2+w_2^2}$ is the the magnitude of the position
+vector of the source star.
 The total magnification is then the sum of the two magnifications, and is given by
@@
 \end{equation}
-where $u = \sqrt{y_1^2+y_2^2}$ is the the magnitude of the position
-vector of the source star.
 The critical curve of the point lens is the Einstein
-ring corresponding to $v=\sqrt{x_1^2+x_2^2}=1$ and the caustic curve is mapped to a
+ring corresponding to $v=\sqrt{z_1^2+z_2^2}=1$ and the caustic curve is mapped to a
 a single point in the source plane at $u=0$.
@@
 observer, lens and the source is rectilinear (acceleration can be neglected).
-We take $\vec{x}_S$ to be the angular position of the source star on the plane
+We take $\vec{w}_S$ to be the angular position of the source star on the plane
 of the sky and $\boldsymbol{\mu}_L$ to be its proper motion vector. Likewise
@@
 \begin{align}
-    \vec{x}_S(t) & =\vec{x}_{S, 0}+
+    \vec{w}_S(t) & =\vec{w}_{S, 0}+
     \left(t-t_{0}\right) \boldsymbol{\mu}_S
     \label{eq:source_position}      \\
-    \vec{x}_L(t) & =\vec{x}_{L, 0}
+    \vec{w}_L(t) & =\vec{w}_{L, 0}
     +\left(t-t_{0}\right) \boldsymbol{\mu}_L\quad,
@@
 \begin{equation}
-    \boldsymbol{u}(t) \equiv \frac{\vec{x}_L(t)
-        -\vec{x}_S(t)}{\theta_E}=
-    \frac{\vec{x}_{LS, 0}}{\theta_E}
+    \boldsymbol{u}(t) \equiv \frac{\vec{w}_L(t)
+        -\vec{w}_S(t)}{\theta_E}=
+    \frac{\vec{w}_{LS, 0}}{\theta_E}
     +\frac{t-t_{0}}{\theta_E} \boldsymbol{\mu}_{LS}
@@
 where
-$\vec{x}_{L S, 0}\equiv\vec{x}_{L,0}-\vec{x}_{S, 0}$
+$\vec{w}_{L S, 0}\equiv\vec{w}_{L,0}-\vec{w}_{S, 0}$
 is the relative position at $t_0$ and
@@
 relative proper motion.
-Since $\boldsymbol{\mu}_{LS}$ and $\vec{x}_{L S, 0}$ are
+Since $\boldsymbol{\mu}_{LS}$ and $\vec{w}_{L S, 0}$ are
 perpendicular to each other, it follows that the magnitude of the relative
@@
 \end{equation}
-where we have defined $u_0\equiv |\vec{x}_{L S, 0}|/\theta_E$
+where we have defined $u_0\equiv |\vec{w}_{L S, 0}|/\theta_E$
 and $t_E\equiv \theta_E/|\boldsymbol{\mu}_{LS}|$.
-The magnification is then a function of three parameters $(t_0, u_0, t_E)$
+The magnification is a function of three parameters $(t_0, u_0, t_E)$
 and the resulting curve as a function of time is often called the Paczy\'nski curve
@@
 curve \citep{2009MNRAS.393..816D}.
+
+\subsection{Astrometric microlensing}
+\label{ssec:astrometric_microlensing}
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=0.7\linewidth]{figures/single_lens_astrometric_shift.pdf}
+        \caption{
+            Astrometric shift in the centroid of light for a point source point lens
+            model.
+            The various curves correspond to different impact parameters $u_0$.}
+        \label{fig:single_lens_astrometric_shift}
+    \end{centering}
+\end{figure}
+
+In addition to the photometric microlensing effect, there is the astrometric
+effect -- the relative shift in the \emph{position} of the source star as the
+star is being magnified. In case of a point source and a single lens we can
+obtain the centroid of the two images by weighing their positions
+(Equation~\ref{eq:images_location}) with their magnifications
+(Equation~\ref{eq:magnification_per_image}):
+\begin{equation}
+    \vec w_c=\frac{A_+\vec w_+ + A_-\vec w_-}{A_+ + A_-}
+\end{equation}
+where $\vec w_c$ is the position of the centroid. It follows that the magnitude of the
+shift is
+\begin{equation}
+    w_c=\frac{1}{2}\left(\frac{u(u^2 + 4)}{u^2+2}+u\right)
+\end{equation}
+and the magnitude $\delta w_c$ of the shift
+$\delta \vec w_c\equiv\vec w_c - \vec u$ relative to the source star is
+\begin{equation}
+    \delta w_c = \frac{u}{u^2 + 2}
+\end{equation}
+The centroidal shift is plotted in Figure~\ref{fig:single_lens_astrometric_shift} for
+different values of the impact parameter $u_0$. The trajectories trace out ellipses
+\citep{1995ApJ...453...37W}.
+Whereas the photometric effect is
+strongest for
+$u\ll 1$  the astrometric shift peaks at $u=\sqrt{2}$ with
+$\delta w_{c,\max}\approx 0.354$ in units of the angular Einstein radius.
+Importantly, the astrometric shift falls of as $1/u$ for $u\gg \sqrt{2}$  which is
+much slower than $A\propto 1/u^4$ for the photometric effect.
 
@@
 brightness source) involving solutions to elliptic integrals of the first,
-second and third kind. A general solution involving Fast Fourier Transforms
-which is applicable to any source profile was recently proposed by
-\citet{2022arXiv220306637S}.
+second and third kinds. \citet{2009ApJ...695..200L} proposed a simple method for
+numerically integrating the point source magnification over the extent of the source
+disc which is fast and accurate for arbitrary intensity profiles.
+Another solution involving Fast Fourier Transforms any source profile was
+recently proposed by \citet{2022arXiv220306637S}.
 
@@
 source disc in the source plane. I will discuss all of this in far greater
-detail In Chapter~\ref{ch:microlensing} but for now I should mention a few key
-points:
+detail In Chapter~\ref{ch:microlensing_modeling} but for now I should mention a
+few key points:
 \begin{itemize}
@@
 \begin{align}
-    \vec{x}_S(t) & = \vec{x}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
+    \vec{w}_S(t) & = \vec{w}_{S,0}+(t-t_0)\boldsymbol{\mu}_S
     +\pi_S\,\boldsymbol{\zeta}(t)                            \\
-    \vec{x}_L(t) & = \vec{x}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
+    \vec{w}_L(t) & = \vec{w}_{L,0}+(t-t_0)\boldsymbol{\mu}_L
     +\pi_L\,\boldsymbol{\zeta}(t)
@@
 \subsection{A system with $N$ lenses}
+\label{ssec:N_lenses}
 Microlensing with more than one lens is considerably more complex than the case
@@
 source plane, given by the the dimensionless vector
-$\mathbf{y}=\boldsymbol\beta/(D_\textrm{S}\theta_E)$ and the position of the
+$\mathbf{z}=\boldsymbol\beta/(D_\textrm{S}\theta_E)$ and the position of the
 images in the lens plane, given by given by
-$\mathbf{x}=\boldsymbol\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now
+$\mathbf{w}=\boldsymbol\theta/(D_\textrm{L}\theta_E)$, where $\theta_E$ now
 refers to the angular Einstein radius corresponding to the total mass $M$. The
@@
 \begin{equation}
-    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\boldsymbol\alpha_i(\mathbf{x},\mathbf{x}_i)
+    \mathbf{z}=\mathbf{w}-\sum_{i=1}^N\boldsymbol\alpha_i(\mathbf{w},\mathbf{w}_i)
 \end{equation}
@@
 \begin{equation}
-    \mathbf{y}=\mathbf{x}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{x}-\mathbf{x}_i}
-    {\lvert \mathbf{x}-\mathbf{x}_i\rvert^2}
+    \mathbf{z}=\mathbf{w}-\sum_{i=1}^N\epsilon_i\frac{\mathbf{w}-\mathbf{w}_i}
+    {\lvert \mathbf{w}-\mathbf{w}_i\rvert^2}
     \label{eq:lens_equation_general}
@@
 It is useful to rewrite Equation~\ref{eq:lens_equation_general} in complex form
-by introducing complex variables \citep{1990A&A...236..311W}
-\begin{equation}
-    w= y_1+iy_2,\quad z=x_1+ix_2
+using complex variables instead of 2D vectors $\vec w$ and $\vec z$
+\citep{1990A&A...236..311W}:
+\begin{equation}
+    w\equiv w_1+iw_2,\quad z\equiv w_1+iw_2
 \end{equation}
@@
 \end{equation}
-The mapping from the image plane $z$ to the source plane $w$ is one-to-one and it is
-straightforward to compute using Equation~\ref{eq:lens_equation_complex}. The inverse
-mapping is not as easy. It turns out that it is possible to rewrite Equation~\ref{eq:lens_equation_complex}
-as a complex polynomial of degree $N^2 + 1$ (see Appendix~\ref{app:complex_poly}).
-From the Fundamental Theorem of Algebra, it follows that such a polynomial has exactly $N^2+1$ complex roots.
-However, not all of these roots are necessarily solutions to the lens equations so in practice one first has to
-compute all of the polynomial roots and then discard those which do not satisfy
-Equation~\ref{eq:lens_equation_complex}.
-In the case of binary lens, there are always either 3 or 5 real images and for $N\geq 2$ the maximum number
-of images is $5(N-1)$ \citep{arXiv:astro-ph/0103463,astro-ph/0305166,arXiv:math/0401188v2}.
-As before, the magnification is given by the inverse Jacobian determinant of the mapping $z\rightarrow w$
-evaluated at the images. We have
+Throughout the rest of the thesis I will be using the complex  for of the lens
+equation  for reasons that will become obvious shortly.
+
+The mapping from the image plane $z$ to the source plane $w$ is one-to-one and
+it is straightforward to compute using Equation~\ref{eq:lens_equation_complex}.
+The inverse mapping is not as easy. It turns out that it is possible to rewrite
+Equation~\ref{eq:lens_equation_complex} as a complex polynomial of degree $N^2
+    + 1$ (see Appendix~\ref{app:complex_poly}). From the Fundamental Theorem of
+Algebra, it follows that such a polynomial has exactly $N^2+1$ complex roots.
+However, not all of these roots are necessarily solutions to the lens equations
+so in practice one first has to compute all of the polynomial roots and then
+discard those which do not satisfy Equation~\ref{eq:lens_equation_complex}. In
+the case of binary lens, there are always either 3 or 5 real images and for
+$N\geq 2$ the maximum number of images is $5(N-1)$
+\citep{arXiv:astro-ph/0103463,astro-ph/0305166,arXiv:math/0401188v2}. As
+before, the magnification is given by the inverse Jacobian determinant of the
+mapping $z\rightarrow w$ evaluated at the images. We have
 \begin{equation}
@@
     \mathrm{det} \,\mathbf J=1-\left|\sum_{i=0}^{N} \frac{\epsilon_{i}}{\left(\bar{z}-\bar{z}_i\right)^{2}}\right|^{2}
+    \label{eq:lens_eq_det_jac}
 \end{equation}
@@
 Figure~\ref{fig:planetary_caustics} we show point source \emph{magnification
-    maps} (computed using the \textsf{caustics} code which is the subject of
-Chapter~\ref{ch:microlensing}) for a binary lens with $q=5\times 10^{-3}$,
-roughly corresponding to a gas giant planet orbiting an M dwarf. For the close
-topology (first two panels from the top in Figure~\ref{fig:planetary_caustics})
-we see one caustic centered on the star which is called a \emph{central
-    caustic} and two additional caustics on the opposite side of the planet,
-symmetric with respect to the star-planet axis, which are called
-\emph{planetary caustics} because they are associated with the planet rather
-than the star. Notice that the magnification in the vicinity of these planetary
-caustics is significantly smaller than the magnification around the central
-caustic. Because of this it is easier to search for binary events with
-trajectory passing close to the central caustic rather than the planetary
+    maps} (computed using the \ssf{caustics} code which is the subject of
+Chapter~\ref{ch:microlensing_modeling}) for a binary lens with $q=5\times
+    10^{-3}$, roughly corresponding to a gas giant planet orbiting an M dwarf. For
+the close topology (first two panels from the top in
+Figure~\ref{fig:planetary_caustics}) we see one caustic centered on the star
+which is called a \emph{central caustic} and two additional caustics on the
+opposite side of the planet, symmetric with respect to the star-planet axis,
+which are called \emph{planetary caustics} because they are associated with the
+planet rather than the star. Notice that the magnification in the vicinity of
+these planetary caustics is significantly smaller than the magnification around
+the central caustic. Because of this it is easier to search for binary events
+with trajectory passing close to the central caustic rather than the planetary
 caustics. As the separation $s$ approaches $s_c$, the critical value between
@@
           of a limb-darkened extended source is highly non-trivial, this is the subject
-          of Chapter~\ref{ch:microlensing}.
+          of Chapter~\ref{ch:microlensing_modeling}.
     \item \textbf{Multiple sources:} instead of a single source star we could have
@@
 
-\section{Occultation and phase curve mapping}
-\label{sec:occultations}
-In this section I will briefly review the theory behind modeling phase curves and
-occultation light curves of spherical bodies for the case of emitted light
-(isotropic blackbody emission from the body itself) and reflected light
-(starlight scattered from the surface or atmosphere of the body). The following is
-mostly a short summary of the papers introducing the \textsf{starry} framework
-\citep{2019AJ....157...64L,2021arXiv210306275L}.
-\subsection{The starry algorithm}
+%\section{Occultation and phase curve mapping}
+%\label{sec:occultations}
+%In this section I will briefly review the theory behind modeling phase curves and
+%occultation light curves of spherical bodies for the case of emitted light
+%(isotropic blackbody emission from the body itself) and reflected light
+%(starlight scattered from the surface or atmosphere of the body). The following is
+%mostly a short summary of the papers introducing the \ssf{starry} framework
+%\citep{2019AJ....157...64L,2021arXiv210306275L}.
+%\subsection{The starry algorithm}
+%\begin{figure}[t]
+%    \begin{centering}
+%        \includegraphics[width=\linewidth]{figures/spherical_harmonics.pdf}
+%        \caption{Real spherical harmonics up to degree $l=5$ computed from
+%            Equation~\ref{eq:spherical_harmonics}. Figure adapted from Figure 1 in
+%            \citet{2019AJ....157...64L}.}
+%        \label{fig:spherical_harmonics}
+%    \end{centering}
+%\end{figure}
+%\subsubsection{Emitted light}
+%Consider a spherical body of unit radius emitting light isotropically (a
+%Lambertian emitter) at each point $(\theta,\phi)$ where $\theta$ is the
+%inclination angle and $\phi$ is the azimuthal angle. To compute a phase curve
+%or an occultation light curve we need to be able to integrate the flux over the
+%entire projected disc of the body given the distribution of specific intensity
+%$I(\theta, \phi)$, the 3D orientation of the body in space, and the location of
+%a possible occultor relative to the body. These two-dimensional integrals can
+%always be computed numerically (see for instance the approaches presented in
+%\citet{2018AJ....156..146F,2018MNRAS.477.2613L}) but
+%\citet{2019AJ....157...64L} showed that it is possible to compute them
+%analytically if the specific intensity is first expanded in a basis of
+%spherical harmonics and then mapped to a different basis more suitable for
+%evaluating the integrals.
+%
+%Following \citet{2019AJ....157...64L}, we start by setting up a cartesian
+%coordinate system on the units sphere, such that
+%\begin{align}
+%     & x=\sin \theta \cos \phi \\
+%     & y=\sin \theta \sin \phi \\
+%     & z=\cos \theta
+%\end{align}
+%and the observer is located on the $z$-axis at $\infty$ such that the projected disc
+%of the body is centered at the origin of the $xy$ plane with the unit vector
+%$\hat{\mathbf{x}}$ pointing to the right and $\hat{\mathbf{y}}$ pointing up.
+%We introduce spherical harmonics $Y_{l m}(\theta, \phi)$ of degree $l\geq 0$ and order
+%$m\in [-l, l]$ defined  as
+%\begin{equation}
+%    Y_{l m}(\theta, \phi)= \begin{cases}\bar{P}_{l m}(\cos \theta) \cos (m \phi) & m \geqslant 0 \\ \bar{P}_{l|m|}(\cos \theta) \sin (|m| \phi) & m<0\end{cases}
+%    \label{eq:spherical_harmonics}
+%\end{equation}
+%When the spherical harmonics defined above are rewritten in terms of $x$, $y$, and $z$
+%they become polynomials in these variables (see Appendix A in \citet{2019AJ....157...64L}).
+%We can expand the specific intensity distribution $I(x,y)$ in a spherical harmonic basis
+%as
+%\begin{equation}
+%    I(x, y)=\tilde{\mathbf{y}}^{\intercal}(x, y) \mathbf{y}
+%    \label{eq:sh_expansion}
+%\end{equation}
+%where $\tilde{\mathbf{y}}$ is the basis vector of spherical harmonics  arranged in
+%increasing order:
+%\begin{align}
+%     & \tilde{\mathbf{y}}=\left(\begin{array}{lllll}
+%                                    Y_{0,0} & Y_{1,-1} & Y_{1,0} & Y_{1,1} & Y_{2,-2}
+%                                \end{array}\right. \\
+%     & \left.\begin{array}{lllll}
+%                 Y_{2,-1} & Y_{2,0} & Y_{2,1} & Y_{2,2} & \cdots
+%             \end{array}\right)^{\intercal} \quad,
+%\end{align}
+%and $\mathbf{y}$ is a vector of scalar \emph{spherical harmonic coefficients}.
+%$\mathbf{y}$ is a central quantity which defines the \emph{map} (the specific
+%intensity at every point) of the body.
+%\citet{2019AJ....157...64L} shows that we can represent $y$ in a polynomial basis (see
+%\citet{2019AJ....157...64L} for the general expression)
+%\begin{equation}
+%    \tilde{\boldsymbol{p}}=\left(\begin{array}{llllllllll}
+%        1 & x & z & y & x^{2} & x z & x y & y z & y^{2} & \cdots
+%    \end{array}\right)^{\intercal}
+%\end{equation}
+%such that
+%\begin{align}
+%    I(x, y) & =\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{p}                    \\
+%            & =\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_1 \mathbf{y}\quad,
+%    \label{eq:intensity_poly_basis}
+%\end{align}
+%and also in the Green's basis
+%\begin{equation}
+%    \tilde{\boldsymbol{g}}=\left(\begin{array}{llllllllll}
+%        1 & 2 x & z & y & 3 x^{2} & -3 x z & 2 x y & 3 y z & y^{2} & \cdots
+%    \end{array}\right)^{\intercal}
+%\end{equation}
+%such that
+%\begin{align}
+%    I(x, y) & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{g}              \\
+%            & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{A}_2 \mathbf{p} \\
+%            & =\tilde{\mathbf{g}}^{\intercal}(x, y) \mathbf{A} \mathbf{y}
+%\end{align}
+%where $\mathbf{A}_1$ is the change-of-basis matrix from $\mathbf{y}$ to $\mathbf{p}$,
+%$\mathbf{A}_2$ is the change-of-basis matrix from $\mathbf{p}$ to $\mathbf{g}$, and
+%$\mathbf{A}\equiv \mathbf{A}_1 \mathbf{A}_2$.
+%
+%To compute rotational light curves we also need to be able to specify the
+%orientation of a surface map with coefficients $\mathbf y$. We can rotate the
+%map using the Wigner rotation matrix $\mathbf{R}(\mathrm{I}, \Lambda, \Theta)$
+%that rotates $y$ given the body's inclination $\mathrm{I}$, obliquity $\Lambda$
+%and rotational phase $\Theta$. The rotated map is then $\mathbf{R}(\mathrm{I},
+%    \Lambda, \Theta) \mathbf{y}$ and Equation~\ref{eq:intensity_poly_basis} can be
+%rewritten more generally as
+%\begin{equation}
+%    I(x, y)=\tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_1 \mathbf{R}\mathbf{y}
+%\end{equation}
+%The total flux measured by an observer is is given by an integral of the specific
+%intensity over a region S of the projected disc of the body:
+%\begin{align}
+%    F & =\oiint I(x, y) \ud S                                                                 \\
+%      & =\oiint \tilde{\mathbf{p}}^{\intercal}(x, y) \mathbf{A}_{1} \mathbf{R} \mathbf{y} d S \\
+%      & =\mathbf{r}^{\intercal} \mathbf{A}_{1} \mathbf{R} \mathbf{y}
+%\end{align}
+%where $\mathbf{r}$ is a column vector whose $n$-th element is \citep{2019AJ....157...64L}
+%\begin{equation}
+%    r_{n} \equiv \oiint \tilde{p}_{n}(x, y) \ud S
+%\end{equation}
+%When the entire disc is visible (no occultor)
+%\begin{equation}
+%    r_{n}=\int_{-1}^{1} \int_{-\sqrt{1-x^{2}}}^{\sqrt{1+x^{2}}} \tilde{p}_{n}(x, y) \mathrm{d}y \mathrm{d} x
+%\end{equation}
+%and the solution to this integral can be expressed in terms of Gamma functions
+%\citep[Equation 20 in ][]{2019AJ....157...64L}.
+%$\mathbf{r}$ and $\mathbf{A}_1$ are independent of the map coefficients $\mathbf{y}$
+%so they can be pre-computed.
+%
+%Computing the occultation light curves is more complicated because the the body
+%is partially occulted by an occultor of radius $r$, centered at $(x_o, y_o)$
+%and the exposed portion of the disc is a function of $(r, x_o, y_o)$. The
+%general expression for the flux is
+%\begin{align}
+%    F & =\oiint I(x, y) \mathrm{d} S                              \\
+%      & =\mathbf{s}^\intercal  \mathbf{A} \mathbf{R} \mathbf{y} .
+%\end{align}
+%where the vector
+%\begin{equation}
+%    \mathbf{s}^\intercal \equiv \oiint \tilde{\mathbf{g}}^{\intercal}(x, y) \mathrm{d} S
+%\end{equation}
+%is defined to be the solution to the integral.
+%\citet{2019AJ....157...64L} solves this integral in two steps.  First, we rotate
+%the coordinate system about the  $z$-axis  so that the occultor lies at along the
+%$+y$ axis and its center is at
+%a distance $b=\sqrt{x_o^{2}+y_o^{2}}$ from the origin. This substantially simplifies
+%the limits of integration.
+%The second step is to use Green's theorem to express the surface integral of
+%$\tilde{\mathbf{g}}_n$ as a line integral of a vector function $\mathbf{G}_n$
+%along the boundary of the visible portion of the occulted disc.
+%The $n$-th element of the solution vector $\mathbf{s}^\intercal$ is
+%\begin{equation}
+%    s_{n}=\oiint \tilde{g}_{n}(x, y) \mathrm{d} S=\oint \mathbf{G}_{n}(x, y) \cdot \ud \mathbf{r}
+%\end{equation}
+%where
+%$\mathbf{G}_{n}(x, y)=G_{n x}(x, y) \hat{\mathbf{x}}+G_{n y}(x, y) \hat{\mathbf{y}}$
+%is constructed such that
+%\begin{equation}
+%    \mathbf{D} \wedge \mathbf{G}_{n}=\tilde{g}_{n}(x, y)
+%    \label{eq:G_n}
+%\end{equation}
+%were $\mathbf{D} \wedge \mathbf{G}_{n}$ is the exterior derivative of $\mathbf{G}_n$
+%given by
+%\begin{equation}
+%    \mathbf{D} \wedge \mathbf{G}_{n} \equiv \frac{\ud G_{n y}}{\ud x}-\frac{\ud G_{n_{x}}}{\ud y}
+%\end{equation}
+%in Cartesian coordinates.
+%\citet{2019AJ....157...64L} provides one possible expression for $\mathbf{G}_n$
+%which satisfies Equation~\ref{eq:G_n} and shows that the final solution can be written as
+%\begin{equation}
+%    s_{n}=\mathcal{Q}\left(\mathbf{G}_{n}\right)-\mathcal{P}\left(\mathbf{G}_{n}\right)
+%\end{equation}
+%where $\mathcal{P}(\mathbf{G}_n)$ is the  line integral along the arc of the occultor
+%of radius $r$ and $\mathcal{Q}(\mathbf{G}_n)$ is the line integral along the arc of the
+%occulted body of unit radius. Solutions to these integrals are long but they consist of
+%analytic functions such as sines, cosines and complete elliptic integrals and importantly
+%they only need to be evaluated once for an arbitrary map with fixed geometry of an
+%occultation \citep{2019AJ....157...64L}.
+%
+%The final expression for the total flux for a particular geometrical
+%arrangement of the occulted body and the occultor can then be written
+%succinctly as
+%\begin{equation}
+%    f = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{y}
+%    \label{eq:starry}
+%\end{equation}
+%Where $\mathbf{R}'$ is a rotation matrix which rotates the map such that the occultor
+%is placed symmetrically on the $+y$ axis at distance $b$.
+%Notice that Equation~\ref{eq:starry} defines a linear operation acting on a vector
+%of spherical harmonic coefficients $\mathbf{y}$.
+%It is also trivial to generalize this expression to a case of \emph{spectral map}. In
+%that case we define a matrix $Y$ whose columns are the spherical harmonic coefficients
+%at different wavelengths and the total flux (spectrum) is then
+%\begin{equation}
+%    \mathbf{f} = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{Y}
+%    \label{eq:starry_mw}
+%\end{equation}
+%where $\mathbf{f}$ is a vector of fluxes, one per wavelength bin.
+%
+%The formalism above can also be extended to account for limb-darkening which is
+%important when modeling transits of planets across stars or the light curves of
+%eclipsing binaries. \citet{2020AJ....159..123A} show how a limb-darkening
+%profile that is an order $l$ polynomial function of
+%$\mu=z=\sqrt{1-x^{2}-y^{2}}$ can be expressed in terms of the $m=0$ spherical
+%harmonics up to order $l$. In the case of quadratic limb-darkening
+%$(l_\mathrm{max}=2)$, the limb darkening profile is
+%\begin{equation}
+%    \frac{I(\mu)}{I(1)}=1-u_{1}(1-\mu)-u_{2}(1-\mu)^{2}
+%\end{equation}
+%which can be expressed as a sum of spherical harmonics
+%\citep[Equation 38 in][]{2019AJ....157...64L}. Limb-darkening needs to be
+%treated separately from the map coefficients $\mathbf{y}$ because it does not rotate along with the map
+%when rotations $\mathbf{R}$ and $\mathbf{R}'$ are applied. The limb-darkening
+%coefficients are applied to the map $\mathbf{y}$ as a multiplicative filter
+%following any rotations which results in another set of spherical harmonic
+%coefficients because products of spherical harmonics are also spherical
+%harmonics. Applying limb-darkening raises the degree of the map by the degree
+%of the limb-darkening.
+%
+%\subsubsection{Reflected light}
+%The results derived in the previous section are valid only for light emitted
+%from the occulted body. Equation~\ref{eq:starry} is a good model for stellar
+%light curves and secondary eclipse light curves and phase curves of planets
+%observed in the mid to far infrared wavelengths where the contribution of
+%scattered starlight to total flux is negligible. Modeling reflected light phase
+%curves and occultations is far more complicated because one has to take into
+%account the the nonuniform illumination of the body and the possible presence
+%of a terminator line (boundary between day and night).
+%\citet{2021arXiv210306275L} extends the formalism described in the previous
+%section to the case of reflected light phase curves and occultations for a body
+%whose spatial \emph{albedo} (specifically the \emph{spherical albedo} -- the
+%fraction of power incident on a body at a given wavelength that is scattered
+%into space in all directions) distribution is expanded in a basis of spherical
+%harmonics. Thus, analogous to Equation~\ref{eq:sh_expansion} we have
+%\begin{equation}
+%    A(x, y)=\tilde{\mathbf{y}}^{\intercal}(x, y) \mathbf{y}
+%\end{equation}
+%where $A$ is the albedo. We work under the assumption of isotropic (Lambertian)
+%scattering of light at every point on the body which means that the illumination
+%profile $\mathcal{I}$ of the surface is given by Lambert's law:
+%\begin{equation}
+%    \mathcal{I}\left(\vartheta_{\mathrm{i}}\right)=\mathcal{I}_{0} \max \left(0, \cos \vartheta_{\mathrm{i}}\right)
+%\end{equation}
+%where $\vartheta_i$ is the angle between the incident radiation at the surface normal
+%and $\mathcal{I}_0$ is the maximum illumination. In case of, for example, a planet illuminated
+%by its host star \citep[Appendix A.2 in][]{2021arXiv210306275L}
+%\begin{equation}
+%    \mathcal{I}_{0}=\frac{f_{s}}{\pi r_{\mathrm{s}}^{2}}
+%\end{equation}
+%where $r_s$ is the distance between the planet and the star in units of planet's radius
+%and $f_s$ is the stellar flux measured at the observer in some arbitrary units.
+%We assume that $f_s=1$ so that all fluxes are defined as the fraction of the flux
+%of the illumination source at the observer.
+%It is the presence of the day/night terminator that complicates the calculation
+%of total flux the most. The limits of integration end up depending on a
+%solution to a quartic equation specifying the points of intersection between
+%the occultor and the day/night terminator line and the solution to those
+%integrals are a function of \emph{incomplete} elliptic integrals as opposed to
+%complete elliptic integrals as was the case for occultations in emitted light.
+%The full derivation of the total flux is provided by \citet{2021arXiv210306275L} and
+%I will only briefly summarize  the result here for completeness.
+%
+%Under the Lambertian scattering assumption the observed intensity at any point
+%of the surface of the body is is proportional to the cosine of the angle
+%$\vartheta_i$ between the incident light rays and the surface normal. All
+%points for which $\vartheta_i\geq \pi/2$ have intensity of zero (they're
+%unilluminated). Let's assume that the point source is placed at sky coordinates
+%$\left(x_{\mathrm{s}}, y_{\mathrm{s}}, z_{\mathrm{s}}\right)$ in units of the
+%radius of the illuminated body. The day/night terminator on the body is a
+%half-ellipse with a semi-major axis equal to unity and a (signed) semi-minor
+%axis equal to
+%\begin{equation}
+%    b=-\frac{z_{\mathrm{s}}}{r_{\mathrm{s}}}
+%\end{equation}
+%where $r_{\mathrm{s}}=\sqrt{x_{\mathrm{s}}^{2}+y_{\mathrm{s}}^{2}+z_{\mathrm{s}}^{2}}$
+%is the distance to the source. The angle by which the semi-major axis of this ellipse is
+%rotated away from the $+x$-axis is  is
+%\begin{equation}
+%    \theta=-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
+%\end{equation}
+%Under the assumption that $rs \gg 1$, the illumination $\mathcal{I}$ at point
+%$(x,y)$ on the projected disc is \citep{2021arXiv210306275L}
+%\begin{equation}
+%    \mathcal{I}\left(b, \theta, r_{\mathrm{s}} ; x, y\right)=\max \left(0, I\left(b, \theta, r_{\mathrm{s}} ; x, y\right)\right)
+%    \label{eq:illumination}
+%\end{equation}
+%where
+%\begin{align}
+%    I\left(b, \theta, r_{\mathrm{s}} ; x, y\right) & =\frac{1}{\pi r_{\mathrm{s}}^{2}} \cos \vartheta_{\mathrm{i}}                                                      \\
+%                                                   & =\frac{1}{\pi r_{\mathrm{s}}^{2}}\left(-b_{\mathrm{c}} \sin \theta x+b_{\mathrm{c}} \cos \theta y-b z(x, y)\right)
+%\end{align}
+%with  $b_{\mathrm{c}} \equiv \sqrt{1-b^{2}}$  and $z(x, y)=\sqrt{1-x^{2}-y^{2}}$.
+%The illumination $\mathcal{I}$ is a dimensionless quantity normalized such that the
+%integral of $A\,\mathcal{I}$ over the units disc is equal to the flux measured by the
+%observer as a fraction of the flux of the illumination source.
+%To compute the total flux, we need to weigh each of the terms in the Green's basis
+%and integrate them over the visible portion of the body's disc to obtain
+%the reflected light solution vector $\mathbb{s}^\intercal$.
+%Evaluating these integrals is unfortunately intractable because of the piecewise
+%function in Equation~\ref{eq:illumination}. It is more tractable to weight the basis
+%terms by the function $I$ and modify the limits of integration such that the nightside
+%of the body is excluded. \citet{2021arXiv210306275L} shows that $I$ can be expressed
+%in the form of a linear operator $\mathbf{I}$ to weight a map vector  in the polynomial
+%basis $\tilde{\mathbf{p}}$ by the illumination profile. The total flux is then
+%\begin{equation}
+%    f=\mathbb{s}^\intercal\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right) \mathbf{A}_{\mathbf{2}} \mathbf{I}\left(b, \theta^{\prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime}\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+%\end{equation}
+%where
+%\begin{equation}
+%    \theta^{\prime}=\arctan 2\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right)-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
+%    \label{eq:starry_ref}
+%\end{equation}
+%is the angle of the terminator in frame $\mathcal{F}^\prime$.
+%In case the illuminated body is not occulted we have
+%\begin{equation}
+%    f_{0}=\mathbb{r}^{\intercal}(b) \mathbf{I}\left(b, \theta^{\prime \prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime \prime}\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
+%    \label{eq:starry_ref_phase}
+%\end{equation}
+%where $\theta^{\prime\prime}=0$ is the angle of the terminator in frame
+%$\mathcal{F}^{\prime\prime}$ by construction. $\mathbf{R}^{\prime\prime}$ rotates the
+%body through an angle $\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)$ so the
+%semi-major axis of the terminator is aligned with the $x^{\prime \prime}$ axis.
+%The solutions to integrals
+%$\mathbb{r}^{\intercal}(b)$ and $\mathbb{s}^{\intercal}\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right)$
+%are anything but trivial, they are derived in the appendices of
+%\citet{2021arXiv210306275L}. The key point is that the model is once again linear and
+%the matrices in Equations~\ref{eq:starry_ref} and \ref{eq:starry_ref_phase} need to be
+%computed only once for a specific occultation/illumination geometry.
+%\citet{2021arXiv210306275L} also derives a solution for the total flux in case the
+%light source is extended (the assumption that $r_s\gg 1$ is not satisfied) and
+%in case the body is not a perfect Lambertian reflector.
+%
+%\subsection{The starry code}
+%The method described in the previous section is implemented in the
+%\ssf{Python} package
+%\ssf{starry}\footnote{\url{https://starry.readthedocs.io/en/latest/}}.
+%\ssf{starry} is many orders of magnitude faster and more accurate than
+%similar codes which rely on numerical integration. It enables computation of
+%rotational light curves, planetary transits, secondary eclipse light curves and
+%planet-planet occultations in both emitted and reflected light. Recent
+%extensions to \ssf{starry}, not mentioned in Section~\ref{sec:occultations},
+%extended the spherical harmonic formalism to the problem of Doppler imaging
+%\citep{2021arXiv211006271L} and add support for computing occultation light
+%curves of oblate bodies \citep{2022ApJ...925..185D}. In
+%Chapters~\ref{ch:mapping_io} and \ref{ch:mapping_exoplanets} we make extensive
+%use of \ssf{starry} to explore solutions to the \emph{inverse problem} --
+%how can we obtain an estimate of the map coefficients $\mathbf{y}$ which best
+%explain the observed light curve? \ssf{starry} abstracts the complexities of
+%computing the total flux and allows us to treat this problem as a linear
+%problem of the form
+%\begin{equation}
+%    f = \mathbf{X}\mathbf{y}
+%    \label{eq:starry_linear_model}
+%\end{equation}
+%where  $\mathbf{X}$ is computed by \ssf{starry} depending on the geometry of an
+%occultation or transit event. The inverse problem is highly degenerate
+%and making progress requires  imposing some prior structure on the solution
+%$\mathbf{y}$.
+%
+%In Chapter~\ref{ch:mapping_io} we use \ssf{starry} to model occultation
+%light curves of a Solar System object -- Jupiter's moon Io, as it is occulted
+%by Jupiter. Thanks to the comparatively close distance to Io relative to
+%objects outside of the Solar System, these light curves are very high quality
+%we are able to fit very high order maps $l=20$ and test our models in a high
+%signal-to-noise regime. We also discuss how possible approaches to modeling
+%time-dependent maps in which case the spherical harmonic coefficients are
+%time-dependent: $\mathbf{y}=\mathbf{y}(t)$. In
+%Chapter~\ref{ch:mapping_exoplanets} we tackle a very similar problem but in a
+%very different regime of exoplanet eclipse mapping where the signal-to-noise is
+%orders of magnitude worse than in the case of Io. In both cases we mostly focus
+%on thermal light curves rather than light curves in reflected light.
+%
+%\section{Statistical inference -- introduction}
+%\label{sec:inference_intro}
+%Having covered the basic physics behind gravitational lensing and occultation mapping I
+%now turn to the subject of statistical inference in general and more specifically --
+%Bayesian statistical inference. Bayesian (as opposed to frequentist)
+%provides a straightforward and elegant theoretical framework for solving inverse
+%problems of the kind we mentioned
+%in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}. These are problems
+%where we have little data per object of interest, the data is very noisy,
+%the physical models are relatively well understood but their parameters don't map
+%to observables in a straightforward way and there are multiple competing, sometimes
+%physically  quite different, models which provide a good explanation for the data.
+%Having said that, I am not particularly dogmatic about the use of Bayesian methods
+%and I will occasionally steer away from the classical approach and point out where
+%a less Bayesian approach is often superior in practice.  What is most important is the
+%use of \emph{probabilities} for encoding uncertainty about the physical world.
+%
+%I start by reviewing basic probability theory and briefly mention the
+%frequentist interpretation. I then introduce two key concepts, the likelihood
+%function -- a recipe for generating plausible datasets given model parameters,
+%and priors -- what we know about the model parameters prior to gathering any
+%data. Next, I briefly summarize fundamental concepts such as the curse of
+%dimensionality and the no-free-lunch-theorem. Finally I end the section by
+%comparing machine learning to statistical inference within the physical
+%sciences.
+%
+%\subsection{Probability theory}
+%There are two different interpretations of probability. In the
+%\emph{frequentist} interpretation probabilities are \emph{frequencies of
+%    events} which are repeated in a large number of trials (think of a repeated
+%coin tosses). An alternative to the frequentist interpretation is the
+%\emph{Bayesian} interpretation in which probabilities represent \emph{degrees
+%    of belief} about the subject of interest. In both cases probabilities are
+%necessarily positive numbers between zero and one. The major advantage of the
+%Bayesian interpretation is that it allows one to reason about the probability
+%of one-off events such as ``person X wins the election'' or ``this microlensing
+%event is a binary lens event''. The differences between the two interpretations
+%are not just philosophical because which interpretation one subscribes to can
+%determine how one approaches a particular statistical inference problem and
+%which sets of tools one uses. Both approaches are commonly used within the
+%physical sciences depending on the field and the problem at hand. For instance,
+%in case of particle physics frequentist methods are much more common because
+%particle physics experiments often involve billions of repeated experiments in
+%particle accelerators. One the other hand, Bayesian methods are more popular in
+%astronomy because we cannot do true repeated experiments with astrophysical
+%objects.
+%
+%The mathematical rules for manipulating probabilities are quite
+%straightforward. In probability theory we use \emph{random variables} which can
+%be either discrete or continuous and their possible values are described with
+%probability distributions over all possible values the variable is allowed to
+%take. For discrete random variables we can talk of probabilities that a given
+%variable assumes a particular value, say $5$, but with continuous random
+%variables we can only talk of probabilities that a given variable is in a
+%particular range of values, say $[0,1]$. Now let's consider a continuous
+%\emph{random variable} $x$ with an associated probability density function
+%(pdf) $p(x)$\footnote{The notation $p(x)$ we use is somewhat overloaded. In
+%    statistics, the proper notation for a probability density function is
+%    $p_{X}(x)$ where $X$ is a random variable and $x$ is a particular realization
+%    of that random variable.} Since $x$ has to take on \emph{some values} within
+%its domain we have the requirement that
+%\begin{equation}
+%    \int p(x)\,\ud x=1
+%    \label{eq:normalization_condition}
+%\end{equation}
+%where the integral is taken over the entire domain of $x$.
+%In order for Equation~\ref{eq:normalization_condition} to hold,
+%$p(x)$ has to have units of $x^{-1}$ so that
+%the integral is a dimensionless number.
+%
+%In general, we are usually dealing with several parameters at once in which
+%case we are interested in \emph{joint} probability distributions over all the
+%parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
+%density function $p(x_1,x_2)$ which when integrated over a particular interval
+%gives the probability that both $x_1$ and $x_2$ are contained within that
+%interval. These joint probabilities can be decomposed into \emph{conditional
+%    probabilities} using the so called product rule:
+%\begin{align}
+%    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
+%    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
+%    \label{eq:product_rule}
+%\end{align}
+%where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
+%    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
+%valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
+%Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
+%\begin{equation}
+%    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
+%    \label{eq:bayes_theorem}
+%\end{equation}
+%Bayes's theorem is simply a rule for converting one kind of a conditional probability
+%into another.
+%Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
+%\emph{marginalize} parameters that we are not interested in to obtain a distribution over
+%parameters of interest:
+%\begin{equation}
+%    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
+%    x_2)\,p(x_2)\textrm{d}x_2
+%    \label{eq:marginalization}
+%\end{equation}
+%where we applied Bayes's theorem in the second inequality.
+%In cases where the parameter space is high dimensional these integrals might be
+%impossible to calculate analytically but they can (sometimes) be estimated using sampling
+%algorithms as we shall se in subsequent sections.
+%In some cases, we may be able to factor out $p(x_1,x_2)$ as
+%If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
+%\begin{equation}
+%    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
+%\end{equation}
+%Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
+%of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
+%of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
+%\begin{equation}
+%    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
+%    \label{eq:likelihood_indep}
+%\end{equation}
+%If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
+%identically distributed (``iid''). If the data is independent but not identically distributed
+%we say that the uncertainties are \emph{heteroscedastic}.
+%The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
+%uncertainties.
+%
+%\subsubsection{Expectation values and the mode}
+%The \emph{mean} or the \emph{expected value} of a distribution $p(\vec{x})$,
+%usually denoted by $\mu$, is a measure of the ``center'' of a distribution and
+%it is defined as
+%\begin{equation}
+%    \mathbb{E}[x] \equiv\int x \,p(x) \ud x
+%\end{equation}
+%Similarly, the \emph{variance} is a measure of the ``spread'' of a distribution:
+%\begin{align}
+%    \mathbb{V}[x] & \equiv\mathbb{E}\left[(x-\mu)^{2}\right]=\int(x-\mu)^{2} p(x) d x                                           \\
+%                  & =\int x^{2} p(x) \ud x+\mu^{2} \int p(x) \ud x-2 \mu \int x p(x) \ud x=\mathbb{E}\left[x^{2}\right]-\mu^{2}
+%\end{align}
+%The variance is usually denoted by $\sigma^2$ and the square root of $\sigma^2$ is the
+%\emph{standard deviation}.
+%For a \emph{multivariate} random variable with $D$ dimensions then the variance generalizes to
+%a symmetric, positive semi definite matrix called the \emph{covariance matrix}:
+%\begin{align}
+%    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\intercal}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                           \\
+%                                & =\left(\begin{array}{cccc}
+%                                             \mathbb{V}\left[x_{1}\right]                & \operatorname{Cov}\left[x_{1}, x_{2}\right] & \cdots & \operatorname{Cov}\left[x_{1}, x_{D}\right] \\
+%                                             \operatorname{Cov}\left[x_{2}, x_{1}\right] & \mathbb{V}\left[x_{2}\right]                & \cdots & \operatorname{Cov}\left[x_{2}, x_{D}\right] \\
+%                                             \vdots                                      & \vdots                                      & \ddots & \vdots                                      \\
+%                                             \operatorname{Cov}\left[x_{D}, x_{1}\right] & \operatorname{Cov}\left[x_{D}, x_{2}\right] & \cdots & \mathbb{V}\left[x_{D}\right]
+%                                         \end{array}\right)
+%\end{align}
+%
+%In the general case, we can compute expectation values of an arbitrary function
+%$g(\vec{x})$ under the probability distribution $p(\vec{x})$:
+%\begin{equation}
+%    \mathbb{E}\left[g(\vec{x})\right]=\int g(\vec{x})\,p(\vec{x})\ud x
+%    \ud\theta
+%\end{equation}
+%Some examples of $g$ include higher moments of the distribution, the median and
+%quantile estimates. The \emph{mode} of $p(\vec{x})$
+%\begin{equation}
+%    \vec{x}^{*}=\underset{\vec{x}}{\operatorname{argmax}}\,p(\vec{x})
+%\end{equation}
+%on the other hand is not an expectation value. It is the global maximum of the
+%function $p(\vec{x})$.
+%Estimating the mode involves computing the \emph{derivatives} of the density
+%$p(\vec{x})$, whereas expectation values require computing
+%\emph{integrals} over the space $\vec{x}$. Computing either
+%the mode or expectation values
+%for nontrivial densities $p(\vec{x})$ is a very hard and often impossible problem
+%(more on this in Section~\ref{sec:inference_practice}).
+%
+%\subsubsection{Central Limit Theorem}
+%Consider $N$ \emph{independent and identically distributed} random variables
+%with pdf's $p_n(x)$ which are not necessarily Gaussian. Let $S_N$ be the sum
+%the random variables. The \emph{Central Limit Theorem} (CLT) says that as $N$
+%increases, the distributions of this sum approaches the Gaussian distribution
+%\begin{equation}
+%    p\left(S_{N}=u\right)=\frac{1}{\sqrt{2 \pi N \sigma^{2}}} \exp \left(-\frac{(u-N \mu)^{2}}{2 N \sigma^{2}}\right)
+%\end{equation}
+%and the distribution of the quantity
+%\begin{equation}
+%    Z_{N} \equiv \frac{S_{N}-N \mu}{\sigma \sqrt{N}}=\frac{\bar{x}-\mu}{\sigma / \sqrt{N}}
+%\end{equation}
+%where $\bar{x}$ is the \emph{sample mean} $\frac{1}{N}\sum_{i=1}^Nx_i=S_N/N$
+%converges to the Gaussian distribution with zero mean and unit variance.
+%Because lots of random variables representing real world processes can be seen as sums
+%of i.i.d. independent random variables the Gaussian distribution is all over the place
+%in statistics. One should keep in mind that the assumptions underlying the CLT are
+%not always satisfied and also the convergence towards a Gaussian is
+%more rapid in the bulk of the distribution around the mean than in the tails of the
+%distribution.
+%
+%\subsubsection{Change of variables}
+%Given a univariate random variable $x$ the effect of some deterministic
+%transformation $y(x)$ is to stretch and shift the distribution $p(x)$. For an
+%infinitesimal interval the \emph{probability mass} is the same in both
+%variables so $p(x) \ud x=p(y) \ud y$ which yields the \emph{change of variables
+%    formula}
+%\begin{equation}
+%    p_{y}(y)=p_{x}(x)\left|\frac{\ud x}{\ud y}\right|
+%\end{equation}
+%In the multivariate case, if $\vec{f}$ is an invertible function mapping $\mathbb{R}^n$
+%to $\mathbb{R}^n$ with inverse $\vec{g}$, the pdf of $\vec{y}=\vec{f}(\vec{x})$ is
+%\citep{murphy_book_2022}
+%\begin{equation}
+%    p_{y}(\vec{y})=p_{x}(\vec{g}(\vec{y}))\left|\operatorname{det}\mathbf{J}_{g}(\vec{y})\right|
+%\end{equation}
+%where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\intercal}}$ is the Jacobian of $\vec{g}$
+%and $\operatorname{det}|\mathbf{J}(\vec{y})|$ is the determinant of the Jacobian.
+%
+%\subsubsection{Kullback Leibler divergence}
+%A common way of measuring ``similarity'' between two distributions $p(x)$ and
+%$q(x)$ is using the \emph{Kullback-Leibler divergence} (KL divergence for
+%short):
+%\begin{equation}
+%    D_{\mathrm{KL}}(p \| q)=\int_{-\infty}^{\infty} p(x) \ln \left(\frac{p(x)}{q(x)}\right) \ud x
+%    \label{eq:kl_divergence}
+%\end{equation}
+%where $D_{\mathrm{KL}}(p \| q)$ should be read as ``the KL divergence from p to q''.
+%This is not a measure of ``distance'' between the two distributions because
+%it is not a symmetric quantity. The KL divergence is best understood as a quantity which
+%quantifies \emph{how surprised one would be if one expected to encounter distribution $p(x)$
+%    but instead saw distribution
+%    $q(x)$}\footnote{\url{https://wiki.santafe.edu/images/a/a8/IT-for-Intelligent-People-DeDeo.pdf}}.
+%This quantity is important in the context of variational inference which is covered in
+%Section~\ref{sec:inference_practice}.
+%
+%\subsection{Bayes' theorem in practice}
+%\label{ssec:likelihood_function}
+%Consider a general inference problem where we collect some data $\mathcal{D}$ and we are interested
+%in the joint probability distribution for parameters $\boldsymbol\theta$ of particular model.
+%We can then rewrite Bayes' theorem as
+%\begin{equation}
+%    p(\vec{\theta} | \mathcal{D})=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{p(\mathcal{D})}=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{\int p\left(\vec{\theta}^{\prime}\right) p\left(\mathcal{D} |\vec{\theta}^{\prime}\right) d \vec{\theta}^{\prime}}
+%    \label{eq:bayes_theorem2}
+%\end{equation}
+%In Equation~\ref{eq:bayes_theorem2} the term $p(\mathcal{D}|\vec{\theta})$  is a probability
+%distribution over the data $\mathcal{D}$ given the parameters of the model $\vec{\theta}$.
+%When this term is seen as a function of the parameters given a fixed dataset, it is
+%called the \emph{likelihood function }  or the \emph{likelihood} for short. $p(\boldsymbol\theta)$
+%is to so called \emph{prior} distribution over $\vec{\theta}$ and it represents our beliefs
+%about $\vec{\theta}$ prior to having observed the data $\mathcal{D}$. When the prior distribution is
+%multiplied by the likelihood we obtain, up to a normalizing constant $p(\mathcal{D})$, the posterior
+%distribution over $\vec{\theta}$ given the data $\mathcal{D}$.
+%The best way to think about Equation~\ref{eq:bayes_theorem2} is seeing at as a rule
+%for updating beliefs. Our prior beliefs about the parameters of the model
+%$p(\mathrm{parameters})$ should be updated to posterior beliefs $
+%    p(\mathrm{parameters} | \mathrm{data})$ by making use of the likelihood
+%$p(\mathrm{data} | \mathrm{parameters})$. If the data is useful, the likelihood is generally
+%much narrower than the prior pdf and the choice of priors does not significantly
+%influence the final results.
+%
+%\subsubsection{The likelihood}
+%Let's cover each of these three terms in more detail, starting with the most
+%important one -- the likelihood. The likelihood can be seen as a \emph{recipe
+%    for generating plausible datasets $\mathcal{D}$} (a \emph{generative model})
+%given a specific realization of the parameters $\vec{\theta}$. For the kinds of
+%models we work with in astronomy, the likelihood consists of two parts. The
+%first part is (usually deterministic) function which describes the physical
+%system of interest. This is for instance the predicted flux for a binary
+%microlensing event with a particular trajectory. The second part is the
+%\emph{noise model} which describes the statistical properties of the noise
+%inherent in any physical measurement. We often assume that the noise is
+%independent and Gaussian. It is very common to use the likelihood as is and
+%optimize it with respect to the model parameters $\vec{\theta}$ in a procedure
+%called \emph{maximum likelihood estimation} (MLE) to find a single point in the
+%parameter space $\vec{\theta}^\star$ which maximizes the likelihood (more on
+%this in Section~\ref{ssec:mle_estimation}).
+%
+%We can also marginalize the likelihood over a subset of parameters. Say that we
+%have a likelihood function $p(\vec{\theta} ,\vec{\alpha}|\mathcal{D})$ where
+%$\vec{\alpha}$ is a set of \emph{nuisance parameters} which we are not
+%particularly interested in. To marginalize over these nuisance parameters we
+%need to introduce a prior\footnote{Not introducing a prior leads to a
+%    dimensionally incorrect result.} $p(\vec{\theta}|\vec{\theta})$ which may
+%potentially also depend on the $\vec{\theta}$ and the marginalized likelihood
+%is then
+%\begin{equation}
+%    p(\mathcal{D} | \vec{\theta})=\int p(\mathcal{D} | \vec{\theta}, \vec{\alpha})\,p(\vec{\alpha} |\vec{\theta}) \ud \vec{\alpha}\quad .
+%\end{equation}
+%The problem structure where where $\vec{\alpha}$ is some set of parameters
+%we don't particularly care about such and $\vec{\theta}$ are the physical
+%parameters of interest is very common in astronomy. For instance, in single lens gravitational microlensing estimating
+%the distribution $p(t_E)$ over the parameter of interest $t_E$ requires marginalizing over
+%parameters such as $F_S, F_B, t_0$ and $u_0$.
+%
+%Not all statistical models can be written in terms of the likelihood function.
+%Most machine learning models for example do not have a simple probabilistic
+%interpretation\footnote{Indeed, finding probabilistic interpretations of
+%    complex machine learning models such as deep neural networks is an active
+%    research area.}. If we can write down the likelihood function for a certain
+%problem, we say that we have a \emph{generative model for the data} because the
+%likelihood can be used to generate mock data sets.
+%
+%\subsubsection{Priors}
+%The second term of interest is the prior $p(\vec{\theta})$. In principle, the
+%prior pdf-s should capture all of the prior information we might have about the
+%model parameters prior to gathering new data; in practice, we don't want to
+%bias our results so we often try to use ``weakly informative'' priors -- priors
+%that are informative enough that they exclude unreasonable parameter values but
+%aren't so narrow that they rule out sensible values which may be favored by the
+%likelihood. For example, we might have prior information such as ``the mass of
+%the planet should be positive'' in which case a suitable weakly informative
+%prior on the mass parameter should first of all exclude negative values but
+%also potentially exclude values which are nonsensically large for the problem
+%at hand. There are also the so called ``non-informative prior'' such as
+%Jeffrey's prior and maximum entropy priors. Although these priors can be of use
+%in some cases, in general \textbf{there no such thing as a truly non
+%    informative prior}. Although a prior for a particular parameter $\theta$ might
+%be uninformative in one set of coordinates, it is often informative for a
+%certain transformation of the parameter $g(\theta)$. For example, a uniform
+%prior in $\theta$ is not uniform in $\ln\theta$. Having said that, it is often
+%worth the effort to construct priors which respect certain symmetries of the
+%problem the we know to be true.
+%
+%Priors should not be seen in isolation, rather, they should be seen in the
+%context of the likelihood \citep{arXiv:1708.07487}. To see that this is the
+%case consider drawing samples from the prior $\vec{\theta}\sim
+%    p(\vec{\theta})$, plugging those values into the likelihood distribution
+%$p(\mathcal{D}|\vec{\theta})$ and then generating mock datasets, a process
+%called \emph{prior predictive checking}. This process makes it easy to spot
+%priors which in combination with the likelihood yield predictions of what the
+%data should like like which do not match our expectations. An excellent highly
+%in depth study on choosing priors is available
+%\href{https://betanalpha.github.io/assets/case_studies/prior_modeling.html}{here}.
+%
+%Depending on how informative the data is we differentiate between two different
+%regimes. We say that the inference process is \emph{data driven} if the data
+%constrains the model parameters so well that the choice of the priors is
+%effectively irrelevant. The other regime is said to be \emph{prior driven},
+%when the data is so weakly informative that the shape of the posterior pdf-s is
+%similar to the prior pdf. In general, in any statistical analysis involving
+%priors, one should check the sensitivity of the scientific conclusion to
+%assumptions of different priors. No statistical inference procedure is free of
+%assumptions but the use of priors and likelihoods makes those assumptions more
+%explicit.
+%
+%Priors are sometimes the goal of the inference. For example, if the prior on
+%the prior pdf $p(\vec{\theta})$ itself depends on some other set of parameters
+%$\vec{\beta}$, then we can marginalize out all of the $\vec{\theta}$ parameters
+%and we are left with a likelihood for the parameters $\vec{\beta}$ of the
+%prior.
+%\begin{equation}
+%    p(\mathcal{D}\lvert\vec{\beta})=\int p(\mathcal{D}\lvert\vec{\theta})\,p(\vec{\theta}\lvert
+%    \vec{\beta})\ud \vec{\theta}
+%\end{equation}
+%This can then be used, using Bayes's theorem, to calculate the posterior
+%over the prior parameters $\bm\beta$.
+%\begin{equation}
+%    p(\bm\beta\lvert\mathcal{D})\propto p(\mathcal{D}\lvert\vec{\beta})
+%    \,p(\bm\beta)
+%\end{equation}
+%The parameters $\bm\beta$ characterizing the prior are usually called
+%\emph{hyperparameters} or \emph{population level parameters}. Similarly, the priors
+%for those parameters are called \emph{hyperpriors}.
+%The whole process is called \emph{hierarchical Bayesian inference}
+%(in astronomy literature),  or \emph{multilevel modelling, hierarchical modelling,
+%    nested modelling, mixed models or random effects models}
+%(in statistics and adjacent fields).
+%Hierarchical modelling is an very useful method for estimating
+%population-wide properties given data about individual objects. For example,
+%in microlensing observations of exoplanets the individual parameters could be the
+%masses of individual planets while the hyperparameters would describe the
+%population the shape of the planetary mass function. We shall return to this topic
+%in Chapter~\ref{ch:microlensing_modeling}.
+%
+%\subsubsection{The evidence}
+%The final component of Equation~\ref{eq:bayes_theorem2} is the normalization
+%constant $ p(\mathcal{D})$ in the denominator:
+%\begin{equation}
+%    \mathcal{Z}\equiv p(\mathcal{D})=\int p(\mathcal D\lvert\boldsymbol\theta)\,p(\boldsymbol\theta)\textrm{d}\boldsymbol\theta
+%    \label{eq:evidence}
+%\end{equation}
+%is often called the \textsl{fully marginalized likelihood} because
+%it is a likelihood marginalized over all of the parameters of the model.
+%Sometimes
+%it is also called the \emph{Bayesian evidence}.
+%$\mathcal{Z}$ is general extremely difficult to calculate because it is an
+%integral over a potentially very high dimensional parameter space $\vec{\theta}$.
+%Fortunately, in most cases, we can obtain the posterior distribution without having
+%to calculate $\mathcal{Z}$.
+%
+%%% mention the likelihood principle
+%
+%\subsection{Maximum likelihood estimation}
+%One of the most popular inference methods is called \emph{maximum likelihood
+%    estimation} (MLE). Maximum likelihood estimation answers the question:
+%\textbf{what choice of parameters $\vec{x}$ makes my data most likely} as
+%opposed to the question \textbf{what are the most likely values of the
+%    parameters $\vec{x}$ having observed the data $\mathcal{D}$}\footnote{The
+%    distinction between the two questions is subtle, but important. The parameters
+%    which maximize the likelihood might be very different from those which are most
+%    ``sensible'' and which best describe the data.} which necessitates the
+%multiplication of the likelihood with a prior. Since we don't incorporate the
+%prior, it is not a Bayesian method. The output of MLE is a \emph{point
+%    estimate} of the parameters $\hat{\vec{\theta}}$. In a Bayesian viewpoint, the
+%MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.
+%The MLE estimate is obtained by maximizing the logarithm\footnote{Because
+%    probabilities in most region of the parameter space are very small numbers we
+%    always use the log transform when optimizing likelihoods or posteriors.} of the
+%likelihood $p(\mathcal{D}\lvert\vec{\theta})$ (or equivalently, minimizing the
+%negative log likelihood) with respect to the parameters $\vec{\theta}$:
+%\begin{equation}
+%    \hat{\boldsymbol{\theta}}_{\mathrm{mle}}=\underset{\boldsymbol{\theta}}{\operatorname{argmax}}
+%    \ln p\left(\mathcal{D}\lvert \boldsymbol{\theta}\right)
+%\end{equation}
+%
+%\subsubsection{MLE for a multivariate Gaussian}
+%The log-likelihood of multivariate Gaussian distribution without the
+%unnecessary constants is
+%\begin{equation}
+%    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\ln p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \ln |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\intercal} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
+%\end{equation}
+%where $\vec{y}_n$ are the data points and $\vec{\Sigma}^{-1}$ is
+%the inverse covariance matrix, also knowns as the \emph{precision matrix}.
+%The MLE for the mean parameter $\hat{\vec{\mu}}$
+%is given by the sample mean (empirical mean) of the data \citep{murphy_book_2022}:
+%\begin{equation}
+%    \hat{\boldsymbol{\mu}}=\frac{1}{N} \sum_{n=1}^{N} \boldsymbol{y}_{n}=\overline{\boldsymbol{y}}\quad .
+%\end{equation}
+%Similarly, the MLE of the the covariance matrix $\hat{\vec{\Sigma}}$ is equal to
+%the empirical covariance matrix
+%\begin{equation}
+%    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\intercal}
+%\end{equation}
+%Thus, whenever we compute an empirical mean and the covariance for a given datasets we
+%are implicitly assuming that the data is normally distributed and obtaining an MLE
+%estimate of that distribution.
+%
+%\subsection{Least squares linear regression}
+%\label{ssec:least_squares}
+%In this section we cover the classical non Bayesian formulation of linear
+%regression, also known as \emph{least squares regression} and in the next
+%section we generalize to the Bayesian case. In a (multiple) linear regression
+%model we have (noisy) measurements $\vec{y}\in \mathbb{R}^n$ (\emph{dependent
+%    variable}), an $n\times p$ matrix $\vec{M}\in \mathbb{R}^{n\times p}$ of inputs
+%(\emph{design matrix} consisting of \emph{independent variables},
+%\emph{covariates}, or \emph{features}) and a vector of \emph{regression
+%    coefficients} or \emph{weights} $\vec{w}\in \mathbb{R}^p$. This is a very
+%general model because the matrix $\vec{M}$ can consist of basis functions such
+%as polynomials, spherical harmonics, Fourier modes, wavelets etc. so although
+%the model is linear in the parameters (regression coefficients), it is not
+%generally linear in the raw input data. If we assume that the observations
+%$\vec{y}$ have Gaussian noise we can write down the likelihood of this model as
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{w})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{w}, \vec{C})
+%\end{equation}
+%where $\vec{C}$ is the data covariance matrix and $\mathcal{N}$ is shorthand notation for
+%the multivariate normal distribution. In astronomical application it is usually
+%assumed to be diagonal and we might have some idea for the scale of the variance
+%of the diagonal elements (the ``errorbars'').
+%The negative log likelihood (NLL) is then equal (ignoring constants)  to
+%\begin{equation}
+%    \operatorname{NLL}(\vec{w})=\frac{1}{2}(\vec{M} \vec{w}-\vec{y})^{\intercal}\vec{C}^{-1}(\vec{M} \vec{w}-\vec{y})
+%\end{equation}
+%which is just the well known least squares objective or (half the) $\chi^2$ as it is known
+%in astronomy.
+%
+%\subsubsection{Underdetermined case ($p < n$)}
+%For now, let's assume that the inverse data covariance matrix $\vec{C}^{-1}$ is
+%a diagonal matrix with the the same variance $\sigma^2$ for every data point.
+%In that case the MLE solution for the weights $\vec{w}$ is the solution to the
+%following (convex) optimization problem:
+%\begin{equation}
+%    \hat{\vec{w}}=
+%    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2}
+%    \label{eq:ols_objective}
+%\end{equation}
+%There is a closed form solution for Equation~\ref{eq:ols_objective} as long as the number
+%of parameters and features $p$ is less than the number of data points $n$ and the solution
+%is \citep{arXiv:2101.07256}
+%\begin{equation}
+%    \hat{\vec{w}}=(\vec{M}^{\intercal} \vec{M})^{-1} \vec{M}^{\intercal} \vec{y}
+%    \label{eq:ols_solution}
+%\end{equation}
+%under the assumption that $\vec{M}\vec{M}^\intercal$ is invertible which is usually the case.
+%Equation~\ref{eq:ols_solution} is called the \emph{ordinary least squares} (OLS) solution.
+%For a nontrivial data covariance matrix $\vec{C}$ Equation~\ref{eq:ols_solution} generalizes
+%to
+%\begin{equation}
+%    \hat{\vec{w}}=(\vec{M}^{\intercal}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\intercal}\vec{C}^{-1} \vec{y}
+%    \label{eq:wls_solution}
+%\end{equation}
+%which is the solution to the \emph{weighted least squares} (WLS) problem.
+%It is common to modify the least squares objective functions with the addition of a
+%\emph{regularization} term to prevent overfitting. For instance, we can modify the OLS
+%objective (Equation~\ref{eq:ols_objective}) by adding an \emph{L2 regularization} term:
+%\begin{equation}
+%    \hat{\vec{w}}=
+%    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2} + \lambda \left|\vec{w}\right|^2
+%    \label{eq:ols_objective_l2}
+%\end{equation}
+%where $\lambda >0$ is a regularization parameter
+%which penalizes large magnitudes of the coefficients $\vec{w}$. This is also a convex
+%optimization problem and the solution is
+%\begin{equation}
+%    \hat{\vec{w}}=(\vec{M}^{\intercal} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\intercal} \vec{y}
+%    \label{eq:ols_solution_l2}
+%\end{equation}
+%where $\vec{I}$ is the identity matrix. Regularized OLS regression with this particular
+%regularization term is also known as \emph{ridge regression} in the statistics
+%literature\footnote{One of the many things that make statistics confusing is that very
+%    similar methods often have special names wich obscures the big picture. The vast majority
+%    of statistical methods can be seen as a linear model with the data represented in a particular
+%    basis, and with a particular choice of the likelihood (noise model), priors (regularization)
+%    and  a method for obtaining the solution (usually a convex optimization problem).}
+%L2 regularization (as well as most other forms of regularization terms) can be seen
+%as a particular choice of priors for the regression coefficients in the Bayesian paradigm.
+%The above also naturally generalizes to the case of weighted least squares.
+%
+%\subsubsection{Overdetermined case ($p > n$)}
+%Contrary to popular belief within the astronomy community it is possible and
+%often desirable to have a model with more parameters than data points (see
+%Chapter~\ref{ch:mapping_io} for a real world example). In the overparametrized
+%case of linear regression $(p > n)$ there are multiple choices for $\vec{w}$
+%which perfectly fit the data. The solution in that case is defined to be a
+%minimum-norm parameter vector $\vec{w}$ that interpolates the data
+%($\vec{y}=\vec{M} \vec{w}$), out of possibly many degenerate solutions. The OLS
+%objective in Equation~\ref{eq:ols_objective} can be modified such that it is
+%valid for both the $p<n$ and the $p>n$ case by making use of a limit in which
+%an L2 regularization term tends to zero \citep{arXiv:2101.07256}:
+%\begin{equation}
+%    \hat{\vec{w}}=\lim _{\lambda \rightarrow 0^{+}}\left[
+%    \underset{\boldsymbol{w}}{\operatorname{argmin}}\left|\vec{y}-\vec{M} \vec{w}\right|^{2} + \lambda \left|\vec{w}\right|^2\right]\quad .
+%\end{equation}
+%In this case the OLS solution becomes
+%\begin{equation}
+%    \hat{\vec{w}}=\vec{M}^{\intercal} (\vec{M} \vec{M}^{\intercal})^{-1} \vec{y}
+%    \label{eq:ols_solution_overdet}
+%\end{equation}
+%Equations~\ref{eq:ols_solution} and \ref{eq:ols_solution_overdet} can be unified into a
+%single equation by making use of the \emph{Moore-Penrose pseudoinverse} $\vec{M}^\dagger$.
+%The pseudoinverse is defined by taking the singular-value decomposition (SVD)
+%of $\vec{M}$:
+%\begin{equation}
+%    \vec{M}\equiv\vec{U} \vec{S} \vec{V}
+%\end{equation}
+%Then $\vec{M}^\dagger =  \vec{V}^{\intercal} \vec{S}^{\dagger} \vec{U}^{\intercal}$.
+%
+%\subsection{Bayesian linear regression}
+%In the Bayesian paradigm we can write down the posterior over the regression
+%weights $\vec{w}$ assuming a Gaussian likelihood and a Gaussian prior over the
+%weights. The posterior is then
+%\begin{align}
+%    p(\vec{w}|\vec{y}) & = \frac{p(\vec{y}\lvert \vec{w})\,p(\vec{w})}{\int p(\vec{y}\lvert \vec{w})p(\vec{w}) \ud\vec{w}}                                                                                                                                                 \\
+%                       & = \frac{    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{w}, \vec{C}) \,\mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda})}{\int \mathcal{N}(\vec{y} \lvert\vec{M} \vec{w}, \vec{C}) \,\mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda}) \ud\vec{w}}
+%    \label{eq:bayes_lin_regression}
+%\end{align}
+%where $\vec{C}$ is the data covariance matrix as before, $\vec{\mu}$ is the mean
+%of the Gaussian prior on coefficients and $\vec{\Lambda}$ is the the prior covariance matrix.
+%Using the fact that the product of two multivariate Gaussian distributions is also a
+%multivariate Gaussian distribution, one can show that
+%the solution to Equation~\ref{eq:bayes_lin_regression} is
+%\citep[see for example][]{arXiv:2005.14199}
+%\begin{equation}
+%    p(\vec{w}|\vec{y}) =\mathcal{N}(\vec{w} \lvert\vec{a}, \vec{A})
+%    \label{eq:linear_regression_posterior}
+%\end{equation}
+%where
+%\begin{align}
+%     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\intercal} \vec{C}^{-1} \vec{M}                                                     \\
+%     & \vec{a}     =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\intercal} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
+%\end{align}
+%Since all \ssf{starry} models for occultation light curves are linear
+%(Equation~\ref{eq:starry_linear_model}) we can use Equation~\ref{eq:linear_regression_posterior}
+%to directly obtain the posterior over the map coefficients $\mathbf{y}$ but the cost
+%of doing so is imposing a Gaussian prior on the map coefficients.
+%
+%\subsection{Marginalizing a likelihood over linear parameters}
+%A model structure which is quite common in astronomy and physics is a model
+%which depends on some linear, and some nonlinear parameters. This is the
+%structure of the equation for the observed flux in microlensing
+%(Equations~\ref{eq:flux_observed_micro} or \ref{eq:flux_dominik}) where the
+%flux parameters are linear and the magnification parameters are nonlinear. The
+%likelihoods for such models can be written as
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{w}, \vec{\theta})
+%\end{equation}
+%where $\vec{w}$ are the linear parameters and $\vec{\theta}$ are the nonlinear parameters.
+%In some circumstances, it is possible to analytically marginalize over the linear
+%parameters $\vec{w}$. Doing so is often advantageous because it reduces the size of the
+%parameter space  and the computational cost of doing inference.
+%The marginalized likelihood is given by (Equation~\ref{eq:marginalization})
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{\theta})=\int p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta}) \ud\vec{w}
+%\end{equation}
+%Marginalization requires the introduction of a prior  $p(\vec{w}\lvert \vec{\theta})$
+%which can potentially also depend on the nonlinear parameters $\vec{\theta}$.
+%The alternative to doing this integral is re-factorizing the expression in the integrand
+%using Bayes' theorem:
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta})  = p(\vec{w}\lvert \vec{y},\vec{\theta})\,p(\vec{y}\lvert \vec{\theta})
+%\end{equation}
+%which yields the marginalized likelihood without us having to do the integral:
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{\theta}) = \frac{    p(\vec{y}\lvert \vec{w}, \vec{\theta})p(\vec{w}\lvert \vec{\theta})}{p(\vec{w}\lvert \vec{y},\vec{\theta})}
+%    \label{eq:likelihood_factorization}
+%\end{equation}
+%If both the prior and the likelihood are multivariate Gaussians with the following form:
+%\begin{align}
+%    p(\vec{y}\lvert \vec{w}, \vec{\theta}) & = \mathcal{N}(\vec{y}\lvert \vec{M}\vec{w},\vec{C})  \\
+%    p(\vec{w}\lvert \vec{\theta})          & = \mathcal{N}(\vec{w}\lvert \vec{\mu},\vec{\Lambda})
+%\end{align}
+%then Equation~\ref{eq:likelihood_factorization} becomes
+%\citep[for the proof see][]{arXiv:2005.14199}
+%\begin{equation}
+%    \mathcal{N}(\vec{y} \lvert \vec{M} \vec{w}, \vec{C})\, \mathcal{N}(\vec{w} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{w} \lvert\vec{a}, \vec{A})\, \mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
+%\end{equation}
+%where
+%\begin{align}
+%     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\intercal} \vec{C}^{-1} \vec{M}                                                     \\
+%     & \vec{a}     =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\intercal} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
+%     & \mathbf{B}=\mathbf{C}+\vec{M} \vec{\Lambda} \vec{M}^{\intercal}                                                                 \\
+%     & \vec{b}=\vec{M} \vec{\mu}
+%\end{align}
+%Thus the marginalized likelihood is
+%\begin{equation}
+%    p(\vec{y}\lvert \vec{\theta}) = \mathcal{N}(\vec{y} \lvert \vec{b}, \vec{B})
+%    \label{eq:likelihood_lin_params_marginalized}
+%\end{equation}
+%
+%\subsubsection{Application to microlensing}
+%Let's apply this trick to the microlensing models
+%(Equation~\ref{eq:flux_dominik}). In the general case, the dataset consists of
+%$J$ flux vectors $\vec{f}_j$, each of lengths $N_j$ where $j$ indexes different
+%spectral bands, often from different observatories as well.
+%Equation~\ref{eq:flux_dominik} can then be written in vector form as
+%\begin{equation}
+%    \vec{f}_j=\vec{M}_j\vec{w}_j
+%    \label{eq:microlensing_flux_general}
+%\end{equation}
+%where the matrix $\vec{M}$ is defined as
+%\begin{equation}
+%    \vec{M}_j\equiv \begin{pmatrix}
+%        \tilde{A}(t_1;\vec{\theta})     & 1      \\
+%        \tilde{A}(t_2;\vec{\theta})     & 1      \\
+%        \vdots                          & \vdots \\
+%        \tilde{A}(t_{N_j};\vec{\theta}) & 1
+%    \end{pmatrix}
+%\end{equation}
+%and $t=t_1,\ldots, t_{N_j}$ are the times of observations in band $j$,
+%$\tilde{A}(t_i) \equiv (A(t_i)-1) /(A\left(t_{0}\right)-1)$, $\vec{\theta}$ are the
+%nonlinear parameters, and the weights vector $\vec{w}_j$ is defined as
+%\begin{equation}
+%    \vec{w}_j\equiv    \left( \Delta F_j\; F_{\mathrm{base},j}\right)^\intercal
+%\end{equation}
+%Equation~\ref{eq:microlensing_flux_general} is general enough to encompass
+%either single or multiple lens microlensing models. Assuming independence of
+%noise properties between observations in different spectral bands\footnote{This
+%    is certainly true if we have observations from different telescopes. It is not
+%    strictly true if the observations are from the same telescope in different
+%    spectral bands but it is often a good enough assumption.} the likelihood for
+%the complete dataset
+%$\mathcal{D}=\left\{f_{j}\right\}_{j=1}^{J}$ factorizes as
+%\begin{equation}
+%    p(\mathcal{D}\lvert \vec{w}, \vec{\theta})=\prod_{j=1}^J p(\vec{f}_j\lvert \vec{w}_j, \vec{\theta})
+%\end{equation}
+%Using Equation~\ref{eq:likelihood_lin_params_marginalized} we can then write down the
+%negative log-likelihood (ignoring constants) for the $j$-th band as
+%\begin{align}
+%    \operatorname{NLL}(\vec{\theta}) & =\frac{1}{2}\left( \vec{f}_j-\vec{M}_j\vec{\mu}\right)^\intercal\left(\vec{C}_j + \vec{M}_j\vec{\Lambda}\vec{M}_j^\intercal\right)^{-1}
+%    \left( \vec{f}_j-\vec{M}_j\vec{\mu}\right)                                                                                                                                 \\
+%                                     & + \frac{1}{2}\ln\left| \vec{C}_j + \vec{M}_j\vec{\Lambda}\vec{M}_j^\intercal\right|
+%    \label{eq:ll_marginalized_microlensing}
+%\end{align}
+%where we have assumed a common mean $\vec{\mu}$ and covariance $\vec{\Lambda}$ for the
+%prior for the linear flux parameters.
+%To avoid computing an inverse and a determinant of $J$ matrices with shape $N_j\times N_j$,
+%we can simplify the terms in Equation~\ref{eq:ll_marginalized_microlensing} using the
+%matrix inversion lemma \citep[see for example Appendix A3 in][]{rasmussen2006}
+%\begin{align}
+%     & \left(\mathbf{C}_j+\vec{M}_j \boldsymbol{\Lambda} \vec{M}_j^{\intercal}\right)^{-1}=\mathbf{C}_j^{-1}-\mathbf{C}_j^{-1} \vec{M}_j\left(\boldsymbol{\Lambda}^{-1}+\vec{M}_j^{\intercal} \mathbf{C}^{-1} \vec{M_j}\right)^{-1} \vec{M}_j^{\intercal} \mathbf{C}_j^{-1} \\
+%     & \ln\left|\mathbf{C}_j+\vec{M}_j \vec{\Lambda} \vec{M}_j^{\intercal}\right|=\ln|\mathbf{C}_j| +\ln|\vec{\Lambda}| + \ln\left|\boldsymbol{\Lambda}^{-1}+\vec{M}_j^{\intercal} \mathbf{C}_j^{-1} \vec{M}_j\right|
+%\end{align}
+%In most cases we can assume that $\vec{\mu}=\vec{0}$, $\vec{\Lambda}$ is diagonal, and
+%in absence of correlated noise all matrices $\vec{C}_j$ are also diagonal so the expression
+%in Equation~\ref{eq:ll_marginalized_microlensing} isn't usually expensive to compute.
+%
+%Although Equation~\ref{eq:ll_marginalized_microlensing} is the formally correct
+%way to marginalize the linear parameters of a general microlensing model, it is
+%not used in the microlensing literature. What most people do instead is using
+%MLE to maximize the joint likelihood $p(\vec{f}\lvert\vec{w},\vec{\theta})$
+%with respect to the linear parameters by solving a least squares problem at
+%every MCMC step in the nonlinear parameters (we'll cover MCMC in the next
+%section). This procedure is approximately equivalent to using the marginalized
+%likelihood from Equation~\ref{eq:ll_marginalized_microlensing} with two
+%important caveats. The first it that one shouldn't optimize for the maximum
+%likelihood flux parameters conditional on nonlinear parameters, but rather the
+%MAP values of those parameters under Gaussian priors (although, the difference
+%between the MLE and MAP estimates is negligible under broad priors). The second
+%caveat is that the optimization can be cast as a linear least squares problem
+%under the assumption that the covariance matrix $\vec{C}$ for the data is
+%diagonal. This is a fairly restrictive assumptions because it implies that
+%there is no correlated noise in the light curve which isn't true for most real
+%world microlensing datasets.
+%
+%Although we have marginalized the likelihood over the linear parameters which
+%substantially reduces the complexity of the inference problem we can still
+%compute the posterior distribution for those parameters \emph{conditional on
+%    the value of the nonlinear parameters}:
+%\begin{equation}
+%    p(\vec{w}\lvert \vec{y}, \vec{\theta})=\mathcal{N}(\vec{w}\lvert \vec{y}, \vec{\theta})
+%\end{equation}
+%which we can use to obtain the full posterior over the linear parameters if we happen
+%to have samples from the posterior over the nonlinear parameters $p(\vec{\theta}|\vec{y})$
+%(obtained using MCMC for example).
+%
+%\subsection{Gaussian processes}
+%% https://peterroelants.github.io/posts/gaussian-process-tutorial/
+%\begin{figure}[t]
+%    \begin{centering}
+%        \includegraphics[width=\linewidth]{figures/gp_samples.pdf}
+%        \caption{A few samples from a Gaussian process prior with a squared exponential
+%            kernel function for two different values of the length scale parameter $l$.}
+%        \label{fig:gp_samples}
+%    \end{centering}
+%\end{figure}
+%A final topic that we need to cover in this section is that of \textsl{Gaussian
+%    Processes}. In Section~\ref{ssec:least_squares} we have discussed linear
+%regression with $p$ components (basis functions) and we have mentioned that one
+%fit many more components than there are data points and that these models can
+%be quite useful. A Gaussian process (GP) is what happens when let the number of
+%components $p$ go to infinity\footnote{See \citet{arXiv:2101.07256} for a
+%    demonstration of this point.} which is why they belong to a class of methods
+%called \textsl{nonparametric regression} methods, although a better name would
+%perhaps be infinite-parameter regression methods. This is not the only
+%formulation of GPs. GPs can also be seen as \emph{prior distributions over the
+%    space of functions} such that each sample from that GP prior is a function.
+%More formally, \emph{a Gaussian process is a collection of random variables,
+%    any finite number of which have a joint multivariate Gaussian
+%    distribution} \citep{rasmussen2006}.
+%
+%GPs are completely specified by a \textsl{mean function} $\mu (x)$ and a
+%positive-definite \textsl{covariance function} $k(x,x^\prime)$ which are
+%functions of the input locations $x\in \mathbb{R}^D$ (for example, a time
+%coordinate):
+%\begin{align}
+%    \mu (x)        & =\mathbb{E}[f(x)]                                                    \\
+%    k(x, x^\prime) & =\mathbb{E}\left[(f(x) -\mu (x))(f(x^\prime) -\mu (x^\prime))\right]
+%\end{align}
+%where $x$ and $x^\prime$ are any two different points in the input space.
+%We write
+%\begin{equation}
+%    f(x) \sim \mathcal{G} \mathcal{P}\left(\mu(x), k\left(x, x^{\prime}\right)\right)
+%\end{equation}
+%Usually the mean function $\mu(x)$  is taken to be zero for simplicity and all the
+%magic happens in the covariance function $k(x,x^\prime)$ which describes
+%\emph{how  any two points in the input space related to each other}.
+%
+%One very popular choice for the kernel function is the \textsl{squared
+%    exponential} kernel:
+%\begin{equation}
+%    k(x, x^\prime) = \exp\left(-\frac{1}{2}\frac{\left|x-x^\prime\right|^2}{l^2}\right)\hquad .
+%    \label{eq:sqexp}
+%\end{equation}
+%The above choice of a covariance function implies that the covariance of the outputs
+%$f(x)$ evaluated at any two points  will decay exponentially with the square of
+%the distance between the two points.
+%In other words, function values evaluated at points close to each tend to go hand in hand
+%while function values evaluated at points far away from each other are practically
+%independent.
+%The parameters $l$ sets the length scale which determines how quickly the covariance decays
+%to zero with increasing distance between the two points.
+%Figure~\ref{fig:gp_samples} shows samples from a GP prior with the squared exponential
+%kernel function in the one-dimensional case for two choices of the length sale parameter
+%$l$. One special property of this kernel is that it is \textsl{stationary}, meaning
+%that the mean and the variance of the samples from the GP prior don't depend on the
+%absolute value of the input coordinate $x$, they are the same everywhere because
+%the covariance only depends on the difference between any two points $x$ and
+%$x^\prime$\footnote{Stationarity is also a general property of time series.
+%    A stationary  time series is one whose properties (mean, variance) do not depend
+%    on the time at which the series is observed. A GP with a stationary kernel is a very flexible
+%    model for stationary time series.}. Not all kernels are stationary. An example of
+%a non-stationary kernel would be a kernel similar to the squared exponential but with
+%a variance growing linearly with time.
+%Kernels can be combined in any number of ways through for example addition and
+%multiplication and the result is also a valid kernel.
+%\href{https://www.cs.toronto.edu/~duvenaud/cookbook/}{This} pages provides a
+%nice visualization of different kernels and their combinations.
+%
+%The key property of a GP is that for any finite subset
+%$\vec{X}=\left\{\mathbf{x}_{1} \ldots \mathbf{x}_{n}\right\}$ of the domain of
+%$x$ (that is, any marginal distribution of $x$) is a multivariate Gaussian:
+%\begin{equation}
+%    f(\vec{x})\sim \mathcal{N}\,(\vec{\mu}(\vec{X}), \vec K(\vec{X}, \vec{X}))
+%\end{equation}
+%where $\vec K$ is a covariance matrix of shape $n\times n$ which is constructed by
+%evaluating  a kernel such as the one defined in Equation~\ref{eq:sqexp} elementwise.
+%This is precisely how we plotted
+%the functions in Figure~\ref{fig:gp_samples}, we simply chose a dense number of linearly
+%spaced grid points, computed the covariance matrix $K$ and then sampled the multivariate
+%Gaussian distribution.
+%\subsubsection{Predictions from posterior distribution}
+%\begin{figure}[t]
+%    \begin{centering}
+%        \includegraphics[width=\linewidth]{figures/gp_samples_conditioned.pdf}
+%        \caption{Samples from a Gaussian process with squared exponential kernel $l=1$
+%            conditioned on observations simulated from a simple sinusoid.
+%            The left panel shows the case when we assume that
+%            the observations are noiseless in which case the GP perfectly interpolates the data.
+%            The right panel shows the case when we assume that the observations are generated
+%            from the GP with the addition of some white Gaussian noise with variance $0.1^2$.}
+%        \label{fig:gp_samples_conditioned}
+%    \end{centering}
+%\end{figure}
+%We can go a step beyond drawing samples from the GP prior and condition the GP
+%on a set of observations in order fit the GP to data. Let $\vec{f}(\vec{X})$ be
+%the vector of observed data points evaluated at $\vec{X}$ with length $n_1$. If
+%we are interested in predicting the function values $\vec{f}^*(\vec{X}^*)$ at a
+%new vector of data points $\vec{X}^*$ with length $n_2$ (so we could, for
+%example, make a plot), then we need to compute the posterior distribution
+%$p(\vec{f}^*\lvert \vec{f}, \vec{X},\vec{X}^*)$. In order to compute the
+%conditional distribution, we start with the join distribution over $\vec{f}$
+%and $\vec{f}^*$. Since $\vec{f}$ and $\vec{f}^*$ both come from the same
+%multivariate Gaussian distribution, the joint distribution $p(\vec{f},
+%    \vec{f}^*)$ is also a multivariate Gaussian:
+%\begin{equation}
+%    \begin{pmatrix}
+%        \vec{f} \\
+%        \vec{f}^*
+%    \end{pmatrix} \sim \mathcal{N}\left[\begin{pmatrix}
+%            \vec{\mu} \\
+%            \vec{\mu}^*
+%        \end{pmatrix},\begin{pmatrix}
+%            \vec K(\vec{X},\vec{X})    & \vec K(\vec{X},\vec{X}^*)   \\
+%            \vec K(\vec{X}^*, \vec{X}) & \vec K(\vec{X}^*,\vec{X}^*)
+%        \end{pmatrix}\right]
+%\end{equation}
+%From this joint distribution, we can obtain the conditional  distribution
+%\citep[see appendix A2 in ][]{rasmussen2006} which is also a multivariate Gaussian\footnote{
+%    This is one of many special properties of Gaussians.
+%}:
+%\begin{align}
+%    p(\mathbf{f}^{*} \lvert \vec X^{*}, \vec X, \mathbf{f} )\sim \mathcal{N}( & \vec K\left(\vec X^*, \vec X\right) \vec K(\vec X, \vec X)^{-1} \mathbf{f}                                                          \\
+%                                                                              & \vec K\left(\vec X^*, \vec X^*\right)  -\vec K\left(\vec X^*, \vec X \right) \vec K(\vec X, \vec X)^{-1} \vec K (\vec X, \vec X^*))
+%    \label{eq:gp_posterior}
+%\end{align}
+%
+%Samples from this posterior GP distribution given by
+%Equation~\ref{eq:gp_posterior} with a squared exponential kernel conditioned on
+%a set of simulated observations $\vec{f}$ generated from a simple sinusoid are
+%shown in the left panel of Figure~\ref{fig:gp_samples_conditioned}. Each draw
+%from the distribution perfectly interpolates all the data points. The variance
+%is naturally larger where the data is sparse and the variance of the
+%predictions grows rapidly in the extrapolative regime outside of the range of
+%the data. The squared exponential kernel is not a great choice here because the
+%covariance rapidly decays to zero with the distance from the first and last
+%data point so the predictions will look flat futher away from the range of the
+%data. A much better choice would be a kernel with a periodic covariance
+%structure which would not have this problem.
+%
+%The predictions shown in the left panel assume that the observations are
+%noisless which is why each sample perfectly interpolates the data. It is more
+%common to assume that there is an additional white Gaussian noise component in
+%the observed data. In that case we add a diagonal matrix (because white noise
+%is independent Gaussian noise) to the covariance matrix
+%$\vec{K}(\vec{X},\vec{X})$ and we have
+%\begin{equation}
+%    \vec{K}(\vec{X},\vec{X}) + \sigma^2\vec{I}
+%\end{equation}
+%where $\vec{I}$ is the identity matrix. Predictions from a GP with this additional
+%noise are shown in the right panel of Figure~\ref{fig:gp_samples_conditioned}.
+%
+%\subsubsection{Fitting for the parameters of a GP kernel}
+%We shown how to sample a GP prior and how to compute predictions from a GP
+%conditioned on a set of observations but we haven't discssed the parameters
+%which define the kernel function, such as the length scale $l$ in the squared
+%exponential kernel. These parameters are often called the
+%\textsl{hyperparameters} of the kernel. Given a set of observations $\vec{f}$
+%evaluated at $\vec{X}$ with length $n$, the likelihood of a GP model for the
+%kernel hyperparameters $\vec{\theta}$ is
+%\begin{equation}
+%    p(\vec{f}\lvert \vec{\theta})\sim\,
+%    \mathcal{N}(\vec{f}\lvert \vec{\mu}(\vec{\theta}),
+%    \vec{K}(\vec X, \vec X; \vec\theta))
+%    \label{eq:gp_marginal_likelihood}
+%\end{equation}
+%where $\vec{\theta}$ is a set of parameters which define the kernel function. We
+%can treat these kernel parameters as we would any other parameters, for example we
+%can optimize the likelihood in Equation~\ref{eq:gp_marginal_likelihood} and find the
+%MLE parameters $\hat{\vec{\theta}}_\mathrm{MLE}$. The likelihood is not linear in these
+%hyperaparamters and the the optimization problem is generally not convex.
+%
+%The major challenge with GPs and the reason they still aren't as popular as for
+%instance linear regression is that sampling the GP prior or computing the
+%marginal likelihood (Equation~\ref{gp:marginal_likelihood}) requires an
+%inversion of a dense $n\times n$ covariance matrix, an operation which scales
+%as $\mathcal{O}(n^3)$. This operation becomes prohibitively expensive for
+%datasets with a few thousand data points. Thankfully, there has been a lot of
+%work on faster algorithms for inverting GP covariance matrices. Most of these
+%algorithms rely on the assumption of some special structure for the covariance
+%matrix which can be exploited to speed up the linear algebra operations. The
+%most notable example is the \ssf{celerite} algorithm
+%\citep{2017AJ....154..220F} which enables the use of 1D GPs with linear
+%$\mathcal{O}(n)$ scaling as long as we restrict ourselved to kernels which are
+%sums of complex exponential functions. These kernels are sufficiently flexible
+%for modeling physical time-series and the sums of complex exponentials also
+%have a physical interpretation as mixtures of stochastically drive, dampled
+%simple harmonic oscillators. Thanks to \ssf{celerite} it is possible to fit
+%GPs to astronomical light curves with thousands of datapoints.
+%
+%\subsubsection{Use of GPs in astronomy}
+%The use of Gaussian process models within astronomy has skyrocketed in recent
+%years. GPs within astronomy are by far most commonly used to model correlated
+%noise in astronomical light curves. Practically, this is a small step from just
+%assuming independent Gaussian noise. All we have to do is to replace a
+%likelihood with a diagonal covariance matrix whose entries are given by the
+%variance estimates for individual data points (``errorbars'') with a dense
+%covariance matrix whose generated using some kernel function (while still
+%including the independent noise term). The mean of the multivariate Gaussian
+%likelihood is usually an astrophysical model such as a transit, a microlensing
+%flux prediction or a secondary eclipse prediction. In some cases, the mean is
+%set to zero and the hyperparameters of the kernel have physical meaning. For
+%instance, GPs have been used to model stellar rotation periods using a
+%quasi-periodic kernel specified with multiple parameters which describe the
+%stochastic variability of the star \citep{2018MNRAS.474.2094A}.
+%\citet{2021AJ....162..124L} went a step beyond, starting with properties of
+%distributions of stellar spots expressed in the \ssf{starry} formalism they
+%derived a closed-form expression for a GP which describes a light curve of a
+%rotating, evolving stellar surface conditioned on a distribution of stellar
+%spot sizes, contrasts and latitudes.
+%
+%The advantage of GP models relative to alternatives such as linear regression
+%with a fixed number of basis functions is that they are far more flexible and
+%allow us to make less rigid assumptions about the phenomenon of interest. This
+%section was just a brief summary of their key properties and there is much more
+%to GPs that we haven't discussed (for example, fitting GPs to multi-dimensional
+%data). In Chapter~\ref{ch:mapping_exoplanets} we will show two very different
+%use cases of GPs in the context of occultation mapping.
+%
+%\section{Statistical inference - computation}
+%\label{sec:inference_practice}
+%In the previous section I briefly summarized some key aspects of probability theory
+%and Bayesian statistics, I discussed maximum likelihood estimation and
+%linear models and GPs for which inference can be done analytically.
+%In most most cases, models of real world interest are not so well behaved and maximum
+%likelihood estimation or obtaining samples from the posterior
+%distribution becomes very challenging because of complex likelihood and posterior
+%landscapes. In those cases computation needs goes hand-in-hand with model building and it
+%is very important to understand how the strucuture of the model and the data maps to the
+%structure of the likelihood.
+%I start by introducing some very fundamental
+%concepts such as the \emph{curse of dimensionality} and the \emph{no free lunch theorems}.
+%Next, I discuss various computational inference methods such
+%as optimization methods, Markov chain Monte Carlo and Variational Inference.
+%Finally, I discuss model comparison through Bayes factors and cross validation.
+%
+%\subsection{The curse of dimensionality and the no-free-lunch theorem}
+%The \emph{curse of dimensionality} refers to a phenomenon where the
+%\emph{volume} of \emph{possible} explanations for the data grows exponentially
+%with increasing dimension of the problem. Consider the following
+%example\footnote{ Taken from the excellent paper by \citet{arXiv:2104.00008}.
+%}. There are images consisting of $n$ pixels where each of the pixels is either
+%a black (0) or white (1). The number of possible images is $2^n$ which grows
+%exponentially. If each image out of the $2^n$ possible images is described by
+%one of two labels $A$, or $B$ then the total number of possible ways of
+%labeling the set of images is $2^{2^n}$ which is a stupidly large number. If
+%$n=9$, the number of ways to classify the set of all 9-bit black and white
+%images is greater than the number of atoms in the observable universe. If the
+%labels don't correlate with the properties of the images then one strategy for
+%solving this problem would be the lookup table approach -- memorize each image
+%with its label. The famous \emph{no-free-lunch theorem} (NFL) proved by the
+%mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
+%impossible to construct an algorithm which does better than the lookup table
+%approach.
+%
+%Of course the assumption that the labels aren't correlated to features in the
+%images is silly for any meaningful problem. For instance, all images of cats
+%have certain things in common as do images of galaxies. The no-free-lunch
+%theorem doesn't say that it is impossible to construct an algorithm to classify
+%images of cats, it only says that there is no best algorithm \emph{if we
+%    average over all possible inputs to the problem}. To construct an algorithm
+%which works in practice we need to impose structure on the problem, what
+%machine learning researchers call \emph{inductive bias}. A set of assumptions
+%which work in one domain may not work well in another domain and no single
+%model will work well in all domains.
+%
+%\subsection{Sampling vs. optimization}
+%%When fitting a model to a scientific dataset we are usually interested in
+%%finding some ``best'' estimate of the parameters of that model\footnote{This is
+%%    opposed to machine learning where in the vast majority of cases the goal is
+%%    \emph{prediction} and the parameters of the model (for instance, weights of a
+%%    neural network) are not of interest.}. In science, simply having an estimate of
+%%the parameters is not sufficient, we also require estimates of the
+%%\emph{uncertainty} associated with those values. In the Bayesian formalism this 
+%%means estimating the posterior distribution $p(\vec \theta \lvert \mathcal{D})$.
+%
+%Broadly speaking, in statistical inference we can differentiate between two
+%classes of methods for fitting models to data. Optimization methods cast the
+%problem of estimating the likelihood or the posterior as an (generally
+%non-convex) optimization problem. The goal in optimization problems is to find
+%a single point which minimizes or maximizes some scalar function called a
+%\textsl{cost function} or an \textsl{objective function}. The space of possible
+%cost functions is infinite but in probabilitic inference these will be (usually
+%unnormalized) probability distributions such as likelihood and posteriors. An
+%example would be MLE and MAP estimation, the Laplace approximation and
+%variational inference.
+%
+%Sampling methods are a fundamentally different class of algorithms whose
+%purpose is accurately computing multidimensional integrals such as with the
+%goal of numerically computing \emph{integrals} such as expectation values over
+%the distribution $p(\vec{\theta})$:
+%\begin{equation}
+%    \mu\equiv \mathbb{E}_{p(\vec \theta)}[g(\vec \theta)]=\int g(\vec \theta) p(\vec \theta) \ud \vec \theta\hquad ,
+%    \label{eq:general_expectation}
+%\end{equation}
+%where $g(\vec \theta)$ is some arbitrary function such as the mean.
+%These integrals are intractable when the dimension of $\vec{\theta}$ is greater than
+%a few because  the volume of the parameter space grows exponentially and finding
+%the the regions in parameter space where the integral is nonzero becomes ever harder.
+%We can get around this problem by generating samples from the probability distribution
+%$\vec{\theta}^{(s)}\sim p(\vec{\theta})$ and computing the integral using
+%\textsl{Monte Carlo Integration} in regions where there is non-negligible probability:
+%\begin{equation}
+%    \hat{\mu}_\mathrm{MC}=\mathbb{E}_{p(\vec \theta)}[g(\vec \theta)] \approx \frac{1}{S} \sum_{i=1}^{S} g\left(\vec \theta^{(s)}\right)\hquad .
+%    \label{eq:monte_carlo_estimator}
+%\end{equation}
+%where $\hat{\mu}_\mathrm{MC}$ is the simple Monte Carlo estimator of the expectation
+%$\mu$.
+%It can be shown that if the expectation $\mu$ exists, the estimator
+%$\hat{\mu}_\mathrm{MC}$ is consistent (it asymptotically converges toward $\mu$ with
+%increasing $S$) and unbiased (the mean of the sampling distribution of the estimator
+%is equal to $\mu$). If the expectation of $g^2$ is also finite the central limit
+%theorem holds and the error of the simple  Monte Carlo estimator decreases as
+%$\sqrt{S}$. This error is independent of the dimension!
+%The major challenge is how to generate those samples from $p(\vec{\theta})$. This
+%is usually accomplished
+%sampling numbers from probability distributions that are easy to sample from
+%(most commonly multivariate normal distributions) and then reweighting those
+%samples at each iteration of the algorithm so that they match the target
+%distribution. Examples of sampling methods include Markov chain Monte Carlo
+%(MCMC), importance sampling and rejection sampling.
+%
+%%The difference between optimization and sampling algorithms can often be fuzzy
+%%because sampling problems can sometimes be cast as optimization problems. In
+%%both cases the NFL theorem holds, meaning that no sampling or optimization
+%%method works best across all problems. The fundamental problem in both cases as
+%%that we can only evaluate the target density some finite number of times and
+%%the challenge is to find high probability regions in a multi-dimensional
+%%parameter space. Even in moderately high-dimensional spaces (on the order of 10
+%%dimensions) this is an extremely challening problem which is why methods which
+%%work well in low dimensions often fail spectacularly in high dimensions. Both
+%%classes of methods often use information about the gradient of target
+%%distribution to aid the exploration of the parameter space. In fact, optimizing
+%%or sampling high dimensional distributions is often impossible without access
+%%to (exacy) gradients of the target density.
+%%
+%\begin{figure}[t]
+%    \begin{centering}
+%        \includegraphics[width=\linewidth]{figures/typical_sets.pdf}
+%        \caption{Histograms showing the distance from the MAP point for a set of 
+%        samples from $D$-dimensional multivariate normal distribution with 
+%        zero mean and unit variance. With increasing number of dimensions the 
+%        probability mass concentrates in a thin shell centered at the MAP point 
+%        called the typical set. For $D=50$ the posterior samples are coming from 
+%        a region more than 5 sigma away from the MAP point.}     
+%               \label{fig:typical_sets}
+%    \end{centering}
+%\end{figure}
+%Contrary to what we might intuitively think, in all but the lowest dimensional models
+%samples from the posterior distribution are not located anywhere 
+%near a point estimate such as the MAP point in the parameter space, despite the fact 
+%that this is where the posterior density is highest.
+%This happens because expectation values  depend not only on the density of the pdf 
+%$p(\vec\theta)$ but also on the volume term 
+%$\ud \vec\theta$ and the value of the integral is non-negligible only when both are 
+%jointly maximized. In high dimensions the volume grows exponentially and most 
+%of the volume ends up being concentrated far away from the mode so that 
+%the product of the density and the  volume, the \textsl{posterior mass}, ends up 
+%being concentrated in a thin shell around the mode called the \textsl{typical set} 
+%\citep[for a good overview of this issue see][]{arXiv:1701.02434}.
+%This point is illustrated in Figure~\ref{fig:typical_sets} using samples from 
+%a $D$-dimensional multivariate normal distribution with zero mean and unit 
+%variance. 
+%The reason this happens is that  although the most likely points are located 
+%in the neighbourhood of the mode, 
+%there are many more points further away in the typical set and those are the 
+%ones that contribute to the integral involved in the calculation of expectation 
+%values.
+%For a toy example shown in Figure~\ref{fig:typical_sets} the fact this does not 
+%matter much because the posterior has spherical symmetry but  for non Gaussian
+%posteriors we can get very different results  when computing the mean (via 
+%sampling) and the MAP point (via optimization).
+%
+%
+%\subsection{Optimization methods}
+%In this section we provide a brief overview of two most relevant methods for
+%estimating the posterior distribution $p(\vec{\theta}\lvert \mathcal{D})$ using
+%optimization algorithms.
+%
+%\subsubsection{The Laplace approximation}
+%One of the simplest way of approximating a posterior distribution is the
+%\textsl{Laplace approximation}. We start with the general posterior density of
+%the form
+%\begin{equation}
+%    p(\vec{\theta}\lvert\mathcal{D})=\frac{1}{\mathcal{Z}}e^{-\mathcal{E}(\vec{\theta})}
+%\end{equation}
+%where $\mathcal{E}(\vec{\theta}) \equiv \ln\,p(\vec{\theta}, \mathcal{D})$
+%is called an \textsl{energy function} and $\mathcal{Z}$ is the fully marginalized
+%likelihood. Following \citet{murphy_book_2022}, we Taylor expand $\mathcal{E}(\vec{\theta})$ around the mode
+%$\hat{\vec{\theta}}$ as follows
+%\begin{equation}
+%    \mathcal{E}(\vec{\theta}) =\ln\,p(\vec{\theta}, \mathcal{D})\approx \mathcal{E}(\hat{\vec{\theta}})+(\vec{\theta}-\hat{\vec{\theta}})^{\intercal} \vec{g}+\frac{1}{2}(\vec{\theta}-\hat{\vec{\theta}})^{\intercal} \mathbf{H}(\vec{\theta}-\hat{\vec{\theta}})
+%\end{equation}
+%where $\vec{g}$ is the gradient at the mode (the Jacobian matrix), and $\vec{H}$
+%is the \textsl{Hessian} evaluated at the mode. The gradient term by definition
+%vanishes at the mode $\hat{\vec{\theta}}$ so we are left with
+%\begin{equation}
+%    \hat{p}(\vec{\theta}, \mathcal{D})=e^{-\mathcal{E}(\hat{\vec{\theta}})} \exp \left[-\frac{1}{2}(\boldsymbol{\theta}-\hat{\vec{\theta}})^{\intercal} \mathbf{H}(\vec{\theta}-\hat{\vec{\theta}})\right]
+%\end{equation}
+%from which it follows that
+%\begin{equation}
+%    \hat{p}(\vec{\theta} \lvert\mathcal{D})=\frac{1}{\mathcal{Z}}\, \hat{p}(\vec{\theta}, \mathcal{D})=\mathcal{N}\left(\vec{\theta} \lvert\hat{\vec{\theta}}, \mathbf{H}^{-1}\right)\hquad .
+%\end{equation}
+%Thus, we obtain the well known result that a Taylor expansion of an arbitrary
+%distribution  $p(\vec{\theta} \lvert\mathcal{D})$ around the mode $\hat{\vec{\theta}}$
+%truncated at the first non-vanishing term is simply a multivariate normal distribution
+%centered at the mode whose covariance matrix is given by the inverse Hessian (which specifies
+%the \emph{curvature} of the distribution) evaluated at the mode.
+%
+%Laplace approximation works very well with posteriors which are well described
+%by a multivariate normal distribution. It is straightforward to compute because
+%we can use an optimizer to find the MAP point and then estimate the Hessian
+%numerically using finite difference approximation for the gradient or use
+%automatic differentiation to obtain the exact Hessian (we'll cover automatic
+%differentiation in the next section).
+%
+%\subsubsection{Variational inference}
+%Another method which uses optimization to obtain an approximation to the true
+%posterior is \emph{Variational inference} (VI). The advantage of VI over the
+%Laplace approximation is that we can fit any distribution we like to the
+%posterior. The key idea is to model an intractable density $p(\vec{\theta}
+%    \lvert\mathcal{D})$ with some off-the-shelf distribution $q(\vec{\theta})$ such
+%that \emph{we minimize a measure of discrepancy between the two distributions}.
+%We optimize for the parameters of the distribution $q(\vec{\theta})$
+%$\vec{\psi}$ which are called the \textsl{variational parameters}. The most
+%common discrepancy measure is the KL divergence
+%(Equation~\ref{eq:kl_divergence})from $q$ to $p$. Thus, the optimization
+%problem is to find the parameters $\vec{\psi}^*$ which minimize the KL
+%divergence:
+%\begin{equation}
+%    \boldsymbol{\psi}^{*}=\underset{\boldsymbol{\psi}}{\operatorname{argmin}}\,D_{\mathbb{K} \mathbb{L}}(q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})\, \|\, p(\boldsymbol{\theta} \lvert\mathcal{D}))
+%    \label{eq:variational_inference}
+%\end{equation}
+%Making use of Equation~\ref{eq:kl_divergence}, we can expand the above equation as \citep{murphy_book_2022}
+%\begin{align}
+%    \vec{\psi}^* & =\underset{\vec{\psi}}{\operatorname{argmin}}\, \mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}\left[\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})-\ln \left(\frac{p(\mathcal{D} \lvert\boldsymbol{\theta}) p(\boldsymbol{\theta})}{p(\mathcal{D})}\right)\right]                           \\
+%                 & =\underset{\vec{\psi}}{\operatorname{argmin}} \,\underbrace{\mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}[-\ln p(\mathcal{D} \lvert\boldsymbol{\theta})-\ln p(\boldsymbol{\theta})+\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})]}_{-\mathrm{ELBO}(\boldsymbol{\psi})}+\ln p(\mathcal{D})
+%\end{align}
+%Since the evidence $\ln\,p(\mathcal{D})$ is not a function of the variational parameters
+%$\vec{\psi}$, the problem of solving Equation~\ref{eq:variational_inference} reduces
+%to maximizing the term
+%\begin{equation}
+%    \mathrm{ELBO}(\boldsymbol{\psi}) \equiv \mathbb{E}_{q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})}[\ln p(\mathcal{D} \lvert\boldsymbol{\theta})+\ln p(\boldsymbol{\theta})-\ln q(\boldsymbol{\theta} \lvert\boldsymbol{\psi})]
+%\end{equation}
+%which is called the \textsl{evidence lower bound} (ELBO) because
+%$D_{\mathbb{K} \mathbb{L}}(q \| p) \geq 0$ implies that
+%$\mathrm{ELBO}(\boldsymbol{\psi}) \leq \ln p(\mathcal{D})$.
+%The most common choice for the distribution $q$ is, unsurprisingly, a multivariate
+%Gaussian. VI with a multivariate Gaussian differs from  the Laplace aproximation
+%because  in VI we are optimizing for both the mean and the covariance matrix rather
+%than equating the covariance matrix to the inverse Hessian at the MAP point.
+%VI is most commonly used with very large datasets and high dimensional parameter
+%spaces because it is computationally cheaper than MCMC (although VI and MCMC are not
+%doing the same thing). It has seen little use within astronomy.
+%
+%\subsection{Sampling methods}
+%\label{ssec:sampling_methods}
+%\subsubsection{Rejection sampling}
+%Rejection sampling is one of the simplest algorithms for generating samples
+%from a probability distribution. Consider a normalized pdf
+%\begin{equation}
+%    p(\vec{\theta})=\frac{\tilde{p}(\vec{\theta})}{\mathcal{Z}}
+%\end{equation}
+%where $\tilde{\vec{\theta}}$ is the unnormalized pdf and $\mathcal{Z}$ is normalization
+%constant which isn't necessarily known.
+%If a proposal distribution $q(\vec{\theta})$ (which we know how to sample from)
+%satisfies the inequality
+%\begin{equation}
+%    Cq(\vec{\theta})\geq \tilde{p}(\vec{\theta})
+%    \label{eq:rejection_sampling_constraint}
+%\end{equation}
+%for some constant $C$, the function $Cq(\vec{\theta})$ provides an upper envelope for
+%$\tilde{p}$. We can use the proposal distribution to generate samples from the target
+%distribution using the following algorithm:
+%\begin{itemize}
+%    \item Sample $\vec{\theta}_0\sim q(\vec{\theta})$ from the proposal distribution.
+%    \item Sample $u_0\sim \mathrm{Unif}(0, Cq(\vec{\theta}_0))$ which corresponds to
+%          uniformly drawing a height under the envelope.
+%    \item If $u_0>\tilde{p}(\vec{\theta}_0)$ reject the sample, otherwise accept
+%\end{itemize}
+%See \citet{murphy_book_2023} for the proof that this procedure works.
+%If $\tilde{p}$ is a normalized target distribution the acceptence probability at
+%each iteration is  $1/C$ (TODO: show this) so  $C$ should be as small as possible
+%while satisfying the constraint in Equation~\ref{eq:rejection_sampling_constraint}.
+%In practice, if the distribution $p(\vec{\theta})$ represents a posterior distribution
+%$p(\vec{\theta}\lvert \mathcal{D})$ rejection sampling can be accomplished by using
+%the prior as the proposal distribution, sampling a very large number of parameter
+%vectors from the prior, computing the value of the likeliood
+%$p(\mathcal{D} \lvert\vec{\theta})$ for each and then setting $C$ to be equal to the
+%maximum value of the likelihood across all samples.
+%
+%The major drawback of rejection sampling is that it scales badly with
+%increasing dimensionality of the target distribution. For instance, if
+%$p(\vec{\theta})=\mathcal{N}(\vec{0}, \sigma_p^2\vec{I})$ and
+%$q(\vec{\theta})=\mathcal{N}(\vec{0}, \sigma_q^2\vec{I})$ we need to have
+%$\sigma_q^2\sigma_p^2>\sigma_p^2$ to satisfy the bound. In $D$ dimensions the
+%optimal value of $C$ is $C=(\sigma_q/\sigma_p)^D$ and since the acceptance
+%probability is $1/C$ this decreases exponentially fast with dimension $D$. If
+%the the proposal distribution $q(\vec{\theta})$ approaches $0$ in the tails at
+%a faster rate than $p(\vec{\theta})$ as $\vec{\theta}\rightarrow \infty$ than
+%their ratio $p(\vec{\theta})/q(\vec{\theta})$ approaches $\infty$ and the
+%constraint in Equation~\ref{eq:rejection_sampling_constraint} will not
+%satisfied. Thus, we must ensure that the target distribution does not have
+%heavier tails than the proposal distribution.
+%
+%Even though it is the simplest sampling algorithm, rejection sampling can be
+%very useful for problems with small numbers of parameters and small data sets
+%because it can deal with very pathological distributions (for example highly
+%multi-modal distributions). The most notable application of rejection sampling
+%in astronomy is the \ssf{The Joker} sampler \citep{2017ApJ...837...20P} for
+%sampling posterior pdfs for with very sparse binary-star and exoplanet radial
+%velocity measurements. \citet{2020ApJ...895....2P} apply this sampler to obtain
+%posterior pdfs in Keplerian parameters for $232 495$ sources from the APOGEE
+%catalog. Becuase these measurements are very sparse (a few data points per
+%star), the posterior pdf over the period prameter is highly multi-modal. The
+%brute-force approach of rejection sampling works very well and produces
+%independent samples from the distribution. For any individual object this pdf
+%doesn't tell us much, however, when the analysis is repeated for thousands of
+%objects in the framework of \textsl{hierarchical Bayesian inference} it is
+%nevertheless possible to obtain very tight constrain on the population-level
+%parameters. This is because even a highly multi-modal posterior pdf represents
+%a major contraction of the prior parameter space and thus excludes certain
+%parameter values.
+%
+%The radial veclocity model in \citet{2017ApJ...837...20P} interesting because
+%structurally it very closely models the single lens microlensing model with
+%parallax effects. Both models have few parameters 7 default parameters, two of
+%which are linear and in both cases the posterior can be multi-modal (see
+%Chapter~\ref{ch:microlensing_modeling} for more on this point). In
+%\citet{2017ApJ...837...20P} they use the marginalized likelihood over the
+%linear parameters and perform rejection sampling with the dimensionally reduced
+%model. However, this requires an extremely large number of prior samples, on
+%the order of hundreds of millions! Evaluating the likelihood this many times is
+%much more computationally expensive in the microlensing case primarily because
+%microlensing light curves have a lot more data points than stellar RV
+%measurements. Neverthless, as I will show in Chapter~\ref{ch:microlensing_modeling}
+%rejection sampling is intractable with the parallax microlensing model but it
+%is tractable with the classic Paczynski model with 3 nonlinear parameters
+%$(t_0, t_E, u_0)$.
+%
+%\subsubsection{Importance sampling}
+%\textsl{Importance sampling} is not a an algorithm for generating samples from
+%$p(\vec{\theta})$ but rather a method for estimating expectations
+%$\mathbb{E}_p[g(\vec{\theta})]$ given an existing set of samples from
+%some other distribution $q(\vec{\theta})$. We start by rewriting the integral 
+%in Equation~\ref{eq:general_expectation} as 
+%\begin{equation}
+%    \mu = \mathbb{E}_{p}[g(\vec \theta)]=\int g(\vec \theta)\,\frac{p(\vec \theta)}{q(\vec \theta)}\,q(\vec \theta) \ud \vec \theta\hquad ,
+%\end{equation}
+%where we have introduced some other distribution $q(\vec \theta)$.
+%The imporance sampling estimator of $\mu$ is then defined as
+%\begin{align}
+%    \hat{\mu}_{\mathrm{IS}}&=\frac{1}{S} \sum_{s=1}^{S} \frac{p\left(\boldsymbol{\theta}^{(s)}\right)}{q\left(\boldsymbol{\theta}^{(s)}\right)} g\left(\boldsymbol{\theta}^{(s)}\right)\\
+%    &=\frac{1}{S} \sum_{s=1}^{S} w^{(s)} g\left(\boldsymbol{\theta}^{(s)}\right)
+%\end{align}
+%where  $\boldsymbol{\theta}^{(s)} \sim q(\boldsymbol{\theta})$ are samples from the 
+%proposal distribution $q$ and $w^{(s)}$ are the \textsl{importance weights}. 
+%$w^{(s)}$  are the ratios between the target distribution $p(\vec \theta)$ and 
+%the proposal distribution $q(\vec \theta)$ evaluated at each $\vec{\theta}^{(s)}$.
+%As in the case of rejection sampling, the requirement for the proposal distribution $q$
+%is that it has support (is nonzero) wherever $p(\boldsymbol{\theta}) g(\boldsymbol{\theta})$
+%is nonzero.
+%Similarly to $\hat{\mu}_{\mathrm{MC}}$,  $\hat{\mu}_{\mathrm{IS}}$ is also a 
+%consistent and  unbiased estimator of $\mu$ but its variance is heavily dependent on 
+%how well the proposal distribution $q$ matches $p$. 
+%If we do not know the normalization constants of $p$ and $q$ we can instead use the 
+%self-normalized importance sampling (SNIS)  estimator 
+%\begin{equation}
+%    \hat{\mu}_{\mathrm{SNIS}}=\frac{\sum_{s=1}^{S} w^{(s)} h\left(\boldsymbol{\theta}^{(s)}\right)}{\sum_{s=1}^{S} w^{(s)}}
+%\label{eq:importance_sampling_estimator}
+%\end{equation}
+%which is also consistent but it has a small bias of order $\mathcal{O}(1/S)$.
+%There are also adaptive versions of importance sampling  which all rely on iteratively 
+%adapting the poposal distribution $q$ based on the importance weights at each iteration. 
+%
+%When the proposal distribution $q$ is a poor approximation of $p$ the distribution of 
+%importance weights can have a heavy right tail which results in a large and 
+%large and possibly infinite variance of $\hat{\mu}_{\mathrm{IS}}$.
+%This is especially true in high dimensions when the importance weights $w^{(s)}$ can 
+%vary by several orders of magnitude and the estimator in 
+%Equation~\ref{eq:importance_sampling_estimator} becomes dominated by a few draws from 
+%proposal distribution. One solution to this  problem is to fit a generalized 
+%Pareto distribution\footnote{Pareto distributions are commonly used in \textsl{Extreme 
+%Value Analysis} (EVA). EVA is a branch of statistics dealing with heavy and fat-tailed 
+%distributions, a regime where many classical methods built around Gaussian assumptions 
+%fall apart.} 
+%to the distribution of importance weights and then use that distribution to produce 
+%a uniformly spaced set of importance weights. This procedure is called 
+%\textsl{Pareto Smoothed Importance Sampling} (PSIS) \citep{arXiv:1507.02646}.
+%PSIS forms the basis of a very popular \ssf{R} package \ssf{loo} for 
+%high-dimensional leave-one-out cross validation \citep{arXiv:1507.04544}.
+%The PSIS algorithm also reports the estimate of the smoothness parameter in the
+%generalized Pareto distribution (\textsl{Pareto $\hat{k}$}). $\hat{k}$ acts as a 
+%diagnostic  because the PSIS estimate is likely to be unstable  if $\hat{k}$ is 
+%too large \citep{arXiv:1507.02646}.
+%
+%What if we are not just interested in computing expectations but also want to obtain 
+%posterior samples from $p(\vec{\theta})$ in order to, say, plot a historgram? We can 
+%use \textsl{importance resampling} which works by resampling the original samples 
+%$\vec{\theta}^{(1)} \dots \vec{\theta}^{(S)}$
+%with replacament with probabilities proportional to  the importance weights.
+%Practically, this meanse drawing a set of indices $i_1, \dots, i_S$ from a 
+%multinomial distribution  where the event probabilities are the normalized importance 
+%weights $\tilde{w}_{i}=w_{i} / \sum_{j} w_{j}$:
+%\begin{equation}
+%i_{1}, \ldots, i_{S} \sim \operatorname{Multinomial}\left(1, \tilde{w}_{1}, \ldots, \tilde{w}_{S}\right)\hquad .
+%\end{equation}
+%Samples with larger weights will appear more often than those with smaller weights. 
+%We can use Pareto smoothing when computing the weights. 
+%
+%In the context of astronomy, one  interesting (and relevant for the work presented 
+%in thesis) application of importance sampling is for simplifying Bayesian hierarchical 
+%inference. \citet{2010ApJ...725.2166H} showed that given posterior samples for a 
+%collection of objects (obtained using MCMC for example) which were obtained with some 
+%set of vague priors on the parameters of interest, we can re-weight those samples using 
+%importance resampling under a different hierarchical prior which depends on some 
+%population level parameters and in the process obtain a posterior over the population 
+%level parameters. \citet{2010ApJ...725.2166H} applied this method to infer parameters 
+%describing the eccentricity  distribution for a catalog of $N$ exoplanets given 
+%posterior samples for the eccentricity for each individual planet. I will describe 
+%this method in more detail in Chapter~\ref{ch:microlensing_modeling} and show how it can be
+%applied to model the distribution of the $t_E$ parameter for a catalog of microlensing 
+%events.
+%
+%\subsubsection{Rosenbluth-Metropolis-Hastings algorithm}
+%The most popular class of algorithms used for generating samples 
+%from a general pdf are 
+%\textsl{Markov Chain Monte Carlo} algorithms or MCMC for short. The simplest 
+%MCMC algorithm and the very first one that was developed is the
+%\textsl{Rosenbluth-Metropolis-Hasting algorithm} (RMH) (also known as the 
+%\textsl{Metropolis algorithm}). The RMH algorithm is one of 
+%the most famous algorithms developed in the 20th century. The original algorithm was 
+%published in \citet{1953JChPh..21.1087M} and it was extended by 
+%\citet{1970Bimka..57...97H}. The original version is extremely simple. 
+%Given a most recent sample $\vec\theta^{(s)}$, to generate the next sample
+%\begin{itemize}
+%    \item Draw a sample $\vec{\theta}^\prime$ from a proposal distribution $q(\bm\theta^\prime\lvert\vec\theta^{(s)})$ 
+%    \item Draw a uniform random number $r\sim\mathcal{U}(0,1)$
+%    \item If $\frac{p(\bm\theta^\prime)}{p(\vec\theta^{(k)})}$ then $\bm\theta^{(k+1)}\leftarrow \bm\theta^\prime$, else $\bm\theta^{(k+1)}\leftarrow\bm\theta^{(k)}$
+%\end{itemize}
+%The main characteristic of this algorithm is that it defines a biased random walk
+%through the parameter space (usually moving in only one parameter at the time) 
+%in such that the amount of time spent at a particular location $\vec\theta$ is 
+%proportional to the target density $p(\vec\theta)$.
+%These samples can then be used to approximate some expectation value of interest 
+%by means of Equation~\ref{eq:monte_carlo_estimator}. The samples are always correlated 
+%to en extent so estimates of the expectations don't converge as quickly with increasing 
+%number of samples as they would if the samples were independent.
+%
+%The reason this algorithm works is because it defines a process called a 
+%\textsl{Markov Chain} in which the probability of moving from a current state to a
+%a successive state in the chain depends only on the current state and not any
+%past states.
+%Markov chains have stationary probability distributions over states and the stationary
+%distribution for the Markov chain defined by the RMH algorithm turns out to be
+%the target distribution $p(\bm\theta)$.
+%For this to happen, the proposal distribution $q$ must satisfied a property
+%called \textsl{detailed balance} which is defined as
+%\begin{equation}
+%    q(\bm\theta^\prime\lvert\bm\theta)=q(\bm\theta\lvert\bm\theta^\prime)\hquad .
+%\end{equation}
+%That is, the proposal has to be reversible so that the probability of going in either 
+%direction is the same. 
+%
+%The proposal distribution $q$ is one of the main tunable parameters
+%of the RMH algoritm.
+%The simplest choice for $q$ is a multivariate Gaussian distribution
+% $q(\bm\theta^\prime\lvert\bm\theta)\sim\mathcal{N}(\bm\theta,\vec \Sigma)$.
+% The covariance matrix $\vec \Sigma$ sets a characteristic 
+%\textsl{step size} in the parameter space. If the step size is
+%too large most proposals will get rejected and the chain might miss regions with high
+%probability mass.
+%If on the other hand the step size is too small, most steps are likely to be accepted
+%but we might not explore the parameter space fully.
+%RMH (and MCMC samplers in general) don't converge to the stationary distribution  
+%of the target distribution immediately, instead there is always an initial exploration 
+%phase before the chain reaches the relevant parts of the parameter space 
+%(typical set).  The time it 
+%takes to reach stationarity is called the \textsl{mixing time} of the sampler 
+%or the \textsl{burn-in} period.
+%
+%All MCMC algorithms have the same goal, namely, generating samples from a
+%target pdf such that the least number of samples $S$ are needed to accurately
+%approximate the integral in Equation~\ref{eq:general_expectation}. Metrics
+%used to judge the quality of different samplers include 
+%\begin{itemize}
+%    \item Efficiency -- what is the acceptance rate of the proposals?
+%    \item Coverage -- is the target distribution explored completely?
+%    \item Correlation -- how correlated are individual samples with past samples?
+%    \item High dimensions -- can the sampler handle high dimensional spaces?
+%    \item Derivatives -- does the sampler require gradients of the target density? 
+%\end{itemize}
+%In real world problems MCMC samplers require very careful tuning and initalization 
+%to work well. Although RMH and other MCMC samplers should in principle converge to the posterior distribution 
+%independent of the initialization point in the parameter space and the particular 
+%choice of parametrization; in practice intialization, careful tuning of the proposal 
+%distribution and how we parametrize the model is crucial for the performance 
+%of the sampler. It is not an usual that a minor reparametrization of a given model leads 
+%to an increase in sampling efficiency of several orders of magnitude\footnote{For this
+%reason it is often much better to invest time into making a given model more friendly 
+%to MCMC than to improve the evaluation time of the model by a factor of a few.}.
+%
+%
+%\subsubsection{Hamiltonian Monte Carlo}
+%Hamiltonian Monte Carlo (HMC) \citep{1987PhLB..195..216D,arXiv:1206.1901} is a 
+%variant of  Markov Chain Monte Carlo which uses the gradient of the target 
+%density to substantially  improve the sampling efficiency compared to 
+%gradient-free samplers such as the RMH sampler, especially in high dimensions.
+%The gradient information is especially useful for high-dimensional densities 
+%because it enables the sampler to better 
+%explore the typical set of parameter space by making proposals along the 
+%the thin shell comprising the typical set\footnote{In high dimensions and 
+%petrurbation away from the thin shell of the typical sets is likely to be 
+%outside the tpical set which is why samplers which are not aware of the 
+%geometry of the target density such as RMH work very poorly in high dimension.}.
+%
+%HMC works by transforming the sampling problem to a classical mechanics 
+%problem expressed using the theory of Hamiltonian mechanics. 
+%We start by transforming the parameter space $\boldsymbol{\theta}$ into phase space by 
+%introducing a set of momentum parameters $\vec{p}$ (thus doubling the number of 
+%parameters) and define a \textsl{Hamiltonian} of the system to be the negative logarithm 
+%of the  joint density over $p(\vec{\theta}, \vec{p})$:
+%\begin{equation}
+%    \mathcal{H}(\boldsymbol{\theta},\mathbf p)\equiv -\ln p(\boldsymbol{\theta},\mathbf p)=-\ln p(\boldsymbol{\theta})-\ln p(\mathbf p\lvert\boldsymbol{\theta})\hquad .
+%\end{equation}
+%We can rewrite this Hamiltonian as
+%\begin{equation}
+%    \mathcal{H}(\boldsymbol{\theta},\mathbf p)= \mathcal{U}(\boldsymbol{\theta})  + 
+%    \mathcal{K}(\boldsymbol{\theta},\mathbf p)
+%\end{equation}
+%where $\mathcal{U}(\boldsymbol{\theta})=-\ln p(\vec \theta)$ is a potential energy term,
+% $\mathcal{K}(\boldsymbol{\theta},\vec p)$ is a kinetic energy term and 
+% $p(\vec\theta)$ is the target  density.
+%By mapping a posterior density to a physical system, we can use the machinery of 
+%classical mechanics by first sampling the momenta conditional on the parameters 
+%$\mathbf{p}\sim p(\mathbf{p}\lvert\boldsymbol{\theta})$
+% and then solving Hamilton's equations
+%\begin{align}
+%  \dot{\theta}_i&=\frac{\partial\mathcal{H}}{\partial p_i}=\frac{\partial\mathcal{K}}{\partial p_i}\\
+%  \dot{p}_i&=-\frac{\partial\mathcal{H}}{\partial q_i}=-\frac{\partial\mathcal{U}}{\partial q_i} - 
+%  \frac{\partial\mathcal{K}}{\partial q_i}
+%\end{align}
+%\textsl{Symplectic integrators} (integrators which respect Liouville's theorem of 
+%classical mechanics) are used to integrate Hamilton's equation because of their 
+%high accuracy. The most common choice for an integrator is the symplectic 
+%(and reversible) \textsl{Leapfrog integrator}, which is defined by 
+%\begin{align}
+%\vec{p}_{t+1 / 2} &=\vec{p}_{t}-\frac{\eta}{2} \frac{\partial \mathcal{U}\left(\boldsymbol{\theta}_{t}\right)}{\partial \boldsymbol{\theta}} \\
+%\boldsymbol{\theta}_{t+1} &=\boldsymbol{\theta}_{t}+\eta \frac{\partial \mathcal{K}\left(\boldsymbol{p}_{t+1 / 2}\right)}{\partial \boldsymbol{p}} \\
+%\vec{p}_{t+1} &=\boldsymbol{p}_{t+1 / 2}-\frac{\eta}{2} \frac{\partial \mathcal{U}\left(\boldsymbol{\theta}_{t+1}\right)}{\partial \boldsymbol{\theta}}
+%\end{align}
+%where $\eta$ is the \textsl{step size}. The Leapfrog integrator works by performing a
+%half-update of the momentum, followed by a full update of position and another half-update 
+%of momentum.
+%Because no integrator t preserve energy perfectly an RMH acceptence criterion of the form 
+%$\alpha=\mathrm{min}\{1,\exp (\mathcal{H}(\boldsymbol{\theta}, \mathbf p) -\mathcal{H}(\boldsymbol{\theta}^\prime, \vec p^\prime))\}$ 
+%is needed at the end of the inegration step in order to restore the detailed balance 
+%property of Markov chains.
+%For the pseudocode of a complete algorithm using the Leapfrog integrator see 
+%for example \citet{murphy_book_2023}.
+%
+%
+%The most challenging part of implementing HMC is tuning the various hyperaparamters of 
+%the algorithm, namely, the number of leapfrog integration steps $L$, the step size
+%$\eta$ and the kinetic energy term $\mathcal{K}$.
+%The most common choice for the kinetic energy term is the Euclidian-Gaussian kinetic 
+%energy distribution where  the conditional momentum distribution is simply a 
+%multivariate normal distribution independent of the position:
+%\begin{equation}
+%\mathcal{K}(\vec\theta, \vec p) = \mathcal{N}(\mathbf p ; 0, \mathbf M)
+%\end{equation}
+%where the covariance matrix $\mathbf M$ is usually called the \textsl{mass matrix}.
+%The choice of the mass matrix $\mathbf{M}$ is important for efficient sampling because 
+%every transformation of the parameter space induces an inverse transformation on 
+%momentum space \citep{arXiv:1701.02434}. Consider for example a a cigar shaped target 
+%density.
+%We can either transform the posterior such that it looks more Gaussian and make it
+% easier to sample from, or equivalently we can choose the momentum distribution such 
+% that the momenta are aligned with the shape of the cigar by having higher velocity 
+% proposals along the longer axis.
+%The optimal choice for $\mathbf{M}$ is to choose the matrix to be as close as 
+%possible to the covariance matrix of the parameter space.
+%The mass matrix is usually estimated empirically from the covariance matrix of the 
+%parameters during a burn-in phase of the sampling. 
+%
+%As for the leapfrog parameters, we want to choose the number of leapfrog steps $L$ to be 
+%large enough so that the algorithm explores large swathes of the parameter space.
+%The most popular variant of HMC, the No U-Turn Sampler (NUTS) algorithm 
+%\citep{arXiv:1111.4246} adaptively chooses $L$ such that Leapfrog integration is 
+%automatically halted when the trajectory starts going back to its starting point (the
+%the momentum vector points in the direction of the starting point).
+%This trick ensures that parameter space updates are as large as possible and it 
+%reduces autocorrelation between successive samples.
+%In addition to automatically selecting $L$, \citet{arXiv:1111.4246} also propose a 
+%scheme for adapting the step size $\eta$ on the fly. Adding the automatic adaptation
+%of the mass matrix by running short chains during the burn-in period means that 
+%NUTS removes any need to tuning. The lack of complicated tuning parameters and 
+%the fact that NUTS works very well across a large set of problems are one of the 
+%reasons why NUTS (and its derivations) is the work-horse algorithm 
+%in many popular statistical modeling packages such as \ssf{Stan}
+%\footnote{\url{https://mc-stan.org/}} \citep{2017JSS....76....1C}, 
+%\ssf{PyMC}\footnote{\url{https://www.pymc.io/welcome.html}}, \ssf{numpyro}
+%\footnote{\url{https://num.pyro.ai/}} and \ssf{TensorFlow Probability}
+%\footnote{\url{https://www.tensorflow.org/probability}}. 
+%
+%How well MCMC works depends almost entirely on the geometry of the target density
+%(which is parametrization dependent). If the target density is non-Gaussian 
+%HMC requires lots of leapfrog integration steps per iteration. It is entirely 
+%possible to have a 3 dimensional density with poor geometry (for example 
+%a multi-modal density with well seperated modes) for which HMC will 
+%fail but if the geometry of the distribution is well behaved HMC can scale to 
+%even millions of parameters.
+%
+%
+%One other notable variant of Hamiltonian Monte Carlo which works better for 
+%densities with highly complex geometries is 
+%Riemannian Manifold Hamiltonian Monte Carlo (RMHMC) \citep{girolami2011}.
+%RMHMC uses a position dependent kinetic energy term $\mathcal{K}$ and second order 
+%gradient information (quantifying the curvature of the target density) to improve 
+%sampling efficiency with highly correlated densities by adapting to the local geometry 
+%of the target density at every point. RMHMC is very rarely used and there are no 
+%good open-source implementations of the algorithm. The reason for this is that 
+%it is  difficult to construct numerically stable integrators for RMHMC
+%\footnote{\href{https://discourse.mc-stan.org/t/riemann-manifold-hmc-in-stan/19466/5}{This}
+%discussion on the \ssf{Stan} forums is particularly illuminating}.
+%Instead of using RMHMC, we can find better parametrizations for the target density
+%and use NUTS to sample it which can end up being equivalent to RMHMC 
+%\citep{arXiv:1910.09407}.
+%
+%For an excellent visualization of different MCMC samplers see 
+%\href{http://chi-feng.github.io/mcmc-demo/app.html}{here}.
+%
+%
+%\subsubsection{Diagnosing MCMC chains}
+%How can we tell that our MCMC chains have converged to target density?
+%Unfortunately there is no point at which we can say that the chain definitively
+%converged to the target density and that we are sampling from the typical set of 
+%the posterior density faithfully. At best, we can
+%tell when the chain has not converged. The prime diagnostic for the convergence
+%of MCMC chains is the \textsl{integrated autocorrelation time}.
+%Given $S$ independent samples, the variance $\sigma^2$ for a Monte Carlo 
+%estimate of an expectation value $g(\vec\theta)$  is 
+%\begin{equation}
+%    \sigma^2= \frac{1}{S} \mathbb{V}_{p}\left[g(\vec\theta)\right]
+%    \label{eq:mcmc_error_indep}
+%\end{equation}
+%and the standard deviation of this estimator will decrease
+%$\sim 1/\sqrt{S}$ with the number of samples.
+%However, samples from $p(\vec\theta)$ produced by MCMC are not truly independent.
+%The way we measure this correlation is by computing the integrated autocorrelation
+%time $\tau_f$. If the samples are correlated, the the variance in 
+%Equation~\ref{eq:mcmc_error_indep} becomes
+%\begin{equation}
+%    \sigma^2= \frac{\tau_f}{S} \mathbb{V}_{p}\left[g(\vec\theta)\right]
+%    \label{eq:mcmc_error}
+%\end{equation}
+%where $\tau_f$ is the average number of steps needed before the chain ``forgets''
+%its initial position. It is defined as
+%\begin{equation}
+%    \tau_f = \sum_{\tau=-\infty}^\infty\rho_f(\tau)
+%\end{equation}
+%where $\rho_f(\tau)$ is the normalized \textsl{autocorrelation function} (ACF) of 
+%the stochastic process which  generated the chain for
+%$f$ \footnote{See this blog post by Daniel Foreman-Mackey for more details on 
+%autocorrelation functions: \url{https://dfm.io/posts/autocorr/}.}.
+%Estimating integrated autocorrelation times is difficult and in general
+%requires long chains because the estimate of $\tau_f$ is slow to converge.
+%
+%Using an estimate of integrated autocorrelation time  we can compute the 
+%\textsl{Effective Sample Size} (ESS) for a given chain \citep{sokal1997}
+%\begin{equation}
+%\mathrm{ESS}\equiv\frac{S}{\tau_f}
+%\end{equation}
+%The integrated autocorrelation time (and hence also the ESS) depends on the function 
+%$g(\vec\theta)$ and will be different depending which expectation value we 
+%are interested in.
+%For example if we are interested in calculating the mean and require a certain
+%fractional precision we need to know the integrated autocorrelation time to estimate 
+%how many effective samples are necessary to achieve this precision.
+%\citet{arXiv:1903.08008} suggest computing ESS for both the bulk of the distribution 
+%(captured by the median) and \textsl{tail-ESS} which they define to be the minimum 
+%of the ESS for the 5\% and 95\% quantiles of the distribution. Tail-ESS will generally 
+%be smaller  than the ESS for the median because MCMC converges at a slower rate 
+%in the tail of the distribution.
+%
+%Since MCMC samples are correlated, some users discard all except every $n-th$ sample in a process called
+%\textsl{thinning} the chains. Except for reasons of storage in memory, this
+%procedure should be discouraged because discarding information never helps with
+%better estimates and the Monte Carlo error in Equation~\ref{eq:mcmc_error} 
+%already incorporates the fact that the samples are correlated.
+%
+%Another popular statistic for the diagnostic of MCMC chains is the
+%\textsl{Gelman-Rubin} $\hat R$ statistic \citep{1992StaSc...7..457G} which uses 
+%information pooled from multiple chains to judge convergence. It is 
+%defined as
+%\begin{equation}
+%    \hat R= \frac{\hat V}{W}
+%\end{equation}
+%where $W$ is the sample variance within a single chain, and $\hat V$ is the
+%variance estimate pooled across several independent chains.
+%The $\hat R$ statistic converges to 1 when each of the chains accurately 
+%sample the target density. A value greater than 1 indicates that one or more 
+%chains have not yet converged.
+%
+%Finally, the most sophisticated diagnostic for the convergence of MCMC chains
+%attempts to detect the the breakdown of the geometric ergodicity property of
+%MCMC. Geometric ergodicity is a necessary condition for MCMC estimators to
+%follow the central limit theorem ensures that the estimator in 
+%Equation~\ref{eq:monte_carlo_estimator} is consistent and unbiased.
+%HMC algorithms enable the detection of so called
+%\textsl{divergences} 
+%\footnote{\url{http://mc-stan.org/users/documentation/case-studies/divergences_and_bias.html}.}.
+% in the parameter space -- points in the parameter space 
+%where the target density has very large gradient indicating 
+%a suspected  breakdown of geometric ergodicity.
+%The presence of a large number of divergences in a model can lead to 
+%biased estimators. 
+%Modern statistical modeling packages include functions for computing ESS 
+%estimators,  $\hat R$ and divergences. Without the use of these diagnostic 
+%tools we cannot be sure that the MCMC chains have converged and that our results 
+%are correct.
+%
+%
+%
+%\subsubsection{Failure modes for samplers}
+%MCMC methods work really well for distributions with relatively well-behaved 
+%geometries but tend to fail with more complex distributions, particularlly 
+%multi-modal distributions with well seperated modes. 
+% \citet{arXiv:2101.09675} lists 
+%key properties of densities for which MCMC methods tend to fail:
+%\begin{itemize}
+%    \item \textbf{Highly non-Gaussian distributions}: For example banana shaped
+%    distributions, distributions with heavy tails.
+%    \item \textbf{Multimodal distributions}: If there are multiple well 
+%    separated  modes in the target densities, even if each mode is individually 
+%    well described by a Gaussian MCMC methods will tend to get stuck in one of
+%    the modes and fail to jump between modes.
+%    \item \textbf{High dimensionality}: High dimensional problems can break pretty 
+%    much every algorithm, they are however less of a problem for HMC like algorithms.
+%    \item \textbf{Highly informative distributions}: If the data is particularly 
+%    informative the posterior occupies an extremely small portion of the total 
+%    prior volume and it may be challenging to discover those regions. Even HMC 
+%    can fail here if the posterior mass is concentrated  in a very narrow 
+%    ``canyon'' if the momentum of the proposals is too large.
+%    \item \textbf{Phase transitions}: These are abrupt changes in the structure 
+%    of the likelihood as we move up constant likelihod contours.
+%\end{itemize}
+%The first and second failure modes are particularly important for this thesis 
+%because likelihoods for microlensing models are highly non-Gaussian and 
+%multimodal. In the next section we cover a class of algorithms which (on paper!)
+%addresses almost all of these challenges.
+%
+%\subsubsection{Nested Sampling}
+%\textsl{Nested sampling} \citep{2004AIPC..735..395S}  is fundamentally different 
+%from MCMC algorithms  because it was designed for the purpose  of estimating the
+%Bayesian evidence  $\mathcal{Z}$ (Equation~\ref{eq:evidence}) instead of drawing samples 
+%from  a target distribution.
+%Since $\mathcal{Z}$ is just a normalization constant, it is usually ignored 
+%in traditional MCMC algorithms such as RMH which don't require the target 
+%density to be properly normalized.
+%However, $\mathcal{Z}$  becomes important in the context of model comparison 
+%using \textsl{Bayes factors} (see Section~\ref{ssec:model_comparison}) which
+%are proportional to the ratios of the evidences of two models. 
+%
+%In addition to estimating $\mathcal{Z}$ Nested Sampling produces samples 
+%from posterior $p(\vec\theta\lvert \mathcal{D})$ as a sort of byproduct. 
+%The major advantage of NS over MCMC methods is that acts as a sort of 
+%global search algorithm working from the outside-in which improves its 
+%performance on highly multi-modal distributions\footnote{
+%    The extent to which this claim is true will be investigated in 
+%    Chapter~\ref{ch:microlensing_modeling}.
+%}. Nested Sampling has been extensively used in cosmology in particular, mostly 
+%thanks to the popularity of the \ssf{MultiNest} package \citep{arXiv:0809.3437}. 
+%\citet{arXiv:2205.15570} is great introductory review to these methods while 
+%\citet{arXiv:2101.09675} goes into a lot more depth.
+%
+%We start by rewriting the evience $\mathcal{Z}$ from Equation~\ref{eq:evidence}
+%as 
+%\begin{equation}
+%\mathcal{Z}=\int \mathcal{L}(\vec\theta) \pi(\vec \theta)\ud\vec\theta
+%\label{eq:evidence2}
+%\end{equation}
+%where  $\mathcal{L}\equiv p(\mathcal{D}\lvert\vec \theta)$ is the likelihood 
+%and $\pi(\vec \theta)$ is the prior. The key idea behind NS is that it is possible 
+%to replacec the high-dimensional integral in Equation~\ref{eq:evidence2} with 
+%a one-dimensional integral over the entire prior volume: 
+%\begin{equation}
+%    \mathcal{Z}=\int_{0}^{1} \mathcal{L}(X) \ud X
+%    \label{eq:nested_sampling}
+%\end{equation}
+%where $\ud X$ is the volume element of points in the parameters space which 
+%share the same likelihood $\mathcal{L}(X)$ weighted by the prior $\pi(\vec\theta)$:
+%\begin{equation}
+%X\left(\mathcal{L}^{\star}\right)=\int_{\mathcal{L}>\mathcal{L}^{\star}} \pi(\vec\theta) \ud \vec\theta\hquad .
+%\end{equation}
+%$\mathcal{L}=\mathcal{L}^\star$ defines a contour of constant likelihood and 
+%$X(\mathcal{L}^\star)$ is the prior volume enclosed by that contour.
+%\emph{Nested sampling thus replaces the original problem of sampling from the posterior 
+%to  sampling from prior with a likelihood constraint}.
+%For a derivation of Equation~\ref{eq:nested_sampling} see \citet{arXiv:2205.15570}.
+%
+%The standard version of the NS algorithm works by drawing $K$ \textsl{live points} 
+%(points in the parameter space) from the prior at each iteration $i$, removing 
+%the point with the lowest likelihood value $\mathcal{L}_i$ and replacing it 
+%with a new live point sampled from the prior and subject to the 
+%constraint $\mathcal{L}_{i+1}>\mathcal{L}_i$ (see \citet{arXiv:2205.15570} 
+%for details). This process continuously shrinks the volume of the prior by an 
+%approximately constant factor, effectively scanning the parameter space from 
+%contours with a lowest likelihood to contours with a highest likelihood.
+%The algorithm terminates when some fractional error tolerance on the estimation 
+%of the evidence is reached. 
+%At the end of the iterative procedure we are left with 
+%estimates of the volume $X(\mathcal{L}^\star)$ at each iteration and we can 
+%estimate the integral in Equation~\ref{eq:nested_sampling} using the trapezoidal 
+%rule:
+%\begin{equation}
+%\mathcal{Z}=\sum_{i=1}^{K}w_i\,\mathcal{L}_i^\star\hquad ,
+%\end{equation}
+%where the weights $w_i$ are 
+%\begin{equation}
+%w_{i}=\frac{1}{2}\left(X_{i-1}-X_{i+1}\right)\hquad .
+%\end{equation}
+%The posterior samples can be obtained from the from the discarded live points 
+%by assigning weights $\hat{w}_i$ to each live point:
+%\begin{equation}
+%    \hat{w}_i\equiv \frac{\mathcal{L}_i\,w_i}{\hat{\mathcal{Z}}}
+%\end{equation}
+%To compute the ESS for these samples, we can use the exectation value dependent 
+%estimator provided by \citet{arXiv:1809.04129}:
+%\begin{equation}
+%    \widehat{\mathrm{ESS}}=\frac{1}{\sum_{s=1}^S\tilde{w}_i^2}
+%\end{equation}
+%where
+%\begin{equation}
+%\tilde{w}_{i}\equiv\frac{\left|g\left(\vec{\theta}^{(s)}\right)\right| \hat w_{s}}{\sum_{i=1}^{S}\left|g\left(\vec{\theta}^{(i)}\right)\right| \hat w_{i}}
+%\end{equation}
+%
+%
+%\cite{arXiv:2101.09675} identifies three distinct phases in the behavior 
+%of the NS algorithm. In the first phase the prior volume shrinks towards the 
+%bulk of the posterior (the typical set) and during this phase the live points 
+%vary in likelihood by many orders of magnitude. All of the probability mass is 
+%associated with a single live point because the live points are all associated 
+%with an equal volume. 
+%The typical set is resolved in the second phase and the live points have 
+%comparable wieghts.  In the final stage NS reaches the region around the MLE 
+%with large values of the likelihood but very small volume.
+%Unlike MCMC, NS has a well defined termination criterion and running the algorithm 
+%for a longer time does not produce more posterior samples (there are other ways 
+%of producing more samples after an NS run has terminated).
+%NS is different compared to HMC in that HMC explores level sets of energy where 
+%the Hamiltonian is constant while NS scans through a range of energy levels. 
+%It's this feature that enables it to discover multiple modes in the density.
+%
+%
+%Historically NS has been  extensively used by physicists and almost completely 
+%ignored by statisticians. This has been changing in recent years and there 
+%is now more work from both communities aimed at understanding and improving the 
+%algorithm and better diagnosing its failure modes (see for example 
+%\citet{ arXiv:1407.5459,arXiv:1804.06406,arXiv:1704.03459,arXiv:2101.09675}).
+%For example, a recent paper by \citet{arXiv:1805.03924}  reframes the NS 
+%algorithm as a special case of \textsl{Sequantial Monte Carlo}, a class of 
+%algorithms that have been extensively studied by statisticians for decades.
+%There are also lots of modern open-source implementations of NS such as 
+%\ssf{dynesty} \citep{arXiv:1904.02180}, \ssf{JAXNS} \citep{arXiv:2012.15286}
+%and \ssf{UltraNest} \citep{2021JOSS....6.3001B}. 
+%\ssf{UltraNest} in particular is the most developed modeling framework 
+%focused on correctness of posterior inferences. Ultimately, the only way 
+%to check whether NS is doing what it promises to do is to apply it to a real 
+%world problem. In Chapter~\ref{ch:microlensing_modeling} we do just that and compare 
+%the results and performance of NS to HMC.
+%
+%\subsubsection{Ensemble samplers with affine invariance}
+%By far the most popular sampler in physics and astronomy is an MCMC method 
+%called \textsl{affine invariant ensamble sampler} (AIES)
+%\citep{2010CAMCS...5...65G} which was popularized by an excellent \ssf{Python}
+%implementation of the algorithm \ssf{emcee} 
+%\citep{arXiv:1202.3665,arXiv:1911.07688}. The method is similar 
+%to RMH  in the sense that it does not use gradient information and it requires 
+%only that the evaluation of the log-probability function. The special property 
+%of affine invariant MCMC is that it is invariante to 
+%\textsl{affine transformations} (linear transformations such as 
+%scale transformations and rotations) of the parameter space. It uses multiple 
+%MCMC chains (called \textsl{walkers}) to explore the parameter space which 
+%are correlated by construction. 
+%
+%\ssf{emcee} is so popular that the 
+%original paper has been cited almost 8000 times so far.  
+%The success of \ssf{emcee} has less to do with its actual suitability 
+%to a wide range  of problems and more to do with the fact that it is very easy to use with the 
+%kinds of models astronomers and physicists work with. \ssf{emcee}  works with
+%black-box physics simulators and it is also easily parallelizable. 
+%This feature also makes it easy to misuse and apply to problems for which it is 
+%not suitable. For example, \citet{arXiv:1509.02230} showed that \ssf{emcee} 
+%does not work well in high dimensions (greater than about 10) and it can appear 
+%as if the sampler converged when in reality it has not. Nevertheless 
+%\ssf{emcee} can still be very useful and fast for very low dimensional 
+%problems. As with other MCMC samplers, it will fail with multi-modal 
+%likelihoods.
+%
+%
+%\subsubsection{Other methods}
+%That there are many many algorithms in the  literature besides the ones I've covered here. 
+%The vast majority of these  algorithms are variants of RHMC, HMC or  
+%NS. Most of these algorithms don't have well tested opens-source implementations, 
+%the examples presented in the papers are usually low-dimensional toy 
+%problems, and the algorithms require very careful tuning to work at all. This 
+%is why the development of general purpose statistical modeling 
+%(probabilistic programming) packages such as \ssf{Stan} and 
+%\ssf{PyMC} has been so important. They demonstrated that methods such as 
+%NUTS work well for a wide variety of problems.
+%
+%\subsection{Model validation and comparison}
+%\label{ssec:model_comparison}
+%
+%So far I have discussed inference of model parameters using either optimization 
+%methods such as maximum likelihood or sampling methods such as MCMC and Nested 
+%Sampling.
+%Using these methods we can obtain some estimates of the parameters of a specific  
+%model but this  does not tell us if the model is any good in the first place.
+%Every time we are doing parameter inference we assume that the particular model
+%is true.
+%A major challenge in statistical inference is comparing different 
+%models which can explain the same dataset, so called \textsl{model comparison}.
+%What we mean by different models is somewhat fuzzy. We could for example have 
+%a model specified by the same likelihood function and a set of continous 
+%parameters and then call different realizations of those parameters different 
+%models. This does not make much sense in the Bayesian context. In most cases
+%when a model comparison problem can be cast as a parameter inference problem it 
+%is easier to treat it as a parameter inference problem\footnote{Although, as we
+%will see in Chapter~\ref{ch:microlensing_modeling} it may be advantageous  to use model 
+%comparison methods when comparing different modes in a multi-modal posterior.}.
+%This happens when two \emph{models are nested}, meaning they share one or 
+%several parameters.
+%For example, let's say we are fitting a microlensing model and trying to decide 
+%whether  or not to include the finite source effect parametrized by a continuous 
+%parameter $\rho_\star$.
+%We could fit two separate models, a model which uses a point source 
+%aproximation for the magnification and one which includes the finite source 
+%effects via the $\rho_\star$ parameter and then decide which model better 
+%describes the data. The alternative is to simply fit the more complex model, 
+%obtain a posterior for the $\rho_\star$ parameter and check if it differs from 
+%zero in a meaningful way. Although the former approach is more elegant sometimes
+%it is practically intractable because we might end up having trouble sampling 
+%the posterior if the additional parameter is not well constrained by the data\footnote{In those
+%cases is may nake sense to put a strong prior pushing the parameter to zero which 
+%helps with inference. If the parameter is well constrained by the data the 
+%likelihood will push the posterior away from zero.} or if the model with the 
+%extra parameter is much more computationally expensive. 
+%If the models are nested but the paramer which differentiates between models is 
+%discrete it is sometimes still possible to cast the comparison problem as a parameter 
+%inference problem using \emph{dimension jumping} methods such as 
+%\textsl{Reversible-jump MCMC} (RJ-MCMC)
+%\citep[See][for an application of a method similar to RJ-MCMC to an 
+%astronomy problem]{2015MNRAS.448.3206B}.
+%However, these methods are computationally very difficult to implement.
+%
+%When the models are not nested we have no choice but to do model comparison. 
+%An example would be fitting a single lens microlensing model vs. a binary lens 
+%or a binary source star model. Statistics alone does not tell us what to put in the 
+%abstract of a paper after we have completed the analysis for multiple competing 
+%models\footnote{This is the subject of decision theory which relies on 
+%specifying \textsl{utility functions} (which depend on our posterior beliefs) 
+%to quantify expected loss given a certain decision}. Best we can hope for is 
+%assigning probabilities to each of the models taking into account how well they 
+%fit the data and how parsimonious they are. Model 
+%comparison is one of the key problems with analyzing microlensing light curves 
+%because there are often multiple very different models which make near-identical 
+%predictions  
+%\citep[see][for examples from microlensing]{ 2017AJ....153..129J,2020AJ....160...17H,2021AJ....162...59R}.
+%Especially problematic is the fact that when multiple models equally well 
+%describe the data, \emph{some models will be more interesting than others}.
+%For example one model might be a triple lens model implying the discovery of 
+%an \textsl{exomoon} in a planetary system. The other model could be a binary 
+%source star model. The major issue here is that the incentives of researchers 
+%are not always aligned with truth-seeking and the outcome of a model 
+%comparison test can be a strong function of how much effort was invested into 
+%a given model\footnote{This is one of the causes of the replication crisis in 
+%many scientific disciplines.}. In the extreme case, if we only consider one interesting model then
+%there is no chance we can conclude that  a more mundane model describes the 
+%data equally well. This is why it is very important to have a clearly defined 
+%framework for assesing different models.
+%
+%\subsubsection{Frequentist model comparison (hypothesis testing)}
+%The question of model comparison is different depending on whether
+%one uses the frequentist or Bayesian statistics. In classical frequentist
+%statistics, one has to form the \textsl{null hypothesis} $\textrm H_0$ which 
+%represents a baseline (null) model and an \textsl{alternative hypothesis} which is 
+%the hypothesis of interest. For instance, the two hypotheses could be
+%\begin{align}
+%    \textrm H_0: & \; \textrm{there is no planet in the data} \notag\\
+%    \textrm H_1: & \;\textrm{there is a planet in the data}\notag
+%\end{align}
+%We then form a \textsl{hypothesis test} under some \textsl{test statistic} 
+%and based on the outcome of that hypothesis test we either reject the null 
+%hypothesis or we don't, depending on an arbitrary threshold called the 
+%\textsl{p-value}.
+%
+%The p-value is the probability of obtaining data equal to or more extreme than 
+%what is observed, \emph{assuming that the null hypothesis $\textrm H_0$ is true}.
+%It is the probability for obtaining the observed
+%dataset in an infinite set of draws of the data, \emph{it is not a statement 
+%about the probability of the hypothesis}\footnote{The definition of the p-value
+%is so confusing that even some statistics textbooks get it wrong.}.
+%All hypothesis testing does is answering the following question:
+%assuming there's nothing interesting in the data, how likely was I to observe 
+%the data I was given.
+%If a certain threshold p-value is reached ($0.05$ in most scientific disciplines, 
+%substantially less in physics and astronomy), the null hypothesis hypothesis 
+%is rejected.
+%The fact the null hypothesis is rejected cannot be interpreted as evidence in
+%favour of the alternative hypothesis because such a statement would involve
+%using Bayes' theorem to invert the probabilities. However, it is almost always
+%(mis)interpreted in that way.
+%Given a set of models $\{\mathcal{M}_1,\mathcal{M}_2,\dots,\mathcal{M}_n\}$, all
+%we are formally allowed to do is to  reject some of those models based on 
+%a p-value, and keep the rest.
+%
+%
+%One might wonder if there is as an absolute way of judging the quality of a
+%single isolated model. In astronomical literature, this usually falls under the term of
+%a \textsl{goodness of fit} statistic. These statistics implicitly use the 
+%hypothesis testing procedure described above, rejecting the null hypothesis 
+%under some p-value. Popular choices are the \textsl{Chi-squared test} 
+%(not to be confused with the $\chi^2$ loss function),
+%\textsl{Kolmogorov-Smirnov test}, \textsl{Anderson-Darling test}, etc.
+%These tests rely on strong assumptions about the data generating process 
+%and they are often misused. 
+%It is often preferable (if possible) to just use visual checks on the predictions 
+%of the model 
+%\citep[see][for a visualization guide in the Bayesian paradigm]{arXiv:1709.01449}. 
+%For example plotting the predictions of the model in data space, plotting 
+%the residuals, etc.  Probabilistic modeling is almost always an iterative 
+%process and visualization helps us spot where the model is failing to describe 
+%the data adequatelly.
+%
+%\subsubsection{Bayesian model comparison using Bayes factors}
+%In the Bayesian paradigm, the most natural way of comparing different 
+%models is by making use of  Bayes' theorem. We can assign the probability of a 
+%given model $\mathcal{M}$ as 
+%\begin{equation}
+%    p(\mathcal{M}\lvert\mathcal D)\propto p(\mathcal{M})\,p(\mathcal D\lvert\mathcal{M})
+%\end{equation}
+%where $p(\mathcal{M})$ is the prior probability of the model $\mathcal{M}$ and
+%$p(\mathcal D\lvert\mathcal{M})$ is the Bayesian evidence. If we are comparing two 
+%models, $\mathcal{M}_o$ with a set of parameters $\vec\theta_0$ and
+%and $\mathcal{M}_1$ with a set of parameters $\vec\theta_1$, we can form the ratio of 
+%the their posterior probabilities given the data
+%\begin{equation}
+%    \frac{p(\mathcal{M}_0\lvert\mathcal D)}{p(\mathcal{M}_1\lvert\mathcal D)} =
+%    \frac{\mathcal{Z}_0}{\mathcal{Z}_1} \, \frac{p(\mathcal{M}_0)}{p(\mathcal{M}_1)}\hquad .
+%\end{equation}
+%The ratio of the two evidences on the right hand side is called the 
+%\textsl{Bayes factor}, and the ratio of prior probabilities is the 
+%\textsl{prior odds}.
+%Prior odds can be set to unity under the assmuption that both models are equally 
+%likely a priori.
+%A large Bayes factor $\textrm{B}_{01}\equiv \mathcal{Z}_0/\mathcal{Z}_1$ should 
+%be interpreted as evidence in favour of model 0 versus model 1, given the 
+%observed data $\mathcal D$.
+%As in the case of hypothesis testing, one can then define arbitrary criteria for
+%the strength of the evidence in favour of model 0 conditional on the specific value
+%of the Bayes factor.
+%The popular choice is using Jeffreys scale \citep{1939thpr.book.....J}.
+%Unlike the p-value, Bayes factors (multiplied by prior odds)
+%are actual probabilities for the models
+%themselves, rather than probabilities for the data in an infinite set of trials.
+%One useful aspect of Bayesian model comparison is the fact that it automatically
+%implements \emph{Occam's razor}, penalizing overly complicated models if there
+%is a simpler alternative \citep[see Chapter 28 of][]{2003itil.book.....M}.
+%The goal when using Bayes factor is to either select a single ``best'' model
+%$\mathcal{M}_i$ or, the more Bayesian option, to average over a set of 
+%multiple models using their  posterior probabilities 
+%$p(\mathcal{M}_i\lvert\mathcal D)$.
+%
+%There are two major problems with Bayes' factor. The first one is the fact that
+%whereas the Bayesian evidence $\mathcal{Z}$ could often be ignored when inferring the 
+%parameters of the models, it is a crucial quantity in Bayesian model comparison. 
+%Computing $\mathcal{Z}$ is very computationally expensive, error prone, and it
+% requires methods such as Nested Sampling.
+%The other problem with Bayes
+%factors is that the estimates of $\mathcal{Z}$ are particularly sensitive to the 
+%width and the shape of te prior distribution for the parameters. For example, if 
+%the prior for parameter $\theta_k$
+%in model $\mathcal{M}_0$ is $\theta_k\sim \mathcal{N}(0, s^2)$, where $s$ is
+%very large, the evidence will scale as $\mathcal{Z}\sim 1/s$ practically
+%independently of the data. This kind of dependence of the evidences on the
+%parameter priors is not unexpected, it means that models which ``waste'' prior
+%parameter space will be penalized, but it is often undesirable\footnote{See the 
+%excellent essay on Bayes factors by \citet{navarro_2020}.}. 
+%
+%Thus, in order to use Bayes factors for model comparison, one has to be able to
+%compute evidences, and think very carefully about the choice of priors. Despite
+%these issues, Bayes factors have been used extensively in staistics and the 
+%natural sciences, most notably in cosmology. For example,\citet{arXiv:1312.3529} 
+%used Bayesian model comparison to estimate Bayes factors of 193 different 
+%inflation models using the Cosmic Microwave Background
+%data from the Planck satellite and Higgs inflation as the reference model. 
+%
+%\subsubsection{Cross validation}
+%The more tractable and robust alternative to estimating Bayes factors is using 
+%\textsl{cross validation}. The key idea behind cross validation (CV) is to judge 
+%different models based on how well they predict unseen data\footnote{This is 
+%principle method of model comparison in machine learning. In machine learning 
+%it is customary to split the original dataset into a \textsl{training dataset} 
+%and a \textsl{validation dataset}. The training dataset is used to optimize 
+%the main parameters of the model and the loss function evaluated on the 
+%validation dataset is used to compare different models or set the hyperparameters
+%of a given model.}. 
+%CV can be used to assess the predictive power of a single model (as a sort of 
+%goodness of fit metric) to compare or average multiple models. 
+%The idea behind CV is to split the dataset into $K$ parts or \textsl{folds}, 
+%re-fit the model including $K-1$ folds, evaluate the objective function (log 
+%likelihood or posterior) and then repeat the procedure for each fold. We 
+%can then for example add up the $K$ values and compute an estimate 
+%the predictive performance of the model. 
+%If we set $K$ to be equal to the number of data points $N$, then we 
+%are talking about \textsl{leave-one-out cross validation} (LOO-CV).
+%
+%One measure of the predictive performance of a given model $\mathcal{M}$ which has 
+%good theoretical properties is ther so-called 
+%\textsl{exptected log predictive density} (ELPD) 
+%\citep[see for example][]{10.1214/12-SS102}. It is defined as 
+%\begin{equation}
+%\operatorname{ELPD}(\mathcal{M})\equiv \int \ln p\left(\tilde{\vec y}_{i} \lvert \vec y, \mathcal{M}\right)\,p_{t}\left(\tilde{\vec y}_{i}\right) \ud \tilde{\vec y}_{i}
+%\end{equation}
+%where 
+%\begin{equation}
+%p(\tilde{\vec y}_i\lvert\vec y)=\int p\left(\tilde{\vec y}_{i} \lvert \vec \theta,\mathcal{M}\right)\,p(\vec\theta \lvert \vec y) \ud \vec\theta
+%\end{equation}
+%is the posterior predictive distribution for a new observation $\tilde{\vec y}_i$ that 
+%was generated from some unknown true distribution $p_t(\tilde{\vec y}_i)$.
+%ELPD captures the expected predictive performance of the model for some future data
+%$\tilde{\vec y}$.  It tells us how well the model will generalize to new data.
+%
+%One method for estimating ELPD is using LOO-CV. In LOO-CV we compute $N$ leave-one-out 
+%posterior distributions $p(\vec\theta\lvert \vec{y}_{-i})$ where $\vec y_{-i}$ 
+%denotes all of the observations except the $i$-th observation. Using these 
+%posterior estimates we can estimate the ELPD as 
+% \begin{align}
+%\mathrm{ELPD}_{\mathrm{LOO}} &\approx\frac{1}{N} \sum_{i=1}^{N} \ln p\left(\vec y_{i} \lvert \vec y_{-i},\mathcal{M}\right) \\
+%\label{eq:elpd_loo}
+%&=\frac{1}{N} \sum_{i=1}^{L} \ln\int p\left(\vec y_{i} \lvert \vec \theta, \mathcal{M}\right) p\left(\vec\theta \lvert \vec y_{-i},\mathcal{M}\right) \ud \vec\theta \\
+%\end{align}
+%where $p(\vec y_i\lvert \vec \theta)$ is the likelihood for the $i$-th observation 
+%and $p(\vec\theta\lvert \vec y_{-i},\mathcal{M})$ is the posterior for $\vec\theta$ 
+%obtained by fitting the full dataset except the held-out observation $\vec y_{-i}$.
+%
+%A naive implementation LOO-CV is very computationally expensive because it 
+%involves re-fitting the model $N$ times. Fortunately, there are reliable methods for 
+%approximating it. 
+%The most popular approximation uses importance sampling such that the target 
+%distribution is the LOO posterior and the proposal distribution is the full
+%posterior. Given $S$ draws from the full posterior 
+%$\vec\theta\sim p(\vec\theta|\vec y, \mathcal{M})$ we can estimate 
+%$p(\vec y_i\lvert \vec y_{-i},\mathcal{M})$ as 
+%\begin{align}
+%\ln p\left(\vec y_{i} \lvert \vec y_{-i},\mathcal{M}\right)&\approx\ln\left(\frac{\frac{1}{S} \sum_{s=1}^{S} p\left(\vec y_{i} \lvert\vec \theta^{(s)},\mathcal{M}\right) w\left(\vec\theta^{(s)}\right)}{\frac{1}{S} \sum_{s=1}^{S} w\left(\vec\theta^{(s)}\right)}\right) \\
+%\label{eq:prob_loocv_pointwise}
+%w\left(\vec\theta_{s}\right)&=\frac{p\left(\vec\theta^{(s)} \lvert \vec y_{-i},\mathcal{M}\right)}{p\left(\vec\theta^{(s)} \lvert\vec y,\mathcal{M}\right)} \propto \frac{1}{p\left(\vec y_{i} \lvert\vec\theta^{(s)},\mathcal{M}\right)}
+%\end{align}
+%If particular observations are highly influential (removing one data point 
+%noticeably changes the results) the target LOO posterior can differ 
+%from the full posterior which is why Pareto smoothed importance sampling (PSIS) 
+%(see Section~\ref{ssec:sampling_methods}) is used to improve the estimate. 
+%Using Equation~\ref{eq:prob_loocv_pointwise} and Equation~\ref{eq:elpd_loo}
+%we define \citep{2015arXiv150704544V} 
+%\begin{equation}
+%\mathrm{ELPD}_\mathrm{PSIS-LOO}\equiv\sum_{i=1}^N\ln\left(\frac{\frac{1}{S} \sum_{s=1}^{S} p\left(\vec y_{i} \lvert\vec \theta^{(s)},\mathcal{M}\right) w\left(\vec\theta^{(s)}\right)}{\frac{1}{S} \sum_{s=1}^{S} w\left(\vec\theta^{(s)}\right)}\right) \hquad .
+%\label{eq:elpd_psis_loo}
+%\end{equation}
+%We can use the value of the Pareto $\hat k$ from PSIS to determine if the 
+%PSIS-LOO ELPD estimator of $p(\vec\theta\lvert\vec y_{-i})$ for the $i$-th data 
+%point potentially do a full re-fit of the model only for the most problematic 
+%points. \citet{arXiv:1004.2316} showed that LOO is a consistent and unbiased estimator of 
+%the ELPD.
+%
+%Equation~\ref{eq:elpd_psis_loo} thus enables us to estimate the expected log pointwise 
+%predictive density for unseen data using only posterior samples without having 
+%to re-fit the model (unless Paretko $\hat k$ is large for certain data points). 
+%This equation forms the basis for the widely popular \ssf{loo} statistical package 
+%\citep{loo_package}.
+%$\mathrm{ELPD}_\mathrm{PSIS-LOO}$ is closely related to the WAIC information 
+%criterion \citep{arXiv:1004.2316} which is an alternative approach to estimating 
+%ELPD. LOO and WAIC are asymptotically equivalent but LOO is a more robust and less 
+%biased estimator with finite data \citep{arXiv:1507.04544}.
+%Other information criteria such as AIC, BIC and DIC are inferior to LOO and WAIC 
+%estimates of ELPD because they use point estimates (maximum likelihood) to assign 
+%a score to a model and they attempt to quantify model complexity in an ad-hoc 
+%way (by including a term proportional to the number of parameters in the model). 
+%LOO-CV is also less sensitive to the choice of priors and often more reliable 
+%for model comparison than Bayes factors \citep{10.1214/17-BA1091,arXiv:2003.04026}.
+%
+%What if our model includes a Gaussian Process and the likelihood is not 
+%factorizable? In that case we cannot use Equation~\ref{eq:elpd_psis_loo} but 
+%\citet{arXiv:1810.10559} showed how to evaluate 
+%$\ln p(\vec y_i\lvert\vec y_{-i},\vec\theta,\mathcal{M})$ if the likelihood is 
+%multivariate normal with a dense covariance or a Student-t likelihood so it is 
+%still possible to use LOO-CV cheaply. 
+%If we are dealing with time-series data and we are only interested in predicting 
+%the next few points conditional on the previous data points then it may not 
+%much sense to use LOO-CV because we are using future data points when evaluating 
+%predictions on past data points. \textsl{leave-future-out cross validation} 
+%(LFO-CV) \citep{arXiv:1902.06281} is a generalization of LOO which takes into 
+%account the time-ordering of the data points in a time-series. 
+%
+%For a more information about Cross Validation and its approximations see the excellent
+%\href{https://avehtari.github.io/modelselection/CV-FAQ.html}{Cross-validation FAQ}
+%by Aki Vehtari.
+%In Chapter~\ref{ch:microlensing_modeling} I will demonstrate the utility of these methods by applying 
+%them to very practical problems in the analysis of microlensing light curves.
+%
+%\subsection{What about machine learning?}
+%How is \textsl{machine learning} (ML) different from the topics we have discussed so far? 
+%I would say the key difference is that ML and particularly \textsl{deep learning} (DL) 
+%\emph{ML is much more focused on predicting future data} conditional on observing 
+%past data  than on estimating the parameters of some model, which is the subject of 
+%statistical inference. Classic example is that of neural networks which 
+%today routinely consist of millions, even billions of parameters (weights) but 
+%the  values of these weights are not important, only the prediction error 
+%(on unseen data) is what matters. For this reason optimization methods such as 
+%\textsl{stochastic gradient descent} (SGD) which are computationally much 
+%cheaper to implement are dominant over sampling methods such as MCMC.
+%
+%Despite the fact that the deep learning revolution started in 2012, 
+%deep learning has not had as much impact  in the physical sciences as many have 
+%expected  despite the fact that  there are many papers on the topic.
+%DL methods tend to be black boxes which are very efficient at 
+%the task they were trained to do, but their internal structure is usually 
+%illegible. This is not as much of an issue in industry applications as it is 
+%in the sciences where we care deeply about the precise values of very few 
+%parameters.
+%Nevertheless, all of this is changing rapidly and there is  a lot of work 
+%going on the boundary of physics and machine learning. One example is the field 
+%of \textsl{likelihood free inference} (LFI) which aims to emulate expensive 
+%physics simulators (such as massive hydrodynamics simulations) with neural 
+%networks so that one can do inference with much faster emulators 
+%\citep[see for example the review paper by]{2020PNAS..11730055C}. There has 
+%also been a lot of work on neural networks which respect certain physical 
+%symmetries (see for example Hamiltonian and Lagrangian neural 
+%networks\footnote{Or \textsl{geometric deep learning} more generally.})
+%and even automated discovery of succinct mathematical descriptions of 
+%the dynamical behavior of a physical system through observations of its 
+%behaviour \citep{arXiv:2006.11287}.
+%
+%However, the biggest significance of the atmospheric rise of machine learning 
+%in my opinion is that the computational infrastructure developed for training 
+%large neural networks can also be used for scientific computing. In particular,
+%two of the key technologies responsible for the deep learning  revolution
+%are \textsl{Graphical Processing Units} (GPUs) and 
+%\textsl{automatic differentiation} (AD), are also very useful 
+%for building scientific models. 
+%
+%\section{Automatic differentiation}
+%\label{sec:autodiff}
+%\textsl{Automatic differentiation} (AD or autodiff for short)\citep{10.1145/355586.364791} is an old idea 
+%that has revolutionized the fields of machine learning, AI and optimization in the 
+%recent decade. 
+%The key idea behind automatic differentiation 
+%is that any model implemented in computer code (such as \ssf{C++} or 
+%\ssf{Python}) is fundamentally a composition 
+%of simple functions such as additions, multiplications and trig functions 
+%which can be easily differentiated and these gradients evaluated at each step 
+%of the computation can be combined using the chain rule from calculus to 
+%yield the exact (up to machine precision) derivative of the complete model.
+%Specialized automatic differentiation libraries build a representation of the 
+%code as graph, evaluate the derivatives at each node and propagate them through 
+%the entire graph. 
+%
+%A variant of automatic differentiation called reverse mode automatic differentiation 
+%is widely used in machine learning (where it goes under the name of 
+%\emph{backpropagation}) to differentiate scalar functions with respect to 
+%millions if not billions of parameters. 
+%Backpropagation and autodiff in general tells us how to change the inputs of a 
+%model (even if there are billions of inputs) to obtain a particular output.
+%Autodiff has been used to differentiate through all sorts of things, including 
+%molecular dynamics simulations, \citep{arXiv:1912.04232}, cosmological simulations 
+%\citep{2016MNRAS.463.2273F}, simulations of the magnetic field in nuclear fusion 
+%reactors \citep{arXiv:2009.00196}, ODE and PDE solvers etc. It has even been 
+%applied to differentiating code at 
+% the level of a compiler using the LLVM compiler infrastructure \citep{arXiv:2010.01709}.
+% Because the wide variatey of autodiff applications
+%the famous AI researcher Yann LeCun recently popularized the term 
+%\textsl{differentiable programming} to refer to any computer code which is 
+%automatically differentiable.
+%
+%The reason why I am discussing this topic in the first place is that implementing AD 
+%for astrophysical models enables the use  of advanced MCMC methods such as 
+%Hamiltonian Monte Carlo and much faster and more stable numerical optimization  
+%(for obtaining the MLE estimate for example).
+%
+%\subsection{Three ways of differentiating computer programs}
+%There are three ways of computing derivatives:
+%\begin{itemize}
+%    \item \textbf{Symbolic differentiation}: computing \emph{exact} derivatives of 
+%    functions using either a pen and paper or a computer algebra system such as 
+%    \ssf{Wolfram Mathematica} or \ssf{SymPy} \citep{10.7717/peerj-cs.103}.
+%    \item \textbf{Numerical differentiation}: computing \emph{approximate} derivatives 
+%    using \textsl{finite difference}  methods.
+%    \item \textbf{Automatic differentiation}: computing \emph{exact} derivatives of entire 
+%    computer programs containing things like conditionals, loops, etc.
+%\end{itemize}
+%Symbolic differentiation is familiar to everyone from basic calculus. Using either 
+%pen and paper  or Computer Algebra Systems (CAS) we can compute exact 
+%derivatives of functions. 
+%If our model is expressible as a straightforward closed-form function it is 
+%often not to difficult to compute the derivatives of interest and indeed this 
+%is what people have been doing for decades to improve the performance of their
+%models\footnote{Before the advent of easy to use open-source AD libraries 
+%the contribution of many papers in applied AI would be to compute the analytic 
+%derivatives for some specific model.}
+%
+%The challenge with symbolic differentiation is that it is not always possible 
+%to easily compute the gradients, especially if the model is implemented in terms 
+%of a code which contains control flow constructs such as loops and conditionals, 
+%or iterative solvers. In those cases gradients of the output of the model 
+%are with respect to the input are usually computed using finite-difference 
+%methods. In finite-difference methods the derivative of some function $f(x)$
+%is defined as 
+%\begin{equation}
+%    f^\prime(x)=\lim _{h \rightarrow 0} \frac{f(x+h / 2)-f(x-h / 2)}{h}
+%\end{equation}
+%for some small step size $h$. Numerical gradients are easy to compute even with 
+%complicated models but they have two key drawbacks. First, obtaining a 
+%good approximation of the derivatives requires a small step size $h$ which is 
+%can lead to errors due to numerical cancellations.
+%The second is that if we want to evaluate the Jacobian of a function with lots 
+%of input parameters we would need to compute a finite difference gradient 
+%for each partial derivative which is prohibitively expensive in high dimensions.
+%
+%Automatic differentiation solves these problems. If AD is properly implemented 
+%evaluating the exact gradient is only a factor of a few slower than evaluating 
+%the function itself, even for models with millions of parameters.
+%The drawback is that it is usually
+%necessary to express our model in terms of constructs which are implemented in 
+%specialized AD libraries. It also requires a slightly different programming style
+%than what most people are used to. Fortunately there are now lots very flexible 
+%and easy to use high performance AD libraries. 
+%
+%\subsection{Forward and reverse-mode autodiff and the chain rule}
+%What follows is mostly a summary of the chapter on autodiff 
+%from \citet{murphy_book_2023}. First, we define the notation
+%\begin{equation}
+%\partial x_1 f\left(x_{1}, x_{2}, \ldots, x_{n}\right)
+%\end{equation} 
+%which denotes the partial derivative of function 
+%$f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}$ with respect to argument $x_1$ 
+%evaluated at the point $(x_{1}, x_{2}, \ldots, x_{n})$.
+%We define the derivative operator $\partial$ and denote the derivative of a 
+%differentiable function $f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^m$ as 
+%\begin{equation}
+%    \partial f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^{m\times n}
+%    \label{eq:general_derivative}
+%\end{equation}
+%The function $\partial f$ is a \textsl{linear map} which maps a point 
+%$\vec x\in \mathbb{R}^n$ to the \textsl{Jacobian} of all partial derivatives 
+%evaluated at $\vec x$.  $\partial f$ evaluated at point $\vec x$,
+%$\partial f (\vec x)$ is the \textsl{linearization} of function $f$ at point $\vec x$.
+%The Jacobian can be be represent as a matrix $\vec J$ whose 
+%elements $J_{ij}$ are given by 
+%\begin{equation}
+%J_{i j}=\frac{\partial f_{i}}{\partial x_{j}}
+%\end{equation}
+%  To make this slighly less abstract, consider the case when
+%$m=n=1$, the function takes a scalar and returns a scalar, then 
+%Equation~\ref{eq:general_derivative} is just the familiar derivative $f'(x)$:
+%\begin{equation}
+%    f^\prime\,:\,\mathbb{R} \rightarrow \mathbb{R}
+%\end{equation}
+%If the $f$ takes a vector and returns a scalar ($m=1$) then derivative operator is 
+%the \textsl{gradient} $\nabla f$:
+%\begin{equation}
+%    \partial f\,:\,\mathbb{R}^{n} \rightarrow \mathbb{R}^{1\times n}
+%\end{equation}
+%
+%We now introduce two key ingredients for an automatic differentiation framework 
+%which enable us to efficiently compute the Jacobian for functions of arbitrary 
+%complexity.
+%
+%\subsubsection{Jacobian-vector products (JVPs)}
+%Now consider consider a perturbation $\vec v\in\mathbb{R}^n$ to the function input 
+%$\vec x$. The  mapping 
+%\begin{equation}
+%(\vec{x}, \vec{v}) \mapsto \partial f(\vec{x})\,\vec v
+%\label{eq:jvp}
+%\end{equation}
+%is called the \textsl{Jacobian-vector product} (JVP). For example, if $f=\cos(x)$ 
+%for a scalar $x$ and $v$ is also a scalar, the JVP would be 
+%\begin{equation}
+%(x, v) \mapsto \partial \cos (x) v=-v \sin (x)\hquad .
+%\end{equation}
+%Why is this useful? With a function $f\,:\,\mathbb{R}^n\rightarrow\mathbb{R}^m$
+%and a \textsl{one-hot} vector $\vec v=(1, 0, \ldots, 0)^\intercal$ we can use  the JVP to 
+%evaluate the Jacobian matrix one column at a time:
+%\begin{equation}
+%    \partial f(\vec x)\,\vec v = 
+%    \begin{pmatrix}
+%\frac{\partial f_{1}}{\partial x_{1}} & \cdots & \frac{\partial f_{1}}{\partial x_{n}} \\
+%\vdots & \ddots & \vdots \\
+%\frac{\partial f_{m}}{\partial x_{1}} & \cdots & \frac{\partial f_{m}}{\partial x_{n}}
+%    \end{pmatrix}
+%    \begin{pmatrix}
+%        1\\
+%        0\\
+%        \vdots\\
+%        0
+%    \end{pmatrix}
+%    = \begin{pmatrix}
+%        \frac{\partial f_{1}}{\partial x_{1}}\\
+%        \frac{\partial f_{2}}{\partial x_{1}}\\
+%        \vdots\\
+%        \frac{\partial f_{m}}{\partial x_{1}}
+%    \end{pmatrix}
+%\end{equation}
+%Evaluating one column of the Jacobian in this way costs about the same as one function evaluation
+%so this is efficient for fucntions with ``tall Jacobians'' (functions with many outputs and few
+%inputs) but innefficient for functions with ``wide Jacobians''.
+%
+%\subsubsection{Vector-Jacobian products (VJPs)}
+%Similarly, if we consider 
+%a perturbation to $\vec u\in\mathbb{R}^m$ to the output of the $f$, we obtain
+%the \textsl{vector-Jacobian product} (VJP):
+%\begin{equation}
+%(\vec{x}, \vec{u}) \mapsto \partial f(\vec{x})^{\intercal}\,\vec u
+%\label{eq:vjp}
+%\end{equation}
+%Whereas JVPs are useful for computing the Jacobian matrix one column at a time, we can use VJPs
+%to compute the Jacobian one row at a time:
+%\begin{equation}
+%    \vec v^\intercal\,\partial f(\vec x)= 
+%    (1\,0\, \cdots \, 0)\,
+%    \begin{pmatrix}
+%\frac{\partial f_{1}}{\partial x_{1}} & \cdots & \frac{\partial f_{1}}{\partial x_{n}} \\
+%\vdots & \ddots & \vdots \\
+%\frac{\partial f_{m}}{\partial x_{1}} & \cdots & \frac{\partial f_{m}}{\partial x_{n}}
+%    \end{pmatrix}
+%        = \begin{pmatrix}
+%        \frac{\partial f_{1}}{\partial x_{1}}\\
+%        \frac{\partial f_{1}}{\partial x_{2}}\\
+%        \vdots\\
+%        \frac{\partial f_{1}}{\partial x_{n}}
+%    \end{pmatrix}
+%\end{equation}
+%For a scalar function $f\,:\,\mathbb{R}^n\rightarrow\mathbb{R}$, the VJP is simply the gradient 
+%$\nabla f$:
+%\begin{equation}
+%    \vec v^\intercal\,\partial f(\vec x)\equiv \nabla f(\vec x)=
+%    \left(1\right)\,\left(  \frac{\partial f}{\partial x_{1}} \,\cdots\,\frac{\partial f}{\partial x_{n}}   \right)
+%    =\begin{pmatrix}
+%        \frac{\partial f}{\partial x_{1}}\\
+%        \vdots\\
+%        \frac{\partial f}{\partial x_{n}}
+%    \end{pmatrix}
+%\end{equation}
+%
+%\subsubsection{Chain rule}
+%Let's now consider a function $f\,:\, \mathbb{R}^n\rightarrow\mathbb{R}^m$ which is a composition
+%three functions $a$, $b$ and $c$:
+%\begin{equation}
+%f=c \circ b \circ a
+%\end{equation}
+%The \textsl{chain rule} from calculus says that we can differentiate $f$ if we know how to differentiate 
+%each of its constituent parts:
+%\begin{align}
+%\partial f(\vec{x})&=\partial c(b(a(\vec{x}))) \circ \partial b(a(\vec{x})) \circ \partial a(\vec{x})\\
+%&=\partial c(b(a(\vec{x})))[\partial b(a(\vec{x}))[\partial a(\vec{x})]]
+%\label{eq:chain_rule}
+%\end{align}
+%The idea behind AD is to decompose complicated functions $f$ into constituent parts (sometimes 
+%called \textsl{primitive functions}) which we know how to differentiate  and then use the chain 
+%rule to obtain the derivative of $f$.
+%Equation~\ref{eq:chain_rule} consists of a series of matrix multiplications. Although the order
+%in which we multiply those matrices does not matter because of the associativity of matrix 
+%multiplication, it matters a great deal when we are implementing AD on a computer.
+%Let's say that we want to compute the JVP $\partial f(\vec{x})\,\vec v$ of this composite 
+%function:
+%\begin{equation}
+% \partial f(\vec{x})\,\vec v = \partial c(b(a(\vec{x})))[\partial b(a(\vec{x}))[\partial a(\vec{x})\vec v]]
+% \label{eq:jvp_chain_rule}
+%\end{equation}
+%Right-to-left evaluation of the expression on the right-hand side of 
+%Equation~\ref{eq:jvp_chain_rule} is known  as \textsl{forward-mode automatic differentiation}.
+%In forward-mode AD we evaluate \textsl{primal terms} $\vec{x}, a(\vec{x}), b(a(\vec{x}))$ 
+%alongside the \textsl{tangent terms} 
+%$\partial a(\vec{x}), \partial b(a(\vec{x})), \partial c(b(a(\vec{x})))$.
+%
+%If instead we are interested in evaluating the VJP the order of operations is reversed:
+%\begin{equation}
+%\partial f(\vec{x})^{\intercal}\,\vec u=\partial a(\vec{x})^{\intercal}\left[\partial b(a(\vec{x}))^\intercal\left[\partial c(b(a(\vec{x})))^{\intercal}\vec u\right]\right]
+%\label{eq:vjp_chain_rule}
+%\end{equation}
+%To evaluate  Equation~\ref{eq:vjp_chain_rule} we first have to traverse the whole chain 
+%computing the primals $\vec{x}, a(\vec{x}), b(a(\vec{x}))$ as we go and then we proceed in reverse order 
+%computing the tangents 
+%$\partial c(b(a(\vec{x})))^{\intercal}, \partial b(a(\vec{x}))^{\intercal}, \partial a(\vec{x})^{\intercal}$.
+%This process is known as \textsl{reverse-mode automatic differentiation} or 
+%\textsl{backpropagation} (since we are propagating the gradients backwards from the function output).
+%
+%Using JVPs and VJPs is much more memory efficient than computing the Jacobian matrix directly 
+%because all intermediate results computed when applying the chain rule are 1D vectors that 
+%don't take up a lot of memory. The key difference between reverse mode and  forward mode is that 
+%in the case of forward-mode AD we can just evaluate the JVPs as we evaluate the function itself 
+%whereas in reverse-mode AD we need to store the primal terms for the entire chain before 
+%backpropagating so reverse-mode AD is more memory intensive.
+%Reverse-mode is usually far more popular than forward-mode in optimization and machine learning 
+%because the function $f$ is some cost function with  a high-dimensional input and a scalar 
+%output so we only need one VJP to obtain the entire Jacobian matrix. There is also mixed-mode 
+%automatic differentiation which combines forward-mode and reverse-mode to optimize memory use.
+%
+%In the general case functions can be more complex than a simple chain of compositions and in AD 
+%frameworks they are represented as \textsl{Directed Acyclic Graphs} (DAGs), sometimes also 
+%called \textsl{computation graphs} or \textsl{circuits}. We point the reader to the Section 
+%6.2 in \citep{murphy_book_2023} for details on how to generalize AD to graphs and also higher 
+%order AD which is also possible using the same formalism. 
+%
+%\subsection{AD libraries}
+%\label{ssec:ad_libraries}
+%The big challenge in automatic differentiation is creating a library wich can knows how to compute 
+%JVPs and VJPs for each primitive operation (for example, multiplication, division, 
+%matrix multiplication, array slicing, etc.).
+%Because the primitive operations need to be coded 
+%manually, expressing a given computation (say an astrophysics code written in \ssf{C++}) in 
+%an AD library often requires some effort and it is a bit like programming in a different 
+%language. 
+%Popular AD libraries are generally developed by large groups of researchers and with the backing 
+%of major tech companies.
+%
+%There are two main kinds of AD libraries:
+%\begin{itemize}
+%    \item \textbf{Static graph libraries} 
+%    \item \textbf{Dynamic graph libraries}
+%\end{itemize}
+%The static graph libraries first build the entire computation graph for a given program before 
+%the graph is evaluated. Examples include \ssf{Theano}\footnote{Theano is no longer 
+%maintained but it has recently been ressurected and turned into \ssf{aesara} 
+%(\url{https://github.com/aesara-devs/aesara})} \citep{arXiv:1605.02688}, 
+%\ssf{TensorFlow}\footnote{\url{https://www.tensorflow.org/}} \citep{arXiv:1603.04467}, 
+%and the \ssf{Stan Math Library} \citep{arXiv:1509.07164}.
+%Dynamic graph libraries build the graph as the program is is being executed. Examples are 
+%\ssf{Tensorflow 2}, \ssf{JAX}\footnote{\url{https://jax.readthedocs.io/en/latest/}.} 
+%\citep{jax2018github} and \ssf{PyTorch}\footnote{\url{https://pytorch.org/}.} \citep{pytorch}.
+%These libraries have many features in addition to autodiff, most notably support for
+%for GPUs. They are written in Python but the primitive operations are usually 
+%expressed in a lower level language like \ssf{C++} and \ssf{CUDA C++} 
+%for GPU support. 
+%
+%The most recent and most research oriented of these libraries is Google's \ssf{JAX} 
+%library.
+%\ssf{JAX} is written in \ssf{Python} and it offers nearly identical functionallity 
+%to the widely popular \ssf{NumPy} \ssf{Python} library for manipulating 
+%multi-dimensional arrays. \ssf{JAX}takes Python code as an input, converts it 
+%to an intermediate language (IR) representation called \ssf{jaxpr} which is 
+%used to perform autodiff, and finally the resulting code for evaluating functions 
+%and their gradients is \textsl{Just-In-Time} (JIT) compiled to a lower level 
+%assembly like language called \ssf{HLO-IR} using the 
+%\ssf{XLA}\footnote{\url{https://www.tensorflow.org/xla}.} compiler in such a way 
+%that the code can be executed on different hardware architecutres such  as 
+%CPUs, GPUs and TPUs (Tensor Processing Units).
+%In practice this means that one can write \ssf{Python} code using a \ssf{NumPy} 
+%like library, easily compute gradients of the code through autodiff and get 
+%the performance of a compiled language like $C++$ through on-the-fly JIT 
+%compilation. \ssf{JAX} natively supports \ssf{Python} language constructs such 
+%as loops, conditionals and data structures such as tuples and dictionaries.
+%Almost everything you could write in \ssf{Python} is differentiable with 
+%\ssf{JAX} although there is no guarantee that the gradients are meaningful 
+%if the function represented by the code is not itself differentiable.
+%Implementing a real-world model in an AD framework often requires making individual 
+%functions ``smoother'' so they are more easily differentiable (see 
+%Chapter~\ref{ch:microlensing_modeling2} for more on this point).
+%
+%
+%Many of the statistical algorithms I have mentioned  in 
+%Section~\ref{ssec:sampling_methods} are now implemented in \ssf{JAX} and 
+%\ssf{JAX} has also been used within astronomy. \citet{2022ApJS..258...31K} built 
+%an autodifferentiable model for exoplanet spectrum modeling, 
+%the \ssf{exoplanet} code \citet{2021JOSS....6.3285F} for modeling transit light 
+%curves and radial velocity time series uses \ssf{JAX} as a backend, and 
+%\citet{2022arXiv220207663G} used \ssf{JAX} to build a differentiable model 
+%for strong lensing  which produces samples from a Bayesian posterior by  first
+%exploring the posterior using an optimizer, running variational inference to 
+%obtain an estimate of the posterior covariance followed by Hamiltonian Monte Carlo
+%to obtain the final samples. Importantly, \ssf{JAX} makes it relatively 
+%straightforward to implement custom primitives. One of the contributions of this 
+%thesis is an implementation of a custom JAX primitive for computing the roots 
+%of complex polynomials (see Chapter~\ref{ch:microlensing_modeling2} for more details).
+%The process of building custom primitives involves specifying the JVP rule 
+%(Equation~\ref{eq:jvp}). Interestingly, to support reverse-mode AD we it is 
+%not necessary to also specify the VJP rule (Equation~\ref{eq:vjp}) because 
+%\ssf{JAX} is able to deduce the transpose of the JVP automatically\footnote{
+%    Essentially, this is possible because the implementation of the JVP rule 
+%    in \ssf{Python} consists of several \ssf{JAX} primitives such as addition
+%    and multiplication and \ssf{JAX} already knows how to transpose those 
+%    primitives.} (see \citet{arXiv:2105.09469} for details). 
+%
+%\subsection{Probabilistic modeling + programming + autodiff = probabilistic programming}
+%The final topic in this chapter nicely ties up everything we have discussed so far.
+%The term \textsl{probabilistic programming} refers to the use of software libraries 
+%which  provide a high-level API for constructing statistical models (usually in 
+%the Bayesian paradigm). Specifically, 
+%the key components of modern probabilistic programming languages (PPLs) include a backend
+%autodiff framework, methods for initializing \emph{random variables} with 
+%associated pdf or pmf distribution functions and evaluation the (log) probability, 
+%and finally an one or several optimizers and samplers. The idea behind these 
+%frameworks is to provide an easy to use interface for constructing probabilistic 
+%models and to abstract away the inference as much as possible.
+%Why is this useful for scientists? Because these frameworks simplify the process  
+%of building probabilistic  models and because they include well-tested and high
+%performance implementations of state-of-the-art inference methods such as 
+%Hamiltonian Monte Carlo. 
+%
+%The most popular such library is \ssf{Stan} \citep{2017JSS....76....1C}. 
+%Stan is built in \ssf{C++}, it uses the \ssf{Stan Math Library} for automatic 
+%differentiation, it uses a variant of the NUTS HMC sampler and also has some 
+%support for Variational Inference. \ssf{Stan} is oriented primarily towards 
+%applications in the social sciences and it is not as flexible and easily extensible 
+%as some other PPLs. \ssf{PyMC} is another PPL which is very similar to 
+%\ssf{Stan} except that it is implemented in \ssf{Python} and it uses 
+%the \ssf{aesara}\footnote{\url{https://github.com/aesara-devs/aesara}.} library 
+%for AD (it used to be called \ssf{PyMC3} and use \ssf{Theano} for AD) and it 
+%also supports \ssf{JAX}. 
+%\ssf{PyMC} is somewhat more popular in the physical sciences and easier to 
+%extend  with custom functionality. The \ssf{exoplanet} code 
+%\citep{2021JOSS....6.3285F} is a good example of an astrophysics code which 
+%interfaces with \ssf{PyMC}.  There is also 
+%\ssf{numpyro}\footnote{\url{https://num.pyro.ai/}.}
+%\citep{arXiv:1912.11554} which is also built in \ssf{Python} and uses 
+%\ssf{JAX} for AD and \ssf{Tensorflow Probability} \citep{arXiv:1711.10604} 
+%which supports bot \ssf{TensorFlow} and \ssf{JAX} for AD. \ssf{PyMC}, \ssf{numpyro}
+%and \ssf{Tensorflow Probability} all use a variant of NUTS similar to the one 
+%implemented in \ssf{Stan} as the core algorithm for inference but also 
+%include support for Variational Inference and additional samplers. 
+%I use PPLs (mostly \ssf{numpyro}) extensively throughout the rest of this thesis.
+
+\chapter{\ssf{caustics} -- a differentiable code for computing the
+  magnification of single, binary and triple lens systems}
+\label{ch:caustics}
+
+\section{Introduction}
+\subsection{Overview of methods for computing the magnification}
+In order to fit microlensing light curves and use statistical methods such as
+MCMC we first need to be able to accurately and efficiently compute the
+microlensing magnification for an extended, possibly limb-darkened source star
+located at arbitrary positions in the source plane. There are no analytical
+solutions to this problem so we have to estimate the magnification numerically.
+We impose three main requirements on a prospective method which achievies this
+goal:
+\begin{enumerate}
+    \item The method needs to compute the magnification with high precision. In this work
+          by high precision we mean at most a $10^{-3}$ relative error for the
+          manification at any given point. This requirement is imposed by the photometric
+          precision of the light curves. The needed precision might be higher for data of
+          very high photometric quality.
+    \item The method needs to be fast. A necessary requirement for using statistical
+          optimization or sampling methods is that one can evaluate the model 10s of
+          thousands to millions of times in order to explore the parameter space. This
+          means that we should be able to compute the magnification for an entire
+          lightcurve (with potentially thousands of data points) in \emph{miliseconds}
+          rather than \emph{seconds} or \emph{minutes}.
+    \item The method should return the exact gradients of the magnification with respect
+          to all input parameters, obtained through automatic differentiation. This
+          enables the use of advanced statistical methods such as Hamiltonian MCMC and
+          also enables much faster optimization.
+    \item  The method should support both binary and triple lensing.
+\end{enumerate}
+The first and the second points are widely agreed upon but the third point did not
+get enough attention in the literature. The major contribution of this work is
+implementing a method which satisfies the third requirement in addition to the
+other three and demonstarting that this is worthwhile.
+
+There are three kinds main approaches to computing the magnification of
+extended sources:
+\begin{enumerate}
+    \item A brute-force aproach called \textsf{inverse ray shooting} (IRS)
+          \citep{1986A&A...166...36K,1997MNRAS.284..172W} which relies on evaluating the
+          lens equation in the direction $z\rightarrow w$ where the mapping is one-to-one
+          and then counting the the points which fall inside the source disc.
+    \item Computing the two-dimensional integral over the images using a Cartesian or
+          Polar grid covering the full extent of the images in the image plane. This
+          method is something also called \ssf{Image Centered Ray Shooting} (ICRS)
+          \citep{1996ApJ...472..660B,2010ApJ...716.1408B} and it is conecptually
+          different from IRS.
+    \item  Using \textsf{Green's theorem} to convert the two-dimensional integral over the
+          extent of the images into a one-dimensional integral over the boundaries of the
+          images
+          \citep{1986A&A...166...36K,1995A&AS..109..597D,1997ApJ...477..580G,1998A&A...333L..79D,2007MNRAS.377.1679D}.
+\end{enumerate}
+Ray shooting is by far the most conceptually and algorithmically simple approach and
+it does not require solving the lens equation $w=f(z)$ for the images $z$. The
+idea behind ray shooting is to first evaluate the lens equation $z \rightarrow w$ on
+a uniform density grid in the image plane $z$ which covers all the images. This
+is called `shooting rays' from the image plane to the source plane.
+The number of points (or \emph{rays}) which land within the source disc
+divided by the total number of points evaluated is then proportional to the
+magnification. The IRS method works with any number of lenses and source and
+because of this it is widely used in quasar microlensing.
+When it comes to planetary microlensing, the drawback of IRS is that the images
+are often very long thin arcs covering a vanishingly small fraction of the area of
+any rectangular grid in the image plane and thus requiring very dense and
+computationally expensive grids. Another drawback of ray shooting is that
+the counting operation (a 2D histogram) is a non-differentiable top-hat like
+function\footnote{Although one could replace the 2D histogram operation with a differentiable \ssf{Gaussian Kernel Density Estimation}.}
+
+The second method, ICRS, was first introduced by \citet{1996ApJ...472..660B}.
+The idea is quite straightforward, since the magnification of an extended
+source is the equal to the surface brigthness wieghted area of the images in
+the image plane divided by the the total flux of the source in absence of
+lensing, we can simply compute the two-dimensional integrals in the image plane
+numerically. \citet{1996ApJ...472..660B} used a rectangular grid to compute the
+integrals while \citet{2010ApJ...716.1408B} improved performance by switching
+to a Polar grid and deriving a custom numerical quadrature rule for computing
+the relevant integrals for a limb-darkened source. The drawback of this method
+is the same as in the case of IRS. Although the integration grids in ICRS are
+designed to be as small as they can be while still encompassing the images,
+often more than half of the pixels in a given grid do not belong to the image.
+In addition, using the higher order numerical quadrature rule derived in
+\citet{2010ApJ...716.1408B} for computing the integral in the radial direction
+requires solving an optimization problem at every grid point in the angular
+coordinate in order to determine the boundary of the image to a precision
+greater than the grid spacing in the radial direction. I have initially pursued
+this approach as the main algorithm in \ssf{caustics} but I found that the
+method is very convoluted, prone to failure when the source limb crosses one or
+multiple caustics, and computationally expensive\footnote{The main problem with
+    this method is accurately determining boundaries of the extended images,
+    especially when two images are nearly adjacent. Once you solve this problem
+    efficiently and robustly you might as well switch to contour integration using
+    Green's theorem and avoid the complexity and cost of evaluating two-dimensional
+    integrals.}.
+
+Finally, the contour integration method avoids constructing 2D integration
+grids entirely by making use of Green's theorem to convert an integral over an
+area into an integral over a 1D contour which encloses that area. The challenge
+is in constructing the contours. \ssf{VBBinaryLensing}
+\citep{2010MNRAS.408.2188B}, the most widely use code for binary microlensing
+uses contour integration. They construct the contours by solving the lens
+equation at multiple point on the source limb in the source plane which then
+map to locations on the extended image limbs in the image plane.
+\citet{2007MNRAS.377.1679D} uses an alternative approach, called \ssf{adaptive
+    contouring} which avoids solving the lens equation. Adaptive contouring
+recursively subdivides squares in the image plane using a \ssf{quadtree} data
+structure until points lying on the contours are located. The exact method used
+\citep{2010MNRAS.408.2188B} in does not easily generalize to the case of triple
+lensing, while adaptive contouring in principle works for any number of lenses
+although it has never been applied to the triple lens case.
+\citet{2021MNRAS.503.6143K} were the first to apply contour integration to
+triple lensing and this method (with some important changes and adjustments)
+forms the basis for for the algorithm in \ssf{caustics}.
+
+\subsection{\ssf{caustics}}
+
+\ssf{caustics}, a highly optimized and
+automatically differentiable \ssf{Python} code for computing the
+magnification of extended limb-darkened sources for single, binary and triple lens
+microlensing systems using contour integration. The code builds on previous work by
+\citet{2021MNRAS.503.6143K}, \citet{1998A&A...333L..79D},
+\citet{2007MNRAS.377.1679D} and \citet{2018MNRAS.479.5157B}.
+Novel features of \ssf{caustics} include the following:
+\begin{itemize}
+    \item \ssf{caustics} is the first microlensing code supporting automatic
+          differentiation which enables efficient evaluation of the exact gradients of
+          magnification with respect to any input. This is made possible by writing the
+          entire code in the \ssf{JAX} library (see Section~\ref{ssec:ad_libraries}).
+    \item Orders of magnitude faster execution for triple lens magnification than
+          alternatives and performance roughly on-par with alternatives in case of binary
+          lensing. \ssf{caustics} supports both CPU and GPU architectures thanks to
+          \ssf{JAX}.
+    \item  Novel implementation of a complex polynomial root solver using the
+          Ehrlich-Aberth algorithm which is faster than the widely used version of
+          Laguerre method from \citet{2012arXiv1203.1034S}. The algorithm is also
+          differentiable and it can be executed on a GPU in parallel for solving the lens
+          equation hundreds of thousands of times at once.
+    \item  The code is highly modular and easily extensible. Almost the exact same code is
+          used to compute the magnification of single, binary and triple lenses on both
+          CPUs and GPUs. The code could easily be extended to quadruple lensing as well
+          and it is very straightforward to plug-in an arbitrary source brightness
+          distribution (such as a \ssf{starry} map) instead of a simple limb-darkening
+          law.
+\end{itemize}
+
+Although \ssf{caustics} is written in \ssf{Python}, at runtime the entire code
+is JIT compiled with \ssf{JAX} to an intermediate representation which is then
+further optimized by \ssf{XLA} (see Section~\ref{ssec:ad_libraries}) which then
+executes the compiled code on a CPU or an (Nvidia) GPU. This enables \ssf{JAX}
+code to execute at speeds comparable and often better than classical compiled
+languages such as \ssf{C} and \ssf{C++} but it comes at a cost of having to
+statically specify the types and the shapes of the inputs and outputs to all
+functions. Thus we cannot have arrays whose shapes depend on the \emph{values}
+of function arguments. This was a design decision made by the \ssf{XLA} team in
+order to maximize the performance of \ssf{JAX} code. The way JIT compilation
+works is that an empty abstract array (called \ssf{ShapedArray} in \ssf{JAX})
+with a static shape and data type is first passed to a given function and then
+the entire computation graph of the function (specifying the shapes and data
+types of all intermediate values) is traced with this abstract array so the XLA
+compiler knows exactly what to expect for any given input to the function of
+the same shape and data type.
+
+Another complication is implementing control flow operations. \ssf{Python}
+constructs such as \ssf{if} statements and \ssf{for} and \ssf{while} loops have
+to be replaced with corresponding \ssf{JAX} operators \ssf{lax.cond},
+\ssf{lax.scan} and \ssf{lax.while\_loop} (\ssf{lax} is a subset of the
+\ssf{JAX} library). Using a \ssf{Python} \ssf{if} statement will automatically
+trigger a new JIT compilation at every branching point. A \ssf{for} loop will
+be automatically unrolled which is ok for small loops but prohibitively
+expensive in memory use for large loops which is it is preferable to use
+\ssf{lax.scan} instead. \ssf{lax.cond} and \ssf{lax.scan} are both forward and
+reverse-mode differentiable but \ssf{lax.while\_loop} is only forward-mode
+differentiable. One other key complication is that although at runtime
+\ssf{lax.cond} will execute only one branch depending on some condition (like a
+classic \ssf{if} statement), XLA requires allocating memory for \emph{both}
+branches. A result of this constraint is that forward mode autodiff is a better
+choice for this application than reverse mode autodiff.
+
+This rest of this chapter is structured as follows. In
+Section~\ref{sec:single_lens_mag} I cover the case of single lense
+magnification using the same formalism I later use for binary and triple
+lenses. I do this because it is easier to test that the algorithm works
+properly in that case and because it is instructive to understand how the
+single lens case differs from binary and triple lens cases.
+Section~\ref{ssec:root_solver} covers the root solver which is a necessary
+component for the binary and triple lens magnification. Finally,
+Section~\ref{sec:binary_and_triple_mag} covers the considerably more
+complicated case of binary and triple lensing.
+
+\section{Single lens magnification}
+\label{sec:single_lens_mag}
+
+Consider a single lens located at the origin. The lens equation in complex
+coordinates (Equation~\ref{eq:lens_equation_complex}) for the $N=1$ case
+reduces to the following simple expression:
+\begin{equation}
+    w = z - \frac{1}{\bar{z}}
+    \label{eq:lens_eq_single}
+\end{equation}
+By conjugating Equation~\ref{eq:lens_eq_single}, solving the resulting expression
+for $\bar{z}$ and plugging it back into Equation~\ref{eq:lens_eq_single}, we
+obtain a 2nd degree complex polynomial equation:
+\begin{equation}
+    \bar{w}z^2-|w|^2 z - w = 0
+    \label{eq:complex_poly_single}
+\end{equation}
+The quadratic polynomial has exactly two roots and both roots are valid images
+of the point source located at $w$. Using the quadratic formula, the roots are
+\begin{equation}
+    z_{1,2} = \frac{w}{2}\left(1 \pm \sqrt{1 +\frac{4}{|w|^2}}\right)
+    \label{eq:images_single_lens}
+\end{equation}
+Thes are the \emph{point source images} in complex plane $z$. In case of an
+extended circular source
+with radius $\rho_\star$ located at $w$ we do not observe these point source
+images but rather the \emph{physical images}\footnote{These images are what we would
+    see if we could resolve the microlensing event.}. I will use the word images
+to refer to both the point source images which are points in the $z$ planes and
+the images of an extended source which are \emph{closed curves} in the $z$ plane.
+The source
+brightness weighted area of these physical images divided by the the total flux
+of the unlensed source is the magnification we are interested in.
+If we parametrize the source  position as $w=w_0+\rho_\star e^{i\phi}$ we
+can obtain a version of Equation~\ref{eq:images_single_lens} parametrized by
+the position on the source limb $\phi$ \citep{1994ApJ...430..505W}:
+\begin{equation}
+    z_{1,2}(\phi)=\frac{w_0+\rho_\star e^{i \phi}}{2}\left[1 \pm \sqrt{1+\frac{4}{w_0^{2}+2 \rho_\star
+    w_0 \cos \phi+\rho_\star^{2}}}\;\right], \quad 0 \leq \phi\leq 2 \pi\hquad .
+\end{equation}
+Following \citet{1994ApJ...430..505W} differentiate between  three different
+regimes (see Figure~\ref{fig:single_lens_images} for visuals):
+\begin{enumerate}
+    \item 1. For $|w_0| < \rho_\star$ the extende images are merged in a ring.
+    \item 2. For $\rho_\star < |w_0| \ll 2$ the extended images form elongated arcs.
+    \item 3. For $|w_0| \gtrsim  2$ the images are slightly deformed  circles.
+\end{enumerate}
+\citet{1994ApJ...430..505W} shows that if we assume that the source brightness is
+uniform across the disc one can solve for the extended source magnification analytically
+and the solution is composed of complete elliptical integrals of the first, second
+and third kinds.
+
+There is no analytical solution for an arbitrary brightness profile but
+\citet{2009ApJ...695..200L} showed that it is straightforward to simply compute
+the two-dimensional integral in the source plane over the source disc:
+\begin{equation}
+    A_\mathrm{tot}(|w_0|,\rho_\star)\equiv \frac{\oiint A_\mathrm{ps}(w)\,I(w - w_0)\ud S}{
+        \oiint I(w - w_0)\ud S}
+    \label{eq:mag_direct_integration}
+\end{equation}
+$A_\mathrm{tot}$ is the extended source magnification, $S$ is the projected
+disc of the source, $A_\mathrm{ps}$ is the point source magnification
+(Equation~\ref{eq:magnification_point_lens}) and $I$ is the source brightness.
+The integral in Equation~\ref{eq:mag_direct_integration} is straightforward to
+compute numerically because the caustic for
+a single lens (curve where $A_\mathrm{ps}$ diverges) is a single point and
+the magnification pattern is azimuthally symmetric so we can integrate over
+the half-disc and avoid the singularity(see \citet{2009ApJ...695..200L}).
+
+\section{A differentiable complex polynomial root solver}
+No analytical solution exists for the lens equation with $N>1$ lenses. To obtain the point
+source images we have to numerically estimate the roots of a complex polynomial 
+equation of degree $N$:
+\begin{equation}
+    p(z)=\sum_{i=0}^{N} a_i z^i
+    \label{eq:general_complex_poly}
+\end{equation}
+where $a_i$ are the coefficients of the polynomial which are derived from the lens equation
+(see Appendix~\ref{app:complex_poly} for details). Those roots which satisfy the lens 
+equation (a subset of all roots) are valid images.
+In case of a binary lens this complex polynomial is 5th order  and for a triple lens it is 10th 
+order. Solving for the roots of a general polynomial is a well-known problem in numerical methods 
+and there exists a variety of algorithms built for this purpose.
+We can divide these algorithms in two two groups. 
+Those which cast the root solving problem as an eigenvalue problem using the fact that 
+an arbitrary complex polynomial can be written as a characteristic polynomial of a 
+particular nonsymmetric square matrix, and those which search for the roots directly making 
+use of the gradients of the polynomial. \ssf{JAX} includes a general algorithm for computing
+the roots of a complex polynomial \ssf{jax.numpy.roots} which uses the eigenvalue 
+approach but it is too slow for microlensing purposes so we decided to implement a faster
+algorithm using the direct approach.
+
+The simplest method for updating a numerical estimate  of the $j$-th root $z_j$ given 
+some initial estimate is the first order Newton's method:
+\begin{equation}
+    z_j \leftarrow z_j - \frac{p(z_j)}{p'(z_j)}
+\end{equation}
+which makes use of the first derivative of the polynomial.
+A more sophisticated method with faster convergence is the second order Lageuerre's method
+which makes use of the second derivative of the polynomial:
+\begin{equation}
+    z_j\leftarrow z_j - \frac{n}{G_{j} \pm \sqrt{(n-1)\left(n H_{j}-G_{j}^{2}\right)}}\hquad ,
+    \label{eq:laguerre_update}
+\end{equation}
+where $G_j=p^\prime (z_j)/p(z_j)$ and $H_j=-(p^\prime (z_j)/p(z_j))^\prime$ and the sign
+in the denominator is chosen such that it maximizes the absolute value of the denominator.
+Applying these update rules to the full polynomial can result in convergence to the same 
+root repeatedly. For this reason the roots are found  one by one using a so 
+called \emph{deflation strategy}. With \emph{explicit deflation}, after 
+each root is found the polynomial is divided by that root using polynomial division so 
+that we  obtain a new  polynomial with degree one less than the original which still 
+shares the other roots with the original polynomial.
+The procedure is repeated until all roots have been found. Because the division
+process introduces some numerical noise, explicit deflation requires a final ``polishing''
+step after all roots have been found in order to fine tune the estimates of the roots. This 
+polishing step  consists of additional applications of the Laguerre or Newton updates to 
+each of the roots.
+
+An alternative to an explicit deflation strategy is an \emph{implicit deflation}
+strategy\footnote{As far as I am aware this strategy has not been previously mentioned in the 
+microlensing literature.} in which the polynomial stays the same but the 
+method is modified. The simplest approach to implementing implicit deflation starts with
+the application of Newton's method to a reduced polynomial $p_j(z)$:
+\begin{equation}
+    p_{j}(z) \equiv \frac{p(z)}{\left(z-z_1\right) \cdots\left(z-z_{j-1}\right)}\hquad ,
+    \label{eq:reduced_poly}
+\end{equation}
+where $z_1\cdots z_{j-1}$ are the roots of the polynomial we have already found and
+$z_j$ is the root we are searching for.
+The derivative of this reduced polynomial, $p^\prime_j(z)$, is
+\begin{equation}
+    p_{j}^{\prime}(z)=\frac{p^{\prime}(z)}{\left(z-z_1\right) \cdots\left(z-z_{j-1}\right)}-\frac{p(z)}{\left(z-z_1\right) \cdots\left(z-z_{j-1}\right)} \sum_{i=1}^{j-1}\left(z-z_{i}\right)^{-1}\hquad .
+\end{equation}
+We can then write down the Newton update for the reduced polynomial 
+$-p_j(z)/p_j^{\prime}(z)$
+as
+\begin{equation}
+    z_j \leftarrow z_j - \frac{1}{\frac{p^\prime (z_j)}{p(z_j)} - \sum_{i=1}^{j-1}\left(z_j-z_{i}\right)^{-1}}\hquad .
+    \label{eq:newton_maehly}
+\end{equation}
+Equation~\ref{eq:newton_maehly} contains only evaluations of the original polynomial 
+and a sum over the previous roots $z_i$, thus avoiding the need for doing explicit polynomial 
+division.  This is known as \ssf{Maehly's procedure} and root solver implementing the 
+Maehly procedure is sometimes called the \ssf{Newton-Maehly} method.
+The key feature of the implicit deflation strategy is that it creates poles at the
+values of the previous roots (via the $(z_j-z_i)^{-1}$ term), thus avoiding multiple
+convergences to the same root.
+
+It is also possible to obtain all roots simultaneously using a procedure very similar to 
+the Newton-Maehly method called the \ssf{Aberth-Ehrlich} (\ssf{B\"{o}rsch-Supan}) method 
+(AE for short). The only difference is that instead of applying Newton's method to a 
+polynomial divided by the first $j-1$ roots like in Equation~\ref{eq:reduced_poly}, we 
+apply it to a polynomial divided by the estimates of all the other roots except $z_j$:
+\begin{equation}
+    \tilde{p}_{j}(z) \equiv \frac{p(z)}{\prod_{i=1;i\neq j}^{n}(z-z_i)}\hquad .
+    \label{eq:reduced_poly2}
+\end{equation}
+It follows that the update rule is then 
+\begin{equation}
+    z_j \leftarrow z_j - \frac{1}{\frac{p^\prime (z_j)}{p(z_j)} - \sum_{i=1\,;i\neq j}^{n}\left(z_j-z_{i}\right)^{-1}}\hquad .
+    \label{eq:aberth_ehrlich_update}
+\end{equation}
+Starting from some initial estimate of all the roots we simply apply the update in 
+Equation~\ref{eq:aberth_ehrlich_update} to each of the roots in parallel until all 
+roots satisfy some convergence criterion. The AE algorithm is superior to the Newton-Maehly
+method because we can evaluate the roots in parallel.
+
+What if instead of applying Newton's rule  to the reduced polynomial in 
+Equation~\ref{eq:reduced_poly2} we applied Laguerre's rule instead?  \citet{cameron2019} 
+do exactly that to obtain a modified Lauguerre's method with an update rule that is 
+identical to Equation~\ref{eq:laguerre_update} except that factors $G_j$ and $H_j$ are 
+replaced with
+\begin{equation}
+    G_{j}=\frac{p^{\prime}\left(z_{j}\right)}{p\left(z_{j}\right)}-
+    \sum_{\substack{i=1 \\ i \neq j}}^{n} \frac{1}{\left(z_{j}-z_{i}\right)} \text { and } H_{j}=-\left(\frac{p^{\prime}\left(z_{j}\right)}{p\left(z_{j}\right)}\right)^{\prime}-\sum_{\substack{i=1 \\ i \neq j}}^{n} \frac{1}{\left(z_{j}-z_{i}\right)^{2}}
+\end{equation}
+The advantage of this modified Lagueerre method over the AE method is faster convergence 
+thanks to the second order Lageurre update (though the Laguerre update is more 
+computationally costly than the first order Newton update in the AE algorithm).
+
+The two most popular implementations of complex polynomial root solver alogorithms are
+the \ssf{zroots} routine from Numerical Recipes \citep{1992nrca.book.....P} and the
+\ssf{C02AFF} routine from the NAG library. The two routines used to both use
+Laguerre's method with an explicit deflation strategy but recently switched to the 
+modified Laguerre's method from \citet{cameron2019}\footnote{In fact, the name
+    \ssf{C02AFF} refers to the old routine and the new one is called \ssf{C02AAF}:
+    \url{https://www.nag.com/numeric/nl/nagdoc_latest/flhtml/c02/c02aaf.html}}. 
+   \ssf{zroots}  has been used in microlensing codes in the past but it has  been replaced
+   with the algorithm presented in \citet{2012arXiv1203.1034S}
+which I will refer to as \ssf{SGroots} from now on. \ssf{SGroots} is a slight
+modification of the old \ssf{zroots} routine and it uses the explicit deflationary
+strategy. It switches between Laguerre's method and Newton's method to improve
+performance and it includes some optimizations tailored specifically to the
+properties of $5$th order complex polynomial derived from the binary lens equation, 
+namely, the fact that two of the five roots are always well separated in the complex plane. 
+\citet{2012arXiv1203.1034S} claim that their algorithm is
+1.6--3 times faster than the (old) \ssf{zroots}. One other key difference between 
+\ssf{SGroots} and \ssf{zroots} is that \ssf{zroots}, like all other Numerical Recipes 
+algorithms, is released under an extremely restrictive proprietary license while 
+\ssf{SGroots} is open-source.
+
+In \ssf{caustics} I chose to use the Aberth-Ehrlich algorithm, for several 
+reasons. First, there exists a well-tested open-source implementation of the method written 
+in \ssf{C}\footnote{\url{https://github.com/trcameron/CompEA}} which accompanies the paper 
+by \citet{cameron2021}. Second, the AE algorithm uses implicit deflation thus  avoiding the 
+need for polishing, in contrast to \ssf{SGroots}. Third, if we have a very good initial 
+estimate for the roots (which is the case when solving for the images at closesly spaced 
+location in the source plane) it will be cheaper than the  modified Laguerre method because, 
+as pointed out by \citet{2012arXiv1203.1034S}, in the limit of vanishing $p(z)$ the 
+Laguerre update is identical to Newton's update while being twice as expensive.
+For this reason \ssf{SGroots} switches to Newton's method in the polishing step. Finally,
+\citet{cameron2021} introduced a version of the Aberth-Ehrlich method which uses
+compensated floating point arithmetic\footnote{Compensated arithmetic reduces  numerical
+error in additions of floating point numbers by tracking a separate variable which tracks 
+the accumulation of the errors.}, most importantly the compensated Horner's method 
+for evaluating a polynomial and its derivative, in order to effectively double the 
+working numerical precision from 64 bits to 128 bits while still using 64 bit variables.
+\citet{2010ApJ...716.1408B} mentions that quadruple precision may be necessary
+for triple lens systems with extreme mass ratios such as a system with an Earth
+mass planet and a Moon like satellite.
+
+There are two ways we could implement this algorithm in \ssf{JAX}. The first
+approach is to write the whole algorithm from scratch in \ssf{Python} using \ssf{JAX} 
+constructs. This is most often the simplest approach and it often results in code that 
+(when JIT-compiled) is as fast as \ssf{C} or \ssf{C++} code. The second approach is to 
+write the algorithm in \ssf{C} or \ssf{C++} and then incorporate it into \ssf{JAX} as a 
+custom primitive. Although it time consuming, I opted for the second approach because 
+I could use the existing code from \citet{cameron2021}. The Python wrapper around the 
+\ssf{C/C++} code tells \ssf{JAX} (more specifically, \ssf{XLA}) how the program transforms 
+under \ssf{JAX} program transformations \ssf{jit} (JIT compilation),
+\ssf{grad} (reverse mode autodiff) and \ssf{vmap} (vectorization over a multi-dimensional 
+input).
+
+\subsubsection{Differentiating through a root solver using the implicit function theorem}
+
+In order to support forward and reverse mode automatic differentiation for a code containing
+an iterative root solving algorithm it is necessary to specify a custom JVP product rule 
+for the primitive operation (in this case, a complex polynomial root solver) because 
+\ssf{JAX} doesn't know how to propagate gradients through the root solver writte in 
+\ssf{C/C++}.  We could technically avoid doing this by using an autodiff library which can
+differentiate \ssf{C/C++} code\footnote{The recent
+    \href{https://enzyme.mit.edu/}{Enzyme} library can differentiate through
+    programs written in compiled languages such as \ssf{C/C++}. It works by
+    differentiatating through LLVM IR representation of the code and it can even differentiate 
+    through \ssf{CUDA} kernels.} but it turns out there is a far better approach. 
+    Instead of differentiating through all the unrolled operations of the iterative 
+root solver which consumes a lot of memory, we can make use of the \ssf{implicit function 
+theorem} to obtain the derivative of the function outputs (the roots) with respect to the 
+inputs (the polynomial coefficients) by linearizing the polynomial at the solution point.
+Implicit function theorem is often used in autodiff frameworks and ML libraries  to simplify 
+differentiation through iterative solvers such as root solvers, nonlinear equation solvers
+and ODE solvers\footnote{
+    For more details see these notes on the applications of implicit differentiation in Deep 
+    Learning: \url{http://implicit-layers-tutorial.org/implicit_functions/}.
+} under the condition that we can linearize the function in the neighborhood of the solution
+to the iterative algorithm.
+
+Let $f(z, \vec{a})$ be the polynomial from Equation~\ref{eq:general_complex_poly} where 
+$\vec{a}$ is a vector of coefficients $a_i$ and let $z^*$ be a particular root of the 
+polynomial which we can compute using the AE algorithm. We cannot write down an
+expression for the roots in closed form but we can say that there is some
+implicit function $h(\vec a)\,:\,\mathbb{C}^{n+1}\rightarrow\mathbb{C}$ which maps the 
+polynomial coefficients to a given root $z^*$:
+\begin{equation}
+    h(\vec{a})=z^*\hquad .
+\end{equation}
+ We are interested in writing down the
+Jacobian $\partial_{\vec a} h(\vec{a})=(\partial_{a_0} h(\vec a),\ldots,\partial_{a_n} h(\vec{a}))^\intercal$ 
+containing derivatives of $h(\vec{a})$ with respect to the
+coefficients $a_i$  evaluated at a root $z^*$. This is the derivative we need to be
+to evaluate in order to evaluate the derivative of the magnification
+(which is a function of the roots) with respect to the physical parameters
+(which determine the polynomial coefficients). The implicit function theorem
+states that
+\begin{equation}
+    \partial_{\vec a}h(\vec{a})=-\left[\partial_z\,f(z^*,\vec{a})\right]^{-1}\,\partial_{\vec a}\,f (z^*,\vec a) \hquad .
+    \label{eq:implicit_function_theorem}
+\end{equation}
+The first part of the equation is the first derivative of the polynomial evaluated at the
+root (which is just another polynomial with degree one less than the original) and the second part 
+is the derivative of the polynomial with respect to each of the
+coefficients evaluated at the root. To obtain the full Jacobian matrix  with shape 
+$(n+1, n)$ containing the  derivatives of each of the $n$ roots with respect to each of 
+the $n+1$ coefficients  we can simply apply Equation~\ref{eq:implicit_function_theorem} 
+to each of the roots $z^*$.
+
+I have ported the code from \citet{cameron2021} into \ssf{C++} and wrote a custom \ssf{JAX}
+primitive\footnote{Following this excellent tutorial by Daniel Foreman-Mackey: 
+\url{https://dfm.io/posts/extending-jax/}.} which implements a JVP rule 
+$\left(\vec{a}, \vec v\right) \mapsto\left(z^{*}\left(\vec a\right), \partial z^{*}\left(\vec a\right) \vec v\right)$
+using Equation~\ref{eq:implicit_function_theorem}. As mentioned
+in Section~\ref{sec:autodiff}, \ssf{JAX} can derive the corresponding VJP rule for 
+reverse mode automatic differentiation automatically. In addition to implementing a 
+custom JVP rule, in order to support GPUs (which almost all native \ssf{JAX} primitives 
+support by default), I needed to port the \ssf{C++} code  for the AE solver to Nvidia's 
+\ssf{CUDA} framework\footnote{\ssf{CUDA} enables execution 
+of \ssf{C} and \ssf{C++} code on Nvidia GPUs.}. I wrote the code such that the root 
+sovler is parallelized over multiple polynomials and each \ssf{CUDA} thread is 
+responsible for solving a single polynomial. The main code implementing the root solver is
+called on both CPUs and GPUs and the only difference is that in the former case it is 
+executed sequentially and in the latter case it is parallelized over multiple polynomials.
+
+The code also alows the user to switch between the compensated  and the non-compensated 
+versions of the AE algorithm. The code uses the non-compensated version by default because 
+it is about two times faster than the non-compensated version. The GPU version of the 
+solver enables solving for the roots of hundreds of thousands of binary or triple lens 
+polynomials (and therefore also the point source magnification) in 10s of miliseconds.  
+This enables the computation of 1000 X 1000 point source magnification maps 
+in a fraction of a second. Unfortunately, despite this speedup, the GPU version of the 
+root solver isn't particularly useful for the root solving stage of the contour integration 
+algorithm because  contour integration requires solving for the roots sequentially 
+which makes it unsuitable for parallel processors.
+\citet{mcdougall2014} implemented a \ssf{CUDA} version of the \ssf{SGroots} algorithm 
+on a GPU but they used  a (highly inaccurate) direct integration approach for 
+computing binary lens magnification which does not require solving for the roots 
+sequentially.
+
+I have compared the CPU version of the algorithm to the implementation of \ssf{SGroots} 
+in \ssf{VBBinaryLensing} by solving a 5th order complex polynomial and found that 
+it is about 40\% slower using the default initialization strategy and the default 
+stopping criteria in each algorithm. This difference in speed is likely due to additional 
+optimizations in \ssf{SGroots} specific to the binary lens polynomial and perhaps also due 
+to  a less stringent stopping criterion. Ultimately this difference in performance is 
+irrelavant since, as we shall see in the following section, we only need to run the 
+root solver once  with the default initialization strategy while all subsequent evaluations 
+use a starting  point for the roots which is very close the the actual roots. 
+Since both algorithms use Newton updates in that regime I expect that the practical 
+difference in speed is negiligble, although I haven't done extensive tests. It is also 
+worth pointing out that \citet{2022MNRAS.514.4379F} also implemented an AE solver for 
+the purpose of solving the binary lens equation and found that their implementation 
+is about twice as fast as \ssf{SGroots}.
+
+I have tested the validity of code extensively by comparing its output to the 
+default polynomial root solver in \ssf{JAX}, \ssf{jax.numpy.roots}. The \ssf{JAX} 
+implementation of the AE algorithm is up to an order of magnitude faster than 
+\ssf{jax.numpy.roots}. The default solver also does not have support for GPUs because 
+at the time of writing \ssf{JAX} does not support a GPU solver for an eigendecomposition 
+of a nonsymmeric matrix.
+I have also tested that the gradients obtained using the implicit function theorem are 
+correct by comparing them to gradients obtained through the finite 
+difference approximation. All these tests are implemented in the directory
+\ssf{tests/test\_ehrlich\_aberth\_primitive.py}.
+
+\section{Constructing the image contours}
+
+For $N>1$ the caustics are complex two-dimensional curves in the source plane
+$w$ and the direct integration approach is not feasible because computing
+two-dimensional integrals with a diverging integrand is highly inaccurate.
+Because of this, all microlensing software packages compute integrals in the
+image plane $z$:
+\begin{equation}
+    A_\mathrm{tot}(|w_0(z)|,\rho_\star)\equiv \frac{\oiint I(w(z) - w_0(z))\ud S^\prime}{
+        \oiint I(w(z) - w_0(z))\ud S}\hquad,
+    \label{eq:mag_integration_image_plane}
+\end{equation}
+wher $w(z)$ is the lens equation (Equation~\ref{eq:lens_equation_complex}).
+The integral in Equation~\ref{eq:mag_integration_image_plane} does not have the
+pathological function $A_\mathrm{ps}$ but the challenge is in computing the
+integral over a the image region $S^\prime$.
+
+As mentioned in the introduction, different methods are used to integrate the
+over the region $S^\prime$ covering the images. In this work we make use of
+Green's theorem\footnote{Green's theorem is the two-dimensional special case of
+    Stokes' theorem.} to reduce the dimensionality of the problem and transform the
+integral over $S^\prime$ to an integral over the \emph{boundary} of $\partial S^\prime$. For two continous functions $P(z_1, z_2)$ and $Q(z_1, z_2)$
+with continous partial derivatives $\partial_{z_2}P$ and $\partial_{z_1}Q$, where $z_1\equiv\mathrm{Re}(z)$ and $z_2\equiv\mathrm{Im}(z)$, we
+have:
+\begin{equation}
+    \oiint_{{S}^\prime}\left(\frac{\partial Q}{\partial z_{1}}-\frac{\partial P}{\partial z_{2}} \right) \ud z_{1} \ud z_{2}= \oint_{C} (P \ud z_{1}+Q \ud
+    z_{2})\hquad . \label{eq:green_theorem}
+\end{equation}
+The path of the integration along the curve $C$ is defined to be
+\textbf{counterclockwise} meaning that if we have a curve that goes in the clockwise
+direction we have to flip the sign of the integral to get the correct result.
+
+For now we assume that the source brightness is uniform across the disc. In
+this case the magnfication is simply given by the total area of the images
+divided by $\pi\rho_\star^2$. This means that the expression in the parenthesis
+of Equation~\ref{eq:green_theorem} has to be equal to 1, a requirement
+satisfied by functions $P(z_1, z_2)\equiv -z_2/2$ and $Q(z_1, z_2)\equiv
+    z_1/2$. It follows that the total magnification is then
+\begin{equation}
+    A_\mathrm{tot}\left(w_0, \rho_{\star}\right)=\frac{1}{\pi \rho_{\star}^{2}}
+    \int_{\{C_k\}} \frac{1}{2}(z_{1} \ud z_{2}-z_{2} \ud z_{1})\hquad ,
+    \label{eq:mag_green_theorem_unif}
+\end{equation}
+where $\{C_k\}$ a set of all $k=1,\dots,k$ closed contours comprising the images.
+When computing the  above intergral for each of the $K$ images  we also have to
+take into account the \textsl{parity} of each image $p_k=\pm 1$ which is defined to
+be equal to the sign of the Jacobian determinant of the lens mapping. The expanded
+form of Equation~\ref{eq:mag_green_theorem_unif} is then
+\begin{equation}
+    A_\mathrm{tot}\left(w_0, \rho_{\star}\right)=\frac{1}{\pi \rho_{\star}^{2}}
+    \sum_{k=1}^Kp_k\int_{C_k} \frac{1}{2}(z_{1} \ud z_{2}-z_{2} \ud z_{1})\hquad ,
+    \label{eq:mag_green_theorem_unif_expanded}
+\end{equation}
+
 \begin{figure}[t]
     \begin{centering}
-        \includegraphics[width=\linewidth]{figures/spherical_harmonics.pdf}
-        \caption{Real spherical harmonics up to degree $l=5$ computed from
-            Equation~\ref{eq:spherical_harmonics}. Figure adapted from Figure 1 in
-            \citet{2019AJ....157...64L}.}
-        \label{fig:spherical_harmonics}
+        \includegraphics[width=\linewidth]{figures/limb_sampling2.pdf}
+        \caption{The effect of using the iterative evaluation of the point
+            source images (see Figure~\ref{fig:limb_sampling1}) on the density of
+            extended image contours (orange and blue curves).  The left panel shows
+            the contours corresponding to a uniform sampling of the source limb while
+            the right panel shows the contours after 15 iterations of the adaptive
+            sampling algorithm. We display the ordering of contour points within a
+            single contour using  progressively smaller circles. The colour indicates
+            the parity of the extended image (1 for blue, -1 for orange).
+            The orange contour starts close
+            to the point (0.3, -0.7) and continues in a clockwise direction back
+            to the starting point. The blue contour starts close to the point
+            (-0.7, 0.7) and continues in a counter-clockwise direction back to its
+            starting point. The density map shows the (limb-darkened) source
+            intensity evaluated on a fine grid in the image plane. Finally, the small
+            inset plots display the (point source) magnification map in the source plane
+            and the extent of the source (black circle).
+        }
+        \label{fig:limb_sampling1}
     \end{centering}
 \end{figure}
-\subsubsection{Emitted light}
-Consider a spherical body of unit radius emitting light isotropically (a
-Lambertian emitter) at each point $(\theta,\phi)$ where $\theta$ is the
-inclination angle and $\phi$ is the azimuthal angle. To compute a phase curve
-or an occultation light curve we need to be able to integrate the flux over the
-entire projected disc of the body given the distribution of specific intensity
-$I(\theta, \phi)$, the 3D orientation of the body in space, and the location of
-a possible occultor relative to the body. These two-dimensional integrals can
-always be computed numerically (see for instance the approaches presented in
-\citet{2018AJ....156..146F,2018MNRAS.477.2613L}) but
-\citet{2019AJ....157...64L} showed that it is possible to compute them
-analytically if the specific intensity is first expanded in a basis of
-spherical harmonics and then mapped to a different basis more suitable for
-evaluating the integrals.
-
-Following \citet{2019AJ....157...64L}, we start by setting up a cartesian
-coordinate system on the units sphere, such that
-\begin{align}
-     & x=\sin \theta \cos \phi \\
-     & y=\sin \theta \sin \phi \\
-     & z=\cos \theta
-\end{align}
-and the observer is located on the $z$-axis at $\infty$ such that the projected disc
-of the body is centered at the origin of the $xy$ plane with the unit vector
-$\hat{\mathbf{x}}$ pointing to the right and $\hat{\mathbf{y}}$ pointing up.
-We introduce spherical harmonics $Y_{l m}(\theta, \phi)$ of degree $l\geq 0$ and order
-$m\in [-l, l]$ defined  as
-\begin{equation}
-    Y_{l m}(\theta, \phi)= \begin{cases}\bar{P}_{l m}(\cos \theta) \cos (m \phi) & m \geqslant 0 \\ \bar{P}_{l|m|}(\cos \theta) \sin (|m| \phi) & m<0\end{cases}
-    \label{eq:spherical_harmonics}
-\end{equation}
-When the spherical harmonics defined above are rewritten in terms of $x$, $y$, and $z$
-they become polynomials in these variables (see Appendix A in \citet{2019AJ....157...64L}).
-We can expand the specific intensity distribution $I(x,y)$ in a spherical harmonic basis
-as
-\begin{equation}
-    I(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
-    \label{eq:sh_expansion}
-\end{equation}
-where $\tilde{\mathbf{y}}$ is the basis vector of spherical harmonics  arranged in
-increasing order:
-\begin{align}
-     & \tilde{\mathbf{y}}=\left(\begin{array}{lllll}
-                                    Y_{0,0} & Y_{1,-1} & Y_{1,0} & Y_{1,1} & Y_{2,-2}
-                                \end{array}\right. \\
-     & \left.\begin{array}{lllll}
-                 Y_{2,-1} & Y_{2,0} & Y_{2,1} & Y_{2,2} & \cdots
-             \end{array}\right)^{\top} \quad,
-\end{align}
-and $\mathbf{y}$ is a vector of scalar \emph{spherical harmonic coefficients}.
-$\mathbf{y}$ is a central quantity which defines the \emph{map} (the specific
-intensity at every point) of the body.
-\citet{2019AJ....157...64L} shows that we can represent $y$ in a polynomial basis (see
-\citet{2019AJ....157...64L} for the general expression)
-\begin{equation}
-    \tilde{\boldsymbol{p}}=\left(\begin{array}{llllllllll}
-        1 & x & z & y & x^{2} & x z & x y & y z & y^{2} & \cdots
-    \end{array}\right)^{\top}
-\end{equation}
-such that
-\begin{align}
-    I(x, y) & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{p}                    \\
-            & =\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{y}\quad,
-    \label{eq:intensity_poly_basis}
-\end{align}
-and also in the Green's basis
-\begin{equation}
-    \tilde{\boldsymbol{g}}=\left(\begin{array}{llllllllll}
-        1 & 2 x & z & y & 3 x^{2} & -3 x z & 2 x y & 3 y z & y^{2} & \cdots
-    \end{array}\right)^{\top}
-\end{equation}
-such that
-\begin{align}
-    I(x, y) & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{g}              \\
-            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A}_2 \mathbf{p} \\
-            & =\tilde{\mathbf{g}}^{\top}(x, y) \mathbf{A} \mathbf{y}
-\end{align}
-where $\mathbf{A}_1$ is the change-of-basis matrix from $\mathbf{y}$ to $\mathbf{p}$,
-$\mathbf{A}_2$ is the change-of-basis matrix from $\mathbf{p}$ to $\mathbf{g}$, and
-$\mathbf{A}\equiv \mathbf{A}_1 \mathbf{A}_2$.
-
-To compute rotational light curves we also need to be able to specify the
-orientation of a surface map with coefficients $\mathbf y$. We can rotate the
-map using the Wigner rotation matrix $\mathbf{R}(\mathrm{I}, \Lambda, \Theta)$
-that rotates $y$ given the body's inclination $\mathrm{I}$, obliquity $\Lambda$
-and rotational phase $\Theta$. The rotated map is then $\mathbf{R}(\mathrm{I},
-    \Lambda, \Theta) \mathbf{y}$ and Equation~\ref{eq:intensity_poly_basis} can be
-rewritten more generally as
-\begin{equation}
-    I(x, y)=\tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_1 \mathbf{R}\mathbf{y}
-\end{equation}
-The total flux measured by an observer is is given by an integral of the specific
-intensity over a region S of the projected disc of the body:
-\begin{align}
-    F & =\oiint I(x, y) \ud S                                                            \\
-      & =\oiint \tilde{\mathbf{p}}^{\top}(x, y) \mathbf{A}_{1} \mathbf{R} \mathbf{y} d S \\
-      & =\mathbf{r}^{\top} \mathbf{A}_{1} \mathbf{R} \mathbf{y}
-\end{align}
-where $\mathbf{r}$ is a column vector whose $n$-th element is \citep{2019AJ....157...64L}
-\begin{equation}
-    r_{n} \equiv \oiint \tilde{p}_{n}(x, y) \ud S
-\end{equation}
-When the entire disc is visible (no occultor)
-\begin{equation}
-    r_{n}=\int_{-1}^{1} \int_{-\sqrt{1-x^{2}}}^{\sqrt{1+x^{2}}} \tilde{p}_{n}(x, y) \mathrm{d}y \mathrm{d} x
-\end{equation}
-and the solution to this integral can be expressed in terms of Gamma functions
-\citep[Equation 20 in ][]{2019AJ....157...64L}.
-$\mathbf{r}$ and $\mathbf{A}_1$ are independent of the map coefficients $\mathbf{y}$
-so they can be pre-computed.
-
-Computing the occultation light curves is more complicated because the the body
-is partially occulted by an occultor of radius $r$, centered at $(x_o, y_o)$
-and the exposed portion of the disc is a function of $(r, x_o, y_o)$. The
-general expression for the flux is
-\begin{align}
-    F & =\oiint I(x, y) \mathrm{d} S                         \\
-      & =\mathbf{s}^\top  \mathbf{A} \mathbf{R} \mathbf{y} .
-\end{align}
-where the vector
-\begin{equation}
-    \mathbf{s}^\top \equiv \oiint \tilde{\mathbf{g}}^{\top}(x, y) \mathrm{d} S
-\end{equation}
-is defined to be the solution to the integral.
-\citet{2019AJ....157...64L} solves this integral in two steps.  First, we rotate
-the coordinate system about the  $z$-axis  so that the occultor lies at along the
-$+y$ axis and its center is at
-a distance $b=\sqrt{x_o^{2}+y_o^{2}}$ from the origin. This substantially simplifies
-the limits of integration.
-The second step is to use Green's theorem to express the surface integral of
-$\tilde{\mathbf{g}}_n$ as a line integral of a vector function $\mathbf{G}_n$
-along the boundary of the visible portion of the occulted disc.
-The $n$-th element of the solution vector $\mathbf{s}^\top$ is
-\begin{equation}
-    s_{n}=\oiint \tilde{g}_{n}(x, y) \mathrm{d} S=\oint \mathbf{G}_{n}(x, y) \cdot \ud \mathbf{r}
-\end{equation}
-where
-$\mathbf{G}_{n}(x, y)=G_{n x}(x, y) \hat{\mathbf{x}}+G_{n y}(x, y) \hat{\mathbf{y}}$
-is constructed such that
-\begin{equation}
-    \mathbf{D} \wedge \mathbf{G}_{n}=\tilde{g}_{n}(x, y)
-    \label{eq:G_n}
-\end{equation}
-were $\mathbf{D} \wedge \mathbf{G}_{n}$ is the exterior derivative of $\mathbf{G}_n$
-given by
-\begin{equation}
-    \mathbf{D} \wedge \mathbf{G}_{n} \equiv \frac{\ud G_{n y}}{\ud x}-\frac{\ud G_{n_{x}}}{\ud y}
-\end{equation}
-in Cartesian coordinates.
-\citet{2019AJ....157...64L} provides one possible expression for $\mathbf{G}_n$
-which satisfies Equation~\ref{eq:G_n} and shows that the final solution can be written as
-\begin{equation}
-    s_{n}=\mathcal{Q}\left(\mathbf{G}_{n}\right)-\mathcal{P}\left(\mathbf{G}_{n}\right)
-\end{equation}
-where $\mathcal{P}(\mathbf{G}_n)$ is the  line integral along the arc of the occultor
-of radius $r$ and $\mathcal{Q}(\mathbf{G}_n)$ is the line integral along the arc of the
-occulted body of unit radius. Solutions to these integrals are long but they consist of
-analytic functions such as sines, cosines and complete elliptic integrals and importantly
-they only need to be evaluated once for an arbitrary map with fixed geometry of an
-occultation \citep{2019AJ....157...64L}.
-
-The final expression for the total flux for a particular geometrical
-arrangement of the occulted body and the occultor can then be written
-succinctly as
-\begin{equation}
-    f = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{y}
-    \label{eq:starry}
-\end{equation}
-Where $\mathbf{R}'$ is a rotation matrix which rotates the map such that the occultor
-is placed symmetrically on the $+y$ axis at distance $b$.
-Notice that Equation~\ref{eq:starry} defines a linear operation acting on a vector
-of spherical harmonic coefficients $\mathbf{y}$.
-It is also trivial to generalize this expression to a case of \emph{spectral map}. In
-that case we define a matrix $Y$ whose columns are the spherical harmonic coefficients
-at different wavelengths and the total flux (spectrum) is then
-\begin{equation}
-    \mathbf{f} = \mathbf{s}^T\mathbf{A}\mathbf{R}'\mathbf{R}\mathbf{Y}
-    \label{eq:starry_mw}
-\end{equation}
-where $\mathbf{f}$ is a vector of fluxes, one per wavelength bin.
-
-The formalism above can also be extended to account for limb-darkening which is
-important when modeling transits of planets across stars or the light curves of
-eclipsing binaries. \citet{2020AJ....159..123A} show how a limb-darkening
-profile that is an order $l$ polynomial function of
-$\mu=z=\sqrt{1-x^{2}-y^{2}}$ can be expressed in terms of the $m=0$ spherical
-harmonics up to order $l$. In the case of quadratic limb-darkening
-$(l_\mathrm{max}=2)$, the limb darkening profile is
-\begin{equation}
-    \frac{I(\mu)}{I(1)}=1-u_{1}(1-\mu)-u_{2}(1-\mu)^{2}
-\end{equation}
-which can be expressed as a sum of spherical harmonics
-\citep[Equation 38 in][]{2019AJ....157...64L}. Limb-darkening needs to be
-treated separately from the map coefficients $\mathbf{y}$ because it does not rotate along with the map
-when rotations $\mathbf{R}$ and $\mathbf{R}'$ are applied. The limb-darkening
-coefficients are applied to the map $\mathbf{y}$ as a multiplicative filter
-following any rotations which results in another set of spherical harmonic
-coefficients because products of spherical harmonics are also spherical
-harmonics. Applying limb-darkening raises the degree of the map by the degree
-of the limb-darkening.
-
-\subsubsection{Reflected light}
-The results derived in the previous section are valid only for light emitted
-from the occulted body. Equation~\ref{eq:starry} is a good model for stellar
-light curves and secondary eclipse light curves and phase curves of planets
-observed in the mid to far infrared wavelengths where the contribution of
-scattered starlight to total flux is negligible. Modeling reflected light phase
-curves and occultations is far more complicated because one has to take into
-account the the nonuniform illumination of the body and the possible presence
-of a terminator line (boundary between day and night).
-\citet{2021arXiv210306275L} extends the formalism described in the previous
-section to the case of reflected light phase curves and occultations for a body
-whose spatial \emph{albedo} (specifically the \emph{spherical albedo} -- the
-fraction of power incident on a body at a given wavelength that is scattered
-into space in all directions) distribution is expanded in a basis of spherical
-harmonics. Thus, analogous to Equation~\ref{eq:sh_expansion} we have
-\begin{equation}
-    A(x, y)=\tilde{\mathbf{y}}^{\top}(x, y) \mathbf{y}
-\end{equation}
-where $A$ is the albedo. We work under the assumption of isotropic (Lambertian)
-scattering of light at every point on the body which means that the illumination
-profile $\mathcal{I}$ of the surface is given by Lambert's law:
-\begin{equation}
-    \mathcal{I}\left(\vartheta_{\mathrm{i}}\right)=\mathcal{I}_{0} \max \left(0, \cos \vartheta_{\mathrm{i}}\right)
-\end{equation}
-where $\vartheta_i$ is the angle between the incident radiation at the surface normal
-and $\mathcal{I}_0$ is the maximum illumination. In case of, for example, a planet illuminated
-by its host star \citep[Appendix A.2 in][]{2021arXiv210306275L}
-\begin{equation}
-    \mathcal{I}_{0}=\frac{f_{s}}{\pi r_{\mathrm{s}}^{2}}
-\end{equation}
-where $r_s$ is the distance between the planet and the star in units of planet's radius
-and $f_s$ is the stellar flux measured at the observer in some arbitrary units.
-We assume that $f_s=1$ so that all fluxes are defined as the fraction of the flux
-of the illumination source at the observer.
-It is the presence of the day/night terminator that complicates the calculation
-of total flux the most. The limits of integration end up depending on a
-solution to a quartic equation specifying the points of intersection between
-the occultor and the day/night terminator line and the solution to those
-integrals are a function of \emph{incomplete} elliptic integrals as opposed to
-complete elliptic integrals as was the case for occultations in emitted light.
-The full derivation of the total flux is provided by \citet{2021arXiv210306275L} and
-I will only briefly summarize  the result here for completeness.
-
-Under the Lambertian scattering assumption the observed intensity at any point
-of the surface of the body is is proportional to the cosine of the angle
-$\vartheta_i$ between the incident light rays and the surface normal. All
-points for which $\vartheta_i\geq \pi/2$ have intensity of zero (they're
-unilluminated). Let's assume that the point source is placed at sky coordinates
-$\left(x_{\mathrm{s}}, y_{\mathrm{s}}, z_{\mathrm{s}}\right)$ in units of the
-radius of the illuminated body. The day/night terminator on the body is a
-half-ellipse with a semi-major axis equal to unity and a (signed) semi-minor
-axis equal to
-\begin{equation}
-    b=-\frac{z_{\mathrm{s}}}{r_{\mathrm{s}}}
-\end{equation}
-where $r_{\mathrm{s}}=\sqrt{x_{\mathrm{s}}^{2}+y_{\mathrm{s}}^{2}+z_{\mathrm{s}}^{2}}$
-is the distance to the source. The angle by which the semi-major axis of this ellipse is
-rotated away from the $+x$-axis is  is
-\begin{equation}
-    \theta=-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
-\end{equation}
-Under the assumption that $rs \gg 1$, the illumination $\mathcal{I}$ at point
-$(x,y)$ on the projected disc is \citep{2021arXiv210306275L}
-\begin{equation}
-    \mathcal{I}\left(b, \theta, r_{\mathrm{s}} ; x, y\right)=\max \left(0, I\left(b, \theta, r_{\mathrm{s}} ; x, y\right)\right)
-    \label{eq:illumination}
-\end{equation}
-where
-\begin{align}
-    I\left(b, \theta, r_{\mathrm{s}} ; x, y\right) & =\frac{1}{\pi r_{\mathrm{s}}^{2}} \cos \vartheta_{\mathrm{i}}                                                      \\
-                                                   & =\frac{1}{\pi r_{\mathrm{s}}^{2}}\left(-b_{\mathrm{c}} \sin \theta x+b_{\mathrm{c}} \cos \theta y-b z(x, y)\right)
-\end{align}
-with  $b_{\mathrm{c}} \equiv \sqrt{1-b^{2}}$  and $z(x, y)=\sqrt{1-x^{2}-y^{2}}$.
-The illumination $\mathcal{I}$ is a dimensionless quantity normalized such that the
-integral of $A\,\mathcal{I}$ over the units disc is equal to the flux measured by the
-observer as a fraction of the flux of the illumination source.
-To compute the total flux, we need to weigh each of the terms in the Green's basis
-and integrate them over the visible portion of the body's disc to obtain
-the reflected light solution vector $\mathbb{s}^\top$.
-Evaluating these integrals is unfortunately intractable because of the piecewise
-function in Equation~\ref{eq:illumination}. It is more tractable to weight the basis
-terms by the function $I$ and modify the limits of integration such that the nightside
-of the body is excluded. \citet{2021arXiv210306275L} shows that $I$ can be expressed
-in the form of a linear operator $\mathbf{I}$ to weight a map vector  in the polynomial
-basis $\tilde{\mathbf{p}}$ by the illumination profile. The total flux is then
-\begin{equation}
-    f=\mathbb{s}^\top\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right) \mathbf{A}_{\mathbf{2}} \mathbf{I}\left(b, \theta^{\prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime}\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
-\end{equation}
-where
-\begin{equation}
-    \theta^{\prime}=\arctan 2\left(x_{\mathrm{o}}, y_{\mathrm{o}}\right)-\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)
-    \label{eq:starry_ref}
-\end{equation}
-is the angle of the terminator in frame $\mathcal{F}^\prime$.
-In case the illuminated body is not occulted we have
-\begin{equation}
-    f_{0}=\mathbb{r}^{\top}(b) \mathbf{I}\left(b, \theta^{\prime \prime}, r_{\mathrm{s}}\right) \mathbf{A}_{\mathbf{1}} \mathbf{R}^{\prime \prime}\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right) \mathbf{R}(\mathrm{I}, \Lambda, \Theta) \mathbf{y}
-    \label{eq:starry_ref_phase}
-\end{equation}
-where $\theta^{\prime\prime}=0$ is the angle of the terminator in frame
-$\mathcal{F}^{\prime\prime}$ by construction. $\mathbf{R}^{\prime\prime}$ rotates the
-body through an angle $\arctan 2\left(x_{\mathrm{s}}, y_{\mathrm{s}}\right)$ so the
-semi-major axis of the terminator is aligned with the $x^{\prime \prime}$ axis.
-The solutions to integrals
-$\mathbb{r}^{\top}(b)$ and $\mathbb{s}^{\top}\left(b, \theta^{\prime}, b_{\mathrm{o}}, r_{\mathrm{o}}\right)$
-are anything but trivial, they are derived in the appendices of
-\citet{2021arXiv210306275L}. The key point is that the model is once again linear and
-the matrices in Equations~\ref{eq:starry_ref} and \ref{eq:starry_ref_phase} need to be
-computed only once for a specific occultation/illumination geometry.
-\citet{2021arXiv210306275L} also derives a solution for the total flux in case the
-light source is extended (the assumption that $r_s\gg 1$ is not satisfied) and
-in case the body is not a perfect Lambertian reflector.
-
-\subsection{The starry code}
-The method described in the previous section is implemented in the
-\textsf{Python} package
-\textsf{starry}\footnote{\url{https://starry.readthedocs.io/en/latest/}}.
-\textsf{starry} is many orders of magnitude faster and more accurate than
-similar codes which rely on numerical integration. It enables computation of
-rotational light curves, planetary transits, secondary eclipse light curves and
-planet-planet occultations in both emitted and reflected light. Recent
-extensions to \textsf{starry}, not mentioned in Section~\ref{sec:occultations},
-extended the spherical harmonic formalism to the problem of Doppler imaging
-\citep{2021arXiv211006271L} and add support for computing occultation light
-curves of oblate bodies \citep{2022ApJ...925..185D}. In
-Chapters~\ref{ch:mapping_io} and \ref{ch:mapping_exoplanets} we make extensive
-use of \textsf{starry} to explore solutions to the \emph{inverse problem} --
-how can we obtain an estimate of the map coefficients $\mathbf{y}$ which best
-explain the observed light curve? \textsf{starry} abstracts the complexities of
-computing the total flux and allows us to treat this problem as a linear
-problem of the form
-\begin{equation}
-    f = \mathbf{X}\mathbf{y}
-\end{equation}
-where  $\mathbf{X}$ is computed by \textsf{starry} depending on the geometry of an
-occultation or transit event. The inverse problem is highly degenerate
-and making progress requires  imposing some prior structure on the solution
-$\mathbf{y}$.
-
-In Chapter~\ref{ch:mapping_io} we use \textsf{starry} to model occultation
-light curves of a Solar System object -- Jupiter's moon Io, as it is occulted
-by Jupiter. Thanks to the comparatively close distance to Io relative to
-objects outside of the Solar System, these light curves are very high quality
-we are able to fit very high order maps $l=20$ and test our models in a high
-signal-to-noise regime. We also discuss how possible approaches to modeling
-time-dependent maps in which case the spherical harmonic coefficients are
-time-dependent: $\mathbf{y}=\mathbf{y}(t)$. In
-Chapter~\ref{ch:mapping_exoplanets} we tackle a very similar problem but in a
-very different regime of exoplanet eclipse mapping where the signal-to-noise is
-orders of magnitude worse than in the case of Io. In both cases we mostly focus
-on thermal light curves rather than light curves in reflected light.
-
-\section{Statistical inference -- theory}
-\label{sec:inference_theory}
-Having covered the basic physics behind gravitational lensing and occultation mapping I
-now turn to the subject of statistical inference in general and more specifically --
-Bayesian statistical inference. Bayesian (as opposed to frequentist)
-provides a straightforward and elegant theoretical framework for solving inverse
-problems of the kind we mentioned
-in Sections~\ref{sec:microlensing}  and \ref{sec:occultations}. These are problems
-where we have little data per object of interest, the data is very noisy,
-the physical models are relatively well understood but their parameters don't map
-to observables in a straightforward way and there are multiple competing, sometimes
-physically  quite different, models which provide a good explanation for the data.
-Having said that, I am not particularly dogmatic about the use of Bayesian methods
-and I will occasionally steer away from the classical approach and point out where
-a less Bayesian approach is often superior in practice.  What is most important is the
-use of \emph{probabilities} for encoding uncertainty about the physical world.
-
-I start by reviewing basic probability theory and briefly mention the
-frequentist interpretation. I then introduce two key concepts, the likelihood
-function -- a recipe for generating plausible datasets given model parameters,
-and priors -- what we know about the model parameters prior to gathering any
-data. Next, I briefly summarize fundamental concepts such as the curse of
-dimensionality and the no-free-lunch-theorem. Finally I end the section by
-comparing machine learning to statistical inference within the physical
-sciences.
-
-\subsection{Probability theory}
-There are two different interpretations of probability. In the
-\emph{frequentist} interpretation probabilities are \emph{frequencies of
-    events} which are repeated in a large number of trials (think of a repeated
-coin tosses). An alternative to the frequentist interpretation is the
-\emph{Bayesian} interpretation in which probabilities represent \emph{degrees
-    of belief} about the subject of interest. In both cases probabilities are
-necessarily positive numbers between zero and one. The major advantage of the
-Bayesian interpretation is that it allows one to reason about the probability
-of one-off events such as ``person X wins the election'' or ``this microlensing
-event is a binary lens event''. The differences between the two interpretations
-are not just philosophical because which interpretation one subscribes to can
-determine how one approaches a particular statistical inference problem and
-which sets of tools one uses. Both approaches are commonly used within the
-physical sciences depending on the field and the problem at hand. For instance,
-in case of particle physics frequentist methods are much more common because
-particle physics experiments often involve billions of repeated experiments in
-particle accelerators. One the other hand, Bayesian methods are more popular in
-astronomy because we cannot do true repeated experiments with astrophysical
-objects.
-
-The mathematical rules for manipulating probabilities are quite
-straightforward. In probability theory we use \emph{random variables} which can
-be either discrete or continuous and their possible values are described with
-probability distributions over all possible values the variable is allowed to
-take. For discrete random variables we can talk of probabilities that a given
-variable assumes a particular value, say $5$, but with continuous random
-variables we can only talk of probabilities that a given variable is in a
-particular range of values, say $[0,1]$. Now let's consider a continuous
-\emph{random variable} $x$ with an associated probability density function
-(pdf) $p(x)$\footnote{The notation $p(x)$ we use is somewhat overloaded. In
-    statistics, the proper notation for a probability density function is
-    $p_{X}(x)$ where $X$ is a random variable and $x$ is a particular realization
-    of that random variable.} Since $x$ has to take on \emph{some values} within
-its domain we have the requirement that
-\begin{equation}
-    \int p(x)\,\ud x=1
-    \label{eq:normalization_condition}
-\end{equation}
-where the integral is taken over the entire domain of $x$.
-In order for Equation~\ref{eq:normalization_condition} to hold,
-$p(x)$ has to have units of $x^{-1}$ so that
-the integral is a dimensionless number.
-
-In general, we are usually dealing with several parameters at once in which
-case we are interested in \emph{joint} probability distributions over all the
-parameters. Consider random variables $x_1$ and $x_2$ with a joint probability
-density function $p(x_1,x_2)$ which when integrated over a particular interval
-gives the probability that both $x_1$ and $x_2$ are contained within that
-interval. These joint probabilities can be decomposed into \emph{conditional
-    probabilities} using the so called product rule:
-\begin{align}
-    p(x_1,x_2) & =p(x_1)\,p(x_2\lvert x_1) \\
-    p(x_1,x_2) & =p(x_1\lvert x_2)\,p(x_2)
-    \label{eq:product_rule}
-\end{align}
-where $p(x_2\lvert x_1)$ is a pdf for $x_2$ \emph{conditional on
-    $x_1$ assuming a certain value.} $p(x_2\lvert x_1)$ is every bit as
-valid pdf as $p(x_2)$, it has the same units and it has to integrate to 1.
-Combining \ref{eq:product_rule} results in the famous \emph{Bayes's theorem}:
-\begin{equation}
-    p(x_1\lvert x_2)= \frac{p(x_2\lvert x_1)\,p(x_1)}{p(x_2)}
-    \label{eq:bayes_theorem}
-\end{equation}
-Bayes's theorem is simply a rule for converting one kind of a conditional probability
-into another.
-Given a pdf $p(x_1,x_2)$, we can \emph{integrate out} or
-\emph{marginalize} parameters that we are not interested in to obtain a distribution over
-parameters of interest:
-\begin{equation}
-    p(x_1)=\int p(x_1,x_2)\textrm{d}x_2=\int p(x_1\lvert
-    x_2)\,p(x_2)\textrm{d}x_2
-\end{equation}
-where we applied Bayes's theorem in the second inequality.
-In cases where the parameter space is high dimensional these integrals might be
-impossible to calculate analytically but they can (sometimes) be estimated using sampling
-algorithms as we shall se in subsequent sections.
-In some cases, we may be able to factor out $p(x_1,x_2)$ as
-If $x_1$ and $x_2$ are \emph{statistically independent} we can factor out $p(x_1,x_2)$ as
-\begin{equation}
-    p(x_1,x_2)=p(x_1)\,p(x_2)\,.
-\end{equation}
-Independence is a very common assumption in statistical modelling. For instance, when modeling light curves
-of microlensing events or planetary transits we often assume that flux $f_i$ measured at time $t_i$ is independent
-of flux $f_j$ measured at time $t_j$  for all measurement times $t_i,t_j$ so that
-\begin{equation}
-    p(\{f\}_{i=1}^N)=\prod_{i=1}^Np(f_i)
-    \label{eq:likelihood_indep}
-\end{equation}
-If the pdf-s $p(f_i)$ also happen to be equal, we say that the data is independent and
-identically distributed (``iid''). If the data is independent but not identically distributed
-we say that the uncertainties are \emph{heteroscedastic}.
-The iid assumption is not that common in astronomy because we tend to have some idea of per-data-point
-uncertainties.
-
-\subsubsection{Expectation values and the mode}
-The \emph{mean} or the \emph{expected value} of a distribution $p(\vec{x})$,
-usually denoted by $\mu$, is a measure of the ``center'' of a distribution and
-it is defined as
-\begin{equation}
-    \mathbb{E}[x] \equiv\int x \,p(x) \ud x
-\end{equation}
-Similarly, the \emph{variance} is a measure of the ``spread'' of a distribution:
-\begin{align}
-    \mathbb{V}[x] & \equiv\mathbb{E}\left[(x-\mu)^{2}\right]=\int(x-\mu)^{2} p(x) d x                                           \\
-                  & =\int x^{2} p(x) \ud x+\mu^{2} \int p(x) \ud x-2 \mu \int x p(x) \ud x=\mathbb{E}\left[x^{2}\right]-\mu^{2}
-\end{align}
-The variance is usually denoted by $\sigma^2$ and the square root of $\sigma^2$ is the
-\emph{standard deviation}.
-For a \emph{multivariate} random variable with $D$ dimensions then the variance generalizes to
-a symmetric, positive semi definite matrix called the \emph{covariance matrix}:
-\begin{align}
-    \operatorname{Cov}[\vec{x}] & \equiv\mathbb{E}\left[(\vec{x}-\mathbb{E}[\vec{x}])(\vec{x}-\mathbb{E}[\vec{x}])^{\top}\right] \equiv \boldsymbol{\Sigma}                                                                                                                                                                                \\
-                                & =\left(\begin{array}{cccc}
-                                             \mathbb{V}\left[x_{1}\right]                & \operatorname{Cov}\left[x_{1}, x_{2}\right] & \cdots & \operatorname{Cov}\left[x_{1}, x_{D}\right] \\
-                                             \operatorname{Cov}\left[x_{2}, x_{1}\right] & \mathbb{V}\left[x_{2}\right]                & \cdots & \operatorname{Cov}\left[x_{2}, x_{D}\right] \\
-                                             \vdots                                      & \vdots                                      & \ddots & \vdots                                      \\
-                                             \operatorname{Cov}\left[x_{D}, x_{1}\right] & \operatorname{Cov}\left[x_{D}, x_{2}\right] & \cdots & \mathbb{V}\left[x_{D}\right]
-                                         \end{array}\right)
-\end{align}
-
-In the general case, we can compute expectation values of an arbitrary function
-$g(\vec{x})$ under the probability distribution $p(\vec{x})$:
-\begin{equation}
-    \mathbb{E}\left[g(\vec{x})\right]=\int g(\vec{x})\,p(\vec{x})\ud x
-    \ud\theta
-\end{equation}
-Some examples of $g$ include higher moments of the distribution, the median and
-quantile estimates. The \emph{mode} of $p(\vec{x})$
-\begin{equation}
-    \vec{x}^{*}=\underset{\vec{x}}{\operatorname{argmax}}\,p(\vec{x})
-\end{equation}
-on the other hand is not an expectation value. It is the global maximum of the
-function $p(\vec{x})$.
-Estimating the mode involves computing the \emph{derivatives} of the density
-$p(\vec{x})$, whereas expectation values require computing
-\emph{integrals} over the space $\vec{x}$. Computing either
-the mode or expectation values
-for nontrivial densities $p(\vec{x})$ is a very hard and often impossible problem
-(more on this in Section~\ref{sec:inference_practice}).
-
-\subsubsection{Central Limit Theorem}
-Consider $N$ \emph{independent and identically distributed} random variables
-with pdf's $p_n(x)$ which are not necessarily Gaussian. Let $S_N$ be the sum
-the random variables. The \emph{Central Limit Theorem} (CLT) says that as $N$
-increases, the distributions of this sum approaches the Gaussian distribution
-\begin{equation}
-    p\left(S_{N}=u\right)=\frac{1}{\sqrt{2 \pi N \sigma^{2}}} \exp \left(-\frac{(u-N \mu)^{2}}{2 N \sigma^{2}}\right)
-\end{equation}
-and the distribution of the quantity
-\begin{equation}
-    Z_{N} \equiv \frac{S_{N}-N \mu}{\sigma \sqrt{N}}=\frac{\bar{x}-\mu}{\sigma / \sqrt{N}}
-\end{equation}
-where $\bar{x}$ is the \emph{sample mean} $\frac{1}{N}\sum_{i=1}^Nx_i=S_N/N$
-converges to the Gaussian distribution with zero mean and unit variance.
-Because lots of random variables representing real world processes can be seen as sums
-of i.i.d. independent random variables the Gaussian distribution is all over the place
-in statistics. One should keep in mind that the assumptions underlying the CLT are
-not always satisfied and also the convergence towards a Gaussian is
-more rapid in the bulk of the distribution around the mean than in the tails of the
-distribution.
-
-\subsubsection{Change of variables}
-Given a univariate random variable $x$ the effect of some deterministic
-transformation $y(x)$ is to stretch and shift the distribution $p(x)$. For an
-infinitesimal interval the \emph{probability mass} is the same in both
-variables so $p(x) \ud x=p(y) \ud y$ which yields the \emph{change of variables
-    formula}
-\begin{equation}
-    p_{y}(y)=p_{x}(x)\left|\frac{\ud x}{\ud y}\right|
-\end{equation}
-In the multivariate case, if $\vec{f}$ is an invertible function mapping $\mathbb{R}^n$
-to $\mathbb{R}^n$ with inverse $\vec{g}$, the pdf of $\vec{y}=\vec{f}(\vec{x})$ is
-\citep{murphy_book_2022}
-\begin{equation}
-    p_{y}(\vec{y})=p_{x}(\vec{g}(\vec{y}))\left|\operatorname{det}\mathbf{J}_{g}(\vec{y})\right|
-\end{equation}
-where $\vec{J}_{g}=\frac{\ud \vec{g}(\vec{y})}{\ud \vec{y}^{\top}}$ is the Jacobian of $\vec{g}$
-and $\operatorname{det}|\mathbf{J}(\vec{y})|$ is the determinant of the Jacobian.
-
-\subsubsection{Kullback Leibler divergence}
-A common way of measuring ``similarity'' between two distributions $p(x)$ and
-$q(x)$ is using the \emph{Kullback-Leibler divergence} (KL divergence for
-short):
-\begin{equation}
-    D_{\mathrm{KL}}(p \| q)=\int_{-\infty}^{\infty} p(x) \ln \left(\frac{p(x)}{q(x)}\right) \ud x
-    \label{eq:kl_divergence}
-\end{equation}
-where $D_{\mathrm{KL}}(p \| q)$ should be read as ``the KL divergence from p to q''.
-This is not a measure of ``distance'' between the two distributions because
-it is not a symmetric quantity. The KL divergence is best understood as a quantity which
-quantifies \emph{how surprised one would be if one expected to encounter distribution $p(x)$
-    but instead saw distribution
-    $q(x)$}\footnote{\url{https://wiki.santafe.edu/images/a/a8/IT-for-Intelligent-People-DeDeo.pdf}}.
-This quantity is important in the context of variational inference which is covered in
-Section~\ref{sec:inference_practice}.
-
-\subsection{Bayes' theorem in practice}
-\label{ssec:likelihood_function}
-Consider a general inference problem where we collect some data $\mathcal{D}$ and we are interested
-in the joint probability distribution for parameters $\boldsymbol\theta$ of particular model.
-We can then rewrite Bayes' theorem as
-\begin{equation}
-    p(\vec{\theta} | \mathcal{D})=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{p(\mathcal{D})}=\frac{p(\vec{\theta}) p(\mathcal{D} | \vec{\theta})}{\int p\left(\vec{\theta}^{\prime}\right) p\left(\mathcal{D} |\vec{\theta}^{\prime}\right) d \vec{\theta}^{\prime}}
-    \label{eq:bayes_theorem2}
-\end{equation}
-In Equation~\ref{eq:bayes_theorem2} the term $p(\mathcal{D}|\vec{\theta})$  is a probability
-distribution over the data $\mathcal{D}$ given the parameters of the model $\vec{\theta}$.
-When this term is seen as a function of the parameters given a fixed dataset, it is
-called the \emph{likelihood function }  or the \emph{likelihood} for short. $p(\boldsymbol\theta)$
-is to so called \emph{prior} distribution over $\vec{\theta}$ and it represents our beliefs
-about $\vec{\theta}$ prior to having observed the data $\mathcal{D}$. When the prior distribution is
-multiplied by the likelihood we obtain, up to a normalizing constant $p(\mathcal{D})$, the posterior
-distribution over $\vec{\theta}$ given the data $\mathcal{D}$.
-The best way to think about Equation~\ref{eq:bayes_theorem2} is seeing at as a rule
-for updating beliefs. Our prior beliefs about the parameters of the model
-$p(\mathrm{parameters})$ should be updated to posterior beliefs $
-    p(\mathrm{parameters} | \mathrm{data})$ by making use of the likelihood
-$p(\mathrm{data} | \mathrm{parameters})$. If the data is useful, the likelihood is generally
-much narrower than the prior pdf and the choice of priors does not significantly
-influence the final results.
-
-\subsubsection{The likelihood}
-Let's cover each of these three terms in more detail, starting with the most
-important one -- the likelihood. The likelihood can be seen as a \emph{recipe
-    for generating plausible datasets $\mathcal{D}$} (a \emph{generative model})
-given a specific realization of the parameters $\vec{\theta}$. For the kinds of
-models we work with in astronomy, the likelihood consists of two parts. The
-first part is (usually deterministic) function which describes the physical
-system of interest. This is for instance the predicted flux for a binary
-microlensing event with a particular trajectory. The second part is the
-\emph{noise model} which describes the statistical properties of the noise
-inherent in any physical measurement. We often assume that the noise is
-independent and Gaussian. It is very common to use the likelihood as is and
-optimize it with respect to the model parameters $\vec{\theta}$ in a procedure
-called \emph{maximum likelihood estimation} (MLE) to find a single point in the
-parameter space $\vec{\theta}^\star$ which maximizes the likelihood (more on
-this in Section~\ref{ssec:mle_estimation}).
-
-We can also marginalize the likelihood over a subset of parameters. Say that we
-have a likelihood function $p(\vec{\theta} ,\vec{\alpha}|\mathcal{D})$ where
-$\vec{\alpha}$ is a set of \emph{nuisance parameters} which we are not
-particularly interested in. To marginalize over these nuisance parameters we
-need to introduce a prior\footnote{Not introducing a prior leads to a
-    dimensionally incorrect result.} $p(\vec{\theta}|\vec{\theta})$ which may
-potentially also depend on the $\vec{\theta}$ and the marginalized likelihood
-is then
-\begin{equation}
-    p(\mathcal{D} | \vec{\theta})=\int p(\mathcal{D} | \vec{\theta}, \vec{\alpha})\,p(\vec{\alpha} |\vec{\theta}) \ud \vec{\alpha}\quad .
-\end{equation}
-The problem structure where where $\vec{\alpha}$ is some set of parameters
-we don't particularly care about such and $\vec{\theta}$ are the physical
-parameters of interest is very common in astronomy. For instance, in single lens gravitational microlensing estimating
-the distribution $p(t_E)$ over the parameter of interest $t_E$ requires marginalizing over
-parameters such as $F_S, F_B, t_0$ and $u_0$.
-
-Not all statistical models can be written in terms of the likelihood function.
-Most machine learning models for example do not have a simple probabilistic
-interpretation\footnote{Indeed, finding probabilistic interpretations of
-    complex machine learning models such as deep neural networks is an active
-    research area.}. If we can write down the likelihood function for a certain
-problem, we say that we have a \emph{generative model for the data} because the
-likelihood can be used to generate mock data sets.
-
-\subsubsection{Priors}
-The second term of interest is the prior $p(\vec{\theta})$. In principle, the
-prior pdf-s should capture all of the prior information we might have about the
-model parameters prior to gathering new data; in practice, we don't want to
-bias our results so we often try to use ``weakly informative'' priors -- priors
-that are informative enough that they exclude unreasonable parameter values but
-aren't so narrow that they rule out sensible values which may be favored by the
-likelihood. For example, we might have prior information such as ``the mass of
-the planet should be positive'' in which case a suitable weakly informative
-prior on the mass parameter should first of all exclude negative values but
-also potentially exclude values which are nonsensically large for the problem
-at hand. There are also the so called ``non-informative prior'' such as
-Jeffrey's prior and maximum entropy priors. Although these priors can be of use
-in some cases, in general \textbf{there no such thing as a truly non
-    informative prior}. Although a prior for a particular parameter $\theta$ might
-be uninformative in one set of coordinates, it is often informative for a
-certain transformation of the parameter $g(\theta)$. For example, a uniform
-prior in $\theta$ is not uniform in $\ln\theta$. Having said that, it is often
-worth the effort to construct priors which respect certain symmetries of the
-problem the we know to be true.
-
-Priors should not be seen in isolation, rather, they should be seen in the
-context of the likelihood \citep{arXiv:1708.07487}. To see that this is the
-case consider drawing samples from the prior $\vec{\theta}\sim
-    p(\vec{\theta})$, plugging those values into the likelihood distribution
-$p(\mathcal{D}|\vec{\theta})$ and then generating mock datasets, a process
-called \emph{prior predictive checking}. This process makes it easy to spot
-priors which in combination with the likelihood yield predictions of what the
-data should like like which do not match our expectations. An excellent highly
-in depth study on choosing priors is available
-\href{https://betanalpha.github.io/assets/case_studies/prior_modeling.html}{here}.
-
-Depending on how informative the data is we differentiate between two different
-regimes. We say that the inference process is \emph{data driven} if the data
-constrains the model parameters so well that the choice of the priors is
-effectively irrelevant. The other regime is said to be \emph{prior driven},
-when the data is so weakly informative that the shape of the posterior pdf-s is
-similar to the prior pdf. In general, in any statistical analysis involving
-priors, one should check the sensitivity of the scientific conclusion to
-assumptions of different priors. No statistical inference procedure is free of
-assumptions but the use of priors and likelihoods makes those assumptions more
-explicit.
-
-Priors are sometimes the goal of the inference. For example, if the prior on
-the prior pdf $p(\vec{\theta})$ itself depends on some other set of parameters
-$\vec{\beta}$, then we can marginalize out all of the $\vec{\theta}$ parameters
-and we are left with a likelihood for the parameters $\vec{\beta}$ of the
-prior.
-\begin{equation}
-    p(\mathcal{D}\lvert\vec{\beta})=\int p(\mathcal{D}\lvert\vec{\theta})\,p(\vec{\theta}\lvert
-    \vec{\beta})\ud \vec{\theta}
-\end{equation}
-This can then be used, using Bayes's theorem, to calculate the posterior
-over the prior parameters $\bm\beta$.
-\begin{equation}
-    p(\bm\beta\lvert\mathcal{D})\propto p(\mathcal{D}\lvert\vec{\beta})
-    \,p(\bm\beta)
-\end{equation}
-The parameters $\bm\beta$ characterizing the prior are usually called
-\emph{hyperparameters} or \emph{population level parameters}. Similarly, the priors
-for those parameters are called \emph{hyperpriors}.
-The whole process is called \emph{hierarchical Bayesian inference}
-(in astronomy literature),  or \emph{multilevel modelling, hierarchical modelling,
-    nested modelling, mixed models or random effects models}
-(in statistics and adjacent fields).
-Hierarchical modelling is an very useful method for estimating
-population-wide properties given data about individual objects. For example,
-in microlensing observations of exoplanets the individual parameters could be the
-masses of individual planets while the hyperparameters would describe the
-population the shape of the planetary mass function. We shall return to this topic
-in Chapter~\ref{ch:microlensing}.
-
-\subsubsection{The evidence}
-The final component of Equation~\ref{eq:bayes_theorem2} is the normalization
-constant $ p(\mathcal{D})$ in the denominator:
-\begin{equation}
-    \mathcal{Z}\equiv p(\mathcal{D})=\int p(\mathcal D\lvert\boldsymbol\theta)\,p(\boldsymbol\theta)\textrm{d}\boldsymbol\theta
-    \label{eq:evidence}
-\end{equation}
-is often called the \emph{fully marginalized likelihood} because
-it is a likelihood marginalized over all of the parameters of the model.
-Sometimes
-it is also called the \emph{Bayesian evidence}.
-$\mathcal{Z}$ is general extremely difficult to calculate because it is an
-integral over a potentially very high dimensional parameter space $\vec{\theta}$.
-Fortunately, in most cases, we can obtain the posterior distribution without having
-to calculate $\mathcal{Z}$.
-
-%% mention the likelihood principle
-\section{Statistical inference}
-\label{sec:inference_practice}
-In the previous section I briefly summarized some key aspects of probability theory
-and Bayesian statistics but I haven't discussed how to apply these ideas in practice. That
-is the subject of statistical inference. I start by introducing some very fundamental
-concepts such as the \emph{curse of dimensionality} and the \emph{no free lunch theorems}.
-I then cover linear models which are the foundation of all statistics and also relevant
-for the occultation mapping work since the problem is linear.
-Next, I discuss various computational inference methods such
-as optimization methods, Markov chain Monte Carlo and Variational Inference.
-Finally, I discuss model comparison and evaluation.
-
-\subsection{The curse of dimensionality and the no-free-lunch theorem}
-The \emph{curse of dimensionality} refers to a phenomenon where the
-\emph{volume} of \emph{possible} explanations for the data grows exponentially
-with increasing dimension of the problem. Consider the following
-example\footnote{ Taken from the excellent paper by \citet{arXiv:2104.00008}.
-}. There are images consisting of $n$ pixels where each of the pixels is either
-a black (0) or white (1). The number of possible images is $2^n$ which grows
-exponentially. If each image out of the $2^n$ possible images is described by
-one of two labels $A$, or $B$ then the total number of possible ways of
-labeling the set of images is $2^{2^n}$ which is a stupidly large number. If
-$n=9$, the number of ways to classify the set of all 9-bit black and white
-images is greater than the number of atoms in the observable universe. If the
-labels don't correlate with the properties of the images then one strategy for
-solving this problem would be the lookup table approach -- memorize each image
-with its label. The famous \emph{no-free-lunch theorem} (NFLT) proved by the
-mathematician David Wolpert \citep{wolpert1996} demonstrates that it is
-impossible to construct an algorithm which does better than the lookup table
-approach.
-
-Of course the assumption that the labels aren't correlated to features in the
-images is silly for any meaningful problem. For instance, all images of cats
-have certain things in common as do images of galaxies. The no-free-lunch
-theorem doesn't say that it is impossible to construct an algorithm to classify
-images of cats, it only says that there is no best algorithm \emph{if we
-    average over all possible inputs to the problem}. To construct an algorithm
-which works in practice we need to impose structure on the problem, what
-machine learning researchers call \emph{inductive bias}. A set of assumptions
-which work in one domain may not work well in another domain and no single
-model will work well in all domains.
-
-\subsection{Maximum likelihood estimation}
-One of the most popular inference methods is called \emph{maximum likelihood
-    estimation} (MLE). Maximum likelihood estimation answers the question:
-\textbf{what choice of parameters $\vec{x}$ makes my data most likely} as
-opposed to the question \textbf{what are the most likely values of the
-    parameters $\vec{x}$ having observed the data $\mathcal{D}$}\footnote{The
-    distinction between the two questions is subtle, but important. The parameters
-    which maximize the likelihood might be very different from those which are most
-    ``sensible'' and which best describe the data.} which necessitates the
-multiplication of the likelihood with a prior. Since we don't incorporate the
-prior, it is not a Bayesian method. The output of MLE is a \emph{point
-    estimate} of the parameters $\hat{\vec{\theta}}$. In a Bayesian viewpoint, the
-MLE is equivalent to MAP estimation under a uniform prior $p(\vec{\theta})$.
-The MLE estimate is obtained by maximizing the logarithm\footnote{Because
-    probabilities in most region of the parameter space are very small numbers we
-    always use the log transform when optimizing likelihoods or posteriors.} of the
-likelihood $p(\mathcal{D}\lvert\vec{\theta})$ (or equivalently, minimizing the
-negative log likelihood) with respect to the parameters $\vec{\theta}$:
-\begin{equation}
-    \hat{\boldsymbol{\theta}}_{\mathrm{mle}}=\underset{\boldsymbol{\theta}}{\operatorname{argmax}}
-    \ln p\left(\mathcal{D}\lvert \boldsymbol{\theta}\right)
-\end{equation}
-
-\subsubsection{MLE for a multivariate Gaussian}
-The log-likelihood of multivariate Gaussian distribution without the
-unnecessary constants is
-\begin{equation}
-    L L(\boldsymbol{\mu}, \boldsymbol{\Sigma})=\log p(\mathcal{D} \lvert \boldsymbol{\mu}, \boldsymbol{\Sigma})=\frac{N}{2} \log |\vec{\Sigma}^{-1}|-\frac{1}{2} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)^{\top} \vec{\Sigma}^{-1}\left(\boldsymbol{y}_{n}-\boldsymbol{\mu}\right)
-\end{equation}
-where $\vec{y}_n$ are the data points and $\vec{\Sigma}^{-1}$ is
-the inverse covariance matrix, also knowns as the \emph{precision matrix}.
-The MLE for the mean parameter $\hat{\vec{\mu}}$
-is given by the sample mean (empirical mean) of the data \citep{murphy_book_2022}:
-\begin{equation}
-    \hat{\boldsymbol{\mu}}=\frac{1}{N} \sum_{n=1}^{N} \boldsymbol{y}_{n}=\overline{\boldsymbol{y}}\quad .
-\end{equation}
-Similarly, the MLE of the the covariance matrix $\hat{\vec{\Sigma}}$ is equal to
-the empirical covariance matrix
-\begin{equation}
-    \hat{\boldsymbol{\Sigma}}=\frac{1}{N} \sum_{n=1}^{N}\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)\left(\boldsymbol{y}_{n}-\overline{\boldsymbol{y}}\right)^{\top}
-\end{equation}
-Thus, whenever we compute an empirical mean and the covariance for a given datasets we
-are implicitly assuming that the data is normally distributed and obtaining an MLE
-estimate of that distribution.
-
-\subsection{Least squares linear regression}
-In this section we cover the classical non Bayesian formulation of linear
-regression, also known as \emph{least squares regression} and in the next
-section we generalize to the Bayesian case. In a (multiple) linear regression
-model we have (noisy) measurements $\vec{y}\in \mathbb{R}^n$ (\emph{dependent
-    variable}), an $n\times p$ matrix $\vec{M}\in \mathbb{R}^{n\times p}$ of inputs
-(\emph{independent variables}, \emph{covariates}, or \emph{features}) and a
-vector of \emph{regression coefficients} or \emph{weights} $\vec{\theta}\in
-    \mathbb{R}^p$. This is a very general model because the matrix $\vec{M}$ can
-consist of basis functions such as polynomials, spherical harmonics, Fourier
-modes, wavelets etc. so although the model is linear in the parameters
-(regression coefficients), it is not generally linear in the raw input data. If
-we assume that the observations $\vec{y}$ have Gaussian noise we can write down
-the likelihood of this model as
-\begin{equation}
-    p(\vec{y}\lvert \vec{\theta})=\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta}, \vec{C})
-\end{equation}
-where $\vec{C}$ is the data covariance matrix and $\mathcal{N}$ is shorthand notation for
-the multivariate normal distribution. In astronomical application it is usually
-assumed to be diagonal and we might have some idea for the scale of the variance
-of the diagonal elements (the ``errorbars'').
-The negative log likelihood (NLL) is then equal (ignoring constants)  to
-\begin{equation}
-    \operatorname{NLL}(\vec{\theta})=\frac{1}{2}(\vec{M} \vec{\theta}-\vec{y})^{\top}\vec{C}^{-1}(\vec{M} \vec{\theta}-\vec{y})
-\end{equation}
-which is just the well known least squares objective or (half the) $\chi^2$ as it is known
-in astronomy.
-
-\subsubsection{Underdetermined case ($p < n$)}
-For now, let's assume that the inverse data covariance matrix $\vec{C}^{-1}$ is
-a diagonal matrix with the the same variance $\sigma^2$ for every data point.
-In that case the MLE solution for the weights $\vec{\theta}$ is the solution to
-the following (convex) optimization problem:
-\begin{equation}
-    \hat{\vec{\theta}}=
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2}
-    \label{eq:ols_objective}
-\end{equation}
-There is a closed form solution for Equation~\ref{eq:ols_objective} as long as the number
-of parameters and features $p$ is less than the number of data points $n$ and the solution
-is \citep{arXiv:2101.07256}
-\begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M})^{-1} \vec{M}^{\top} \vec{y}
-    \label{eq:ols_solution}
-\end{equation}
-under the assumption that $\vec{M}\vec{M}^\top$ is invertible which is usually the case.
-Equation~\ref{eq:ols_solution} is called the \emph{ordinary least squares} (OLS) solution.
-For a nontrivial data covariance matrix $\vec{C}$ Equation~\ref{eq:ols_solution} generalizes
-to
-\begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top}\vec{C}^{-1}\vec{M})^{-1} \vec{M}^{\top}\vec{C}^{-1} \vec{y}
-    \label{eq:wls_solution}
-\end{equation}
-which is the solution to the \emph{weighted least squares} (WLS) problem.
-It is common to modify the least squares objective functions with the addition of a
-\emph{regularization} term to prevent overfitting. For instance, we can modify the OLS
-objective (Equation~\ref{eq:ols_objective}) by adding an \emph{L2 regularization} term:
-\begin{equation}
-    \hat{\vec{\theta}}=
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2
-    \label{eq:ols_objective_l2}
-\end{equation}
-where $\lambda >0$ is a regularization parameter
-which penalizes large magnitudes of the coefficients $\vec{\theta}$. This is also a convex
-optimization problem and the solution is
-\begin{equation}
-    \hat{\vec{\theta}}=(\vec{M}^{\top} \vec{M} + \lambda\vec{I})^{-1} \vec{M}^{\top} \vec{y}
-    \label{eq:ols_solution_l2}
-\end{equation}
-where $\vec{I}$ is the identity matrix. Regularized OLS regression with this particular
-regularization term is also known as \emph{ridge regression} in the statistics
-literature\footnote{One of the many things that make statistics confusing is that very
-    similar methods often have special names which obscures the big picture. The vast majority
-    of statistical methods can be seen as a linear model with the data represented in a particular
-    basis, and with a particular choice of the likelihood (noise model), priors (regularization)
-    and  a method for obtaining the solution (usually a convex optimization problem).}
-L2 regularization (as well as most other forms of regularization terms) can be seen
-as a particular choice of priors for the regression coefficients in the Bayesian paradigm.
-The above also naturally generalizes to the case of weighted least squares.
-
-\subsubsection{Overdetermined case ($p > n$)}
-Contrary to popular belief within the astronomy community it is possible and
-often desirable to have a model with more parameters than data points (see
-Chapter~\ref{ch:mapping_io} for a real world example). In the overparametrized
-case of linear regression $(p > n)$ there are multiple choices for
-$\vec{\theta}$ which perfectly fit the data. The solution in that case is
-defined to be a minimum-norm parameter vector $\vec{\theta}$ that interpolates
-the data ($\vec{y}=\vec{M} \vec{\theta}$), out of possibly many degenerate
-solutions. The OLS objective in Equation~\ref{eq:ols_objective} can be modified
-such that it is valid for both the $p<n$ and the $p>n$ case by making use of a
-limit in which an L2 regularization term tends to zero
-\citep{arXiv:2101.07256}:
-\begin{equation}
-    \hat{\vec{\theta}}=\lim _{\lambda \rightarrow 0^{+}}\left[
-    \underset{\boldsymbol{\theta}}{\operatorname{argmin}}\|\vec{y}-\vec{M} \vec{\theta}\|_{2}^{2} + \lambda \|\vec{\theta}\|_2^2\right]\quad .
-\end{equation}
-In this case the OLS solution becomes
-\begin{equation}
-    \hat{\vec{\theta}}=\vec{M}^{\top} (\vec{M} \vec{M}^{\top})^{-1} \vec{y}
-    \label{eq:ols_solution_overdet}
-\end{equation}
-Equations~\ref{eq:ols_solution} and \ref{eq:ols_solution_overdet} can be unified into a
-single equation by making use of the \emph{Moore-Penrose pseudoinverse} $\vec{M}^\dagger$.
-The pseudoinverse is defined by taking the singular-value decomposition (SVD)
-of $\vec{M}$:
-\begin{equation}
-    \vec{M}^\dagger\equiv\vec{U} \vec{S}^\dagger \vec{V}^\dagger
-\end{equation}
-Then $\vec{M}^\dagger\equiv  \vec{V}^{\top} \vec{S}^{\dagger} \vec{U}^{\top}$.
-
-\subsection{Bayesian linear regression}
-In the Bayesian paradigm we can write down the posterior over the regression
-weights $\vec{\theta}$ assuming a Gaussian likelihood and a Gaussian prior over
-the weights. The posterior is then proportional to
-\begin{align}
-    p(\vec{\theta}|\vec{y}) & \propto p(\vec{y}\lvert \vec{\theta})\,p(\vec{\theta})                                                          \\
-                            & =\mathcal{N}(\vec{y}\lvert\vec{M}\vec{\theta},\vec{C})\,\mathcal{N}(\vec{\theta}\lvert \vec{\mu},\vec{\Lambda})
-    \label{eq:bayes_lin_regression}
-\end{align}
-where $\vec{C}$ is the data covariance matrix as before, $\vec{\mu}$ is the mean
-of the Gaussian prior on coefficients and $\vec{\Lambda}$ is the the prior covariance matrix.
-The product of two multivariate Gaussian distributions is also a multivariate Gaussian distribution
-and the solution to Equation~\ref{eq:bayes_lin_regression} is \citep{arXiv:2005.14199}
-\begin{equation}
-    \mathcal{N}(\vec{y} \lvert\vec{M} \vec{\theta}, \vec{C}) \,\mathcal{N}(\vec{\theta} \lvert\vec{\mu}, \vec{\Lambda})=\mathcal{N}(\vec{\theta} \lvert\vec{a}, \vec{A}) \,\mathcal{N}(\vec{y} \lvert\vec{b}, \vec{B})
-\end{equation}
-where
-\begin{align}
-     & \vec{A}^{-1}   =\vec{\Lambda}^{-1}+\vec{M}^{\top} \vec{C}^{-1} \vec{M}                                                         \\
-     & \vec{a}         =\vec{A} \left(\boldsymbol{\Lambda}^{-1} \boldsymbol{\mu}+\vec{M}^{\top} \mathbf{C}^{-1} \boldsymbol{y}\right) \\
-     & \mathbf{B}      =\mathbf{C}+\vec{M} \boldsymbol{\Lambda} \vec{M}^{\top}                                                        \\
-     & \boldsymbol{b}  =\vec{M} \boldsymbol{\mu}
-\end{align}
-
-%\subsection{Machine learning and the difference between inference and prediction}
-%% mention that even if we have a model focused on inference, it might be best if we judge 
-%% its quality through its predictive performance
-%
-\subsection{Sampling and optimization methods}
-\subsubsection{The Laplace approximation}
-\subsubsection{Variational inference}
-\subsubsection{Importance sampling}
-\subsubsection{Rejection sampling}
-\subsubsection{Markov Chain Monte Carlo}
-\subsubsection{Hamiltonian Monte Carlo}
-%\subsubsection{Other }
-%
-%\subsection{Model validation and comparison}
-%\label{ssec:model_comparisonGaussian linear }
-%
-``Furthermore, in the future, a blind search for planetary signals will not
-turn up all potential competing models unless those models
-are specifically considered.''\citep{2017AJ....153..129J}.
-\citet{2021AJ....162...59R}
-\citet{2020AJ....160...17H}
-
-%\subsubsection{Cross validation}
-%% discuss vanilla k-fold CV, stress out just how important and central CV is to science
-%\subsubsubsection{Pareto smoothed importance sampling LOO}
-%\subsubsection{Model comparison}
-%% don't do it!
-%
-%\section{Putting it all together -- \emph{probabilistic programming}}
-%\label{sec:programming}
-%\subsection{The success of machine learning}
-%% that quote from Yann LeCun on differentiable code 
-%\subsection{Automatic differentiation}
-%\subsubsection{Forward mode autodiff}
-%\subsubsection{Reverse mode autodiff or backpropagation}
-%\subsubsection{Jacobian vector products}
-%\subsection{JAX}
-%\subsection{Numpyro}
-
-% CHAPTER 3: Modeling microlensing events 
+
+Thus, to compute the magnification of an extended uniform source at an
+arbitrary position $w_0$ via Equation~\ref{eq:mag_green_theorem_unif} we need
+to obtain the complex contour points $z_k^{(n)}$ where $n 1\dots
+    n_\mathrm{limb}$ indexes the points along the $k$-th contour. We start by
+evaluating Equation~\ref{eq:images_single_lens} at successive points
+\begin{equation}
+    w^{(n)}=w_0 + \rho_\star e^{i\phi_n}
+\end{equation}
+on the limb of the source star where $\phi_n$  is the discretization of the
+angle $\phi\in[-\pi,\pi)$. A closed contour specified by complex points $w^{(n)}$ in the
+source plane $w$ maps to two closed contours in the image plane $z^{(n)}_k$ ($k=1,2$).
+Uniform sampling in $\phi$ such that $\phi_n=-\pi+n\Delta\phi$ is the simplest strategy
+for obtaining these contours. The outcome of this strategy is shown in the left panel
+of Figure~\ref{fig:limb_sampling1}. The small inset plot shows the source plane $w$
+and the circular disc of the source (black circle) on top of the point source
+magnification map. The source in this case is very close to the infinitesimal caustic
+point.  The density plot (black arcs) shows the two extended images. The two complex
+contours $z^{(n)}_k$ (orange and blue lines) are displayed using a connected dots.
+The directionality of the contours is encoded using the marker size
+such that the first point in each contour $z^{(0)}_k$ has the largest marker size
+and the marker size decreases linearly with index $n$. Thus, the orange contour
+proceeeds in a clockwise direction and the blue contour proceeds in a counter-clockwise
+direction. In addition to the location of each contour point we also store the
+parity of each point which is given by the sign of the determinant $\mathrm{det}\,\vec J$
+(Equation~\ref{eq:lens_eq_det_jac}). All points within a contour share the same sign of
+the determinant so we define the contour parity $p_k$ to be the sign corresponding
+to the first point in each contour $z^{(0)}_k$. In the case shown in
+Figure~\ref{fig:limb_sampling1} the orange contour has clockwise directionality
+meaning that it should acquire a negative sign in
+Equation~\ref{eq:mag_green_theorem_unif_expanded} but it also has negative parity
+$p_k$ so the the total magnification is proportional to the sum of the two areas
+which is what we expect in this case.
+
+\subsubsection{Adaptive sampling along the source limb}
+
+It is is clear that that a uniform distribution of points on the limb of the
+source in the source plane does not map to a uniform distribution of points
+along the contours in the image plane. There parts of the contours where the
+points are not densly distributed correspond to points on the limb of the
+source which are close to the caustic. A uniform sampling strategy leads to
+prohibitively innaccurate integration when the source is close to the caustic.
+This is especially problematic in case of binary and triple lenses because the
+limb of the source star can intersect multiple caustic curves. To solve this
+problem I have devised an adaptive sampling scheme which increases the density
+of points along the limb of the source star in regions where the gradient of
+the point source magnification is large. The algorithm is as follows:
+\begin{enumerate}
+    \item Start by uniformply sampling $N_\mathrm{limb}$ points on the limb of the source
+          and set $N=N_\mathrm{limb}$. Evaluate the point source magnification
+          $\mu_k^{(n)}$ at each point and estimate its derivative
+          $\mu^\prime(\phi^{(n)})$ along the limb using the central finite difference
+          approximation\footnote{
+              \begin{equation}
+                  \mu^\prime(\phi^{(n)}) \approx \frac{
+                      (\Delta\phi^{(n-1)})^2\mu(\phi^{(n)} + \Delta \phi^{(n)}) +
+                      \left[(\Delta\phi^{(n)})^2-(\Delta\phi^{(n-1)})^2\right]
+                      \mu(\phi^{(n)})-(\Delta\phi^{(n)})^2\mu(\phi^{(n)} - \Delta\phi^{(n-1)})
+                  }{\Delta\phi^{(n)}\Delta \phi^{(n-1)}(\Delta\phi^{(n)} + \Delta\phi^{(n-1)})}
+              \end{equation}
+              where $\Delta\phi^{(n)}=\phi^{(n+1)}-\phi^{(n)}$ and $\Delta\phi^{(n-1)}=\phi^{(n)}
+                  -\phi^{(n-1)}$.
+          } \citep{10.2307/2008770}
+    \item  Sort points $\phi^{(n)}$ according to the magnitude of the gradient
+          $\mu^\prime(\phi^{(n)})$ in decreasing order.
+    \item Set $N=\mathrm{int}(C_\mathrm{geom}\times N)$ where $\mathrm{int}()$ denotes
+          casting to integer and select $N$ points with the largest gradient from the
+          sorted list. Evaluate the images, magnification and magnification gradient at
+          new locations $0.5(\phi^{(n)} + \phi^{(n+1)})$ (the midpoints of the intervals
+          with largest gradient). Append new points to previous points.
+    \item Go back to step 2. and repeat until $C_\mathrm{geom}\times N < 2$. Set $N=2$
+          for the final iteration.
+\end{enumerate}
+Figure~\ref{fig:limb_sampling2} shows multiple iterations of the algorithm in the single
+lens case for a source whose limb is just outside the caustic. The point source magnification
+is plotted as a function of $\phi$. we see that the caustic is located at
+$\phi\approx -160^\circ$ and the density of points near the caustic is lower than
+everywhere else. At subsequent iterations additional points are placed only where
+the magnification gradient is large which is close to the caustic. The effect of this
+procedure on the contours in the image plane is shown in the right panel of
+Figure~\ref{fig:limb_sampling1}. Although the contours shown in the left and the right
+panels have the same total number of points, the adaptive sampling ensures that the
+points are far more evenly spread around the contour. The reason we place multiple new
+points at each iteration rather than a single point is that in case of binary
+and triple lenses there are generally multiple peaks in the magnification pattern
+along the limb  because the limb can intersect multiple caustics and placing one
+point at the time would result in denser sampling for only a single peak.
+
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/limb_sampling1.pdf}
+        \caption{Iterative evaluation of the point source magnification and images
+            along the source star limb. At iteration 0 we
+            uniformly distribute a set of points on the limb of the source
+            parametrized by $\theta_\mathrm{limb}$. At each subsequent iteration
+            we evaluate the point source magnification at every point, and then place
+            fewer and fewer points in geometric progression at locations with largest
+            magnification gradient. This procedure ensures dense sampling of the region
+            around the caustic.}
+        \label{fig:limb_sampling2}
+    \end{centering}
+\end{figure}
+
+\section{Magnification of a uniform brigthness source}
+We can evaluate the 1D integral in Equation~\ref{eq:mag_green_theorem_unif}
+using the (composite) \ssf{trapezoidal rule} \footnote{\begin{equation}\int_{x_0}^{x_n}\,f(x)\ud x\approx \sum_{i=1}^{n-1} \frac{1}{2}(f(x_i) + f(x_{i+1}))\,(x_{i+1} - x_i)\hquad .\end{equation}}:
+\begin{equation}
+    A_\mathrm{tot}\left(w_0, \rho_{\star}\right)\approx  \frac{1}{ \pi \rho_{\star}^{2}}
+    \frac{1}{4}\sum_{i=1}^{n}\left[\left(z_1^{(i)}+z_1^{(i+1)}\right)\Delta z_2^{(i)}
+    -\left(z_2^{(i)}+z_2^{(i+1)}\right)\Delta z_1^{(i)}\right]
+\end{equation}
+where $\Delta z_1^{(i)}= z_1^{(i+1)}-z_1^{(i)}$ and $\Delta z_2^{(i)}= z_2^{(i+1)}-z_2^{(i)}$.
+
+\section{Magnification of a limb-darkened source}
+
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/p_and_q_integrals.pdf}
+        \caption{}
+        \label{fig:p_and_q_integrals}
+    \end{centering}
+\end{figure}
+
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=\linewidth]{figures/gauss_legendre_vs_simpson.pdf}
+        \caption{}
+        \label{fig:gauss_legendre_vs_simpson}
+    \end{centering}
+\end{figure}
+
+\section{Computing lightcurves}
+\begin{figure}[t]
+    \begin{centering}
+        \includegraphics[width=0.6\linewidth]{figures/single_lens_hex.pdf}
+        \caption{}
+        \label{fig:single_lens_hex}
+    \end{centering}
+\end{figure}
+
+\section{Automatic differentiation}
+
+% CHAPTER 4: Modeling binary and triple microlensing events
 \chapter{Modeling microlensing events}
-\label{ch:microlensing}
+\label{ch:microlensing_modeling}
+%% PLOT POINTWISE LOO
+
 \section{Single lens events}
-\section{Modeling single lens events}
-\section{Numerical solutions to the lens equation}
-\section{\textsf{caustics} -- computing the magnification of an extended limb-darkened source}
-\section{Comparison with previous work}
-\section{Modeling binary lens events}
+%Although the majority of effort dedicated to modeling microlensing events in recent 
+%years has been dedicated to binary and triple lens events, single lens events 
+%exhibit many of the same problems (such as degeneracies) as binary and triple 
+%lens events and they share common parameters. It is thus instructive to understand
+%the challenges in modeling single lens events before tackling events with more 
+%lenses. Single lens events are also interesting scientifically because we can use 
+%them to constrain the occurance rates of free-floating planets (CITE) and study 
+%isolated Black Hole populations. They're also about an order of magnitude more 
+%common than binary and triple lens events.
+%
+%
+%Models with multiple lenses share all of the parameters with the single lens model,
+%to make progress with these often pathological models, we need to understand the simpler model first.
+%The single lens model is also not useless by itself since one of the most well known results from microlensing 
+%in the past decade, the reported discovery of a large population of a large population 
+%of unbound free floating planets in the Milky Way by \cite{sumi_unbound_2011} (and 
+%the subsequent correction by \cite{mroz_no_2017}) relies entirely on inferences 
+%drawn using a single lens model in a regime particularly  sensitive to modeling 
+%assumptions.
+
+\section{\ssf{caustics} -- computing the magnification of an extended limb-darkened source}
+\section{Binary and triple lens events}
 \section{Population-level inference using hierarchical modeling}
 
-% CHAPTER 4: Mapping the surface of Io 
+% CHAPTER 5: Mapping the surface of Io 
 \chapter{Mapping the surface of Io}
@@
 Equation~\ref{eq:complex_poly} is complex polynomial of degree $N^2+1$. To obtain coefficients of this
-polynomial in terms of fundamental parameters we use a computer algebra software package \textsf{SymPy}\footnote{
+polynomial in terms of fundamental parameters we use a computer algebra software package \ssf{SymPy}\footnote{
     \url{https://www.sympy.org}
@@
 \href{https://github.com/fbartolic/caustics/blob/main/notebooks/ComplexPolynomialCoefficients.ipynb}{here}.
-I have managed to derive the coefficient for the binary and triple lens case but \textsf{SymPy} fails to
+I have managed to derive the coefficient for the binary and triple lens case but \ssf{SymPy} fails to
 converge for the quadruple lens case because the number of coefficients becomes very large.
@@
 Equation~\ref{eq:complex_poly_critical} is a complex polynomial of degree $2N$. We can
-obtain its coefficients with \textsf{SymPy} as before and their values are available
+obtain its coefficients with \ssf{SymPy} as before and their values are available
 \href{https://github.com/fbartolic/caustics/blob/main/notebooks/ComplexPolynomialCoefficients.ipynb}{here}.

### 2022-08-10 src/figures/gauss_legendre_vs_simpson.py
Switch to new showyourwork version, Chapter 3 almost done

@@ -1,234 +1,0 @@
-from functools import partial
-
-import numpy as np
-from jax.config import config
-import jax.numpy as jnp
-from jax import random, jit, vmap, lax
-
-import matplotlib.pyplot as plt
-
-from caustics.utils import *
-from caustics.extended_source_magnification import (
-    _images_of_source_limb,
-)
-from caustics.integrate import _brightness_profile
-
-from scipy.special import roots_legendre
-
-config.update("jax_enable_x64", True)
-config.update("jax_platform_name", "cpu")
-
-
-@jit
-def simpson_quadrature(x, y):
-    """
-    Compute an integral using Simpson's 1/3 rule assuming that x is uniformly
-    distributed. `len(x)` has to be an odd number.
-    """
-    # len(x) must be odd
-    h = (x[-1] - x[0]) / (len(x) - 1)
-    return h / 3.0 * jnp.sum(y[0:-1:2] + 4 * y[1::2] + y[2::2], axis=0)
-
-
-@partial(jit, static_argnames=("npts"))
-def compute_p_and_q(
-    w_center,
-    rho,
-    contours,
-    parity,
-    tail_idcs,
-    u1=0.0,
-    npts=201,
-):
-    def P(_, y0, xl, yl):
-        # Construct grid in z2 and evaluate the brightness profile at each point
-        y = jnp.linspace(y0 * jnp.ones_like(xl), yl, npts)
-
-        integrands = _brightness_profile(xl + 1j * y, rho, w_center, u1=u1, nlenses=1)
-        I = simpson_quadrature(y, integrands)
-        return -0.5 * I
-
-    def Q(x0, _, xl, yl):
-        # Construct grid in z1 and evaluate the brightness profile at each point
-        x = jnp.linspace(x0 * jnp.ones_like(xl), xl, npts)
-
-        integrands = _brightness_profile(x + 1j * yl, rho, w_center, u1=u1, nlenses=1)
-        I = simpson_quadrature(x, integrands)
-        return 0.5 * I
-
-    # We choose the centroid of each contour to be lower limit for the P and Q
-    # integrals
[…]

### 2022-08-10 src/figures/limb_sampling1.py

@@ -1,115 +1,0 @@
-import numpy as np
-from jax.config import config
-import jax.numpy as jnp
-from jax import random
-
-import matplotlib.pyplot as plt
-from matplotlib.ticker import AutoMinorLocator
-
-
-from caustics import images_point_source
-from caustics.point_source_magnification import lens_eq_det_jac
-from caustics.utils import *
-
-config.update("jax_enable_x64", True)
-config.update("jax_platform_name", "cpu")
-
-
-def _images_of_source_limb(
-    w_center,
-    rho,
-    nlenses=2,
-    npts_init=500,
-    niter=2,
-    roots_itmax=2500,
-    roots_compensated=False,
-    **params,
-):
-    def images_and_mag(theta):
-        z, z_mask = images_point_source(
-            rho * jnp.exp(1j * theta) + w_center,
-            nlenses=nlenses,
-            roots_itmax=roots_itmax,
-            roots_compensated=roots_compensated,
-            **params,
-        )
-        det = lens_eq_det_jac(z, nlenses=nlenses, **params)
-        parity = jnp.sign(det)
-        mag = jnp.sum((1.0 / jnp.abs(det)) * z_mask, axis=0)
-        return z, z_mask, mag, parity
-
-    # Initial sampling on the source limb
-    theta = jnp.linspace(-np.pi, np.pi, npts_init - 1, endpoint=False)
-    theta = jnp.pad(theta, (0, 1), constant_values=np.pi - 1e-05)
-    z, z_mask, mag, parity = images_and_mag(theta)
-
-    # Refine sampling by placing geometrically fewer points each iteration
-    # in the regions where the magnification gradient is largest
-    npts_list = np.geomspace(2, npts_init, niter, endpoint=False, dtype=int)[::-1]
-    key = random.PRNGKey(42)
-
-    thetas = [theta]
-    mags = [mag]
-
-    for _npts in npts_list:
-        # Resample theta
-        delta_mag = jnp.gradient(mag)
-        idcs_maxdelta = jnp.argsort(jnp.abs(delta_mag))[::-1][:_npts]
-        theta_patch = 0.5 * (theta[idcs_maxdelta] + theta[idcs_maxdelta + 1])
-
[…]

### 2022-08-10 src/figures/limb_sampling2.py

@@ -1,157 +1,0 @@
-from functools import partial
-
-import numpy as np
-from jax.config import config
-import jax.numpy as jnp
-from jax import random, jit, vmap, lax
-
-import matplotlib.pyplot as plt
-from matplotlib.ticker import AutoMinorLocator
-from matplotlib import colors
-
-from mpl_toolkits.axes_grid1.inset_locator import inset_axes
-
-from caustics import lens_eq, mag_point_source
-from caustics.utils import *
-from caustics.extended_source_magnification import (
-    _images_of_source_limb,
-)
-
-
-config.update("jax_enable_x64", True)
-config.update("jax_platform_name", "cpu")
-
-
-@partial(jit, static_argnames=("nlenses"))
-def _brightness_profile(z, rho, w_center, u1=0.0, nlenses=2, **params):
-    w = lens_eq(z, nlenses=nlenses, **params)
-    r = jnp.abs(w - w_center) / rho
-
-    def safe_for_grad_sqrt(x):
-        return jnp.sqrt(jnp.where(x > 0.0, x, 0.0))
-
-    B_r = jnp.where(
-        r <= 1.0,
-        1 + safe_for_grad_sqrt(1 - r**2),
-        0.01,
-    )
-
-    I = 3.0 / (3.0 - u1) * (u1 * B_r + 1.0 - 2.0 * u1)
-    return I
-
-
-def get_contours(w_center, rho, npts_limb=200, niter_limb=1):
-    images, images_mask, images_parity = _images_of_source_limb(
-        w_center,
-        rho,
-        npts_init=npts_limb,
-        niter=niter_limb,
-        nlenses=1,
-    )
-    # Per image parity
-    parity = images_parity[:, 0]
-
-    # Set last point to be equal to first point
-    contours = jnp.hstack([images, images[:, 0][:, None]])
-    return contours, parity
-
-
-def plot_contour(ax, x, y, parity):
-    if parity == 1.0:
-        color = "C0"
-    else:
-        color = "C1"
-
-    s_min = 1.0
-    s_max = 12
-    s_vals = (np.linspace(s_min, s_max, len(x)) ** 2)[::-1]
[…]

### 2022-08-10 src/scripts/adaptive_sampling.py

@@ -1,0 +1,253 @@
+import paths
+import numpy as np
+from jax.config import config
+import jax.numpy as jnp
+from jax import random
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
+from matplotlib import colors
+
+from caustics import images_point_source, critical_and_caustic_curves, mag_point_source
+from caustics.point_source_magnification import lens_eq_det_jac
+from caustics.extended_source_magnification import _eval_images_sequentially
+
+from caustics.utils import *
+
+
+config.update("jax_enable_x64", True)
+config.update("jax_platform_name", "cpu")
+
+def _images_of_source_limb(
+    w0,
+    rho,
+    nlenses=2,
+    npts=300,
+    niter=10,
+    roots_itmax=2500,
+    roots_compensated=False,
+    **params,
+):
+    key = random.PRNGKey(0)
+    key1, key2 = random.split(key)
+
+    def fn(theta, z_init):
+        # Add a small perturbation to z_init to avoid situations where
+        # there is convergence to the exact same roots when difference in theta
+        # is very small and z_init is very close to the exact roots
+        u1 = random.uniform(key1, shape=z_init.shape, minval=-1e-6, maxval=1e-6)
+        u2 = random.uniform(key2, shape=z_init.shape, minval=-1e-6, maxval=1e-6)
+        z_init = z_init + u1 + u2 * 1j
+
+        z, z_mask = images_point_source(
+            rho * jnp.exp(1j * theta) + w0,
+            nlenses=nlenses,
+            roots_itmax=2500,
+            roots_compensated=roots_compensated,
+            z_init=z_init.T,
+            custom_init=True,
+            **params,
+        )
+        det = lens_eq_det_jac(z, nlenses=nlenses, **params)
+        z_parity = jnp.sign(det)
+        mag = jnp.sum((1.0 / jnp.abs(det)) * z_mask, axis=0)
[…]

### 2022-08-10 src/scripts/ads_keyword_bayesian.py

@@ -1,8 +1,9 @@
+import paths
 import numpy as np
 from matplotlib import pyplot as plt
 
 
-data = np.loadtxt("../data/misc/ads_keyword_bayes.csv", delimiter=",")
+data = np.loadtxt(paths.data/"misc/ads_keyword_bayes.csv", delimiter=",")
 year, count, count_ref = data.T
 
 fig, ax = plt.subplots(figsize=(9, 5))
@@ -13,5 +14,5 @@ ax.set_ylabel("Number of entries for the keyword\n 'Bayesian' on NASA ADS")
 ax.set_xlabel("year")
 
 # Save as pdf
-fig.savefig("ads_keyword_bayesian.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"ads_keyword_bayesian.pdf", bbox_inches="tight")
 

### 2022-08-10 src/scripts/binary_lens_topology.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import jax.numpy as jnp
 from jax.config import config
@@ -101,4 +102,4 @@ for _a in ax[:, 1]:
     _a.set_yticklabels([])
 
 
-fig.savefig("binary_lens_topology.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"binary_lens_topology.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/close_wide_degeneracy.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import jax.numpy as jnp
 from jax.config import config
@@ -102,4 +103,4 @@ ax_maps[1].set_ylabel(r"$\mathrm{Im}(w)$")
 
 
 # Save to file
-fig.savefig("close_wide_degeneracy.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"close_wide_degeneracy.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/gp_samples.py

@@ -1,0 +1,35 @@
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+import jax
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+from tinygp import kernels
+from tinygp import GaussianProcess
+
+
+# Evaluation points
+X = np.sort(np.random.default_rng(1).uniform(0, 10, 300))
+gp1 = GaussianProcess(kernels.ExpSquared(scale=0.5), X)
+gp2 = GaussianProcess(kernels.ExpSquared(scale=2.0), X)
+
+y1 = gp1.sample(jax.random.PRNGKey(4), shape=(10,))
+y2 = gp2.sample(jax.random.PRNGKey(4), shape=(10,))
+
+fig, ax = plt.subplots(1, 2, figsize=(12, 3), sharey=True, gridspec_kw={"wspace": 0.1})
+ax[0].plot(X, y1.T, color="k", lw=1.0, alpha=0.7)
+ax[1].plot(X, y2.T, color="k", lw=1.0, alpha=0.7)
+
+for a in ax:
+    a.set(
+        xlabel="$x$", ylim=(-3, 3), xlim=(0, 10),
+    )
+
+ax[0].set(ylabel="$y=f(x)$")
+ax[0].set_title("$l=0.5$")
+ax[1].set_title("$l=2.0$")
+
+# Save as pdf
+fig.savefig(paths.figures/"gp_samples.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/gp_samples_conditioned.py

@@ -1,0 +1,55 @@
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+import jax
+from jax.config import config
+
+config.update("jax_enable_x64", True)
+
+from tinygp import kernels
+from tinygp import GaussianProcess
+
+# Test points
+X = np.sort(np.random.default_rng(1).uniform(0, 10, 300))
+
+# Simulate data
+random = np.random.default_rng(2)
+X_obs = np.sort(random.uniform(2, 8, 8))
+y_obs1 = np.sin(2 * X_obs) + 1e-4 * random.normal(size=X_obs.shape)
+y_obs2 = np.sin(2 * X_obs) + 1e-1 * random.normal(size=X_obs.shape)
+
+
+# Condition
+kernel = 0.5 * kernels.ExpSquared(scale=1.0)
+gp1 = GaussianProcess(kernel, X_obs, diag=1e-4)
+gp2 = GaussianProcess(kernel, X_obs, diag=1e-01 ** 2)
+cond1 = gp1.condition(y_obs1)
+cond2 = gp2.condition(y_obs2)
+_, cond_gp1 = gp1.condition(y_obs1, X)
+_, cond_gp2 = gp2.condition(y_obs2, X)
+y_samp1 = cond_gp1.sample(jax.random.PRNGKey(1), shape=(12,))
+y_samp2 = cond_gp2.sample(jax.random.PRNGKey(1), shape=(12,))
+
+fig, ax = plt.subplots(1, 2, figsize=(12, 3), sharey=True, gridspec_kw={"wspace": 0.1})
+
+ax[0].plot(X, y_samp1[0], "k", lw=1.0, alpha=0.5, label="samples")
+ax[0].plot(X, y_samp1[1:].T, "k", lw=1.0, alpha=0.5)
+ax[0].plot(X_obs, y_obs1, "ko")
+
+ax[1].plot(X, y_samp2[0], "k", lw=1.0, alpha=0.5, label="samples")
+ax[1].plot(X, y_samp2[1:].T, "k", lw=1.0, alpha=0.5)
+ax[1].plot(X_obs, y_obs2, "ko", label="data")
+ax[1].legend(prop={"size": 12}, loc="lower right")
+
+for a in ax:
+    a.set(
+        xlabel="$x$", ylim=(-2, 2), xlim=(0, 10),
+    )
+
+ax[0].set(ylabel="$y=f(x)$")
+ax[0].set_title("noiseless observations")
[…]

### 2022-08-10 src/scripts/initial_sampling.py

@@ -1,0 +1,113 @@
+import paths
+import numpy as np
+from jax.config import config
+import jax.numpy as jnp
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics import  critical_and_caustic_curves
+from caustics.extended_source_magnification import _eval_images_sequentially
+
+from caustics.utils import *
+
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+
+def get_images(
+    w0,
+    rho,
+    nlenses=2,
+    npts=300,
+    roots_itmax=2500,
+    roots_compensated=False,
+    **params,
+):
+    # Initial sampling on the source limb
+    npts_init = int(0.5 * npts)
+    theta = jnp.linspace(-np.pi, np.pi, npts_init - 1, endpoint=False)
+    theta = jnp.pad(theta, (0, 1), constant_values=np.pi - 1e-8)
+    z, z_mask, z_parity = _eval_images_sequentially(
+        theta, w0, rho, nlenses, roots_compensated, roots_itmax, **params
+    )
+
+    return z, z_mask, z_parity
+
+
+def plot_track(ax, x, y, mask, parity, color, connect=True, label=''):
+    s_min = 3.
+    s_max = 12
+    s_vals = (np.linspace(s_min, s_max, len(x))**2)[::-1]
+
+    for _x, _y, m, p, s in zip(x, y, mask, parity, s_vals):
+        if m == True:
+            marker ='o' 
+        else:
+            marker ='x'
+
+        ax.scatter(_x, _y, s=s, alpha=0.6,color=color, marker=marker, zorder=-1)
+
+    if connect:
+        ax.plot(x, y, color=color, linewidth=1.5, alpha=0.5, zorder=-1, label=label)
+
+
+
+
+w0 = -0.67 + -0.027j
+a, e1, e2, r3 = 0.698, 0.02809, 0.9687, -0.0197 - 0.95087j
+rho = 5e-2
+
+z, z_mask, z_parity = get_images(
+    w0,
+    rho,
+    npts=200,
+    nlenses=3,
+    a=a,
+    r3=r3,
+    e1=e1,
+    e2=e2,
+)
[…]

### 2022-08-10 src/scripts/p_and_q_integrals.py

@@ -1,10 +1,11 @@
+import paths
 import numpy as np
 from jax.config import config
 import jax.numpy as jnp
-from jax import random, jit, vmap, lax
+from jax import vmap
 
 import matplotlib.pyplot as plt
-import matplotlib.colors as colors
+from matplotlib.ticker import AutoMinorLocator
 
 from caustics.utils import *
 from caustics.extended_source_magnification import (
@@ -16,17 +17,15 @@ from caustics.integrate import _brightness_profile
 config.update("jax_enable_x64", True)
 config.update("jax_platform_name", "cpu")
 
-npts_limb = 200
-niter_limb = 1
-u1 = 0.3
-rho = 0.1
+npts_limb = 80
+u1 = 0.5
+rho = 0.05
 w_center = 1.5 * rho * jnp.exp(1j * jnp.pi / 4)
 
 images, images_mask, images_parity = _images_of_source_limb(
     w_center,
     rho,
-    npts_init=npts_limb,
-    niter=niter_limb,
+    npts=npts_limb,
     nlenses=1,
 )
 # Per image parity
@@ -46,7 +45,7 @@ I_eval = _brightness_profile(zgrid, rho, w_center, u1=u1, nlenses=1)
 # Evaluate the P and Q integrals for a single image and point on contour
 z0 = vmap(lambda idx, contour: contour.sum() / (idx + 1))(tail_idcs, contours)
 idx_image = 0
-idx_point = 160
+idx_point = 70
 z_origin = z0[idx_image]
 z_limb = contours[idx_image, idx_point]
 
@@ -61,6 +60,7 @@ Q_integrand = _brightness_profile(
     x_linspace + 1j * jnp.imag(z_limb), rho, w_center, u1=0.2, nlenses=1
 )
 
+
 axd = plt.figure(constrained_layout=True, figsize=(13, 8)).subplot_mosaic(
     """
     AB
@@ -71,13 +71,23 @@ axd = plt.figure(constrained_layout=True, figsize=(13, 8)).subplot_mosaic(
 im = axd["A"].pcolormesh(xgrid_im, ygrid_im, I_eval, cmap="Greys", zorder=-1)
 # plt.colorbar(im, ax=ax)
 axd["A"].set_aspect(1)
[…]

### 2022-08-10 src/scripts/paczynski_curve.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import matplotlib.pyplot as plt
 
@@ -51,4 +52,5 @@ ax.set_xticks([-1.0, -0.5, 0.0, 0.5, 1.0])
 ax.legend(prop={"size": 16})
 
 # Save to disk
-fig.savefig("paczynski_curve.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"paczynski_curve.pdf", bbox_inches="tight")
+

### 2022-08-10 src/scripts/paths.py

@@ -1,0 +1,25 @@
+"""
+Exposes common paths useful for manipulating datasets and generating figures.
+"""
+from pathlib import Path
+
+# Absolute path to the top level of the repository
+root = Path(__file__).resolve().parents[2].absolute()
+
+# Absolute path to the `src` folder
+src = root / "src"
+
+# Absolute path to the `src/data` folder (contains datasets)
+data = src / "data"
+
+# Absolute path to the `src/static` folder (contains static images)
+static = src / "static"
+
+# Absolute path to the `src/scripts` folder (contains figure/pipeline scripts)
+scripts = src / "scripts"
+
+# Absolute path to the `src/tex` folder (contains the manuscript)
+tex = src / "tex"
+
+# Absolute path to the `src/tex/figures` folder (contains figure output)
+figures = tex / "figures"

### 2022-08-10 src/scripts/planetary_caustics.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import jax.numpy as jnp
 from jax.config import config
@@ -75,4 +76,4 @@ plt.colorbar(
 )
 
 # Save to file
-fig.savefig("planetary_caustics.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"planetary_caustics.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/quadrature_comparison.py

@@ -1,0 +1,211 @@
+
+import paths
+from functools import partial
+import numpy as np
+from jax.config import config
+import jax.numpy as jnp
+from jax import  jit, vmap
+
+import matplotlib.pyplot as plt
+
+from caustics import lens_eq
+from caustics.utils import *
+from caustics.extended_source_magnification import (
+    _images_of_source_limb,
+)
+from caustics.integrate import (
+    _brightness_profile, _integrate_gauss_legendre
+)
+import scipy
+
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+
+@jit
+def simpson_quadrature(x, y):
+    """
+    Compute an integral using Simpson's 1/3 rule assuming that x is uniformly
+    distributed. `len(x)` has to be an odd number.
+    """
+    # len(x) must be odd
+    h = (x[-1] - x[0]) / (len(x) - 1)
+    return h / 3.0 * jnp.sum(y[0:-1:2] + 4 * y[1::2] + y[2::2], axis=0)
+
+
+def compute_P_integrand_scipy(
+    w_center, rho, u1=0.0, nlenses=1,  npts_limb=100, 
+):
+    z, z_mask, z_parity = _images_of_source_limb(
+        w_center,
+        rho,
+        npts=npts_limb,
+        nlenses=1,
+    )
+
+    # Set last point to be equal to first point
+    z= jnp.hstack([z, z[:, 0][:, None]])
+    tail_idcs = jnp.array([z.shape[1] - 1, z.shape[1] - 1])
+
+    # Compute the Legendre roots and weights for use in Gaussian quadrature
+    def P(_, y0, xl, yl):
+        # Construct grid in z2 and evaluate the brightness profile at each point
+        a, b = y0 * jnp.ones_like(xl), yl  # lower and upper limits
+
+        # Integrate
+        I = []
+        for i in range(len(xl)):
+            f = lambda y: np.array(_brightness_profile(
+                xl[i] + 1j * y, rho, w_center, u1=u1, nlenses=nlenses, 
+            ))
+            val, err = scipy.integrate.quad(f, a[i], b[i], limit=1000, epsrel=1e-12)
+            I.append(val)
+        return -0.5 * jnp.array(I)
+
[…]

### 2022-08-10 src/scripts/segments_and_contours.py

@@ -1,0 +1,166 @@
+import paths
+import numpy as np
+from jax.config import config
+import jax.numpy as jnp
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
+
+from caustics import critical_and_caustic_curves
+from caustics.extended_source_magnification import (
+    _images_of_source_limb,
+    _get_segments,
+    _contours_from_closed_segments,
+    _contours_from_open_segments,
+)
+from caustics.utils import *
+
+config.update("jax_enable_x64", True)
+config.update("jax_platform_name", "cpu")
+
+def plot_contour(ax, x, y,  color):
+    s_min = 1.
+    s_max = 12
+    s_vals = (np.linspace(s_min, s_max, len(x))**2)[::-1]
+    ax.scatter(x, y, color=color, s=s_vals, marker='o', alpha=0.8, zorder=-1)
+    ax.plot(x, y, color=color, linewidth=1., alpha=0.5, zorder=-1)
+
+w0 = -0.67 + -0.027j
+a, e1, e2, r3 = 0.698, 0.02809, 0.9687, -0.0197 - 0.95087j
+rho = 5e-2
+
+z, z_mask, z_parity = _images_of_source_limb(
+    w0,
+    rho,
+    npts=200,
+    niter=10,
+    nlenses=3,
+    a=a,
+    r3=r3,
+    e1=e1,
+    e2=e2,
+)
+
+segments_closed, segments_open, _ = _get_segments(z, z_mask, z_parity, nlenses=3)
+cc, _ = critical_and_caustic_curves(
+    nlenses=3, npts=1000, a=a, e1=e1, e2=e2, r3=r3, rho=rho
+)
+contours1, contours1_p = _contours_from_closed_segments(segments_closed)
+contours2, contours2_p = _contours_from_open_segments(segments_open)
+
+segments_list = []
+for i in range(len(segments_open)):
+    if not jnp.all(segments_open[i] == 0 + 0j):
+        segments_list.append(segments_open[i])
+for i in range(len(segments_closed)):
+    if not jnp.all(segments_closed[i] == 0 + 0j):
+        segments_list.append(segments_closed[i])
+
[…]

### 2022-08-10 src/scripts/single_lens_astrometric_shift.py

@@ -1,0 +1,30 @@
+import paths
+import numpy as np
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+tau = np.linspace(-1.1, 1.1, 1000)
+u0 = np.array([0.1, 0.3, 0.5, 0.7, 0.9])
+
+fig, ax = plt.subplots(figsize=(8, 8))
+
+ax.axhline(0, alpha=0.5, color="grey", lw=1)
+ax.axvline(0, alpha=0.5, color="grey", lw=1)
+
+for _u0 in u0:
+    u = np.linspace(-50.0, 50.0, 5000) + _u0 * 1j
+
+    delta_c_x = (u / (np.abs(u) ** 2 + 2)).real
+    delta_c_y = (u / (np.abs(u) ** 2 + 2)).imag
+
+    ax.plot(delta_c_x, delta_c_y, label=f"$u_0=${_u0}", lw=2)
+
+ax.set_aspect(1)
+ax.xaxis.set_minor_locator(AutoMinorLocator())
+ax.yaxis.set_minor_locator(AutoMinorLocator())
+ax.set(xlabel=r"$\delta w_{c,1}$", ylabel="$\delta w_{c, 2}$")
+ax.legend(prop={"size": 14})
+
+
+fig.savefig(paths.figures/"single_lens_astrometric_shift.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/single_lens_hex.py

@@ -1,3 +1,4 @@
+import paths
 from functools import partial
 
 from jax.config import config
@@ -106,4 +107,4 @@ for a in ax:
     a.xaxis.set_minor_locator(AutoMinorLocator())
     a.yaxis.set_minor_locator(AutoMinorLocator())
 
-fig.savefig("single_lens_hex.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"single_lens_hex.pdf", bbox_inches="tight")

### 2022-08-10 src/scripts/single_lens_images.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import jax.numpy as jnp
 from jax import jit, vmap
@@ -73,4 +74,4 @@ for a in ax.reshape(-1):
     a.set_rasterization_zorder(0)
 
 # Save to disk
-fig.savefig("single_lens_images.pdf", bbox_inches="tight", dpi=200)
+fig.savefig(paths.figures/"single_lens_images.pdf", bbox_inches="tight", dpi=200)

### 2022-08-10 src/scripts/single_lens_images_and_contours.py

@@ -1,0 +1,138 @@
+import paths
+from functools import partial
+
+import numpy as np
+from jax.config import config
+import jax.numpy as jnp
+from jax import jit
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+from matplotlib import colors
+
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
+
+from caustics import (
+    lens_eq,
+    mag_point_source,
+)
+from caustics.utils import *
+from caustics.extended_source_magnification import (
+    _images_of_source_limb,
+)
+
+config.update("jax_enable_x64", True)
+config.update("jax_platform_name", "cpu")
+
+
+@partial(jit, static_argnames=("nlenses"))
+def _brightness_profile(z, rho, w_center, u1=0.0, nlenses=2, **params):
+    w = lens_eq(z, nlenses=nlenses, **params)
+    r = jnp.abs(w - w_center) / rho
+
+    def safe_for_grad_sqrt(x):
+        return jnp.sqrt(jnp.where(x > 0.0, x, 0.0))
+
+    B_r = jnp.where(
+        r <= 1.0,
+        1 + safe_for_grad_sqrt(1 - r**2),
+        0.01,
+    )
+
+    I = 3.0 / (3.0 - u1) * (u1 * B_r + 1.0 - 2.0 * u1)
+    return I
+
+
+def get_contours(w_center, rho, npts_limb=200):
+    images, images_mask, images_parity = _images_of_source_limb(
+        w_center,
+        rho,
+        npts_init=npts_limb,
+        geom_factor=0.0,
+        nlenses=1,
+    )
+    # Per image parity
+    parity = images_parity[:, 0]
+
+    # Set last point to be equal to first point
+    contours = jnp.hstack([images, images[:, 0][:, None]])
+    return contours, parity
+
+
+npts_limb = 200
+u1 = 0.3
+rho = 0.3
+w_center = 1.05 * rho * jnp.exp(1j * jnp.pi / 4)
+l = 1.5
+
+contours1, parity1 = get_contours(w_center, rho, npts_limb=npts_limb)
+
+# Surface brightness in image plane
[…]

### 2022-08-10 src/scripts/spherical_harmonics.py

@@ -1,3 +1,4 @@
+import paths
 import starry
 import matplotlib.pyplot as plt
 
@@ -59,4 +60,4 @@ for i, l in enumerate(range(ydeg + 1)):
         ax[i, j].set_rasterization_zorder(0)
 
 # Save to file
-plt.savefig("spherical_harmonics.pdf", bbox_inches="tight", dpi=200)
+plt.savefig(paths.figures/"spherical_harmonics.pdf", bbox_inches="tight", dpi=200)

### 2022-08-10 src/scripts/triple_lens_caustics.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import jax.numpy as jnp
 from jax.config import config
@@ -45,5 +46,5 @@ plt.colorbar(im, label="magnification")
 ax.set_rasterization_zorder(0)
 
 # Save to disc
-fig.savefig("triple_lens_caustics.pdf", bbox_inches="tight", dpi=200)
+fig.savefig(path.figures/"triple_lens_caustics.pdf", bbox_inches="tight", dpi=200)
 

### 2022-08-10 src/scripts/typical_sets.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 import matplotlib.pyplot as plt
 
@@ -28,5 +29,5 @@ for a in ax.ravel():
     a.set_xlabel("Distance from MAP")
 
 # Save as pdf
-fig.savefig("typical_sets.pdf", bbox_inches="tight")
+fig.savefig(paths.figures/"typical_sets.pdf", bbox_inches="tight")
 

### 2022-09-24 src/scripts/adaptive_sampling.py
major changes

@@ -11,7 +11,7 @@ from matplotlib import colors
 
 from caustics import images_point_source, critical_and_caustic_curves, mag_point_source
 from caustics.point_source_magnification import lens_eq_det_jac
-from caustics.extended_source_magnification import _eval_images_sequentially
+from caustics.extended_source_magnification import _images_point_source_sequential 
 
 from caustics.utils import *
 
@@ -58,10 +58,11 @@ def _images_of_source_limb(
     npts_init = int(0.5 * npts)
     theta = jnp.linspace(-np.pi, np.pi, npts_init - 1, endpoint=False)
     theta = jnp.pad(theta, (0, 1), constant_values=np.pi - 1e-8)
-    z, z_mask, z_parity = _eval_images_sequentially(
-        theta, w0, rho, nlenses, roots_compensated, roots_itmax, **params
+    z, z_mask = _images_point_source_sequential(
+        rho * jnp.exp(1j * theta) + w0, nlenses=nlenses, roots_itmax=roots_itmax, **params
     )
     det = lens_eq_det_jac(z, nlenses=nlenses, **params)
+    z_parity = jnp.sign(det)
     mag = jnp.sum((1.0 / jnp.abs(det)) * z_mask, axis=0)
 
     theta_init, z_init, z_mask_init, mag_init = theta, z, z_mask, mag
@@ -150,7 +151,7 @@ theta_init, z_init, z_mask_init, mag_init, theta_new, z_new, z_mask_new, mag_new
     e2=e2,
 )
 cc, _ = critical_and_caustic_curves(
-    nlenses=3, npts=1000, a=a, e1=e1, e2=e2, r3=r3, rho=rho
+    nlenses=3, npts=300, a=a, e1=e1, e2=e2, r3=r3, rho=rho
 )
 
 fig, ax = plt.subplot_mosaic(
@@ -243,7 +244,9 @@ ax['A'].set(xlabel=r"$\mathrm{Re}(w)$", ylabel=r"$\mathrm{Im}(w)$")
 ax['C'].set(xlabel=r"$\mathrm{Re}(z)$", ylabel=r"$\mathrm{Im}(z)$")
 
 # Critical curves
-ax['C'].scatter(cc.real, cc.imag, marker='.', color='k', s=1 ,alpha=0.6, zorder=-2)
+for z in cc:
+    ax['C'].plot(z.real, z.imag, color='k', lw=0.7, zorder=-2)
+
 
[…]

### 2022-09-24 src/scripts/binary_lens_topology.py

@@ -7,7 +7,7 @@ config.update("jax_enable_x64", True)
 
 import matplotlib.pyplot as plt
 
-from caustics import critical_and_caustic_curves_binary
+from caustics import critical_and_caustic_curves
 
 s_array = jnp.array([0.6, 1 / np.sqrt(2), 1.2, 2.0, 3.2])
 
@@ -15,15 +15,13 @@ fig, ax = plt.subplots(5, 2, figsize=(10, 18), sharex=True)
 fig.subplots_adjust(hspace=0.15, wspace=0.3)
 
 for i, s_ in enumerate(s_array):
-    critical_curves, caustic_curves = critical_and_caustic_curves_binary(
-        0.5 * s_, 0.5, npts=1000
+    critical_curves, caustic_curves = critical_and_caustic_curves(
+        a=0.5 * s_, e1=0.5, npts=3000, nlenses=2
     )
-
-    ax[i, 0].scatter(
-        critical_curves.real, critical_curves.imag, s=0.8, c="k", alpha=0.8
-    )
-    ax[i, 1].scatter(caustic_curves.real, caustic_curves.imag, s=0.8, c="k", alpha=0.8)
-
+    for z in critical_curves:
+        ax[i, 0].plot(z.real, z.imag, "k-", lw=0.7, zorder=-1)
+    for z in caustic_curves:
+        ax[i, 1].plot(z.real, z.imag, "k-", lw=0.7, zorder=-1)
 
 for a in ax.ravel():
     #    a.grid(True)

### 2022-09-24 src/scripts/extended_source_test.py

@@ -1,0 +1,86 @@
+import paths
+
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics import (
+    critical_and_caustic_curves,
+    images_point_source,
+)
+
+from caustics.utils import *
+from caustics.multipole import mag_hexadecapole
+from caustics.lightcurve import _caustics_proximity_test
+
+import MulensModel as mm
+
+def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=5e-05):
+    e2 = 1 - e1
+    x_cm = (e1 - e2)*a
+    bl = mm.BinaryLens(e2, e1, 2*a)
+    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
+
+# System parameters
+rho = 5e-03
+a = 0.8
+e1 = 0.95
+e2 = 1. - e1
+
+critical_curves, caustic_curves = critical_and_caustic_curves(nlenses=2, a=a, e1=e1, npts=100)
+
+
+_x = jnp.linspace(-0.5, -0.5 + 1.5, int(3.7*200))
+_y = jnp.linspace(-0.2, -0.2 + 0.4, 200)
+xgrid, ygrid = jnp.meshgrid(_x, _y)
+wgrid = xgrid + 1j*ygrid
+
+# Full calculation with VBB
+mags_ref = np.zeros_like(wgrid).astype(float)
+
+for i in range(wgrid.shape[0]):
+    for j in range(wgrid.shape[1]):
+        w_center = wgrid[i, j]
+        mags_ref[i, j]  = mag_vbb_binary(w_center, rho, a, e1, u1=0.0)
+
+# Evaluate hex approx. and the test
+z, z_mask = images_point_source(wgrid, nlenses=2, a=a, e1=e1)
+mu_multi, delta_mu_multi = mag_hexadecapole(z, z_mask, rho, nlenses=2, a=a,e1=e1)
+err_hex = jnp.abs(mu_multi - mags_ref)/mags_ref
+
+test = _caustics_proximity_test(
[…]

### 2022-09-24 src/scripts/fit_single_lens_model.py

@@ -1,0 +1,202 @@
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import random, jit, vmap, pmap
+import jax.scipy as jsp
+
+from astropy.table import Table
+from astropy.coordinates import SkyCoord
+from astropy import units as u
+
+from caustics.linalg import weighted_least_squares 
+from caustics.trajectory import AnnualParallaxTrajectory
+
+from scipy.signal import medfilt
+
+import ultranest
+
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import MCMC, NUTS
+from numpyro.infer import *
+numpyro.set_host_device_count(4)
+
+import arviz as az
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path_to_file):
+    t = Table.read(path_to_file, format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+lc = get_data("../data/microlensing/starBLG234.6.I.218982.dat")
+
+# Trim
+t = lc['HJD']
+fobs, ferr = magnitudes_to_fluxes(lc["mag"], lc["mag_err"])
+t, fobs, ferr = jnp.array(t), jnp.array(fobs), jnp.array(ferr)
+
+RA = "18:04:45.70"
+Dec = "-26:59:15.5"
+coords = SkyCoord(
+    RA, Dec, unit=(u.hourangle, u.deg, u.arcminute)
+)
+trajectory = AnnualParallaxTrajectory(t, coords)
+t0_est = t[np.argmax(medfilt(fobs, 9))]
+
+def model(t, t0_est, fobs, ferr, trajectory):
+    ln_t0 = numpyro.sample("ln_t0", dist.Normal(jnp.log(t0_est), jnp.log(10.)))
[…]

### 2022-09-24 src/scripts/lightcurve_gradients.py

@@ -1,0 +1,121 @@
+import paths
+
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import jit, jacfwd 
+
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics import (
+    mag,
+    critical_and_caustic_curves,
+)
+
+# Parameters
+a = 0.45
+e1 = 0.8
+e2 = 1. - e1
+q = e1/e2
+rho = 1e-02
+alpha = jnp.deg2rad(45.)
+u0 = 0.15 # COM frame
+t0 = 50.
+tE = 23.
+
+t = jnp.linspace(30, 70, 2000) # days
+
+@jit
+def get_mag(params):
+    s, q, rho, alpha, u0, t0, tE = params
+    a = 0.5*s
+    e1 = q/(1 + q)
+
+    x_cm = (e1 - e2)*a # b
+    u_mp = u0 - x_cm*jnp.sin(alpha)
+    t_mp = t0 - tE*(u0/jnp.tan(alpha) - u_mp/jnp.tan(alpha))
+
+    w_x = (t - t_mp)/tE*jnp.cos(alpha) - u_mp*jnp.sin(alpha)
+    w_y = (t - t_mp)/tE*jnp.sin(alpha) + u_mp*jnp.cos(alpha)
+    w_points = w_x + 1j*w_y
+
+    return w_points, mag(w_points, rho, nlenses=2, a=a, e1=e1)
+
+s = 2*a
+params = jnp.array([s, q, rho, alpha, u0, t0, tE])
+w_points, A = get_mag(params)
+
+# Evaluate the Jacobian at every point
+mag_jac = jit(jacfwd(lambda params: get_mag(params)[1]))
+jac_eval = mag_jac(params)
+
+fig, ax = plt.subplots(
+    8, 1,
+    figsize=(14, 14),
+    gridspec_kw={'height_ratios': [4, 1, 1, 1, 1, 1,1, 1], 'wspace':0.3},
+    sharex=True,
+)
+# Inset axes for images
+ax_in = inset_axes(ax[0],
+    width="60%", # 
+    height="60%", 
+    bbox_transform=ax[0].transAxes,
[…]

### 2022-09-24 src/scripts/mag_limbdark_accuracy.py

@@ -1,0 +1,114 @@
+import paths
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import random, lax
+
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics import (
+    critical_and_caustic_curves,
+    mag_extended_source,
+)
+
+from caustics.utils import *
+from caustics.integrate import *
+
+import MulensModel as mm
+
+def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=1e-05):
+    e2 = 1 - e1
+    x_cm = (e1 - e2)*a
+    bl = mm.BinaryLens(e2, e1, 2*a)
+    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
+
+
+
+@partial(jit, static_argnames=("npts_limb", "npts_ld", "limb_darkening"))
+def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
+    def body_fn(_, w):
+        mag = mag_extended_source(
+            w,
+            rho,
+            nlenses=2,
+            npts_limb=npts_limb,
+            limb_darkening=limb_darkening,
+            npts_ld=npts_ld,
+            u1=u1,
+            a=a,
+            e1=e1,
+        )
+        return 0, mag
+
+    _, mags = lax.scan(body_fn, 0, w_points)
+    return mags
+
+
+# Parameters
+a, e1 = 0.45, 0.8
+u1 = 0.7
+
+# 1000  points on caustic curve
+npts = 25 
+critical_curves, caustic_curves = critical_and_caustic_curves(
+    npts=npts, nlenses=2, a=a, e1=e1
+)
+caustic_curves = caustic_curves.reshape(-1)
+
+# Accuracy of integration
+acc_vbb = 1e-05
+npts_limb = 400
+
+
+# Compute relative error w.r.t. VBB for different values of rho 
+mags_vbb_list = []
+mags_list = []
+
+rho_list = [1.,1e-01, 1e-02, 1e-03]
+
+for rho in rho_list:
+    print(f"rho = {rho}")
+
[…]

### 2022-09-24 src/scripts/mag_uniform_accuracy.py

@@ -1,0 +1,117 @@
+import paths
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import random, lax
+
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics import (
+    critical_and_caustic_curves,
+    mag_extended_source,
+)
+
+from caustics.utils import *
+from caustics.integrate import *
+
+import MulensModel as mm
+
+def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=1e-05):
+    e2 = 1 - e1
+    x_cm = (e1 - e2)*a
+    bl = mm.BinaryLens(e2, e1, 2*a)
+    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
+
+
+@partial(jit, static_argnames=("npts_limb", "npts_ld", "limb_darkening"))
+def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
+    def body_fn(_, w):
+        mag = mag_extended_source(
+            w,
+            rho,
+            nlenses=2,
+            npts_limb=npts_limb,
+            limb_darkening=limb_darkening,
+            npts_ld=npts_ld,
+            u1=u1,
+            a=a,
+            e1=e1,
+        )
+        return 0, mag
+
+    _, mags = lax.scan(body_fn, 0, w_points)
+    return mags
+
+
+
+# Parameters
+a, e1, = (
+    0.45,
+    0.8,
+)
+
+# 1000  points on caustic curve
+npts = 250
+critical_curves, caustic_curves = critical_and_caustic_curves(
+    npts=npts, nlenses=2, a=a, e1=e1
+)
+caustic_curves = caustic_curves.reshape(-1)
+
+acc_vbb = 1e-05
+npts_limb = 400
+
+
+# Compute relative error w.r.t. VBB for different values of rho 
+mags_vbb_list = []
+mags_list = []
+
+rho_list = [1.,1e-01, 1e-02, 1e-03, 1e-04]
+
+for rho in rho_list:
+    print(f"rho = {rho}")
+
[…]

### 2022-09-24 src/scripts/performance_comparison.py

@@ -1,0 +1,184 @@
+import paths
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+from jax import random
+
+import matplotlib.pyplot as plt
+
+from caustics import (
+    critical_and_caustic_curves,
+    mag_extended_source,
+)
+
+from caustics.utils import *
+from caustics.integrate import *
+
+import MulensModel as mm
+
+import timeit
+
+def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=1e-05):
+    e2 = 1 - e1
+    x_cm = (e1 - e2)*a
+    bl = mm.BinaryLens(e2, e1, 2*a)
+    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
+
+def mag_ac_binary(w0, rho, a, e1, u1=0., accuracy=0.05, ld_accuracy=1e-03):
+    e2 = 1 - e1
+    x_cm = (e1 - e2)*a
+    bl = mm.BinaryLens(e2, e1, 2*a)
+    if u1==0.:
+        return bl.adaptive_contouring_magnification(
+            w0.real - x_cm, w0.imag, rho, accuracy=accuracy, 
+        )
+    else:
+        return bl.adaptive_contouring_magnification(
+            w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1, ld_accuracy=ld_accuracy
+        )
+
+def time(func, number=5, repeats=2):
+    times = 1000*np.array((timeit.Timer(func).repeat(repeat=repeats, number=number)))/number
+    return np.mean(times)
+
+# Parameters
+a, e1 = 0.45, 0.8
+u1 = 0.7
+
+npts = 3 # dozen points
+critical_curves, caustic_curves = critical_and_caustic_curves(
+    npts=npts, nlenses=2, a=a, e1=e1
+)
+caustic_curves = caustic_curves.reshape(-1)
+
+# Adjusted such that all codes are roughly equivalent in terms of accuracy
+acc_vbb = 6e-03
+acc_ac = 1e-02
+acc_ac_ld = 1e-03
+
+npts_limb = 500
+npts_ld = 100
+
+rho_list = [1., 1e-01, 1e-02, 1e-03]
+
+t_list = []
[…]

### 2022-09-24 src/scripts/planetary_caustics.py

@@ -9,7 +9,7 @@ config.update("jax_enable_x64", True)
 import matplotlib.pyplot as plt
 import matplotlib.colors as colors
 
-from caustics import mag_point_source_binary
+from caustics import mag_point_source
 
 q = 5e-03
 e1 = q / (1 + q)
@@ -26,7 +26,7 @@ for i, s_ in enumerate(s_array):
     wgrid = xgrid + 1j * ygrid
     a = 0.5 * s_
     x_cm = (2 * e1 - 1) * a
-    mag = mag_point_source_binary(wgrid, a, e1)
+    mag = mag_point_source(wgrid, a=a, e1=e1, nlenses=2)
     im = ax[i].pcolormesh(
         xgrid,
         ygrid,

### 2022-09-24 src/scripts/segments_and_contours.py

@@ -44,7 +44,7 @@ z, z_mask, z_parity = _images_of_source_limb(
 
 segments_closed, segments_open, _ = _get_segments(z, z_mask, z_parity, nlenses=3)
 cc, _ = critical_and_caustic_curves(
-    nlenses=3, npts=1000, a=a, e1=e1, e2=e2, r3=r3, rho=rho
+    nlenses=3, npts=300, a=a, e1=e1, e2=e2, r3=r3, rho=rho
 )
 contours1, contours1_p = _contours_from_closed_segments(segments_closed)
 contours2, contours2_p = _contours_from_open_segments(segments_open)
@@ -87,8 +87,9 @@ for i in range(len(segments_list)):
             f'C{i%10}',
         )
 
+for z in cc:
+    ax[0].plot(z.real, z.imag, color='k', lw=0.7) 
 
-ax[0].scatter(cc.real, cc.imag, marker='.', color='k', s=1 ,alpha=0.6, zorder=-2)
 ax[0].set_title("Contour segments")
 
 # BOTTOM PANEL 

### 2022-09-24 src/scripts/single_lens_fit_plots.py

@@ -1,0 +1,466 @@
+import paths
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import random
+
+from astropy.table import Table
+from astropy.coordinates import SkyCoord
+from astropy import units as u
+
+from caustics.linalg import weighted_least_squares 
+from caustics.trajectory import AnnualParallaxTrajectory
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+from scipy.signal import medfilt
+
+import arviz as az
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path_to_file):
+    t = Table.read(path_to_file, format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+
+def predict_flux(t, t_dense, samples, nsamples=20):
+    fpred_list = []
+    fpred_dense_list = []
+    w_list = []
+
+    for i in random.randint(random.PRNGKey(0), (nsamples,), 0, samples['u0'].values.shape[0]):
+        # Compute trajectory
+        w = trajectory.compute(
+            t, t0=jnp.exp(samples['ln_t0'].values[i]), tE=jnp.exp(samples['ln_tE'].values[i]), 
+            u0=samples['u0'].values[i], piEE=samples['piEE'].values[i], piEN=samples['piEN'].values[i]
+        )
+        w_dense = trajectory.compute(
+            t_dense, t0=jnp.exp(samples['ln_t0'].values[i]), tE=jnp.exp(samples['ln_tE'].values[i]), 
+            u0=samples['u0'].values[i], piEE=samples['piEE'].values[i], piEN=samples['piEN'].values[i]
+        )
+ 
[…]

### 2022-09-24 src/scripts/single_lens_hex.py

@@ -1,110 +1,0 @@
-import paths
-from functools import partial
-
-from jax.config import config
-import jax.numpy as jnp
-from jax import jit, lax
-
-import matplotlib.pyplot as plt
-from matplotlib.ticker import AutoMinorLocator
-
-from caustics import mag_extended_source, images_point_source
-from caustics.multipole import mag_hexadecapole
-
-
-config.update("jax_enable_x64", True)
-config.update("jax_platform_name", "cpu")
-
-
-@partial(
-    jit,
-    static_argnames=("nlenses", "npts_limb", "niter_limb", "limb_darkening", "npts_ld"),
-)
-def mag_extended_source_vectorized(
-    w_points,
-    rho,
-    limb_darkening=False,
-    u1=0.0,
-    nlenses=2,
-    npts_limb=200,
-    niter_limb=4,
-    npts_ld=50,
-):
-    # Iterate over w_points and execute either the hexadecapole  approximation
-    # or the full extended source calculation. `vmap` cannot be used here because
-    # `lax.cond` executes both branches within vmap.
-    mag_full = lambda w: mag_extended_source(
-        w,
-        rho,
-        limb_darkening=limb_darkening,
-        u1=u1,
-        nlenses=nlenses,
-        npts_limb=npts_limb,
-        niter_limb=niter_limb,
-        npts_ld=npts_ld,
-    )
-
-    def body_fn(_, x):
-        w = x
-        m = mag_full(w)
-        return 0, m
-
-    return lax.scan(body_fn, 0, w_points)[1]
-
-
-# Parameters
-u1 = 0.3
-rho = 0.05
-
-npts_limb = 150
-niter_limb = 15
-npts_ld = 100
-
-w_points = jnp.linspace(0.0, 3.0 * rho, 200)
-
-# Hex approx.
-z, mask_z = images_point_source(w_points, nlenses=1)
-mu_ps, mu_quad, mu_hex = mag_hexadecapole(z, mask_z, rho, u1=u1, nlenses=1)
-mag_multi = mu_ps + mu_quad + mu_hex
-
-# Full computation
-mag_full = mag_extended_source_vectorized(
-    w_points,
-    rho,
-    limb_darkening=True,
-    u1=u1,
-    nlenses=1,
-    npts_limb=npts_limb,
[…]

### 2022-09-24 src/scripts/single_lens_laplace_comparison.py

@@ -1,0 +1,362 @@
+import paths
+from functools import partial
+import numpy as np
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import random, vmap, jit
+import jax.scipy as jsp
+
+from astropy.table import Table
+from astropy.coordinates import SkyCoord
+from astropy import units as u
+
+from caustics.linalg import weighted_least_squares 
+from caustics.trajectory import AnnualParallaxTrajectory
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from scipy.signal import medfilt
+
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer.util import  log_likelihood, init_to_value
+from numpyro.infer import *
+from numpyro.infer.autoguide import AutoLaplaceApproximation
+from numpyro.infer.elbo import Trace_ELBO
+numpyro.set_host_device_count(4)
+
+from numpyro_ext.optim import optimize as optimize_numpyro_ext
+
+import arviz as az
+
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path_to_file):
+    t = Table.read(path_to_file, format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+def model(t, t0_est, fobs, ferr, trajectory):
+    ln_t0 = numpyro.sample("ln_t0", dist.Normal(jnp.log(t0_est), jnp.log(10.)))
+    ln_tE = numpyro.sample("ln_tE", dist.Normal(2.5, 1.5))
+    u0 = numpyro.sample("u0", dist.Normal(0.0, 1.))
+    piEE = numpyro.sample("piEE", dist.Normal(0.0, 1.))
+    piEN = numpyro.sample("piEN", dist.Normal(0.0, 1.))
+    numpyro.deterministic("piE", jnp.sqrt(piEE**2 + piEN**2))
+    t0, tE = jnp.exp(ln_t0), jnp.exp(ln_tE)
+    
[…]

### 2022-09-24 src/scripts/triple_lens_caustics.py

@@ -9,7 +9,7 @@ config.update("jax_enable_x64", True)
 import matplotlib.pyplot as plt
 import matplotlib.colors as colors
 
-from caustics import mag_point_source_triple
+from caustics import mag_point_source
 
 # Lens postions
 a = 0.4
@@ -20,6 +20,12 @@ e1 = 0.9
 e2 = 0.05
 e3 = 1.0 - e1 - e2
 
+psi = jnp.arctan2(r3.imag, r3.real)
+s = 0.5*a
+q = e2/e1
+q3 = (1 - e1-e2)/e1
+
+
 # Define 2D grid in the source plane
 npts = 1000
 
@@ -28,7 +34,7 @@ y = jnp.linspace(-0.45, 0.5, npts)
 xgrid, ygrid = jnp.meshgrid(x, y)
 wgrid = xgrid + 1j * ygrid
 
-mag = mag_point_source_triple(wgrid, a, r3, e1, e2)
+mag = mag_point_source(wgrid, s=s, q=q, q3=q3, r3=jnp.abs(r3), psi=psi, nlenses=3)
 
 fig, ax = plt.subplots(figsize=(12, 8))
 im = ax.pcolormesh(
@@ -46,5 +52,5 @@ plt.colorbar(im, label="magnification")
 ax.set_rasterization_zorder(0)
 
 # Save to disc
-fig.savefig(path.figures/"triple_lens_caustics.pdf", bbox_inches="tight", dpi=200)
+fig.savefig(paths.figures/"triple_lens_caustics.pdf", bbox_inches="tight", dpi=200)
 

### 2022-11-16 src/scripts/chapter_1/paths.py
add scripts

@@ -4,7 +4,7 @@ Exposes common paths useful for manipulating datasets and generating figures.
 from pathlib import Path
 
 # Absolute path to the top level of the repository
-root = Path(__file__).resolve().parents[2].absolute()
+root = Path(__file__).resolve().parents[3].absolute()
 
 # Absolute path to the `src` folder
 src = root / "src"

### 2022-11-16 src/scripts/chapter_2/triple_lens_caustics.py

@@ -1,5 +1,4 @@
 import paths
-import numpy as np
 import jax.numpy as jnp
 from jax.config import config
 
@@ -12,25 +11,21 @@ import matplotlib.colors as colors
 from caustics import mag_point_source
 
 # Lens postions
-a = 0.4
-r3 = 0.3 - 0.8j
-
-# Lens mass ratios
 e1 = 0.9
 e2 = 0.05
 e3 = 1.0 - e1 - e2
 
+s = 0.8
+r3 = 0.3 - 0.8j
 psi = jnp.arctan2(r3.imag, r3.real)
-s = 0.5*a
 q = e2/e1
 q3 = (1 - e1-e2)/e1
 
-
 # Define 2D grid in the source plane
-npts = 1000
+npts = 1200 
 
-x = jnp.linspace(-0.1, 1, npts)
-y = jnp.linspace(-0.45, 0.5, npts)
+x = jnp.linspace(-0.1, 1.2, npts)
+y = jnp.linspace(-1.3, -0.2, npts)
 xgrid, ygrid = jnp.meshgrid(x, y)
 wgrid = xgrid + 1j * ygrid
 
@@ -45,6 +40,8 @@ im = ax.pcolormesh(
     norm=colors.LogNorm(vmax=150),
     zorder=-1,
 )
+
+
 
 ax.set_aspect(1)
 ax.set(xlabel="$\mathrm{Re}(w)$", ylabel="$\mathrm{Im}(w)$")

### 2022-11-16 src/scripts/chapter_3/mag_limbdark_accuracy.py

@@ -20,15 +20,16 @@ from caustics.integrate import *
 
 import MulensModel as mm
 
-def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=1e-05):
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    e1 = 1/(1 + q)
     e2 = 1 - e1
-    x_cm = (e1 - e2)*a
-    bl = mm.BinaryLens(e2, e1, 2*a)
-    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
-
+    bl = mm.BinaryLens(e1, e2, s)
+    return bl.vbbl_magnification(
+        w0.real, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1
+    )
 
 @partial(jit, static_argnames=("npts_limb", "npts_ld", "limb_darkening"))
-def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
+def mag_binary(w_points, rho, s, q, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
     def body_fn(_, w):
         mag = mag_extended_source(
             w,
@@ -38,29 +39,27 @@ def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_dar
             limb_darkening=limb_darkening,
             npts_ld=npts_ld,
             u1=u1,
-            a=a,
-            e1=e1,
+            s=s,
+            q=q,
         )
         return 0, mag
-
     _, mags = lax.scan(body_fn, 0, w_points)
     return mags
 
 
 
 # Parameters
-a, e1, = (
-    0.45,
-    0.8,
-)
+s, q = 0.9, 0.2
+u1 = 0.7
 
 # 1000  points on caustic curve
-npts = 250
+npts = 25 
 critical_curves, caustic_curves = critical_and_caustic_curves(
-    npts=npts, nlenses=2, a=a, e1=e1
+    npts=npts, nlenses=2, s=s, q=q
 )
 caustic_curves = caustic_curves.reshape(-1)
 
+# Accuracy of integration
 acc_vbb = 1e-05
 npts_limb = 400
 
@@ -69,13 +68,13 @@ npts_limb = 400
 mags_vbb_list = []
 mags_list = []
 
[…]

### 2022-11-16 src/scripts/chapter_3/mag_uniform_accuracy.py

@@ -20,16 +20,17 @@ from caustics.integrate import *
 
 import MulensModel as mm
 
-def mag_vbb_binary(w0, rho, a, e1, u1=0., accuracy=1e-05):
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    e1 = 1/(1 + q)
     e2 = 1 - e1
-    x_cm = (e1 - e2)*a
-    bl = mm.BinaryLens(e2, e1, 2*a)
-    return bl.vbbl_magnification(w0.real - x_cm, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1)
-
+    bl = mm.BinaryLens(e1, e2, s)
+    return bl.vbbl_magnification(
+        w0.real, w0.imag, rho, accuracy=accuracy, u_limb_darkening=u1
+    )
 
 
 @partial(jit, static_argnames=("npts_limb", "npts_ld", "limb_darkening"))
-def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
+def mag_binary(w_points, rho, s, q, u1=0., npts_limb=300, npts_ld=100, limb_darkening=False):
     def body_fn(_, w):
         mag = mag_extended_source(
             w,
@@ -39,27 +40,25 @@ def mag_binary(w_points, rho, a, e1, u1=0., npts_limb=300, npts_ld=100, limb_dar
             limb_darkening=limb_darkening,
             npts_ld=npts_ld,
             u1=u1,
-            a=a,
-            e1=e1,
+            s=s,
+            q=q,
         )
         return 0, mag
-
     _, mags = lax.scan(body_fn, 0, w_points)
     return mags
 
 
+
 # Parameters
-a, e1 = 0.45, 0.8
-u1 = 0.7
+s, q = 0.9, 0.2
 
 # 1000  points on caustic curve
-npts = 25 
+npts = 250
 critical_curves, caustic_curves = critical_and_caustic_curves(
-    npts=npts, nlenses=2, a=a, e1=e1
+    npts=npts, nlenses=2, s=s, q=q
 )
 caustic_curves = caustic_curves.reshape(-1)
 
-# Accuracy of integration
 acc_vbb = 1e-05
 npts_limb = 400
 
@@ -68,7 +67,7 @@ npts_limb = 400
 mags_vbb_list = []
[…]

### 2022-11-16 src/scripts/chapter_4/fit_single_lens_model.py

@@ -1,3 +1,4 @@
+import paths
 import numpy as np
 from jax.config import config
 config.update("jax_enable_x64", True)
@@ -148,7 +149,7 @@ samples_mcmc_az_list = [
 samples_mcmc_az = az.concat(samples_mcmc_az_list, dim='chain')
 
 # Save to disk 
-az.to_netcdf(samples_mcmc_az, 'output/single_lens/samples_mcmc_az.nc')
+az.to_netcdf(samples_mcmc_az, paths.data/'output/single_lens/samples_mcmc_az.nc')
 
 # NESTED SAMPLING
 @jit
@@ -178,25 +179,24 @@ def prior_transform(u):
 
     return x
 
-#log_likelihood_fn = lambda x: numpyro.infer.log_likelihood(
-#    model,
-#    {'ln_t0': x[0], 'ln_tE': x[1], 'u0': x[2], 'piEE': x[3], 'piEN': x[4]},
-#    t=t, t0_est=t0_est, fobs=fobs,ferr=ferr, trajectory=trajectory
-#)['obs'].sum()
-#log_likelihood_fn = jit(log_likelihood_fn)
-#
-#log_likelihood_vectorized = lambda x: np.array(vmap(log_likelihood_fn)(x))
-#prior_transform_vectorized = lambda x: np.array(vmap(prior_transform)(x))
-#
-#param_names = ['ln_t0', 'ln_tE', 'u0', 'pi_EE', 'pi_EN']
-#sampler = ultranest.ReactiveNestedSampler(
-#    param_names,
-#    log_likelihood_vectorized,
-#    prior_transform_vectorized, 
-#    resume=True,
-#    vectorized=True,
-#    log_dir='output/single_lens/ultranest',
-#);
-#result = sampler.run()
-#
-#
+log_likelihood_fn = lambda x: numpyro.infer.log_likelihood(
+    model,
+    {'ln_t0': x[0], 'ln_tE': x[1], 'u0': x[2], 'piEE': x[3], 'piEN': x[4]},
+    t=t, t0_est=t0_est, fobs=fobs,ferr=ferr, trajectory=trajectory
+)['obs'].sum()
+log_likelihood_fn = jit(log_likelihood_fn)
+
+log_likelihood_vectorized = lambda x: np.array(vmap(log_likelihood_fn)(x))
+prior_transform_vectorized = lambda x: np.array(vmap(prior_transform)(x))
+
[…]

### 2022-11-16 src/scripts/chapter_4/single_lens_fit_plots.py

@@ -150,7 +150,7 @@ fig.savefig(paths.figures/"single_lens_lightcurve.pdf", bbox_inches="tight")
 
 
 # Load samples from disk with arviz
-samples_mcmc_az = az.from_netcdf(paths.scripts/"output/single_lens/samples_mcmc_az.nc")
+samples_mcmc_az = az.from_netcdf(paths.data/"output/single_lens/samples_mcmc_az.nc")
 samples_mcmc_az.posterior['piE'] = np.sqrt(samples_mcmc_az.posterior.piEE**2 + samples_mcmc_az.posterior.piEN**2)
 
 
@@ -364,13 +364,25 @@ samples_mcmc_reweighted_stacking = mixture_draws(samples_mcmc_az.posterior, weig
 
 params_names = ['ln_t0', 'ln_tE', 'u0', 'piEE', 'piEN']
 chains_equal_weighted = np.loadtxt(
-    paths.scripts/"output/single_lens/ultranest/chains/equal_weighted_post.txt", skiprows=1
+    paths.data/"output/single_lens/ultranest/chains/equal_weighted_post.txt", skiprows=1
 )
 nsamples = samples_mcmc_az.posterior.isel(chain=[0])['ln_tE'].shape[-1]
 idcs = np.random.randint(0, chains_equal_weighted.shape[0], nsamples)
 chains_equal_weighted = chains_equal_weighted[idcs, :]
 samples_ultranest = {k: chains_equal_weighted[:, i] for i, k in enumerate(params_names)}
 samples_ultranest['piE'] = np.sqrt(samples_ultranest['piEE']**2 + samples_ultranest['piEN']**2)
+
+# Print weights
+print("Pseudo-BMA weights:")
+print(az.compare(compare_dict_mcmc, method='pseudo-BMA')['weight'])
+print("\n")
+
+print("Pseudo-BMA+ weights:")
+print(az.compare(compare_dict_mcmc, method='BB-pseudo-BMA')['weight'])
+print("\n")
+
+print("Stacking weights:")
+print(az.compare(compare_dict_mcmc)['weight'])
 
 
 # Plot re-weighted samples
@@ -407,7 +419,7 @@ ax['D'].hist(
 # Stacking
 ax['C'].scatter(
     jnp.exp(samples_mcmc_reweighted_stacking['ln_tE']),
-    samples_mcmc_reweighted['piE'],
[…]

### 2022-11-16 src/scripts/chapter_4/single_lens_laplace_comparison.py

@@ -72,14 +72,15 @@ def model(t, t0_est, fobs, ferr, trajectory):
 
     numpyro.sample("obs", dist.Normal(fpred, ferr), obs=fobs)
 
-
 def initialization_model(t0_est):
-    t0 = numpyro.sample('t0', dist.Normal(t0_est, 1.))
+    t0 = numpyro.sample('t0', dist.Normal(t0_est, 10.))
     numpyro.deterministic('ln_t0', jnp.log(t0))
-    numpyro.sample('ln_tE', dist.Uniform(jnp.log(20.0), jnp.log(365.0)))
-    numpyro.sample('u0', dist.Normal(0., 0.4))
-    numpyro.sample('piEE', dist.Normal(0., 0.05))
-    numpyro.sample('piEN', dist.Normal(0., 0.05))
+    tE = numpyro.sample('tE', dist.Uniform(10., 400.))
+    numpyro.deterministic('ln_tE', jnp.log(tE))
+    numpyro.sample('u0', dist.Uniform(-1., 1.))
+    numpyro.sample('piEE', dist.Normal(0., 0.1))
+    numpyro.sample('piEN', dist.Normal(0., 0.1))
+
 
 def fit_map(key, t0_est, params_init):
     opt = optimize_numpyro_ext(model, start=params_init, num_steps=3, include_deterministics=False)
@@ -238,8 +239,8 @@ samples_init = prior_predictive(subkey, t0_est=t0_est)
 
 key, subkey = random.split(key)
 map_params_dict = fit_map_batched(subkey, samples_init)
+map_params_dict = {key: map_params_dict[key][::-1] for key in map_params_dict.keys()}
 params_names = list(map_params_dict.keys())
-
 
 # Get samples from Laplace approximation  
 samples_laplace_dict, samples_laplace_unconstrained_dict = vmap(get_laplace_posterior_samples)(map_params_dict)
@@ -255,8 +256,16 @@ log_weights_list, ll_pointwise_list =\
 compare_dict_laplace = {
     f'mode_{i+1}': compute_loo(log_weights_list[i], ll_pointwise_list[i]) for i in range(len(log_weights_list))
 }
-az.compare(compare_dict_laplace, method='BB-pseudo-BMA')
+elpd_data_list = [compute_loo(log_weights_list[i], ll_pointwise_list[i]) for i in range(len(log_weights_list))]
 
[…]

### 2022-11-16 src/scripts/chapter_5/binary_lens_likelihood.py

@@ -1,0 +1,371 @@
+import paths
+import numpy as np
+
+import os
+from astropy.table import Table
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from caustics.linalg import weighted_least_squares
+
+import VBBinaryLensing
+VBBL = VBBinaryLensing.VBBinaryLensing()
+
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    return VBBL.BinaryMag2(s, q, w0.real, w0.imag, rho)
+
+def ll_fn_vbb(params, pointwise=False):
+    ln_s, ln_q, ln_t0, ln_u0, ln_tE, alpha, ln_rho =(
+        params['ln_s'], params['ln_q'], params['ln_t0'], params['ln_u0'], params['ln_tE'], params['alpha'], params['ln_rho']
+    )
+    s, q, t0, u0, tE, rho = np.exp(ln_s), np.exp(ln_q), np.exp(ln_t0),np.exp(ln_u0), np.exp(ln_tE), np.exp(ln_rho)
+
+    w_x = (t - t0)/tE*np.cos(alpha) - u0*np.sin(alpha)
+    w_y = (t - t0)/tE*np.sin(alpha) + u0*np.cos(alpha)
+    w_points = w_x + 1j*w_y
+
+    A = np.array([mag_vbb_binary(w0, rho, s, q) for w0 in w_points])
+
+    M = np.stack([A - 1., np.ones_like(A)]).T
+    beta = weighted_least_squares(fobs, ferr, M)
+    fpred = (M @ beta).reshape(-1)
+
+    if pointwise:
+        return -0.5 * (fobs - fpred)**2 / ferr**2
+    else:
+        return -0.5 * np.sum((fobs - fpred)**2 / ferr**2)
+
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
[…]

### 2022-11-16 src/scripts/chapter_5/binary_lens_modes.py

@@ -1,0 +1,284 @@
+import paths
+import numpy as np
+
+import os
+from astropy.table import Table
+
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+from matplotlib.ticker import AutoMinorLocator
+
+from jax import random
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+
+from caustics import (
+    mag,
+    mag_point_source,
+    critical_and_caustic_curves,
+)
+from caustics.linalg import weighted_least_squares
+
+import numpyro
+import numpyro.distributions as dist
+from numpyro_ext.optim import optimize as optimize_numpyro_ext
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+lc = get_data(paths.data/"microlensing/blg-0039/")
+
+# Trim
+lc['HJD'] = lc['HJD'] - 2450000
+lc = lc[lc['HJD'] > 7400]
+fobs, ferr = magnitudes_to_fluxes(lc["mag"], lc["mag_err"])
+t = lc['HJD']
+t, fobs, ferr = jnp.array(t), jnp.array(fobs), jnp.array(ferr)
+t0_est = 7832
+
+def model(t, t0_est, fobs, ferr):
+    ln_s = numpyro.sample("ln_s", dist.Uniform(jnp.log(1e-02), jnp.log(3.)))
+    ln_q = numpyro.sample("ln_q", dist.Uniform(jnp.log(1e-05), jnp.log(1.)))
[…]

### 2022-11-16 src/scripts/chapter_5/binary_lens_ns_analysis.py

@@ -1,0 +1,211 @@
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+
+import arviz as az 
+
+param_names = ['ln_t0', 'ln_tE', 'ln_u0', 'ln_s', 'ln_q', 'alpha', 'ln_rho']
+
+def get_samples_from_ns(path):
+    chains_equal_weighted = np.loadtxt(path, skiprows=1)[-5000:, :]
+#    ll_vals_pointwise = np.stack([ll_fn_vbb(s, pointwise=True) for s in chains_equal_weighted])
+    samples_ns = {name: chains_equal_weighted[-5000:, i] for i, name in enumerate(param_names)}
+    samples_ns_az = az.from_dict(samples_ns)
+#    samples_ns_az.add_groups(
+#        log_likelihood=xr.DataArray(ll_vals_pointwise[None, :, :], dims=['chain', 'draw', 'y_dim_0'])
+#    )
+    return samples_ns_az
+
+samples_2k = get_samples_from_ns(paths.data/"output/binary_lens/nlive_2000/chains/equal_weighted_post.txt")
+samples_5k = get_samples_from_ns(paths.data/"output/binary_lens/nlive_5000/chains/equal_weighted_post.txt")
+samples_10k = get_samples_from_ns(paths.data/"output/binary_lens/nlive_10000/chains/equal_weighted_post.txt")
+
+for samples in (samples_2k, samples_5k, samples_10k):
+    samples.posterior['s'] = np.exp(samples.posterior['ln_s'])
+    samples.posterior['q'] = np.exp(samples.posterior['ln_q'])
+    samples.posterior['t0'] = np.exp(samples.posterior['ln_t0'])
+    samples.posterior['u0'] = np.exp(samples.posterior['ln_u0'])
+    samples.posterior['tE'] = np.exp(samples.posterior['ln_tE'])
+    samples.posterior['rho'] = np.exp(samples.posterior['ln_rho'])
+
+fig, ax = plt.subplots(3, 5, figsize=(16, 13), gridspec_kw={
[…]

### 2022-11-16 src/scripts/chapter_5/compute_likelihood_on_grid.py

@@ -1,0 +1,178 @@
+import paths
+import numpy as np
+
+import os
+from astropy.table import Table
+
+from caustics.linalg import weighted_least_squares
+
+import VBBinaryLensing
+VBBL = VBBinaryLensing.VBBinaryLensing()
+
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    return VBBL.BinaryMag2(s, q, w0.real, w0.imag, rho)
+
+def ll_fn_vbb(params, pointwise=False):
+    ln_s, ln_q, ln_t0, ln_u0, ln_tE, alpha, ln_rho =(
+        params['ln_s'], params['ln_q'], params['ln_t0'], params['ln_u0'], params['ln_tE'], params['alpha'], params['ln_rho']
+    )
+    s, q, t0, u0, tE, rho = np.exp(ln_s), np.exp(ln_q), np.exp(ln_t0),np.exp(ln_u0), np.exp(ln_tE), np.exp(ln_rho)
+
+    w_x = (t - t0)/tE*np.cos(alpha) - u0*np.sin(alpha)
+    w_y = (t - t0)/tE*np.sin(alpha) + u0*np.cos(alpha)
+    w_points = w_x + 1j*w_y
+
+    A = np.array([mag_vbb_binary(w0, rho, s, q) for w0 in w_points])
+
+    M = np.stack([A - 1., np.ones_like(A)]).T
+    beta = weighted_least_squares(fobs, ferr, M)
+    fpred = (M @ beta).reshape(-1)
+
+    if pointwise:
+        return -0.5 * (fobs - fpred)**2 / ferr**2
+    else:
+        return -0.5 * np.sum((fobs - fpred)**2 / ferr**2)
+
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
[…]

### 2022-11-16 src/scripts/chapter_5/compute_likelihood_on_grid_backup.py

@@ -1,0 +1,218 @@
+import paths
+import numpy as np
+
+import os
+from astropy.table import Table
+
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import jax.numpy as jnp
+from jax import jit, vmap, lax
+
+from caustics import  mag
+from caustics.linalg import weighted_least_squares
+
+
+import VBBinaryLensing
+VBBL = VBBinaryLensing.VBBinaryLensing()
+
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    return VBBL.BinaryMag2(s, q, w0.real, w0.imag, rho)
+
+def ll_fn_vbb(x, pointwise=False):
+    ln_t0, ln_tE, ln_u0, ln_s, ln_q, alpha, ln_rho = x
+    s, q, t0, u0, tE, rho = np.exp(ln_s), np.exp(ln_q), np.exp(ln_t0),np.exp(ln_u0), np.exp(ln_tE), np.exp(ln_rho)
+
+    w_x = (t - t0)/tE*np.cos(alpha) - u0*np.sin(alpha)
+    w_y = (t - t0)/tE*np.sin(alpha) + u0*np.cos(alpha)
+    w_points = w_x + 1j*w_y
+
+    A = np.array([mag_vbb_binary(w0, rho, s, q) for w0 in w_points])
+
+    M = np.stack([A - 1., np.ones_like(A)]).T
+    beta = weighted_least_squares(fobs, ferr, M)
+    fpred = (M @ beta).reshape(-1)
+
+    if pointwise:
+        return -0.5 * (fobs - fpred)**2 / ferr**2
+    else:
+        return -0.5 * np.sum((fobs - fpred)**2 / ferr**2)
+
+
+import numpyro
+import numpyro.distributions as dist
+numpyro.set_host_device_count(4)
+
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
[…]

### 2022-11-16 src/scripts/chapter_5/data_plot.py

@@ -1,0 +1,51 @@
+import paths
+import numpy as np
+import os
+from astropy.table import Table
+
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+
+lc = get_data(paths.data/"microlensing/blg-0039/")
+
+# Trim
+lc['HJD'] = lc['HJD'] - 2450000
+lc = lc[lc['HJD'] > 7400]
+fobs, ferr = magnitudes_to_fluxes(lc["mag"], lc["mag_err"])
+t = lc['HJD']
+t, fobs, ferr = np.array(t), np.array(fobs), np.array(ferr)
+#
+fig, ax = plt.subplots(1, 2, figsize=(12, 5),gridspec_kw={'width_ratios': [2, 1], 'wspace':0.1})
+
+for a in ax:
+    a.errorbar(lc['HJD'], lc['mag'], lc['mag_err'],  marker='o', color='k', linestyle='None', alpha=0.2) 
+    a.invert_yaxis()
+    a.grid(alpha=0.5)
+    a.set(xlabel='Time [HJD - 2450000 days]')
+    a.xaxis.set_minor_locator(AutoMinorLocator())
+    a.yaxis.set_minor_locator(AutoMinorLocator())
+
+    
+ax[0].set(ylabel='I magnitude')
+ax[0].set_title('OGLE-2016-BLG-0039', pad=20, horizontalalignment='left')
+ax[1].set_xlim(7815, 7850)
+ax[1].set_yticklabels([])
+
+fig.savefig(paths.figures/"blg-0039.pdf", bbox_inches='tight')

### 2022-11-16 src/scripts/chapter_5/fit_binary_lens_nested_sampling.py

@@ -1,0 +1,99 @@
+import paths
+import numpy as np
+import os
+from astropy.table import Table
+
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+import numpy as np
+
+from caustics.linalg import weighted_least_squares
+
+import ultranest
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+
+lc = get_data(paths.data/"microlensing/blg-0039/")
+
+# Trim
+lc['HJD'] = lc['HJD'] - 2450000
+lc = lc[lc['HJD'] > 7400]
+fobs, ferr = magnitudes_to_fluxes(lc["mag"], lc["mag_err"])
+t = lc['HJD']
+t, fobs, ferr = np.array(t), np.array(fobs), np.array(ferr)
+
+def prior_transform(u):
+    """
+    Transforms the uniform random variables `u ~ Unif[0., 1.)`
+    to the parameters of interest.
+    """
+    x = np.array(u)  # copy u
+
+    x[0] = np.log(7820.) + u[0] * (np.log(7860.) - np.log(7820.)) # ln_t0
+    x[1] = np.log(10) + u[1] * (np.log(365.) - np.log(10.)) # ln_tE
+    x[2] = np.log(0.001) + u[2] * (np.log(2.) - np.log(0.001)) # ln_u0
+    x[3] = np.log(1e-02) + u[3] * (np.log(3.) - np.log(1e-02)) # ln_s
+    x[4] = np.log(1e-05) + u[4] * (np.log(1.) - np.log(1e-05)) # ln_q
+    x[5] = -np.pi + u[5] * (np.pi - (-np.pi)) # alpha
+    x[6] = np.log(5e-04) + u[6] * (np.log(0.5) - np.log(5e-04)) # ln_rho
+
+    return x
[…]

### 2022-11-16 src/scripts/chapter_5/fit_emcee.py

@@ -1,0 +1,105 @@
+import numpy as np
+import os
+from astropy.table import Table
+
+import emcee
+from caustics.linalg import weighted_least_squares
+
+import arviz as az 
+os.environ["OMP_NUM_THREADS"] = "1"
+
+def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
+    flux = 10. ** (0.4 * (zeropoint - mag))
+    err_flux = mag_err * flux * np.log(10.) * 0.4
+    return flux, err_flux
+
+def get_data(path):
+    t = Table.read(os.path.join(path, "phot.dat"), format="ascii")
+
+    # Remove additional columns
+    t.columns[0].name = "HJD"
+    t.columns[1].name = "mag"
+    t.columns[2].name = "mag_err"
+    t.keep_columns(("HJD", "mag", "mag_err"))
+    return t
+
+
+lc = get_data("../../../../../data.nosync/OGLE_ews/2017/blg-0039/")
+
+# Trim
+lc['HJD'] = lc['HJD'] - 2450000
+lc = lc[lc['HJD'] > 7400]
+fobs, ferr = magnitudes_to_fluxes(lc["mag"], lc["mag_err"])
+t = lc['HJD']
+t, fobs, ferr = np.array(t), np.array(fobs), np.array(ferr)
+
+import VBBinaryLensing
+VBBL = VBBinaryLensing.VBBinaryLensing()
+
+def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
+    return VBBL.BinaryMag2(s, q, w0.real, w0.imag, rho)
+
+def log_likelihood(x):
+    ln_t0, ln_tE, ln_u0, ln_s, ln_q, alpha, ln_rho = x
+    s, q, t0, u0, tE, rho = np.exp(ln_s), np.exp(ln_q), np.exp(ln_t0),np.exp(ln_u0), np.exp(ln_tE), np.exp(ln_rho)
+
+    w_x = (t - t0)/tE*np.cos(alpha) - u0*np.sin(alpha)
+    w_y = (t - t0)/tE*np.sin(alpha) + u0*np.cos(alpha)
+    w_points = w_x + 1j*w_y
+
+    A = np.array([mag_vbb_binary(w0, rho, s, q) for w0 in w_points])
+
+    M = np.stack([A - 1., np.ones_like(A)]).T
+    beta = weighted_least_squares(fobs, ferr, M)
+    fpred = (M @ beta).reshape(-1)
+
[…]

### 2022-11-16 src/scripts/chapter_6/information_content.py

@@ -1,0 +1,387 @@
+import paths
+import numpy as np
+import starry
+import astropy.units as u
+from astropy.time import Time
+import sys
+import os
+import scipy
+from scipy.linalg import cho_factor, cho_solve
+
+from matplotlib import pyplot as plt
+from matplotlib.legend_handler import HandlerTuple
+
+from io_mapping_utils import get_body_ephemeris, get_body_vectors
+
+np.random.seed(42)
+starry.config.lazy = False
+
+# Get 10 1hr observations of Io in eclipse/sunlight with 1min cadence
+start = Time("2009-01-01", format="isot")
+stop = Time("2010-01-05", format="isot")
+start_times = np.random.uniform(start.mjd, stop.mjd, 100)
+
+eph_phase_em = []
+
+# This takes a while
+nlightcurves = 100
+
+# Phase curves
+for i in range(nlightcurves):
+    t = start_times[i]
+    t_end = t + 2
+    npts = int((t_end - t) * 24)  # 1 hour cadence
+    times = Time(np.linspace(t, t_end, npts), format="mjd")
+
+    eph = get_body_ephemeris(
+        times, body_id="501", step="12h", return_orientation=True
+    )
+
+    # Get phase curves in eclipse
+    mask_ecl = np.all([eph["ecl_tot"], ~eph["occ_umbra"]], axis=0)
+    if mask_ecl.sum() > 0.0:
+        eph_phase_em.append(eph[mask_ecl])
+
+
+# Occultations by Jupiter
+eph_io_occ_em = []
+eph_jup_occ_em = []
+
+for i in range(nlightcurves):
+    t = start_times[i]
+    t_end = t + 2.0
+    npts = int((t_end - t) * 24 * 60 * 15)  # 4sec cadence
+    times = Time(np.linspace(t, t_end, npts), format="mjd")
+
+    eph_io = get_body_ephemeris(
+        times, body_id="501", step="30m", return_orientation=True
+    )
+
+    eph_jup = get_body_ephemeris(
+        times, body_id="599", step="30m", return_orientation=False
+    )
+
+    # Get occultations in eclipse
+    mask = eph_io["occ_umbra"]
+
+    if mask.sum() > 0.0:
+        eph_io_occ_em.append(eph_io[mask])
+        eph_jup_occ_em.append(eph_jup[mask])
+
+
[…]

### 2022-11-16 src/scripts/chapter_6/io_mapping_utils.py

@@ -1,0 +1,635 @@
+import numpy as np
+from astropy import units as u
+from astropy.time import Time
+from astropy.timeseries import TimeSeries
+from astroquery.jplhorizons import Horizons
+from scipy.interpolate import interp1d
+
+import starry
+
+starry.config.lazy = False
+
+
+def get_body_ephemeris(
+    times, body_id="501", step="1m", return_orientation=True
+):
+    """
+    Given a NAIF code of a Solar System body, this function computes the
+    position and orientation of the body in equatorial coordinates at requested
+    times using ephemeris from JPL Horizons. The relevant JPL Horizons data is
+    the following:
+
+    'RA', 'DEC'
+    ###########
+    Position on the plane of the sky.
+
+    'ang_width'
+    ###########
+    The angle subtended by the disk of the target seen by the observer, if
+    it was fully illuminated. The target diameter is taken to be the IAU2009
+    equatorial diameter. Oblateness aspect is not currently included.
+    Units: ARCSECONDS
+
+    'sat_vis'
+    #########
+    The angle between the center of a non-lunar target body and the center
+    of the primary body it revolves around, as seen by the observer.
+    Units: ARCSECONDS
+
+    Non-lunar natural satellite visibility codes (limb-to-limb):
+
+    /t = Transitting primary body disk, /O = Occulted by primary body disk,
+    /p = Partial umbral eclipse,        /P = Occulted partial umbral eclipse,
+    /u = Total umbral eclipse,          /U = Occulted total umbral eclipse,
+    /- = Target is the primary body,    /* = None of above ("free and clear")
+
+    ... the radius of major bodies is taken to be the equatorial value (max)
+    defined by the IAU2009 system. Atmospheric effects and oblateness aspect
+    are not currently considered in these computations. Light-time is included.
+
+    'r'
+    ###
+    Heliocentric range ("r", light-time compensated) and range-rate ("rdot")
+    of the target point at the instant light later seen by the observer at
+    print-time would have left the target (at the instant print-time minus
+    down-leg light-time); the Sun-to-target distance traveled by a ray of
+    light emanating from the center of the Sun that reaches the target at some
+    instant and is recordable by the observer one down-leg light-time later at
+    print-time. "rdot" is a projection of the velocity vector along this ray,
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_dataset.py

@@ -1,0 +1,57 @@
+import paths
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+import pickle as pkl
+import astropy.units as u
+
+np.random.seed(42)
+
+# The 1998 pair of light curves
+with open(paths.data/"output/mapping_io/irtf_processed/lc_1998-08-27.pkl", "rb") as handle:
+    lc1_in = pkl.load(handle)
+
+with open(paths.data/"output/mapping_io/irtf_processed/lc_1998-11-29.pkl", "rb") as handle:
+    lc1_eg = pkl.load(handle)
+
+
+## The 2017 pair of light curves
+# with open("../../data/irtf_processed/lc_2017-03-31.pkl", "rb") as handle:
+#    lc2_in = pkl.load(handle)
+#
+# with open("../../data/irtf_processed/lc_2017-05-11.pkl", "rb") as handle:
+#    lc2_eg = pkl.load(handle)
+
+fig, ax = plt.subplots(1, 2, figsize=(9, 4), sharey=True)
+
+ax[0].plot(
+    (lc1_in.time.mjd - lc1_in.time.mjd[0]) * u.d.to(u.min),
+    lc1_in["flux"],
+    "ko",
+    alpha=0.4,
+)
+ax[1].plot(
+    (lc1_eg.time.mjd - lc1_eg.time.mjd[0]) * u.d.to(u.min),
+    lc1_eg["flux"],
+    "ko",
+    alpha=0.4,
+)
+
+ax[0].set(xticks=np.arange(0, 5, 1), yticks=np.arange(0, 60, 10))
+ax[1].set(xticks=np.arange(0, 6, 1))
+
+ax[0].set_title(lc1_in.time.datetime[0].strftime("%Y-%m-%d %H:%M"))
+ax[1].set_title(lc1_eg.time.datetime[0].strftime("%Y-%m-%d %H:%M"))
+
+ax[0].set(ylabel="Spectral intensity [ GW/um/sr]")
+
+
+for a in ax:
+    a.xaxis.set_minor_locator(AutoMinorLocator())
+    a.yaxis.set_minor_locator(AutoMinorLocator())
+
+for a in ax.flatten():
+    a.grid(alpha=0.5)
+
+fig.text(0.5, -0.02, "Duration [minutes]", ha="center", va="center")
+fig.savefig(paths.figures/"irtf_1998_occultations.pdf", bbox_inches="tight", dpi=500)

### 2022-11-16 src/scripts/chapter_6/irtf_errorbar_posteriors.py

@@ -1,0 +1,39 @@
+import numpy as np
+from matplotlib import pyplot as plt
+import pickle as pkl
+
+import seaborn as sns
+
+with open("scripts/irtf_1998_samples.pkl", "rb") as handle:
+    samples = pkl.load(handle)
+
+fig, ax = plt.subplots(1, 2, figsize=(12, 4), sharex=True)
+fig.subplots_adjust(hspace=0.3)
+
+for i in range(len(samples["f_err_in_mod"][0])):
+    sns.distplot(
+        samples["f_err_in_mod"][:, i],
+        hist=False,
+        norm_hist=True,
+        kde_kws={"alpha": 0.7, "lw": 1},
+        ax=ax[0],
+    )
+for i in range(len(samples["f_err_in_mod"][0])):
+    sns.distplot(
+        samples["f_err_eg_mod"][:, i],
+        hist=False,
+        norm_hist=True,
+        kde_kws={"alpha": 0.7, "lw": 1},
+        ax=ax[1],
+    )
+
+ax[0].set_xlabel("Ingress errorbars [GW/um/sr]")
+ax[1].set_xlabel("Egress errorbars [GW/um/sr]")
+fig.text(0.07, 0.5, "Posterior probability", va="center", rotation="vertical")
+
+for a in ax:
+    a.grid()
+    a.set_xlim(-0.1, 2.0)
+    a.set_ylabel("")
+
+fig.savefig("irtf_1998_errorbars.pdf", bbox_inches="tight", dpi=400)

### 2022-11-16 src/scripts/chapter_6/irtf_fit.py

@@ -1,0 +1,314 @@
+import numpy as np
+import pickle as pkl
+
+import starry
+import jax.numpy as jnp
+from jax import random
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+
+import celerite2.jax
+from celerite2.jax import terms as jax_terms
+from celerite2 import terms, GaussianProcess
+from exoplanet.distributions import estimate_inverse_gamma_parameters
+
+from volcano.utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+numpyro.enable_x64()
+
+
+def fit_model(ydeg_inf, lc_in, lc_eg):
+    # Compute ephemeris
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
+    eph_jup_eg = eph_list_jup[1]
+
+    t_in = (lc_in.time.mjd - lc_in.time.mjd[0]) * 24 * 60
+    t_eg = (lc_eg.time.mjd - lc_eg.time.mjd[0]) * 24 * 60
+
+    f_obs_in = lc_in["flux"].value
+    f_err_in = lc_in["flux_err"].value
+    f_obs_eg = lc_eg["flux"].value
+    f_err_eg = lc_eg["flux_err"].value
+
+    f_obs = np.concatenate([f_obs_in, f_obs_eg])
+    f_err = np.concatenate([f_err_in, f_err_eg])
+
+    (xo_in, yo_in, ro_in, occ_lat_in,) = get_occultor_position_and_radius(
+        eph_io_in,
+        eph_jup_in,
+        occultor_is_jupiter=True,
+        rotate=True,
+        return_occ_lat=True,
+        method="",
+    )
+    (xo_eg, yo_eg, ro_eg, occ_lat_eg,) = get_occultor_position_and_radius(
+        eph_io_eg,
+        eph_jup_eg,
+        occultor_is_jupiter=True,
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_fit_param_err.py

@@ -1,0 +1,314 @@
+import numpy as np
+import pickle as pkl
+
+import starry
+import jax.numpy as jnp
+from jax import random
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+
+import celerite2.jax
+from celerite2.jax import terms as jax_terms
+from celerite2 import terms, GaussianProcess
+from exoplanet.distributions import estimate_inverse_gamma_parameters
+
+from volcano.utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+numpyro.enable_x64()
+
+
+def fit_model(ydeg_inf, lc_in, lc_eg):
+    # Compute ephemeris
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
+    eph_jup_eg = eph_list_jup[1]
+
+    t_in = (lc_in.time.mjd - lc_in.time.mjd[0]) * 24 * 60
+    t_eg = (lc_eg.time.mjd - lc_eg.time.mjd[0]) * 24 * 60
+
+    f_obs_in = lc_in["flux"].value
+    f_err_in = lc_in["flux_err"].value
+    f_obs_eg = lc_eg["flux"].value
+    f_err_eg = lc_eg["flux_err"].value
+
+    f_obs = np.concatenate([f_obs_in, f_obs_eg])
+    f_err = np.concatenate([f_err_in, f_err_eg])
+
+    (xo_in, yo_in, ro_in, occ_lat_in,) = get_occultor_position_and_radius(
+        eph_io_in,
+        eph_jup_in,
+        occultor_is_jupiter=True,
+        rotate=True,
+        return_occ_lat=True,
+        method="",
+    )
+    (xo_eg, yo_eg, ro_eg, occ_lat_eg,) = get_occultor_position_and_radius(
+        eph_io_eg,
+        eph_jup_eg,
+        occultor_is_jupiter=True,
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_ingress_egress_fit_classic_horseshoe.py

@@ -1,0 +1,304 @@
+import numpy as np
+import pickle as pkl
+
+import starry
+import jax.numpy as jnp
+from jax import random, jit, vmap, lax
+import numpyro
+import numpyro.distributions as dist
+from numpyro.diagnostics import print_summary
+from numpyro.infer import *
+
+import celerite2.jax
+from celerite2.jax import terms as jax_terms
+from celerite2 import terms, GaussianProcess
+from exoplanet.distributions import estimate_inverse_gamma_parameters
+
+from astropy import units as u
+from astropy.table import Table
+from astropy.timeseries import TimeSeries
+
+from volcano.utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+numpyro.enable_x64()
+
+
+def fit_model(ydeg_inf, lc_in, lc_eg):
+    # Compute ephemeris
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
+    eph_jup_eg = eph_list_jup[1]
+
+    t_in = (lc_in.time.mjd - lc_in.time.mjd[0]) * 24 * 60
+    t_eg = (lc_eg.time.mjd - lc_eg.time.mjd[0]) * 24 * 60
+
+    f_obs_in = lc_in["flux"].value
+    f_err_in = lc_in["flux_err"].value
+    f_obs_eg = lc_eg["flux"].value
+    f_err_eg = lc_eg["flux_err"].value
+
+    f_obs = np.concatenate([f_obs_in, f_obs_eg])
+    f_err = np.concatenate([f_err_in, f_err_eg])
+
+    def get_pos_rot(eph_io, eph_jup, method=""):
+        """
+        Get sky position of Io relative to Jupiter rotated such that the
+        obliquity of Jupiter is zero.
+        """
+        # Get occultor position
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_ingress_egress_main_plot_no_GP.py

@@ -1,0 +1,409 @@
+import numpy as np
+from matplotlib import pyplot as plt
+import pickle as pkl
+
+import starry
+import celerite2.jax
+from celerite2.jax import terms as jax_terms
+from celerite2 import terms, GaussianProcess
+from exoplanet.distributions import estimate_inverse_gamma_parameters
+
+from matplotlib import colors
+import matplotlib.gridspec as gridspec
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+from matplotlib.lines import Line2D
+import seaborn as sns
+
+from volcano.utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def get_pos_rot(eph_io, eph_jup, method=""):
+    # Get occultor position
+    obl = eph_io["obl"]
+    inc = np.mean(eph_jup["inc"])
+    theta = np.array(eph_io["theta"])
+
+    xo_unrot, yo_unrot, zo, ro = get_occultor_position_and_radius(
+        eph_io, eph_jup, occultor_is_jupiter=True, method=method
+    )
+
+    # Rotate to coordinate system where the obliquity of Io is 0
+    theta_rot = -obl.to(u.rad)
+    xo_rot, yo_rot = rotate_vectors(xo_unrot, yo_unrot, theta_rot)
+
+    return xo_rot, yo_rot, ro
+
+
+def make_plots(
+    lc_in,
+    lc_eg,
+    samples,
+    yticks,
+    ylim,
+    xticks_in,
+    xticks_eg,
+    res_yticks,
+    res_ylim,
+    cmap_norm=colors.Normalize(vmin=0.0, vmax=1500),
+):
+    #  Compute epheremis
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_main_plots.py

@@ -1,0 +1,576 @@
+import paths
+import numpy as np
+from matplotlib import pyplot as plt
+import pickle as pkl
+
+import starry
+import celerite2.jax
+from celerite2 import terms, GaussianProcess
+
+from matplotlib import colors
+import matplotlib.gridspec as gridspec
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+from matplotlib.lines import Line2D
+import matplotlib.cm as cm
+import seaborn as sns
+
+from io_mapping_utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def make_plots(
+    lc_in,
+    lc_eg,
+    samples,
+    yticks,
+    ylim,
+    xticks_in,
+    xticks_eg,
+    res_yticks,
+    res_ylim,
+    gp=True,
+    cmap_norm=colors.Normalize(vmin=0.0, vmax=1500),
+):
+    #  Compute epheremis
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
+    eph_jup_eg = eph_list_jup[1]
+
+    t_in = (lc_in.time.mjd - lc_in.time.mjd[0]) * 24 * 60
+    t_eg = (lc_eg.time.mjd - lc_eg.time.mjd[0]) * 24 * 60
+
+    f_obs_in = lc_in["flux"].value
+    f_err_in = lc_in["flux_err"].value
+    f_obs_eg = lc_eg["flux"].value
+    f_err_eg = lc_eg["flux_err"].value
+
+#    f_obs = np.concatenate([f_obs_in, f_obs_eg])
+#    f_err = np.concatenate([f_err_in, f_err_eg])
+
+    xo_in, yo_in, ro_in = get_occultor_position_and_radius(
+        eph_io_in, eph_jup_in, occultor_is_jupiter=True
+    )
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_main_plots_param_error.py

@@ -1,0 +1,564 @@
+import numpy as np
+from matplotlib import pyplot as plt
+import pickle as pkl
+
+import starry
+import celerite2.jax
+from celerite2.jax import terms as jax_terms
+from celerite2 import terms, GaussianProcess
+from exoplanet.distributions import estimate_inverse_gamma_parameters
+
+from matplotlib import colors
+import matplotlib.gridspec as gridspec
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+from matplotlib.lines import Line2D
+import seaborn as sns
+
+from volcano.utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def make_plots(
+    lc_in,
+    lc_eg,
+    samples,
+    yticks,
+    ylim,
+    xticks_in,
+    xticks_eg,
+    res_yticks,
+    res_ylim,
+    gp=True,
+    cmap_norm=colors.Normalize(vmin=0.0, vmax=1500),
+):
+    #  Compute epheremis
+    eph_list_io = []
+    eph_list_jup = []
+
+    for lc in (lc_in, lc_eg):
+        times = lc.time
+
+        eph_io = get_body_ephemeris(
+            times, body_id="501", step="1m", return_orientation=True
+        )
+        eph_jup = get_body_ephemeris(
+            times, body_id="599", step="1m", return_orientation=True
+        )
+
+        eph_list_io.append(eph_io)
+        eph_list_jup.append(eph_jup)
+
+    eph_io_in = eph_list_io[0]
+    eph_jup_in = eph_list_jup[0]
+    eph_io_eg = eph_list_io[1]
+    eph_jup_eg = eph_list_jup[1]
+
+    t_in = (lc_in.time.mjd - lc_in.time.mjd[0]) * 24 * 60
+    t_eg = (lc_eg.time.mjd - lc_eg.time.mjd[0]) * 24 * 60
+
+    f_obs_in = lc_in["flux"].value
+    f_err_in = lc_in["flux_err"].value
+    f_obs_eg = lc_eg["flux"].value
+    f_err_eg = lc_eg["flux_err"].value
+
+    f_obs = np.concatenate([f_obs_in, f_obs_eg])
+    f_err = np.concatenate([f_err_in, f_err_eg])
+
+    xo_in, yo_in, ro_in = get_occultor_position_and_radius(
+        eph_io_in, eph_jup_in, occultor_is_jupiter=True
[…]

### 2022-11-16 src/scripts/chapter_6/irtf_spots_plots.py

@@ -1,0 +1,343 @@
+import paths
+import numpy as np
+from matplotlib import pyplot as plt
+from PIL import Image
+import pickle as pkl
+
+import starry
+
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+from mpl_toolkits.axes_grid1.anchored_artists import AnchoredSizeBar
+import matplotlib.font_manager as fm
+import matplotlib.ticker as ticker
+
+from scipy.integrate import dblquad
+
+from io_mapping_utils import get_median_map
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def integrand(theta, phi, map):
+    lat = theta * 180 / np.pi - 90
+    lon = phi * 180 / np.pi
+
+    return map.intensity(lat=lat, lon=lon) * np.sin(theta)
+
+
+def get_emitted_power(map, lat_center, lon_center, delta):
+    # Rotate so the spot is at the North pole
+    map.rotate([0, 1, 0], -lon_center)
+    map.rotate([1, 0, 0], -(90 - lat_center))
+
+    phi_min = -180 * np.pi / 180
+    phi_max = 180 * np.pi / 180
+    theta_min = (90 - delta) * np.pi / 180 + np.pi / 2
+    theta_max = 90 * np.pi / 180 + np.pi / 2
+
+    res, _ = dblquad(
+        integrand,
+        phi_min,
+        phi_max,
+        lambda x: theta_min,
+        lambda x: theta_max,
+        epsabs=1e-4,
+        epsrel=1e-4,
+        args=(map,),
+    )
+
+    # Rotate back
+    map.rotate([1, 0, 0], (90 - lat_center))
+    map.rotate([0, 1, 0], lon_center)
+
+    return res
+
+
+def find_centroid(lats, lons, img, bounds):
+    mask_lat = np.logical_and(lats > bounds[0][0], lats < bounds[0][1])
+    mask_lon = np.logical_and(lons > bounds[1][0], lons < bounds[1][1])
+
+    img_bounded = img[mask_lat][:, mask_lon]
+
+    I_min = np.nanpercentile(img_bounded, [90])
+    mask = img_bounded > I_min
+
+    lon_grid, lat_grid = np.meshgrid(lons[mask_lon], lats[mask_lat])
+    return np.mean(lon_grid[mask]), np.mean(lat_grid[mask])
+
+
+def get_contour_levels(lats, lons, img, bounds, percentiles):
[…]

### 2022-11-16 src/scripts/chapter_6/paths.py

@@ -1,0 +1,25 @@
+"""
+Exposes common paths useful for manipulating datasets and generating figures.
+"""
+from pathlib import Path
+
+# Absolute path to the top level of the repository
+root = Path(__file__).resolve().parents[3].absolute()
+
+# Absolute path to the `src` folder
+src = root / "src"
+
+# Absolute path to the `src/data` folder (contains datasets)
+data = src / "data"
+
+# Absolute path to the `src/static` folder (contains static images)
+static = src / "static"
+
+# Absolute path to the `src/scripts` folder (contains figure/pipeline scripts)
+scripts = src / "scripts"
+
+# Absolute path to the `src/tex` folder (contains the manuscript)
+tex = src / "tex"
+
+# Absolute path to the `src/tex/figures` folder (contains figure output)
+figures = tex / "figures"

### 2022-11-16 src/scripts/chapter_6/pixels_to_spherical_harmonics.py

@@ -1,0 +1,217 @@
+import paths
+import numpy as np
+import theano.tensor as tt
+import pymc3 as pm
+import starry
+from starry._plotting import (
+    get_moll_latitude_lines,
+    get_moll_longitude_lines,
+)
+
+from matplotlib import pyplot as plt
+from matplotlib import colors
+from scipy import optimize
+
+np.random.seed(42)
+starry.config.lazy = True
+
+ydeg = 20
+map = starry.Map(ydeg)
+lat, lon, Y2P, P2Y, Dx, Dy = map.get_pixel_transforms(oversample=4)
+npix = Y2P.shape[0]
+
+std_p = 1.62
+
+with pm.Model() as model:
+    p = pm.Exponential("p", 1 / std_p, shape=(npix,))
+    x = tt.dot(P2Y, p)
+    pm.Deterministic("x", x)
+    p_back = tt.dot(Y2P, x)
+    pm.Deterministic("p_back", p_back)
+
+    trace_pp = pm.sample_prior_predictive(10)
+
+
+# Convert lat, lon to x,y coordinates in Mollewiede projection
+def lon_lat_to_mollweide(lon, lat):
+    lat *= np.pi / 180
+    lon *= np.pi / 180
+
+    f = lambda x: 2 * x + np.sin(2 * x) - np.pi * np.sin(lat)
+    theta = optimize.newton(f, 0.3)
+
+    x = 2 * np.sqrt(2) / np.pi * lon * np.cos(theta)
+    y = np.sqrt(2) * np.sin(theta)
+
+    return x, y
+
+
+x_mol = np.zeros(npix)
+y_mol = np.zeros(npix)
+
+for idx, (lo, la) in enumerate(zip(lon, lat)):
+    x_, y_ = lon_lat_to_mollweide(lo, la)
+    x_mol[idx] = x_
+    y_mol[idx] = y_
+
+
+def plot_grid_lines(ax, alpha=0.6):
+    """
+    Code from https://github.com/rodluger/starry/blob/0546b4e445f6570b9a1cf6e33068e01a96ecf20f/starry/maps.py.
+    """
+    ax.axis("off")
+
+    borders = []
+    x = np.linspace(-2 * np.sqrt(2), 2 * np.sqrt(2), 10000)
+    y = np.sqrt(2) * np.sqrt(1 - (x / (2 * np.sqrt(2))) ** 2)
[…]

### 2022-11-16 src/scripts/chapter_6/pixels_vs_harmonics.py

@@ -1,0 +1,474 @@
+import paths
+import numpy as np
+import os
+import sys
+
+from matplotlib import pyplot as plt
+from matplotlib import colors
+import matplotlib.gridspec as gridspec
+from matplotlib.ticker import FormatStrFormatter, AutoMinorLocator
+import matplotlib.cm as cm
+
+
+import starry
+import jax.numpy as jnp
+from jax import random, jit, vmap, lax
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+
+from io_mapping_utils import get_smoothing_filter
+
+from scipy import optimize
+
+numpyro.enable_x64()
+numpyro.set_host_device_count(4)
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def get_median_map(
+    ydeg_inf,
+    samples,
+    projection="Mollweide",
+    inc=90,
+    theta=0.0,
+    nsamples=100,
+    resol=250,
+):
+    imgs = []
+    map = starry.Map(ydeg=ydeg_inf)
+    map.inc = inc
+
+    for n in np.random.randint(0, len(samples), nsamples):
+        x = samples[n]
+        map.amp = x[0]
+        map[1:, :] = x[1:] / map.amp
+
+        if projection == "Mollweide":
+            im = map.render(projection="Mollweide", res=resol)
+        else:
+            im = map.render(theta=theta, res=resol)
+        imgs.append(im)
+
+    return np.median(imgs, axis=0)
+
+
+npts = 150
+xo_eg = np.linspace(37.15, 39.43, npts)
+yo_eg = np.linspace(-8.284, -8.27, npts)
+
+xo_in = np.linspace(-39.5, -37.3, npts)
+yo_in = np.linspace(-7.714, -7.726, npts)
+
+xo_com = np.concatenate([xo_in, xo_eg])
+yo_com = np.concatenate([yo_in, yo_eg])
+
+theta_in = 350.0
+theta_eg = 10.0
+ro = 39.1
+
+ydeg_true = 30
+map_true = starry.Map(ydeg_true)
+spot_ang_dim = 5 * np.pi / 180
[…]

### 2022-11-16 src/scripts/chapter_6/process_irtf_lightcurves.py

@@ -1,0 +1,135 @@
+import paths
+import numpy as np
+import os
+import xml.etree.ElementTree as ET
+import sys
+import pickle as pkl
+
+from astropy import units as u
+from astropy.time import Time
+from astropy.timeseries import TimeSeries
+from astropy.table import vstack
+from astropy.table import Table
+from astropy.stats import sigma_clip, mad_std
+
+from scipy.signal import savgol_filter
+
+np.random.seed(42)
+
+
+def get_data_from_xml(path):
+    tree = ET.parse(path)
+    root = tree.getroot()
+
+    namespaces = {"pds": "http://pds.nasa.gov/pds4/pds/v1"}
+
+    description = root.findall(".//pds:description", namespaces=namespaces)[
+        0
+    ].text
+    start_date = root.findall(".//pds:start_date_time", namespaces=namespaces)[
+        0
+    ].text
+    end_date = root.findall(".//pds:stop_date_time", namespaces=namespaces)[
+        0
+    ].text
+    instrument = root.findall(
+        ".//pds:Observing_System_Component", namespaces=namespaces
+    )[0][0].text
+    facility = root.findall(
+        ".//pds:Observing_System_Component", namespaces=namespaces
+    )[1][0].text
+
+    dic = {
+        "description": description,
+        "start_date": start_date,
+        "end_date": end_date,
+        "instrument": instrument,
+        "facility": facility,
+    }
+    return dic
+
+
+directory = paths.data/"mapping_io/irtf_raw"
+
+lcs = []
+i = 0
+for filename in os.listdir(directory):
+    if filename.endswith(".xml"):
+        metadata = get_data_from_xml(os.path.join(directory, filename))
+        data = np.genfromtxt(
+            os.path.join(directory, filename[:-4] + "tab.txt")
+        )
+
+        # Remove the zeros at the beginning of some files
+        mask = data[:, 0] == 0
+        data = data[~mask]
+
+        start_time = Time(
+            metadata["start_date"][:-1], format="isot", scale="utc"
+        ).to_value("mjd", "long")
+        times = start_time + data[:, 0] / 24.0
+        times = Time(times, format="mjd")
+
+        # Convert to astropy.TimeSeries
+        ts = TimeSeries(time=times)
[…]

### 2022-11-16 src/scripts/chapter_6/simulated_ingress_egress.py

@@ -1,0 +1,601 @@
+import paths
+import numpy as np
+from matplotlib import pyplot as plt
+import matplotlib.cm as cm
+
+import starry
+from scipy import optimize
+
+import jax.numpy as jnp
+from jax import random
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+
+from matplotlib import colors
+from matplotlib.ticker import  AutoMinorLocator
+
+from io_mapping_utils import *
+
+np.random.seed(42)
+starry.config.lazy = False
+numpyro.enable_x64()
+numpyro.set_host_device_count(4)
+
+
+xo_eg = np.linspace(37.15, 39.43, 150)
+yo_eg = np.linspace(-8.284, -8.27, 150)
+
+xo_in = np.linspace(-39.5, -37.3, 150)
+yo_in = np.linspace(-7.714, -7.726, 150)
+
+xo_com = np.concatenate([xo_in, xo_eg])
+yo_com = np.concatenate([yo_in, yo_eg])
+
+theta_in = 350.0
+theta_eg = 10.0
+ro = 39.1
+
+ydeg_true = 30
+map_true = starry.Map(ydeg_true)
+spot_ang_dim = 10 * np.pi / 180
+spot_sigma = 1 - np.cos(spot_ang_dim / 2)
+map_true.add_spot(
+    amp=1.0, sigma=spot_sigma, lat=13.0, lon=51.0, relative=False
+)
+map_true.add_spot(
+    amp=0.3, sigma=spot_sigma, lat=-15.0, lon=-40.0, relative=False
+)
+map_true.amp = 20
+map_true_img = map_true.render(res=250, projection="Mollweide")
+
+# Smooth the true map
+sigma_s = 2 / ydeg_true
+S_true = get_smoothing_filter(ydeg_true, sigma_s)
+x = map_true.amp * map_true.y
+xsmooth = (S_true @ x[:, None]).reshape(-1)
+map_true[:, :] = xsmooth / xsmooth[0]
+map_true.amp = xsmooth[0]
+
+# Generate mock ingress and egress light curves
+f_true_in = map_true.flux(ro=ro, xo=xo_in, yo=yo_in, theta=theta_in)
+f_true_eg = map_true.flux(ro=ro, xo=xo_eg, yo=yo_eg, theta=theta_eg)
+
+# S/N = 50
+f_err_in_50 = np.max(np.concatenate([f_true_in, f_true_eg])) / 50
+f_err_eg_50 = f_err_in_50
[…]

### 2022-11-16 src/scripts/chapter_6/smoothing_kernel.py

@@ -1,0 +1,85 @@
+import paths
+import numpy as np
+import starry
+from matplotlib import pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from  io_mapping_utils import get_smoothing_filter
+
+np.random.seed(42)
+starry.config.lazy = False
+
+# Dummy featureless map
+map0 = starry.Map(ydeg=1)
+map0.show()
+
+# Compute the intensity along the equator
+# Remove the baseline intensity and normalize it
+lon = np.linspace(-90, 90, 300)
+baseline = 1.0 / np.pi
+
+# Compute the intensity of a normalized gaussian
+# in cos(longitude) with the same standard deviation
+coslon = np.cos(lon * np.pi / 180)
+spot_ang_dim = 5 * np.pi / 180
+spot_sigma = 1 - np.cos(spot_ang_dim / 2)
+I_gaussian = np.exp(-((coslon - 1) ** 2) / (2 * spot_sigma ** 2))
+
+degs = [10, 15, 20]
+
+Is = np.zeros((3, len(degs), 300))
+smoothing_sigmas = np.array([0.0, 0.1, 0.2])
+
+
+def get_map(deg, sig_smooth, sig_size):
+    if sig_smooth == 0:
+        map = starry.Map(deg)
+        map.add_spot(amp=2.0, sigma=sig_size, lat=0.0, lon=0, relative=False)
+        map.amp = 20
+    else:
+        map = starry.Map(deg)
+        map.add_spot(amp=2.0, sigma=sig_size, lat=0.0, lon=0, relative=False)
+        map.amp = 20
+        S = get_smoothing_filter(deg, sig_smooth)
+        x = map.amp * map.y
+        x_smooth = (S @ x[:, None]).reshape(-1)
+        map[:, :] = x_smooth / x_smooth[0]
+        map.amp = x_smooth[0]
+
+    return map
+
+
+for j in range(len(smoothing_sigmas)):
+    for l in range(len(degs)):
+        map = get_map(degs[l], smoothing_sigmas[j], spot_sigma)
+        I = (map.intensity(lon=lon) - baseline) / (
+            map.intensity(lon=0) - baseline
+        )
+        Is[j, l] = I
+
+
+fig, ax = plt.subplots(1, 3, figsize=(10, 3), sharey=True)
+fig.subplots_adjust(wspace=0.15, hspace=0.1)
[…]

### 2022-11-16 src/scripts/chapter_7/compute_fluxes_from_ylm.py

@@ -1,0 +1,231 @@
+import paths
+import numpy as np
+from matplotlib import pyplot as plt
+from matplotlib import colors
+
+import starry
+import astropy.units as u
+import pickle as pkl
+import yaml
+
+import os
+import sys
+
+from utils import (
+    load_filter,
+    planck,
+    starry_intensity_to_bbtemp,
+    integrate_planck_over_filter,
+    inverse_integrate_planck_over_filter,
+)
+
+
+np.random.seed(42)
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def compute_simulated_flux(
+    t, map_star, map_planet, params_s, params_p, filt, wavelength_grid, texp,
+):
+    # Interpolate filter throughput
+    thr_interp = np.interp(wavelength_grid, filt[0], filt[1])
+
+    # Ratio of star and planet map *ampliudes* needs to be proportional to
+    # (Rp/Rs)**2 so we need to multiply the planet map amplitude with that factor
+    radius_ratio = params_p["r"] * u.Rjupiter.to(u.Rsun) / params_s["r"]
+    map_planet.amp *= radius_ratio ** 2
+
+    # Initialize system
+    star = starry.Primary(map_star, r=params_s["r"] * u.Rsun, m=params_s["m"] * u.Msun)
+
+    planet = starry.Secondary(
+        map_planet,
+        r=params_p["r"] * (u.Rjupiter.to(u.Rsun)) * u.Rsun,
+        porb=params_p["porb"] * u.d,
+        prot=params_p["prot"] * u.d,
+        t0=0.5 * params_p["porb"] * u.d,
+        inc=params_p["inc"] * u.deg,
+        theta0=180,
+    )
+
+    sys = starry.System(star, planet, texp=(texp.to(u.d)).value, oversample=9, order=0)
+
+    # Compute flux
+    A = sys.design_matrix(t)
+    x = np.concatenate([map_star.amp * map_star._y, map_planet.amp * map_planet._y])
+    fsim_spectral = np.tensordot(A, x, axes=1)
+
+    wav_filt = filt[0]
+    throughput = filt[1]
+
+    # Interpolate filter throughput to map wavelength grid
+    throughput_interp = np.interp(wavelength_grid, wav_filt, throughput)
+
+    # Integrate flux over bandpass
+    fsim = np.trapz(
[…]

### 2022-11-16 src/scripts/chapter_7/eclipse_snr.py

@@ -1,0 +1,226 @@
+"""Scripts taken from https://github.com/rodluger/planetplanet/blob/master/planetplanet/detect/jwst.py"""
+import numpy as np
+import astropy.units as u
+
+from matplotlib import colors
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+from utils import (
+    load_filter,
+    planck,
+)
+
+
+def jwst_background(wl):
+    """
+    Calculates the JWST thermal background from (Glasse et al. 2015)
+
+    Args:
+        wl (ndarray): Wavelengths [:math:`\mu \mathrm{m}`]
+
+    Returns:
+        ndarray: Background flux [:math:`\mathrm{W/m}^2 / \mu \mathrm{m}`]
+    """
+
+    # Solid angle
+    omega_background = np.pi * (0.42 / 206265.0 * wl / 10.0) ** 2.0
+
+    # Number of background grey body components (see table 1 of Glasse et al.):
+    nback = 7
+    emiss = [4.2e-14, 4.3e-8, 3.35e-7, 9.7e-5, 1.72e-3, 1.48e-2, 1.31e-4]
+    tb = [5500.0, 270.0, 133.8, 71.0, 62.0, 51.7, 86.7]
+
+    # Set up a vector for computing the background flux emissivity:
+    Fback = np.zeros_like(wl)
+
+    # Sum over the background components (insert 1 steradian for omega):
+    for i in range(nback):
+        Fback += emiss[i] * planck(tb[i], wl).value / 1e06 * omega_background
+
+    return Fback
+
+
+def photon_rate(lam, flux, wl, throughput, atel=25.0):
+    """
+    Compute the photon count rate registered by the detector.
+
+    Args:
+        lam (ndarray): High-res wavelengths [:math:`\mu \mathrm{m}`]
+        flux (ndarray): Spectral flux density \
+           [:math:`\mathrm{W/m}^2 / \mu \mathrm{m}`]
+        wl (ndarray): Filter wavelengths [:math:`\mu \mathrm{m}`]
+        throughput (ndarray): Filter throughput.
+        atel (float): Aperture area [:math:`\mathrm{m}^2`]. 
+    
+    Returns:
[…]

### 2022-11-16 src/scripts/chapter_7/fit_mcmc.py

@@ -1,0 +1,303 @@
+import paths
+import numpy as np
+import yaml
+import starry
+import astropy.units as u
+import os
+import arviz as az
+
+from utils import (
+    load_filter, planck, 
+    integrate_planck_over_filter, 
+    inverse_integrate_planck_over_filter, 
+)
+
+import jax.numpy as jnp
+from jax import random
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+numpyro.set_host_device_count(4)
+
+import arviz as az
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def initialize_featureless_map(T_star, wavelength_grid, ydeg=1):
+    # Initialize star map
+    map_star = starry.Map(ydeg=1, nw=len(wavelength_grid))
+    Llam = (4 * np.pi) * np.pi * planck(T_star, wavelength_grid).value
+    map_star.amp = Llam / 4
+    return map_star
+
+
+def get_lower_order_map(map, ydeg=2):
+    assert map.ydeg > ydeg
+    x = map._y*map.amp
+    x = x[:(ydeg + 1)**2]
+    map = starry.Map(ydeg, nw=map.nw)
+    map[1:, :, :] = x[1:, :]/x[0]
+    map.amp = x[0]
+    
+    return map
+
+def draw_sample_lightcurve(t, fsim, sigma=None, snr=None):
+    eclipse_depth = np.max(fsim) - np.min(fsim)
+    sigma = eclipse_depth / snr
+    fobs = fsim + np.random.normal(0, sigma, size=len(t))
+    return fobs, sigma*np.ones_like(fobs)
+
+
+def inferred_intensity_to_bbtemp(I_planet_raw, filt, params_s, params_p):
+    """
+    Convert inferred starry intensity map to a BB temperature map.
+    """
+    # Star spectral radiance integrated over solid angle and bandpass
+    I_star = np.pi * integrate_planck_over_filter(params_s["Teff"], filt)
+
+    # Rescale the intensity of the planet map to physical units
+    I_planet = I_planet_raw * I_star * (params_s["r"] / (params_p["r"]*u.Rjupiter.to(u.Rsun))) ** 2
+
[…]

### 2022-11-16 src/scripts/chapter_7/flux_variation.py

@@ -1,0 +1,50 @@
+import paths
+import os
+import numpy as np
+from matplotlib import pyplot as plt
+from matplotlib.ticker import AutoMinorLocator
+
+
+input_dir = paths.data/"output/mapping_exo/T341_1bar_fluxes/hd209/"
+fsim = np.load(os.path.join(input_dir, "fsim_snapshots.npy"))
+fsim_ref = np.load(os.path.join(input_dir, "fsim_reference_snapshots.npy"))
+t_eclipse = np.load(os.path.join(input_dir, "t_eclipse.npy"))
+
+# Normalize
+norm = np.max(fsim_ref, axis=1)
+fsim_ref /= norm[:, None]
+fsim /= norm[:, None]
+
+# Rescale 
+fsim = (fsim - 1)*1e06
+fsim_ref = (fsim_ref - 1)*1e06
+
+fig, ax = plt.subplots(1, 2, figsize=(14, 5), gridspec_kw={"wspace":0.4})
+
+# Plot eclipses
+fsim_median = np.median(fsim, axis=0)
+colors = plt.cm.viridis(np.linspace(0, 1, len(fsim)))
+
+for fsim_, fsim_ref_ in zip(fsim, fsim_ref):
+    ax[0].plot(t_eclipse/3.5, fsim_ - fsim_median, color="k", alpha=0.3, lw=0.8)
+for fsim_, fsim_ref_ in zip(fsim, fsim_ref):
+    ax[1].plot(t_eclipse/3.5, fsim_ - fsim_ref_, color="k", alpha=0.3, lw=0.8)
+
+for a in ax:
+    a.set_xlabel("Orbital phase")
+    a.grid(alpha=0.5)
+    a.xaxis.set_minor_locator(AutoMinorLocator())
+    a.yaxis.set_minor_locator(AutoMinorLocator())
+#    a.set(xlim=(-0.25, 0.25))
+    a.set(xlim=(-0.06, 0.06))
+
+ax[0].set(ylim=(-70, 70))
+ax[0].set(ylabel=r"$F - F_\mathrm{median}$ [ppm]")
+ax[1].set(ylabel=r"$F_{l=20} - F_{l=2}$ [ppm]")
+ax[0].set_title("Secondary eclipse variability between orbits")
+ax[1].set_title("Deviation from baseline map (an $l=2$ map)")
+ax[0].set_yticks(np.arange(-50, 75, 25))
[…]

### 2022-11-16 src/scripts/chapter_7/models.py

@@ -1,0 +1,149 @@
+import numpy as np
+import starry
+
+import jax.numpy as jnp
+from jax import random
+import numpyro
+import numpyro.distributions as dist
+from numpyro.infer import *
+import arviz as az
+
+import os
+
+numpyro.enable_x64(True)
+numpyro.set_host_device_count(4)
+
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+HOMEPATH = os.path.abspath(os.path.split(__file__)[0])
+
+
+def fit_model_num(
+    A,
+    fobs,
+    ferr,
+    ydeg_inf=5,
+    tune=500,
+    draws=500,
+    chains=2,
+    sigmoid_constraint=True,
+    sigmoid_steepness=1e6,
+    sd=1e-02,
+):
+    ncoeff = (ydeg_inf + 1) ** 2
+
+    # First we solve the least squares problem to get an estimate of stellar flux fs
+    # This parameter is problematic for the sampler for some reason so we fit for a
+    # small deviation from the least squares estimate
+    ncoeff_prim = 4  # stellar map is l=1
+    A_full = A[:, : ncoeff_prim + (ydeg_inf + 1) ** 2]
+    A_sec = A_full[:, ncoeff_prim:]
+
+    # Primary
+    L_prim = np.ones(ncoeff_prim)
+    L_prim[1:] = 1e-10 ** 2
+    L_sec = 1e-02 ** 2 * np.ones((ydeg_inf + 1) ** 2)
+    L_sec[1:] = 1e-03 ** 2
+    L = np.concatenate([L_prim, L_sec])
+
+    x_lsq, _ = starry.linalg.solve(A_full, fobs, C=ferr ** 2, L=L,)
+    fs_lsq = x_lsq[0]
+
+    # Transform the Ylm coefficients to simplify the structure of the covariance matrix
+    # Polar transform
+    R = starry.Map(ydeg_inf).ops.dotR(
+        np.eye(ncoeff),
+        np.array(1.0),
+        np.array(0.0),
+        np.array(0.0),
+        np.array(-np.pi / 2),
+    )
+
+    # Transform to group Ylms by order
+    idx = [[] for m in range(2 * ydeg_inf + 1)]
+    for m in range(ydeg_inf + 1):
+        l = np.arange(abs(m), ydeg_inf + 1)
+        idx[2 * m] = l ** 2 + l + m
+        if m > 0:
+            idx[2 * m - 1] = l ** 2 + l - m
[…]

### 2022-11-16 src/scripts/chapter_7/old_hydro_sim_snapshots_fits.py

@@ -1,0 +1,482 @@
+import numpy as np
+import yaml
+import starry
+import astropy.units as u
+
+from matplotlib import colors
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+
+from utils import (
+    load_filter,
+    planck,
+    starry_intensity_to_bbtemp,
+    simulation_snapshot_to_ylm,
+    integrate_planck_over_filter,
+    inverse_integrate_planck_over_filter,
+    fit_model_num,
+)
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def simulation_snapshot_to_ylm(path, wavelength_grid, ydeg=20, temp_offset=-450):
+    data = np.loadtxt(path)
+
+    nlat = 512
+    nlon = 1024
+
+    lons = np.linspace(-180, 180, nlon)
+    lats = np.linspace(-90, 90, nlat)
+
+    lon_grid, grid = np.meshgrid(lons, lats)
+
+    temp_grid = np.zeros_like(lon_grid)
+
+    for i in range(nlat):
+        for j in range(nlon):
+            temp_grid[i, j] = data.reshape((nlat, nlon))[i, j]
+
+    temp_grid = np.roll(temp_grid, int(temp_grid.shape[1] / 2), axis=1) + temp_offset
+
+    x_list = []
+    map_tmp = starry.Map(ydeg)
+
+    # Evaluate at fewer points for performance reasons
+    idcs = np.linspace(0, len(wavelength_grid) - 1, 10).astype(int)
+    for i in idcs:
+        I_grid = np.pi * planck(temp_grid, wavelength_grid[i])
+        map_tmp.load(I_grid, force_psd=True)
+        x_list.append(map_tmp._y * map_tmp.amp)
+
+    # Interpolate to full grid
+    x_ = np.vstack(x_list).T
+    x_interp_list = [
+        np.interp(wavelength_grid, wavelength_grid[idcs], x_[i, :])
+        for i in range(x_.shape[0])
+    ]
+    x = np.vstack(x_interp_list)
+
+    return x
+
+
+def compute_simulated_lightcurve(
+    t, map_star, map_planet, params_s, params_p, filt, wavelength_grid, texp,
+):
+    # Interpolate filter throughput
+    thr_interp = np.interp(wavelength_grid, filt[0], filt[1])
+
+    # Ratio of star and planet map *ampliudes* needs to be proportional to
[…]

### 2022-11-16 src/scripts/chapter_7/preimages.py

@@ -1,0 +1,443 @@
+import paths
+import numpy as np
+import matplotlib
+from matplotlib import pyplot as plt
+from matplotlib import colors
+
+import starry
+import astropy.units as u
+
+from utils import add_band
+
+
+np.random.seed(42)
+starry.config.lazy = False
+
+
+def get_preimage(map_planet, b, obl=0.0, include_phase_curves=False):
+    map = starry.Map(map_planet.ydeg)
+    map[1:, :] = map_planet[1:, :]
+    map.amp = map_planet.amp
+
+    # Initialize starry system
+    Porb = 1 * u.d
+    Mstar = 1 * u.Msun
+    Rstar = 1 * u.Rsun
+    a = (Porb.to(u.yr).value ** 2 * Mstar.value) ** (1 / 3.0) * u.au
+    i = np.arccos(b * Rstar / a.to(u.Rsun))
+    radius_ratio = 0.1
+
+    map_star = starry.Map(ydeg=0)
+    map_star.amp = 1
+    star = starry.Primary(map_star, r=Rstar,)
+
+    planet = starry.Secondary(
+        map,
+        r=radius_ratio * Rstar,
+        porb=Porb,
+        prot=Porb,
+        inc=i.to(u.deg),
+        t0=Porb / 2,
+        theta0=180.0,
+    )
+
+    # Tidally locked planet -> equatorial plane same as orbital plane
+    planet.map.inc = planet.inc
+    planet.map.obl = obl
+
+    sys = starry.System(star, planet)
+
+    # Generate high cadence light excluding transit
+    t0 = 0.0 * u.d
+    texp = 5 * u.s
+    delta_t = 0.5 * u.d
+    npts = int((2 * delta_t.to(u.s)) / (texp))  # total number of data points
+    t = np.linspace(-delta_t.value, +delta_t.value, npts)
+
+    # Masks for eclipse, transit and phase curves
+    mask_ecl = np.logical_and(t < 0.07, t > -0.07)
+
+    if not include_phase_curves:
+        t = t[mask_ecl]
+
+    A = sys.design_matrix(t)
+    x_com = np.concatenate([map_star._y * map_star.amp, map._y * map.amp])
+    flux = (A @ x_com[:, None]).reshape(-1)
+
[…]

### 2022-11-16 src/scripts/chapter_7/process_mcmc_samples.py

@@ -1,0 +1,120 @@
+import paths
+import numpy as np
+import yaml
+import starry
+import astropy.units as u
+import arviz as az
+
+from utils import (
+    load_filter, planck, 
+    integrate_planck_over_filter, 
+    inverse_integrate_planck_over_filter, 
+    starry_intensity_to_bbtemp,
+)
+
+from jax.config import config
+config.update("jax_enable_x64", True)
+config.update('jax_platform_name', 'cpu')
+
+import arviz as az
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+def initialize_map(ydeg, nw, x):
+    map = starry.Map(ydeg, nw=nw)
+    map[1:, :, :] = x[1:, :] / x[0]
+    map.amp = x[0]
+    return map
+
+def inferred_intensity_to_bbtemp(I_planet_raw, filt, params_s, params_p):
+    """
+    Convert inferred starry intensity map to a BB temperature map.
+    """
+    # Star spectral radiance integrated over solid angle and bandpass
+    I_star = np.pi * integrate_planck_over_filter(params_s["Teff"], filt)
+
+    # Rescale the intensity of the planet map to physical units
+    I_planet = I_planet_raw * I_star * (params_s["r"] / (params_p["r"]*u.Rjupiter.to(u.Rsun))) ** 2
+
+    # Plot temperature map of the planet
+    bbtemp_map_inf = np.copy(I_planet[:, :].value)
+
+    for i in range(I_planet.shape[0]):
+        for j in range(I_planet.shape[1]):
+            bbtemp_map_inf[i, j] = inverse_integrate_planck_over_filter(
+                I_planet[i, j].value, filt
+            )
+    return bbtemp_map_inf
+
+# System parameters
+planet = "hd209"
+filter_name = "f444w"
+
+# Load orbital and system parameters
+with open(f"../../data/mapping_exo/system_parameters/{planet}/orbital_params_planet.yaml", "rb") as handle:
+    params_p = yaml.safe_load(handle)
+with open(f"../../data/mapping_exo/system_parameters/{planet}/orbital_params_star.yaml", "rb") as handle:
+    params_s = yaml.safe_load(handle)
+    
+# Load filter
+filt = load_filter("../../data/mapping_exo/filter_files", name=f"{filter_name}")
+mask = filt[1] > 0.002
+
+# Wavelength grid for starry map (should match filter range)
+wavelength_grid = np.linspace(4.5 - 1.2, 4.5 + 1.2, 50)
+
[…]

### 2022-11-16 src/scripts/chapter_7/simulated_vs_inferred.py

@@ -1,0 +1,270 @@
+import paths
+import yaml
+import numpy as np
+import starry
+import matplotlib.pyplot as plt
+from matplotlib import colors
+import matplotlib as mpl
+import astropy.units as u
+import arviz as az
+
+from utils import (
+    starry_intensity_to_bbtemp,
+    load_filter,
+    integrate_planck_over_filter,
+    inverse_integrate_planck_over_filter
+)
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+def initialize_map(ydeg, nw, x):
+    map = starry.Map(ydeg, nw=nw)
+    map[1:, :, :] = x[1:, :] / x[0]
+    map.amp = x[0]
+    return map
+
+def inferred_intensity_to_bbtemp(I_planet_raw, filt, params_s, params_p):
+    """
+    Convert inferred starry intensity map to a BB temperature map.
+    """
+    # Star spectral radiance integrated over solid angle and bandpass
+    I_star = np.pi * integrate_planck_over_filter(params_s["Teff"], filt)
+
+    # Rescale the intensity of the planet map to physical units
+    I_planet = I_planet_raw * I_star * (params_s["r"] / (params_p["r"]*u.Rjupiter.to(u.Rsun))) ** 2
+
+    # Plot temperature map of the planet
+    bbtemp_map_inf = np.copy(I_planet[:, :].value)
+
+    for i in range(I_planet.shape[0]):
+        for j in range(I_planet.shape[1]):
+            bbtemp_map_inf[i, j] = inverse_integrate_planck_over_filter(
+                I_planet[i, j].value, filt
+            )
+    return bbtemp_map_inf
+
+
+def get_design_matrix(t, ydeg, params_s, params_p, texp):
+    map_planet = starry.Map(ydeg)
+    map_star = starry.Map(1)
+
+    # Initialize system
+    star = starry.Primary(map_star, r=params_s["r"] * u.Rsun, m=params_s["m"] * u.Msun)
+    planet = starry.Secondary(
+        map_planet,
+        r=params_p["r"] * (u.Rjupiter.to(u.Rsun)) * u.Rsun,
+        porb=params_p["porb"] * u.d,
[…]

### 2022-11-16 src/scripts/chapter_7/single_fit.py

@@ -1,0 +1,172 @@
+import paths
+import numpy as np
+import starry
+import matplotlib.pyplot as plt
+from matplotlib import colors
+import matplotlib as mpl
+from matplotlib.ticker import AutoMinorLocator
+import arviz as az
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+# Load data
+fsim = np.load(paths.data/"output/mapping_exo/T341_1bar_fluxes/hd209/fsim_snapshots.npy")[0]
+fsim_ref = np.load(paths.data/"output/mapping_exo/T341_1bar_fluxes/hd209/fsim_reference_snapshots.npy")[0]
+fobs1, ferr1 = np.load(
+   paths.data/"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_15/lightcurve.npy"
+)
+fobs2, ferr2 = np.load(
+   paths.data/"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_50/lightcurve.npy"
+)
+
+t_eclipse = np.load(paths.data/"output/mapping_exo/T341_1bar_fluxes/hd209/t_eclipse.npy")
+
+# Normalize
+norm = np.max(fsim_ref)
+fsim_ref /= norm
+fsim /= norm
+
+l = 4
+samples1 = az.from_netcdf(
+   paths.data/f"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_15/l_{l}/samples.nc"
+)
+samples2 = az.from_netcdf(
+   paths.data/f"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_50/l_{l}/samples.nc"
+)
+map_sim_temp = np.load(
+    paths.data/"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/sim_map_temp.npy",
+)
+dat = np.load(
+   paths.data/f"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_15/l_{l}/map_inf_temp.npz"
+) 
+map_inf_temp_mean1 = dat["mean"]
+map_inf_temp_samples1 = dat["samples"]
+
+dat = np.load(
+   paths.data/f"output/mapping_exo/T341_1bar_mcmc/hd209/time_0/snr_50/l_{l}/map_inf_temp.npz"
+) 
+map_inf_temp_mean2 = dat["mean"]
[…]

### 2022-11-16 src/scripts/chapter_7/snapshots_multiple_orbits.py

@@ -1,0 +1,62 @@
+import paths
+import numpy as np
+import starry
+import matplotlib.pyplot as plt
+from matplotlib import colors
+import matplotlib as mpl
+
+from utils import starry_intensity_to_bbtemp
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+def initialize_map(ydeg, nw, x):
+    map = starry.Map(ydeg, nw=nw)
+    map[1:, :, :] = x[1:, :] / x[0]
+    map.amp = x[0]
+    return map
+
+
+# Wavelength grid for starry map (should match filter range)
+wavelength_grid = np.linspace(4.5 - 1.2, 4.5 + 1.2, 50)
+
+xs_sim = np.load(paths.data/"output/mapping_exo/T341_1bar_ylms/coefficients.npy")
+
+maps_sim_temp = []
+for i in range(30):
+    x_sim = xs_sim[i]
+    map = initialize_map(20, len(wavelength_grid), x_sim)
+    map_sim_temp = starry_intensity_to_bbtemp(
+        map.render(res=200, projection="Orthogonal"), wavelength_grid
+    ) 
+    maps_sim_temp.append(map_sim_temp)
+
+fig, ax = plt.subplots(3, 10, figsize=(12, 4),  gridspec_kw={'hspace':0.4})
+ax = ax.reshape(-1)
+
+vmin, vmax= 1550, 1750
+for i in range(len(maps_sim_temp)):
+    map.show(
+        image=maps_sim_temp[i], projection="orthogonal", cmap="OrRd", 
+        norm=colors.Normalize(vmin=vmin, vmax=vmax),
+        ax=ax[i],
+        zorder=-1,
+    )
+    ax[i].set_aspect(1)
+    ax[i].set_title(f"$t/P = {i}$")
+
+# Colorbar
+cax = fig.add_axes([0.36, 0.02, 0.3, 0.03])
+fig.colorbar(
+mpl.cm.ScalarMappable(norm=colors.Normalize(vmin=vmin, vmax=vmax), cmap="OrRd"),
+    cax=cax,
+        orientation='horizontal',
+        label="blackbody temperature [K]", fraction=1.
+)
+
+for a in ax:
+    a.set_rasterization_zorder(0)
+
+fig.savefig(paths.figures/"snapshots_multiple_orbits.pdf", bbox_inches="tight")

### 2022-11-16 src/scripts/chapter_7/snapshots_within_orbit.py

@@ -1,0 +1,138 @@
+import paths
+import numpy as np
+import yaml
+import starry
+
+from matplotlib import colors
+import matplotlib as mpl
+import matplotlib.pyplot as plt
+
+from utils import (
+    load_filter,
+    planck,
+    starry_intensity_to_bbtemp,
+)
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def initialize_featureless_map(T, wavelength_grid, ydeg=1):
+    # Initialize star map
+    map = starry.Map(ydeg=1, nw=len(wavelength_grid))
+    Llam = (4 * np.pi) * np.pi * planck(T, wavelength_grid).value
+    map.amp = Llam / 4
+    return map
+
+
+def get_lower_order_map(map, ydeg=2):
+    assert map.ydeg > ydeg
+    x = map._y * map.amp
+    x = x[: (ydeg + 1) ** 2]
+    map = starry.Map(ydeg, nw=map.nw)
+    map[1:, :, :] = x[1:, :] / x[0]
+    map.amp = x[0]
+
+    return map
+
+
+def initialize_map(ydeg, nw, x):
+    map = starry.Map(ydeg, nw=nw)
+    map[1:, :, :] = x[1:, :] / x[0]
+    map.amp = x[0]
+    return map
+
+
+# System parameters
+planet = "hd209"
+filter_name = "f444w"
+
+# Load orbital and system parameters
+with open(
+    paths.data/f"mapping_exo/system_parameters/{planet}/orbital_params_planet.yaml", "rb"
+) as handle:
+    params_p = yaml.safe_load(handle)
+with open(
+    paths.data/f"mapping_exo/system_parameters/{planet}/orbital_params_star.yaml", "rb"
+) as handle:
+    params_s = yaml.safe_load(handle)
+
+# Load filter
+filt = load_filter(paths.data/"mapping_exo/filter_files", name=f"{filter_name}")
+mask = filt[1] > 0.002
+
+# Wavelength grid for starry map (should match filter range)
+wavelength_grid = np.linspace(4.5 - 1.2, 4.5 + 1.2, 50)
+
+xs = np.load(paths.data/"output/mapping_exo/T341_1bar_ylms/coefficients_single_orbit.npy")
[…]

### 2022-11-16 src/scripts/chapter_7/spot_residuals.py

@@ -1,0 +1,286 @@
+import paths
+import numpy as np
+import starry
+import astropy.units as u
+from scipy.optimize import brent
+from matplotlib import colors
+import matplotlib as mpl
+from matplotlib import pyplot as plt
+from mpl_toolkits.axes_grid1 import make_axes_locatable
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def compute_simulated_lightcurve(
+    t, map_star, map_planet, params_s, params_p, texp, radius_ratio=0.1,
+):
+
+    # Ratio of star and planet map *amplitudes* needs to be proportional to
+    # (Rp/Rs)**2 so we need to multiply the planet map amplitude with that factor
+    map_planet.amp *= (radius_ratio) ** 2
+
+    # Initialize system
+    star = starry.Primary(map_star, r=params_s["r"])
+
+    planet = starry.Secondary(
+        map_planet,
+        r=radius_ratio * params_s["r"],
+        porb=params_p["porb"],
+        prot=params_p["porb"],
+        t0=t0,
+        inc=params_p["inc"],
+        theta0=180,
+    )
+
+    sys = starry.System(star, planet, texp=(texp.to(u.d)).value, oversample=9, order=0)
+
+    # Compute flux
+    A = sys.design_matrix(t)
+    x = np.concatenate([map_star.amp * map_star._y, map_planet.amp * map_planet._y])
+    fsim = (A @ x[:, None]).reshape(-1)
+
+    # Rescale the amplitude of the planet map back to its original value
+    map_planet.amp *= 1 / (radius_ratio) ** 2
+
+    return fsim, sys
+
+
+def get_map_amplitude_given_constraint(
+    radius_ratio=0.1,
+    target_Fp_Fs=2e-03,
+    ydeg=20,
+    spot_contrast=-0.25,
+    radius=30.0,
+    lat=30.0,
+    lon=30.0,
+):
+    map_star = starry.Map(1)
+    Fs = map_star.flux()[0]
+
+    def cost_fn(map_amplitude):
+        map = starry.Map(ydeg)
+        map.amp = map_amplitude
+
+        # Add spot
+        map.spot(
[…]

### 2022-11-16 src/scripts/chapter_7/temperature_snapshots_to_ylms.py

@@ -1,0 +1,97 @@
+import paths
+import numpy as np
+import starry
+
+import os
+from utils import planck
+
+np.random.seed(42)
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def simulation_snapshot_to_ylm(path, wavelength_grid, ydeg=20):
+    data = np.loadtxt(path)
+
+    nlat = 512
+    nlon = 1024
+
+    temp_grid = data.reshape(nlat, nlon)
+    temp_grid = np.roll(temp_grid, int(temp_grid.shape[1] / 2), axis=1)
+
+    x_list = []
+    map_tmp = starry.Map(ydeg)
+
+    # Evaluate at fewer points in wavelength for performance reasons
+    idcs = np.linspace(0, len(wavelength_grid) - 1, 10).astype(int)
+    for i in idcs:
+        I_grid = np.pi * planck(temp_grid, wavelength_grid[i])
+        map_tmp.load(I_grid, force_psd=False)
+        x_list.append(map_tmp._y * map_tmp.amp)
+
+    # Interpolate to full grid
+    x_ = np.vstack(x_list).T
+    x_interp_list = [
+        np.interp(wavelength_grid, wavelength_grid[idcs], x_[i, :])
+        for i in range(x_.shape[0])
+    ]
+    x = np.vstack(x_interp_list)
+
+    return x
+
+
+input_dir = paths.data/"mapping_exo/hydro_snapshots_raw/T341_1bar/"
+output_dir = paths.data/"output/mapping_exo/T341_1bar_ylms/"
+fnames = os.listdir(input_dir)
+
+# Sort according to time
+vals = np.array([float(fname.split("_")[3]) for fname in fnames])
+sorted_idcs = np.argsort(vals)
+fnames = np.array(fnames)[sorted_idcs]
+vals = vals[sorted_idcs]
+
+
+# Wavelength grid for starry map (should match filter range)
+wavelength_grid = np.linspace(4.5 - 1.2, 4.5 + 1.2, 50)
+ydeg = 20
+
+
+
+# Iterate over files in order and only select snapshots at the same orbital
+# phase 
+xs = []
+times = []
+for fname in fnames:
+    time =  float(fname.split("_")[3]) 
+    if (time % 1 == 0.) and (len(times) <= 30):
+        print(fname)
+        times.append(time)
+        # Load simulation snapshots as starry maps
+        x = simulation_snapshot_to_ylm(
[…]

### 2022-11-16 src/scripts/chapter_7/utils.py

@@ -1,0 +1,284 @@
+import paths
+import numpy as np
+import os
+
+import astropy.units as u
+from astropy import constants as const
+
+import pickle as pkl
+from scipy.optimize import brent
+from scipy.special import legendre as P
+from numba import jit
+
+import starry
+
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+
+def get_smoothing_filter(ydeg, sigma=0.1):
+    """
+    Returns a smoothing matrix which applies an isotropic Gaussian beam filter
+    to a spherical harmonic coefficient vector. This helps suppress ringing
+    artefacts around spot like features. The standard deviation of the Gaussian
+    filter controls the strength of the smoothing. Features on angular scales
+    smaller than ~ 1/sigma are strongly suppressed.
+
+    Args:
+        ydeg (int): Degree of the map.
+        sigma (float, optional): Standard deviation of the Gaussian filter.
+            Defaults to 0.1.
+
+    Returns:
+        ndarray: Diagonal matrix of shape (ncoeff, ncoeff) where ncoeff = (l + 1)^2.
+    """
+    l = np.concatenate([np.repeat(l, 2 * l + 1) for l in range(ydeg + 1)])
+    s = np.exp(-0.5 * l * (l + 1) * sigma ** 2)
+    S = np.diag(s)
+    return S
+
+
+def load_filter(path, name="f444w"):
+    if name == "f356w":
+        path = os.path.join(path, "F356W_ModAB_mean.csv")
+        dat = np.loadtxt(path, skiprows=1, delimiter=",")
+        return dat.T
+    elif name == "f322w2":
+        path = os.path.join(path, "F322W2_filteronly_ModAB_mean.txt")
+        dat = np.loadtxt(path)
+        return dat.T
+    elif name == "f444w":
+        path = os.path.join(path, "F444W_ModAB_mean.csv")
+        dat = np.loadtxt(path, skiprows=1, delimiter=",")
+        return dat.T
+    else:
+        raise ValueError("Filter name not recognized.")
+
+
+def planck(T, lam):
+    """
+    Planck function.
+
+    Args:
+        T (float): Blackbody temperature in Kelvin. 
+        lam (float or ndarray): Wavelength in um (microns).
+
+    Returns:
[…]

### 2022-11-16 src/scripts/chapter_7/varying_contrast_spot.py

@@ -1,0 +1,461 @@
+import paths
+import numpy as np
+import starry
+import astropy.units as u
+from scipy.optimize import brent
+from scipy.stats import binned_statistic
+from matplotlib import colors
+from matplotlib import pyplot as plt
+from exo_mapping_utils import fit_model_num
+
+np.random.seed(42)
+
+starry.config.lazy = False
+starry.config.quiet = True
+
+
+def compute_simulated_lightcurve(
+    map_star, map_planet, params_s, params_p, texp, radius_ratio=0.1,
+):
+
+    # Initialize star map
+    map_star = starry.Map(ydeg=1)
+
+    # Ratio of star and planet map *amplitudes* needs to be proportional to
+    # (Rp/Rs)**2 so we need to multiply the planet map amplitude with that factor
+    map_planet.amp *= (radius_ratio) ** 2
+
+    # Generate observation times excluding transit
+    porb = params_p["porb"]
+    t0 = 0.5 * params_p["porb"]
+
+    t_ = np.linspace(-t0.value, +t0.value, int(0.5 * porb.to(u.s) / texp))
+
+    # Select only times around eclipse
+    mask_ecl = np.abs(t_) < 0.45
+    t = t_[mask_ecl]
+
+    # Initialize system
+    star = starry.Primary(map_star, r=params_s["r"])
+
+    planet = starry.Secondary(
+        map_planet,
+        r=radius_ratio * params_s["r"],
+        porb=params_p["porb"],
+        prot=params_p["porb"],
+        t0=t0,
+        inc=params_p["inc"],
+        theta0=180,
+    )
+    sys = starry.System(star, planet, texp=(texp.to(u.d)).value, oversample=9, order=0)
+
+    # Compute flux
+    A = sys.design_matrix(t)
+    x = np.concatenate([map_star.amp * map_star._y, map_planet.amp * map_planet._y])
+    fsim = (A @ x[:, None]).reshape(-1)
+
+    # Rescale the amplitude of the planet map back to its original value
+    map_planet.amp *= 1 / (radius_ratio) ** 2
+
+    return t, fsim, sys
+
+
+def get_map_amplitude_given_constraint(
+    radius_ratio=0.08,
+    target_Fp_Fs=2e-03,
+    ydeg=25,
+    spot_contrast=-0.25,
+    radius=30.0,
+    lat=30,
+    lon=30.0,
+):
+    map_star = starry.Map(1)
[…]

### 2023-02-25 src/scripts/chapter_1/ads_keyword_bayesian.py
Keith's corrections

@@ -1,18 +1,0 @@
-import paths
-import numpy as np
-from matplotlib import pyplot as plt
-
-
-data = np.loadtxt(paths.data/"misc/ads_keyword_bayes.csv", delimiter=",")
-year, count, count_ref = data.T
-
-fig, ax = plt.subplots(figsize=(9, 5))
-
-ax.bar(year, count, color="white", edgecolor="k", linewidth=1.5)
-ax.set_xlim(1980, 2021)
-ax.set_ylabel("Number of entries for the keyword\n 'Bayesian' on NASA ADS")
-ax.set_xlabel("year")
-
-# Save as pdf
-fig.savefig(paths.figures/"ads_keyword_bayesian.pdf", bbox_inches="tight")
-

### 2023-02-25 src/scripts/chapter_1/paths.py

@@ -1,25 +1,0 @@
-"""
-Exposes common paths useful for manipulating datasets and generating figures.
-"""
-from pathlib import Path
-
-# Absolute path to the top level of the repository
-root = Path(__file__).resolve().parents[3].absolute()
-
-# Absolute path to the `src` folder
-src = root / "src"
-
-# Absolute path to the `src/data` folder (contains datasets)
-data = src / "data"
-
-# Absolute path to the `src/static` folder (contains static images)
-static = src / "static"
-
-# Absolute path to the `src/scripts` folder (contains figure/pipeline scripts)
-scripts = src / "scripts"
-
-# Absolute path to the `src/tex` folder (contains the manuscript)
-tex = src / "tex"
-
-# Absolute path to the `src/tex/figures` folder (contains figure output)
-figures = tex / "figures"

### 2023-02-28 src/scripts/chapter_3/mag_limbdark_accuracy.py
final updateds

@@ -53,7 +53,7 @@ s, q = 0.9, 0.2
 u1 = 0.7
 
 # 1000  points on caustic curve
-npts = 25 
+npts = 20
 critical_curves, caustic_curves = critical_and_caustic_curves(
     npts=npts, nlenses=2, s=s, q=q
 )
@@ -61,7 +61,8 @@ caustic_curves = caustic_curves.reshape(-1)
 
 # Accuracy of integration
 acc_vbb = 1e-05
-npts_limb = 400
+npts_limb = 500
+npts_ld = 150
 
 
 # Compute relative error w.r.t. VBB for different values of rho 
@@ -74,7 +75,7 @@ for rho in rho_list:
     print(f"rho = {rho}")
 
     # Generate 1000 random test points within 2 source radii away from the caustic points 
-    key = random.PRNGKey(3)
+    key = random.PRNGKey(5)
     key, subkey1, subkey2 = random.split(key, num=3)
     phi = random.uniform(subkey1, caustic_curves.shape, minval=-np.pi, maxval=np.pi)
     r = random.uniform(subkey2, caustic_curves.shape, minval=0., maxval=2*rho)
@@ -86,7 +87,7 @@ for rho in rho_list:
             for w in w_test
         ]
     )
-    mags = mag_binary(w_test, rho, s, q, u1=u1, npts_limb=npts_limb, npts_ld=npts_limb, limb_darkening=True)
+    mags = mag_binary(w_test, rho, s, q, u1=u1, npts_limb=npts_limb, npts_ld=npts_ld, limb_darkening=True)
     mags_vbb_list.append(mags_vbb)
     mags_list.append(mags)
 

### 2023-02-28 src/scripts/chapter_5/compute_likelihood_on_grid_backup.py

@@ -1,218 +1,0 @@
-import paths
-import numpy as np
-
-import os
-from astropy.table import Table
-
-from jax.config import config
-config.update("jax_enable_x64", True)
-config.update('jax_platform_name', 'cpu')
-import jax.numpy as jnp
-from jax import jit, vmap, lax
-
-from caustics import  mag
-from caustics.linalg import weighted_least_squares
-
-
-import VBBinaryLensing
-VBBL = VBBinaryLensing.VBBinaryLensing()
-
-def mag_vbb_binary(w0, rho, s, q, u1=0.0, accuracy=1e-05):
-    return VBBL.BinaryMag2(s, q, w0.real, w0.imag, rho)
-
-def ll_fn_vbb(x, pointwise=False):
-    ln_t0, ln_tE, ln_u0, ln_s, ln_q, alpha, ln_rho = x
-    s, q, t0, u0, tE, rho = np.exp(ln_s), np.exp(ln_q), np.exp(ln_t0),np.exp(ln_u0), np.exp(ln_tE), np.exp(ln_rho)
-
-    w_x = (t - t0)/tE*np.cos(alpha) - u0*np.sin(alpha)
-    w_y = (t - t0)/tE*np.sin(alpha) + u0*np.cos(alpha)
-    w_points = w_x + 1j*w_y
-
-    A = np.array([mag_vbb_binary(w0, rho, s, q) for w0 in w_points])
-
-    M = np.stack([A - 1., np.ones_like(A)]).T
-    beta = weighted_least_squares(fobs, ferr, M)
-    fpred = (M @ beta).reshape(-1)
-
-    if pointwise:
-        return -0.5 * (fobs - fpred)**2 / ferr**2
-    else:
-        return -0.5 * np.sum((fobs - fpred)**2 / ferr**2)
-
-
-import numpyro
-import numpyro.distributions as dist
-numpyro.set_host_device_count(4)
-
-
-def magnitudes_to_fluxes(mag, mag_err, zeropoint=22):
-    flux = 10. ** (0.4 * (zeropoint - mag))
-    err_flux = mag_err * flux * np.log(10.) * 0.4
-    return flux, err_flux
-
-def get_data(path):
[…]
